### 2017-09-06 README.Rmd
added travis etc :)

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# dynanalysis

The goal of dynanalysis is to ...

### Citation

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Title of compendium_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### Installation

You can install dynanalysis from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Zouter/dynanalysis")
```

### Licenses

**Text and figures:**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code:** See the [DESCRIPTION](DESCRIPTION) file

$\textbf{Data:}$ [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2017-09-06 analysis/paper/paper.Rmd
added travis etc :)

---
title: "Title Goes Here"
author:
  - author 1
  - author 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: journal-of-archaeological-science.csl
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(dynanalysis)
```

# Introduction

Here is a citation [@Marwick2017]

# Background

```{r demo, eval = FALSE}
1 + 1
```

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

# References

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```

### 2017-09-06 README.Rmd
changed name to dynalysis

@@
 
-# dynanalysis
+# dynalysis
 
-The goal of dynanalysis is to ...
+The goal of dynalysis is to ...
 
@@
 
-You can install dynanalysis from github with:
+You can install dynalysis from github with:
 
@@
 # install.packages("devtools")
-devtools::install_github("Zouter/dynanalysis")
+devtools::install_github("Zouter/dynalysis")
 ```

### 2017-09-06 analysis/1-methods-toys.Rmd
added 1-methods-toys

---
title: "1-methods-toys"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Testing methods on simple toy datasets

```{r}
library(tidyverse)
library(dyneval)

set.seed(1)

tasks <- generate_toy_datasets()
tasks <- tasks[1, ]
```

We choose certain parameters for which we know that the method will find (or focus on finding) a linear trajectory, and will do this quickly
```{r}
method_descriptions <- list(
  waterfall=list(),
  scorpius=list(),
  slingshot=list(),
  gpfates=list(nfates=1),
  stemid=list(clustnr=10, bootnr=10, pdishuf=10),
  tscan=list(),
  embeddr=list(nn_pct = 2),
  celltree_gibbs=list(sd_filter = 0),
  celltree_maptpx=list(sd_filter = 0),
  celltree_vem=list(sd_filter = 0),
  scuba=list(),
  slicer = list(min_branch_len=50),
  monocle_ddrtree=list(),
  wishbone = list(branch=F),
  pseudogp = list(iter=100, initialise_from="principal_curve"),
  mpath = list(numcluster=6),
  random_linear=list()
)
```

```{r}
results <- parallel::mclapply(names(method_descriptions), function(method_name) {
  library(dyneval)

  cat("Processing ", method_name, "\n", sep="")
  method <- get(paste0("description_", method_name))()
  method_params <- method_descriptions[[method_name]]

  factory <- function(fun) {
    function(...) {
      warn <- err <- NULL
      res <- withCallingHandlers(
        tryCatch(fun(...), error=function(e) {
          err <<- conditionMessage(e)
          NULL
        }), warning=function(w) {
          warn <<- append(warn, conditionMessage(w))
          invokeRestart("muffleWarning")
        })
      list(res, warn=warn, err=err, method_name=method_name)
    }
  }

  factory(function() {
    dyneval:::execute_evaluation(tasks, method, method_params, metrics = metric_names, suppress_output = F)
  }
  )()
}, mc.cores=4)
```

Gather the output of the different methods, and check whether there were any errors
```{r}
scores <- purrr::map(results, function(.) {
  bind_cols(
    tibble(error=list(.$err), warning=list(.$warning)),
    attr(.[[1]], "extras")$.summary,
    tibble(method_name=.$method_name)
  )
}) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null)) %>% dplyr::select(-contains("method_name1"))
```

```{r}
scores %>% filter(errored)
```

```{r}
scores %>%
  dplyr::select(-starts_with("time")) %>%
  gather(score_id, score, -method_name, -method_short_name, -task_id, -error, -warning) %>%
  ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + facet_wrap(~score_id) + coord_flip()
```

### 2017-09-06 analysis/1-methods-toys.Rmd
save toy scores data

@@
 }) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null)) %>% dplyr::select(-contains("method_name1"))
+
+scores %>% saveRDS("data/derived_data/toy_method_scores.rds")
 ```

### 2017-09-06 analysis/paper/paper.Rmd
changed name to dynalysis
added method toys figure

@@
 
-library(dynanalysis)
+library(dynalysis)
+library(tidyverse)
 ```
@@
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
+```{r get-data}
+scores <- read_rds("../data/derived_data/toy_method_scores.rds")
+
+scores %>%
+  dplyr::select(-starts_with("time")) %>%
+  gather(score_id, score, -method_name, -method_short_name, -task_id, -error, -warning) %>%
+  ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + facet_wrap(~score_id) + coord_flip()
 ```

### 2017-09-06 analysis/1-methods-toys.Rmd
added method toys figure

@@
 )
+
+metric_names <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score")
 ```
@@
 ```{r}
-results <- parallel::mclapply(names(method_descriptions), function(method_name) {
+results <- purrr::map(names(method_descriptions), function(method_name) {
   library(dyneval)
@@
   )()
-}, mc.cores=4)
+})
 ```

### 2017-09-06 analysis/paper/paper.Rmd
split up dockers, first dependencies, than analysis

@@
 author:
-  - author 1
-  - author 2
+  - Robrecht Cannoodt
+  - Wouter Saelens
+  - Helena Todorov
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-```{r, setup, echo = FALSE}
+```{r, setup, echo = FALSE, message=FALSE}
 knitr::opts_chunk$set(

### 2017-09-07 analysis_output/1-methods-toys.Rmd
fixed...... a lot......

--- 2017-09-06 analysis/1-methods-toys.Rmd
@@
 ```{r}
-method_descriptions <- list(
+method_params <- list(
   waterfall=list(),
@@
 )
+method_names <- names(method_params)
+
+method_descriptions <- map(method_names, ~get(paste0("description_", .))()) %>% set_names(method_names)
 
@@
 ```{r}
-results <- purrr::map(names(method_descriptions), function(method_name) {
+results <- purrr::map(method_names, function(method_name) {
   library(dyneval)
@@
   cat("Processing ", method_name, "\n", sep="")
-  method <- get(paste0("description_", method_name))()
-  method_params <- method_descriptions[[method_name]]
+  method <- method_descriptions[[method_name]]
+  params <- method_params[[method_name]]
 
@@
   factory(function() {
-    dyneval:::execute_evaluation(tasks, method, method_params, metrics = metric_names, suppress_output = F)
+    dyneval:::execute_evaluation(tasks, method, params, metrics = metric_names, suppress_output = F)
   }
@@
 scores %>% filter(errored)
+
+print(scores$error)
 ```

### 2017-09-07 analysis/1-methods-toys.Rmd
added method plotting

@@
   bind_cols(
-    tibble(error=list(.$err), warning=list(.$warning)),
+    tibble(error=list(.$err), warning=list(.$warning), model=list(attr(.[[1]], "extras")$.models[[1]])),
     attr(.[[1]], "extras")$.summary,
@@
   )
-}) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null)) %>% dplyr::select(-contains("method_name1"))
+}) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null), method_id=method_names) %>% dplyr::select(-contains("method_name1"))
 

### 2017-09-12 analysis/1-methods-toys.Rmd
added separate toy_params file and plotting

@@
 ```{r}
-method_params <- list(
-  waterfall=list(),
-  scorpius=list(),
-  slingshot=list(),
-  gpfates=list(nfates=1),
-  stemid=list(clustnr=10, bootnr=10, pdishuf=10),
-  tscan=list(),
-  embeddr=list(nn_pct = 2),
-  celltree_gibbs=list(sd_filter = 0),
-  celltree_maptpx=list(sd_filter = 0),
-  celltree_vem=list(sd_filter = 0),
-  scuba=list(),
-  slicer = list(min_branch_len=50),
-  monocle_ddrtree=list(),
-  wishbone = list(branch=F),
-  pseudogp = list(iter=100, initialise_from="principal_curve"),
-  mpath = list(numcluster=6),
-  random_linear=list()
-)
+method_params <- yaml::yaml.load_file("conf/toy_params.yml")$linear
 method_names <- names(method_params)
+method_names <- c("topslam")
 
@@
 ```{r}
+# results <- PRISM::qsub_lapply(names(method_descriptions), function(method_name) {
 results <- purrr::map(method_names, function(method_name) {
@@
 })
+# }, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1")))
 ```
@@
 
-scores %>% saveRDS("data/derived_data/toy_method_scores.rds")
+scores %>% saveRDS("results/toy_method_scores.rds")
 ```
@@
 
-```{r}
+```{r fig.height=10, fig.width=10}
 scores %>%
   dplyr::select(-starts_with("time")) %>%
-  gather(score_id, score, -method_name, -method_short_name, -task_id, -error, -warning) %>%
+  gather(score_id, score, -method_name, -method_short_name, -method_id, -task_id, -error, -warning, -model) %>%
+  filter(score_id %in% c("errored", "correlation", "robbie_network_score", "mean_R_nx")) %>% 
   ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + facet_wrap(~score_id) + coord_flip()

### 2017-09-12 analysis/method-qc.Rmd
added method qc

---
title: "method-qc"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(googlesheets)
library(cowplot)
rules_info <- gs_title("Trajectory Inference Methods") %>% gs_read(skip=1)
```

```{r}
ggplot(rules_info,aes(log(Citations+1),`QC score`))+
  geom_smooth(method='lm',formula=y~x) +
  geom_point() +
  ggrepel::geom_label_repel(aes(label=Name))

ggplot(rules_info,aes(log(Citations+1),`QC score`))+
  geom_smooth(method='lm',formula=y~x) +
  geom_point() +
  ggrepel::geom_label_repel(aes(label=Name))

rules_info$date <- pmax(
  lubridate::as_date(lubridate::ymd(rules_info$PubDate)),
  lubridate::as_date(lubridate::ymd(rules_info$Preprint)),
  na.rm=T
)

ggplot(rules_info,aes(date,`QC score`))+
  geom_smooth(method='lm',formula=y~x) +
  geom_point() +
  ggrepel::geom_label_repel(aes(label=Name))

rules_info$published <- !is.na(rules_info$PubDate)

ggplot(rules_info,aes(published,`QC score`))+
  geom_violin() +
  geom_point() +
  ggrepel::geom_text_repel(aes(label=Name))

```

### 2017-09-12 analysis/1-methods-toys.Rmd
added method qc

@@
 method_names <- names(method_params)
-method_names <- c("topslam")
+# method_names <- c("celltree_maptpx")
 
@@
 ```{r}
-# results <- PRISM::qsub_lapply(names(method_descriptions), function(method_name) {
-results <- purrr::map(method_names, function(method_name) {
+results <- PRISM::qsub_lapply(names(method_descriptions), function(method_name) {
+# results <- purrr::map(method_names, function(method_name) {
   library(dyneval)
@@
   )()
-})
-# }, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1")))
+# })
+}, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1")))
 ```

### 2017-09-14 analysis/1-methods-toys.Rmd
aggregating scores

@@
 method_names <- names(method_params)
-# method_names <- c("celltree_maptpx")
+# method_names <- c("ouija", "slice")
 
@@
 # })
-}, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1")))
+}, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
 ```
@@
 ```{r}
+results <- PRISM::qsub_retrieve(results)
 scores <- purrr::map(results, function(.) {

### 2017-09-14 analysis/analyses/score_rules/1-generate-toys.Rmd
aggregating scores

---
title: "1-generate-toys"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generate toy datasets for comparing scores
```{r}
library(tidyverse)

source("generation.R")
source("perturbation.R")
```

```{r}
toys_blueprint <- tribble(
  ~generator_id, ~perturbator_id,
  "linear", "gs",
  "linear", "switch_two_cells",
  "linear", "switch_all_cells",
  "linear", "join_linear",
  "linear", "split_linear",
  "linear", "warp",
  "linear", "hairy",
  "linear", "hairy_small",
  "linear", "hairy_large",
  "linear", "remove_cells",
  "linear", "add_distant_edge",
  "bifurcating", "gs",
  "bifurcating", "switch_two_cells",
  "bifurcating", "switch_all_cells",
  "bifurcating", "warp",
  "bifurcating", "hairy",
  "bifurcating", "remove_cells",
  "bifurcating", "add_distant_edge",
  "cycle", "gs",
  "cycle", "switch_all_cells",
  "cycle", "break_cycles",
  "cycle", "warp",
  "cycle", "hairy",
  "cycle", "remove_cells",
  "cycle", "add_distant_edge"
) %>% bind_rows(tibble(generator_id="linear", perturbator_id=paste0("switch_", percs * 100))) %>% 
  rowwise() %>% 
  mutate(
  generator=list(get(paste0("generate_", generator_id))),
  perturbator=list(get(paste0("perturb_", perturbator_id)))
)

# replicate
nreplicates <- 1
toys <- toys_blueprint %>% slice(rep(1:n(), each=nreplicates)) %>% mutate(
  replicate=seq_len(nrow(.))%%nreplicates,
  toy_category=paste0(generator_id, "-", perturbator_id),
  toy_id=paste0(toy_category, "-", replicate)
) %>%
  mutate(ncells=sample(50:500, n(), replace=TRUE))

# generate gold standards and toys, can take some time (for computing the geodesic distances I presume)
# I choose to not do this using mutate because it is much easier to debug using loops
toys$gs <- toys %>% split(seq_len(nrow(toys))) %>% parallel::mclapply(function(row) {
  row$generator[[1]](row$ncells)
}, mc.cores=8)
#toys$gs <- map2(toys$generator, toys$ncells, ~.x(.y))

toys$toy <- toys %>% split(seq_len(nrow(toys))) %>% parallel::mclapply(function(row) {
  row$perturbator[[1]](row$gs[[1]])
}, mc.cores=8)
# toys$toy <- map2(toys$perturbator, toys$gs, ~.x(.y))
toys$toy <- map2(toys$toy, toys$toy_id, ~rename_toy(.x, .y))

# plot toys
toyplots <- toys %>% group_by(toy_category) %>% filter(row_number()==1) %>%
  {split(., seq_len(nrow(.)))} %>% parallel::mclapply(function(row) dynplot::plot_strip_connections(row$gs[[1]], row$toy[[1]]), mc.cores = 8)

write_rds(toys, "toys.rds")
```

```{r}
toys <- read_rds("toys.rds")

# get the scores when comparing the gs to toy
metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score", "mantel_pval")
compare_toy <- function(gs, toy, id=toy$id) {
  scores <- dyneval:::calculate_metrics(gs, toy, metrics=metrics)$summary
  scores %>% mutate(toy_id=id)
}

# get the scores when both comparing gs to gs and comparing gs to toy
compare_gs_toy <- function(gs, toy) {
  bind_rows(
    compare_toy(gs, toy) %>% mutate(comparison="gs-toy"),
    compare_toy(gs, gs, id=toy$id) %>% mutate(comparison="gs-gs")
  )
}

scores <- toys %>% rowwise() %>% do(compare_gs_toy(.$gs, .$toy)) %>% select(-starts_with("time"))

write_rds(scores, "scores.rds")
```

### 2017-09-14 analysis/analyses/score_rules/2-rules.Rmd
aggregating scores

---
title: "2-rules"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
scores <- read_rds("scores.rds")
toys <- read_rds("toys.rds")

# summarise the scores
# get gs-toy score
# get difference and fractions between gs-toy and gs-gs
scores_summary <- scores %>% gather(score_id, score, -toy_id, -comparison) %>%
  spread(comparison, score) %>% mutate(
    score=`gs-toy`,
    diff= `gs-toy`-`gs-gs`,
    frac=`gs-toy`/`gs-gs`
  ) %>% select(-`gs-toy`, -`gs-gs`) %>% left_join(toys, by="toy_id")

scores_summary %>%
  ggplot() +
  geom_boxplot(aes(toy_category, score, color=perturbator_id)) +
  facet_wrap(~score_id) +
  coord_flip()

scores_summary %>%
  ggplot() +
  geom_boxplot(aes(toy_category, diff, color=perturbator_id)) +
  facet_wrap(~score_id) +
  coord_flip()
```

```{r}
library(googlesheets)
rules_info <- gs_title("Metrics rules") %>% gs_read() %>% mutate(id=factor(id, levels=id))

source("rules.R")
rules_info$check_func <- map(paste0("check_", rules_info$id), get)

rules <- map2(
  rules_info$check_func,
  rules_info$id,
  ~.x(scores_summary) %>%
    mutate(rule_id=.y)
  ) %>% bind_rows() %>% left_join(rules_info, by=c("rule_id"="id"))

## Plot all rules
rules %>% ggplot(aes(rule_id, score_id)) +
  geom_raster(aes(fill = c("red", "green")[as.numeric(rule)+1])) +
  geom_text(aes(label = c("▼", "▲")[as.numeric(rule)+1]), color="white") +
  scale_fill_identity() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
## Possible score combinations satisfying all rules
allscores <-unique(rules$score_id)
score_ids_combinations <- map(1:length(allscores), function(x) map(combn(length(allscores), x, simplify = FALSE), ~allscores[.])) %>% unlist(recursive=FALSE)

combination_scores <- map(score_ids_combinations, function(score_ids_combination) {
  rules %>% filter(score_id %in% score_ids_combination) %>%
    group_by(rule_id) %>%
    summarise(any=any(rule), number=sum(rule, na.rm=TRUE)) %>%
    group_by() %>%
    summarise(all=all(any), avg_rules_retrieved=sum(number)/length(score_ids_combination), nscores = length(score_ids_combination)) %>%
    mutate(score_ids_combination=list(score_ids_combination))
}) %>% bind_rows()

combination_scores %>% filter(all) %>% arrange(nscores, -avg_rules_retrieved) %>% head(5)

combination_scores %>% filter(all) %>% arrange(nscores, -avg_rules_retrieved) %>% pull(score_ids_combination) %>% head(5)
```

## Aggregating metrics to get one final metric following all scores

In the end, we want to compare methods using one aggregated number which follows all the rules of a good evaluation metric. We will aggregate all combinations of metrics using different aggregation functions.
```{r}
score_ids_combinations <- map(1:length(allscores), function(x) map(combn(length(allscores), x, simplify = FALSE), ~allscores[.])) %>% unlist(recursive=FALSE)
new_scores_combinations <- tibble(
  combination = score_ids_combinations, 
  nscores = map_int(combination, length),
  combination_id = map_chr(combination, ~paste0(., collapse=","))
)
new_scores_aggregations <- tibble(
  aggregation_id = c("harmonic_mean", "arithmetic_mean", "geometric_mean", "max", "min"),
  aggregation_func = list(
    function(x) {if(any(x == 0)) {return(0)} else {length(x) /sum(1/x)}}, 
    mean, 
    function(x) {if(any(x == 0)) {return(0)} else {prod(x)^(1/length(x))}},
    max,
    min
  )
)

new_scores <- merge(new_scores_combinations, new_scores_aggregations) %>% as_tibble() %>% 
  mutate(score_id = paste0(aggregation_id, "(", combination_id, ")"))
```

```{r}
scores2 <- new_scores %>% rowwise() %>% do(
  tibble(score = apply(scores[.$combination], 1, .$aggregation_func), score_id=.$score_id, i=seq_len(nrow(scores)))
) %>% spread(score_id, score) %>% bind_cols(scores %>% select(toy_id, comparison))

scores2_summary <- scores2 %>% gather(score_id, score, -toy_id, -comparison) %>%
  spread(comparison, score) %>% mutate(
    score=`gs-toy`,
    diff= `gs-toy`-`gs-gs`,
    frac=`gs-toy`/`gs-gs`
  ) %>% select(-`gs-toy`, -`gs-gs`) %>% left_join(toys, by="toy_id")

rules <- map2(
  rules_info$check_func,
  rules_info$id,
  ~.x(scores2_summary) %>%
    mutate(rule_id=.y)
  ) %>% bind_rows() %>% left_join(rules_info, by=c("rule_id"="id"))

rules %>% group_by(score_id) %>% 
  summarise(allpassed = all(rule), npassed=sum(rule)) %>% 
  left_join(new_scores, by="score_id") %>% 
  filter(allpassed) %>% 
  arrange(nscores)
```

### 2017-09-14 analysis/analyses/score_rules/2-rules.Rmd
grouped rules, added rules, added perturbations, added comparison with grouped cells

@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
 ```{r}
@@
 toys <- read_rds("toys.rds")
+timings <- read_rds("timings.rds")
 
@@
     diff= `gs-toy`-`gs-gs`,
-    frac=`gs-toy`/`gs-gs`
-  ) %>% select(-`gs-toy`, -`gs-gs`) %>% left_join(toys, by="toy_id")
+    frac=`gs-toy`/`gs-gs`,
+    score_grouped = `gs_groupid-toy`
+  ) %>% select(-`gs-toy`, -`gs-gs`, -`gs_groupid-toy`) %>% left_join(toys, by="toy_id")
 
@@
 library(googlesheets)
-rules_info <- gs_title("Metrics rules") %>% gs_read() %>% mutate(id=factor(id, levels=id))
+rules_info <- gs_title("Metrics rules") %>% gs_read("rules") %>% mutate(id=factor(id, levels=id))
 
@@
 
+metric_types <- rep("positional", length(unique(scores_summary$score_id))) %>% setNames(unique(scores_summary$score_id))
+metric_types[c("ged", "robbie_network_score", "isomorphic")] <- "structural"
+
+rules$metric_type <- metric_types[rules$score_id]
+rules$rule_type <- rules$type
+
 ## Plot all rules
@@
 
+## Best combinations of scores
 ```{r}
@@
 combination_scores <- map(score_ids_combinations, function(score_ids_combination) {
-  rules %>% filter(score_id %in% score_ids_combination) %>%
+  rules %>% 
+    filter(rule_id != "structure_and_position") %>% # do not look at this rule, none of the scores can satisfy it alone
+    filter(score_id %in% score_ids_combination) %>%
     group_by(rule_id) %>%
@@
 
-## Aggregating metrics to get one final metric following all scores
+## Best aggregations of scores
 
@@
 ```{r}
+# score combinations
 score_ids_combinations <- map(1:length(allscores), function(x) map(combn(length(allscores), x, simplify = FALSE), ~allscores[.])) %>% unlist(recursive=FALSE)
@@
 )
+
+# aggregation functions
 new_scores_aggregations <- tibble(
@@
 
+# all combinations of score combinations and aggregation functions
 new_scores <- merge(new_scores_combinations, new_scores_aggregations) %>% as_tibble() %>% 
@@
 ```{r}
+# calculate the scores
 scores2 <- new_scores %>% rowwise() %>% do(
@@
 
+# summarize the scores
 scores2_summary <- scores2 %>% gather(score_id, score, -toy_id, -comparison) %>%
@@
 
+# check whether a score follows a rule
 rules <- map2(
@@
   ) %>% bind_rows() %>% left_join(rules_info, by=c("rule_id"="id"))
-
+```
+We want a score to follow all rules
+```{r}
 rules %>% group_by(score_id) %>% 
@@
 ```
+
+## Grouping cells
+```{r}
+scores_summary %>% ggplot() + geom_point(aes(score, score_grouped, color=generator_id)) + facet_wrap(~score_id)
+
+scores_summary %>% group_by(score_id) %>% summarise(
+  cor = cor(score, score_grouped),
+  rms = sqrt(sum((score - score_grouped)^2)),
+  rms_shifted = sqrt(sum((score - mean(score) - score_grouped + mean(score_grouped))^2))
+)
+```

### 2017-09-14 analysis/analyses/score_rules/1-generate-toys.Rmd
grouped rules, added rules, added perturbations, added comparison with grouped cells

@@
 library(tidyverse)
+library(dyneval)
 
@@
   "linear", "add_distant_edge",
+  "linear", "structure_and_position",
   "bifurcating", "gs",
+  "bifurcating", "change_lengths",
+  "bifurcating", "change_terminal_lengths",
   "bifurcating", "switch_two_cells",
@@
   "bifurcating", "add_distant_edge",
+  "bifurcating", "structure_and_position",
   "cycle", "gs",
@@
   "cycle", "add_distant_edge"
-) %>% bind_rows(tibble(generator_id="linear", perturbator_id=paste0("switch_", percs * 100))) %>% 
+) %>% 
+  # bind_rows(tibble(generator_id="linear", perturbator_id=paste0("switch_", percs * 100))) %>% 
   rowwise() %>% 
@@
 
+toys$gs_grouped <- parallel::mclapply(toys$gs, group_task, mc.cores=8)
+
 # plot toys
@@
 # get the scores when comparing the gs to toy
-metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score", "mantel_pval")
 compare_toy <- function(gs, toy, id=toy$id) {
@@
 # get the scores when both comparing gs to gs and comparing gs to toy
-compare_gs_toy <- function(gs, toy) {
+compare_gs_toy <- function(gs, toy, gs_grouped) {
   bind_rows(
     compare_toy(gs, toy) %>% mutate(comparison="gs-toy"),
-    compare_toy(gs, gs, id=toy$id) %>% mutate(comparison="gs-gs")
+    compare_toy(gs, gs, id=toy$id) %>% mutate(comparison="gs-gs"),
+    compare_toy(gs_grouped, toy) %>% mutate(comparison="gs_groupid-toy")
   )
@@
 
-scores <- toys %>% rowwise() %>% do(compare_gs_toy(.$gs, .$toy)) %>% select(-starts_with("time"))
+metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score", "mantel_pval", "ged")
+
+results <- toys %>% rowwise() %>% do(compare_gs_toy(.$gs, .$toy, .$gs_grouped))# %>% select(-starts_with("time"))
+scores <- results %>% select(-starts_with("time"))
 
@@
 ```
+
+```{r}
+timings <- results %>% 
+  select(starts_with("time")) %>% 
+  rename_all(list(function(x) substring(x, 6))) %>% 
+  gather(metric_group_id, time) %>% 
+  group_by(metric_group_id) %>% 
+  summarise(time = mean(time))
+
+metric_groups <- c("mean_R_nx"="coranking", "auc_R_nx"="coranking", "Q_local"="coranking", "Q_global"="coranking", "correlation"="correlation", "isomorphic"="isomorphic", "robbie_network_score"="robbie_network_score", "mantel_pval"="mantel", "ged"="ged") %>%
+  enframe("metric_id", "metric_group_id")
+
+timings %>% 
+  left_join(metric_groups, by="metric_group_id") %>% 
+  write_rds("timings.rds")
+```

### 2017-09-14 analysis/analyses/score_rules/1-generate-toys.Rmd
added grouping rule

@@
 ) %>% 
-  # bind_rows(tibble(generator_id="linear", perturbator_id=paste0("switch_", percs * 100))) %>% 
+  bind_rows(tibble(generator_id="linear", perturbator_id=paste0("switch_", percs * 100))) %>%
   rowwise() %>% 

### 2017-09-15 analysis/analyses/evaluate_toy/2-methods-plots.Rmd
added method qc, moved analyses in separate folders

---
title: "1b-methods-toys-plots"
output: html_document
---

## Arrange all method plots in a grid

```{r loading}
library(tidyverse)
library(dyneval)
```

```{r}
plot_method <- function(model, method_name) {
  print(method_name)
  plot_fun <- method_descriptions[[method_name]]$plot_fun

  pdf(paste0("figures/toy_method_plots/", method_name, ".pdf"), width=5, height=5)

  plot <- plot_fun(model)
  if("gg" %in% class(plot)) {
    print(plot)
  }

  dev.off()
}

scores <- readRDS("results/toy_method_scores.rds") %>% filter(method_name != "random_linear")

method_descriptions <- map(scores$method_id, ~get(paste0("description_", .))()) %>% set_names(scores$method_id)
plots <- map2(scores$model, scores$method_id, plot_method)
```

```{r}
system("montage -geometry 600x600+0+0 figures/toy_method_plots/*.pdf figures/toy_method_plots/grid.png")
```

### 2017-09-15 analysis/paper/paper.Rmd
added separate toy_params file and plotting
added method qc, moved analyses in separate folders

@@
 
+## Evaluation workflow
+Overview of methods...
+Overview of parameter optimalisation...
+Nicely illustrated using the toy datasets: Simpele toy datasets. Theoretische mogelijkheden van elke methode + bewijs van waarde van parameteroptimalisatie
+
+## Evaluation on synthetic data
+Initial evaluation..
+Robustness..
+Gegenereerde synthetische data. Parameter tuning voor echte data + initiële evaluatie (tot wat is de methode in staat) + robustness etc testing
+
+## Evaluation on real data
+Biological relevance..
+Echte data. Evaluatie (biologische relevantie van de methode)
+
 ```{r get-data}
-scores <- read_rds("../data/derived_data/toy_method_scores.rds")
+scores <- read_rds("../results/toy_method_scores.rds")
 

### 2017-09-17 analysis/method-qc.Rmd
added method qc, moved analyses in separate folders
add support for number of citations

@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(cowplot)
-rules_info <- gs_title("Trajectory Inference Methods") %>% gs_read(skip=1)
+library(tidyverse)
+rules_info <- gs_title("Trajectory Inference Methods") %>% 
+  gs_read(ws = "Software", col_types = cols(GScholarClusterID = "c"), skip = 1)
+```
+
+Update number of citations
+```{r}
+# script_file <- "analysis/scholar.py"
+# download.file("https://raw.githubusercontent.com/ckreibich/scholar.py/master/scholar.py", destfile = script_file)
+
+num_citations_by_clusterid <- function(clusterid, scholar_file = script_file) {
+  tryCatch({
+    command <- paste0("python ", scholar_file, " -C ", clusterid, " --csv-header")
+    output <- system(command, intern = T)
+    tab <- readr::read_delim(paste(gsub("\n", " ", output), collapse = "\n"), delim = "|")
+    sum(tab$num_citations)
+  }, error = function(e) NA)
+}
+
+# If you're connected to UGentNet:
+# rules_info <- rules_info %>%
+#   mutate(Citations = pbapply::pbsapply(GScholarClusterID, num_citations_by_clusterid))
 ```
@@
 
-ggplot(rules_info,aes(log(Citations+1),`QC score`))+
+cit_breaks <- c(0, 1, 10, 100, max(rules_info$Citations))
+ggplot(rules_info,aes(Citations+1,`QC score`))+
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name))
+  ggrepel::geom_label_repel(aes(label=Name)) +
+  scale_x_log10(breaks = cit_breaks+1, labels = cit_breaks)
 
@@
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name))
+  ggrepel::geom_label_repel(aes(label=Name)) +
+  scale_y_continuous(breaks = seq_len(10))
 
@@
 ```
+
+```{r}
+
+```

### 2017-09-18 analysis/method-qc.Rmd
added kinds of output

--- 2017-09-12 analysis/method-qc.Rmd
@@
 library(cowplot)
+library(tidyverse)
 rules_info <- gs_title("Trajectory Inference Methods") %>% gs_read(skip=1)
@@
 ```{r}
-ggplot(rules_info,aes(log(Citations+1),`QC score`))+
+ggplot(rules_info,aes(log(Citations+1), qc_score))+
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name))
+  ggrepel::geom_label_repel(aes(label=name))
 
-ggplot(rules_info,aes(log(Citations+1),`QC score`))+
+ggplot(rules_info,aes(log(Citations+1), qc_score))+
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name))
+  ggrepel::geom_label_repel(aes(label=name))
 
@@
 
-ggplot(rules_info,aes(date,`QC score`))+
+ggplot(rules_info,aes(date, qc_score))+
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name))
+  ggrepel::geom_label_repel(aes(label=name))
 
@@
 
-ggplot(rules_info,aes(published,`QC score`))+
+ggplot(rules_info,aes(published, qc_score))+
   geom_violin() +
   geom_point() +
-  ggrepel::geom_text_repel(aes(label=Name))
+  ggrepel::geom_text_repel(aes(label=name))
 
 ```
+
+Different methods generate different kinds of output...
+```{r}
+rules_info %>% filter(wrapper == "Done") %>% select(ids, output) %>% 
+  separate_rows(ids, sep=",") %>% 
+  separate_rows(output, sep=",") %>% 
+  mutate(output=trimws(output)) %>% 
+  ggplot() + 
+    geom_bar(aes(output)) +
+    coord_flip()
+```

### 2017-09-18 analysis/method-qc.Rmd

--- 2017-09-17 analysis/method-qc.Rmd
@@
 ```{r}
-ggplot(rules_info,aes(log(Citations+1),`QC score`))+
+ggplot(rules_info,aes(log(Citations+1), qc_score))+
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name))
+  ggrepel::geom_label_repel(aes(label=name))
 
 cit_breaks <- c(0, 1, 10, 100, max(rules_info$Citations))
-ggplot(rules_info,aes(Citations+1,`QC score`))+
+ggplot(rules_info,aes(Citations+1, qc_score))+
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name)) +
+  ggrepel::geom_label_repel(aes(label=name)) +
   scale_x_log10(breaks = cit_breaks+1, labels = cit_breaks)
@@
 
-ggplot(rules_info,aes(date,`QC score`))+
+ggplot(rules_info,aes(date, qc_score)) +
   geom_smooth(method='lm',formula=y~x) +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label=Name)) +
+  ggrepel::geom_label_repel(aes(label=name)) +
   scale_y_continuous(breaks = seq_len(10))
@@
 
-ggplot(rules_info,aes(published,`QC score`))+
+ggplot(rules_info,aes(published, qc_score))+
   geom_violin() +
   geom_point() +
-  ggrepel::geom_text_repel(aes(label=Name))
+  ggrepel::geom_text_repel(aes(label=name))
 
@@
 
+Different methods generate different kinds of output...
 ```{r}
-
+rules_info %>% filter(wrapper == "Done") %>% select(ids, output) %>% 
+  separate_rows(ids, sep=",") %>% 
+  separate_rows(output, sep=",") %>% 
+  mutate(output=trimws(output)) %>% 
+  ggplot() + 
+    geom_bar(aes(output)) +
+    coord_flip()
 ```

### 2017-10-03 analysis/paper/paper.Rmd
added kinds of output
nmeth csl

@@
   - Helena Todorov
+  - Yvan Saeys
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 bibliography: references.bib
-csl: journal-of-archaeological-science.csl
+csl: nature-methods.csl
 abstract: |
@@
 
+*Points to maybe raise*:
+- Some methods can handle different trajectories which are not directly described in the main paper: eg.
+circular trajectories (for a lot of principal curves algorithms using `periodic.lowess`), trees (for a lot of bifurcation methods)
+
 ## Evaluation on synthetic data

### 2017-10-03 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
added method qc, moved analyses in separate folders
updated toy params

--- 2017-09-14 analysis/1-methods-toys.Rmd
@@
 library(tidyverse)
+library(dyntoy)
 library(dyneval)
@@
 
-tasks <- generate_toy_datasets()
-tasks <- tasks[1, ]
+tasks <- generate_toy_datasets(num_cells=500)
+#tasks <- tasks[1, ]
+```
+
+```{r}
+task <- dynutils::extract_row_to_list(tasks, 2)
+pheatmap::pheatmap(task$counts)
 ```
@@
 ```{r}
-method_params <- yaml::yaml.load_file("conf/toy_params.yml")$linear
+method_params <- yaml::yaml.load_file("toy_params.yml")$linear
 method_names <- names(method_params)

### 2017-11-20 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
reordered analyses

@@
 
-tasks <- generate_toy_datasets(num_cells=500)
-#tasks <- tasks[1, ]
+tasks <- generate_toy_datasets(num_replicates = 1, num_cells=300)
 ```
@@
 ```{r}
+tasks$counts <- map(tasks$counts, function(counts) {
+  cbind2(counts, matrix(sample(1:100, nrow(counts)*100, replace=TRUE), nrow=nrow(counts))) %>% {magrittr::set_colnames(., paste0("G", 1:ncol(.)))}
+})
 task <- dynutils::extract_row_to_list(tasks, 2)
 pheatmap::pheatmap(task$counts)
+
+source("~/thesis/projects/dynverse/dynmodular/dimred_wrappers.R")
+space <- task$counts %>% dimred_dp(2, neigen=2) %>% as.data.frame
+# space <- task$counts %>% dimred_mds(2) %>% as.data.frame
+# space <- task$counts %>% dimred_lle(2) %>% as.data.frame
+space <- space %>% bind_cols(task$cell_grouping %>% slice(match(rownames(task$counts), cell_id)))
+space %>%  ggplot() + geom_point(aes(Comp1, Comp2, color=group_id))
 ```
@@
 metric_names <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score")
+
+method_names <- c("monocle_ddrtree", "scorpius")
 ```
@@
 ```{r}
-results <- PRISM::qsub_lapply(names(method_descriptions), function(method_name) {
-# results <- purrr::map(method_names, function(method_name) {
+# results <- PRISM::qsub_lapply(names(method_descriptions), function(method_name) {
+results <- purrr::map(method_names, function(method_name) {
   library(dyneval)
@@
   factory(function() {
-    dyneval:::execute_evaluation(tasks, method, params, metrics = metric_names, suppress_output = F)
+    dyneval:::execute_evaluation(tasks, method, params, metrics = metric_names, timeout = 99999999999999)
   }
   )()
-# })
-}, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
+})
+# }, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
 ```
@@
 ```{r}
-results <- PRISM::qsub_retrieve(results)
+# results <- PRISM::qsub_retrieve(results)
 scores <- purrr::map(results, function(.) {
   bind_cols(
-    tibble(error=list(.$err), warning=list(.$warning), model=list(attr(.[[1]], "extras")$.models[[1]])),
-    attr(.[[1]], "extras")$.summary,
-    tibble(method_name=.$method_name)
-  )
-}) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null), method_id=method_names) %>% dplyr::select(-contains("method_name1"))
+    tibble(
+      error=list(.$err), 
+      warning=list(.$warning), 
+      model=attr(.[[1]], "extras")$.models
+    ),
+    attr(.[[1]], "extras")$.summary
+  ) %>% mutate(method_name=.$method_name)
+}) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null), method_id=rep(method_names, each=nrow(tasks))) %>% dplyr::select(-contains("method_name1"))
 
-scores %>% saveRDS("results/toy_method_scores.rds")
+# scores %>% saveRDS("results/toy_method_scores.rds")
 ```
@@
 ```{r fig.height=10, fig.width=10}
-scores %>%
+scores %>% left_join(tasks %>% select(id, ti_type), by=c("task_id"="id")) %>% 
   dplyr::select(-starts_with("time")) %>%
-  gather(score_id, score, -method_name, -method_short_name, -method_id, -task_id, -error, -warning, -model) %>%
+  gather(score_id, score, -method_name, -method_short_name, -method_id, -task_id, -error, -warning, -model, -ti_type) %>%
   filter(score_id %in% c("errored", "correlation", "robbie_network_score", "mean_R_nx")) %>% 
-  ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + facet_wrap(~score_id) + coord_flip()
+  ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + 
+  facet_grid(ti_type~score_id) + coord_flip()
 ```

### 2017-11-20 analysis/analyses/3-score_rules/2-rules.Rmd

--- 2017-09-14 analysis/analyses/score_rules/2-rules.Rmd
@@
 ---
+
+```{r}
+library(tidyverse)
+library(dyneval)
+```
 
@@
     diff= `gs-toy`-`gs-gs`,
-    frac=`gs-toy`/`gs-gs`
+    frac=`gs-toy`/`gs-gs`,
+    score_grouped = `gs_groupid-toy`
   ) %>% select(-`gs-toy`, -`gs-gs`) %>% left_join(toys, by="toy_id")
@@
 ```{r}
-scores_summary %>% ggplot() + geom_point(aes(score, score_grouped, color=generator_id)) + facet_wrap(~score_id)
+check_grouping(scores_summary)
 
-scores_summary %>% group_by(score_id) %>% summarise(
-  cor = cor(score, score_grouped),
-  rms = sqrt(sum((score - score_grouped)^2)),
-  rms_shifted = sqrt(sum((score - mean(score) - score_grouped + mean(score_grouped))^2))
-)
+plot_grouping(scores_summary)
 ```

### 2017-11-21 analysis/paper/paper.Rmd
cleanup analysis and paper (for now)

@@
 abstract: |
-  Text of abstract
+  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cell differentiation.
+Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories.
+However, the performance of these methods have not yet been comprehensively evaluated.
+Here we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets.
+We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness.
+We conclude with practical guidelines for method users, as well as future .
+Our whole evaluation pipeline is fully reproducible and available at www.github.com/dynverse.
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
 highlights: |
@@
 # Introduction
-
-Here is a citation [@Marwick2017]
-
-# Background
+Single-cell omics technologies
 
@@
 ```{r get-data}
-scores <- read_rds("../results/toy_method_scores.rds")
-
-scores %>%
-  dplyr::select(-starts_with("time")) %>%
-  gather(score_id, score, -method_name, -method_short_name, -task_id, -error, -warning) %>%
-  ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + facet_wrap(~score_id) + coord_flip()
 ```

### 2017-11-21 analysis/paper/paper.Rmd
added references, added abstract, added some introduction

@@
 ---
-title: "Title Goes Here"
+title: "An evaluation framework for trajectory inference methods"
 author:
-  - Robrecht Cannoodt
-  - Wouter Saelens
+  - Robrecht Cannoodt*
+  - Wouter Saelens*
   - Helena Todorov
@@
 abstract: |
-  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cell differentiation.
+  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
 Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories.
 However, the performance of these methods have not yet been comprehensively evaluated.
-Here we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets.
+Here, we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets.
 We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness.
-We conclude with practical guidelines for method users, as well as future .
-Our whole evaluation pipeline is fully reproducible and available at www.github.com/dynverse.
+We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
+Our whole evaluation pipeline is fully reproducible, can be easily extended and is available at www.github.com/dynverse.
 keywords: |
@@
 ---
+\* Equal contribution
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-```{r, setup, echo = FALSE, message=FALSE}
+```{r setup, include=FALSE}
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
 knitr::opts_chunk$set(
@@
   comment = "#>",
-  fig.path = "../figures/"
+  fig.path = "analysis/figures/"
 )
@@
 library(tidyverse)
+
+figs <- list()
+sfigs <- list()
 ```
@@
 # Introduction
-Single-cell omics technologies
+<!-- Three paragraph introduction, typical for Nat Biotech papers -->
+<!-- General into about single-cell & trajectories -->
+Single-cell -omics technologies now make it possible to more accuratly model biological systems than ever before.
+One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
+When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
+Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation <!-- refs -->.
 
-```{r demo, eval = FALSE}
-1 + 1
-```
+<!-- Discussion of TI, history, different methods, brief future perspectives -->
+Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month.
+At first, most trajectory inference was done for linear trajectories, but several methods which can detect complex branching or cyclical behaviour
+
+<!-- Context and general overview of our evaluation -->
+Given this plethora of available trajectory inference methods,
+
+Trajectories can then be used to delineate the different stages in the dynamic process and their associated gene signatures <!-- REF -->,
 
@@
 # Results
+## Evaluation workflow
+### Method characterization
+```{r}
+# Load method characteristics
+method_df <- readRDS(derived_file("method_df.rds", experiment_name="method_characteristics"))
+```
 
-## Evaluation workflow
-Overview of methods...
-Overview of parameter optimalisation...
-Nicely illustrated using the toy datasets: Simpele toy datasets. Theoretische mogelijkheden van elke methode + bewijs van waarde van parameteroptimalisatie
+<!-- I would move some stuff from this paragraph to the supplementary if necessary -->
+We gathered a total of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, based on free availability and being able to be used within a programming context.
+We characterized the methods from both a computational and a user perspective.
+Computationally, methods use a variety of algorithms and subcomponents to reach a final ordering.
+However, most use a common framework consisting of three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
+<!-- -> add supplementary information -->  We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extendability.
+We first contacted all the authors, to allow them to improve their method before publishing our evaluation.
+While most methods lack at several metrics, code quality tends to increase over time.
 
+### Data sources
+We used three different data sources
+
+### Score metrics and parameter optimisation
+
+## Overall method comparison
 *Points to maybe raise*:
@@
 
-# References
-
-### Colophon
+# Colophon
 
@@
 # if your Rmd is not in analysis/paper/
-git2r::repository("../..")
+git2r::repository(".")
 ```
+
+# References

### 2017-11-21 analysis/paper/paper.Rmd
nature biotechnology reference style

@@
 bibliography: references.bib
-csl: nature-methods.csl
+csl: nature-biotechnology.csl
 abstract: |

### 2017-11-21 analysis/paper/paper.Rmd
updated introduction

@@
 We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
-Our whole evaluation pipeline is fully reproducible, can be easily extended and is available at www.github.com/dynverse.
+Our evaluation pipeline is fully reproducible, can be easily extended and is available at www.github.com/dynverse.
 keywords: |
@@
 library(tidyverse)
+library(cowplot)
 
-figs <- list()
-sfigs <- list()
+# refering to figures & supplementary figures
+refs <- tibble(ref_id = character(), name = character(), ref_type = character())
+ref <- function(ref_type, ref_id, suffix="") {
+  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
+    refs <<- refs %>% bind_rows(tibble(
+      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
+      ref_type = ref_type,
+      ref_id = ref_id
+    ))
+  }
+  refs %>% filter(ref_id == !!ref_id) %>% pull(name) %>% paste0("**", ., suffix, "**")
+}
+create_names <- list(
+  sfig = function(i) {glue::glue("Supplementary Figure {i}")},
+  fig = function(i) {glue::glue("Figure {i}")}
+)
 ```
@@
 When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation <!-- refs -->.
+Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation<!-- refs -->.
 
 <!-- Discussion of TI, history, different methods, brief future perspectives -->
-Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month.
-At first, most trajectory inference was done for linear trajectories, but several methods which can detect complex branching or cyclical behaviour
+Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
+Initially most trajectory inference was focused on linear trajectories, with one start and one end point.
+However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
+Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes hapening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+
+```{r, fig.width = 15, fig.height = 10}
+n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_name="method_characteristics"))
+trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_name="method_characteristics"))
+
+cowplot::plot_grid(plotlist=list(n_methods_over_time, trajectory_components_over_time), nrow=2, rel_heights = c(0.7, 0.3), labels="auto")
+```
+`r ref("sfig", "methods_over_time")` New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
+
+<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
+Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
+These components are frequently interchangeable.
+Depending on the way the cells are ordered, some prior information about the dynamic process can also be required, or optional.
+<!---- TODO --->
 
 <!-- Context and general overview of our evaluation -->
-Given this plethora of available trajectory inference methods,
-
-Trajectories can then be used to delineate the different stages in the dynamic process and their associated gene signatures <!-- REF -->,
-
-# Methods
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
+Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
+In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods.
+We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from current knowledge.
+We include a rigorous parameter optimisation.
+To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
 
@@
 We characterized the methods from both a computational and a user perspective.
-Computationally, methods use a variety of algorithms and subcomponents to reach a final ordering.
-However, most use a common framework consisting of three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
 <!-- -> add supplementary information -->  We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extendability.
@@
 
+# Online Methods
+
 # Acknowledgements

### 2017-11-22 analysis/paper/paper.Rmd
added google doc support

@@
 # Introduction
-<!-- Three paragraph introduction, typical for Nat Biotech papers -->
-<!-- General into about single-cell & trajectories -->
-Single-cell -omics technologies now make it possible to more accuratly model biological systems than ever before.
+
+Single-cell[a] -omics technologies now make it possible to more accurately model biological systems than ever before.
 One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
 When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation<!-- refs -->.
-
-<!-- Discussion of TI, history, different methods, brief future perspectives -->
-Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
-Initially most trajectory inference was focused on linear trajectories, with one start and one end point.
-However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
-Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes hapening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Because they offer an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
 
@@
 ```
-`r ref("sfig", "methods_over_time")` New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
 
-<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
+`r ref("sfig", "methods_over_time", anchor=TRUE)` New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
+
+Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
+Initially most trajectory inference was focused on linear trajectories, with one start and one end point.
+However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
+Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes hapening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+
+Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
 These components are frequently interchangeable.
 Depending on the way the cells are ordered, some prior information about the dynamic process can also be required, or optional.
-<!---- TODO --->
 
-<!-- Context and general overview of our evaluation -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
+Given[d] this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
@@
 
-<!-- I would move some stuff from this paragraph to the supplementary if necessary -->
-We gathered a total of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, based on free availability and being able to be used within a programming context.
+[e]
+We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface.
 We characterized the methods from both a computational and a user perspective.
-<!-- -> add supplementary information -->  We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extendability.
+From a computational perspective, each method uses different interchangeable components We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extensibility.
 We first contacted all the authors, to allow them to improve their method before publishing our evaluation.
@@
 Echte data. Evaluatie (biologische relevantie van de methode)
-
-```{r get-data}
-```
 
@@
 # References
+
+[a]General info about single-cell & trajectories
+[b]History of TI, different methods
+[c]Common framework of TI, different components, naturally progressing towards the need for an evaluation
+[d]Context and general overview of our evaluation
+[e]I would move some stuff from this paragraph to supplementary if possible

### 2017-11-22 analysis/paper/paper_base.Rmd
added google doc support

--- 2017-11-21 analysis/paper/paper.Rmd
@@
 library(tidyverse)
+library(cowplot)
 
-figs <- list()
-sfigs <- list()
+# refering to figures & supplementary figures
+refs <- tibble(ref_id = character(), name = character(), ref_type = character())
+ref <- function(ref_type, ref_id, suffix="", anchor=FALSE) {
+  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
+    refs <<- refs %>% bind_rows(tibble(
+      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
+      ref_type = ref_type,
+      ref_id = ref_id
+    ))
+  }
+  ref <- refs %>% filter(ref_id == !!ref_id) %>% 
+    pull(name) %>% 
+    {glue::glue("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
+
+  if (anchor) ref <- glue::glue("{ref}<a name='{ref_type}_{ref_id}'></a>")
+  ref
+}
+create_names <- list(
+  sfig = function(i) {glue::glue("Supplementary Figure {i}")},
+  fig = function(i) {glue::glue("Figure {i}")}
+)
 ```
-
-# Introduction
-<!-- Three paragraph introduction, typical for Nat Biotech papers -->
-<!-- General into about single-cell & trajectories -->
-Single-cell -omics technologies now make it possible to more accuratly model biological systems than ever before.
-One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
-When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation <!-- refs -->.
-
-<!-- Discussion of TI, history, different methods, brief future perspectives -->
-Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month.
-At first, most trajectory inference was done for linear trajectories, but several methods which can detect complex branching or cyclical behaviour
-
-<!-- Context and general overview of our evaluation -->
-Given this plethora of available trajectory inference methods,
-
-Trajectories can then be used to delineate the different stages in the dynamic process and their associated gene signatures <!-- REF -->,
-
-# Methods
-
-# Results
-## Evaluation workflow
-### Method characterization
-```{r}
-# Load method characteristics
-method_df <- readRDS(derived_file("method_df.rds", experiment_name="method_characteristics"))
-```
-
-<!-- I would move some stuff from this paragraph to the supplementary if necessary -->
-We gathered a total of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, based on free availability and being able to be used within a programming context.
-We characterized the methods from both a computational and a user perspective.
-Computationally, methods use a variety of algorithms and subcomponents to reach a final ordering.
-However, most use a common framework consisting of three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
-<!-- -> add supplementary information -->  We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extendability.
-We first contacted all the authors, to allow them to improve their method before publishing our evaluation.
-While most methods lack at several metrics, code quality tends to increase over time.
-
-### Data sources
-We used three different data sources
-
-### Score metrics and parameter optimisation
-
-## Overall method comparison
-*Points to maybe raise*:
-- Some methods can handle different trajectories which are not directly described in the main paper: eg.
-circular trajectories (for a lot of principal curves algorithms using `periodic.lowess`), trees (for a lot of bifurcation methods)
-
-## Evaluation on synthetic data
-Initial evaluation..
-Robustness..
-Gegenereerde synthetische data. Parameter tuning voor echte data + initiële evaluatie (tot wat is de methode in staat) + robustness etc testing
-
-## Evaluation on real data
-Biological relevance..
-Echte data. Evaluatie (biologische relevantie van de methode)
-
-```{r get-data}
-```
-
-# Discussion
-
-# Conclusion
-
-# Acknowledgements
-
-# Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-devtools::session_info()
-```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at? You may need to change the path value
-# if your Rmd is not in analysis/paper/
-git2r::repository(".")
-```
-
-# References

### 2017-11-23 analysis/paper/paper.Rmd
paper update

@@
 refs <- tibble(ref_id = character(), name = character(), ref_type = character())
-ref <- function(ref_type, ref_id, suffix="") {
+ref <- function(ref_type, ref_id, suffix="", anchor=FALSE) {
   if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
@@
   }
-  refs %>% filter(ref_id == !!ref_id) %>% pull(name) %>% paste0("**", ., suffix, "**")
+  ref <- refs %>% filter(ref_id == !!ref_id) %>% 
+    pull(name) %>% 
+    {glue::glue("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
+
+  if (anchor) ref <- glue::glue("{ref}<a name='{ref_type}_{ref_id}'></a>")
+  ref
 }
@@
 ```
-
 # Introduction
@@
 However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
-Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes hapening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
@@
 We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface.
-We characterized the methods from both a computational and a user perspective.
-From a computational perspective, each method uses different interchangeable components We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extensibility.
+We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures.
+From a computational perspective, each method uses different interchangeable components We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extensibility (`r ref("fig", "overview")`).
 We first contacted all the authors, to allow them to improve their method before publishing our evaluation.
@@
 
+```{r, fig.width = 12, fig.height = 9}
+method_overview <- readRDS(figure_file("method_overview.rds", experiment_name="method_characteristics"))
+
+print(method_overview)
+```
+
+`r ref("fig", "overview", anchor=TRUE)` Evaluation overview and method overview.
+
 ### Data sources
-We used three different data sources
 

### 2017-11-23 analysis/paper/paper.Rmd
paper update

@@
 author:
-  - Robrecht Cannoodt*
-  - Wouter Saelens*
-  - Helena Todorov
-  - Yvan Saeys
+  - Robrecht Cannoodt* ¹ ² ³
+  - Wouter Saelens* ¹ ²
+  - Helena Todorov¹ ²
+  - Yvan Saeys¹ ²
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 \* Equal contribution  
+¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium  
+² Department of ???, Ghent University, Ghent, Belgium  
+³ Center for Medical Genetics, Ghent University
 
@@
 We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface.
-We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures.
-From a computational perspective, each method uses different interchangeable components We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extensibility (`r ref("fig", "overview")`).
+We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures (`r ref("fig", "method_characteristics")`).
+The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms.
+Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?.
+While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it easier for other developers to adapt the method and extend it further.
+We therefore looked at the implementation of each method, and assessed their implementation quality using a transparent scoring scheme sNOTE.
 We first contacted all the authors, to allow them to improve their method before publishing our evaluation.
-While most methods lack at several metrics, code quality tends to increase over time.
+While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), code quality tends to increase over time sFIG.
+However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).
 
 ```{r, fig.width = 12, fig.height = 9}
-method_overview <- readRDS(figure_file("method_overview.rds", experiment_name="method_characteristics"))
+method_characteristics <- readRDS(figure_file("method_characteristics.rds", experiment_name="method_characteristics"))
 
-print(method_overview)
+print(method_characteristics)
 ```
 
-`r ref("fig", "overview", anchor=TRUE)` Evaluation overview and method overview.
+`r ref("fig", "method_characteristics", anchor=TRUE)` Evaluation overview and method overview.
 

### 2017-11-23 analysis/paper/paper.Rmd
it's not that I don't like glue::glue, but it's one dependency less

--- 2017-11-23 analysis/paper/paper.Rmd
@@
     pull(name) %>% 
-    {glue::glue("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
+    {pritt("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
 
-  if (anchor) ref <- glue::glue("{ref}<a name='{ref_type}_{ref_id}'></a>")
+  if (anchor) ref <- pritt("{ref}<a name='{ref_type}_{ref_id}'></a>")
   ref
@@
 create_names <- list(
-  sfig = function(i) {glue::glue("Supplementary Figure {i}")},
-  fig = function(i) {glue::glue("Figure {i}")}
+  sfig = function(i) {pritt("Supplementary Figure {i}")},
+  fig = function(i) {pritt("Figure {i}")}
 )

### 2017-11-23 scripts/update_paper_from_google_drive.R
it's not that I don't like glue::glue, but it's one dependency less

@@ -3,7 +3,7 @@ library(tidyverse)
 library(rmarkdown)
 
 drive <- drive_download(as_id("1BCCaP21N2PXfzhj9H09yEpZz9lLY2HXd_LxTSsJ_wro"), type="text/plain", overwrite=TRUE, path = tempfile())
-system(glue::glue("sed -i '1s/^.//' {drive$local_path}")) # remove first character, because this is some strang eunicode character
-system(glue::glue("cat analysis/paper/paper_base.Rmd {drive$local_path} > analysis/paper/paper.Rmd"))
+system(pritt("sed -i '1s/^.//' {drive$local_path}")) # remove first character, because this is some strang eunicode character
+system(pritt("cat analysis/paper/paper_base.Rmd {drive$local_path} > analysis/paper/paper.Rmd"))
 
 render("analysis/paper/paper.Rmd", "html_document")

### 2017-11-23 analysis/paper/paper.Rmd

--- 2017-11-23 analysis/paper/paper.Rmd
@@
 
-`r ref("fig", "method_characteristics", anchor=TRUE)` Evaluation overview and method overview.
+`r ref("fig", "method_characteristics", anchor=TRUE)` FIrst characterization of each method.
 

### 2017-11-24 analysis/paper/paper.Rmd
add back the exp_id parameter!!!!!

@@
 ```{r, fig.width = 15, fig.height = 10}
-n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_name="method_characteristics"))
-trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_name="method_characteristics"))
+n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", exp_id="method_characteristics"))
+trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", exp_id="method_characteristics"))
 
@@
 In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods.
-We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from current knowledge.
-We include a rigorous parameter optimisation.
+We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
+We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
 To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
-
 # Results
@@
 # Load method characteristics
-method_df <- readRDS(derived_file("method_df.rds", experiment_name="method_characteristics"))
+method_df <- readRDS(derived_file("method_df.rds", exp_id="method_characteristics"))
 ```
@@
 ```{r, fig.width = 12, fig.height = 9}
-method_characteristics <- readRDS(figure_file("method_characteristics.rds", experiment_name="method_characteristics"))
+method_characteristics <- readRDS(figure_file("method_characteristics.rds", exp_id="method_characteristics"))
 
@@
 ## Overall method comparison
-*Points to maybe raise*:
 - Some methods can handle different trajectories which are not directly described in the main paper: eg.

### 2017-11-24 analysis/paper/paper.Rmd
update paper

@@
 ```{r, fig.width = 15, fig.height = 10}
-n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", exp_id="method_characteristics"))
-trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", exp_id="method_characteristics"))
+n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
+trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_id="method_characteristics"))
 
@@
 # Load method characteristics
-method_df <- readRDS(derived_file("method_df.rds", exp_id="method_characteristics"))
+method_df <- readRDS(derived_file("method_df.rds", experiment_id="method_characteristics"))
 ```
 
-[e]
 We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface.
@@
 Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?.
-While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it easier for other developers to adapt the method and extend it further.
+While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
 We therefore looked at the implementation of each method, and assessed their implementation quality using a transparent scoring scheme sNOTE.
-We first contacted all the authors, to allow them to improve their method before publishing our evaluation.
-While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), code quality tends to increase over time sFIG.
+After initial scoring we contacted all the authors, to allow them to improve their method before publishing.
+While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), recent methods tend to have higher code quality sFIG.
 However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).
@@
 ```{r, fig.width = 12, fig.height = 9}
-method_characteristics <- readRDS(figure_file("method_characteristics.rds", exp_id="method_characteristics"))
+method_characteristics <- readRDS(figure_file("method_characteristics.rds", experiment_id="method_characteristics"))
 
@@
 ### Data sources
+![evaluation_overview](../../analysis/figures/evaluation_overview.png)
+`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation
 
+We use three types of datasets for our evaluation
 ### Score metrics and parameter optimisation
@@
 [d]Context and general overview of our evaluation
-[e]I would move some stuff from this paragraph to supplementary if possible

### 2017-11-24 analysis/paper/paper_base.Rmd
recompile paper

@@
   comment = "#>",
-  fig.path = "analysis/figures/"
+  fig.path = ".scratch/"
 )

### 2017-11-28 analysis/paper/paper.Rmd

@@
 Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
-Initially most trajectory inference was focused on linear trajectories, with one start and one end point.
+Initially, most trajectory inference was focused on linear trajectories, with one start and one end point.
 However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
@@
 While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
-We therefore looked at the implementation of each method, and assessed their implementation quality using a transparent scoring scheme sNOTE.
+We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme sNOTE.
 After initial scoring we contacted all the authors, to allow them to improve their method before publishing.
@@
 
-### Data sources
+### Evaluation overview
 ![evaluation_overview](../../analysis/figures/evaluation_overview.png)
@@
 
-We use three types of datasets for our evaluation
+Our evaluation was structured in three main phases.
+In the first phase, we used toy data as a positive control for each method, assessing whether it can detect the correct trajectories in very simplistic data.
+Next, we generated synthetic data from a realistic model of gene regulation, combined with a simulation of the technical variation induced by single-cell RNA sequencing.
+We used this synthetic data in a parameter optimization procedure, where we used cross-validation between datasets to avoid overfitting on particular data set structures.
+Finally, in the third phase, we used the optimized parameters to score the performance of each method on real data.
 ### Score metrics and parameter optimisation
@@
 # Discussion
-
 # Conclusion
@@
 # Online Methods
+
+## Trajectory inference methods
+## Toy data generation
+## Synthetic data generation
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012], but required several additions 
+## Real data processing
+## Normalization
+## Evaluation metrics
+## Parameter optimization
 

### 2017-11-30 analysis/paper/paper.Rmd
remove technologies

@@
 
-Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
+Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
 These components are frequently interchangeable.
-Depending on the way the cells are ordered, some prior information about the dynamic process can also be required, or optional.
+Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method
 
@@
 To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
+
 # Results
+
 ## Evaluation workflow
+
 ### Method characterization
+
 ```{r}
@@
 ## Overall method comparison
-- Some methods can handle different trajectories which are not directly described in the main paper: eg.
-circular trajectories (for a lot of principal curves algorithms using `periodic.lowess`), trees (for a lot of bifurcation methods)
 
@@
 ## Synthetic data generation
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012], but required several additions 
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012], but required several additions.
+
 ## Real data processing

### 2017-11-30 README.Rmd
added running instruction
added method plotting
add travis badge
update readme

@@
 
-# dynalysis
+# An evaluation framework for trajectory inference methods
+
+![travis](https://travis-ci.com/Zouter/dynalysis.svg?token=pzNqJjk3yapU1Hk7sDg8&branch=master)
+
+![overview](analysis/figures/toy_method_plots/grid.png)
 
@@
 ```{r gh-installation, eval = FALSE}
-# install.packages("devtools")
-devtools::install_github("Zouter/dynalysis")
+# install.packages("remotes")
+remotes::install_github("Zouter/dynalysis", dependencies = TRUE, upgrade_dependencies = TRUE)
 ```
+
+### Running
+
+To run locally, install `docker` and run `docker build -t dynalysis .`
 

### 2017-12-02 README.Rmd

@@
 
+### Dependencies
+
+You might have to install a few packages.
+
+Ubuntu/Debian:
+```
+sudo apt-get install libudunits2-dev libgsl-dev libsdl1.2-dev libreadline-dev imagemagick libfftw3-dev libudunits2-dev -y
+sudo apt-get install python2.7-dev python3-dev python3-pip
+pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn
+```
+
+Fedora/CentOS:
+```
+sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip
+pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn
+Rscript -e 'install.packages("udunits2", configure.args =  c(udunits2 = '--with-udunits2-include=/usr/include/udunits2'))'
+```
+
 ### Installation
@@
 ```{r gh-installation, eval = FALSE}
-# install.packages("remotes")
-remotes::install_github("Zouter/dynalysis", dependencies = TRUE, upgrade_dependencies = TRUE)
+install.packages("remotes")
+options(repos = c("https://cran.rstudio.com/", "http://bioconductor.org/packages/release/bioc/", "http://bioconductor.org/packages/release/data/annotation", "http://bioconductor.org/packages/release/data/experiment", "http://bioconductor.org/packages/release/extra"))
+remotes::install_github("Zouter/dynalysis", dependencies = TRUE)
 ```

### 2017-12-05 analysis/analyses/3-score_rules/1-generate-toys.Rmd
reordered analyses
update to toy rules

--- 2017-09-14 analysis/analyses/score_rules/1-generate-toys.Rmd
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup}
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+library(dynalysis)
+library(tidyverse)
+library(dyneval)
 ```
@@
 ```{r}
-library(tidyverse)
-library(dyneval)
+source("analysis/analyses/3-score_rules/generation.R")
+source("analysis/analyses/3-score_rules/perturbation.R")
 
-source("generation.R")
-source("perturbation.R")
+experiment("score_rules")
 ```
@@
 toys_blueprint <- tribble(
-  ~generator_id, ~perturbator_id,
-  "linear", "gs",
-  "linear", "switch_two_cells",
-  "linear", "switch_all_cells",
-  "linear", "join_linear",
-  "linear", "split_linear",
-  "linear", "warp",
-  "linear", "hairy",
-  "linear", "hairy_small",
-  "linear", "hairy_large",
-  "linear", "remove_cells",
-  "linear", "add_distant_edge",
-  "linear", "structure_and_position",
+  ~trajectory_type, ~perturbator_id,
+  "simple_linear", "gs",
+  "simple_linear", "switch_two_cells",
+  "simple_linear", "switch_all_cells",
+  "simple_linear", "join_linear",
+  "simple_linear", "split_linear",
+  "simple_linear", "warp",
+  "simple_linear", "hairy",
+  "simple_linear", "hairy_small",
+  "simple_linear", "hairy_large",
+  "simple_linear", "remove_cells",
+  "simple_linear", "add_distant_edge",
+  "simple_linear", "structure_and_position",
   "bifurcating", "gs",
@@
 ) %>% 
-  bind_rows(tibble(generator_id="linear", perturbator_id=paste0("switch_", percs * 100))) %>%
+  bind_rows(tibble(trajectory_type="simple_linear", perturbator_id=paste0("switch_", percs * 100))) %>%
+  bind_rows(crossing(trajectory_type=trajectory_types, perturbator_id=paste0("change_network_", trajectory_types)) %>% mutate(ncells = 5)) %>% 
   rowwise() %>% 
   mutate(
-  generator=list(get(paste0("generate_", generator_id))),
   perturbator=list(get(paste0("perturb_", perturbator_id)))
@@
   replicate=seq_len(nrow(.))%%nreplicates,
-  toy_category=paste0(generator_id, "-", perturbator_id),
+  toy_category=paste0(trajectory_type, "-", perturbator_id),
   toy_id=paste0(toy_category, "-", replicate)
@@
 toys$gs <- toys %>% split(seq_len(nrow(toys))) %>% parallel::mclapply(function(row) {
-  row$generator[[1]](row$ncells)
+  generate_dataset(row$trajectory_type, row$ncells)
 }, mc.cores=8)
@@
 
-write_rds(toys, "toys.rds")
+write_rds(toys, derived_file("toys.rds"))
 ```
@@
 ```{r}
-toys <- read_rds("toys.rds")
+toys <- read_rds(derived_file("toys.rds"))
 
@@
 
-metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score", "mantel_pval", "ged")
+metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "node_edit_score", "node_edge_edit_score", "mantel_pval", "ged")
 
-results <- toys %>% rowwise() %>% do(compare_gs_toy(.$gs, .$toy, .$gs_grouped))# %>% select(-starts_with("time"))
+library(PRISM)
+
+qsub_run(
+  function(x) {
+    library(tidyverse)
+    parallel::mclapply(seq_len(nrow(toys)), function(row_id) {
+      row <- dynutils::extract_row_to_list(toys, row_id)
+      compare_gs_toy(.$gs, .$toy, .$gs_grouped)
+    })
+  },
+  qsub_config = override_qsub_config(num_cores = 24),
+  qsub_environment = list2env(list(compare_toy, compare_gs_toy, toys)
+)
+
 scores <- results %>% select(-starts_with("time"))
 
-write_rds(scores, "scores.rds")
+write_rds(scores, result_file("scores.rds"))
 ```
@@
   left_join(metric_groups, by="metric_group_id") %>% 
-  write_rds("timings.rds")
+  write_rds(result_file("timings.rds"))
 ```

### 2017-12-05 analysis/analyses/3-score_rules/2-rules.Rmd
update to toy rules

@@
 
-```{r}
+```{r setup}
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+library(dynalysis)
 library(tidyverse)
@@
 ```{r}
-scores <- read_rds("scores.rds")
-toys <- read_rds("toys.rds")
-timings <- read_rds("timings.rds")
+experiment("score_rules")
+
+scores <- read_rds(result_file("scores.rds"))
+toys <- read_rds(derived_file("toys.rds"))
+timings <- read_rds(result_file("timings.rds"))
 
@@
 
-source("rules.R")
+source("analysis/analyses/3-score_rules/rules.R")
 rules_info$check_func <- map(paste0("check_", rules_info$id), get)
@@
 metric_types <- rep("positional", length(unique(scores_summary$score_id))) %>% setNames(unique(scores_summary$score_id))
-metric_types[c("ged", "robbie_network_score", "isomorphic")] <- "structural"
+metric_types[c("ged", "robbie_network_score", "lies_network_score", "isomorphic")] <- "structural"
 

### 2017-12-05 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
score rules

@@
 library(dyneval)
+library(dynmethods)
 
@@
 
-tasks <- generate_toy_datasets(num_replicates = 1, num_cells=300)
+data("toy_tasks")
+tasks <- toy_tasks
 ```
@@
 })
-task <- dynutils::extract_row_to_list(tasks, 2)
+task <- dynutils::extract_row_to_list(tasks, 3)
 pheatmap::pheatmap(task$counts)
@@
 source("~/thesis/projects/dynverse/dynmodular/dimred_wrappers.R")
-space <- task$counts %>% dimred_dp(2, neigen=2) %>% as.data.frame
-# space <- task$counts %>% dimred_mds(2) %>% as.data.frame
+# space <- task$counts %>% dimred_dp(2, neigen=2) %>% as.data.frame
+space <- task$counts %>% dimred_mds(2) %>% as.data.frame
 # space <- task$counts %>% dimred_lle(2) %>% as.data.frame
-space <- space %>% bind_cols(task$cell_grouping %>% slice(match(rownames(task$counts), cell_id)))
+space <- space %>% bind_cols(task$prior_information$grouping_assignment %>% slice(match(rownames(task$counts), cell_id)))
 space %>%  ggplot() + geom_point(aes(Comp1, Comp2, color=group_id))
@@
 
-metric_names <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "robbie_network_score")
+metric_names <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation")
 
-method_names <- c("monocle_ddrtree", "scorpius")
+method_names <- c("scorpius")
 ```
@@
 
-  factory(function() {
-    dyneval:::execute_evaluation(tasks, method, params, metrics = metric_names, timeout = 99999999999999)
-  }
-  )()
+  # factory(function() {
+    dyneval:::execute_evaluation(tasks, method, params, metrics = metric_names, timeout = 100)
+  # }
+  # )()
 })
@@
 ```{r}
-# results <- PRISM::qsub_retrieve(results)
-scores <- purrr::map(results, function(.) {
+scores <- purrr::map(results, function(result) {
   bind_cols(
     tibble(
-      error=list(.$err), 
-      warning=list(.$warning), 
-      model=attr(.[[1]], "extras")$.models
+      model = attr(result, "extras")$.models
     ),
-    attr(.[[1]], "extras")$.summary
-  ) %>% mutate(method_name=.$method_name)
-}) %>% bind_rows() %>% mutate(errored = !map_lgl(error, is.null), method_id=rep(method_names, each=nrow(tasks))) %>% dplyr::select(-contains("method_name1"))
+    attr(result, "extras")$.summary
+  )
+}) %>% bind_rows()
 

### 2017-12-05 README.Rmd
update readme, script and pdfs

@@
 ```
-sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip
+sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip python3-matplotlib-tk
 pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn

### 2017-12-05 analysis/analyses/1-datasets/real/dataset_pancreatic-cell-maturation_zhang.R
Major refactoring due to rcannood/dynutils/commit/ababaecdd886d3898c52a91912119b174f318f85

@@ -59,7 +59,7 @@ settings <- list(
       "α-cell P15", "α-cell P18", 3,
       "α-cell P18", "α-cell P60", 42
     ) %>% mutate(directed = TRUE),
-    ti_type = "linear"
+    trajectory_type = "linear"
   ),
   list(
     id = "real/pancreatic-beta-cell-maturation_zhang",
@@ -72,7 +72,7 @@ settings <- list(
       "β-cell P15", "β-cell P18", 3,
       "β-cell P18", "β-cell P60", 42
     ) %>% mutate(directed = TRUE),
-    ti_type = "linear"
+    trajectory_type = "linear"
   )
 )
 
@@ -93,7 +93,7 @@ for (setting in settings) {
   feature_info <- tibble(feature_id = colnames(counts))
 
   datasetpreproc_normalise_filter_wrap_and_save(
-    ti_type = setting$ti_type,
+    trajectory_type = setting$trajectory_type,
     counts = counts,
     cell_ids = cell_ids,
     milestone_ids = milestone_ids,

### 2017-12-05 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
Major refactoring due to rcannood/dynutils/commit/ababaecdd886d3898c52a91912119b174f318f85

--- 2017-11-20 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
@@
 ```{r fig.height=10, fig.width=10}
-scores %>% left_join(tasks %>% select(id, ti_type), by=c("task_id"="id")) %>% 
+scores %>% left_join(tasks %>% select(id, trajectory_type), by=c("task_id"="id")) %>% 
   dplyr::select(-starts_with("time")) %>%
-  gather(score_id, score, -method_name, -method_short_name, -method_id, -task_id, -error, -warning, -model, -ti_type) %>%
+  gather(score_id, score, -method_name, -method_short_name, -method_id, -task_id, -error, -warning, -model, -trajectory_type) %>%
   filter(score_id %in% c("errored", "correlation", "robbie_network_score", "mean_R_nx")) %>% 
   ggplot() + geom_bar(aes(method_name, score, fill=score_id), stat="identity") + 
-  facet_grid(ti_type~score_id) + coord_flip()
+  facet_grid(trajectory_type~score_id) + coord_flip()
 ```

### 2017-12-05 analysis/analyses/1-datasets/real/dataset_aging-hsc_kowalczyk.R

@@ -46,7 +46,7 @@ settings <- lapply(c("old", "young"), function(age) {
       "LT-HSC", "ST-HSC",
       "ST-HSC", "MPP"
     ) %>% mutate(length = 1, directed = TRUE),
-    ti_type = "linear",
+    trajectory_type = "linear",
     age = age
   )
 })
@@ -68,7 +68,7 @@ for (setting in settings) {
   feature_info <- tibble(feature_id = colnames(expression))
 
   datasetpreproc_normalise_filter_wrap_and_save(
-    ti_type = setting$ti_type,
+    trajectory_type = setting$trajectory_type,
     counts = 2^expression - 1, # todo: fix this
     cell_ids = cell_ids,
     milestone_ids = milestone_ids,

### 2017-12-06 analysis/paper/paper.Rmd
traj types

@@
 However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
-Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Some methods can already handle more complex, cyclical and branching behaviour [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
-Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation.
-These components are frequently interchangeable.
-Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method
+Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
+Furthermore, with the ongoing strong decreases in single-cell -omics technologies, it becomes
 
@@
 
-```{r, fig.width = 12, fig.height = 9}
-method_characteristics <- readRDS(figure_file("method_characteristics.rds", experiment_id="method_characteristics"))
-
-print(method_characteristics)
-```
+![](`r figure_file("figure_methods.svg", experiment_id="method_characteristics")`)
 
@@
 ### Evaluation overview
+
 ![evaluation_overview](../../analysis/figures/evaluation_overview.png)
@@
 Finally, in the third phase, we used the optimized parameters to score the performance of each method on real data.
+We split the real datasets in two groups based on the type of reference trajectory available.
+The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.
 ### Score metrics and parameter optimisation
@@
 ## Synthetic data generation
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012], but required several additions.
+
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+At every step, great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
+
+### Network generation
+
+One of the main process involved in cellular dynamic processes is through gene regulation, where regulatory cascades and feedback loops lead to .
+The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
+One driver of bifurcations seem to be mutual antognism, where two genes  or gene modules strongly inhibit eachother, forcing one of the two to become inactive.
+The choice of which gene(s) survive can be both due to stochasticity [ref] or driven by environmental cues.Of Modules?
+[Th17]
 

### 2017-12-06 analysis/analyses/4-method_characterization/characterize.R
commit svg & Rmd & R

@@ -65,7 +65,7 @@ trajectory_components <- method_df_evaluated %>%
     binary_tree = fork & n_forks > 1,
     single_multifurcation = fork & n_fork_paths > 2,
     non_binary_tree = binary_tree & single_multifurcation,
-    cycle = loop,
+    single_cycle = loop,
     simple_graph = binary_tree & single_multifurcation & convergence
   )
 
@@ -77,17 +77,18 @@ method_df_evaluated <- read_rds(derived_file("method_df_evaluated.rds"))
 
 # Comparing number of citations, qc score, and time -------------------------------------
 g1 <- ggplot(method_df_evaluated, aes(date, qc_score)) +
-  geom_smooth() +
+  geom_smooth(method="lm") +
   geom_point() +
-  ggrepel::geom_label_repel(aes(label = name), nudge_y = .25) +
+  ggrepel::geom_label_repel(aes(label = name)) +
   cowplot::theme_cowplot() +
   scale_y_continuous(breaks = c(0, 2, 4, 6, 8), limits = c(0, 8)) +
   labs(x = "Time", y = "QC score", title = "Code quality score over time")
 g1
+write_rds(g1, figure_file("ncitations_over_time.rds"))
 ggsave(figure_file("time_vs_qcscore.png"), g1, width = 16, height = 8)
 
 g2 <- ggplot(method_df_evaluated, aes(date, Citations+1)) +
-  geom_smooth() +
+  geom_smooth(method="lm") +
   geom_point() +
   ggrepel::geom_label_repel(aes(label = name), nudge_y = .25) +
   cowplot::theme_cowplot() +
@@ -97,7 +98,7 @@ g2
 ggsave(figure_file("time_vs_citations.png"), g2, width = 16, height = 8)
 
 g3 <- ggplot(method_df_evaluated, aes(Citations+1, qc_score)) +
-  geom_smooth() +
+  geom_smooth(method="lm") +
   geom_point() +
   ggrepel::geom_label_repel(aes(label = name), nudge_y = .25) +
   cowplot::theme_cowplot() +
@@ -246,7 +247,7 @@ add_step <- function(df) {
 }
 
 trajectory_components_step <- add_step(arrange(trajectory_components, date))
-trajectory_types <- c("linear", "single_bifurcation", "binary_tree", "single_multifurcation", "non_binary_tree", "cycle", "simple_graph")
[…]

### 2017-12-06 analysis/analyses/3-score_rules/1-generate-toys.Rmd

@@
   qsub_environment = list2env(list(compare_toy, compare_gs_toy, toys)
-)
+))
 

### 2017-12-07 README.Rmd
update readme

--- 2017-12-02 README.Rmd
@@
 ```
-sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip
+sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip python3-matplotlib-tk openblas-devel lapack-devel
 pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn

### 2017-12-13 README.Rmd
update readme to use devtools again

@@
 ```{r gh-installation, eval = FALSE}
-install.packages("remotes")
-options(repos = c("https://cran.rstudio.com/", "http://bioconductor.org/packages/release/bioc/", "http://bioconductor.org/packages/release/data/annotation", "http://bioconductor.org/packages/release/data/experiment", "http://bioconductor.org/packages/release/extra"))
-remotes::install_github("Zouter/dynalysis", dependencies = TRUE)
+install.packages("devtools")
+setRepositories(ind = 1:2)
+devtools::install_github("Zouter/dynalysis", dependencies = T, upgrade = T)
 ```

### 2017-12-15 analysis/paper/paper.Rmd

@@
 ## Trajectory inference methods
+
 ## Toy data generation
+
 ## Synthetic data generation
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcations seem to be mutual antognism, where two genes  or gene modules strongly inhibit eachother, forcing one of the two to become inactive.
-The choice of which gene(s) survive can be both due to stochasticity [ref] or driven by environmental cues.Of Modules?
+One driver of bifurcations seem to be mutual antagonism, where two genes [@XuRegulationbifurcatingcell2015] or gene modules strongly inhibit each other, forcing one of the two to become inactive [@GrafForcingcellschange2009].
+Which of these two Of Modules?
 [Th17]

### 2018-01-03 analysis/paper/paper_base.Rmd
updated paper

@@
 author:
+  - Wouter Saelens* ¹ ²
   - Robrecht Cannoodt* ¹ ² ³
-  - Wouter Saelens* ¹ ²
-  - Helena Todorov¹ ²
+  - Helena Todorov¹ ² ⁴ ⁵
   - Yvan Saeys¹ ²
@@
 We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
-Our evaluation pipeline is fully reproducible, can be easily extended and is available at www.github.com/dynverse.
+Our evaluation pipeline is fully reproducible, can be easily extended, and is available at www.github.com/dynverse.
 keywords: |
@@
 \* Equal contribution  
-¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium  
-² Department of ???, Ghent University, Ghent, Belgium  
-³ Center for Medical Genetics, Ghent University
+¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
+² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
+³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
+⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
+⁵ Université Claude Bernard Lyon 1, Lyon, France.
 

### 2018-01-03 analysis/paper/paper.Rmd
updated paper

@@
 \* Equal contribution  
-¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium  
-² Department of ???, Ghent University, Ghent, Belgium  
-³ Center for Medical Genetics, Ghent University
+¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
+² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
+³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
+⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.
+⁵ Université Claude Bernard Lyon 1, Lyon, France.
 
@@
 ```
+# Title
+Not all trajectory inference methods are created equal
+Trajectory inference: work in progress
+Trajectory inference: we’re not quite there yet
+Paving the way for trajectories
+Benchmarking trajecotry inference methods
+
 # Introduction
@@
 In the first phase, we used toy data as a positive control for each method, assessing whether it can detect the correct trajectories in very simplistic data.
-Next, we generated synthetic data from a realistic model of gene regulation, combined with a simulation of the technical variation induced by single-cell RNA sequencing.
+Next, we generated synthetic data from a model of gene regulation, combined with a simulation of the technical variation introduced by single-cell RNA sequencing.
 We used this synthetic data in a parameter optimization procedure, where we used cross-validation between datasets to avoid overfitting on particular data set structures.
@@
 ## Real data processing
+
+We gathered
+
 ## Normalization

### 2018-01-16 analysis/analyses/3-score_rules/flip_score_characterisation.Rmd
added flip score characterisation

---
title: "Flip score characterisation"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(dynmethods)
library(dyneval)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
models <- eval(formals(dyntoy:::generate_toy_milestone_network)$model)
models <- models[models != "BA"]
networks_synthetic <- map(models, dyntoy:::generate_toy_milestone_network) %>% tibble(milestone_network = ., model = models)
```

Add SCORPIUS predictions
```{r}
descriptions <- list(description_scorpius(), description_dpt())
data <- dyntoy::generate_toy_datasets(c("linear"), num_replicates=1, num_cells=30, num_genes=10) %>% dynutils::extract_row_to_list(1)

results <- map(descriptions, ~.$run_fun(data$counts))

networks_methods <- tibble(model=map_chr(descriptions, "name"), milestone_network=map(results, "milestone_network"))

```

```{r}
networks <- bind_rows(networks_synthetic, networks_methods)
design <- crossing(networks, networks)
```

```{r fig.height=7}
scores <- design %>% as.list() %>% pmap(function(milestone_network, milestone_network1, model, model1, ...) {
  # print(model)
  # print(model1)

  start <- Sys.time()
  score <- dyneval:::calculate_edge_flip(milestone_network, milestone_network1)
  time <- Sys.time() - start
  tibble(score=score, time=time)
}) %>% bind_rows()
design <- bind_cols(design, scores)

design %>% ggplot() + geom_raster(aes(model, model1, fill=score)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
library(ggraph)
library(tidygraph)

networks$plot <- networks$milestone_network %>% map(function(net) {
  net %>% as_tbl_graph(directed=FALSE) %>%
    ggraph() +
      geom_edge_fan() + geom_node_point() + ggraph::theme_graph(plot_margin = margin(0, 0, 0, 0))
})

design %>%
  group_by(model) %>%
  arrange(-score) %>%
  left_join(networks %>% select(-milestone_network), by=c("model1"="model")) %>%
  summarise(plots = list(map2(plot, score, ~.x + ggtitle(round(.y, 2))))) %>%
  pull(plots) %>% map(~cowplot::plot_grid(plotlist=.))
```

### 2018-01-17 analysis/analyses/4-method_characterization/2_quality_control.Rmd
update to new qc

---
title: "Method quality control"
output:
  html_document:
    df_print: paged
---

```{r message=F}
library(tidyverse)
library(dynalysis)
library(cowplot)

experiment("method_characteristics")

method_qc <- readRDS(derived_file("method_qc.rds"))
checks <- readRDS(derived_file("checks.rds"))

source("0_common.R")
```

```{r}
category_gradients_white <- map(category_colors, function(color) {
  n <- 1000
  ramp <- shades::gradient(c("white", color), n)

  function(x, min=0, max=1) {
    x[x == 0] <- 0.0000000001
    ramp[round((x - min)/(max-min)*(n-1))+1]
  }
})
applications <- c("developer_friendly", "user_friendly", "good_science")
```

Calculate the overall qc scores for each method
```{r}
final_scores <- method_qc %>%
  group_by(method_id) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  arrange(-score)

# set the ordering of method_ids to this ordering
method_qc$method_id <- factor(method_qc$method_id, levels=final_scores$method_id)
final_scores$method_id <- factor(final_scores$method_id, levels=final_scores$method_id)

method_ordering_plot <- final_scores %>%
  ggplot(aes(method_id, score)) +
    geom_bar(aes(fill=score), stat="identity") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
    viridis::scale_fill_viridis(option="A", limits=c(0, 1))

method_ordering_plot
```

Orderin by category or application
```{r}
method_qc %>%
  group_by(method_id, category) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  ggplot(aes(method_id, category)) + geom_raster(aes(fill=score))

method_qc %>%
  gather(application, application_applicable, !!applications) %>%
  filter(application_applicable) %>%
  group_by(method_id, application) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  ggplot(aes(method_id, application)) + geom_raster(aes(fill=score))
```

```{r, fig.width=10, fig.height=12}
question_position <- method_qc %>%
  filter(method_id == "SCORPIUS") %>%
  arrange(question_id) %>%
  group_by(question_id) %>%
  summarise(
    question_width = sum(score * weight),
    name = first(name)
  ) %>%
  mutate(
    question_end = cumsum(question_width),
    question_start = lag(question_end, 1, default=0),
    question_mid = question_start + (question_end - question_start)/2
  )

check_position <- method_qc %>%
  filter(method_id == "SCORPIUS") %>%
  arrange(question_id, check_id) %>%
  group_by(check_id) %>%
  summarise(
    check_width = score * weight,
    question_id = first(question_id)
  ) %>%
  left_join(question_position, by="question_id") %>%
  group_by(question_id) %>%
  mutate(
    check_start = question_start + lag(cumsum(check_width), default=0),
    check_end = question_start + cumsum(check_width)
  )

check_heatmap_plot <- method_qc %>%
  left_join(check_position, by="check_id") %>%
  mutate(method_id = factor(method_id)) %>%
  mutate(answer = ifelse(is.na(answer), 0, answer)) %>%
  ggplot() +
    geom_rect(aes(xmax=check_start, xmin=check_end, ymin=as.integer(method_id), ymax=as.integer(method_id) + 1, fill=category, alpha=answer)) +
    geom_vline(aes(xintercept=check_end), data=check_position, color="#444444", alpha=0.25) +
    geom_vline(aes(xintercept=question_end), data=question_position, color="#444444") +
    scale_x_reverse("Check", breaks = question_position$question_mid, labels = question_position$name, expand = c(0, 0)) +
    scale_y_continuous("Method", breaks = seq_along(levels(method_qc$method_id))+0.5, labels = levels(method_qc$method_id), expand = c(0, 0)) +
    scale_fill_manual(values=category_colors) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

check_heatmap_plot
```

Combined plot
```{r, fig.width=10, fig.height=12}
empty_x <- theme(
  axis.title.x=element_blank(),
  axis.text.x=element_blank(),
  axis.ticks.x=element_blank()
)

cowplot::plot_grid(method_ordering_plot + empty_x, check_heatmap_plot, nrow=2, align = "v", axis = "rl", rel_heights = c(0.2, 0.9))

```

## Order checks

Order by breakpoint (when ordering all methods by score)
```{r}
breakpoint <- function(x) {
  map_dbl(0:length(x), function(i) {
    sum(1-x[seq(1, i, length.out=i)]) +
      sum(x[seq(i, length(x), length.out=length(x)-i)])
  }) %>% which.min()
}
check_breakpoints <- method_qc %>%
  reshape2::acast(check_id~method_id, value.var="answer") %>%
  {
    tibble(
      breakpoint=apply(., 1, breakpoint),
      check_id = as.integer(rownames(.)),
      n_bad_methods = apply(., 1, sum)
    )
  } %>% 
  left_join(checks, by="check_id")
```

```{r fig.width=20}
method_qc %>% 
  mutate(answer = round(answer, 2)) %>% 
  group_by(check_id) %>% 
  count(answer) %>% 
  mutate(check_score = sum(answer * n)) %>% 
  ungroup() %>% 
  left_join(checks, by="check_id") %>% 
  arrange(-check_score) %>% 
  mutate(check_id = factor(check_id, levels=unique(check_id))) %>% 
  ggplot() +
    geom_bar(aes(check_id, n, fill=answer), stat="identity") +
    coord_flip() +
    scale_y_reverse(expand=c(0, 0)) +
    scale_x_discrete(labels=setNames(method_qc$scoring, method_qc$check_id)) +
    scale_fill_distiller(palette="Blues", direction = 1)
```

```{r}
question_info <- method_qc %>% 
  group_by(question_id) %>% 
  summarise(
    name = first(name),
    category = first(category)
  )

question_ordering_plotdata <- method_qc %>%
  group_by(method_id, question_id) %>%
  summarise(
    qc_score=sum(answer * score * weight)/sum(score * weight)
  ) %>%
  ungroup() %>% 
  group_by(question_id) %>% 
  count(qc_score) %>% 
  mutate(question_ordering = sum((qc_score < 1) * n)) %>% 
  ungroup() %>% 
  arrange(question_ordering) %>% 
  left_join(question_info, by="question_id") %>% 
  mutate(question_id = factor(question_id, levels=unique(question_id)))

question_ordering_plot1 <- question_ordering_plotdata %>% 
  ggplot() +
    geom_bar(aes(question_id, n, fill=qc_score), stat="identity") +
    scale_fill_distiller(palette="Blues", direction = -1)
question_ordering_plot1

question_ordering_plotdata2 <- question_ordering_plotdata %>% 
  mutate(category = factor(category, levels=categories)) %>% 
  arrange(-as.numeric(category), question_ordering) %>% 
  mutate(question_id = factor(question_id, levels=as.character(unique(question_id))))

question_ordering_plotdata2_category <- question_ordering_plotdata2 %>% 
  group_by(category, question_id) %>% 
  summarise() %>% 
  ungroup() %>% 
  mutate(position = n()-seq_len(n())) %>% 
  group_by(category) %>% 
  summarise(start = min(position)+0.5, end=max(position)+1.5)

question_ordering_plot2 <- question_ordering_plotdata2 %>% 
  arrange(qc_score) %>% 
  ggplot() +
    geom_bar(aes(question_id, n, fill=category_colors[category], alpha=qc_score), stat="identity") +
    geom_vline(aes(xintercept=start), data=question_ordering_plotdata2_category) +
    scale_fill_identity() +
    scale_alpha_continuous(range=c(1, 0.05), name = "Quality control score", guide = guide_legend())
question_ordering_plot2

category_legend <- question_ordering_plotdata2 %>% 
  group_by(category, question_id) %>% 
  summarise() %>% 
  ungroup() %>% 
  mutate(position = seq_len(n())) %>% 
  group_by(category) %>% 
  summarise(start = -min(position), end=-max(position)-1) %>% 
  ggplot() +
    geom_rect(aes(xmin=0, xmax=1, ymin=start, ymax=end), fill="white") +
    geom_text(aes(1, start + (end-start)/2, label=category_labels[category], color=category), hjust=1, size=5) +
    geom_segment(aes(x=1.1, xend=1.1, y=start-0.2, yend=end+0.2, color=category)) +
    scale_color_manual(values=category_colors) +
    scale_y_continuous(expand=c(0, 0)) +
    theme_void() +
    theme(legend.position="none")
category_legend

n_methods <- length(unique(method_qc$method_id))

question_ordering_plot2 <- question_ordering_plot2 +
    coord_flip() +
    scale_y_continuous(expand=c(0, 0), name = "# methods", c(seq(0, n_methods, 5), n_methods)) +
    scale_x_discrete(labels=setNames(question_ordering_plotdata$name, question_ordering_plotdata$question_id), name="") +
    theme(legend.position="top")

method_qc_ordering_plot <- cowplot::plot_grid(category_legend, question_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
method_qc_ordering_plot
saveRDS(method_qc_ordering_plot, figure_file("method_qc_ordering_plot.rds"))
```

### 2018-01-17 analysis/analyses/4-method_characterization/1_process_methods.R

@@ -38,7 +38,7 @@ method_df <- method_df %>%
 
 # --------------------
 method_df_evaluated <- method_df %>%
-  filter(!is.na(qc_score))
+  filter(wrapper == "Done")
 
 # Trajectory components --------------------------
 method_df_evaluated <- method_df_evaluated %>%

### 2018-01-17 analysis/analyses/4-method_characterization/1_process_qc.R
update to new qc

@@ -3,17 +3,18 @@
 library(tidyverse)
 library(googlesheets)
 library(dynalysis)
-library(cowplot)
 
+experiment("method_characteristics")
 
-# Process qc -------------------------
+# Download qc & initial processing
 method_qc_sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE") %>%
   gs_read(ws = "QualityControl")
 
-categories <- unique(method_qc_sheet$category) %>% {.[!is.na(.)]}
+categories <- unique(method_qc_sheet$category) %>% discard(is.na)
 applications <- c("developer_friendly", "user_friendly", "good_science")
 
 method_qc_converted <- method_qc_sheet %>%
+  filter(active) %>%
   tidyr::fill(question_id) %>%
   mutate(question_id = factor(question_id, levels = method_qc_sheet$question_id %>% unique() %>% keep(~!is.na(.)))) %>%
   group_by(question_id) %>%
@@ -22,11 +23,16 @@ method_qc_converted <- method_qc_sheet %>%
   ungroup()
 
 method_cols <- colnames(method_qc_converted)[seq(which(colnames(method_qc_converted) == "scoring")+1, ncol(method_qc_converted))]
+
+method_qc_converted <- method_qc_converted %>% mutate(check_id = row_number())
+
+# create checks
+checks <- method_qc_converted[, colnames(method_qc_converted) %>% keep(~!(. %in% method_cols))]
+
 method_qc_molten <- method_qc_converted %>%
-  mutate(check_id = row_number()) %>%
   gather("method_id", "answer", !!method_cols)
 
-# process answer
+# process answer, NA coercion warnings are normal here due to check of
 method_qc_processed <- method_qc_molten %>%
   mutate(
     answer = ifelse(is.na(answer), " ", answer),
@@ -39,110 +45,22 @@ method_qc_processed <- method_qc_molten %>%
         answer_first_char == "?",
         NA,
         ifelse(
-          !is.na(as.numeric(answer_first_char)),
-          as.numeric(answer_first_char),
+          suppressWarnings({!is.na(as.numeric(answer_first_char))}),
+          suppressWarnings(as.numeric(answer_first_char)),
           0
         )
       ))
   )
 
-# are all questions answerred?
+# are all questions answered?
[…]

### 2018-01-17 analysis/paper/paper_base.Rmd
update paper & remove scratch

@@
 ---
-title: "An evaluation framework for trajectory inference methods"
+title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
 author:

### 2018-01-17 analysis/paper/paper.Rmd
update paper & remove scratch

@@
 ```
-# Title
+<!---
 Not all trajectory inference methods are created equal
@@
 Paving the way for trajectories
-Benchmarking trajecotry inference methods
+Benchmarking trajectory inference methods
+Bringing single-cell trajectory inference methods to the next level
+Resolving disarray amongst trajectory inference methods with a comprehensive evaluation
+A comparison of trajectory inference methods
+Trajectory inference 2.0
+We’re on the trajectory to better trajectory inference methods
+On the road to more accurate and robust single-cell trajectory inference methods
+Towards more accurate and robust single-cell trajectory inference methods
+On the trajectory to more accurate,robust and trajectory-like single-cell trajectory inference method trajectories
+A comparison of single-cell trajectory inference methods: towards more accurate and robust tools
+--->
 
@@
 # Results
-
 ## Evaluation workflow
-
 ### Method characterization

### 2018-01-17 analysis/paper/paper.Rmd
allow supplementary notes, update figures & paper

@@
 create_names <- list(
-  sfig = function(i) {pritt("Supplementary Figure {i}")},
-  fig = function(i) {pritt("Figure {i}")}
+  sfig = function(i) pritt("Supplementary Figure {i}"),
+  fig = function(i) pritt("Figure {i}"),
+  snote = function(i) pritt("Supplementary Note {i}")
 )
 ```
-<!---
+<!----
+Tips for working in this document:
+* Go to tools -> perferences and remove Use smart quotes
+* Add enters after every paragraph (this is markdown!)
+
+TITLES: 
 Not all trajectory inference methods are created equal
@@
 
-Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
 Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
@@
 # Results
-## Evaluation workflow
-### Method characterization
+## Method characterization
 
@@
 
-We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface.
-We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures (`r ref("fig", "method_characteristics")`).
-The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms.
-Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?.
-While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
-We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme sNOTE.
-After initial scoring we contacted all the authors, to allow them to improve their method before publishing.
-While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), recent methods tend to have higher code quality sFIG.
-However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).
+We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(method_df$wrapper == "Done")` for evaluation, primarily based on their free availability and the presence of a programming interface.
+We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and underlying algorithm(s) (`r ref("fig", "method_characteristics")`).
+
+At the level of trajectory structures, most current methods can handle one or more splits in the trajectory (`r ref("fig", "method_characteristics", "a")`).
+However, only one method (SLICER) can currently handle more complex graph structures which can include loops.
+
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("snote", "qc_scoring")`).
+The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).
+
+After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
+Most methods reach a decent score, fulfilling most of the quality criteria (`r ref("fig", "method_characteristics", "b")`) recent methods tend to have higher code quality sFIG.
+Several aspects are lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation are problematic areas in particular (`r ref("sfig", "method_qc_check_plot")` and `r ref("fig", "method_characteristics", "c")`).
+
+The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
 
@@
 
-`r ref("fig", "method_characteristics", anchor=TRUE)` FIrst characterization of each method.
+`r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.
+
+```{r, fig.width = 10, fig.height = 13}
+readRDS(figure_file("method_qc_check_plot.rds"))
+```
+`r ref("sfig", "method_qc_check_plot", anchor=TRUE)`
+
+```{r, fig.width = 15, fig.height = 7.5}
+readRDS(figure_file("method_components_plot.rds"))
+```
+`r ref("sfig", "method_components", anchor=TRUE)` TI methods frequently share steps
 

### 2018-01-17 analysis/paper/paper_base.Rmd
allow supplementary notes, update figures & paper

@@
 create_names <- list(
-  sfig = function(i) {pritt("Supplementary Figure {i}")},
-  fig = function(i) {pritt("Figure {i}")}
+  sfig = function(i) pritt("Supplementary Figure {i}"),
+  fig = function(i) pritt("Figure {i}"),
+  snote = function(i) pritt("Supplementary Note {i}")
 )

### 2018-01-18 analysis/paper/paper.Rmd
large update to paper generation. Includes supplementary figures at the end. Fix references. Text and figure update

@@
 ⁵ Université Claude Bernard Lyon 1, Lyon, France.  
-
-```{r setup, include=FALSE}
-knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
-  comment = "#>",
-  fig.path = ".scratch/"
-)
-
-library(dynalysis)
-library(tidyverse)
-library(cowplot)
-
-# refering to figures & supplementary figures
-refs <- tibble(ref_id = character(), name = character(), ref_type = character())
-ref <- function(ref_type, ref_id, suffix="", anchor=FALSE) {
-  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
-    refs <<- refs %>% bind_rows(tibble(
-      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
-      ref_type = ref_type,
-      ref_id = ref_id
-    ))
-  }
-  ref <- refs %>% filter(ref_id == !!ref_id) %>% 
-    pull(name) %>% 
-    {pritt("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
-
-  if (anchor) ref <- pritt("{ref}<a name='{ref_type}_{ref_id}'></a>")
-  ref
-}
-create_names <- list(
-  sfig = function(i) pritt("Supplementary Figure {i}"),
-  fig = function(i) pritt("Figure {i}"),
-  snote = function(i) pritt("Supplementary Note {i}")
-)
-```
 <!----
@@
 
+```{r setup, include=FALSE}
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  comment = "#>",
+  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
+)
+
+library(dynalysis)
+library(tidyverse)
+library(cowplot)
+
+# refering to figures & supplementary figures
+refs <- tibble(ref_id = character(), name = character(), ref_type = character())
+ref <- function(ref_type, ref_id, suffix="", anchor=FALSE) {
+  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
+    refs <<- refs %>% bind_rows(tibble(
+      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
+      ref_type = ref_type,
+      ref_id = ref_id
+    ))
+  }
+  ref <- refs %>% filter(ref_id == !!ref_id) %>% 
+    pull(name) %>% 
+    {pritt("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
+
+  if (anchor)  ref <- pritt("{ref}{anchor(ref_type, ref_id)}")
+
+  ref
+}
+anchor <- function(ref_type, ref_id) {
+  pritt("<a name='{ref_type}_{ref_id}'></a>")
+}
+create_names <- list(
+  sfig = function(i) pritt("Supplementary Figure {i}"),
+  fig = function(i) pritt("Figure {i}"),
+  snote = function(i) pritt("Supplementary Note {i}")
+)
+
+sfigs <- tibble(ref_id = character(), fig = list(), caption = character(), width=numeric(), height=numeric())
+add_sfig <- function(fig, ref_id, caption, width=15, height=10) {
+  sfigs <<- sfigs %>% add_row(
+    fig = list(fig), 
+    ref_id = ref_id, 
+    caption=caption, 
+    width=width, 
+    height=height
+  )
+}
+```
+
 # Introduction
@@
 
-```{r, fig.width = 15, fig.height = 10}
+```{r}
+add_sfig({
 n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
@@
 cowplot::plot_grid(plotlist=list(n_methods_over_time, trajectory_components_over_time), nrow=2, rel_heights = c(0.7, 0.3), labels="auto")
+}, 
+"methods_over_time", 
+"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
+", 
+15, 
+10
+)
 ```
-
-`r ref("sfig", "methods_over_time", anchor=TRUE)` New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
 
@@
 We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(method_df$wrapper == "Done")` for evaluation, primarily based on their free availability and the presence of a programming interface.
-We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and underlying algorithm(s) (`r ref("fig", "method_characteristics")`).
+We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
 
@@
 The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).
+After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
 
-After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
-Most methods reach a decent score, fulfilling most of the quality criteria (`r ref("fig", "method_characteristics", "b")`) recent methods tend to have higher code quality sFIG.
-Several aspects are lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation are problematic areas in particular (`r ref("sfig", "method_qc_check_plot")` and `r ref("fig", "method_characteristics", "c")`).
+Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
+Recent methods tend to have higher code quality sFIG.
+However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014].
 
@@
 
+`r anchor("fig", "method_characteristics")`
 ![](`r figure_file("figure_methods.svg", experiment_id="method_characteristics")`)
-
 `r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.
 
-```{r, fig.width = 10, fig.height = 13}
-readRDS(figure_file("method_qc_check_plot.rds"))
+```{r}
+add_sfig({
+readRDS(figure_file("method_qc_check_plot.rds", experiment_id="method_characteristics"))
+}, 
+"method_qc_check", 
+"Caption", 
+10, 
+13
+)
 ```
-`r ref("sfig", "method_qc_check_plot", anchor=TRUE)`
 
-```{r, fig.width = 15, fig.height = 7.5}
-readRDS(figure_file("method_components_plot.rds"))
+```{r}
+add_sfig({
+readRDS(figure_file("method_components_plot.rds", experiment_id="method_characteristics"))
+}, 
+"method_components", 
+"TI methods frequently share steps", 
+15, 
+7.5
+)
 ```
-`r ref("sfig", "method_components", anchor=TRUE)` TI methods frequently share steps
-
 ### Evaluation overview
-
+`r anchor("fig", "evaluation_overview")`
 ![evaluation_overview](../../analysis/figures/evaluation_overview.png)
-`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation
+`r ref("fig", "evaluation_overview")` Overview of the evaluation
 
@@
 
+# Supplementary Figures
+
+```{r, echo=FALSE, results='asis'}
+subchunkify <- function(ref_id, caption, width=5, height=7) {
+  subchunk <- glue::glue("",
+  "",
+  "`r anchor('sfig', '{ref_id}')`",
+  "```{{r {ref_id}, fig.height={height}, fig.width={width}, echo=FALSE}}",
+  "sfigs %>% filter(ref_id == !!ref_id) %>% pull(fig) %>% .[[1]]",
+  "```",
+  "`r ref('sfig', '{ref_id}')` {caption}",
+  "", .sep = "\n")
+
+  cat(knitr::knit(text=subchunk, quiet=TRUE))
+}
+
+pwalk(sfigs, function(ref_id, fig, caption, width, height) {
+  subchunkify(ref_id, caption, width, height)
+})
+```
+
 # Colophon

### 2018-01-18 analysis/paper/update_paper_from_google_drive.R

@@ -4,8 +4,9 @@ library(rmarkdown)
 library(dynutils)
 
 drive <- drive_download(as_id("1BCCaP21N2PXfzhj9H09yEpZz9lLY2HXd_LxTSsJ_wro"), type="text/plain", overwrite=TRUE, path = tempfile())
-system(pritt("sed -i '1s/^.//' {drive$local_path}")) # remove first character, because this is some strang eunicode character
-system(pritt("cat analysis/paper/paper_base.Rmd {drive$local_path} > analysis/paper/paper.Rmd"))
+system(pritt("sed -i '1s/^.//' {drive$local_path}")) # remove first character, because this is some strange unicode character
+system(pritt("cat {drive$local_path} > analysis/paper/paper.Rmd"))
 
-render("analysis/paper/paper.Rmd", "html_document")
+file.remove("analysis/paper/paper.html")
+render("analysis/paper/paper.Rmd", "html_document", output_dir = "analysis/paper/", output_file = "paper.html")
 browseURL("analysis/paper/paper.html")

### 2018-01-23 analysis/paper/paper.Rmd
added paper setup

@@
 ```{r setup, include=FALSE}
-knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
-  comment = "#>",
-  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
-)
-
-library(dynalysis)
-library(tidyverse)
-library(cowplot)
-
-# refering to figures & supplementary figures
-refs <- tibble(ref_id = character(), name = character(), ref_type = character())
-ref <- function(ref_type, ref_id, suffix="", anchor=FALSE) {
-  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
-    refs <<- refs %>% bind_rows(tibble(
-      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
-      ref_type = ref_type,
-      ref_id = ref_id
-    ))
-  }
-  ref <- refs %>% filter(ref_id == !!ref_id) %>% 
-    pull(name) %>% 
-    {pritt("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
-
-  if (anchor)  ref <- pritt("{ref}{anchor(ref_type, ref_id)}")
-
-  ref
-}
-anchor <- function(ref_type, ref_id) {
-  pritt("<a name='{ref_type}_{ref_id}'></a>")
-}
-create_names <- list(
-  sfig = function(i) pritt("Supplementary Figure {i}"),
-  fig = function(i) pritt("Figure {i}"),
-  snote = function(i) pritt("Supplementary Note {i}")
-)
-
-sfigs <- tibble(ref_id = character(), fig = list(), caption = character(), width=numeric(), height=numeric())
-add_sfig <- function(fig, ref_id, caption, width=15, height=10) {
-  sfigs <<- sfigs %>% add_row(
-    fig = list(fig), 
-    ref_id = ref_id, 
-    caption=caption, 
-    width=width, 
-    height=height
-  )
-}
+source("setup.R")
 ```
@@
 
-Single-cell[a] -omics technologies now make it possible to more accurately model biological systems than ever before.
+Single-cell -omics technologies now make it possible to more accurately model biological systems than ever before.
 One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
@@
 
-Dozens[b] of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
+<!--- history of Ti, different methods --->
+Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
 Initially, most trajectory inference was focused on linear trajectories, with one start and one end point.
@@
 
-Although[c] trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+<!--- Common framework of TI, different components, naturally progressing towards the need for an evaluation --->
+Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
 Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
-Furthermore, with the ongoing strong decreases in single-cell -omics technologies, it becomes
+Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become scalable and user friendly .
 
-Given[d] this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
+<!--- Common framework of TI, different components, naturally progressing towards the need for an evaluation --->
+Given[a] this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
@@
 
-At the level of trajectory structures, most current methods can handle one or more splits in the trajectory (`r ref("fig", "method_characteristics", "a")`).
-However, only one method (SLICER) can currently handle more complex graph structures which can include loops.
+At the level of trajectory structures, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`).
+However, only one method (SLICER) can currently handle more complex graph structures which can include loops (`r ref("sfig", "methods_over_time", "b")`).
 
@@
 The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).
+These categorisations can then be used to help current developers to improve their tool, and to guide the selection of users and developers to use these tools for their purpose.
 After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
@@
 However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014].
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding studies pinpointing these as good practices [@wilson_best_2014].
 
-The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
+The underlying algorithms can be divided into several modules, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps.
+Interestingly, modules are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
 
@@
 `r anchor("fig", "evaluation_overview")`
-![evaluation_overview](../../analysis/figures/evaluation_overview.png)
+![evaluation_overview](../figures/ti_evaluation_overview_v1.svg)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 We split the real datasets in two groups based on the type of reference trajectory available.
-The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.
+The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.[b]
 ### Score metrics and parameter optimisation
@@
 
-[a]General info about single-cell & trajectories
-[b]History of TI, different methods
-[c]Common framework of TI, different components, naturally progressing towards the need for an evaluation
-[d]Context and general overview of our evaluation
+[a]Context and general overview of our evaluation
+[b]Outdated

### 2018-01-23 analysis/paper/setup.R

@@ -1,0 +1,47 @@
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  comment = "#>",
+  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
+)
+library(dynalysis)
+library(tidyverse)
+library(cowplot)
+# refering to figures & supplementary figures
+refs <- tibble(ref_id = character(), name = character(), ref_type = character())
+ref <- function(ref_type, ref_id, suffix="", anchor=FALSE) {
+  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
+    refs <<- refs %>% bind_rows(tibble(
+      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
+      ref_type = ref_type,
+      ref_id = ref_id
+    ))
+  }
+  ref <- refs %>% filter(ref_id == !!ref_id) %>%
+    pull(name) %>%
+    {pritt("[**{.}{suffix}**](#{ref_type}_{ref_id})")}
+
+  if (anchor)  ref <- pritt("{ref}{anchor(ref_type, ref_id)}")
+  ref
+}
+anchor <- function(ref_type, ref_id) {
+  pritt("<a name='{ref_type}_{ref_id}'></a>")
+}
+create_names <- list(
+  sfig = function(i) pritt("Supplementary Figure {i}"),
+  fig = function(i) pritt("Figure {i}"),
+  snote = function(i) pritt("Supplementary Note {i}")
+)
+sfigs <- tibble(ref_id = character(), fig = list(), caption = character(), width=numeric(), height=numeric())
+add_sfig <- function(fig, ref_id, caption, width=15, height=10) {
+  sfigs <<- sfigs %>% add_row(
+    fig = list(fig),
+    ref_id = ref_id,
+    caption=caption,
+    width=width,
+    height=height
+  )
+}

### 2018-01-23 analysis/analyses/3-score_rules/1-generate-toys.Rmd
small update to score rules

@@
 toys_blueprint <- tribble(
-  ~trajectory_type, ~perturbator_id,
+  ~trajectory_model, ~perturbator_id,
   "simple_linear", "gs",
@@
 ) %>% 
-  bind_rows(tibble(trajectory_type="simple_linear", perturbator_id=paste0("switch_", percs * 100))) %>%
-  bind_rows(crossing(trajectory_type=trajectory_types, perturbator_id=paste0("change_network_", trajectory_types)) %>% mutate(ncells = 5)) %>% 
+  bind_rows(tibble(trajectory_model="simple_linear", perturbator_id=paste0("switch_", percs * 100))) %>%
+  bind_rows(crossing(trajectory_model=trajectory_models, perturbator_id=paste0("change_network_", trajectory_models)) %>% mutate(ncells = 5)) %>% 
   rowwise() %>% 
@@
   replicate=seq_len(nrow(.))%%nreplicates,
-  toy_category=paste0(trajectory_type, "-", perturbator_id),
+  toy_category=paste0(trajectory_model, "-", perturbator_id),
   toy_id=paste0(toy_category, "-", replicate)
@@
 toys$gs <- toys %>% split(seq_len(nrow(toys))) %>% parallel::mclapply(function(row) {
-  generate_dataset(row$trajectory_type, row$ncells)
+  dyntoy:::generate_dataset("why", row$trajectory_model, row$ncells)
 }, mc.cores=8)

### 2018-01-29 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
large update, include tables, include evaluation overview, update paper, add labels, ...

--- 2018-01-17 analysis/analyses/4-method_characterization/2_quality_control.Rmd
@@
 source("0_common.R")
-```
-
-```{r}
-category_gradients_white <- map(category_colors, function(color) {
-  n <- 1000
-  ramp <- shades::gradient(c("white", color), n)
-
-  function(x, min=0, max=1) {
-    x[x == 0] <- 0.0000000001
-    ramp[round((x - min)/(max-min)*(n-1))+1]
-  }
-})
-applications <- c("developer_friendly", "user_friendly", "good_science")
 ```
@@
 
-cowplot::plot_grid(method_ordering_plot + empty_x, check_heatmap_plot, nrow=2, align = "v", axis = "rl", rel_heights = c(0.2, 0.9))
+method_qc_check_plot <- cowplot::plot_grid(method_ordering_plot + empty_x, check_heatmap_plot, nrow=2, align = "v", axis = "rl", rel_heights = c(0.2, 0.9))
 
+method_qc_check_plot %>% saveRDS(figure_file("method_qc_check_plot.rds"))
 ```
@@
     scale_fill_identity() +
-    scale_alpha_continuous(range=c(1, 0.05), name = "Quality control score", guide = guide_legend())
+    scale_alpha_continuous(range=c(1, 0.05), name = "Quality control score", guide = guide_legend(title.position = "top", title.hjust = 0.5,label.position = "bottom"))
 question_ordering_plot2

### 2018-01-29 analysis/paper/paper.Rmd
large update, include tables, include evaluation overview, update paper, add labels, ...

@@
   Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
-Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories.
-However, the performance of these methods have not yet been comprehensively evaluated.
-Here, we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets.
+Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics.
+This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
+We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available.
+In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
 We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness.
@@
 * Go to tools -> perferences and remove Use smart quotes
-* Add enters after every paragraph (this is markdown!)
+* Add extra enters after every paragraph (this is markdown!)
 
@@
 When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Because they offer an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
+Because they offer an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
 
@@
 <!--- Common framework of TI, different components, naturally progressing towards the need for an evaluation --->
-Given[a] this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
@@
 # Load method characteristics
-method_df <- readRDS(derived_file("method_df.rds", experiment_id="method_characteristics"))
+methods <- readRDS(derived_file("methods.rds", experiment_id="method_characteristics"))
 ```
 
-We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(method_df$wrapper == "Done")` for evaluation, primarily based on their free availability and the presence of a programming interface.
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface.
 We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
+
+`r ref("table", "methods", anchor=TRUE)` Characterisation and quality control of the evaluated trajectory inference methods
+```{r}
+read_rds(figure_file("methods_table.rds", "method_characteristics")) %>%
+        knitr::kable()
+```
 
@@
 `r anchor("fig", "method_characteristics")`
-![](`r figure_file("figure_methods.svg", experiment_id="method_characteristics")`)
+![](`r figure_file("method_characteristics.svg", experiment_id="method_characteristics")`)
 `r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.
@@
 `r anchor("fig", "evaluation_overview")`
-![evaluation_overview](../figures/ti_evaluation_overview_v1.svg)
+![](`r figure_file("ti_evaluation_overview_v1.svg", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 We split the real datasets in two groups based on the type of reference trajectory available.
-The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.[b]
+The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.[a]
 ### Score metrics and parameter optimisation
@@
 
-[a]Context and general overview of our evaluation
-[b]Outdated
+[a]Outdated

### 2018-01-29 analysis/analyses/4-method_characterization/0_common.R
large update, include tables, include evaluation overview, update paper, add labels, ...

@@ -1,69 +1,0 @@
-trajectory_types <- c("undirected_linear","simple_fork","complex_fork","unrooted_tree","undirected_cycle","undirected_graph", "disconnected_undirected_graph")
-trajectory_type_undirected_to_directed<- c(
-  "undirected_linear"="directed_linear",
-  "simple_fork"="bifurcation",
-  "complex_fork"="multifurcation",
-  "unrooted_tree"="rooted_tree",
-  "undirected_cycle"="directed_cycle",
-  "undirected_graph"="directed_acyclic_graph",
-  "disconnected_undirected_graph"="disconnected_directed_graph",
-  "unknown"="unknown"
-)
-trajectory_type_colors <- c(
-  "undirected_linear" = "#af0dc7",
-  "simple_fork" = "#0073d7",
-  "unrooted_tree" = "#5ecd2e",
-  "complex_fork" = "#cfbf00",
-  "undirected_cycle" = "#39cccc",
-  "undirected_graph" = "#ff8821",
-  "disconnected_undirected_graph" = "#ff4237",
-  "unknown" = "#AAAAAA"
-)
-trajectory_type_colors[trajectory_type_undirected_to_directed[names(trajectory_type_colors)]] <- trajectory_type_colors
-
-lighten <- function(color, factor=1.4){
-  map_chr(color, function(color) {
-    col <- col2rgb(color)
-    col <- do.call(rgb2hsv, as.list(col))
-    col[1] <- col[1] * 360
-    col[2] <- 0.3
-    col[3] <- 0.9
-    colorspace::hex(do.call(colorspace::HSV, as.list(col)))
-  })
-}
-trajectory_type_background_colors <- set_names(rep("white", length(trajectory_types)), trajectory_types)
-trajectory_type_background_colors <- lighten(trajectory_type_colors, 0.3)
-
-
-
-applications <- c("developer_friendly", "user_friendly", "good_science")
-application_labels <- setNames(applications %>% gsub("_", " ", .) %>% Hmisc::capitalize(), applications)
-
-categories <- c("availability", "code_quality", "code_assurance", "documentation", "behaviour", "paper")
-category_colors <- c("#3498DB", "#E74C3C", "#A4CC2E", "#FEB308", "#B10DC9", "#85144b", "#EA8F10", "#2ECC49", "#CC2E63") %>% .[1:length(categories)] %>% setNames(categories)
[…]

### 2018-01-30 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
bye bye common.R, hello common in dynalyisis package

@@
 method_qc <- readRDS(derived_file("method_qc.rds"))
+methods_evaluated <- readRDS(derived_file("methods_evaluated.rds"))
 checks <- readRDS(derived_file("checks.rds"))
+```
 
-source("0_common.R")
+Comparing qc and citations overtime     
+```{r, fig.width=16, fig.height=8}
+g1 <- ggplot(methods_evaluated, aes(date, qc_score)) +
+    geom_smooth(span=2) +
+    geom_point() +
+    ggrepel::geom_label_repel(aes(label = name)) +
+    cowplot::theme_cowplot() +
+    labs(x = "Time", y = "QC score", title = "Code quality score over time")
+g1
+
+g2 <- ggplot(methods_evaluated, aes(date, ncitations+1)) +
+    geom_smooth(span=2) +
+    geom_point() +
+    ggrepel::geom_label_repel(aes(label = name)) +
+    cowplot::theme_cowplot() +
+    scale_y_log10() +
+    labs(x = "Time", y = "Citations", title = "Citations over time")
+g2
+
+g3 <- ggplot(methods_evaluated, aes(ncitations+1, qc_score)) +
+    geom_smooth(span=2) +
+    geom_point() +
+    ggrepel::geom_label_repel(aes(label = name)) +
+    cowplot::theme_cowplot() +
+    scale_x_log10() +
+    labs(x = "Citations", y = "QC score", title = "QC score over # citations")
+g3
+
+cowplot::plot_grid(g1, g3, g2, ncol=2, nrow=2, labels = "auto") %>% 
+  write_rds(figure_file("qc_over_time.rds"))
 ```
@@
 
-Orderin by category or application
+Ordering by category or application
 ```{r}
@@
 method_qc %>%
-  gather(application, application_applicable, !!applications) %>%
+  gather(application, application_applicable, !!qc_applications$application) %>%
   filter(application_applicable) %>%
@@
     scale_y_continuous("Method", breaks = seq_along(levels(method_qc$method_id))+0.5, labels = levels(method_qc$method_id), expand = c(0, 0)) +
-    scale_fill_manual(values=category_colors) +
+    scale_fill_manual(values=set_names(qc_categories$color, qc_categories$category)) +
     coord_flip() +
@@
 
-method_qc_check_plot %>% saveRDS(figure_file("method_qc_check_plot.rds"))
+method_qc_check_plot %>% saveRDS(figure_file("method_qc_check.rds"))
 ```
@@
 
-question_ordering_plot1 <- question_ordering_plotdata %>% 
-  ggplot() +
-    geom_bar(aes(question_id, n, fill=qc_score), stat="identity") +
-    scale_fill_distiller(palette="Blues", direction = -1)
-question_ordering_plot1
-
 question_ordering_plotdata2 <- question_ordering_plotdata %>% 
-  mutate(category = factor(category, levels=categories)) %>% 
+  mutate(category = factor(category, levels=qc_categories$category)) %>% 
   arrange(-as.numeric(category), question_ordering) %>% 
@@
   ggplot() +
-    geom_bar(aes(question_id, n, fill=category_colors[category], alpha=qc_score), stat="identity") +
+    geom_bar(aes(question_id, n, fill=set_names(qc_categories$color, qc_categories$category)[category], alpha=qc_score), stat="identity") +
     geom_vline(aes(xintercept=start), data=question_ordering_plotdata2_category) +
@@
     geom_rect(aes(xmin=0, xmax=1, ymin=start, ymax=end), fill="white") +
-    geom_text(aes(1, start + (end-start)/2, label=category_labels[category], color=category), hjust=1, size=5) +
+    geom_text(aes(1, start + (end-start)/2, label=label_long(category), color=category), hjust=1, size=5) +
     geom_segment(aes(x=1.1, xend=1.1, y=start-0.2, yend=end+0.2, color=category)) +
-    scale_color_manual(values=category_colors) +
+    scale_color_manual(values=set_names(qc_categories$color, qc_categories$category)) +
     scale_y_continuous(expand=c(0, 0)) +
@@
 method_qc_ordering_plot
-saveRDS(method_qc_ordering_plot, figure_file("method_qc_ordering_plot.rds"))
+saveRDS(method_qc_ordering_plot, figure_file("method_qc_ordering.rds"))
 ```

### 2018-01-30 analysis/paper/paper.Rmd
bye bye common.R, hello common in dynalyisis package

@@
 <!--- history of Ti, different methods --->
-Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
-Initially, most trajectory inference was focused on linear trajectories, with one start and one end point.
-However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`).
-Some methods can already handle more complex, cyclical and branching behaviour [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface.
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface `r ref("table", "methods")`.
 We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
 
-`r ref("table", "methods", anchor=TRUE)` Characterisation and quality control of the evaluated trajectory inference methods
+`r ref("table", "methods", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
 ```{r}
-read_rds(figure_file("methods_table.rds", "method_characteristics")) %>%
-        knitr::kable()
+read_rds(figure_file("methods_table.rds", "method_characteristics"))
 ```
@@
 At the level of trajectory structures, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`).
-However, only one method (SLICER) can currently handle more complex graph structures which can include loops (`r ref("sfig", "methods_over_time", "b")`).
+However, only a handful of more recent methods can currently handle more complex graph structures which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
-Recent methods tend to have higher code quality sFIG.
+Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
 However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
@@
 add_sfig({
-readRDS(figure_file("method_qc_check_plot.rds", experiment_id="method_characteristics"))
+readRDS(figure_file("method_qc_check.rds", experiment_id="method_characteristics"))
 }, 
@@
 add_sfig({
-readRDS(figure_file("method_components_plot.rds", experiment_id="method_characteristics"))
+readRDS(figure_file("method_components.rds", experiment_id="method_characteristics"))
 }, 
@@
 7.5
+)
+```
+
+```{r}
+add_sfig({
+readRDS(figure_file("qc_over_time.rds", experiment_id="method_characteristics"))
+}, 
+"qc_over_time", 
+"Quality control scores and number of citations over time", 
+10, 
+10
 )
@@
 
-## Toy data generation
-
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods `r ref("table", "methods", anchor=TRUE)`, based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018,@gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
+We believe this to be the most complete list to data, but are happy to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
+For each method we investigated whether it was
 ## Synthetic data generation
@@
 
-pwalk(sfigs, function(ref_id, fig, caption, width, height) {
+pwalk(sfigs %>% arrange(match(ref_id, refs$ref_id)), function(ref_id, fig, caption, width, height) {
   subchunkify(ref_id, caption, width, height)
@@
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
+This report was generated on `r Sys.time()` using `r devtools:::platform_info()$version` and the following packages:
 ```{r colophon, cache = FALSE}
-# which R packages and versions?
-devtools::session_info()
+devtools::session_info()$packages %>% knitr::kable()
 ```

### 2018-01-30 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
large paper update, now include bookdown, now include supplementary figures, method update

@@
 checks <- readRDS(derived_file("checks.rds"))
+```
+
+Table for the different checks
+```{r, results="asis"}
+library(kableExtra)
+table <- checks %>% 
+  mutate(
+    weight = cell_spec(
+      pritt("<span style='display: none;'>{question_id}</span> {weight}"), #add an invisible grouping indicator
+      background = spec_color(weight, end=0.9,option="A",direction=-1),
+      escape=FALSE,
+      color="white"
+    )
+  ) %>%
+  mutate(
+    references = ifelse(!is.na(references), pritt("[{references}]"), ""),
+    references = pritt("<span style='display: none;'>{question_id}</span> {references}")
+  ) %>% 
+  select(question, weight, references, score, scoring, !!qc_applications$application) %>% 
+  mutate_at(vars(!!qc_applications$application), ~ifelse(., "✓", "")) %>% 
+  rename_all(label_long) %>% 
+  knitr::kable("html", escape=FALSE) %>% 
+  kable_styling(bootstrap_options = "striped") %>% 
+  collapse_rows(columns=1:3)
+
+for (category in unique(checks$category)) {
+  color <- qc_categories %>% filter(category == !!category) %>% pull(color)
+  table <- table %>% 
+    group_rows(
+      label_long(category), 
+      min(which(checks$category == category)), 
+      max(which(checks$category == category)),
+      label_row_css = pritt("background-color: {color}; color: #fff;")
+  )
+}
+
+table %>% write_rds(figure_file("qc_checks.rds"))
+
+table
 ```

### 2018-01-30 analysis/paper/paper.Rmd
large paper update, now include bookdown, now include supplementary figures, method update

@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: 
-    bookdown::html_document2:
-    fig_caption: yes
-    reference_docx: templates/template.docx
 bibliography: references.bib
@@
 While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
-We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("snote", "qc_scoring")`).
+We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
 The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).
@@
 After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
+
+```{r}
+add_stable(
+read_rds(figure_file("qc_checks.rds","method_characteristics")), 
+"qc_checks", 
+""
+)
+```
 
@@
 We believe this to be the most complete list to data, but are happy to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
-For each method we investigated whether it was
+Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
+This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "methods")`).
+
+## Method quality control
+
+We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and used several references to weigh each quality check.
+
 ## Synthetic data generation
@@
 
-# Supplementary Figures
+# Supplementary
 
 ```{r, echo=FALSE, results='asis'}
-subchunkify <- function(ref_id, caption, width=5, height=7) {
-  subchunk <- glue::glue("",
-  "",
-  "`r anchor('sfig', '{ref_id}')`",
-  "```{{r {ref_id}, fig.height={height}, fig.width={width}, echo=FALSE}}",
-  "sfigs %>% filter(ref_id == !!ref_id) %>% pull(fig) %>% .[[1]]",
-  "```",
-  "`r ref('sfig', '{ref_id}')` {caption}",
-  "", .sep = "\n")
-
-  cat(knitr::knit(text=subchunk, quiet=TRUE))
-}
-
-pwalk(sfigs %>% arrange(match(ref_id, refs$ref_id)), function(ref_id, fig, caption, width, height) {
-  subchunkify(ref_id, caption, width, height)
-})
+source("analysis/paper/sfigs_stables.R")
 ```

### 2018-01-30 analysis/paper/paper.Rmd
dont need to include libs

@@
 # Discussion
-# Conclusion
 
-# Online Methods
+# Methods
 
@@
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods `r ref("table", "methods", anchor=TRUE)`, based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018,@gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
-We believe this to be the most complete list to data, but are happy to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
+We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
 
-We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and used several references to weigh each quality check.
+We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and weighed each quality check by the number of times it was found in a list of
 

### 2018-01-30 analysis/paper/paper.Rmd
added datasets table, update to methods section

@@
 title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
-author:
-  - Wouter Saelens* ¹ ²
-  - Robrecht Cannoodt* ¹ ² ³
-  - Helena Todorov¹ ² ⁴ ⁵
-  - Yvan Saeys¹ ²
-date: "`r format(Sys.time(), '%d %B, %Y')`"
 bibliography: references.bib
 csl: nature-biotechnology.csl
-abstract: |
-  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
-Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics.
-This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available.
-In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
-We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness.
-We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
-Our evaluation pipeline is fully reproducible, can be easily extended, and is available at www.github.com/dynverse.
-keywords: |
-  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
-highlights: |
-  These are the highlights. 
 ---
-\* Equal contribution  
-¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
-² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
-³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
-⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
-⁵ Université Claude Bernard Lyon 1, Lyon, France.  
+
+*`r format(Sys.time(), '%d %B, %Y')`*
 <!----
@@
 ```
+Wouter Saelens* ¹ ² , Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²
+
+\* Equal contribution  
+¹ Data mining and Modelling for Biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
+² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
+³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
+⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
+⁵ Université Claude Bernard Lyon 1, Lyon, France.
+
+# Abstract
+
+Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
+Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics.
+This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
+We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available.
+In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
+We employ a novel parameter optimisation procedure, and We compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness.
+We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
+Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.
 
@@
 
+```{r}
+add_stable(
+read_rds(figure_file("datasets.rds","dataset_characterisation")), 
+"datasets", 
+"Real datasets used in this study"
+)
+```
+
 Our evaluation was structured in three main phases.
@@
 
-We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and weighed each quality check by the number of times it was found in a list of
+We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and weighed each quality check by the number of times it was found in a list of references.
+
+## Real data processing
+
+We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`) .
 
@@
 
-## Real data processing
-
-We gathered
-
 ## Normalization
@@
 # Acknowledgements
+# References
 
-# Supplementary
+<div id="refs"></div>
+
+# Supplementary Material
 
@@
 ```
-
-# References
-
 [a]Outdated

### 2018-01-30 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
added datasets table, update to methods section

--- 2018-01-30 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
@@
 checks <- readRDS(derived_file("checks.rds"))
+```
+
+Table for the different checks
+```{r, results="asis"}
+library(kableExtra)
+table <- checks %>% 
+  mutate(
+    weight = cell_spec(
+      pritt("<span style='display: none;'>{question_id}</span> {weight}"), # add an invisible grouping indicator
+      background = spec_color(weight, end=0.9,option="A",direction=-1),
+      escape=FALSE,
+      color="white"
+    )
+  ) %>%
+  mutate(
+    references = references %>% 
+      gsub(",", "; ", .) %>% 
+      {ifelse(!is.na(.), pritt("[{.}]"), "")} %>%
+      {pritt("<span style='display: none;'>{question_id}</span> {.}")} # add an invisible grouping indicator
+  ) %>% 
+  select(question, weight, references, score, scoring, !!qc_applications$application) %>% 
+  mutate_at(vars(!!qc_applications$application), ~ifelse(., "✓", "")) %>% 
+  rename_all(label_long) %>% 
+  knitr::kable("html", escape=FALSE) %>% 
+  kable_styling(bootstrap_options = "striped") %>% 
+  collapse_rows(columns=1:3)
+
+for (category in unique(checks$category)) {
+  color <- qc_categories %>% filter(category == !!category) %>% pull(color)
+  table <- table %>% 
+    group_rows(
+      label_long(category), 
+      min(which(checks$category == category)), 
+      max(which(checks$category == category)),
+      label_row_css = pritt("background-color: {color}; color: #fff;")
+  )
+}
+
+table %>% write_rds(figure_file("qc_checks.rds"))
+
+table
 ```

### 2018-01-31 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
rename qc

--- 2018-01-30 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
@@
 checks <- readRDS(derived_file("checks.rds"))
+```
+
+Table for the different checks
+```{r, results="asis"}
+library(kableExtra)
+table <- checks %>% 
+  mutate(
+    weight = cell_spec(
+      pritt("<span style='display: none;'>{aspect_id}</span> {weight}"), # add an invisible grouping indicator
+      background = spec_color(weight, end=0.9,option="A",direction=-1),
+      escape=FALSE,
+      color="white"
+    )
+  ) %>%
+  mutate(
+    references = references %>% 
+      gsub(",", "; ", .) %>% 
+      {ifelse(!is.na(.), pritt("[{.}]"), "")} %>%
+      {pritt("<span style='display: none;'>{aspect_id}</span> {.}")} # add an invisible grouping indicator
+  ) %>% 
+  select(aspect, weight, references, item, item_weight, !!qc_applications$application) %>% 
+  mutate_at(vars(!!qc_applications$application), ~ifelse(., "✓", "")) %>% 
+  rename_all(label_long) %>% 
+  knitr::kable("html", escape=FALSE) %>% 
+  kable_styling(bootstrap_options = "striped") %>% 
+  collapse_rows(columns=1:3)
+
+for (category in unique(checks$category)) {
+  color <- qc_categories %>% filter(category == !!category) %>% pull(color)
+  table <- table %>% 
+    group_rows(
+      label_long(category), 
+      min(which(checks$category == category)), 
+      max(which(checks$category == category)),
+      label_row_css = pritt("background-color: {color}; color: #fff;")
+  )
+}
+
+table %>% write_rds(figure_file("qc_checks.rds"))
+
+table
 ```
@@
 ```{r}
-final_scores <- method_qc %>%
-  group_by(method_id) %>%
-  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
-  arrange(-score)
-
 # set the ordering of method_ids to this ordering
-method_qc$method_id <- factor(method_qc$method_id, levels=final_scores$method_id)
-final_scores$method_id <- factor(final_scores$method_id, levels=final_scores$method_id)
-
-method_ordering_plot <- final_scores %>%
-  ggplot(aes(method_id, score)) +
-    geom_bar(aes(fill=score), stat="identity") +
+method_ordering_plot <- methods_evaluated %>%
+  arrange(-qc_score) %>% 
+  mutate(method_id = factor(name, levels=name)) %>% 
+  ggplot(aes(method_id, qc_score)) +
+    geom_bar(aes(fill=qc_score), stat="identity") +
     theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
@@
 
-Ordering by category or application
-```{r}
-method_qc %>%
-  group_by(method_id, category) %>%
-  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
-  ggplot(aes(method_id, category)) + geom_raster(aes(fill=score))
-
-method_qc %>%
-  gather(application, application_applicable, !!qc_applications$application) %>%
-  filter(application_applicable) %>%
-  group_by(method_id, application) %>%
-  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
-  ggplot(aes(method_id, application)) + geom_raster(aes(fill=score))
-```
-
 ```{r, fig.width=10, fig.height=12}
-question_position <- method_qc %>%
+aspect_position <- method_qc %>%
   filter(method_id == "SCORPIUS") %>%
-  arrange(question_id) %>%
-  group_by(question_id) %>%
+  arrange(aspect_id) %>%
+  group_by(aspect_id) %>%
   summarise(
-    question_width = sum(score * weight),
+    aspect_width = sum(score * weight),
     name = first(name)
@@
   mutate(
-    question_end = cumsum(question_width),
-    question_start = lag(question_end, 1, default=0),
-    question_mid = question_start + (question_end - question_start)/2
+    aspect_end = cumsum(aspect_width),
+    aspect_start = lag(aspect_end, 1, default=0),
+    aspect_mid = aspect_start + (aspect_end - aspect_start)/2
   )
@@
   filter(method_id == "SCORPIUS") %>%
-  arrange(question_id, check_id) %>%
+  arrange(aspect_id, check_id) %>%
   group_by(check_id) %>%
@@
     check_width = score * weight,
-    question_id = first(question_id)
+    aspect_id = first(aspect_id)
   ) %>%
-  left_join(question_position, by="question_id") %>%
-  group_by(question_id) %>%
+  left_join(aspect_position, by="aspect_id") %>%
+  group_by(aspect_id) %>%
   mutate(
-    check_start = question_start + lag(cumsum(check_width), default=0),
-    check_end = question_start + cumsum(check_width)
+    check_start = aspect_start + lag(cumsum(check_width), default=0),
+    check_end = aspect_start + cumsum(check_width)
   )
@@
     geom_vline(aes(xintercept=check_end), data=check_position, color="#444444", alpha=0.25) +
-    geom_vline(aes(xintercept=question_end), data=question_position, color="#444444") +
-    scale_x_reverse("Check", breaks = question_position$question_mid, labels = question_position$name, expand = c(0, 0)) +
+    geom_vline(aes(xintercept=aspect_end), data=aspect_position, color="#444444") +
+    scale_x_reverse("Check", breaks = aspect_position$aspect_mid, labels = aspect_position$name, expand = c(0, 0)) +
     scale_y_continuous("Method", breaks = seq_along(levels(method_qc$method_id))+0.5, labels = levels(method_qc$method_id), expand = c(0, 0)) +
@@
     scale_y_reverse(expand=c(0, 0)) +
-    scale_x_discrete(labels=setNames(method_qc$scoring, method_qc$check_id)) +
+    scale_x_discrete(labels=setNames(method_qc$item_weight, method_qc$check_id)) +
     scale_fill_distiller(palette="Blues", direction = 1)
@@
 ```{r}
-question_info <- method_qc %>% 
-  group_by(question_id) %>% 
+aspect_info <- method_qc %>% 
+  group_by(aspect_id) %>% 
   summarise(
@@
 
-question_ordering_plotdata <- method_qc %>%
-  group_by(method_id, question_id) %>%
+aspect_ordering_plotdata <- method_qc %>%
+  group_by(method_id, aspect_id) %>%
   summarise(
-    qc_score=sum(answer * score * weight)/sum(score * weight)
+    qc_score=sum(answer * item_weight * weight)/sum(item_weight * weight)
   ) %>%
   ungroup() %>% 
-  group_by(question_id) %>% 
+  group_by(aspect_id) %>% 
   count(qc_score) %>% 
-  mutate(question_ordering = sum((qc_score < 1) * n)) %>% 
+  mutate(aspect_ordering = sum((qc_score < 1) * n)) %>% 
   ungroup() %>% 
-  arrange(question_ordering) %>% 
-  left_join(question_info, by="question_id") %>% 
-  mutate(question_id = factor(question_id, levels=unique(question_id)))
+  arrange(aspect_ordering) %>% 
+  left_join(aspect_info, by="aspect_id") %>% 
+  mutate(aspect_id = factor(aspect_id, levels=unique(aspect_id)))
 
-question_ordering_plotdata2 <- question_ordering_plotdata %>% 
+aspect_ordering_plotdata2 <- aspect_ordering_plotdata %>% 
   mutate(category = factor(category, levels=qc_categories$category)) %>% 
-  arrange(-as.numeric(category), question_ordering) %>% 
-  mutate(question_id = factor(question_id, levels=as.character(unique(question_id))))
+  arrange(-as.numeric(category), aspect_ordering) %>% 
+  mutate(aspect_id = factor(aspect_id, levels=as.character(unique(aspect_id))))
 
-question_ordering_plotdata2_category <- question_ordering_plotdata2 %>% 
-  group_by(category, question_id) %>% 
+aspect_ordering_plotdata2_category <- aspect_ordering_plotdata2 %>% 
+  group_by(category, aspect_id) %>% 
   summarise() %>% 
@@
 
-question_ordering_plot2 <- question_ordering_plotdata2 %>% 
+aspect_ordering_plot2 <- aspect_ordering_plotdata2 %>% 
   arrange(qc_score) %>% 
   ggplot() +
-    geom_bar(aes(question_id, n, fill=set_names(qc_categories$color, qc_categories$category)[category], alpha=qc_score), stat="identity") +
-    geom_vline(aes(xintercept=start), data=question_ordering_plotdata2_category) +
+    geom_bar(aes(aspect_id, n, fill=set_names(qc_categories$color, qc_categories$category)[category], alpha=qc_score), stat="identity") +
+    geom_vline(aes(xintercept=start), data=aspect_ordering_plotdata2_category) +
     scale_fill_identity() +
     scale_alpha_continuous(range=c(1, 0.05), name = "Quality control score", guide = guide_legend(title.position = "top", title.hjust = 0.5,label.position = "bottom"))
-question_ordering_plot2
+aspect_ordering_plot2
 
-category_legend <- question_ordering_plotdata2 %>% 
-  group_by(category, question_id) %>% 
+category_legend <- aspect_ordering_plotdata2 %>% 
+  group_by(category, aspect_id) %>% 
   summarise() %>% 
@@
 
-question_ordering_plot2 <- question_ordering_plot2 +
+aspect_ordering_plot2 <- aspect_ordering_plot2 +
     coord_flip() +
     scale_y_continuous(expand=c(0, 0), name = "# methods", c(seq(0, n_methods, 5), n_methods)) +
-    scale_x_discrete(labels=setNames(question_ordering_plotdata$name, question_ordering_plotdata$question_id), name="") +
+    scale_x_discrete(labels=setNames(aspect_ordering_plotdata$name, aspect_ordering_plotdata$aspect_id), name="") +
     theme(legend.position="top")
 
-method_qc_ordering_plot <- cowplot::plot_grid(category_legend, question_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
+method_qc_ordering_plot <- cowplot::plot_grid(category_legend, aspect_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
 method_qc_ordering_plot

### 2018-01-31 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
another fix for method characterisation

@@
 table %>% write_rds(figure_file("qc_checks.rds"))
-
-table
 ```
@@
   summarise(
-    aspect_width = sum(score * weight),
+    aspect_width = sum(item_weight * weight),
     name = first(name)
@@
   summarise(
-    check_width = score * weight,
+    check_width = item_weight * weight,
     aspect_id = first(aspect_id)

### 2018-02-01 analysis/paper/paper.Rmd

@@
 *`r format(Sys.time(), '%d %B, %Y')`*
-<!----
+<!--
 Tips for working in this document:
@@
 A comparison of single-cell trajectory inference methods: towards more accurate and robust tools
---->
+-->
 
@@
 
-<!--- history of Ti, different methods --->
+<!-- history of Ti, different methods -->
+Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
+Initially, most trajectory inference fixed the trajectory structure to a linear or bifurcating structure, by either algorithmic limitations or through user parameters, and focussed on correctly ordering the cells along this fixed structure.
+However, several recent methods now infer both the ordering and the structure at the same time, although certain structural limitations are still imposed(`r ref("sfig", "methods_over_time", "b")`).
+Methods which do not impose limitations on the structure are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
-<!--- Common framework of TI, different components, naturally progressing towards the need for an evaluation --->
+<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
 Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
@@
 
-<!--- Common framework of TI, different components, naturally progressing towards the need for an evaluation --->
+<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
 Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
@@
 # Results
-## Method characterization
+## Method quality control
 
@@
 ```
-### Evaluation overview
+## Evaluation overview
+
 `r anchor("fig", "evaluation_overview")`
@@
 The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.[a]
-### Score metrics and parameter optimisation
 
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods `r ref("table", "methods", anchor=TRUE)`, based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018,@gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods `r ref("table", "methods", anchor=TRUE)`, based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
 We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
@@
 
-We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`) and weighed each quality check by the number of times it was found in a list of references.
+We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`).
+The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
+The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
+The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
+The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
+The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
+The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
+Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
 
-## Real data processing
+For each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "methods")`).
+This was to prevent make sure more important aspect, such as the open source availability of the method, to outweigh other aspects, such as the availability of a graphical user interface.
+For calculating the final score, we however weighed each of the six categories equally.
 
-We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`) .
+## Real datasets
 
-## Synthetic data generation
+We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
+The scripts to download and process these datasets are all available in our repository (http://www.github.com/dynverse/dynalysis).
+When available, we preferred to start from the raw counts data.
+These raw counts were all normalised and filtered using a common pipeline, discussed later.
+We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).
 
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
-At every step, great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
+## Synthetic datasets
+
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
 
@@
 
-One of the main process involved in cellular dynamic processes is through gene regulation, where regulatory cascades and feedback loops lead to .
+One of the main process involved in cellular dynamic processes is through gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcations seem to be mutual antagonism, where two genes [@XuRegulationbifurcatingcell2015] or gene modules strongly inhibit each other, forcing one of the two to become inactive [@GrafForcingcellschange2009].
-Which of these two Of Modules?
-[Th17]
+One driver of bifurcations seem to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009].
+Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; ferrell_bistabilitybifurcationswaddington_2012].
+Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into models) repressing each other [@yosef_dynamicregulatorynetwork_2013].
 
-## Normalization
+To simulate certain trajectory structures, we therefore started from the concept
+
+```{r}
+add_sfig({
+readRDS(figure_file("modulenets.rds", experiment_id="dataset_characterisation/synthetic")) %>%
+walk(cowplot::ggdraw)
+}, 
+"modulenets", 
+"Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
+15, 
+5
+)
+```
+
+### Simulation
+
+### Gold standard extraction
+
+### Simulation of the single-cell RNA-seq experiment
+
+## Common normalisation pipeline
+We used a popular single-cell RNA-seq pipeline [@lun_stepbystepworkflowlowlevel_2016] which uses the scran and scater Bioconductor packages.
+The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed.
+Furthermore, the pipeline can select the most highly variable genes.
 ## Evaluation metrics
-## Parameter optimization
+## Trajectory types
 

### 2018-02-01 analysis/paper/paper.Rmd
update paper

@@
 ```
-Wouter Saelens* ¹ ² , Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²
+`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³")), ", ")`, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²
 
@@
 At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
+For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
 
@@
 One driver of bifurcations seem to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009].
-Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; ferrell_bistabilitybifurcationswaddington_2012].
+Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
 Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into models) repressing each other [@yosef_dynamicregulatorynetwork_2013].
 
-To simulate certain trajectory structures, we therefore started from the concept
+<p class='wip'> Cascades? Cycle?
+</p>
+
+<p class='wip'> To simulate certain trajectory structures, we therefore designed module networks in which the cells will follow a particular trajectory structure given certain parameters (`r ref("sfig", "modulenets")`).
+Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated trajectories which simple undirected graphs (bifurcating loop and bifurcating convergence).
+</p>
+
+<p class='wip'> From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks.
+For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
+5% of genes were assigned to be part of the main, and were randomly distributed among all modules (with at least one gene per module).
+We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module.
+To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1).
+</p>
 
@@
 add_sfig({
-readRDS(figure_file("modulenets.rds", experiment_id="dataset_characterisation/synthetic")) %>%
-walk(cowplot::ggdraw)
+readRDS(figure_file("modulenets.rds", experiment_id="dataset_characterisation/synthetic"))
 }, 
@@
 
-### Simulation
+### Simulation of thermodynamics
+
+To simulate the gene regulatory network, we used a system of differential equations similar to which has been standard in evaluations of network inference [@marbach_wisdom_2012].
+In this model, changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations [@schaffter_genenetweaversilicobenchmark_2011] (ODEs):
+
+$$
+\begin{align}
+\label{eq:mrna_ode}
+\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation} \\
+\label{eq:prot_ode}
+\frac{dy_i}{dt} &= \underbrace{r \times x_i}_\text{production} - \underbrace{\Lambda \times y_i}_\text{degradation}
+\end{align}
+$$
+
+where $m$, $\lambda$, $r$ and $\Lambda$ represent production and degradation rates, the ratio of which determines the maximal gene and protein expression.
+The two types of equations are coupled because the production of protein $y_i$ depends on the amount of gene expression $x_i$, which in turn depends on the amount of other proteins through the activation function $f(y_1, y_2, ...)$.
+
+The activation function is inspired by a thermodynamic model of gene regulation, in which the promoter of a gene can be bound or unbound by a set of transcription factors, each representing a certain state of the promoter.
+Each state is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state.
+The production rate of the gene is calculated by combining the probabilities of the promoter being in each state with the relative activation:
+
+$$
+\begin{align}
+\label{eq:input_funcion_probabilities}
+f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
+\end{align}
+$$
+
+The probability of being in a state is based on the thermodynamics of transcription factor binding. When only one transcription factor is bound in a state:
+$$
+\begin{align}
+P_j \propto \nu = \left(\frac{y}{k}\right)^{n}
+\end{align}
+$$
+
+Where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
+
+$$
+\begin{align}
+P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
+\end{align}
+$$
+
+where $\rho$ represents the cooperativity of binding between the different transcription factors.
+
+$P_i$ is only proportional to $\nu$ because $\nu$ is normalized such that $\sum_{i} P_i = 1$.
+
+To each differential equation, we added an additional stochastic term:
+
+$$
+\begin{align}
+\frac{dx_i}{dt} = m \times f(y_1, y_2, ...) - \lambda \times x_i &+ \eta \times \sqrt{x_i} \times \Delta W_t \\
+\frac{dy_i}{dt} = r \times x_i - \Lambda \times y_i &+ \eta \times \sqrt{y_i} \times \Delta W_t
+\end{align}
+$$
+
+with $\Delta W_t \sim \mathcal{N}(0, h)$.
+
+Similar to [@schaffter_genenetweaversilicobenchmark_2011], we sample the different parameters from random distributions `r ref("stable", "samplers")`.
+
+```{r}
+add_stable(
+read_rds(figure_file("samplers.rds","dataset_characterisation/synthetic")), 
+"samplers", 
+"Distributions from which each parameter in the thermodynamic model was sampled"
+)
+```
+
+These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 2$.
+
+### Simulation of the single-cell RNA-seq experiment
+
+Estimation
+
+Simulation
 
@@
 
-### Simulation of the single-cell RNA-seq experiment
+DTW
 
@@
 
-```{r, echo=FALSE, results='asis'}
+```{r "supplementary", echo=FALSE, results='asis'}
 source("analysis/paper/sfigs_stables.R")

### 2018-02-01 analysis/paper/paper.Rmd
update normalisation + synthetic data paper

@@
 Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
-Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics.
+Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
 This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
@@
 Where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
-
 $$
@@
 To each differential equation, we added an additional stochastic term: 
-
 $$
@@
 
-These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 2$.
+These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$.
+The total simulation time varied between 5 for linear and bifurcating datasets, 10 for consecutive bifurcating, trifurcating and converging datasets, 15 for bifurcating converging datasets and 30 for linear long, cycle and bifurcating loop datasets.
+The burn-in period was for each simulation 2.
+Each network was simulated 32 times.
 
@@
 
-Estimation
-
-Simulation
+For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in.
+Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
+We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$.
+These were combined with the expression levels of the genes simulated within a trajectory.
+Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$.
+Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernouilli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
 
@@
 
-DTW
+Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward.
+Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?.
+To estimate a cell's position during a simulation within the trajectory structure, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
+We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
+We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
+In case of cyclical trajectory structures, the number of possible milestones a backbone could progress through was limited to 20.
 
 ## Common normalisation pipeline
-We used a popular single-cell RNA-seq pipeline [@lun_stepbystepworkflowlowlevel_2016] which uses the scran and scater Bioconductor packages.
-The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed.
-Furthermore, the pipeline can select the most highly variable genes.
+We used a standard single-cell RNA-seq preprocessing pipeline which uses the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016.
+The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3).
+We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05.
+Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5.
+As a final filter, we removed both all zero genes and all zero cells until convergence.
 ## Evaluation metrics

### 2018-02-01 analysis/paper/update_paper_from_google_drive.R
update paper

@@ -20,5 +20,12 @@ render(
 )
 browseURL("analysis/paper/paper.html")
 
+# render(
+#   "analysis/paper/paper.Rmd",
+#   output_format = rmarkdown::pdf_document(),
+#   output_dir = "analysis/paper/",
+#   output_file = "paper.pdf"
+# )
+
 # for conversion to pdf:
-#system(pritt("pandoc analysis/paper/paper.html +RTS -K1024M -RTS --pdf-engine=xelatex -o analysis/paper/paper.pdf"))
+# system(pritt("cd analysis/paper; pandoc paper.html +RTS -K1024M -RTS --pdf-engine=xelatex -o paper.pdf"))

### 2018-02-01 analysis/paper/paper.Rmd
update paper

@@
 
+## Trajectory types
+
+To classify the output of methods and the network `r ref("sfig", "trajectory_type_tree")`.
+
+```{r}
+add_sfig({
+readRDS(figure_file("trajectory_type_tree.rds", experiment_id="dataset_characterisation/trajectory_types"))
+}, 
+"trajectory_type_tree", 
+"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
+10, 
+10
+)
+```
+
 ## Trajectory inference methods
@@
 
-## Method quality control
+<p class='wip'> We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model.
+This model consists of two parts.
+The milestone network represents the overall network structure, and contains of edges between different milestones and the length of the edge between them.
+The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
+Linear methods are the easiest to wrap into this format, with two milestones and all cells in between.
+Methods such as
+</p>## Method quality control
 
@@
 
-<p class='wip'> From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks.
+From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks.
 For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
@@
 To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1).
-</p>
 
@@
 
-Similar to [@schaffter_genenetweaversilicobenchmark_2011], we sample the different parameters from random distributions `r ref("stable", "samplers")`.
+Similar to [@schaffter_genenetweaversilicobenchmark_2011], we sample the different parameters from random distributions, given in `r ref("stable", "samplers")`.
 
@@
 
-Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward.
+<p class='wip'> Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward.
 Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?.
@@
 In case of cyclical trajectory structures, the number of possible milestones a backbone could progress through was limited to 20.
+</p>
 
 ## Common normalisation pipeline
-We used a standard single-cell RNA-seq preprocessing pipeline which uses the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016.
+We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016].
 The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3).
@@
 Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5.
-As a final filter, we removed both all zero genes and all zero cells until convergence.
+As a final filter, we removed both all zero genes and cells until convergence.
+
 ## Evaluation metrics
-## Trajectory types
+
+We used two different metrics to assess the performance of each TI method, one looking at the accuracy of the cellular ordering, and one looking at the accuracy of the detected trajectory structure.
+The correlation metric first calculates the geodesic distances between all cells within the trajectory structure and than calculates the Spearman's correlation between these two distance matrices.
+Cells which are positioned closely to each other in both the prediction and the gold standard will therefore have an equally
 
@@
 ```{r}
-# what commit is this file at? You may need to change the path value
-# if your Rmd is not in analysis/paper/
 git2r::repository(".")

### 2018-02-02 analysis/analyses/4-method_characterisation/2_quality_control.Rmd

@@
 
-g2 <- ggplot(methods_evaluated, aes(date, ncitations+1)) +
+g2 <- ggplot(methods_evaluated %>% mutate(ncitations = ifelse(ncitations == 0, 0.5, ncitations)), aes(date, ncitations)) +
     geom_smooth(span=2) +
@@
 
-g3 <- ggplot(methods_evaluated, aes(ncitations+1, qc_score)) +
+g3 <- ggplot(methods_evaluated %>% mutate(ncitations = ifelse(ncitations == 0, 0.5, ncitations)), aes(ncitations, qc_score)) +
     geom_smooth(span=2) +
@@
 ```{r, fig.width=10, fig.height=12}
-aspect_position <- method_qc %>%
+qc_checks <- method_qc %>%
   filter(method_id == "SCORPIUS") %>%
+  group_by(category) %>% 
+  mutate(normalised_weight = weight/sum(weight))
+
+aspect_position <- qc_checks %>% 
   arrange(aspect_id) %>%
@@
   summarise(
-    aspect_width = sum(item_weight * weight),
+    aspect_width = sum(item_weight * normalised_weight),
     name = first(name)
@@
 
-check_position <- method_qc %>%
-  filter(method_id == "SCORPIUS") %>%
+check_position <- qc_checks %>%
   arrange(aspect_id, check_id) %>%
@@
   summarise(
-    check_width = item_weight * weight,
+    check_width = item_weight * normalised_weight,
     aspect_id = first(aspect_id)

### 2018-02-02 analysis/paper/paper.Rmd

@@
 
-Our evaluation was structured in three main phases.
-In the first phase, we used toy data as a positive control for each method, assessing whether it can detect the correct trajectories in very simplistic data.
-Next, we generated synthetic data from a model of gene regulation, combined with a simulation of the technical variation introduced by single-cell RNA sequencing.
-We used this synthetic data in a parameter optimization procedure, where we used cross-validation between datasets to avoid overfitting on particular data set structures.
-Finally, in the third phase, we used the optimized parameters to score the performance of each method on real data.
-We split the real datasets in two groups based on the type of reference trajectory available.
-The reference in datasets with a silver standard were defined by the author’s based on the expression data itself, while the reference for gold standard was defined using external information, such as fluorescent sorting or time series.[a]
-
-## Overall method comparison
-
-## Evaluation on synthetic data
-Initial evaluation..
-Robustness..
-Gegenereerde synthetische data. Parameter tuning voor echte data + initiële evaluatie (tot wat is de methode in staat) + robustness etc testing
-
-## Evaluation on real data
-Biological relevance..
-Echte data. Evaluatie (biologische relevantie van de methode)
-
 # Discussion
@@
 
-To classify the output of methods and the network `r ref("sfig", "trajectory_type_tree")`.
+To classify the output of methods and the gold standards, we defined a set of basic trajectory types.
+These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph using certain graph properties, progressively becoming more simple until the two basic types: linear and cycles `r ref("sfig", "trajectory_type_tree", "a")`.
+For every type, a corresponding directed graph type can be defined by allowing directedness `r ref("sfig", "trajectory_type_tree", "b")`.
 
@@
 add_sfig({
-readRDS(figure_file("trajectory_type_tree.rds", experiment_id="dataset_characterisation/trajectory_types"))
+readRDS(figure_file("trajectory_type_trees.rds", experiment_id="dataset_characterisation/trajectory_types"))
 }, 
@@
 "Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
-10, 
+15, 
 10
@@
 ```
-[a]Outdated

### 2018-02-06 analysis/paper/paper.Rmd
update paper

@@
 This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available.
+We developed a framework of data topologies for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available.
 In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
@@
 Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
-Initially, most trajectory inference fixed the trajectory structure to a linear or bifurcating structure, by either algorithmic limitations or through user parameters, and focussed on correctly ordering the cells along this fixed structure.
-However, several recent methods now infer both the ordering and the structure at the same time, although certain structural limitations are still imposed(`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the structure are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Initially, most trajectory inference fixed the trajectory topology to a linear or bifurcating topology, by either algorithmic limitations or through user parameters, and focussed on correctly ordering the cells along this fixed topology.
+However, several recent methods now infer both the ordering and the topology at the same time, although certain structural limitations are still imposed(`r ref("sfig", "methods_over_time", "b")`).
+Methods which do not impose limitations on the topology are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
@@
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface `r ref("table", "methods")`.
-We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
+We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
 
@@
 
-At the level of trajectory structures, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`).
-However, only a handful of more recent methods can currently handle more complex graph structures which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`).
+However, only a handful of more recent methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 This model consists of two parts.
-The milestone network represents the overall network structure, and contains of edges between different milestones and the length of the edge between them.
+The milestone network represents the overall network topology, and contains of edges between different milestones and the length of the edge between them.
 The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
@@
 
-<p class='wip'> To simulate certain trajectory structures, we therefore designed module networks in which the cells will follow a particular trajectory structure given certain parameters (`r ref("sfig", "modulenets")`).
+<p class='wip'> To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
 Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated trajectories which simple undirected graphs (bifurcating loop and bifurcating convergence).
@@
 
-<p class='wip'> Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward.
+<p class='wip'> Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward.
 Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?.
-To estimate a cell's position during a simulation within the trajectory structure, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
+To estimate a cell's position during a simulation within the trajectory toplogy, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
 We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
 We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
-In case of cyclical trajectory structures, the number of possible milestones a backbone could progress through was limited to 20.
+In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
 </p>
@@
 ## Evaluation metrics
+We used three different metrics to assess the performance of each TI method.
+The importance of using multiple metrics to compare complex models has been stated repeatedly REFS.
+Indeed, each of our metrics looks at the correspondence between prediction and gold standard in a different way: the correlation between geodesic distances assesses the overall ordering, the RF prediction error assesses the cell neighbourhood, and the edge flip score assesses the milestone network topology (`r ref("sfig", "score_aspects")`).
 
-We used two different metrics to assess the performance of each TI method, one looking at the accuracy of the cellular ordering, and one looking at the accuracy of the detected trajectory structure.
-The correlation metric first calculates the geodesic distances between all cells within the trajectory structure and than calculates the Spearman's correlation between these two distance matrices.
-Cells which are positioned closely to each other in both the prediction and the gold standard will therefore have an equally
+```{r}
+add_sfig({
+readRDS(figure_file("score_aspects.rds", experiment_id="score_rules/score_aspects"))
+}, 
+"score_aspects", 
+"Different scores assess different aspects of the ", 
+15, 
+5
+)
+```
+
+### Correlation between geodesic distances
+The accuracy of cell ordering was assessed by first calculating the geodesic distances $d(c_i, c_j)$ between all cells in both the gold standard and the prediction.
+This geodesic distance is calculated by finding the shortest path between two cells through the milestone network, while taking into account the edge lengths as returned by most methods.
+To calculate this for every pair of cells, we first calculated the shortest path lengths $d(m_a, m_b)$ between all milestones using Dijkstra's algorithm, as implemented in the R *igraph* package.
+We also calculated $d(m_a, c_i)$ for every cell to its neighbouring milestones by calculating the earth's movers distance using the shortlist method [@gottschlich_shortlistmethodfast_2014], as implemented in the R *transport* package.
+Then we again used Dijkstra's algorithm to calculate the shortest path lengths between every pair of cells which minimises:
+
+$$
+d(c_i, c_j) = min_{a \in A, b \in B} d(m_a, c_i) + d(m_a, m_b) + d(m_b, c_j)
+$$
+
+where A and B contain all milestones adjacent to respectively $c_i$ and $c_j$.
+The final score was then calculated the Spearman's rank correlation between all $d(c_i, c_j)$ in gold standard and prediction.
+
+In some cases, methods are unable to place some cells on a trajectory, mainly because they classify these cells as noisy.
+In these cases, we add these cells to a new milestone and put this milestone at a distance of 5 times the maximal distance between two milestones within the network.
+
+### Random forest prediction error
+
+Although the correlation between geodesic distances directly assesses the position of the cells in the trajectory, a bad correlation does not directly imply that similar cells were not grouped together by the method, as illustrated in `r ref("sfig", "score_aspects", c("b", "c"))`.
+Indeed, certain methods will inevitably reach an incorrect ordering because they cannot handle the correct trajectory type, but these methods could still correctly place similar cells next to each other.
+We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard.
+We used a Random Forest regression, implemented in the ranger package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of every cell in the gold standard, using the milestone percentages of these cells in the prediction as features.
+We then used the out-of-bag mean-squared error on these percentages to score each method's capability of predicting the correct neighbourhood of each cell.
+
+### Edge flip score
+
+As a third independent score, we assessed the similarity between the milestone network topology.
+We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that A->A would be converted A->B->C->D, by adding an intermediate node to linear networks.
+Because we are interested , the network was made undirected.
+Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
+This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no suitable [@bahiense_maximumcommonedge_2012].
+We implemented a branch&bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes.
+To further limit the search space, we make sure the degree distributions between the two networks are similar, before assessing whether the two networks are isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
+
+```{r}
+add_sfig({
+readRDS(figure_file("edge_flip_comparison.rds", experiment_id="score_rules/edge_flip"))
+}, 
+"edge_flip_comparison", 
+"Edge flip score values for common trajectory topologies", 
+15, 
+5
+)
+```
 

### 2018-02-06 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
small additions to method characterisation figures

@@
     cowplot::theme_cowplot() +
-    labs(x = "Time", y = "QC score", title = "Code quality score over time")
+    labs(x = label_long("publication_date"), y = "QC score", title = "Code quality score over time")
 g1

### 2018-02-06 analysis/paper/paper.Rmd
update paper

@@
 
-Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
+Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
@@
 To classify the output of methods and the gold standards, we defined a set of basic trajectory types.
-These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph using certain graph properties, progressively becoming more simple until the two basic types: linear and cycles `r ref("sfig", "trajectory_type_tree", "a")`.
-For every type, a corresponding directed graph type can be defined by allowing directedness `r ref("sfig", "trajectory_type_tree", "b")`.
+These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph using certain graph properties, progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
+For every type, a corresponding directed graph type can be defined by allowing directedness (`r ref("sfig", "trajectory_type_tree", "b")`).
 
@@
 
-<p class='wip'> We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model.
+We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its heterogenous output into a common probabilistic trajectory model.
 This model consists of two parts.
@@
 The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
-Linear methods are the easiest to wrap into this format, with two milestones and all cells in between.
-Methods such as
-</p>## Method quality control
+
+Depending on the output of a method, we used different strategies to convert the output to our common model:
+
+* Linear trajectory methods usually return a single pseudotime value, which can be converted to our model using a single edge between two milestones
+* Conversely, methods which assign each cell to a branch together with a pseudotime value and a branch network easily converted into our model
+* Methods returning both a global pseudotime value and a probability for every end state were converted into our model by using one single begin milestone and an end milestone for every end state.
+The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities
+Special conversions were necessary for certain method
+* DPT: 
+* Sincell
+* SLICER
+* SLICE
+*
+
+## Method quality control
 
@@
 15, 
-5
+8
 )

### 2018-02-06 analysis/analyses/3-score_rules/flip_score_characterisation.Rmd

@@
 library(dyneval)
+library(dynalysis)
+
+experiment("score_rules/edge_flip")
+
 knitr::opts_chunk$set(echo = TRUE)
@@
 
+Generate different trajectory topologies
 ```{r}
-models <- eval(formals(dyntoy:::generate_toy_milestone_network)$model)
-models <- models[models != "BA"]
-networks_synthetic <- map(models, dyntoy:::generate_toy_milestone_network) %>% tibble(milestone_network = ., model = models)
+topologies <- eval(formals(dyntoy:::generate_toy_milestone_network)$model)
+topologies <- topologies[topologies != "BA"]
+networks_synthetic <- map(topologies, dyntoy:::generate_toy_milestone_network) %>% tibble(milestone_network = ., topology = topologies)
 ```
@@
 ```{r}
-descriptions <- list(description_scorpius(), description_dpt())
+descriptions <- list(description_scorpius())
 data <- dyntoy::generate_toy_datasets(c("linear"), num_replicates=1, num_cells=30, num_genes=10) %>% dynutils::extract_row_to_list(1)
@@
 
-networks_methods <- tibble(model=map_chr(descriptions, "name"), milestone_network=map(results, "milestone_network"))
-
+networks_methods <- tibble(topology=map_chr(descriptions, "name"), milestone_network=map(results, "milestone_network"))
 ```
@@
 ```{r fig.height=7}
-scores <- design %>% as.list() %>% pmap(function(milestone_network, milestone_network1, model, model1, ...) {
-  # print(model)
-  # print(model1)
+scores <- design %>% as.list() %>% pmap(function(milestone_network, milestone_network1, topology, topology1, ...) {
+  # print(topology)
+  # print(topology1)
 
@@
 
-design %>% ggplot() + geom_raster(aes(model, model1, fill=score)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
+design %>% 
+  ggplot() + 
+  geom_raster(aes(topology, topology1, fill=score)) + 
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+```
+
+```{r}
+exclude_topologies <- c("SCORPIUS", "linear_long")
+edge_flip_comparison <- design %>% 
+  filter(!topology %in% exclude_topologies, !topology1 %in% exclude_topologies) %>% 
+  mutate(topology = factor(topology, levels=topologies)) %>% 
+  mutate(topology1 = factor(topology1, levels=rev(topologies))) %>% 
+  ggplot(aes(topology, topology1)) + 
+  geom_tile(aes(fill=score)) + 
+  geom_text(
+    aes(
+      label=round(score, 2) %>% gsub("0\\.", "\\.", .),
+      color = ifelse(score > 0.5, "white", "black")
+      )
+  ) +
+  viridis::scale_fill_viridis("Edge flip score", option="A", direction = -1) +
+  scale_color_identity() +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
+  scale_x_discrete("", labels=label_long, position="top", expand=c(0, 0)) +
+  scale_y_discrete("", labels=label_long, expand=c(0, 0)) +
+  theme(axis.text.x.top = element_text(hjust=0)) +
+  coord_equal()
+edge_flip_comparison
+write_rds(edge_flip_comparison, figure_file("edge_flip_comparison.rds"))
 ```
@@
 design %>%
-  group_by(model) %>%
+  group_by(topology) %>%
   arrange(-score) %>%
-  left_join(networks %>% select(-milestone_network), by=c("model1"="model")) %>%
+  left_join(networks %>% select(-milestone_network), by=c("topology1"="topology")) %>%
   summarise(plots = list(map2(plot, score, ~.x + ggtitle(round(.y, 2))))) %>%

### 2018-02-06 analysis/paper/paper.Rmd

@@
 # Discussion
+Limitations of the model:
+
+* Robrecht.Cannoodt @robrechtc 10:21
+en heterogene output van alle methode... true, maar sommigen hebben we wel wat moeten tweaken 😛
+en ik kan me output bedenken dat niet in ons model zou kunnen passen
+* Wouter Saelend @wouters 10:21
+oh 😛
+* Robrecht.Cannoodt @robrechtc 10:21
+welja
+* Wouter Saelend @wouters 10:21
+Misschien een kritiekpunt is onzekerheid... maarja 😛
+* Robrecht.Cannoodt @robrechtc 10:22
+onzekerheid, en dat een cell in meerdere niet-neighbouring milestones zou kunnen zitten
+als in, biologisch gezien houdt het mss geen steek, maar een methode zou wel kunnen zeggen, "wete, ik dénk dat het ofwel een DC is of een rode bloedcel"
+* Wouter Saelend @wouters 10:24
+hmmm uhu true, hoort allemaal wat samen met de onzekerheid. Je zou per cell een probabiliteitsdistributie doorheen het milestone netwerk kunnen geven, maar sla dat eens op 😛
+* Robrecht.Cannoodt @robrechtc 10:25
+uhu ^^
+* Wouter Saelend @wouters 10:25
+En bereken dan eens de emd 😛
+* Robrecht.Cannoodt @robrechtc 10:25
+idd...
+deze discussie zou wel ergens een plaatsje kunnen krijgen in het manuscript, vind ik
+* Wouter Saelend @wouters 10:25
+Uhu in de discussie dan wss
+* Robrecht.Cannoodt @robrechtc 10:26
+tiens, een discussie die in de discussie zou plaats vinden, wat een goed idee 😮
+<tease>
 
@@
 
-Depending on the output of a method, we used different strategies to convert the output to our common model:
+Depending on the output of a method, we used different common strategies to convert the output to our common model (in some cases combined with a special conversion denoted by a *):
 
-* Linear trajectory methods usually return a single pseudotime value, which can be converted to our model using a single edge between two milestones
-* Conversely, methods which assign each cell to a branch together with a pseudotime value and a branch network easily converted into our model
-* Methods returning both a global pseudotime value and a probability for every end state were converted into our model by using one single begin milestone and an end milestone for every end state.
-The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities
-Special conversions were necessary for certain method
-* DPT: 
-* Sincell
-* SLICER
-* SLICE
-*
+```{r}
+list_sentence <- function(x) {pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))}
+conversion <- function(conversion_oi) {
+    methods_evaluated %>% filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% mutate(text = paste0(name, ifelse(conversion_special, "\\*", ""))) %>% pull(text) %>% list_sentence()
+}
+```
 
+* Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
+* Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`.
+Similarly, the branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
+* Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`.
+We use a single start state and an add an edge to every end milestone each representing an end state.
+The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
+* Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`.
+The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
+* Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`.
+We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction.
+This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.
+
+Special conversions were necessary for certain methods:
+
+* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to
+* Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
+This was then processed
+* SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network.
+We therefore used the cell graph provided by the method to calculate the shortest paths between each branch, and then projected other cells onto this backbone graph structure, creating a local pseudotime for every cell within its branch.
+The output was then further processed as above.
+* SLICE: As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
+* MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
 ## Method quality control

### 2018-02-06 analysis/paper/paper.Rmd

@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface `r ref("table", "methods")`.
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`).
 We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods `r ref("table", "methods", anchor=TRUE)`, based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "methods", anchor=TRUE)`), based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
 We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
@@
 We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its heterogenous output into a common probabilistic trajectory model.
-This model consists of two parts.
+This model consisted of two parts.
 The milestone network represents the overall network topology, and contains of edges between different milestones and the length of the edge between them.
@@
 
-Depending on the output of a method, we used different common strategies to convert the output to our common model (in some cases combined with a special conversion denoted by a *):
+Depending on the output of a method, we used different common strategies to convert the output to our common model (in some cases combined with a special conversion denoted by a \*):
 
@@
 * MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
+
 ## Method quality control

### 2018-02-06 analysis/paper/paper.Rmd
update paper

@@
 This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a framework of data topologies for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available.
-In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
-We employ a novel parameter optimisation procedure, and We compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness.
+We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
+Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
+We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
 We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
@@
 Because they offer an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
+This both for specific subsets of cells, but in the near future also for whole tissues or organisms [@regev_scienceforumhuman_2017].
 
@@
 methods <- readRDS(derived_file("methods.rds", experiment_id="method_characteristics"))
+methods_evaluated <- readRDS(derived_file("methods_evaluated.rds", experiment_id="method_characteristics"))
 ```
@@
 At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`).
-However, only a handful of more recent methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).
-These categorisations can then be used to help current developers to improve their tool, and to guide the selection of users and developers to use these tools for their purpose.
+These categorisations can help current developers to improve their tool, and to guide the selection of users and developers to use these tools for their purpose.
 After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
@@
 ```
-
-Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
-Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding studies pinpointing these as good practices [@wilson_best_2014].
-
-The underlying algorithms can be divided into several modules, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps.
-Interestingly, modules are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
 
@@
 
-```{r}
-add_sfig({
-readRDS(figure_file("method_components.rds", experiment_id="method_characteristics"))
-}, 
-"method_components", 
-"TI methods frequently share steps", 
-15, 
-7.5
-)
-```
+Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
+Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
+However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 ```
-## Evaluation overview
+
+The underlying algorithms can be divided into several modules, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps.
+Interestingly, modules are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
+
+```{r}
+add_sfig({
+readRDS(figure_file("method_components.rds", experiment_id="method_characteristics"))
+}, 
+"method_components", 
+"TI methods frequently share steps", 
+15, 
+7.5
+)
+```
+## Evaluating the accuracy of the inferred trajectory
 
@@
 
-To classify the output of methods and the gold standards, we defined a set of basic trajectory types.
-These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph using certain graph properties, progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
+We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
+These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph, progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
 For every type, a corresponding directed graph type can be defined by allowing directedness (`r ref("sfig", "trajectory_type_tree", "b")`).
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "methods", anchor=TRUE)`), based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering".
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "methods", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
 We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
@@
 
-We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its heterogenous output into a common probabilistic trajectory model.
+Due to the absence of a common format for trajectory inference methods, most methods produce a heterogeneous output.
+We therefore created wrappers around each method in R (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model.
 This model consisted of two parts.
-The milestone network represents the overall network topology, and contains of edges between different milestones and the length of the edge between them.
+The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
 The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
 
-Depending on the output of a method, we used different common strategies to convert the output to our common model (in some cases combined with a special conversion denoted by a \*):
+Depending on the output of a method, we used different strategies to convert the output to our model (in some cases combined with a special conversion denoted by a \*):
 
@@
 
-We created a transparent scoring scheme to check the quality of each method (`r ref("table", "methods")`).
+We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online.
+The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
 The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
-The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
 The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
@@
 For each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "methods")`).
-This was to prevent make sure more important aspect, such as the open source availability of the method, to outweigh other aspects, such as the availability of a graphical user interface.
-For calculating the final score, we however weighed each of the six categories equally.
+This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
+For calculating the final score, we weighed each of the six categories equally.
 
@@
 
-One of the main process involved in cellular dynamic processes is through gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
+One of the main process involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
@@
 This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no suitable [@bahiense_maximumcommonedge_2012].
-We implemented a branch&bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes.
+We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes.
 To further limit the search space, we make sure the degree distributions between the two networks are similar, before assessing whether the two networks are isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

### 2018-02-07 analysis/analyses/2-dataset_characterisation/2_dataset_plotting.R
rename experiment ids, and allow the number of citations to be updated to the google sheet

@@ -1,320 +1,0 @@
-library(tidyverse)
-library(cowplot)
-library(googlesheets)
-library(dynalysis)
-
-experiment("dataset_characterisation")
-
-tasks <- read_rds(derived_file("tasks.rds"))
-
-
-load_expression <- function(expression) {
-  if(class(expression) == "matrix") {
-    expression
-  } else {
-    expression()
-  }
-}
-
-# Check the tasks ------------------------------------
-# task_checks <- map_prism(task_ids, function(task_id) {
-task_checks <- pbapply::pblapply(tasks$id, function(task_id) {
-  print(task_id)
-  task <- extract_row_to_list(tasks, which(tasks$id == task_id))
-  expression <- load_expression(task$expression)
-  counts <- load_expression(task$counts)
-
-  checks <- list(id = task_id)
-
-  checks$all_milestones_represented <- all(task$milestone_ids %in% task$prior_information$grouping_assignment$group_id)
-
-  checks$marker_feature_ids_in_expression <- all(task$prior_information$marker_feature_ids %in% colnames(expression))
-
-  checks$columns <- all(c("geodesic_dist") %in% names(task))
-
-  checks$n_genes <- nrow(task$feature_info)
-  checks$n_cells <- nrow(task$cell_info)
-
-  checks
-}) %>% bind_rows()
-
-task_checks %>% bind_cols(tasks) %>% ggplot() + geom_point(aes(n_genes, n_cells, color=category))
-
-task_ids_filtered <- task_checks %>% filter(all_milestones_represented) %>% pull(task_id)
-tasks_filtered <- tasks %>% filter(id %in% task_ids_filtered)
-
-load_expression <- function(expr) {
-  if(class(expr) == "function") {
-    expr()
-  } else {
-    expr
-  }
-}
-
-# Score the tasks ---------------------------------
-task_scores <- pmap(as.list(tasks_filtered), function(...) {
-  # task <- extract_row_to_list(tasks, 20)
-  task <- list(...)
-  library(tidyverse)
-  print(paste0("-------------", task$id))
-  print("loading")
-
-  task$expression <- load_expression(task$expression)
-
-  Coi <- rowMeans(task$expression) %>% {(. > (median(.) - IQR(., 0.5)*2)) & (. < (median(.) + IQR(., 0.5)*2))} %>% keep(~.) %>% names
-
[…]

### 2018-02-07 analysis/analyses/4-method_characterisation/1_update_citations.R

@@ -1,0 +1,20 @@
+# This file will update the number of citations in the google sheets
+library(tidyverse)
+library(googlesheets)
+library(dynalysis)
+
+experiment("4-method_characterisation")
+
+# Downloading -----------------------
+# # If it's your first time running this script, run this:
+# gs_auth()
+
+sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE")
+
+methods <- sheet %>%
+  gs_read(ws = "Software", col_types = cols(GScholarClusterID = "c"), skip = 1)
+
+ncitations <- pbapply::pbsapply(methods$GScholarClusterID, dynutils::google_scholar_num_citations)
+ncitations[is.na(ncitations)] <- ""
+
+gs_edit_cells(sheet, input = ncitations, byrow = FALSE, anchor = "AK3")

### 2018-02-07 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
rename experiment ids, and allow the number of citations to be updated to the google sheet

--- 2018-02-06 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
@@
 
+```{r setup}
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+library(dynalysis)
+library(tidyverse)
+library(cowplot)
+```
+
 ```{r message=F}
-library(tidyverse)
-library(dynalysis)
-library(cowplot)
-
-experiment("method_characteristics")
+experiment("4-method_characterisation")
 
@@
 
-g2 <- ggplot(methods_evaluated %>% mutate(ncitations = ifelse(ncitations == 0, 0.5, ncitations)), aes(date, ncitations)) +
+g2 <- ggplot(methods_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(date, Citations)) +
     geom_smooth(span=2) +
@@
 
-g3 <- ggplot(methods_evaluated %>% mutate(ncitations = ifelse(ncitations == 0, 0.5, ncitations)), aes(ncitations, qc_score)) +
+g3 <- ggplot(methods_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(Citations, qc_score)) +
     geom_smooth(span=2) +

### 2018-02-07 analysis/analyses/2-dataset_characterisation/wip-dataset_plotting.R

@@ -1,0 +1,321 @@
+# library(tidyverse)
+# library(cowplot)
+# library(googlesheets)
+# library(dynalysis)
+#
+# experiment("2-dataset_characterisation")
+#
+# tasks <- read_rds(derived_file("tasks.rds"))
+#
+#
+# load_expression <- function(expression) {
+#   if(class(expression) == "matrix") {
+#     expression
+#   } else {
+#     expression()
+#   }
+# }
+#
+# # Check the tasks ------------------------------------
+# # task_checks <- map_prism(task_ids, function(task_id) {
+# task_checks <- pbapply::pblapply(tasks$id, function(task_id) {
+#   print(task_id)
+#   task <- extract_row_to_list(tasks, which(tasks$id == task_id))
+#   expression <- load_expression(task$expression)
+#   counts <- load_expression(task$counts)
+#
+#   checks <- list(id = task_id)
+#
+#   checks$all_milestones_represented <- all(task$milestone_ids %in% task$prior_information$grouping_assignment$group_id)
+#
+#   checks$marker_feature_ids_in_expression <- all(task$prior_information$marker_feature_ids %in% colnames(expression))
+#
+#   checks$columns <- all(c("geodesic_dist") %in% names(task))
+#
+#   checks$n_genes <- nrow(task$feature_info)
+#   checks$n_cells <- nrow(task$cell_info)
+#
+#   checks
+# }) %>% bind_rows()
+#
+# task_checks %>% bind_cols(tasks) %>% ggplot() + geom_point(aes(n_genes, n_cells, color=category))
+#
+# task_ids_filtered <- task_checks %>% filter(all_milestones_represented) %>% pull(task_id)
+# tasks_filtered <- tasks %>% filter(id %in% task_ids_filtered)
+#
+# load_expression <- function(expr) {
+#   if(class(expr) == "function") {
+#     expr()
+#   } else {
+#     expr
+#   }
+# }
+#
+# # Score the tasks ---------------------------------
+# task_scores <- pmap(as.list(tasks_filtered), function(...) {
+#   # task <- extract_row_to_list(tasks, 20)
+#   task <- list(...)
+#   library(tidyverse)
+#   print(paste0("-------------", task$id))
+#   print("loading")
+#
+#   task$expression <- load_expression(task$expression)
+#
[…]

### 2018-02-07 analysis/analyses/4-method_characterisation/3_process_qc.R

@@ -2,7 +2,7 @@ library(tidyverse)
 library(googlesheets)
 library(dynalysis)
 
-experiment("method_characteristics")
+experiment("4-method_characterisation")
 
 
 ##  ............................................................................

### 2018-02-07 analysis/analyses/3-score_rules/1-generate_toys.Rmd
refactor score_rules

--- 2018-01-23 analysis/analyses/3-score_rules/1-generate-toys.Rmd
@@
 
-experiment("score_rules")
+experiment("3-score_rules/1-generate_toys")
 ```
@@
 
-metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "isomorphic", "node_edit_score", "node_edge_edit_score", "mantel_pval", "ged")
+metrics <- c("mean_R_nx", "auc_R_nx", "Q_local", "Q_global", "correlation", "edge_flip", "rf_mse", "rf_rsq")
 
@@
   },
-  qsub_config = override_qsub_config(num_cores = 24),
-  qsub_environment = list2env(list(compare_toy, compare_gs_toy, toys)
-))
+  qsub_config = override_qsub_config(
+    num_cores = 24, 
+    execute_before = "source /scratch/irc/shared/dynverse/module_load_R.sh; export R_MAX_NUM_DLLS=500"
+  )
+)
 
@@
 
-metric_groups <- c("mean_R_nx"="coranking", "auc_R_nx"="coranking", "Q_local"="coranking", "Q_global"="coranking", "correlation"="correlation", "isomorphic"="isomorphic", "robbie_network_score"="robbie_network_score", "mantel_pval"="mantel", "ged"="ged") %>%
+metric_groups <- c("mean_R_nx"="coranking", "auc_R_nx"="coranking", "Q_local"="coranking", "Q_global"="coranking", "correlation"="correlation", "edge_flip"="edge_flip", "rf_mse"="rf", "rf_msq"="rf") %>%
   enframe("metric_id", "metric_group_id")

### 2018-02-07 analysis/paper/paper.Rmd

@@
 
+This is the first overview about ~~deep reinforcement learning~~ trajectory inference publicly available online. It is comprehensive. Comments and criticisms are welcome.
+
 # Introduction
@@
 add_sfig({
-n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
-trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_id="method_characteristics"))
+n_methods_over_time <- read_rds(figure_file("n_methods_over_time.rds", experiment_id="4-method_characterisation"))
+trajectory_components_over_time <- read_rds(figure_file("trajectory_components_over_time.rds", experiment_id="4-method_characterisation"))
 
@@
 <!-- history of Ti, different methods -->
+```{r}
+fixed_linear <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
+fixed_bifurcating <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
+fixed_other <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
+fixed_algorithm <-methods$name[which(methods$fixes_structure == 'algorithm')]
+fixed_parameters <-methods$name[which(methods$fixes_structure == 'parameter')]
+```
+
 Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
-Initially, most trajectory inference fixed the trajectory topology to a linear or bifurcating topology, by either algorithmic limitations or through user parameters, and focussed on correctly ordering the cells along this fixed topology.
-However, several recent methods now infer both the ordering and the topology at the same time, although certain structural limitations are still imposed(`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the topology are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`.
+The focus of these methods was therefore mainly to correctly ordering the cells along this fixed topology.
+However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`).
+Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
@@
 Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
-Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become scalable and user friendly .
+Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
+Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017] and user friendly.
 
@@
 
-```{r}
-# Load method characteristics
-methods <- readRDS(derived_file("methods.rds", experiment_id="method_characteristics"))
-methods_evaluated <- readRDS(derived_file("methods_evaluated.rds", experiment_id="method_characteristics"))
-```
-
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`).
-We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).
+We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "4-method_characterisation")`).
 
@@
 ```{r}
-read_rds(figure_file("methods_table.rds", "method_characteristics"))
+read_rds(figure_file("methods_table.rds", "4-method_characterisation"))
 ```
 
-At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`).
+At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "4-method_characterisation", "a")`).
 However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
@@
 add_stable(
-read_rds(figure_file("qc_checks.rds","method_characteristics")), 
+read_rds(figure_file("qc_checks.rds","4-method_characterisation")), 
 "qc_checks", 
@@
 
-`r anchor("fig", "method_characteristics")`
-![](`r figure_file("method_characteristics.svg", experiment_id="method_characteristics")`)
-`r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.
+`r anchor("fig", "4-method_characterisation")`
+![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
+`r ref("fig", "4-method_characterisation", anchor=TRUE)` First characterization of each method.
 
@@
 add_sfig({
-readRDS(figure_file("method_qc_check.rds", experiment_id="method_characteristics"))
+read_rds(figure_file("method_qc_check.rds", experiment_id="4-method_characterisation"))
 }, 
@@
 
-Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`).
+Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "4-method_characterisation", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "4-method_characterisation", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "4-method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 add_sfig({
-readRDS(figure_file("qc_over_time.rds", experiment_id="method_characteristics"))
+read_rds(figure_file("qc_over_time.rds", experiment_id="4-method_characterisation"))
 }, 
@@
 add_sfig({
-readRDS(figure_file("method_components.rds", experiment_id="method_characteristics"))
+read_rds(figure_file("method_components.rds", experiment_id="4-method_characterisation"))
 }, 
@@
 add_stable(
-read_rds(figure_file("datasets.rds","dataset_characterisation")), 
+read_rds(figure_file("datasets.rds","2-dataset_characterisation")), 
 "datasets", 
@@
 # Discussion
+
+Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
+Indeed, it is for this category that most methods were developed among other single-cell analysis categories, both according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category) and the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis.
+In this preprint, we summarise the initial version of our evaluation of trajectory inference methods, focussing on the quality control
+
 Limitations of the model:
 
+```
 * Robrecht.Cannoodt @robrechtc 10:21
@@
 <tease>
+```
 
@@
 add_sfig({
-readRDS(figure_file("trajectory_type_trees.rds", experiment_id="dataset_characterisation/trajectory_types"))
+read_rds(figure_file("trajectory_type_trees.rds", experiment_id="2-dataset_characterisation/2-trajectory_types"))
 }, 
@@
 * Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`.
-We use a single start state and an add an edge to every end milestone each representing an end state.
+We use a single start state and add an edge to every end milestone each representing an end state.
 The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
@@
 
-* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to
+* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
+This was then processed as described above.
 * Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
-This was then processed
+This was then processed as described above.
 * SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network.
@@
 add_sfig({
-readRDS(figure_file("modulenets.rds", experiment_id="dataset_characterisation/synthetic"))
+read_rds(figure_file("modulenets.rds", experiment_id="2-dataset_characterisation/3-synthetic"))
 }, 
@@
 add_stable(
-read_rds(figure_file("samplers.rds","dataset_characterisation/synthetic")), 
+read_rds(figure_file("samplers.rds","2-dataset_characterisation/3-synthetic")), 
 "samplers", 
@@
 add_sfig({
-readRDS(figure_file("score_aspects.rds", experiment_id="score_rules/score_aspects"))
+read_rds(figure_file("score_aspects.rds", experiment_id="3-score_rules/5-score_aspects"))
 }, 
@@
 add_sfig({
-readRDS(figure_file("edge_flip_comparison.rds", experiment_id="score_rules/edge_flip"))
+read_rds(figure_file("edge_flip_comparison.rds", experiment_id="3-score_rules/4-edge_flip"))
 }, 

### 2018-02-08 analysis/analyses/manual_ti/guide.Rmd
add wip guide

---
title: "Guide for manual trajectory inference"
output: html_document
---

```{r setup, include=FALSE}
library(dynalysis)
library(tidyverse)

tasks_control <- read_rds(derived_file("tasks.rds", "1-datasets/control"))
space <- tasks_control %>% filter(id == "control_bifurcating") %>% pull(space) %>% first()
```

Shown are individual cells within a single-cell RNA-seq dataset in 2D. Cells closer to each other have more similar expression profiles compared to cells farther apart.

```{r cars}
annotation <- tribble(
  ~x1, ~x2, ~y1, ~y2, ~label,
  0.5, 0.8, 0.8, 0.75, "",
  0.5, 0.8, 0.75, 0.75, "Very similar cells",
  0., 0.25, 0., 0.5, "",
  0.5, 0.25, 0.8, 0.5, "Very different cells"
)
space %>% 
  ggplot(aes(Comp1, Comp2)) +
    geom_point() +
    geom_segment(aes(x1, y1, xend=x2, yend=y2), data=annotation) +
    geom_label(aes(x2, y2, label=label), data=annotation %>% filter(label != "")) +
    geom_point(aes(x1, y1), color="red", data=annotation) +
    ggraph::theme_graph()
```

In all datasets, cells were sampled while they were going through a cellular dynamic process, such as the cell cycle, differentiation, activation, ... However, some cells are earlier in this process, others are later.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2018-02-08 analysis/paper/paper.Rmd
update paper

@@
 ```{r}
-fixed_linear <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
-fixed_bifurcating <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
-fixed_other <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
-fixed_algorithm <-methods$name[which(methods$fixes_structure == 'algorithm')]
-fixed_parameters <-methods$name[which(methods$fixes_structure == 'parameter')]
+fixed_linear <-methods$id[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
+fixed_bifurcating <-methods$id[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
+fixed_other <-methods$id[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
+fixed_algorithm <-methods$id[which(methods$fixes_structure == 'algorithm')]
+fixed_parameters <-methods$id[which(methods$fixes_structure == 'parameter')]
 ```
@@
 We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
-The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).
+The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or which purpose(s) they serve (user friendliness, developer friendliness or good science).
 These categorisations can help current developers to improve their tool, and to guide the selection of users and developers to use these tools for their purpose.
@@
 
+We evaluated
+
 `r anchor("fig", "evaluation_overview")`
@@
 # Discussion
-
+<!-- Some intro -->
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
 Indeed, it is for this category that most methods were developed among other single-cell analysis categories, both according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category) and the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis.
-In this preprint, we summarise the initial version of our evaluation of trajectory inference methods, focussing on the quality control
+It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
+In this preprint, we summarise the initial version of our evaluation of these methods, focussing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
 
+<!-- Results  -->
+
+<!-- Our common output model, advantages & limitations -->
+We managed to wrap the output of all methods into one common format.
+This not only allowed us to compare different methods with a gold standard, but could also be useful for users of these tools because it allows the user to test multiple methods without manual conversion of input and output.
+Furthermore, it makes it possible to compare the output of different methods, which opens up possibilities for new comparative visualisation techniques or ensemble methods.
+However, we acknowledge that our model has some limitations.
+It can currently not take into account uncertainty of the cell's position.
+This can be both locally, for example uncertainty of its position within one branch, or globally, for example uncertainty over multiple branches.
+Some current methods already model this uncertainty in some way, mainly locally `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
+
+<!----
 Limitations of the model:
-
-```
 * Robrecht.Cannoodt @robrechtc 10:21
@@
 onzekerheid, en dat een cell in meerdere niet-neighbouring milestones zou kunnen zitten
-als in, biologisch gezien houdt het mss geen steek, maar een methode zou wel kunnen zeggen, "wete, ik dénk dat het ofwel een DC is of een rode bloedcel"
+als in, biologisch gezien houdt het mss geen steek, maar een methode zou wel kunnen zeggen, "wete, ik denk dat het ofwel een DC is of een rode bloedcel"
 * Wouter Saelend @wouters 10:24
-hmmm uhu true, hoort allemaal wat samen met de onzekerheid. Je zou per cell een probabiliteitsdistributie doorheen het milestone netwerk kunnen geven, maar sla dat eens op 😛
+hmmm uhu true, hoort allemaal wat samen met de onzekerheid. Je zou per cell een probabiliteit distributie doorheen het milestone netwerk kunnen geven, maar sla dat eens op 😛
 * Robrecht.Cannoodt @robrechtc 10:25
@@
 <tease>
-```
-
+-->
 # Methods
@@
 ```{r}
+# create functions to list the methods within a particular conversion category
 list_sentence <- function(x) {pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))}

### 2018-02-09 analysis/paper/paper.Rmd
update paper

@@
 * Add extra enters after every paragraph (this is markdown!)
+* Enable automatic zotero syncing to bibtex
+* Make sure to pin the zotero bibtex references after first using them!
 
@@
 
-The underlying algorithms can be divided into several modules, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps.
-Interestingly, modules are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
+The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps.
+Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
+MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).
 
@@
 
-We evaluated
+We wrapped each method and postprocessed its output into a common probabilistic format.
+This format works at two level: at the To evaluate the accuracy of each method, we gathered both synthetic and real datasets, and used these to assess how good a method is at correctly ordering cells, detecting the right trajectory topology and 2
 
 `r anchor("fig", "evaluation_overview")`  
-![](`r figure_file("ti_evaluation_overview_v1.svg", experiment_id="evaluation_overview")`)
+![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)  
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, it is for this category that most methods were developed among other single-cell analysis categories, both according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category) and the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis.
+Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) .
 It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.

### 2018-02-09 analysis/paper/paper.Rmd
update paper

@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`).
-We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "4-method_characterisation")`).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`).
+We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).
 
@@
 
-At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "4-method_characterisation", "a")`).
+`r anchor("fig", "evaluation_overview")`  
+![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)  
+`r ref("fig", "evaluation_overview")` Overview of the evaluation
+
+```{r}
+add_stable(
+read_rds(figure_file("datasets.rds","2-dataset_characterisation")), 
+"datasets", 
+"Real datasets used in this study"
+)
+```
+
+At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`).
 However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
@@
 
-`r anchor("fig", "4-method_characterisation")`
-![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
-`r ref("fig", "4-method_characterisation", anchor=TRUE)` First characterization of each method.
+`r anchor("fig", "method_characterisation")`
+![](`r figure_file("method_characteristics.svg", experiment_id="method_characterisation")`)
+`r ref("fig", "method_characterisation", anchor=TRUE)` First characterization of each method.
 
@@
 
-Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "4-method_characterisation", "b")`).
+Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "4-method_characterisation", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "4-method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characterisation", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 This format works at two level: at the To evaluate the accuracy of each method, we gathered both synthetic and real datasets, and used these to assess how good a method is at correctly ordering cells, detecting the right trajectory topology and 2
-
-`r anchor("fig", "evaluation_overview")`  
-![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)  
-`r ref("fig", "evaluation_overview")` Overview of the evaluation
-
-```{r}
-add_stable(
-read_rds(figure_file("datasets.rds","2-dataset_characterisation")), 
-"datasets", 
-"Real datasets used in this study"
-)
-```
-
 # Discussion

### 2018-02-12 analysis/analyses/manual_ti/guide.Rmd
update manual ti
updated guide

@@
 
-```{r cars}
+```{r, echo=F}
 annotation <- tribble(
@@
 In all datasets, cells were sampled while they were going through a cellular dynamic process, such as the cell cycle, differentiation, activation, ... 
-However, some cells are earlier in this process, others are later.
+At some points during this process, cells make decisions about different paths they can take. At other points, they could converge towards very similar cell types.
 
-You can also embed plots, for example:
+When we assume that cells go through this process gradually, the expression changes will follow a continuum, and this will be reflected in the 2D visualisation of the cells.
+Cells will follow paths within this 2D visualisation, but due to heterogeneity, will not exactly follow the path.
+See for example the red path in the following visualisation;
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r, echo=F}
+graph <- tribble(
+  ~x1, ~y1, ~x2, ~y2,
+  0.5, 1, 0.5, 0.35,
+  0.5, 0.35, 1, 0,
+  0.5, 0.35,0,0
+)
+space %>% 
+  ggplot(aes(Comp1, Comp2)) +
+    geom_point() +
+    geom_segment(aes(x1, y1, xend=x2, yend=y2), data=graph, color="red", size=3,alpha=0.6) +
+  ggraph::theme_graph()
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+Your goal is to use this visualisation to draw the paths the cells follow. These can contain cycles, multiple splits and convergences.
+
+Open the svg file in inkscape and use these tools to draw the lines:
+
+![](pen.png)
+To draw lines
+
+![](select.png)
+To edit lines
+
+![](arrow.png)
+To select lines (and delete them with the delete key)
+
+You don't have to define an end or start point. When drawing cycles, splits or convergences, two points don't have to lie at exactly the same position, as they will be mergen computationally.
+
+When ready, send the svg file back to wouter.saelens@ugent.be

### 2018-02-19 analysis/paper/paper.Rmd
update paper

@@
 ```{r setup, include=FALSE}
-source("setup.R")
+source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
 ```
@@
 Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
-Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
+This makes a direct[a] comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
+We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.[b] Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
 We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
+We conclude with practical guidelines for method[c] users, and also believe our evaluation framework can be used to spearhead the development for new methods.
 Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.
@@
 
-Single-cell -omics technologies now make it possible to more accurately model biological systems than ever before.
+Single-cell -omics technologies now make it possible to more accurately[d] model biological systems than ever before.
 One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
-When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Because they offer an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
-This both for specific subsets of cells, but in the near future also for whole tissues or organisms [@regev_scienceforumhuman_2017].
+When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[e] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
+Because they offer an unbiased and transcriptome-wide understanding of the dynamic process[f], trajectories then allow the identification of new [g](primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
+This both for specific subsets of cells, but in the near future also for whole tissues or organisms[h] [@regev_scienceforumhuman_2017].
 
@@
 Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`.
-The focus of these methods was therefore mainly to correctly ordering the cells along this fixed topology.
+The focus of these methods was therefore mainly to correctly ordering[i] the cells along this[j] fixed topology.
 However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to[k] model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
 <!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Although trajectory inference methods use a variety of algorithms and steps to reach a final ordering, most consist of the following three main steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
 Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
-Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017] and user friendly.
+Furthermore, with the ongoing scalability in single-cell -omics technologies (reaching a million cells [@angerer_singlecellsmake_2017], it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user friendly.
 
 <!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[l].
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods.
-We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
+In this study, we therefore for the first time developed a[m] comprehensive evaluation framework for trajectory inference methods.
+We test[n] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
 We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`).
-We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "methods")`).
+We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).
 
@@
 At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`).
-However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+However, only a handful, mostly recent[o], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 `r anchor("fig", "method_characterisation")`
-![](`r figure_file("method_characteristics.svg", experiment_id="method_characterisation")`)
+![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
 `r ref("fig", "method_characterisation", anchor=TRUE)` First characterization of each method.
@@
 
-Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`).
+Most[p] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
@@
 
-The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building steps.
+The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
 Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
@@
 
-We wrapped each method and postprocessed its output into a common probabilistic format.
-This format works at two level: at the To evaluate the accuracy of each method, we gathered both synthetic and real datasets, and used these to assess how good a method is at correctly ordering cells, detecting the right trajectory topology and 2
+We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
+To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
+In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
+This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
+We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
+We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
+To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
+To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
+
+```{r}
+add_sfig({
+                glue::glue("![]({figure_file('wrapping.svg', experiment_id = '4-method_characterisation')})")
+}, 
+"wrapping", 
+"Our wrapping workflow", 
+15, 
+7.5
+)
+```
+
 # Discussion
@@
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) .
-It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
-In this preprint, we summarise the initial version of our evaluation of these methods, focussing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
+Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) .
+It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice[q].
+In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
 
@@
 * SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network.
-We therefore used the cell graph provided by the method to calculate the shortest paths between each branch, and then projected other cells onto this backbone graph structure, creating a local pseudotime for every cell within its branch.
+We therefore used the cell graph provided by the method to calculate the shortest paths between the start and end cells found by the method, and then used this graph as the backbone for all other cells, creating a local pseudotime for every cell within its branch.
 The output was then further processed as above.
@@
 ```
+[a]This makes it difficult to directly compare between trajectory inference methods , and as a result (...)
+[b]which facilitates the comparison of the results of trajectory inference methods (start using the TI abbreviation at some point?) between each other, and possibly to a gold standard.
+[c]which might come in handy to users. We believe (...)
+[d]move? "make it possible to model biological systems more accurately than ever before".
+[e]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
+Trajectory inference methods can be used to computationally order these cells along trajectories, based on gradual changes in the expression of several genes.
+[f]processes?
+[g]the trajectories can allow to identify new (primed) subsets of cells, to delineate the exact wiring of a differentiation tree and to infer the regulatory interactions which are responsible for a bifurcation.
+[h]This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future.
+[i]order
+[j]a?
+[k]will be able to model any underlying structure in the data, without the need of user provided parameters. These methods might allow to model more complex behaviors, such as (...)
+[l]get an idea of which available method or set of methods would optimally resolve their problem.
+[m]we present the first comprehensive (...)?
+[n]compared 22 TI methods on both synthetic (..)
+[o]a handful of the most recent methods
+[p]Repetition of most: put "The majority of" instead?
+[q]analysis?

### 2018-02-23 analysis/analyses/4-method_characterisation/7_quality_control.Rmd

@@
 ---
-title: "Method quality control"
+title: "implementation quality control"
 output:
@@
 
-method_qc <- readRDS(derived_file("method_qc.rds"))
-methods_evaluated <- readRDS(derived_file("methods_evaluated.rds"))
+implementation_qc <- readRDS(derived_file("implementation_qc.rds"))
+implementations_evaluated <- readRDS(derived_file("implementations_evaluated.rds"))
 checks <- readRDS(derived_file("checks.rds"))
@@
 ```{r, fig.width=16, fig.height=8}
-g1 <- ggplot(methods_evaluated, aes(date, qc_score)) +
+g1 <- ggplot(implementations_evaluated, aes(date, qc_score)) +
     geom_smooth(span=2) +
     geom_point() +
-    ggrepel::geom_label_repel(aes(label = name)) +
+    ggrepel::geom_label_repel(aes(label = implementation_name)) +
     cowplot::theme_cowplot() +
@@
 
-g2 <- ggplot(methods_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(date, Citations)) +
+g2 <- ggplot(implementations_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(date, Citations)) +
     geom_smooth(span=2) +
     geom_point() +
-    ggrepel::geom_label_repel(aes(label = name)) +
+    ggrepel::geom_label_repel(aes(label = implementation_name)) +
     cowplot::theme_cowplot() +
@@
 
-g3 <- ggplot(methods_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(Citations, qc_score)) +
+g3 <- ggplot(implementations_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(Citations, qc_score)) +
     geom_smooth(span=2) +
     geom_point() +
-    ggrepel::geom_label_repel(aes(label = name)) +
+    ggrepel::geom_label_repel(aes(label = implementation_name)) +
     cowplot::theme_cowplot() +
@@
 
-Calculate the overall qc scores for each method
+Calculate the overall qc scores for each implementation
 ```{r}
-# set the ordering of method_ids to this ordering
-method_ordering_plot <- methods_evaluated %>%
+# set the ordering of implementation_ids to this ordering
+implementation_ordering_plot <- implementations_evaluated %>%
   arrange(-qc_score) %>% 
-  mutate(method_id = factor(name, levels=name)) %>% 
-  ggplot(aes(method_id, qc_score)) +
+  mutate(implementation_id = factor(implementation_id, levels=implementation_id)) %>% 
+  ggplot(aes(implementation_id, qc_score)) +
     geom_bar(aes(fill=qc_score), stat="identity") +
@@
 
-method_ordering_plot
+implementation_ordering_plot
 ```
@@
 ```{r, fig.width=10, fig.height=12}
-qc_checks <- method_qc %>%
-  filter(method_id == "SCORPIUS") %>%
+qc_checks <- implementation_qc %>%
+  filter(implementation_id == "scorpius") %>%
   group_by(category) %>% 
@@
 
-check_heatmap_plot <- method_qc %>%
+check_heatmap_plot <- implementation_qc %>%
   left_join(check_position, by="check_id") %>%
-  mutate(method_id = factor(method_id)) %>%
+  mutate(implementation_id = factor(implementation_id)) %>%
   mutate(answer = ifelse(is.na(answer), 0, answer)) %>%
   ggplot() +
-    geom_rect(aes(xmax=check_start, xmin=check_end, ymin=as.integer(method_id), ymax=as.integer(method_id) + 1, fill=category, alpha=answer)) +
+    geom_rect(aes(xmax=check_start, xmin=check_end, ymin=as.integer(implementation_id), ymax=as.integer(implementation_id) + 1, fill=category, alpha=answer)) +
     geom_vline(aes(xintercept=check_end), data=check_position, color="#444444", alpha=0.25) +
@@
     scale_x_reverse("Check", breaks = aspect_position$aspect_mid, labels = aspect_position$name, expand = c(0, 0)) +
-    scale_y_continuous("Method", breaks = seq_along(levels(method_qc$method_id))+0.5, labels = levels(method_qc$method_id), expand = c(0, 0)) +
+    scale_y_continuous("implementation", breaks = seq_along(levels(implementation_qc$implementation_id))+0.5, labels = levels(implementation_qc$implementation_id), expand = c(0, 0)) +
     scale_fill_manual(values=set_names(qc_categories$color, qc_categories$category)) +
@@
 
-method_qc_check_plot <- cowplot::plot_grid(method_ordering_plot + empty_x, check_heatmap_plot, nrow=2, align = "v", axis = "rl", rel_heights = c(0.2, 0.9))
+implementation_qc_check_plot <- cowplot::plot_grid(implementation_ordering_plot + empty_x, check_heatmap_plot, nrow=2, align = "v", axis = "rl", rel_heights = c(0.2, 0.9))
 
-method_qc_check_plot %>% saveRDS(figure_file("method_qc_check.rds"))
+implementation_qc_check_plot %>% saveRDS(figure_file("implementation_qc_check.rds"))
 ```
@@
 
-Order by breakpoint (when ordering all methods by score)
+Order by breakpoint (when ordering all implementations by score)
 ```{r}
@@
 }
-check_breakpoints <- method_qc %>%
-  reshape2::acast(check_id~method_id, value.var="answer") %>%
+check_breakpoints <- implementation_qc %>%
+  reshape2::acast(check_id~implementation_id, value.var="answer") %>%
   {
@@
       check_id = as.integer(rownames(.)),
-      n_bad_methods = apply(., 1, sum)
+      n_bad_implementations = apply(., 1, sum)
     )
@@
 ```{r fig.width=20}
-method_qc %>% 
+implementation_qc %>% 
   mutate(answer = round(answer, 2)) %>% 
@@
     scale_y_reverse(expand=c(0, 0)) +
-    scale_x_discrete(labels=setNames(method_qc$item_weight, method_qc$check_id)) +
+    scale_x_discrete(labels=setNames(implementation_qc$item_weight, implementation_qc$check_id)) +
     scale_fill_distiller(palette="Blues", direction = 1)
@@
 ```{r}
-aspect_info <- method_qc %>% 
+aspect_info <- implementation_qc %>% 
   group_by(aspect_id) %>% 
@@
 
-aspect_ordering_plotdata <- method_qc %>%
-  group_by(method_id, aspect_id) %>%
+aspect_ordering_plotdata <- implementation_qc %>%
+  group_by(implementation_id, aspect_id) %>%
   summarise(
@@
 
-n_methods <- length(unique(method_qc$method_id))
+n_implementations <- length(unique(implementation_qc$implementation_id))
 
@@
     coord_flip() +
-    scale_y_continuous(expand=c(0, 0), name = "# methods", c(seq(0, n_methods, 5), n_methods)) +
+    scale_y_continuous(expand=c(0, 0), name = "# implementations", c(seq(0, n_implementations, 5), n_implementations)) +
     scale_x_discrete(labels=setNames(aspect_ordering_plotdata$name, aspect_ordering_plotdata$aspect_id), name="") +
@@
 
-method_qc_ordering_plot <- cowplot::plot_grid(category_legend, aspect_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
-method_qc_ordering_plot
-saveRDS(method_qc_ordering_plot, figure_file("method_qc_ordering.rds"))
+implementation_qc_ordering_plot <- cowplot::plot_grid(category_legend, aspect_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
+implementation_qc_ordering_plot
+saveRDS(implementation_qc_ordering_plot, figure_file("implementation_qc_ordering.rds"))
 ```

### 2018-02-26 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
small update to method characterisation

@@
   arrange(-qc_score) %>% 
-  mutate(implementation_id = factor(implementation_id, levels=implementation_id)) %>% 
-  ggplot(aes(implementation_id, qc_score)) +
+  mutate(implementation_name = factor(implementation_name, levels=implementation_name)) %>% 
+  ggplot(aes(implementation_name, qc_score)) +
     geom_bar(aes(fill=qc_score), stat="identity") +
@@
     scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
-    viridis::scale_fill_viridis(option="A", limits=c(0, 1))
+    viridis::scale_fill_viridis(option="A", limits=c(0, 1)) +
+    labs(y=label_long("qc_score"), x=label_long("implementation_name")) +
+    theme(legend.position="none")
 
 implementation_ordering_plot
+implementation_ordering_plot %>% write_rds(figure_file("implementation_qc_ordering.rds"))
 ```
@@
 implementation_qc_ordering_plot
-saveRDS(implementation_qc_ordering_plot, figure_file("implementation_qc_ordering.rds"))
+saveRDS(implementation_qc_ordering_plot, figure_file("implementation_qc_issues.rds"))
 ```

### 2018-02-26 analysis/paper/paper.Rmd

@@
 Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-This makes a direct[a] comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.[b] Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
+This makes a comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
+We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.[a] Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
 We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We conclude with practical guidelines for method[c] users, and also believe our evaluation framework can be used to spearhead the development for new methods.
+We conclude with practical guidelines for method[b] users, and also believe our evaluation framework can be used to spearhead the development for new methods.
 Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.
@@
 
-Single-cell -omics technologies now make it possible to more accurately[d] model biological systems than ever before.
+Single-cell -omics technologies now make it possible to more accurately[c] model biological systems than ever before.
 One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
-When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[e] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Because they offer an unbiased and transcriptome-wide understanding of the dynamic process[f], trajectories then allow the identification of new [g](primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
-This both for specific subsets of cells, but in the near future also for whole tissues or organisms[h] [@regev_scienceforumhuman_2017].
+When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[d] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
+Because they offer an unbiased and transcriptome-wide understanding of the dynamic process[e], trajectories then allow the identification of new [f](primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
+This both for specific subsets of cells, but in the near future also for whole tissues or organisms[g] [@regev_scienceforumhuman_2017].
 
@@
 add_sfig({
-n_methods_over_time <- read_rds(figure_file("n_methods_over_time.rds", experiment_id="4-method_characterisation"))
+n_methods_over_time <- read_rds(figure_file("n_implementations_over_time.rds", experiment_id="4-method_characterisation"))
 trajectory_components_over_time <- read_rds(figure_file("trajectory_components_over_time.rds", experiment_id="4-method_characterisation"))
@@
 ```{r}
-fixed_linear <-methods$id[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
-fixed_bifurcating <-methods$id[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
-fixed_other <-methods$id[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
-fixed_algorithm <-methods$id[which(methods$fixes_structure == 'algorithm')]
-fixed_parameters <-methods$id[which(methods$fixes_structure == 'parameter')]
+fixed_linear <-methods$method_id[which(methods$fixes_topology != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
+fixed_bifurcating <-methods$method_id[which(methods$fixes_topology != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
+fixed_other <-methods$method_id[which(methods$fixes_topology != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
+fixed_algorithm <-methods$method_id[which(methods$fixes_topology == 'algorithm')]
+fixed_parameters <-methods$method_id[which(methods$fixes_topology == 'parameter')]
 ```
@@
 Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`.
-The focus of these methods was therefore mainly to correctly ordering[i] the cells along this[j] fixed topology.
+The focus of these methods was therefore mainly to correctly ordering[h] the cells along this[i] fixed topology.
 However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to[k] model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to[j] model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
@@
 <!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[l].
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[k].
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore for the first time developed a[m] comprehensive evaluation framework for trajectory inference methods.
-We test[n] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
+In this study, we therefore for the first time developed a[l] comprehensive evaluation framework for trajectory inference methods.
+We test[m] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
 We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "methods")`).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
 We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).
 
-`r ref("table", "methods", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
+`r ref("table", "implementations", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
 ```{r}
-read_rds(figure_file("methods_table.rds", "4-method_characterisation"))
+read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))
 ```
@@
 At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`).
-However, only a handful, mostly recent[o], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+However, only a handful, mostly recent[n], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 add_sfig({
-read_rds(figure_file("method_qc_check.rds", experiment_id="4-method_characterisation"))
+read_rds(figure_file("implementations_table.rds", experiment_id="4-method_characterisation"))
 }, 
-"method_qc_check", 
+"implementations", 
 "Caption", 
@@
 
-Most[p] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`).
+Most[o] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
 However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characterisation", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 
+When looking across all datasets, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics. Over time, new methods
+
+Top methods tend to contain similar components, with …
+
+Over time, both the average and maximum performance increased with the addition of new methods `r ref("sfig", "top_performers_over_time")`
+
 # Discussion
@@
 Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) .
-It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice[q].
+It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
 In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "methods", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
 We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "methods")`).
+This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).
 
@@
 conversion <- function(conversion_oi) {
-    methods_evaluated %>% filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% mutate(text = paste0(name, ifelse(conversion_special, "\\*", ""))) %>% pull(text) %>% list_sentence()
+    methods_evaluated %>% filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% mutate(text = paste0(method_name, ifelse(conversion_special, "\\*", ""))) %>% pull(text) %>% list_sentence()
 }
@@
 
-For each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "methods")`).
+For each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
 This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
@@
 ```
-[a]This makes it difficult to directly compare between trajectory inference methods , and as a result (...)
-[b]which facilitates the comparison of the results of trajectory inference methods (start using the TI abbreviation at some point?) between each other, and possibly to a gold standard.
-[c]which might come in handy to users. We believe (...)
-[d]move? "make it possible to model biological systems more accurately than ever before".
-[e]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
+[a]which facilitates the comparison of the results of trajectory inference methods (start using the TI abbreviation at some point?) between each other, and possibly to a gold standard.
+[b]which might come in handy to users. We believe (...)
+[c]move? "make it possible to model biological systems more accurately than ever before".
+[d]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
 Trajectory inference methods can be used to computationally order these cells along trajectories, based on gradual changes in the expression of several genes.
-[f]processes?
-[g]the trajectories can allow to identify new (primed) subsets of cells, to delineate the exact wiring of a differentiation tree and to infer the regulatory interactions which are responsible for a bifurcation.
-[h]This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future.
-[i]order
-[j]a?
-[k]will be able to model any underlying structure in the data, without the need of user provided parameters. These methods might allow to model more complex behaviors, such as (...)
-[l]get an idea of which available method or set of methods would optimally resolve their problem.
-[m]we present the first comprehensive (...)?
-[n]compared 22 TI methods on both synthetic (..)
-[o]a handful of the most recent methods
-[p]Repetition of most: put "The majority of" instead?
-[q]analysis?
+[e]processes?
+[f]the trajectories can allow to identify new (primed) subsets of cells, to delineate the exact wiring of a differentiation tree and to infer the regulatory interactions which are responsible for a bifurcation.
+[g]This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future.
+[h]order
+[i]a?
+[j]will be able to model any underlying structure in the data, without the need of user provided parameters. These methods might allow to model more complex behaviors, such as (...)
+[k]get an idea of which available method or set of methods would optimally resolve their problem.
+[l]we present the first comprehensive (...)?
+[m]compared 22 TI methods on both synthetic (..)
+[n]a handful of the most recent methods
+[o]Repetition of most: put "The majority of" instead?

### 2018-02-26 analysis/paper/paper.Rmd
update paper

@@
 This makes a comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.[a] Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
+We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
+Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
 We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We conclude with practical guidelines for method[b] users, and also believe our evaluation framework can be used to spearhead the development for new methods.
+We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
 Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.
@@
 
-Single-cell -omics technologies now make it possible to more accurately[c] model biological systems than ever before.
+Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before.
 One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
-When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[d] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Because they offer an unbiased and transcriptome-wide understanding of the dynamic process[e], trajectories then allow the identification of new [f](primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
-This both for specific subsets of cells, but in the near future also for whole tissues or organisms[g] [@regev_scienceforumhuman_2017].
+When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[a][b] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
+Because they offer an unbiased and transcriptome-wide understanding of dynamic process, trajectories then allow the objective identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
+This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018].
 
@@
 Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`.
-The focus of these methods was therefore mainly to correctly ordering[h] the cells along this[i] fixed topology.
+The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
 However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to[j] model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
@@
 <!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[k].
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[c].
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore for the first time developed a[l] comprehensive evaluation framework for trajectory inference methods.
-We test[m] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
+In this study, we therefore for the first time developed a[d] comprehensive evaluation framework for trajectory inference methods.
+We test[e] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
 We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
@@
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
-We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).
+We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_characterisation")`).
 
@@
 
-At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`).
-However, only a handful, mostly recent[n], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_characterisation", "a")`).
+However, only a handful, mostly recent[f], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 
-`r anchor("fig", "method_characterisation")`
-![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
-`r ref("fig", "method_characterisation", anchor=TRUE)` First characterization of each method.
+`r anchor("fig", "implementation_overview")`
+![](`r figure_file("implementation_overview.svg", experiment_id="4-method_characterisation")`)
+`r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
 
-```{r}
-add_sfig({
-read_rds(figure_file("implementations_table.rds", experiment_id="4-method_characterisation"))
-}, 
-"implementations", 
-"Caption", 
-10, 
-13
-)
-```
-
-Most[o] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`).
+Most[g] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_overview", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characterisation", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "implementation_overview", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "implementation_overview", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 ```
-[a]which facilitates the comparison of the results of trajectory inference methods (start using the TI abbreviation at some point?) between each other, and possibly to a gold standard.
-[b]which might come in handy to users. We believe (...)
-[c]move? "make it possible to model biological systems more accurately than ever before".
-[d]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
+[a]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
 Trajectory inference methods can be used to computationally order these cells along trajectories, based on gradual changes in the expression of several genes.
-[e]processes?
-[f]the trajectories can allow to identify new (primed) subsets of cells, to delineate the exact wiring of a differentiation tree and to infer the regulatory interactions which are responsible for a bifurcation.
-[g]This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future.
-[h]order
-[i]a?
-[j]will be able to model any underlying structure in the data, without the need of user provided parameters. These methods might allow to model more complex behaviors, such as (...)
-[k]get an idea of which available method or set of methods would optimally resolve their problem.
-[l]we present the first comprehensive (...)?
-[m]compared 22 TI methods on both synthetic (..)
-[n]a handful of the most recent methods
-[o]Repetition of most: put "The majority of" instead?
+[b]It is not really due to stochasticity that these cells are in different states of a dynamic process though?
+[c]get an idea of which available method or set of methods would optimally resolve their problem.
+[d]we present the first comprehensive (...)?
+[e]compared 22 TI methods on both synthetic (..)
+[f]a handful of the most recent methods
+[g]Repetition of most: put "The majority of" instead?

### 2018-02-27 analysis/paper/paper.Rmd

@@
 csl: nature-biotechnology.csl
+header-includes:
+  - \usepackage{booktabs}
+  - \usepackage{longtable}
+  - \usepackage{array}
+  - \usepackage{multirow}
+  - \usepackage[table]{xcolor}
+  - \usepackage{wrapfig}
+  - \usepackage{float}
+  - \usepackage{colortbl}
+  - \usepackage{pdflscape}
+  - \usepackage{tabu}
+  - \usepackage{threeparttable}
+  - \usepackage[normalem]{ulem}
+mainfont: Open Sans
+
 ---
@@
 Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-This makes a comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking.
+As a result, comparing between trajectory inference methods often challenging, and a comprehensive assessment of the performance and robustness of each method is still lacking.
 We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
@@
 
-This is the first overview about ~~deep reinforcement learning~~ trajectory inference publicly available online. It is comprehensive. Comments and criticisms are welcome.
+This is the first comprehensive assessment the performance of trajectory inference methods publicly available online. Comments and criticisms are welcome.
 
@@
 One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
-When cells are sampled from a population in which cells are at different unknown points in the dynamic process,[a][b] trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
+When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference (TI) methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
 Because they offer an unbiased and transcriptome-wide understanding of dynamic process, trajectories then allow the objective identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
@@
 ```{r}
-add_sfig({
-n_methods_over_time <- read_rds(figure_file("n_implementations_over_time.rds", experiment_id="4-method_characterisation"))
-trajectory_components_over_time <- read_rds(figure_file("trajectory_components_over_time.rds", experiment_id="4-method_characterisation"))
-
-cowplot::plot_grid(plotlist=list(n_methods_over_time, trajectory_components_over_time), nrow=2, rel_heights = c(0.7, 0.3), labels="auto")
-}, 
+add_sfig(
+figure_file("methods_timeline.pdf", "4-method_characterisation"),
 "methods_over_time", 
-"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
-", 
-15, 
-10
+"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory."
 )
@@
 ```{r}
-fixed_linear <-methods$method_id[which(methods$fixes_topology != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
-fixed_bifurcating <-methods$method_id[which(methods$fixes_topology != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
-fixed_other <-methods$method_id[which(methods$fixes_topology != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
-fixed_algorithm <-methods$method_id[which(methods$fixes_topology == 'algorithm')]
-fixed_parameters <-methods$method_id[which(methods$fixes_topology == 'parameter')]
+fixed_linear <- methods %>%
+ filter(
+   fixes_topology != 'free', 
+   maximal_trajectory_type != 'undirected_linear'
+ ) %>%
+ .$method_id
+fixed_bifurcating <- methods %>%
+ filter(
+   fixes_topology != 'free', 
+   maximal_trajectory_type != 'simple_fork'
+ ) %>%
+ .$method_id
+fixed_other <- methods %>%
+ filter(
+   fixes_topology != 'free', 
+   !maximal_trajectory_type %in% c('undirected_linear', 'simple_fork')
+ ) %>%
+ .$method_id
+fixed_algorithm <- methods %>%
+ filter(fixes_topology == 'algorithm') %>%
+ .$method_id
+fixed_parameters <- methods %>%
+ filter(fixes_topology == 'parameter') %>%
+ .$method_id
+
 ```
@@
 <!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem[c].
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore for the first time developed a[d] comprehensive evaluation framework for trajectory inference methods.
-We test[e] both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
+In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods.
+We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
 We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
@@
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
-We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_characterisation")`).
+We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).
 
@@
 ```{r}
-read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))
+read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]]
 ```
@@
 `r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)  
+![](`r figure_file("ti_evaluation_overview_v2.pdf", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 add_stable(
-read_rds(figure_file("datasets.rds","2-dataset_characterisation")), 
+read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
 "datasets", 
@@
 
-At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_characterisation", "a")`).
-However, only a handful, mostly recent[f], methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
+However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
 
@@
 add_stable(
-read_rds(figure_file("qc_checks.rds","4-method_characterisation")), 
+read_rds(figure_file("qc_checks.rds","4-method_characterisation"))[[params$table_format]], 
 "qc_checks", 
-""
+"QC checks"
 )
@@
 `r anchor("fig", "implementation_overview")`
-![](`r figure_file("implementation_overview.svg", experiment_id="4-method_characterisation")`)
+![](`r figure_file("implementation_overview.pdf", experiment_id="4-method_characterisation")`)
 `r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
 
-Most[g] methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_overview", "b")`).
+Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_overview", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
@@
 ```{r}
-add_sfig({
-read_rds(figure_file("qc_over_time.rds", experiment_id="4-method_characterisation"))
-}, 
+add_sfig(
+figure_file("qc_over_time.pdf","4-method_characterisation"), 
 "qc_over_time", 
-"Quality control scores and number of citations over time", 
-10, 
-10
+"Quality control scores and number of citations over time"
 )
@@
 ```{r}
-add_sfig({
-read_rds(figure_file("method_components.rds", experiment_id="4-method_characterisation"))
-}, 
+add_sfig(
+figure_file("method_components.pdf","4-method_characterisation"), 
 "method_components", 
-"TI methods frequently share steps", 
-15, 
-7.5
+"TI methods frequently share steps"
 )
@@
 add_sfig({
-                glue::glue("![]({figure_file('wrapping.svg', experiment_id = '4-method_characterisation')})")
+        figure_file('wrapping.pdf', '4-method_characterisation')
 }, 
@@
 ```{r}
-add_sfig({
-read_rds(figure_file("trajectory_type_trees.rds", experiment_id="2-dataset_characterisation/2-trajectory_types"))
-}, 
+add_sfig(
+figure_file("trajectory_type_trees.pdf","2-dataset_characterisation/2-trajectory_types"), 
 "trajectory_type_tree", 
@@
 # create functions to list the methods within a particular conversion category
-list_sentence <- function(x) {pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))}
+list_sentence <- function(x) {
+  pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))
+}
 conversion <- function(conversion_oi) {
-    methods_evaluated %>% filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% mutate(text = paste0(method_name, ifelse(conversion_special, "\\*", ""))) %>% pull(text) %>% list_sentence()
+ methods_evaluated %>% 
+   filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% 
+   mutate(text = paste0(method_name, ifelse(conversion_special, "\\*", ""))) %>%
+   pull(text) %>% 
+   list_sentence()
 }
+
 ```
@@
 $$
-\begin{align}
+\begin{aligned}
 \label{eq:mrna_ode}
-\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation} \\
+\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation}
+\end{aligned}
+$$
+$$
+\begin{aligned}
 \label{eq:prot_ode}
 \frac{dy_i}{dt} &= \underbrace{r \times x_i}_\text{production} - \underbrace{\Lambda \times y_i}_\text{degradation}
-\end{align}
+\end{aligned}
 $$
@@
 $$
-\begin{align}
+\begin{aligned}
 \label{eq:input_funcion_probabilities}
 f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
-\end{align}
+\end{aligned}
 $$
@@
 $$
-\begin{align}
+\begin{aligned}
 P_j \propto \nu = \left(\frac{y}{k}\right)^{n}
-\end{align}
+\end{aligned}
 $$
@@
 $$
-\begin{align}
+\begin{aligned}
 P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
-\end{align}
+\end{aligned}
 $$
@@
 $$
-\begin{align}
+\begin{aligned}
 \frac{dx_i}{dt} = m \times f(y_1, y_2, ...) - \lambda \times x_i &+ \eta \times \sqrt{x_i} \times \Delta W_t \\
 \frac{dy_i}{dt} = r \times x_i - \Lambda \times y_i &+ \eta \times \sqrt{y_i} \times \Delta W_t
-\end{align}
+\end{aligned}
 $$
@@
 add_stable(
-read_rds(figure_file("samplers.rds","2-dataset_characterisation/3-synthetic")), 
+read_rds(figure_file("samplers.rds","2-dataset_characterisation/3-synthetic"))[[params$table_format]], 
 "samplers", 
@@
 add_sfig({
-read_rds(figure_file("edge_flip_comparison.rds", experiment_id="3-score_rules/4-edge_flip"))
+figure_file("edge_flip_comparison.pdf", experiment_id="3-score_rules/4-edge_flip")
 }, 
@@
 ```
-[a]When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
+When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
 Trajectory inference methods can be used to computationally order these cells along trajectories, based on gradual changes in the expression of several genes.
-[b]It is not really due to stochasticity that these cells are in different states of a dynamic process though?
-[c]get an idea of which available method or set of methods would optimally resolve their problem.
-[d]we present the first comprehensive (...)?
-[e]compared 22 TI methods on both synthetic (..)
-[f]a handful of the most recent methods
-[g]Repetition of most: put "The majority of" instead?
+It is not really due to stochasticity that these cells are in different states of a dynamic process though?
+insert refs
+get an idea of which available method or set of methods would optimally resolve their problem.
+we present the first comprehensive (...)?
+compared 22 TI methods on both synthetic (..)
+Love this figure! It's a great summary of the /whole/ study methodology
+I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
+I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
+part c: I would explicitly show that the probabilistic trajectory models are being used for both the tasks and for the method outputs
+The spacing is a bit variable; e.g. part b takes up a lot of real estate while parts d and e seem to be packed very tightly together
+part f: I'd explicitly mention that each column is to be interpreted separately; there is no connection horizontally. JUST TO MAKE SURE.
+a handful of the most recent methods
+Repetition of most: put "The majority of" instead?

### 2018-02-27 analysis/paper/paper.Rmd
bump paper + update paper generation

@@
 Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-As a result, comparing between trajectory inference methods often challenging, and a comprehensive assessment of the performance and robustness of each method is still lacking.
+As a result, comparing between trajectory inference methods is often challenging, and a comprehensive assessment of the performance and robustness of each method is still lacking.
 We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
@@
 # Introduction
+<!-- Introduction to single-cell and TI --->
 
@@
 When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference (TI) methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
-Because they offer an unbiased and transcriptome-wide understanding of dynamic process, trajectories then allow the objective identification of new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation refs.
-This is currently applicable to specific subsets of cells, but might also be applicable to whole tissues or organisms in the near future [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018].
+TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017].
+Current applications of TI focus on specific subsets of cells, but current efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
 
@@
 add_sfig(
-figure_file("methods_timeline.pdf", "4-method_characterisation"),
+figure_file("methods_timeline.svg", "4-method_characterisation"),
 "methods_over_time", 
@@
 ```{r}
-fixed_linear <- methods %>%
+fixed_linear <- implementations %>%
  filter(
    fixes_topology != 'free', 
-   maximal_trajectory_type != 'undirected_linear'
+   maximal_trajectory_type == 'directed_linear'
  ) %>%
- .$method_id
-fixed_bifurcating <- methods %>%
+ .$implementation_id
+fixed_bifurcating <- implementations %>%
  filter(
    fixes_topology != 'free', 
-   maximal_trajectory_type != 'simple_fork'
+   maximal_trajectory_type == 'bifurcation'
  ) %>%
- .$method_id
-fixed_other <- methods %>%
+ .$implementation_id
+fixed_other <- implementations %>%
  filter(
    fixes_topology != 'free', 
-   !maximal_trajectory_type %in% c('undirected_linear', 'simple_fork')
+   !maximal_trajectory_type %in% c('directed_linear', 'bifurcation')
  ) %>%
- .$method_id
-fixed_algorithm <- methods %>%
+ .$implementation_id
+fixed_algorithm <- implementations %>%
  filter(fixes_topology == 'algorithm') %>%
- .$method_id
-fixed_parameters <- methods %>%
+ .$implementation_id
+fixed_parameters <- implementations %>%
  filter(fixes_topology == 'parameter') %>%
- .$method_id
+ .$implementation_id
 
@@
 However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
+Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
 <!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
+
+```{r}
+uses_priors <- implementations %>% 
+  gather(prior_id, prior_usage, !!priors$prior_id) %>% 
+  filter(prior_usage %in% c("can_use", "can_root")) %>% 
+  pull(implementation_id) %>% unique()
+
+requires_priors <- implementations %>% 
+  gather(prior_id, prior_usage, !!priors$prior_id) %>% 
+  filter(prior_usage %in% c("required", "required_default")) %>% 
+  pull(implementation_id) %>% unique()
+
+```
+
 Although trajectory inference methods use a variety of algorithms and steps to reach a final ordering, most consist of the following three main steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
-Furthermore, with the ongoing scalability in single-cell -omics technologies (reaching a million cells [@angerer_singlecellsmake_2017], it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user friendly.
+Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used`r cite_methods(uses_priors)` or required`r cite_methods(requires_priors)`, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
 
-<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem.
+<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation →
+
+Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal methods available for their problem.
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods.
-We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge.
-We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
-To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
+In this study, we therefore present the first comprehensive evaluation framework for trajectory inference methods.
+We compared `r sum(methods$evaluated)` methods both on synthetic data, for which the gold standard is known, and real data, for which a gold standard has to be extracted from expert knowledge.
+<!-- We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
+--> To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
 
@@
 `r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.pdf", experiment_id="evaluation_overview")`)
+![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 `r anchor("fig", "implementation_overview")`
-![](`r figure_file("implementation_overview.pdf", experiment_id="4-method_characterisation")`)
+![](`r figure_file("implementation_overview.svg", experiment_id="4-method_characterisation")`)
 `r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
@@
 add_sfig(
-figure_file("qc_over_time.pdf","4-method_characterisation"), 
+figure_file("qc_over_time.svg","4-method_characterisation"), 
 "qc_over_time", 
@@
 add_sfig(
-figure_file("method_components.pdf","4-method_characterisation"), 
+figure_file("method_components.svg","4-method_characterisation"), 
 "method_components", 
@@
 add_sfig({
-        figure_file('wrapping.pdf', '4-method_characterisation')
+        figure_file('wrapping.svg', '4-method_characterisation')
 }, 
@@
 add_sfig(
-figure_file("trajectory_type_trees.pdf","2-dataset_characterisation/2-trajectory_types"), 
+figure_file("trajectory_type_trees.svg","2-dataset_characterisation/2-trajectory_types"), 
 "trajectory_type_tree", 
@@
 add_sfig({
-figure_file("edge_flip_comparison.pdf", experiment_id="3-score_rules/4-edge_flip")
+figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip")
 }, 
@@
 ```
-When sampled, single cells will most probably be in different states of dynamic processes, due to the stochasticity of expression.
-Trajectory inference methods can be used to computationally order these cells along trajectories, based on gradual changes in the expression of several genes.
-It is not really due to stochasticity that these cells are in different states of a dynamic process though?
-insert refs
-get an idea of which available method or set of methods would optimally resolve their problem.
-we present the first comprehensive (...)?
-compared 22 TI methods on both synthetic (..)
+I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
+For example, I talked with Abhishek the other day and his view was really flow oriented.
+I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
 Love this figure! It's a great summary of the /whole/ study methodology

### 2018-02-27 analysis/paper/paper.Rmd
bump paper + pdf generation

@@
 add_sfig(
-figure_file("methods_timeline.svg", "4-method_characterisation"),
+figure_file("methods_timeline.pdf", "4-method_characterisation"),
 "methods_over_time", 
@@
 `r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)
+![](`r figure_file("ti_evaluation_overview_v2.pdf", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 `r anchor("fig", "implementation_overview")`
-![](`r figure_file("implementation_overview.svg", experiment_id="4-method_characterisation")`)
+![](`r figure_file("implementation_overview.pdf", experiment_id="4-method_characterisation")`)
 `r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
@@
 add_sfig(
-figure_file("qc_over_time.svg","4-method_characterisation"), 
+figure_file("qc_over_time.pdf","4-method_characterisation"), 
 "qc_over_time", 
@@
 add_sfig(
-figure_file("method_components.svg","4-method_characterisation"), 
+figure_file("method_components.pdf","4-method_characterisation"), 
 "method_components", 
@@
 add_sfig({
-        figure_file('wrapping.svg', '4-method_characterisation')
+        figure_file('wrapping.pdf', '4-method_characterisation')
 }, 
@@
 add_sfig(
-figure_file("trajectory_type_trees.svg","2-dataset_characterisation/2-trajectory_types"), 
+figure_file("trajectory_type_trees.pdf","2-dataset_characterisation/2-trajectory_types"), 
 "trajectory_type_tree", 
@@
 add_sfig({
-figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip")
+figure_file("edge_flip_comparison.pdf", experiment_id="3-score_rules/4-edge_flip")
 }, 

### 2018-02-27 analysis/analyses/4-method_characterisation/7_quality_control.Rmd

@@
 library(kableExtra)
+
+invisible_grouper <- c(latex="\\iffalse {label_long(aspect_id)} \\fi {.}", html = "<span style='display: none;'>{aspect_id}</span> {.}")
+
+citation_format <- c(
+  latex=function(references) {
+    if(is.na(references)) {
+      ""
+    } else {
+      references %>% 
+        gsub("@", "", .) %>% 
+        str_split("[, \\n]+") %>% 
+        first() %>% 
+        glue::collapse(",") %>% 
+        paste0("\\cite{", ., "}")
+    }
+  },
+  html=function(references) {
+    if(is.na(references)) {
+      ""
+    } else {
+      references %>% 
+        str_split("[, \\n]+") %>% 
+        first() %>% 
+        glue::collapse(";") %>% 
+        {pritt("[{.}]")}
+    }
+  }
+)
+
+collapser <- c(latex="\\newline ", html="\n")
+
+table <- map(c("latex", "html"), function(format) {
   table <- checks %>% 
@@
       weight = cell_spec(
-      pritt("<span style='display: none;'>{aspect_id}</span> {weight}"), # add an invisible grouping indicator
+        weight %>% {pritt(invisible_grouper[[format]])}, # add an invisible grouping indicator
+        format,
         background = spec_color(weight, end=0.9,option="A",direction=-1),
@@
     mutate(
-    references = references %>% 
-      gsub(",", "; ", .) %>% 
-      {ifelse(!is.na(.), pritt("[{.}]"), "")} %>%
-      {pritt("<span style='display: none;'>{aspect_id}</span> {.}")} # add an invisible grouping indicator
+      references = map_chr(references, citation_format[[format]])
     ) %>% 
-  select(aspect, weight, references, item, item_weight, !!qc_applications$application) %>% 
-  mutate_at(vars(!!qc_applications$application), ~ifelse(., "✓", "")) %>% 
+    select(aspect, weight, references, item, item_weight) %>% 
     rename_all(label_long) %>% 
-  knitr::kable("html", escape=FALSE) %>% 
-  kable_styling(bootstrap_options = "striped") %>% 
-  collapse_rows(columns=1:3)
+    knitr::kable(format, escape=FALSE) %>% 
+    kable_styling(bootstrap_options = "striped", latex_options = c("striped", "scale_down")) %>% 
+    landscape() %>% 
+    collapse_rows(1:3)
+  # 
+  # for (category in unique(checks$category)) {
+  #   color <- qc_categories %>% filter(category == !!category) %>% pull(color)
+  #   table <- table %>% 
+  #     group_rows(
+  #       label_long(category), 
+  #       min(which(checks$category == category)), 
+  #       max(which(checks$category == category)),
+  #       label_row_css = pritt("background-color: {color}; color: #fff;")
+  #   )
+  # }
 
-for (category in unique(checks$category)) {
-  color <- qc_categories %>% filter(category == !!category) %>% pull(color)
-  table <- table %>% 
-    group_rows(
-      label_long(category), 
-      min(which(checks$category == category)), 
-      max(which(checks$category == category)),
-      label_row_css = pritt("background-color: {color}; color: #fff;")
-  )
-}
+  table
+}) %>% set_names(c("latex", "html"))
 
@@
 cowplot::plot_grid(g1, g3, g2, ncol=2, nrow=2, labels = "auto") %>% 
-  write_rds(figure_file("qc_over_time.rds"))
+  save_plot(figure_file("qc_over_time.svg"), ., base_width=10, base_height=10)
 ```

### 2018-02-27 analysis/paper/paper.Rmd
finally got the citations to work both in tables and in the paper

@@
   - \usepackage[normalem]{ulem}
+  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript}
 mainfont: Open Sans
@@
 
-We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online.
+```{r}
+qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
+  str_split("[, \n]+") %>% 
+  unlist() %>% 
+  unique() %>% 
+  discard(is.na) %>%
+  glue::collapse(., ";") %>%
+  paste0("[", ., "]")
+```
+
+We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
 The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
@@
 # Acknowledgements
-# References
-
-<div id="refs"></div>
 

### 2018-02-27 analysis/paper/paper.Rmd
bump paper and some figures

@@
   - \usepackage[table]{xcolor}
-  - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
-  - \usepackage{pdflscape}
   - \usepackage{tabu}
-  - \usepackage{threeparttable}
   - \usepackage[normalem]{ulem}
+  - \usepackage{lscape}
   - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript}
@@
 
+### Method wrapping
+
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
@@
 
+### Prior information
+
+TODO: describe which prior information was given and how it was extracted
 ## Method quality control
@@
 Repetition of most: put "The majority of" instead?
+..

### 2018-02-27 analysis/paper/paper.Rmd
bump paper include real synthetic comparison

@@
 
-<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation →
+<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
 
@@
 
+Performance was stable between real and synthetic datasets `r ref("sfig", "real_synthetic_comparison")`
+
+```{r}
+add_sfig(
+        figure_file('real_synthetic_comparison.svg', "5-optimise_parameters/5-real_synthetic_comparison")
+, 
+"real_synthetic_comparison", 
+"Comparison of method performance between real and synthetic datasets. ", 
+12, 
+4
+)
+```
 # Discussion
@@
 
-Due to the absence of a common format for trajectory inference methods, most methods produce a heterogeneous output.
+Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with rarely any overlap.
 We therefore created wrappers around each method in R (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model.
@@
 
-TODO: describe which prior information was given and how it was extracted
+TODO: describe which prior information was given and how it was extracted.
+
 ## Method quality control
@@
 We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
-The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
 The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
@@
 
-For each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
 This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.

### 2018-02-27 analysis/paper/paper_latex.Rmd
every implementation now has a reference, hurray

--- 2018-02-27 analysis/paper/paper.Rmd
@@
 
+Performance was stable between real and synthetic datasets `r ref("sfig", "real_synthetic_comparison")`
+
+```{r}
+add_sfig(
+        figure_file('real_synthetic_comparison.pdf', "5-optimise_parameters/5-real_synthetic_comparison")
+, 
+"real_synthetic_comparison", 
+"Comparison of method performance between real and synthetic datasets. ", 
+12, 
+4
+)
+```
 # Discussion

### 2018-03-01 analysis/paper/paper_latex.Rmd
update paper

@@
 add_sfig(
-figure_file("methods_timeline.pdf", "4-method_characterisation"),
+figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
 "methods_over_time", 
@@
 # Results
+
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
+Currently, only methods published before june 2017 are included in the current evaluation, while more recent methods will be included in the next version of the evaluation.
+We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
+Based on these results, we develop a set of guidelines both for the method users, as well as for method developers.
+
+## Performance on real and synthetic data
+We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
+To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
+In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
+This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
+We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
+We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
+To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
+To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
+
+```{r}
+add_sfig({
+        figure_file('wrapping.tmp.pdf', '4-method_characterisation')
+}, 
+"wrapping", 
+"Our wrapping workflow", 
+15, 
+7.5
+)
+```
+
+`r anchor("fig", "results_overview")`
+![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
+`r ref("fig", "results_overview")` HERE ARE THE RESULTS
+
+Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree CREF.
+These results are consistent between real and synthetic datasets CREF, confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
+However, results are more variable when comparing the three different evaluation metrics,
+## Detectable trajectory topologies
+
+The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. In some 
+## Prior information
+
+## Running time and scalability
+
+## Algorithm components
+
 ## Method quality control
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
 We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).
@@
 `r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.pdf", experiment_id="evaluation_overview")`)
+![](`r figure_file("ti_evaluation_overview_v2.tmp.pdf", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 `r anchor("fig", "implementation_overview")`
-![](`r figure_file("implementation_overview.pdf", experiment_id="4-method_characterisation")`)
+![](`r figure_file("implementation_overview.tmp.pdf", experiment_id="4-method_characterisation")`)
 `r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
@@
 add_sfig(
-figure_file("qc_over_time.pdf","4-method_characterisation"), 
+figure_file("qc_over_time.tmp.pdf","4-method_characterisation"), 
 "qc_over_time", 
@@
 add_sfig(
-figure_file("method_components.pdf","4-method_characterisation"), 
+figure_file("method_components.tmp.pdf","4-method_characterisation"), 
 "method_components", 
@@
 ## Evaluating the accuracy of the inferred trajectory
-
-We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
-To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
-In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
-This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
-We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
-We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
-To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
-To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
-
-```{r}
-add_sfig({
-        figure_file('wrapping.pdf', '4-method_characterisation')
-}, 
-"wrapping", 
-"Our wrapping workflow", 
-15, 
-7.5
-)
-```
 
@@
 add_sfig(
-        figure_file('real_synthetic_comparison.pdf', "5-optimise_parameters/5-real_synthetic_comparison")
+        figure_file('real_synthetic_comparison.tmp.pdf', "5-optimise_parameters/5-real_synthetic_comparison")
 , 
@@
 add_sfig(
-figure_file("trajectory_type_trees.pdf","2-dataset_characterisation/2-trajectory_types"), 
+figure_file("trajectory_type_trees.tmp.pdf","2-dataset_characterisation/2-trajectory_types"), 
 "trajectory_type_tree", 
@@
 add_sfig({
-figure_file("edge_flip_comparison.pdf", experiment_id="3-score_rules/4-edge_flip")
+figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip")
 }, 
@@
 I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
-part c: I would explicitly show that the probabilistic trajectory models are being used for both the tasks and for the method outputs
-The spacing is a bit variable; e.g. part b takes up a lot of real estate while parts d and e seem to be packed very tightly together
-part f: I'd explicitly mention that each column is to be interpreted separately; there is no connection horizontally. JUST TO MAKE SURE.
+This is complex though to put into this figure :p.
+Have a look at 4-method_characteristics/wrapping.tmp.pdf I think it is the least crucial of all parts of the figure now, so..
+But feel free to try it yourself :p
 a handful of the most recent methods

### 2018-03-01 analysis/paper/paper.Rmd
update paper

@@
 # Results
+
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
+Currently, only methods published before june 2017 are included in the current evaluation, while more recent methods will be included in the next version of the evaluation.
+We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
+Based on these results, we develop a set of guidelines both for the method users, as well as for method developers.
+
+## Performance on real and synthetic data
+We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
+To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
+In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
+This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
+We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
+We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
+To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
+To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
+
+```{r}
+add_sfig({
+        figure_file('wrapping.svg', '4-method_characterisation')
+}, 
+"wrapping", 
+"Our wrapping workflow", 
+15, 
+7.5
+)
+```
+
+`r anchor("fig", "results_overview")`
+![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
+`r ref("fig", "results_overview")` HERE ARE THE RESULTS
+
+Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree CREF.
+These results are consistent between real and synthetic datasets CREF, confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
+However, results are more variable when comparing the three different evaluation metrics,
+## Detectable trajectory topologies
+
+The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. In some 
+## Prior information
+
+## Running time and scalability
+
+## Algorithm components
+
 ## Method quality control
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
 We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).
@@
 ## Evaluating the accuracy of the inferred trajectory
-
-We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
-To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
-In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
-This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
-We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
-We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
-To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
-To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
-
-```{r}
-add_sfig({
-        figure_file('wrapping.svg', '4-method_characterisation')
-}, 
-"wrapping", 
-"Our wrapping workflow", 
-15, 
-7.5
-)
-```
 
@@
 I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
-part c: I would explicitly show that the probabilistic trajectory models are being used for both the tasks and for the method outputs
-The spacing is a bit variable; e.g. part b takes up a lot of real estate while parts d and e seem to be packed very tightly together
-part f: I'd explicitly mention that each column is to be interpreted separately; there is no connection horizontally. JUST TO MAKE SURE.
+This is complex though to put into this figure :p.
+Have a look at 4-method_characteristics/wrapping.svg I think it is the least crucial of all parts of the figure now, so..
+But feel free to try it yourself :p
 a handful of the most recent methods

### 2018-03-01 analysis/paper/paper.Rmd
update paper & user guidelines tree

@@
 However, results are more variable when comparing the three different evaluation metrics,
-## Detectable trajectory topologies
+##
 
@@
 `r anchor("fig", "implementation_overview")`
-![](`r figure_file("implementation_overview.svg", experiment_id="4-method_characterisation")`)
+![](`r figure_file("method_components.svg", experiment_id="4-method_characterisation")`)
 `r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
@@
 <!-- Results  -->
+Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
+The main consideration for the choice of a method is based on the user's existing knowledge about what type of trajectory is expected in the data.
+If the user expects a linear, cyclic or bifurcating trajectory, a specific set of methods .
+
+`r anchor("fig", "user_guidelines")`
+![](`r figure_file("tree.svg", experiment_id="7-user_guidelines")`)
+`r ref("fig", "user_guidelines", anchor=TRUE)` Practical guidelines for method users.
 

### 2018-03-01 analysis/paper/paper_latex.Rmd
update paper & user guidelines tree

@@
 `r anchor("fig", "implementation_overview")`
-![](`r figure_file("implementation_overview.tmp.pdf", experiment_id="4-method_characterisation")`)
+![](`r figure_file("method_components.tmp.pdf", experiment_id="4-method_characterisation")`)
 `r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
@@
 <!-- Results  -->
+Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
+The main consideration for the choice of a method is based on the user's existing knowledge about what type of trajectory is expected in the data.
+If the user expects a linear, cyclic or bifurcating trajectory, a specific set of methods .
+
+`r anchor("fig", "user_guidelines")`
+![](`r figure_file("tree.tmp.pdf", experiment_id="7-user_guidelines")`)
+`r ref("fig", "user_guidelines", anchor=TRUE)` Practical guidelines for method users.
 

### 2018-03-01 analysis/paper/paper.Rmd
update paper

@@
 Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
-The main consideration for the choice of a method is based on the user's existing knowledge about what type of trajectory is expected in the data.
-If the user expects a linear, cyclic or bifurcating trajectory, a specific set of methods .
+As method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
+For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
+In each case, a different set of methods performs optimally, with Monocle DDRTree performing best when the data contains a complex tree, while Slingshot performing better when the trajectory could as well be less complex.
+When the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT `r cite_methods("wadot")` and AGA `r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
+If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
+When the users knows the exact .
+For cases where the user does
 

### 2018-03-02 analysis/paper/paper_latex.Rmd
update paper and shizzle

@@
 We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
-Using this framework, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
+Using this framework, we compare the results from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
 We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
@@
 Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore present the first comprehensive evaluation framework for trajectory inference methods.
+In this study, we therefore present the first comprehensive evaluation framework for trajectory inference methods (`r ref("fig", "evaluation_overview", "a")`).
 We compared `r sum(methods$evaluated)` methods both on synthetic data, for which the gold standard is known, and real data, for which a gold standard has to be extracted from expert knowledge.
@@
 
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "b")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
 Currently, only methods published before june 2017 are included in the current evaluation, while more recent methods will be included in the next version of the evaluation.
@@
 
+`r ref("table", "implementations", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
+```{r}
+read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]]
+```
 ## Performance on real and synthetic data
-We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
-To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
-In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
+We evaluated the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "a")`).
+To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "d")` and `r ref("sfig", "wrapping")`.
+In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "d")`).
 This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
 We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
 We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
-To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
+To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
 To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
+
+`r anchor("fig", "evaluation_overview")`
+![](`r figure_file("ti_evaluation_overview_v2.tmp.pdf", experiment_id="evaluation_overview")`)
+`r ref("fig", "evaluation_overview")` Overview of the evaluation
 
@@
 
-`r anchor("fig", "results_overview")`
-![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
-`r ref("fig", "results_overview")` HERE ARE THE RESULTS
-
-Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree CREF.
-These results are consistent between real and synthetic datasets CREF, confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
-However, results are more variable when comparing the three different evaluation metrics,
-##
-
-The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. In some 
-## Prior information
-
-## Running time and scalability
-
-## Algorithm components
-
-## Method quality control
-
-We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).
-
-`r ref("table", "implementations", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
-```{r}
-read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]]
-```
-
-`r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.tmp.pdf", experiment_id="evaluation_overview")`)
-`r ref("fig", "evaluation_overview")` Overview of the evaluation
-
 ```{r}
@@
 
-At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
+Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
+These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
+However, results are more variable when comparing the three different evaluation metrics.
+Some methods, mainly those able to detect more complex trajectory types, have higher edge flip scores.
+Other methods, not necessarily those with the highest edge flip score, are better at preserving the ordering and neighbourhood of the cells.
+
+`r anchor("fig", "results_overview")`
+![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
+`r ref("fig", "results_overview")` HERE ARE THE RESULTS
+## Trajectory types
+
+The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset.
+Most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
 However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+
+Evaluation across trajectory types
+
+## Prior information
+
+## Running time and scalability
+
+## Algorithm components
+
+The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
+Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
+MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).
+
+```{r}
+add_sfig(
+figure_file("method_components.tmp.pdf","4-method_characterisation"), 
+"method_components", 
+"TI methods frequently share steps"
+)
+```
+
+Top methods tend to contain similar components. Principal curves…
+
+## Method quality control
 
@@
 
-`r anchor("fig", "implementation_overview")`
-![](`r figure_file("method_components.tmp.pdf", experiment_id="4-method_characterisation")`)
-`r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
-
 Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_overview", "b")`).
@@
 
-The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
-Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
-MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).
-
-```{r}
-add_sfig(
-figure_file("method_components.tmp.pdf","4-method_characterisation"), 
-"method_components", 
-"TI methods frequently share steps"
-)
-```
-## Evaluating the accuracy of the inferred trajectory
-
-When looking across all datasets, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics. Over time, new methods
-
-Top methods tend to contain similar components, with …
-
-Over time, both the average and maximum performance increased with the addition of new methods `r ref("sfig", "top_performers_over_time")`
-
-Performance was stable between real and synthetic datasets `r ref("sfig", "real_synthetic_comparison")`
-
-```{r}
-add_sfig(
-        figure_file('real_synthetic_comparison.tmp.pdf', "5-optimise_parameters/5-real_synthetic_comparison")
-, 
-"real_synthetic_comparison", 
-"Comparison of method performance between real and synthetic datasets. ", 
-12, 
-4
-)
-```
 # Discussion
@@
 If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
-When the users knows the exact .
-For cases where the user does
+The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
+Here, to our knowledge, only ElPiGraph.R (https://github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
 
@@
 ### Method wrapping
-
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis).
+We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis/issues).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
 TODO: describe which prior information was given and how it was extracted.
+
+## Benchmark
+### Pipeline
+* Look at dyneval methods to write this paragraph
+### Method execution
+* Look at dynmethods::execute_method to write this paragraph
 
@@
 \begin{aligned}
-\label{eq:input_funcion_probabilities}
+\label{eq:input_function_probabilities}
 f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
@@
 The accuracy of cell ordering was assessed by first calculating the geodesic distances $d(c_i, c_j)$ between all cells in both the gold standard and the prediction.
-This geodesic distance is calculated by finding the shortest path between two cells through the milestone network, while taking into account the edge lengths as returned by most methods.
+The geodesic distance is calculated by finding the shortest path between two cells through the milestone network, while taking into account the edge lengths as returned by most methods.
 To calculate this for every pair of cells, we first calculated the shortest path lengths $d(m_a, m_b)$ between all milestones using Dijkstra's algorithm, as implemented in the R *igraph* package.
@@
 Although the correlation between geodesic distances directly assesses the position of the cells in the trajectory, a bad correlation does not directly imply that similar cells were not grouped together by the method, as illustrated in `r ref("sfig", "score_aspects", c("b", "c"))`.
-Indeed, certain methods will inevitably reach an incorrect ordering because they cannot handle the correct trajectory type, but these methods could still correctly place similar cells next to each other.
+For example, certain methods will inevitably reach an incorrect ordering because they cannot handle the correct trajectory type, but these methods could still correctly place similar cells next to each other.
 We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard.
@@
 ```
-
 # Acknowledgements
@@
 But feel free to try it yourself :p
-a handful of the most recent methods
-Repetition of most: put "The majority of" instead?
+Should we create a separate github containing a readme, referencing to the google doc, for this?
 ..
+TODO
+TODO
+to do: add metric figure
+needs to be updated with waypoints
+this is not used anymore

### 2018-03-02 analysis/paper/paper.Rmd

@@
 
+`r ref("table", "implementations", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
+```{r}
+read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]]
+```
 ## Performance on real and synthetic data
-We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`).
-To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`.
-In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`).
+We evaluated the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "a")`).
+To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "d")` and `r ref("sfig", "wrapping")`.
+In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "d")`).
 This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
 We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`).
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
 We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
-To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`).
+To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
 To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
+
+`r anchor("fig", "evaluation_overview")`
+![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)
+`r ref("fig", "evaluation_overview")` Overview of the evaluation
 
@@
 
-`r anchor("fig", "results_overview")`
-![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
-`r ref("fig", "results_overview")` HERE ARE THE RESULTS
-
-Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree CREF.
-These results are consistent between real and synthetic datasets CREF, confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
-However, results are more variable when comparing the three different evaluation metrics,
-##
-
-The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. In some 
-## Prior information
-
-## Running time and scalability
-
-## Algorithm components
-
-## Method quality control
-
-We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).
-
-`r ref("table", "implementations", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
-```{r}
-read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]]
-```
-
-`r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)
-`r ref("fig", "evaluation_overview")` Overview of the evaluation
-
 ```{r}
@@
 
-At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
+Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
+These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
+However, results are more variable when comparing the three different evaluation metrics.
+Some methods, mainly those able to detect more complex trajectory types, have higher edge flip scores.
+Other methods, not necessarily those with the highest edge flip score, are better at preserving the ordering and neighbourhood of the cells.
+
+`r anchor("fig", "results_overview")`
+![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
+`r ref("fig", "results_overview")` HERE ARE THE RESULTS
+## Trajectory types
+
+The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset.
+Most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
 However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+
+Evaluation across trajectory types
+
+## Prior information
+
+## Running time and scalability
+
+## Algorithm components
+
+The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
+Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
+MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).
+
+```{r}
+add_sfig(
+figure_file("method_components.svg","4-method_characterisation"), 
+"method_components", 
+"TI methods frequently share steps"
+)
+```
+
+Top methods tend to contain similar components. Principal curves…
+
+## Method quality control
 
@@
 
-`r anchor("fig", "implementation_overview")`
-![](`r figure_file("method_components.svg", experiment_id="4-method_characterisation")`)
-`r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.
-
 Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_overview", "b")`).
@@
 
-The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
-Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
-MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).
-
-```{r}
-add_sfig(
-figure_file("method_components.svg","4-method_characterisation"), 
-"method_components", 
-"TI methods frequently share steps"
-)
-```
-## Evaluating the accuracy of the inferred trajectory
-
-When looking across all datasets, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics. Over time, new methods
-
-Top methods tend to contain similar components, with …
-
-Over time, both the average and maximum performance increased with the addition of new methods `r ref("sfig", "top_performers_over_time")`
-
-Performance was stable between real and synthetic datasets `r ref("sfig", "real_synthetic_comparison")`
-
-```{r}
-add_sfig(
-        figure_file('real_synthetic_comparison.svg', "5-optimise_parameters/5-real_synthetic_comparison")
-, 
-"real_synthetic_comparison", 
-"Comparison of method performance between real and synthetic datasets. ", 
-12, 
-4
-)
-```
 # Discussion

### 2018-03-02 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
update qc

@@
         format,
-        background = spec_color(weight, end=0.9,option="A",direction=-1),
-        escape=FALSE,
-        color="white"
+        color = spec_color(weight,option="viridis",direction=-1),
+        escape=FALSE
       )
@@
     ) %>% 
-    select(aspect, weight, references, item, item_weight) %>% 
+    mutate(
+      category = cell_spec(
+        label_long(category),
+        format,
+        color = set_names(qc_categories$color, qc_categories$category)[category],
+        escape=FALSE
+      )
+    ) %>% 
+    select(category, aspect, weight, references, item, item_weight) %>% 
     rename_all(label_long) %>% 
     knitr::kable(format, escape=FALSE) %>% 
-    kable_styling(bootstrap_options = "striped", latex_options = c("striped", "scale_down")) %>% 
+    kable_styling(bootstrap_options = "striped", latex_options = c("scale_down")) %>% 
     landscape() %>% 
-    collapse_rows(1:3)
-  # 
+    collapse_rows(1:4)
+
+  # if (format == "html") {
   #   for (category in unique(checks$category)) {
@@
   #   }
+  # }
 
@@
 }) %>% set_names(c("latex", "html"))
-
+table
 table %>% write_rds(figure_file("qc_checks.rds"))

### 2018-03-03 analysis/paper/paper.Rmd
table caption floats with table

@@
   - \usepackage{lscape}
+  - \usepackage{pbox}
+  - \usepackage[labelformat=empty]{caption}
   - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript}
@@
 We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods.
-Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.
+We conclude with practical guidelines for method users .
+We believe our evaluation framework can be used to spearhead the development for new methods.
+Our evaluation pipeline is fully reproducible, can be easily extended, and is available at github.com/dynverse/dynverse.
 
@@
 # Introduction
-<!-- Introduction to single-cell and TI --->
-
+<!-- §1: Introduction to single-cell and TI --->
 Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before.
-One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation.
-When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference (TI) methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016].
+One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
+When cells are profiled from a population in which cells are at different unknown points in the dynamic process, dynamic processes can be modelled using trajectory inference (TI) methods.
+These methods computationally order these cells along a trajectory topology, which can be linear, bifurcating, or more complex [@cannoodt_computational_2016].
 TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017].
-Current applications of TI focus on specific subsets of cells, but current efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
 
@@
 
-<!-- history of Ti, different methods -->
 ```{r}
-fixed_linear <- implementations %>%
-filter(
-  fixes_topology != 'free', 
-  maximal_trajectory_type == 'directed_linear'
-) %>%
+flin <- implementations %>%
+filter(fixes_topology != 'free', maximal_trajectory_type == 'directed_linear') %>% .$implementation_id
+fbif <- implementations %>%
+filter(fixes_topology != 'free', maximal_trajectory_type == 'bifurcation') %>% .$implementation_id
+foth <- implementations %>% 
+ filter(fixes_topology != 'free', !maximal_trajectory_type %in% c('directed_linear', 'bifurcation')) %>%
  .$implementation_id
-fixed_bifurcating <- implementations %>%
-filter(
-  fixes_topology != 'free', 
-  maximal_trajectory_type == 'bifurcation'
-) %>%
-.$implementation_id
-fixed_other <- implementations %>%
-filter(
-  fixes_topology != 'free', 
-  !maximal_trajectory_type %in% c('directed_linear', 'bifurcation')
-) %>%
-.$implementation_id
-fixed_algorithm <- implementations %>%
-filter(fixes_topology == 'algorithm') %>%
-.$implementation_id
-fixed_parameters <- implementations %>%
-filter(fixes_topology == 'parameter') %>%
-.$implementation_id
+falg <- implementations %>% filter(fixes_topology == 'algorithm') %>% .$implementation_id
+fpar <- implementations %>% filter(fixes_topology == 'parameter') %>% .$implementation_id
 
 ```
-
-Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`).
-Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`.
-The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
-However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`).
-Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
-
-<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
 
@@
 ```
+<!-- §2: plethora of new TI methods, with wildly varying characteristics -->
+A plethora of TI methods has been developed over the last years, and more are being published at a rate of about one method every two weeks (`r ref("sfig", "methods_per_year", "a")`).
+Surprisingly, almost all methods have a unique combination of key characteristics in terms of the required inputs, produced outputs and methodology used.
+<!-- §3: TOPOLOGY FIXING VERSUS TOPOLOGY INFERENCE -->
+Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(flin)`, bifurcating `r cite_methods(fbif)` or other `r cite_methods(foth)` topology, by either algorithmic limitations `r cite_methods(falg)` or through user parameters `r cite_methods(fpar)`.
+The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
+However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed.
+Methods which do not impose limitations on the topology at all are now being developed [@welch_slicerinferringbranched_2016; @wolf_graphabstractionreconciles_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
 
+<!-- §4: STEPS AND COMPONENTS -->
 Although trajectory inference methods use a variety of algorithms and steps to reach a final ordering, most consist of the following three main steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+
+<!-- §5: OPTIONAL OR REQUIRED PRIORS -->
 Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used`r cite_methods(uses_priors)` or required`r cite_methods(requires_priors)`, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
 
-<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
-
+<!-- §5: Problem statement -->
 Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal methods available for their problem.
@@
 # Results
+## Method characterisation
 
@@
 We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
-Based on these results, we develop a set of guidelines both for the method users, as well as for method developers.
 
-`r ref("table", "implementations", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
-```{r}
-read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]]
+The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset, with … of the `r sum(methods$evaluated)` evaluated methods fixing the trajectory type either due to algorithmic limitations or parameters.
+Moreover, if a method's topology is free, the topology can still be constrained, limiting the number of trajectory types a method can detect.
+Although a significant number of methods can only predict linear or cyclical trajectories, a majority of methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
+However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+
+```{r, results='asis' }
+add_table(
+"implementations",
+read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
+"Overview of current trajectory inference methods, and whether they were included in this current evaluation study"
+)
 ```
-## Performance on real and synthetic data
+## Evaluation overview
 We evaluated the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "a")`).
@@
 
+## Method benchmark
+
 Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
@@
 ![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
-`r ref("fig", "results_overview")` HERE ARE THE RESULTS
-## Trajectory types
+`r ref("fig", "results_overview")` Overview of characteristics and benchmark of `r sum(methods$evaluated)` evaluated methods
 
-The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset.
-Most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
-However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+### Trajectory types and topology complexity
 
-Evaluation across trajectory types
+Except for multifurcations, acyclic graphs and graphs, the top performing method was also a method which can handle that particular trajectory type in theory (`r ref("sfig", "trajtype_handle_comparison")`).
+This higher performance was mainly due to a higher edge flip and random forest MSE score, as the correlation score was in almost every case (except for linear trajectories) higher for methods which cannot handle the particular topology type.
+This indicates that among the current methods, one has to balance finding the correct topology against optimally ordering the cells along this topology.
+For acyclic graphs and graphs, only one method was included in the study (SLICER), and it did not perform well on any trajectory type.
+For multifurcations, wishbone slightly outperformed slingshot and other methods able to detect multifurcations, ……..
+Ugh this is difficult to explain :p
 
-## Prior information
+```{r}
+add_sfig({
+        figure_file('trajtype_handle_comparison.svg', "8-compare_topology")
+}, 
+"trajtype_handle_comparison", 
+""
+)
+```
 
-## Running time and scalability
+Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
+Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler of structure, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree for example, frequently predicted a tree topology, despite only a cyclic, linear or bifurcating topology being present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories generated by Monocle DDRTree (at default parameters) tend to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an edge on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
+Indeed, when we assessed how often a method can predict the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was the only method correctly predicting tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("fig", "topology_sensitivity")`).
+Therefore, current methods still require a considerable guidance by the end user, to either optimise the parameters or to choose the method which best fits the user's expert knowledge.
 
-## Algorithm components
+`r anchor("fig", "topology_complexity_comparison")`
+![](`r figure_file("topology_complexity_comparison.svg","8-compare_topology")`)
+`r ref("fig", "topology_complexity_comparison")`
 
-The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
+```{r}
+add_sfig({
+        figure_file('topology_sensitivity.svg', "8-compare_topology")
+}, 
+"topology_sensitivity", 
+""
+)
+```
+### Effect of prior information
+
+### Running time and scalability
+
+### Algorithm components
+
+The underlying algorithm of each method can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
 Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
@@
 
-Top methods tend to contain similar components. Principal curves…
+We assessed whether the components of an algorithm can be predictive for method performance, using both random forest regression and statistical testing.
+Methods which include principal curves and clustering (k-means in particular) tend to perform better
 
@@
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
 We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
@@
 ```
+## Practical guidelines
 
-# Discussion
-<!-- Some intro -->
-Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) .
-It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
-In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
-
-<!-- Results  -->
 Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
@@
 `r ref("fig", "user_guidelines", anchor=TRUE)` Practical guidelines for method users.
+
+# Discussion
+<!-- Some intro -->
+Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
+Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
+It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
+In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
 
@@
 
-<p class='wip'> Cascades? Cycle?
-</p>
+§ Cascades? Cycle?
 
-<p class='wip'> To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
+§ To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
 Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated trajectories which simple undirected graphs (bifurcating loop and bifurcating convergence).
-</p>
 
@@
 
-<p class='wip'> Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward.
+§ Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward.
 Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?.
@@
 In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
-</p>
 
@@
 Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
-This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no suitable [@bahiense_maximumcommonedge_2012].
+This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no suitable …?
+[@bahiense_maximumcommonedge_2012].
 We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes.
@@
 ```
+
+### Aggregation of scores
+```{r}
+add_sfig({
+figure_file("metric_aggregation_overview_v1.svg", experiment_id="evaluation_overview")
+}, 
+"aggregation_of_results", 
+"Aggregation methodology of metric scores", 
+15, 
+5
+)
+```
+
 # Acknowledgements
@@
 ```
+Rephrase
 I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
@@
 I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
+I'm still on the fence. I think we don't even discuss TI in enough detail, let alone transcriptomics vs. cytometry
+But TI methods are also central in cytometry
+is this the best place for this citation?
+TODO: explicitly show the number of TI methods each year, to support our claim of '1 method every 2 weeks'.
+
+For example, see:
+figure_file('ti_methods_per_year.svg', '4-method_characterisation')
+Use different word since it was already used in §2
 Love this figure! It's a great summary of the /whole/ study methodology
@@
 But feel free to try it yourself :p
+😉
 Should we create a separate github containing a readme, referencing to the google doc, for this?
@@
 this is not used anymore
+suitable what?

### 2018-03-03 analysis/paper/paper_latex.Rmd
table caption floats with table

@@
 ![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
-`r ref("fig", "results_overview")` HERE ARE THE RESULTS
-## Trajectory types
+`r ref("fig", "results_overview")` Overview of characteristics and benchmark of `r sum(methods$evaluated)` evaluated methods
 
-The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset.
-Most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
-However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+### Trajectory types and topology complexity
 
-Evaluation across trajectory types
+Except for multifurcations, acyclic graphs and graphs, the top performing method was also a method which can handle that particular trajectory type in theory (`r ref("sfig", "trajtype_handle_comparison")`).
+This higher performance was mainly due to a higher edge flip and random forest MSE score, as the correlation score was in almost every case (except for linear trajectories) higher for methods which cannot handle the particular topology type.
+This indicates that among the current methods, one has to balance finding the correct topology against optimally ordering the cells along this topology.
+For acyclic graphs and graphs, only one method was included in the study (SLICER), and it did not perform well on any trajectory type.
+For multifurcations, wishbone slightly outperformed slingshot and other methods able to detect multifurcations, ……..
+Ugh this is difficult to explain :p
 
-## Prior information
+```{r}
+add_sfig({
+        figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology")
+}, 
+"trajtype_handle_comparison", 
+""
+)
+```
 
-## Running time and scalability
+Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
+Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler of structure, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree for example, frequently predicted a tree topology, despite only a cyclic, linear or bifurcating topology being present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories generated by Monocle DDRTree (at default parameters) tend to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an edge on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
+Indeed, when we assessed how often a method can predict the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was the only method correctly predicting tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("fig", "topology_sensitivity")`).
+Therefore, current methods still require a considerable guidance by the end user, to either optimise the parameters or to choose the method which best fits the user's expert knowledge.
 
-## Algorithm components
+`r anchor("fig", "topology_complexity_comparison")`
+![](`r figure_file("topology_complexity_comparison.tmp.pdf","8-compare_topology")`)
+`r ref("fig", "topology_complexity_comparison")`
 
-The underlying algorithms can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
+```{r}
+add_sfig({
+        figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology")
+}, 
+"topology_sensitivity", 
+""
+)
+```
+### Effect of prior information
+
+### Running time and scalability
+
+### Algorithm components
+
+The underlying algorithm of each method can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
 Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
@@
 ```
+
+### Aggregation of scores
+```{r}
+add_sfig({
+figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview")
+}, 
+"aggregation_of_results", 
+"Aggregation methodology of metric scores", 
+15, 
+5
+)
+```
+
 # Acknowledgements
@@
 I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
+I'm still on the fence. I think we don't even discuss TI in enough detail, let alone transcriptomics vs. cytometry
+But TI methods are also central in cytometry
+is this the best place for this citation?
+TODO: explicitly show the number of TI methods each year, to support our claim of '1 method every 2 weeks'.
+
+For example, see:
+figure_file('ti_methods_per_year.tmp.pdf', '4-method_characterisation')
+Use different word since it was already used in §2
 Love this figure! It's a great summary of the /whole/ study methodology

### 2018-03-03 analysis/paper/paper.Rmd
refresh figures and tables

@@
 We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We conclude with practical guidelines for method users .
+We conclude with practical guidelines for users to select appropriate methods for a given use case.
 We believe our evaluation framework can be used to spearhead the development for new methods.
@@
 
-This is the first comprehensive assessment the performance of trajectory inference methods publicly available online. Comments and criticisms are welcome.
-
+We believe this to be the first comprehensive assessment of trajectory inference methods.
+We gladly welcome any discussion and feedback on key decisions made as part of this study, including the metrics used in the benchmark, the quality control checklist, and the implementation of the method wrappers.
+These discussions can be held at github.com/dynverse/dynverse/issues.
 # Introduction
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-When cells are profiled from a population in which cells are at different unknown points in the dynamic process, dynamic processes can be modelled using trajectory inference (TI) methods.
-These methods computationally order these cells along a trajectory topology, which can be linear, bifurcating, or more complex [@cannoodt_computational_2016].
+When cells are profiled from a population in which cells are at different unknown points in the dynamic process, dynamic processes can be modelled using trajectory inference (TI) methods  [@cannoodt_computational_2016].
+These methods computationally order these cells along a trajectory topology, which can be linear, bifurcating, or more complex.
 TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017].
@@
 )
-```
-
-```{r}
-flin <- implementations %>%
-filter(fixes_topology != 'free', maximal_trajectory_type == 'directed_linear') %>% .$implementation_id
-fbif <- implementations %>%
-filter(fixes_topology != 'free', maximal_trajectory_type == 'bifurcation') %>% .$implementation_id
-foth <- implementations %>% 
- filter(fixes_topology != 'free', !maximal_trajectory_type %in% c('directed_linear', 'bifurcation')) %>%
- .$implementation_id
-falg <- implementations %>% filter(fixes_topology == 'algorithm') %>% .$implementation_id
-fpar <- implementations %>% filter(fixes_topology == 'parameter') %>% .$implementation_id
-
 ```
@@
 <!-- §2: plethora of new TI methods, with wildly varying characteristics -->
-A plethora of TI methods has been developed over the last years, and more are being published at a rate of about one method every two weeks (`r ref("sfig", "methods_per_year", "a")`).
-Surprisingly, almost all methods have a unique combination of key characteristics in terms of the required inputs, produced outputs and methodology used.
-<!-- §3: TOPOLOGY FIXING VERSUS TOPOLOGY INFERENCE -->
-Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(flin)`, bifurcating `r cite_methods(fbif)` or other `r cite_methods(foth)` topology, by either algorithmic limitations `r cite_methods(falg)` or through user parameters `r cite_methods(fpar)`.
-The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
-However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed.
-Methods which do not impose limitations on the topology at all are now being developed [@welch_slicerinferringbranched_2016; @wolf_graphabstractionreconciles_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].
-
-<!-- §4: STEPS AND COMPONENTS -->
-Although trajectory inference methods use a variety of algorithms and steps to reach a final ordering, most consist of the following three main steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-
-<!-- §5: OPTIONAL OR REQUIRED PRIORS -->
-Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used`r cite_methods(uses_priors)` or required`r cite_methods(requires_priors)`, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others.
-
-<!-- §5: Problem statement -->
-Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal methods available for their problem.
-Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art.
-In this study, we therefore present the first comprehensive evaluation framework for trajectory inference methods (`r ref("fig", "evaluation_overview", "a")`).
-We compared `r sum(methods$evaluated)` methods both on synthetic data, for which the gold standard is known, and real data, for which a gold standard has to be extracted from expert knowledge.
-<!-- We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets.
---> To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
-
-# Results
-## Method characterisation
-
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "b")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
-Currently, only methods published before june 2017 are included in the current evaluation, while more recent methods will be included in the next version of the evaluation.
-We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
-
-The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset, with … of the `r sum(methods$evaluated)` evaluated methods fixing the trajectory type either due to algorithmic limitations or parameters.
-Moreover, if a method's topology is free, the topology can still be constrained, limiting the number of trajectory types a method can detect.
-Although a significant number of methods can only predict linear or cyclical trajectories, a majority of methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`).
-However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).
+A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`).
+We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` TI methods (`r ref("table", "implementations")`), which we believe to be the most complete list to date.
+It is perhaps surprising that almost all of them have a unique combination of characteristics, in terms of the required inputs (prior information), produced outputs (topology fixing, trajectory type) and methodology used (not shown).
 
@@
 ```
+<!-- §3: TOPOLOGY FIXING VERSUS TOPOLOGY INFERENCE -->
+One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
+Early TI methods typically fixed the topology algorithmically (e.g.
+linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015], bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]) or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
+The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
+Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows a method to be more broadly applicable on more use cases.
+Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).
+Most noticeable of these methods is AGA[@wolf_graphabstractionreconciles_2017], which is the only TI method currently able to predict disconnected graphs.
+It is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
+
+<!-- §4: PRIOR INFORMATION -->
+Another key characteristic is the selection of prior information that it requires or can optionally exploit.
+Examples of prior informations are supplying a starting cell to start the trajectory from, a set of important marker genes, or even a grouping of cells into cell states.
+Giving prior information to a TI method can be both a blessing and a curse.
+In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
+On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, limiting the applicability of such TI methods to only well studied systems.
+
+<!-- §5: COMPONENTS -->
+A reductionist approach to characterising TI methods is dissecting it into a set of algorithmic components, as any one component can have a significant impact on the performance, scalability, and output data structures.
+Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+
+<!-- §5: Problem statement -->
+An important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
+Many attempts at addressing this issue has already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, performing a comprehensive benchmarking evaluation of TI methods requires a lot of work.
+This is problematic, as new users to the field are confronted with a wide array of TI methods without a clue as to what the most optimal method available for their problem could be. 
+Moreover, the strengths and weakness of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
+
+<!-- §6: Our contributions -->
+In this study, we perform a comprehensive evaluation for `r sum(methods$evaluated)` trajectory inference methods.
+As part of our evaluation framework (`r ref("fig", "evaluation_overview", "a")`), we developed a common probabilistic model for representing trajectories from all possible sources, and defined a set of metrics for comparing the likeness of such trajectories.
+The data compendium consists of both synthetic datasets for which we know the gold standard, and real scRNA-seq datasets for which the gold standard was extracted from expert knowledge.
+The generation of synthetic datasets was performed by simulating thermodynamics models of gene regulatory networks, and subsequently simulating a single cell profiling experiment.
+In addition to the quantitative benchmarking, we also assembled a quality control checklist evaluating the corresponding software and manuscript, including measures such as code quality, automated unit testing, and user friendliness.
+Finally, we present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+
+# Results
 ## Evaluation overview
+The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
+Currently, only methods published before june 2017 are included in the evaluation, while more recent methods will be included in the next version.
+We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
+
 We evaluated the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "a")`).
@@
 
+```{r}
+add_sfig(
+figure_file('real_synthetic_comparison.svg', "5-optimise_parameters/5-real_synthetic_comparison"), 
+"real_synthetic_comparison", 
+"Comparison between performance on real and synthetic datasets across trajectory types."
+)
+```
+
 ### Trajectory types and topology complexity
 
-Except for multifurcations, acyclic graphs and graphs, the top performing method was also a method which can handle that particular trajectory type in theory (`r ref("sfig", "trajtype_handle_comparison")`).
-This higher performance was mainly due to a higher edge flip and random forest MSE score, as the correlation score was in almost every case (except for linear trajectories) higher for methods which cannot handle the particular topology type.
-This indicates that among the current methods, one has to balance finding the correct topology against optimally ordering the cells along this topology.
-For acyclic graphs and graphs, only one method was included in the study (SLICER), and it did not perform well on any trajectory type.
-For multifurcations, wishbone slightly outperformed slingshot and other methods able to detect multifurcations, ……..
-Ugh this is difficult to explain :p
+In most cases, methods which can handle a particular trajectory type also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+These methods had typically better edge flip scores - as can be expected - and RF MSE scores compared to methods not able to handle the particular trajectory type.
+However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced a better ordering, irrespective of whether the dataset contained a linear trajectory or more complex
+To further investigate, we divided all methods in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
+While there were no significant difference on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "'linear_split_comparison")`).
+Together, this indicates that current non-linear methods contain less accurate ordering algorithms.
+Inclusion of the ordering algorithms from the top linear methods could therefore improve the performance of the non-linear methods.
 
@@
 
+```{r}
+add_sfig({
+        figure_file('linear_split_comparison.svg', "8-compare_topology")
+}, 
+"linear_split_comparison", 
+"Linear methods have a higher correlation score on non-linear datasets.
+Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
+"
+)
+```
+
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler of structure, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree for example, frequently predicted a tree topology, despite only a cyclic, linear or bifurcating topology being present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree for example, frequently predicted a tree topology, even when only cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
 Trajectories generated by Monocle DDRTree (at default parameters) tend to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an edge on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
-Indeed, when we assessed how often a method can predict the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was the only method correctly predicting tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("fig", "topology_sensitivity")`).
-Therefore, current methods still require a considerable guidance by the end user, to either optimise the parameters or to choose the method which best fits the user's expert knowledge.
+Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
+Therefore, inferring the correct topology without parameter tuning is still an open challenge.
+Conversely, when the data contains a complex trajectory structure, TI currently will still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
 
@@
 ### Effect of prior information
+
+Giving prior information to a TI method can be both a blessing and a curse.
+In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
+On the other hand, incorrect or noisy prior information can bias the trajectory towards a wrong, but perhaps consistent with current knowledge, trajectory.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, limiting the applicability of such TI methods to only well studied systems.
+
+In the current version of the evaluation, we only provided prior information when it was required by a method.
+We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`).
+Rather, methods who received prior information were on average positioned in the middle of the ranking.
+Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.
+
+```{r}
+add_sfig({
+        figure_file('required_priors_performance_comparison.svg', "10-prior_information")
+}, 
+"required_priors_performance_comparison", 
+""
+)
+```
 
@@
 
-Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_overview", "b")`).
+Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "b")`).
 Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "implementation_overview", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "implementation_overview", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "results_overview", "e")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "results_overview", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 ```
-Rephrase
-I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
-For example, I talked with Abhishek the other day and his view was really flow oriented.
-I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
-I'm still on the fence. I think we don't even discuss TI in enough detail, let alone transcriptomics vs. cytometry
-But TI methods are also central in cytometry
-is this the best place for this citation?
-TODO: explicitly show the number of TI methods each year, to support our claim of '1 method every 2 weeks'.
-
-For example, see:
-figure_file('ti_methods_per_year.svg', '4-method_characterisation')
-Use different word since it was already used in §2
-Love this figure! It's a great summary of the /whole/ study methodology
-I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
+This paragraph may be expanded
+Rephrase -_-
+Define in paragraph 1
 I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
@@
 But feel free to try it yourself :p
+Op zoek naar een betere formulering hiervoor, mischien:
+"In most cases, being able to detect a trajectory type leads to a better score on data containing this particular trajectory type"
 😉
+For V2/3: we should check what happens if wrong or noise prior information is given to the methods!
+I wish we were writing V2 instead of this temporary, half-decent version :p
+In V2, "randomise" the prior information so that the positive effect of giving it to a method can be directly assessed, instead of this indirect evidence
 Should we create a separate github containing a readme, referencing to the google doc, for this?

### 2018-03-03 analysis/paper/paper_latex.Rmd
refresh figures and tables

@@
 
+```{r}
+add_sfig(
+figure_file('real_synthetic_comparison.tmp.pdf', "5-optimise_parameters/5-real_synthetic_comparison"), 
+"real_synthetic_comparison", 
+"Comparison between performance on real and synthetic datasets across trajectory types."
+)
+```
+
 ### Trajectory types and topology complexity
 
-Except for multifurcations, acyclic graphs and graphs, the top performing method was also a method which can handle that particular trajectory type in theory (`r ref("sfig", "trajtype_handle_comparison")`).
-This higher performance was mainly due to a higher edge flip and random forest MSE score, as the correlation score was in almost every case (except for linear trajectories) higher for methods which cannot handle the particular topology type.
-This indicates that among the current methods, one has to balance finding the correct topology against optimally ordering the cells along this topology.
-For acyclic graphs and graphs, only one method was included in the study (SLICER), and it did not perform well on any trajectory type.
-For multifurcations, wishbone slightly outperformed slingshot and other methods able to detect multifurcations, ……..
-Ugh this is difficult to explain :p
+In most cases, methods which can handle a particular trajectory type also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+These methods had typically better edge flip scores - as can be expected - and RF MSE scores compared to methods not able to handle the particular trajectory type.
+However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced a better ordering, irrespective of whether the dataset contained a linear trajectory or more complex
+To further investigate, we divided all methods in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
+While there were no significant difference on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "'linear_split_comparison")`).
+Together, this indicates that current non-linear methods contain less accurate ordering algorithms.
+Inclusion of the ordering algorithms from the top linear methods could therefore improve the performance of the non-linear methods.
 
@@
 
+```{r}
+add_sfig({
+        figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology")
+}, 
+"linear_split_comparison", 
+"Linear methods have a higher correlation score on non-linear datasets.
+Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
+"
+)
+```
+
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler of structure, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree for example, frequently predicted a tree topology, despite only a cyclic, linear or bifurcating topology being present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree for example, frequently predicted a tree topology, even when only cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
 Trajectories generated by Monocle DDRTree (at default parameters) tend to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an edge on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
-Indeed, when we assessed how often a method can predict the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was the only method correctly predicting tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("fig", "topology_sensitivity")`).
-Therefore, current methods still require a considerable guidance by the end user, to either optimise the parameters or to choose the method which best fits the user's expert knowledge.
+Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
+Therefore, inferring the correct topology without parameter tuning is still an open challenge.
+Conversely, when the data contains a complex trajectory structure, TI currently will still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
 
@@
 ### Effect of prior information
+
+Giving prior information to a TI method can be both a blessing and a curse.
+In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
+On the other hand, incorrect or noisy prior information can bias the trajectory towards a wrong, but perhaps consistent with current knowledge, trajectory.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, limiting the applicability of such TI methods to only well studied systems.
+
+In the current version of the evaluation, we only provided prior information when it was required by a method.
+We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`).
+Rather, methods who received prior information were on average positioned in the middle of the ranking.
+Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.
+
+```{r}
+add_sfig({
+        figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information")
+}, 
+"required_priors_performance_comparison", 
+""
+)
+```
 
@@
 ```
-Rephrase
-I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
-For example, I talked with Abhishek the other day and his view was really flow oriented.
-I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
-I'm still on the fence. I think we don't even discuss TI in enough detail, let alone transcriptomics vs. cytometry
-But TI methods are also central in cytometry
-is this the best place for this citation?
-TODO: explicitly show the number of TI methods each year, to support our claim of '1 method every 2 weeks'.
-
-For example, see:
-figure_file('ti_methods_per_year.tmp.pdf', '4-method_characterisation')
-Use different word since it was already used in §2
-Love this figure! It's a great summary of the /whole/ study methodology
-I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
+This paragraph may be expanded
+Rephrase -_-
+Define in paragraph 1
 I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model

### 2018-03-04 analysis/paper/paper.Rmd
update pipeline overview figure

@@
 ## Evaluation overview
+WIP
 The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
@@
 `r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)
+![](`r figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
+
+```{r}
+add_sfig({
+        figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview')
+}, 
+"wrapping", 
+"Extended overview ", 
+15, 
+7.5
+)
+```
 
@@
 The underlying algorithm of each method can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
-Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "method_components")`).
-MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "method_components", "b")`).
+Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
 
@@
 add_sfig(
-figure_file("method_components.svg","4-method_characterisation"), 
-"method_components", 
+figure_file("implementation_components.svg","4-method_characterisation"), 
+"implementation_components", 
 "TI methods frequently share steps"
@@
 
-We assessed whether the components of an algorithm can be predictive for method performance, using both random forest regression and statistical testing.
-Methods which include principal curves and clustering (k-means in particular) tend to perform better
+We assessed whether the components of an algorithm can be predictive for method performance, using both random forest classification and statistical testing.
+Methods which include principal curves (which include Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
 
+```{r}
+add_sfig(
+figure_file("method_components_scores.svg","6-developer_guidelines"), 
+"method_components_scores", 
+"Score of methods containing particular algorithmic components"
+)
+```
+
+```{r}
+add_stable(
+read_rds(figure_file("component_importance.rds","6-developer_guidelines"))[[params$table_format]], 
+"component_importance", 
+"Importance values for algorithmic components"
+)
+```
 ## Method quality control
@@
 We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
-The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or which purpose(s) they serve (user friendliness, developer friendliness or good science).
-These categorisations can help current developers to improve their tool, and to guide the selection of users and developers to use these tools for their purpose.
+The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
+These categorisations can help current developers to improve their tool, and guide the selection of users and developers to use these tools for their purpose.
 After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
@@
 "qc_checks", 
-"QC checks"
+"Scoring scheme for method quality control"
 )
@@
 
-Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "b")`).
-Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "results_overview", "e")`) and we believe that these should receive extra attention from developers.
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "b")`).
+Recent methods tended to have higher code quality (`r ref("sfig", "qc_over_time")`).
+However, several aspects were consistently lacking for the majority of the methods (`r ref("fig", "results_overview", "e")`) and we believe that these should receive extra attention from developers.
 Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "results_overview", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
@@
 ```
+
 ## Practical guidelines
@@
 
-* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
+* **DPT** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
 This was then processed as described above.
-* Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
+* **Sincell** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
 This was then processed as described above.
-* SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network.
+* **SLICER** Although this method returns a branch network and a global pseudotime, the method did not return a position of each cell on the individual branches.
 We therefore used the cell graph provided by the method to calculate the shortest paths between the start and end cells found by the method, and then used this graph as the backbone for all other cells, creating a local pseudotime for every cell within its branch.
 The output was then further processed as above.
-* SLICE: As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
-* MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
+* **SLICE** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
+* **MFA** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
 
@@
 
-TODO: describe which prior information was given and how it was extracted.
+All evaluated methods required or could use one of 7 types of prior information. This prior information was extracted from the gold/silver standards as follows:
 
-## Benchmark
-### Pipeline
-* Look at dyneval methods to write this paragraph
-### Method execution
-* Look at dynmethods::execute_method to write this paragraph
-
-## Method quality control
-
-```{r}
-qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
-  str_split("[, \n]+") %>% 
-  unlist() %>% 
-  unique() %>% 
-  discard(is.na) %>%
-  glue::collapse(., ";") %>%
-  paste0("[", ., "]")
-```
-
-We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
-The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
-The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
-The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
-The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
-The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
-The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
-The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
-Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
-
-Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
-This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
-For calculating the final score, we weighed each of the six categories equally.
+* **Start cells** The identity of one or more start cells.
+For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
+For ties, we chose one random cell.
+For cyclic datasets, a random cell was chosen.
+* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
+* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
+* **Grouping** For each cell a label to which state/cluster/branch it belongs.
+For real data, the states from the gold/silver standard.
+For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
+* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
+* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
 
@@
 
+## Benchmark
+
+### Pipeline
+* Look at dyneval methods to write this paragraph
+
+### Method execution
+* Look at dynmethods::execute_method to write this paragraph
+
+### Effect of prior information
+
+To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a
+
+### Effect of algorithm components
+To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score.
+To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test.
+We only investigated algorithm components which were part of at least 4 different methods in our evaluation study.
+
+## Method quality control
+
+```{r}
+qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
+  str_split("[, \n]+") %>% 
+  unlist() %>% 
+  unique() %>% 
+  discard(is.na) %>%
+  glue::collapse(., ";") %>%
+  paste0("[", ., "]")
+```
+
+We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
+The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
+The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
+The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
+The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
+The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
+The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
+Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
+
+Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
+For calculating the final score, we weighed each of the six categories equally.
+
 ## Evaluation metrics
@@
 Define in paragraph 1
-I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
-This is complex though to put into this figure :p.
-Have a look at 4-method_characteristics/wrapping.svg I think it is the least crucial of all parts of the figure now, so..
-But feel free to try it yourself :p
+WIP
+... wip ...
+mention timecourse prior
 Op zoek naar een betere formulering hiervoor, mischien:
@@
 Should we create a separate github containing a readme, referencing to the google doc, for this?
-..
 TODO

### 2018-03-04 analysis/paper/paper.Rmd
update paper

@@
 `r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview")`)
+![](`r figure_file("ti_evaluation_overview_v2.svg", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
-
-```{r}
-add_sfig({
-        figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview')
-}, 
-"wrapping", 
-"Extended overview ", 
-15, 
-7.5
-)
-```
 
@@
 This was then processed as described above.
-* **SLICER** Although this method returns a branch network and a global pseudotime, the method did not return a position of each cell on the individual branches.
+* **SLICER** Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network.
 We therefore used the cell graph provided by the method to calculate the shortest paths between the start and end cells found by the method, and then used this graph as the backbone for all other cells, creating a local pseudotime for every cell within its branch.
@@
 * **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
-
 ## Real datasets
@@
 As a final filter, we removed both all zero genes and cells until convergence.
-
 ## Benchmark
-
 ### Pipeline
 * Look at dyneval methods to write this paragraph
-
 ### Method execution
@@
 
-### Effect of prior information
+### Algorithm components
 
-To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a
-
-### Effect of algorithm components
-To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score.
-To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test.
-We only investigated algorithm components which were part of at least 4 different methods in our evaluation study.
+### Algorithm components
 

### 2018-03-04 analysis/paper/paper_latex.Rmd
update paper

@@
 add_sfig(
-figure_file("method_components.tmp.pdf","4-method_characterisation"), 
-"method_components", 
+figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
+"implementation_components", 
 "TI methods frequently share steps"
@@
 We assessed whether the components of an algorithm can be predictive for method performance, using both random forest regression and statistical testing.
-Methods which include principal curves and clustering (k-means in particular) tend to perform better
+Methods which include principal curves (which include Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not significant.
 
+```{r}
+add_sfig(
+figure_file("method_components_scores.tmp.pdf","6-developer_guidelines"), 
+"method_components_scores", 
+"Score of methods containing particular algorithmic components"
+)
+```
+
+```{r}
+add_stable(
+read_rds(figure_file("component_importance.rds","6-developer_guidelines"))[[params$table_format]], 
+"component_importance", 
+"Importance values for algorithmic components"
+)
+```
 ## Method quality control
@@
 Define in paragraph 1
+WIP
+... wip ...
 I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model

### 2018-03-04 analysis/paper/paper.Rmd

--- 2018-03-04 analysis/paper/paper.Rmd
@@
 We believe our evaluation framework can be used to spearhead the development for new methods.
-Our evaluation pipeline is fully reproducible, can be easily extended, and is available at github.com/dynverse/dynverse.
+Our evaluation pipeline is fully reproducible, can be easily extended, and is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
@@
 We gladly welcome any discussion and feedback on key decisions made as part of this study, including the metrics used in the benchmark, the quality control checklist, and the implementation of the method wrappers.
-These discussions can be held at github.com/dynverse/dynverse/issues.
+These discussions can be held at [github.com/dynverse/dynverse/issues](https://www.github.com/dynverse/dynverse/issues).
+
 # Introduction
-<!-- §1: Introduction to single-cell and TI --->
+<!-- §1: Single-cell omics and TI --->
 Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before.
@@
 
-```{r}
-uses_priors <- implementations %>% 
- gather(prior_id, prior_usage, !!priors$prior_id) %>% 
- filter(prior_usage %in% c("can_use", "can_root")) %>% 
- pull(implementation_id) %>% unique()
-
-requires_priors <- implementations %>% 
- gather(prior_id, prior_usage, !!priors$prior_id) %>% 
- filter(prior_usage %in% c("required", "required_default")) %>% 
- pull(implementation_id) %>% unique()
-
-```
-<!-- §2: plethora of new TI methods, with wildly varying characteristics -->
+<!-- §2: Plethora of new TI methods -->
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`).
@@
 ```
-<!-- §3: TOPOLOGY FIXING VERSUS TOPOLOGY INFERENCE -->
+<!-- §3: Topology inference -->
 One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
@@
 
-<!-- §4: PRIOR INFORMATION -->
+<!-- §4: Prior information -->
 Another key characteristic is the selection of prior information that it requires or can optionally exploit.
@@
 
-<!-- §5: COMPONENTS -->
+<!-- §5: Components -->
 A reductionist approach to characterising TI methods is dissecting it into a set of algorithmic components, as any one component can have a significant impact on the performance, scalability, and output data structures.
@@
 
-<!-- §5: Problem statement -->
+<!-- §6: Problem statement -->
 An important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
@@
 
-<!-- §6: Our contributions -->
-In this study, we perform a comprehensive evaluation for `r sum(methods$evaluated)` trajectory inference methods.
-As part of our evaluation framework (`r ref("fig", "evaluation_overview", "a")`), we developed a common probabilistic model for representing trajectories from all possible sources, and defined a set of metrics for comparing the likeness of such trajectories.
-The data compendium consists of both synthetic datasets for which we know the gold standard, and real scRNA-seq datasets for which the gold standard was extracted from expert knowledge.
-The generation of synthetic datasets was performed by simulating thermodynamics models of gene regulatory networks, and subsequently simulating a single cell profiling experiment.
-In addition to the quantitative benchmarking, we also assembled a quality control checklist evaluating the corresponding software and manuscript, including measures such as code quality, automated unit testing, and user friendliness.
-Finally, we present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
-
 # Results
-## Evaluation overview
-WIP
+In this study, we perform a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (Overview: `r ref("fig", "evaluation_overview", "a")`, Extended overview: `r ref("sfig", "extended_evaluation_overview")`).
 The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
 Currently, only methods published before june 2017 are included in the evaluation, while more recent methods will be included in the next version.
-We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
+The evaluation comprises of two core aspects: the assessment of the accuracy and scalability of each method by comparing its output with a gold standard, and a quality control of the software and documentation provided for each method.
 
-We evaluated the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "a")`).
-To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "d")` and `r ref("sfig", "wrapping")`.
-In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "d")`).
-This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
-We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
-We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
-To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
-To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
+In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In this model the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consists of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+The generation of synthetic datasets was performed by simulating thermodynamics models of gene regulatory networks, and subsequently simulating a single cell profiling experiment.
+
+As part of the quality control evaluation, we assembled a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+
+Finally, we present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
 
@@
 ![](`r figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview")`)
+
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 ```{r}
-add_sfig({
-        figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview')
-}, 
-"wrapping", 
-"Extended overview ", 
-15, 
-7.5
-)
-```
-
-```{r}
-add_sfig({
-        figure_file('wrapping.svg', '4-method_characterisation')
-}, 
-"wrapping", 
-"Our wrapping workflow", 
-15, 
-7.5
-)
-```
-
-```{r}
-add_stable(
-read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
-"datasets", 
-"Real datasets used in this study"
+add_sfig(
+        {figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview')}, 
+"extended_evaluation_overview", "Extended overview ", 15, 7.5
 )
@@
 ## Method benchmark
+An overview of the main results from this study is shown in `r ref("fig", "results_overview")`.
 
-Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
+Overall, we find that Slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
 These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
@@
 
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "b")`).
-Recent methods tended to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects were consistently lacking for the majority of the methods (`r ref("fig", "results_overview", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "results_overview", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_qc_overview")`).
+We found that recent methods had a slightly better quality than older methods (`r ref("sfig", "qc_over_time")`).
+However, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+
+`r anchor("fig", "implementation_qc_overview")`
+![](`r figure_file("implementation_qc_overview.svg","4-method_characterisation")`)
+`r ref("fig", "implementation_qc_overview")` Overview of the quality control results for every method.
 
@@
 Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with rarely any overlap.
-We therefore created wrappers around each method in R (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model.
+We therefore created wrappers around each method in R (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping")`).
 This model consisted of two parts.
@@
 The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
+
+```{r}
+add_sfig(
+        {figure_file('wrapping.svg', '4-method_characterisation')},
+"wrapping",  "Our wrapping workflow", 15, 7.5
+)
+```
 
@@
 ## Real datasets
-
+Overview:  (`r ref("sfig", "supp_real_data")`)
 We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
@@
 
+```{r}
+add_sfig(
+        {figure_file('supp_real_data.svg', 'evaluation_overview')},
+"supp_real_data",  "Overview of the real data", 15, 7.5
+)
+```
+
+```{r}
+add_stable(
+read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
+"datasets", 
+"Real datasets used in this study"
+)
+```
 ## Synthetic datasets
 
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "supp_synthetic_data")`).
 At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
@@
 
+<!-- REMOVED FROM THE MANUSCRIPT; MIGHT NEED TO BE MERGED WITH EXISTING TEXT HERE
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
+→
+
+```{r}
+add_sfig(
+        {figure_file('supp_synthetic_data.svg', 'evaluation_overview')},
+"supp_synthetic_data",  "Synthetic data generation workflow", 15, 7.5
+)
+```
 ### Network generation
@@
 Rephrase -_-
-Define in paragraph 1
-WIP
-... wip ...
 mention timecourse prior
@@
 In V2, "randomise" the prior information so that the positive effect of giving it to a method can be directly assessed, instead of this indirect evidence
-Should we create a separate github containing a readme, referencing to the google doc, for this?
+Make into sentence
+Check whether figure matches the 4 steps listed in the text
 TODO

### 2018-03-04 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
add qc overview figure

@@
 checks <- readRDS(derived_file("checks.rds"))
+
+implementation_order <- implementations_evaluated %>%
+  arrange(-qc_score) %>% 
+  pull(implementation_id)
+
+label_implementation <- function(implementation_id) {implementations_evaluated$implementation_name[match(implementation_id, implementations_evaluated$implementation_id)]}
+```
+
+Calculate the overall qc scores for each implementation
+```{r}
+# set the ordering of implementation_ids to this ordering
+implementation_ordering_plot <- implementations_evaluated %>%
+  mutate(implementation_id = factor(implementation_id, implementation_order)) %>% 
+  ggplot(aes(implementation_id, qc_score)) +
+    geom_bar(aes(fill=qc_score), stat="identity") +
+    geom_text(aes(y=0.025, label=label_implementation(implementation_id), color=ifelse(qc_score > 0.5, "black", "white")), angle=90, hjust=0, size=5) +
+    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
+    scale_y_continuous(NULL, expand=c(0, 0), limits=c(0, 1)) +
+    scale_x_discrete(NULL, breaks=c(0, 100), expand=c(0, 0)) +
+    viridis::scale_fill_viridis(label_long("qc_score"), option="D", guide=guide_colorbar(title.position = "top", title.hjust=0.5, barwidth = unit(2, "inches"))) +
+    scale_color_identity() +
+    annotate("segment", x=-Inf, xend=Inf, y=-Inf, yend=-Inf)+
+    theme(legend.position="top")
+
+implementation_ordering_plot
+implementation_ordering_plot %>% write_rds(figure_file("implementation_qc_ordering.rds"))
+```
+
+```{r, fig.width=10, fig.height=12}
+qc_checks <- implementation_qc %>%
+  filter(implementation_id == "scorpius") %>%
+  group_by(category) %>% 
+  mutate(normalised_weight = weight/sum(weight))
+
+aspect_position <- qc_checks %>% 
+  arrange(aspect_id) %>%
+  group_by(aspect_id) %>%
+  summarise(
+    aspect_width = sum(item_weight * normalised_weight),
+    name = first(name)
+  ) %>%
+  mutate(
+    aspect_end = cumsum(aspect_width),
+    aspect_start = lag(aspect_end, 1, default=0),
+    aspect_mid = aspect_start + (aspect_end - aspect_start)/2
+  )
+
+check_position <- qc_checks %>%
+  arrange(aspect_id, check_id) %>%
+  group_by(check_id) %>%
+  summarise(
+    check_width = item_weight * normalised_weight,
+    aspect_id = first(aspect_id)
+  ) %>%
+  left_join(aspect_position, by="aspect_id") %>%
+  group_by(aspect_id) %>%
+  mutate(
+    check_start = aspect_start + lag(cumsum(check_width), default=0),
+    check_end = aspect_start + cumsum(check_width)
+  )
+
+hline_check <- geom_hline(aes(yintercept=check_end), data=check_position, color="#444444", alpha=0.25)
+hline_aspect <- geom_hline(aes(yintercept=aspect_end), data=aspect_position, color="#444444")
+
+implementation_qc_checks <- implementation_qc %>% 
+  mutate(implementation_id = factor(implementation_id, implementation_order)) %>% 
+  left_join(check_position, by=c("check_id", "name", "aspect_id")) %>%
+  mutate(answer = ifelse(is.na(answer), 0, answer))
+
+check_heatmap_plot <- implementation_qc_checks %>% 
+  ggplot() +
+    geom_rect(aes(ymax=check_start, ymin=check_end, xmin=as.integer(implementation_id), xmax=as.integer(implementation_id) + 1, fill=category, alpha=answer)) +
+  hline_check + hline_aspect +
+    geom_vline(aes(xintercept = as.integer(implementation_id)), color="white", alpha=0.2) +
+    scale_y_reverse(NULL, breaks = aspect_position$aspect_mid, labels = aspect_position$name, expand = c(0, 0)) +
+    # scale_x_continuous("", breaks = seq_along(levels(implementation_qc_checks$implementation_id))+0.5, labels = label_implementation(levels(implementation_qc_checks$implementation_id)), expand = c(0, 0), position="right") +
+    scale_x_continuous("", breaks=NULL, expand = c(0, 0)) +
+    scale_fill_manual("Category", values=set_names(qc_categories$color, qc_categories$category), label=label_long, guide=guide_legend(ncol=2)) +
+    scale_alpha_continuous("Score", guide=guide_legend(ncol=5, label.position = "bottom")) +
+    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.text.y = element_text(size=12), legend.position="bottom")
+
+check_heatmap_plot
+```
+
+```{r}
+check_scores <- implementation_qc_checks %>% group_by(check_id) %>% summarise(score = mean(answer))
+aspect_scores <- implementation_qc_checks %>% group_by(aspect_id) %>% summarise(score = mean(answer))
+
+check_difficulty_data <- checks %>% 
+  left_join(check_scores, "check_id") %>% 
+  left_join(check_position, "check_id")
+
+check_difficulty_heatmap <- check_difficulty_data %>% 
+  ggplot() + 
+    geom_rect(aes(ymin=check_start, ymax=check_end, xmin=0, xmax=1), fill="#DDDDDD") +
+    geom_rect(aes(ymin=check_start, ymax=check_end, xmin=0, xmax=1, fill=score)) +
+    ggrepel::geom_label_repel(
+      aes(
+        1, 
+        (check_end+check_start)/2,
+        label=label_wrap(item, 35)
+      ), 
+      nudge_x = 6, 
+      direction="y", 
+      data=check_difficulty_data %>% 
+        filter(score <= 0.5), 
+      size=5, 
+      lineheight=0.8, 
+      min.segment.length = 0
+    ) +
+    geom_segment(aes(x=0, xend=1, y=check_end, yend=check_end), data=check_position, color="#444444", alpha=0.25) +
+    geom_segment(aes(x=0, xend=1, y=aspect_end, yend=aspect_end), data=aspect_position, color="#444444") +
+    # hline_check + hline_aspect +
+    # scale_fill_manual(values=set_names(qc_categories$color, qc_categories$category)) +
+    # scale_fill_distiller("QC\ndifficulty", palette="Spectral", limits=c(0,1), breaks=c(0,0.5, 1), direction=1) +
+    scale_fill_gradientn("Average QC score across methods", colors = c("#9e0142", "#d53e4f", "#f46d43", "#fee08b", "#abdda4", "#1a9850"), limits=c(0,1), breaks=c(0,0.5, 1), guide=guide_colorbar(title.position = "bottom", title.hjust=0.5, barwidth = unit(2, "inches"))) +
+    # viridis::scale_fill_viridis("QC\ndifficulty", limits=c(0,1), breaks=c(0,0.5, 1)) +
+    scale_x_continuous(NULL, breaks=NULL, limits=c(0, 12), expand=c(0,0)) +
+    scale_y_reverse(NULL, breaks=NULL, expand=c(0,0)) +
+    annotate("segment", x=-Inf, xend=-Inf, y=-Inf, yend=Inf)+
+    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.position = "bottom")
+
+check_difficulty_heatmap
+```
+
+Combined plot
+```{r, fig.width=10, fig.height=12}
+empty_x <- theme(
+  axis.title.x=element_blank(),
+  axis.text.x=element_blank(),
+  axis.ticks.x=element_blank()
+)
+empty_y <- theme(
+  axis.title.y=element_blank(),
+  axis.text.y=element_blank(),
+  axis.ticks.y=element_blank(),
+  axis.line.y = element_blank()
+)
+empty_margin <- theme(plot.margin = margin(0.05, 0.05, 0.05, 0.05, unit="inches"))
+
+a <- cowplot::plot_grid(
+  implementation_ordering_plot + empty_margin,
+  NULL,
+  check_heatmap_plot + empty_margin, 
+  NULL,
+  nrow=2, ncol=2,
+  align = "v", 
+  axis = "rl", 
+  rel_heights = c(0.2, 0.8),
+  rel_widths = c(0.8, 0.3)#,
+  # labels = c("a", "", "b", "c")
+)
+a
+b <- cowplot::plot_grid(
+  NULL,
+  NULL,
+  check_heatmap_plot + empty_margin, 
+  check_difficulty_heatmap +empty_y + empty_margin,
+  nrow=2, ncol=2,
+  align = "h", 
+  axis = "bt", 
+  rel_heights = c(0.2, 0.8),
+  rel_widths = c(0.8, 0.3)#,
+  # labels = c("a", "", "b", "c")
+)
+b
+
+g <- ggdraw(b)+draw_plot(a)
+
+g
+
+g %>% save_plot(figure_file("implementation_qc_overview.svg"), ., base_width=15, base_height=18)
 ```
@@
     geom_point() +
-    ggrepel::geom_label_repel(aes(label = implementation_name)) +
-    cowplot::theme_cowplot() +
+    ggrepel::geom_text_repel(aes(label = label_implementation(implementation_id))) +
     labs(x = label_long("publication_date"), y = "QC score", title = "Code quality score over time")
@@
     geom_point() +
-    ggrepel::geom_label_repel(aes(label = implementation_name)) +
-    cowplot::theme_cowplot() +
+    ggrepel::geom_text_repel(aes(label = label_implementation(implementation_id))) +
     scale_y_log10() +
@@
     geom_point() +
-    ggrepel::geom_label_repel(aes(label = implementation_name)) +
-    cowplot::theme_cowplot() +
+    ggrepel::geom_text_repel(aes(label = label_implementation(implementation_id))) +
     scale_x_log10() +
@@
   save_plot(figure_file("qc_over_time.svg"), ., base_width=10, base_height=10)
-```
-
-Calculate the overall qc scores for each implementation
-```{r}
-# set the ordering of implementation_ids to this ordering
-implementation_ordering_plot <- implementations_evaluated %>%
-  arrange(-qc_score) %>% 
-  mutate(implementation_name = factor(implementation_name, levels=implementation_name)) %>% 
-  ggplot(aes(implementation_name, qc_score)) +
-    geom_bar(aes(fill=qc_score), stat="identity") +
-    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
-    scale_y_continuous(expand=c(0, 0), limits=c(0, 1)) +
-    viridis::scale_fill_viridis(option="A", limits=c(0, 1)) +
-    labs(y=label_long("qc_score"), x=label_long("implementation_name")) +
-    theme(legend.position="none")
-
-implementation_ordering_plot
-implementation_ordering_plot %>% write_rds(figure_file("implementation_qc_ordering.rds"))
-```
-
-```{r, fig.width=10, fig.height=12}
-qc_checks <- implementation_qc %>%
-  filter(implementation_id == "scorpius") %>%
-  group_by(category) %>% 
-  mutate(normalised_weight = weight/sum(weight))
-
-aspect_position <- qc_checks %>% 
-  arrange(aspect_id) %>%
-  group_by(aspect_id) %>%
-  summarise(
-    aspect_width = sum(item_weight * normalised_weight),
-    name = first(name)
-  ) %>%
-  mutate(
-    aspect_end = cumsum(aspect_width),
-    aspect_start = lag(aspect_end, 1, default=0),
-    aspect_mid = aspect_start + (aspect_end - aspect_start)/2
-  )
-
-check_position <- qc_checks %>%
-  arrange(aspect_id, check_id) %>%
-  group_by(check_id) %>%
-  summarise(
-    check_width = item_weight * normalised_weight,
-    aspect_id = first(aspect_id)
-  ) %>%
-  left_join(aspect_position, by="aspect_id") %>%
-  group_by(aspect_id) %>%
-  mutate(
-    check_start = aspect_start + lag(cumsum(check_width), default=0),
-    check_end = aspect_start + cumsum(check_width)
-  )
-
-check_heatmap_plot <- implementation_qc %>%
-  left_join(check_position, by="check_id") %>%
-  mutate(implementation_id = factor(implementation_id)) %>%
-  mutate(answer = ifelse(is.na(answer), 0, answer)) %>%
-  ggplot() +
-    geom_rect(aes(xmax=check_start, xmin=check_end, ymin=as.integer(implementation_id), ymax=as.integer(implementation_id) + 1, fill=category, alpha=answer)) +
-    geom_vline(aes(xintercept=check_end), data=check_position, color="#444444", alpha=0.25) +
-    geom_vline(aes(xintercept=aspect_end), data=aspect_position, color="#444444") +
-    scale_x_reverse("Check", breaks = aspect_position$aspect_mid, labels = aspect_position$name, expand = c(0, 0)) +
-    scale_y_continuous("implementation", breaks = seq_along(levels(implementation_qc$implementation_id))+0.5, labels = levels(implementation_qc$implementation_id), expand = c(0, 0)) +
-    scale_fill_manual(values=set_names(qc_categories$color, qc_categories$category)) +
-    coord_flip() +
-    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
-
-check_heatmap_plot
-```
-
-Combined plot
-```{r, fig.width=10, fig.height=12}
-empty_x <- theme(
-  axis.title.x=element_blank(),
-  axis.text.x=element_blank(),
-  axis.ticks.x=element_blank()
-)
-
-implementation_qc_check_plot <- cowplot::plot_grid(implementation_ordering_plot + empty_x, check_heatmap_plot, nrow=2, align = "v", axis = "rl", rel_heights = c(0.2, 0.9))
-
-implementation_qc_check_plot %>% saveRDS(figure_file("implementation_qc_check.rds"))
 ```

### 2018-03-04 analysis/paper/paper_latex.Rmd
some fixes to qc checks and update paper

@@
 
-<!-- §6: Our contributions -->
-In this study, we perform a comprehensive evaluation for `r sum(methods$evaluated)` trajectory inference methods.
-As part of our evaluation framework (`r ref("fig", "evaluation_overview", "a")`), we developed a common probabilistic model for representing trajectories from all possible sources, and defined a set of metrics for comparing the likeness of such trajectories.
-The data compendium consists of both synthetic datasets for which we know the gold standard, and real scRNA-seq datasets for which the gold standard was extracted from expert knowledge.
+# Results
+In this study, we perform a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (Overview: `r ref("fig", "evaluation_overview", "a")`, Extended overview: `r ref("sfig", "extended_evaluation_overview")`).
+The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
+Currently, only methods published before june 2017 are included in the evaluation, while more recent methods will be included in the next version.
+The evaluation comprises of two core aspects: the assessment of the accuracy and scalability of each method by comparing its output with a gold standard, and a quality control of the software and documentation provided for each method.
+
+In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In this model the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consists of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
 The generation of synthetic datasets was performed by simulating thermodynamics models of gene regulatory networks, and subsequently simulating a single cell profiling experiment.
-In addition to the quantitative benchmarking, we also assembled a quality control checklist evaluating the corresponding software and manuscript, including measures such as code quality, automated unit testing, and user friendliness.
+
+As part of the quality control evaluation, we assembled a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+
 Finally, we present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
 
-# Results
-## Evaluation overview
-WIP
-The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
-Currently, only methods published before june 2017 are included in the evaluation, while more recent methods will be included in the next version.
-We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools.
+`r anchor("fig", "evaluation_overview")`
+![](`r figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="evaluation_overview")`)
 
-We evaluated the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "a")`).
-To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "d")` and `r ref("sfig", "wrapping")`.
-In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "d")`).
-This common trajectory format allowed us to compare the output of each method against gold standard trajectories.
-We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
-We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver).
-To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
-To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.
-
-`r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v2.tmp.pdf", experiment_id="evaluation_overview")`)
 `r ref("fig", "evaluation_overview")` Overview of the evaluation
@@
 ```{r}
-add_sfig({
-        figure_file('wrapping.tmp.pdf', '4-method_characterisation')
-}, 
-"wrapping", 
-"Our wrapping workflow", 
-15, 
-7.5
-)
-```
-
-```{r}
-add_stable(
-read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
-"datasets", 
-"Real datasets used in this study"
+add_sfig(
+        {figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview')}, 
+"extended_evaluation_overview", "Extended overview ", 15, 7.5
 )
@@
 
-We assessed whether the components of an algorithm can be predictive for method performance, using both random forest regression and statistical testing.
+We assessed whether the components of an algorithm can be predictive for method performance, using both random forest classification and statistical testing.
 Methods which include principal curves (which include Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
-On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not significant.
+On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
 
@@
 
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "b")`).
-Recent methods tended to have higher code quality (`r ref("sfig", "qc_over_time")`).
-However, several aspects were consistently lacking for the majority of the methods (`r ref("fig", "results_overview", "e")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "implementation_qc_check")` and `r ref("fig", "results_overview", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_qc_overview")`).
+We found that recent methods had a slightly better quality than older methods (`r ref("sfig", "qc_over_time")`).
+However, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+
+`r anchor("fig", "implementation_qc_overview")`
+![](`r figure_file("implementation_qc_overview.tmp.pdf","4-method_characterisation")`)
+`r ref("fig", "implementation_qc_overview")` Overview of the quality control results for every method.
 
@@
 The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
+
+```{r}
+add_sfig(
+        {figure_file('wrapping.tmp.pdf', '4-method_characterisation')},
+"wrapping",  "Our wrapping workflow", 15, 7.5
+)
+```
 
@@
 
-TODO: describe which prior information was given and how it was extracted.
+All evaluated methods required or could use one of 7 types of prior information. This prior information was extracted from the gold/silver standards as follows:
 
-## Benchmark
-### Pipeline
-* Look at dyneval methods to write this paragraph
-### Method execution
-* Look at dynmethods::execute_method to write this paragraph
-
-## Method quality control
-
-```{r}
-qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
-  str_split("[, \n]+") %>% 
-  unlist() %>% 
-  unique() %>% 
-  discard(is.na) %>%
-  glue::collapse(., ";") %>%
-  paste0("[", ., "]")
-```
-
-We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
-The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
-The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
-The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
-The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
-The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
-The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
-The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
-Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
-
-Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
-This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
-For calculating the final score, we weighed each of the six categories equally.
+* **Start cells** The identity of one or more start cells.
+For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
+For ties, we chose one random cell.
+For cyclic datasets, a random cell was chosen.
+* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
+* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
+* **Grouping** For each cell a label to which state/cluster/branch it belongs.
+For real data, the states from the gold/silver standard.
+For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
+* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
+* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
 
 ## Real datasets
-
+Overview:  (`r ref("sfig", "supp_real_data")`)
 We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
@@
 
+```{r}
+add_sfig(
+        {figure_file('supp_real_data.tmp.pdf', 'evaluation_overview')},
+"supp_real_data",  "Overview of the real data", 15, 7.5
+)
+```
+
+```{r}
+add_stable(
+read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
+"datasets", 
+"Real datasets used in this study"
+)
+```
 ## Synthetic datasets
 
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "supp_synthetic_data")`).
 At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
@@
 
+<!-- REMOVED FROM THE MANUSCRIPT; MIGHT NEED TO BE MERGED WITH EXISTING TEXT HERE
+These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
+We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
+→
+
+```{r}
+add_sfig(
+        {figure_file('supp_synthetic_data.tmp.pdf', 'evaluation_overview')},
+"supp_synthetic_data",  "Synthetic data generation workflow", 15, 7.5
+)
+```
 ### Network generation
@@
 Rephrase -_-
-Define in paragraph 1
-WIP
-... wip ...
-I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
-This is complex though to put into this figure :p.
-Have a look at 4-method_characteristics/wrapping.tmp.pdf I think it is the least crucial of all parts of the figure now, so..
-But feel free to try it yourself :p
+mention timecourse prior
 Op zoek naar een betere formulering hiervoor, mischien:
@@
 In V2, "randomise" the prior information so that the positive effect of giving it to a method can be directly assessed, instead of this indirect evidence
-Should we create a separate github containing a readme, referencing to the google doc, for this?
-..
+Make into sentence
+Check whether figure matches the 4 steps listed in the text
 TODO

### 2018-03-04 analysis/paper/paper_latex.Rmd
update paper

@@
 Currently, only methods published before june 2017 are included in the evaluation, while more recent methods will be included in the next version.
-The evaluation comprises of two core aspects: the assessment of the accuracy and scalability of each method by comparing its output with a gold standard, and a quality control of the software and documentation provided for each method.
+The evaluation comprises of three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) the assessment of the accuracy and scalability by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
 
@@
 In this model the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
-The data compendium consists of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-The generation of synthetic datasets was performed by simulating thermodynamics models of gene regulatory networks, and subsequently simulating a single cell profiling experiment.
+We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets from a variety of single-cell technologies, organisms, trajectory topologies and dynamic processes (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`), which offer the highest biological relevance.
+To generate synthetic datasets, we simulated a gene regulatory networks using a thermodynamic model [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
+
+```{r}
+add_sfig(
+        figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
+"dataset_characterisation",  
+"Characteristics of the real datasets used for the evaluation"
+)
+```
 
@@
 An overview of the main results from this study is shown in `r ref("fig", "results_overview")`.
+This includes an overview of the results obtained from the method characterisation (`r ref("fig", "results_overview", "a")`), the benchmarking evaluation (`r ref("fig", "results_overview", "b")`), and the quality control evaluation (`r ref("fig", "results_overview", "c")`).
 
 Overall, we find that Slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
-These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards.
+These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets.
+
 However, results are more variable when comparing the three different evaluation metrics. 
@@
 ![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
-`r ref("fig", "results_overview")` Overview of characteristics and benchmark of `r sum(methods$evaluated)` evaluated methods
+`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the maximum complexity trajectory they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally make use of.
+b) The
 
@@
 ```
+We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`).
+We could also not find any quality aspect which was significantly associated with method performance.
+
+```{r}
+add_sfig(
+figure_file("qc_performance_comparison.tmp.pdf","5-optimise_parameters/6-qc_performance_comparison"), 
+"qc_performance_comparison", 
+"Comparison between method qc (x-axis) and method performance (y-axis)."
+)
+```
 
@@
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each case, a different set of methods performs optimally, with Monocle DDRTree performing best when the data contains a complex tree, while Slingshot performing better when the trajectory could as well be less complex.
-When the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT `r cite_methods("wadot")` and AGA `r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
-If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
+In each such use case, a different set of methods performs optimally, with Monocle DDRTree performing best when the data contains a complex tree, while Slingshot performing better when the trajectory could as well be less complex.
+For the use case where the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT `r cite_methods("wadot")` and AGA `r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
+If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
@@
 
-<!----
-Limitations of the model:
-* Robrecht.Cannoodt @robrechtc 10:21
-en heterogene output van alle methode... true, maar sommigen hebben we wel wat moeten tweaken 😛
-en ik kan me output bedenken dat niet in ons model zou kunnen passen
-* Wouter Saelend @wouters 10:21
-oh 😛
-* Robrecht.Cannoodt @robrechtc 10:21
-welja
-* Wouter Saelend @wouters 10:21
-Misschien een kritiekpunt is onzekerheid... maarja 😛
-* Robrecht.Cannoodt @robrechtc 10:22
-onzekerheid, en dat een cell in meerdere niet-neighbouring milestones zou kunnen zitten
-als in, biologisch gezien houdt het mss geen steek, maar een methode zou wel kunnen zeggen, "wete, ik denk dat het ofwel een DC is of een rode bloedcel"
-* Wouter Saelend @wouters 10:24
-hmmm uhu true, hoort allemaal wat samen met de onzekerheid. Je zou per cell een probabiliteit distributie doorheen het milestone netwerk kunnen geven, maar sla dat eens op 😛
-* Robrecht.Cannoodt @robrechtc 10:25
-uhu ^^
-* Wouter Saelend @wouters 10:25
-En bereken dan eens de emd 😛
-* Robrecht.Cannoodt @robrechtc 10:25
-idd...
-deze discussie zou wel ergens een plaatsje kunnen krijgen in het manuscript, vind ik
-* Wouter Saelend @wouters 10:25
-Uhu in de discussie dan wss
-* Robrecht.Cannoodt @robrechtc 10:26
-tiens, een discussie die in de discussie zou plaats vinden, wat een goed idee 😮
-<tease>
--->
 # Methods
@@
 ## Real datasets
-Overview:  (`r ref("sfig", "supp_real_data")`)
-We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets are all available in our repository (http://www.github.com/dynverse/dynalysis).
+We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")` and `r ref("sfig", "dataset_characterisation")`).
+The scripts to download and process these datasets will be made available n our repository (http://www.github.com/dynverse/dynalysis).
 When available, we preferred to start from the raw counts data.
@@
 We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).
-
-```{r}
-add_sfig(
-        {figure_file('supp_real_data.tmp.pdf', 'evaluation_overview')},
-"supp_real_data",  "Overview of the real data", 15, 7.5
-)
-```
 
@@
 
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "supp_synthetic_data")`).
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "synthetic_data_generation")`).
 At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
@@
 
-<!-- REMOVED FROM THE MANUSCRIPT; MIGHT NEED TO BE MERGED WITH EXISTING TEXT HERE
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
-→
-
 ```{r}
 add_sfig(
-        {figure_file('supp_synthetic_data.tmp.pdf', 'evaluation_overview')},
-"supp_synthetic_data",  "Synthetic data generation workflow", 15, 7.5
+        figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
+"synthetic_data_generation",  "Synthetic data generation workflow"
 )
@@
 
-§ Cascades? Cycle?
-
-§ To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
+To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
 Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated trajectories which simple undirected graphs (bifurcating loop and bifurcating convergence).
@@
 
-§ Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward.
-Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?.
-To estimate a cell's position during a simulation within the trajectory toplogy, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
+Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward.
+Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable.
+To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
 We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
@@
 We used three different metrics to assess the performance of each TI method.
-The importance of using multiple metrics to compare complex models has been stated repeatedly REFS.
+The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
 Indeed, each of our metrics looks at the correspondence between prediction and gold standard in a different way: the correlation between geodesic distances assesses the overall ordering, the RF prediction error assesses the cell neighbourhood, and the edge flip score assesses the milestone network topology (`r ref("sfig", "score_aspects")`).
@@
 Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
-This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no suitable …?
-[@bahiense_maximumcommonedge_2012].
+This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no scalable solutions [@bahiense_maximumcommonedge_2012].
 We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes.
@@
 mention timecourse prior
-Op zoek naar een betere formulering hiervoor, mischien:
-"In most cases, being able to detect a trajectory type leads to a better score on data containing this particular trajectory type"
 😉
@@
 In V2, "randomise" the prior information so that the positive effect of giving it to a method can be directly assessed, instead of this indirect evidence
-Make into sentence
-Check whether figure matches the 4 steps listed in the text
 TODO
@@
 this is not used anymore
-suitable what?

### 2018-03-04 analysis/paper/paper.Rmd
update paper

@@
 In this model the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "c")`).
-The data compendium consists of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-The generation of synthetic datasets was performed by simulating thermodynamics models of gene regulatory networks, and subsequently simulating a single cell profiling experiment.
+We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets from a variety of single-cell technologies, organisms, trajectory topologies and dynamic processes (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`), which offer the highest biological relevance.
+To generate synthetic datasets, we simulated a gene regulatory networks using a thermodynamic model [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
+
+```{r}
+add_sfig(
+        figure_file('dataset_characterisation.svg', '2-dataset_characterisation'),
+"dataset_characterisation",  
+"Characteristics of the real datasets used for the evaluation"
+)
+```
 
@@
 ![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
-`r ref("fig", "results_overview")` Overview of characteristics and benchmark of `r sum(methods$evaluated)` evaluated methods
+`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the maximum complexity trajectory they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally make use of.
+b) The
 
@@
 ```
+We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`).
+We could also not find any quality aspect which was significantly associated with method performance.
+
+```{r}
+add_sfig(
+figure_file("qc_performance_comparison.svg","5-optimise_parameters/6-qc_performance_comparison"), 
+"qc_performance_comparison", 
+"Comparison between method qc (x-axis) and method performance (y-axis)."
+)
+```
 
@@
 We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).
-
-```{r}
-add_sfig(
-        {figure_file('supp_real_data.svg', 'evaluation_overview')},
-"supp_real_data",  "Overview of the real data", 15, 7.5
-)
-```
 
@@
 
-<!-- REMOVED FROM THE MANUSCRIPT; MIGHT NEED TO BE MERGED WITH EXISTING TEXT HERE
-These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of this dataset as best as possible (`r ref("fig", "evaluation_overview", "d")`).
-We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")` and `r ref("stable", "datasets")`).
-→
-
 ```{r}
 add_sfig(
-        {figure_file('supp_synthetic_data.svg', 'evaluation_overview')},
-"supp_synthetic_data",  "Synthetic data generation workflow", 15, 7.5
+        figure_file('synthetic_data_generation.svg', "2-dataset_characterisation/3-synthetic"),
+"synthetic_data_generation",  "Synthetic data generation workflow"
 )

### 2018-03-05 analysis/paper/paper_latex.Rmd
avoid figures scaling like crazy

@@
 We believe our evaluation framework can be used to spearhead the development for new methods.
-Our evaluation pipeline is fully reproducible, can be easily extended, and is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+Our evaluation pipeline is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
 We believe this to be the first comprehensive assessment of trajectory inference methods.
+For now, we only evaluated the methods on default parameters, but plan to add a detailed parameter tuning procedure in the future.
 We gladly welcome any discussion and feedback on key decisions made as part of this study, including the metrics used in the benchmark, the quality control checklist, and the implementation of the method wrappers.
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-When cells are profiled from a population in which cells are at different unknown points in the dynamic process, dynamic processes can be modelled using trajectory inference (TI) methods  [@cannoodt_computational_2016].
-These methods computationally order these cells along a trajectory topology, which can be linear, bifurcating, or more complex.
+When cells are profiled from a population in which cells are at different unknown points in the dynamic process, this process  processes can be modelled using trajectory inference (TI) methods  [@cannoodt_computational_2016].
+These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex.
 TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017].
@@
 We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` TI methods (`r ref("table", "implementations")`), which we believe to be the most complete list to date.
-It is perhaps surprising that almost all of them have a unique combination of characteristics, in terms of the required inputs (prior information), produced outputs (topology fixing, trajectory type) and methodology used (not shown).
+It is perhaps surprising that almost all of them have a unique combination of characteristics, in terms of the required inputs (prior information), produced outputs (topology fixing and trajectory type) and methodology used.
 
@@
 Early TI methods typically fixed the topology algorithmically (e.g.
-linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015], bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]) or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
+linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]) or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
 The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
@@
 Most noticeable of these methods is AGA[@wolf_graphabstractionreconciles_2017], which is the only TI method currently able to predict disconnected graphs.
-It is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
 
 <!-- §4: Prior information -->
-Another key characteristic is the selection of prior information that it requires or can optionally exploit.
-Examples of prior informations are supplying a starting cell to start the trajectory from, a set of important marker genes, or even a grouping of cells into cell states.
+Another key characteristic is the selection of prior information that a method requires or can optionally exploit.
+Examples of prior informations are supplying a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
 Giving prior information to a TI method can be both a blessing and a curse.
@@
 Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
+
+```{r}
+add_sfig(
+figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
+"implementation_components", 
+"TI methods frequently share steps"
+)
+```
 
 <!-- §6: Problem statement -->
-An important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
-Many attempts at addressing this issue has already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, performing a comprehensive benchmarking evaluation of TI methods requires a lot of work.
+Given this diversity in TI methods, an important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
+Many attempts at addressing this issue has already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, performing a comprehensive benchmarking evaluation of TI methods is still lacking.
 This is problematic, as new users to the field are confronted with a wide array of TI methods without a clue as to what the most optimal method available for their problem could be. 
@@
 # Results
-In this study, we perform a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (Overview: `r ref("fig", "evaluation_overview", "a")`, Extended overview: `r ref("sfig", "extended_evaluation_overview")`).
+In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (Overview: `r ref("fig", "evaluation_overview", "a")`, Extended overview: `r ref("sfig", "extended_evaluation_overview")`).
 The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
-Currently, only methods published before june 2017 are included in the evaluation, while more recent methods will be included in the next version.
-The evaluation comprises of three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) the assessment of the accuracy and scalability by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
+Currently, only methods published before june 2017 are included in the current version of the evaluation, while more recent methods will be included in the next version.
+The evaluation comprises of three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) the assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
 
@@
 In this model the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the likeness in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
-The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets from a variety of single-cell technologies, organisms, trajectory topologies and dynamic processes (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`), which offer the highest biological relevance.
+We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
 To generate synthetic datasets, we simulated a gene regulatory networks using a thermodynamic model [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
@@
 
-As part of the quality control evaluation, we assembled a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+```{r}
+add_sfig(
+        figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
+"synthetic_data_generation",  "Synthetic data generation workflow"
+)
+```
 
-Finally, we present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+We not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation.
+We scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
+
+Using both the benchmark and quality control results, we will finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
 
@@
 
-`r ref("fig", "evaluation_overview")` Overview of the evaluation
+`r ref("fig", "evaluation_overview")` **An overview of several key aspects of the evaluation.** a) .
+b) In order to make any two trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
+c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
 
@@
 
-## Method benchmark
+## Evaluation of trajectory inference methods
+<!-- §1: Introduce figure --> 
 An overview of the main results from this study is shown in `r ref("fig", "results_overview")`.
@@
 
-Overall, we find that Slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
-These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets.
-
-However, results are more variable when comparing the three different evaluation metrics. 
-Some methods, mainly those able to detect more complex trajectory types, have higher edge flip scores.
-Other methods, not necessarily those with the highest edge flip score, are better at preserving the ordering and neighbourhood of the cells.
-
 `r anchor("fig", "results_overview")`
 ![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
+
 `r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the maximum complexity trajectory they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally make use of.
-b) The
+b) The methods are ordered according to the overall score achieved in the benchmark evaluation.
+Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+c) Overall performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
+Also listed are the quality control scores aggregated according to practicality and the different categories.
+
+<!-- §2: Overall and topology types -->
+Ordering all methods by their overall benchmarking score, we found that Slingshot predicts the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
+Observing the benchmark scores per trajectory type, Slingshot seems to be the only method that performs well across most trajectory types.
+However, several methods are seemingly specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, Monocle DDRTree for rooted trees, and, surprisingly Wishbone for multifurcations.
+We will discuss the effect of topology types and constraints on inferrable trajectory types placed by the method in Section SEC_TRAJTYPE.
+
+<!-- §3: real versus synthetic -->
+A high correlation can be observed between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`).
+This confirms both the relevance of the synthetic data and the accuracy of the real datasets.
 
@@
 
+<!-- §4: Compare metrics -->
+As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is unsurprising to observe differences in rankings of TI methods amongst the different metrics.
+While there is no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also seed to obtain a higher RF MSE score (`r ref("sfig", "metric_comparison")`).
+In addition, methods able to detect more complex trajectory types seem to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
+This issue will also be investigated in more detail in Section SEC_TRAJTYPE.
+
+```{r}
+add_sfig(
+figure_file('metrics_comparison.tmp.pdf', "5-optimise_parameters/7-metric_comparison"), 
+"metric_comparison", 
+"Comparison between the different metrics across trajectory types."
+)
+```
+
+During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded (>6h) or an error was produced, a zero score would be returned for that execution.
+If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
+Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
+Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
+Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largests of datasets (reCAT, SCOUP, StemID).
+For a few methods, the time or memory limits were exceeded to frequently that insufficient results were generated to compare it to other methods (SCIMITAR, Ouija, Pseudogp).
+
+Comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
+We hope the identifications of the main strengths and weaknesses of existing TI methods will lead to the next few state-of-the-art improvements.
+
+```{r}
+add_sfig(
+figure_file('leading_method.tmp.pdf', "9-main_figure"), 
+"leading_method", 
+"Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
+)
+```
+
 ### Trajectory types and topology complexity
@@
 These methods had typically better edge flip scores - as can be expected - and RF MSE scores compared to methods not able to handle the particular trajectory type.
-However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced a better ordering, irrespective of whether the dataset contained a linear trajectory or more complex
+However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
 To further investigate, we divided all methods in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
-While there were no significant difference on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "'linear_split_comparison")`).
-Together, this indicates that current non-linear methods contain less accurate ordering algorithms.
-Inclusion of the ordering algorithms from the top linear methods could therefore improve the performance of the non-linear methods.
+While there were no significant differences on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`).
+Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
+A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.
 
@@
 ```
+
 ### Effect of prior information
-
-Giving prior information to a TI method can be both a blessing and a curse.
-In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
-On the other hand, incorrect or noisy prior information can bias the trajectory towards a wrong, but perhaps consistent with current knowledge, trajectory.
-Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, limiting the applicability of such TI methods to only well studied systems.
 
@@
 "required_priors_performance_comparison", 
-""
+"Comparison of method performance "
 )
@@
 
-### Running time and scalability
-
 ### Algorithm components
 
-The underlying algorithm of each method can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
-Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-```{r}
-add_sfig(
-figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
-"implementation_components", 
-"TI methods frequently share steps"
-)
-```
-
 We assessed whether the components of an algorithm can be predictive for method performance, using both random forest classification and statistical testing.
-Methods which include principal curves (which include Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+Methods which include principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
 On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
@@
 "method_components_scores", 
-"Score of methods containing particular algorithmic components"
+"Score of methods containing particular algorithmic components. NS: Not significant. *: "
 )
@@
 ```
+
 ## Method quality control
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, since good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
 We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
@@
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each such use case, a different set of methods performs optimally, with Monocle DDRTree performing best when the data contains a complex tree, while Slingshot performing better when the trajectory could as well be less complex.
-For the use case where the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT `r cite_methods("wadot")` and AGA `r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
+In each such use case, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contains a complex tree, and Slingshot performing better when the trajectory could as well be less complex.
+For the use case where the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA`r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
 If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
-Here, to our knowledge, only ElPiGraph.R (https://github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
+Here, to our knowledge, only ElPiGraph.R [https://github.com/Albluca/ElPiGraph.R](github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
 
@@
 ![](`r figure_file("tree.tmp.pdf", experiment_id="7-user_guidelines")`)
-`r ref("fig", "user_guidelines", anchor=TRUE)` Practical guidelines for method users.
+
+`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.**
 
 # Discussion
-<!-- Some intro -->
+<!-- The context -->
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
+Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
 It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
-In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
+In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters on real and synthetic data.
+
+<!-- The implications of the results -->
+Our evaluation indicated a large difference between current TI methods, with Slingshot, followed by TSCAN and Monocle DDRTree, towering above all other methods.
+Nonetheless, we found that methods which did not perform well across all settings, can still be useful in certain specific use cases.
+Indeed, on data containing more than one bifurcations, Monocle DDRTree clearly performed better than other methods.
+We found that this was mainly because the default parameters of Monocle DDRTree prefer are more likely to detect tree topologies, while those of Slingshot prefer to find linear and bifurcation trajectories.
 
@@
 We managed to wrap the output of all methods into one common format.
-This not only allowed us to compare different methods with a gold standard, but could also be useful for users of these tools because it allows the user to test multiple methods without manual conversion of input and output.
+This not only allowed us to compare different methods with a gold standard, but could also be useful for TI users because it allows the user to test multiple methods without manual conversion of input and output.
 Furthermore, it makes it possible to compare the output of different methods, which opens up possibilities for new comparative visualisation techniques or ensemble methods.
@@
 It can currently not take into account uncertainty of the cell's position.
-This can be both locally, for example uncertainty of its position within one branch, or globally, for example uncertainty over multiple branches.
+This can be both locally, for example uncertainty of its position within one branch, or globally, for example uncertainty over connections between branches.
 Some current methods already model this uncertainty in some way, mainly locally `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
+
+<!-- Synthetic data: observations and possibilities -->
+The use of synthetic data to evaluate methods can be problematic, because the model to generate the data is not necessarily similar to the way a true biological system works.
+This could bias an evaluation towards methods for which the underlying model best fits the model used for generating data.
+Therefore, it is essential that results of an evaluation on synthetic data are confirmed using real data.
+In our study, the overall performance of methods was very similar between real and synthetic datasets, confirming the biological relevance of the synthetic data.
+Given this, we believe our synthetic data can now be used to effectively prototype new TI methods, which use ever more complex models, such as disconnected graphs.
+Furthermore, it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
+Synthetic data based on our workflow could therefore be used to spearhead the development of these methods, given that currently only a limited number of real datasets are available for which the methods could be useful.
+Furthermore, we believe our evaluation workflow could also be used to evaluate other types of single-cell modelling techniques, such as single-cell network inference, clustering and normalisation.
+We sincerely hope that such efforts will lead to a more rapid development of accurate methods, and will in the near future provide a package which can be used to simulate synthetic data for a wide variety of single-cell modelling problems.
 
@@
 
-## Trajectory types
-
-We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
-These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph, progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
-For every type, a corresponding directed graph type can be defined by allowing directedness (`r ref("sfig", "trajectory_type_tree", "b")`).
-
-```{r}
-add_sfig(
-figure_file("trajectory_type_trees.tmp.pdf","2-dataset_characterisation/2-trajectory_types"), 
-"trajectory_type_tree", 
-"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
-15, 
-10
-)
-```
-
 ## Trajectory inference methods
 
-### Method wrapping
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis/issues).
+### Method selection
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynalysis/issues).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
 
+### Method input
+
+As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
+Furthermore, when required, we also provided a maximum of 7 types of prior information.
+This prior information was extracted from the gold/silver standards as follows:
+
+* **Start cells** The identity of one or more start cells.
+For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
+For ties, we chose one random cell.
+For cyclic datasets, a random cell was chosen.
+* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
+* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
+* **Grouping** For each cell a label to which state/cluster/branch it belongs.
+For real data, the states from the gold/silver standard.
+For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
+* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
+* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
+
+### Common output model
+
 Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with rarely any overlap.
-We therefore created wrappers around each method in R (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping")`).
+We therefore created wrappers around each method in R (available at [https://www.github.com/dynverse/dynmethods](github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping")`).
 This model consisted of two parts.
@@
 ```{r}
+
 # create functions to list the methods within a particular conversion category
@@
 
-### Prior information
-
-All evaluated methods required or could use one of 7 types of prior information. This prior information was extracted from the gold/silver standards as follows:
-
-* **Start cells** The identity of one or more start cells.
-For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
-For ties, we chose one random cell.
-For cyclic datasets, a random cell was chosen.
-* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
-* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
-* **Grouping** For each cell a label to which state/cluster/branch it belongs.
-For real data, the states from the gold/silver standard.
-For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
-* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
-* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
-
 ## Real datasets
 We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")` and `r ref("sfig", "dataset_characterisation")`).
-The scripts to download and process these datasets will be made available n our repository (http://www.github.com/dynverse/dynalysis).
+The scripts to download and process these datasets will be made available n our repository ([http://www.github.com/dynverse/dynalysis](github.com/dynverse/dynalysis)).
 When available, we preferred to start from the raw counts data.
@@
 ```
+
 ## Synthetic datasets
@@
 
-```{r}
-add_sfig(
-        figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
-"synthetic_data_generation",  "Synthetic data generation workflow"
-)
-```
 ### Network generation
@@
 We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module.
-To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1).
+To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.
 
@@
 
-## Benchmark
-
-### Pipeline
-* Look at dyneval methods to write this paragraph
-
-### Method execution
-* Look at dynmethods::execute_method to write this paragraph
-
-### Effect of prior information
-
-To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a
-
-### Effect of algorithm components
-To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score.
-To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test.
-We only investigated algorithm components which were part of at least 4 different methods in our evaluation study.
-
-## Method quality control
-
-```{r}
-qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
-  str_split("[, \n]+") %>% 
-  unlist() %>% 
-  unique() %>% 
-  discard(is.na) %>%
-  glue::collapse(., ";") %>%
-  paste0("[", ., "]")
-```
-
-We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
-The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
-The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
-The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
-The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
-The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
-The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
-The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
-Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
-
-Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
-This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
-For calculating the final score, we weighed each of the six categories equally.
-
 ## Evaluation metrics
-We used three different metrics to assess the performance of each TI method.
+We used three different metrics to assess the performance of each TI method (`r ref("fig", "evaluation_overview")`).
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
@@
 "score_aspects", 
-"Different scores assess different aspects of the ", 
+"Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
 15, 
@@
 "edge_flip_comparison", 
-"Edge flip score values for common trajectory topologies", 
+paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets")), 
 15, 
@@
 5
+)
+```
+
+## Benchmark
+
+### Pipeline
+* Look at dyneval methods to write this paragraph
+
+### Method execution
+We ran each method in a controlled environment, where we limited the memory usage (max.
+32GB) and total run time (max 24h).
+Moreover, we disabled the function of the R `set.seed` function, to avoid that methods false make their output deterministic.
+
+### Effect of prior information
+
+To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test.
+P-values were controlled for multiple testing using Benjamini-Hochberg correction.
+
+### Effect of algorithm components
+To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score.
+To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test.
+We only investigated algorithm components which were part of at least 4 different methods in our evaluation study.
+P-values were controlled for multiple testing using Benjamini-Hochberg correction.
+
+## Method quality control
+
+```{r}
+qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
+  str_split("[, \n]+") %>% 
+  unlist() %>% 
+  unique() %>% 
+  discard(is.na) %>%
+  glue::collapse(., ";") %>%
+  paste0("[", ., "]")
+```
+
+We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
+The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
+The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
+The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
+The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
+The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
+The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
+Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
+
+Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions), the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing), or indications that the tool will be broadly applicable on new datasets (such as being open-source, containing a good tutorial and support system, and being thoroughly evaluated in the study where it was published).
+
+Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
+Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("table", "implementations")`).
+For calculating the final score, we weighed each of the six categories equally.
+
+## Trajectory types
+
+We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
+These trajectory types start from the most general trajectory type, a disconnected directed graph, and moves down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
+For every directed trajectory type, a corresponding undirected trajectory type can also be defined (`r ref("sfig", "trajectory_type_tree", "b")`).
+In most cases, a method which can detect a complex trajectory type, was also be able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).
+
+```{r}
+add_sfig(
+figure_file("trajectory_type_trees.tmp.pdf","2-dataset_characterisation/2-trajectory_types"), 
+"trajectory_type_tree", 
+"Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types.", 
+15, 
+10
 )
@@
 This paragraph may be expanded
-Rephrase -_-
-mention timecourse prior
+Done (stolen from results section)
+🤔
+Add the timecourse prior to the figure
+LINK TO
+
+### Trajectory types and topology complexity
+LINK TO
+
+### Trajectory types and topology complexity
+It's already a bit better, though could be improved still!
+Expand, but maybe it should be moved to discussion anyway
 😉
-For V2/3: we should check what happens if wrong or noise prior information is given to the methods!
-I wish we were writing V2 instead of this temporary, half-decent version :p
-In V2, "randomise" the prior information so that the positive effect of giving it to a method can be directly assessed, instead of this indirect evidence
-TODO
-TODO
 to do: add metric figure
@@
 this is not used anymore
+TODO
+More to add here?
+* Look at dynmethods::execute_method to write this paragraph
+Are there more exectuion details?

### 2018-03-05 analysis/paper/paper.Rmd
avoid figures scaling like crazy

@@
 * Go to tools -> perferences and remove Use smart quotes
-* Add extra enters after every paragraph (this is markdown!)
 * Enable automatic zotero syncing to bibtex
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-When cells are profiled from a population in which cells are at different unknown points in the dynamic process, dynamic processes can be modelled using trajectory inference (TI) methods  [@cannoodt_computational_2016].
-These methods computationally order these cells along a trajectory topology, which can be linear, bifurcating, or more complex.
-TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017].
+Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), by profiling cells from a population in which the cells are at different unknown points in the dynamic process  [@cannoodt_computational_2016].
+These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex tree or graph structures.
+Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
@@
   "methods_over_time", 
-"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory."
+  "New TI methods over time. a) Number of methods published or in preprint. b) Number of methods which can handle a particular type of trajectory."
 )
@@
   read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-"Overview of current trajectory inference methods, and whether they were included in this current evaluation study"
+  "Overview of current trajectory inference methods, and whether they were included in this current evaluation study."
 )
@@
 Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
+
+```{r}
+add_sfig(
+  figure_file("implementation_components.svg","4-method_characterisation"), 
+  "implementation_components", 
+  "TI methods frequently share steps."
+)
+```
 
 <!-- §6: Problem statement -->
-An important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
-Many attempts at addressing this issue has already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, performing a comprehensive benchmarking evaluation of TI methods requires a lot of work.
+Given this diversity in TI methods, an important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
+Many attempts at addressing this issue has already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, performing a comprehensive benchmarking evaluation of TI methods is still lacking.
 This is problematic, as new users to the field are confronted with a wide array of TI methods without a clue as to what the most optimal method available for their problem could be. 
@@
   "dataset_characterisation",  
-"Characteristics of the real datasets used for the evaluation"
+  "Characteristics of the real datasets used for the evaluation."
 )
@@
 
-As part of the quality control evaluation, we assembled a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+```{r}
+add_sfig(
+  figure_file('synthetic_data_generation.svg', "2-dataset_characterisation/3-synthetic"),
+  "synthetic_data_generation",
+  "Synthetic data generation workflow."
+)
+```
 
-Finally, we present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+We not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation.
+To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
+Finally, using both the benchmark results and quality control, we finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
 
@@
 add_sfig(
-        {figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview')}, 
-"extended_evaluation_overview", "Extended overview ", 15, 7.5
+  figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview'),
+  "extended_evaluation_overview",
+  "Extended overview", 
+  15, 
+  7.5
 )
@@
 
-Overall, we find that Slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`).
-These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets.
-
-However, results are more variable when comparing the three different evaluation metrics. 
-Some methods, mainly those able to detect more complex trajectory types, have higher edge flip scores.
-Other methods, not necessarily those with the highest edge flip score, are better at preserving the ordering and neighbourhood of the cells.
-
 `r anchor("fig", "results_overview")`
 ![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
+
 `r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the maximum complexity trajectory they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally make use of.
-b) The
+b) The methods are ordered according to the overall score achieved in the benchmark evaluation.
+Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+c) Overall performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
+Also listed are the quality control scores aggregated according to practicality and the different categories.
+
+<!-- §2: Overall and topology types -->
+Ordering all methods by their overall benchmarking score, we found that Slingshot predicts the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
+Observing the benchmark scores per trajectory type, Slingshot seems to be the only method that performs well across most trajectory types.
+However, several methods are seemingly specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, Monocle DDRTree for rooted trees, and, surprisingly Wishbone for multifurcations.
+We will discuss the effect of topology types and constraints on inferrable trajectory types placed by the method in Section SEC_TRAJTYPE.
+
+<!-- §3: real versus synthetic -->
+We observed a high correlation (0.7-0.9) between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`).
+This confirms both the relevance of the synthetic data and the accuracy of the gold standard in the real datasets.
+However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available which contain such topologies (`r ref("stable", "datasets")`).
 
@@
 add_sfig(
-figure_file('real_synthetic_comparison.svg', "5-optimise_parameters/5-real_synthetic_comparison"), 
+  figure_file(
+    "real_synthetic_comparison.svg",
+    "5-optimise_parameters/5-real_synthetic_comparison"
+  ), 
   "real_synthetic_comparison", 
   "Comparison between performance on real and synthetic datasets across trajectory types."
+)
+```
+
+<!-- §4: Compare metrics -->
+As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics.
+While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`).
+In addition, methods able to detect more complex trajectory types seemed to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
+We will explore this issue in more detail in a further section SEC_TRAJTYPE.
+
+```{r}
+add_sfig(
+  figure_file('metrics_comparison.svg', "5-optimise_parameters/7-metric_comparison"), 
+  "metric_comparison", 
+  "Comparison between the different metrics across trajectory types."
+)
+```
+
+During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded (>6h) or an error was produced, a zero score would be returned for that execution.
+If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
+Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
+Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
+Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largests of datasets (reCAT, SCOUP, StemID).
+For a few methods, the time or memory limits were exceeded to frequently that insufficient results were generated to compare it to other methods (SCIMITAR, Ouija, Pseudogp).
+
+Comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
+We hope the identifications of the main strengths and weaknesses of existing TI methods will lead to the next few state-of-the-art improvements.
+
+```{r}
+add_sfig(
+  figure_file('leading_method.svg', "9-main_figure"), 
+  "leading_method", 
+  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
 )
@@
 These methods had typically better edge flip scores - as can be expected - and RF MSE scores compared to methods not able to handle the particular trajectory type.
-However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced a better ordering, irrespective of whether the dataset contained a linear trajectory or more complex
+However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
 To further investigate, we divided all methods in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
-While there were no significant difference on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "'linear_split_comparison")`).
-Together, this indicates that current non-linear methods contain less accurate ordering algorithms.
-Inclusion of the ordering algorithms from the top linear methods could therefore improve the performance of the non-linear methods.
+While there were no significant differences on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`).
+Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
+A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.
 
 ```{r}
-add_sfig({
-        figure_file('trajtype_handle_comparison.svg', "8-compare_topology")
-}, 
+add_sfig(
+  figure_file('trajtype_handle_comparison.svg', "8-compare_topology"), 
   "trajtype_handle_comparison", 
-""
+  "ADD CAPTION"
 )
@@
 ```{r}
-add_sfig({
-        figure_file('linear_split_comparison.svg', "8-compare_topology")
-}, 
+add_sfig(
+  figure_file('linear_split_comparison.svg', "8-compare_topology"), 
   "linear_split_comparison", 
   "Linear methods have a higher correlation score on non-linear datasets.
-Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
-"
+Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods."
 )
@@
 ![](`r figure_file("topology_complexity_comparison.svg","8-compare_topology")`)
-`r ref("fig", "topology_complexity_comparison")`
+
+`r ref("fig", "topology_complexity_comparison")` **INSERT CAPTION.** Caption subdetails.
 
 ```{r}
-add_sfig({
-        figure_file('topology_sensitivity.svg', "8-compare_topology")
-}, 
+add_sfig(
+  figure_file('topology_sensitivity.svg', "8-compare_topology"), 
   "topology_sensitivity", 
-""
+  "ADD CAPTION"
 )
 ```
+
 ### Effect of prior information
-
-Giving prior information to a TI method can be both a blessing and a curse.
-In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
-On the other hand, incorrect or noisy prior information can bias the trajectory towards a wrong, but perhaps consistent with current knowledge, trajectory.
-Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, limiting the applicability of such TI methods to only well studied systems.
 
@@
 ```{r}
-add_sfig({
-        figure_file('required_priors_performance_comparison.svg', "10-prior_information")
-}, 
+add_sfig(
+  figure_file('required_priors_performance_comparison.svg', "10-prior_information"), 
   "required_priors_performance_comparison", 
-""
+  "Comparison of method performance."
 )
@@
 
-### Running time and scalability
-
 ### Algorithm components
 
-The underlying algorithm of each method can be divided into several steps, and these frequently contain soms clustering, dimensionality reduction and/or graph building.
-Interestingly, steps are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-```{r}
-add_sfig(
-figure_file("implementation_components.svg","4-method_characterisation"), 
-"implementation_components", 
-"TI methods frequently share steps"
-)
-```
-
 We assessed whether the components of an algorithm can be predictive for method performance, using both random forest classification and statistical testing.
-Methods which include principal curves (which include Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+Methods which include principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
 On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
@@
   "method_components_scores", 
-"Score of methods containing particular algorithmic components"
+  "Score of methods containing particular algorithmic components. NS: Not significant. *: "
 )
@@
 add_stable(
-read_rds(figure_file("component_importance.rds","6-developer_guidelines"))[[params$table_format]], 
+  read_rds(figure_file(
+    "component_importance.rds",
+    "6-developer_guidelines"
+  ))[[params$table_format]], 
   "component_importance", 
-"Importance values for algorithmic components"
+  "Importance values for algorithmic components."
 )
 ```
+
 ## Method quality control
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, since good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
 We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
@@
 add_stable(
-read_rds(figure_file("qc_checks.rds","4-method_characterisation"))[[params$table_format]], 
+  read_rds(figure_file(
+    "qc_checks.rds",
+    "4-method_characterisation"
+  ))[[params$table_format]], 
   "qc_checks", 
-"Scoring scheme for method quality control"
+  "Scoring scheme for method quality control."
 )
@@
 ![](`r figure_file("implementation_qc_overview.svg","4-method_characterisation")`)
-`r ref("fig", "implementation_qc_overview")` Overview of the quality control results for every method.
+
+`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.**
 
@@
   "qc_over_time", 
-"Quality control scores and number of citations over time"
+  "Quality control scores and number of citations over time."
 )
@@
 add_sfig(
-figure_file("qc_performance_comparison.svg","5-optimise_parameters/6-qc_performance_comparison"), 
+  figure_file(
+    "qc_performance_comparison.svg",
+    "5-optimise_parameters/6-qc_performance_comparison"
+  ), 
   "qc_performance_comparison", 
@@
 ![](`r figure_file("tree.svg", experiment_id="7-user_guidelines")`)
-`r ref("fig", "user_guidelines", anchor=TRUE)` Practical guidelines for method users.
+
+`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.**
 
 # Discussion
-<!-- Some intro -->
+<!-- The context -->
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
+Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
 It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
-In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
+In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters on real and synthetic data.
+
+<!-- The implications of the results -->
+Our evaluation indicated a large difference between current TI methods, with Slingshot, followed by TSCAN and Monocle DDRTree, towering above all other methods.
+Nonetheless, we found that methods which did not perform well across all settings, can still be useful in certain specific use cases.
+Indeed, on data containing more than one bifurcations, Monocle DDRTree clearly performed better than other methods.
+We found that this was mainly because the default parameters of Monocle DDRTree prefer are more likely to detect tree topologies, while those of Slingshot prefer to find linear and bifurcation trajectories.
 
@@
 
-## Trajectory types
-
-We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
-These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph, progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
-For every type, a corresponding directed graph type can be defined by allowing directedness (`r ref("sfig", "trajectory_type_tree", "b")`).
-
-```{r}
-add_sfig(
-figure_file("trajectory_type_trees.svg","2-dataset_characterisation/2-trajectory_types"), 
-"trajectory_type_tree", 
-"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
-15, 
-10
-)
-```
-
 ## Trajectory inference methods
 
-### Method wrapping
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations", anchor=TRUE)`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis/issues).
+### Method selection
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
 
+### Method input
+As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
+Furthermore, when required, we also provided a maximum of 7 types of prior information.
+This prior information was extracted from the gold/silver standards as follows:
+
+* **Start cells** The identity of one or more start cells.
+For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
+For ties, one random cell was chosen.
+For cyclic datasets, a random cell was chosen.
+* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
+* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
+* **Grouping** For each cell a label to which state/cluster/branch it belongs.
+For real data, the states from the gold/silver standard.
+For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
+* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
+* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
+
+### Common output model
+
 Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with rarely any overlap.
-We therefore created wrappers around each method in R (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping")`).
+We therefore created wrappers around each method in R (available at [https://www.github.com/dynverse/dynmethods](github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping")`).
 This model consisted of two parts.
@@
 add_sfig(
-        {figure_file('wrapping.svg', '4-method_characterisation')},
-"wrapping",  "Our wrapping workflow", 15, 7.5
+  figure_file('wrapping.svg', '4-method_characterisation'),
+  "wrapping", 
+  "Our wrapping workflow.", 
+  15, 
+  7.5
 )
@@
 }
-
 ```
@@
   "datasets", 
-"Real datasets used in this study"
+  "Real datasets used in this study."
 )
 ```
+
 ## Synthetic datasets
@@
 
-```{r}
-add_sfig(
-        figure_file('synthetic_data_generation.svg', "2-dataset_characterisation/3-synthetic"),
-"synthetic_data_generation",  "Synthetic data generation workflow"
-)
-```
 ### Network generation
@@
 We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module.
-To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1).
+To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.
 
 ```{r}
-add_sfig({
-read_rds(figure_file("modulenets.rds", experiment_id="2-dataset_characterisation/3-synthetic"))
-}, 
+add_sfig(
+  read_rds(figure_file(
+    "modulenets.rds",
+     experiment_id = "2-dataset_characterisation/3-synthetic"
+   )), 
   "modulenets", 
@@
 add_stable(
-read_rds(figure_file("samplers.rds","2-dataset_characterisation/3-synthetic"))[[params$table_format]], 
+  read_rds(figure_file(
+    "samplers.rds",
+    "2-dataset_characterisation/3-synthetic")
+  )[[params$table_format]], 
   "samplers", 
-"Distributions from which each parameter in the thermodynamic model was sampled"
+  "Distributions from which each parameter in the thermodynamic model was sampled."
 )
@@
 Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$.
-Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernouilli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
+Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
 
@@
 
-Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward.
+Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward.
 Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable.
@@
 add_sfig({
-read_rds(figure_file("score_aspects.rds", experiment_id="3-score_rules/5-score_aspects"))
+  read_rds(figure_file(
+    "score_aspects.rds", 
+    experiment_id="3-score_rules/5-score_aspects"))
   }, 
   "score_aspects", 
-"Different scores assess different aspects of the ", 
+  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.",
   15, 
@@
 As a third independent score, we assessed the similarity between the milestone network topology.
-We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that A->A would be converted A->B->C->D, by adding an intermediate node to linear networks.
+We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks.
 Because we are interested , the network was made undirected.
@@
 ```{r}
-add_sfig({
-figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip")
-}, 
+add_sfig(
+  figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip"), 
   "edge_flip_comparison", 
-"Edge flip score values for common trajectory topologies", 
+  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets")), 
   15, 
@@
 ```{r}
-add_sfig({
-figure_file("metric_aggregation_overview_v1.svg", experiment_id="evaluation_overview")
-}, 
+add_sfig(
+  figure_file("metric_aggregation_overview_v1.svg", experiment_id="evaluation_overview"), 
   "aggregation_of_results", 
-"Aggregation methodology of metric scores", 
+  "Aggregation methodology of metric scores.", 
   15, 
   5
+)
+```
+
+## Benchmark
+
+### Pipeline
+<!-- Look at dyneval methods to write this paragraph -->
+
+### Method execution
+We ran each method in a controlled environment, where we limited the memory usage (max.
+32GB) and total run time (max 24h).
+Moreover, we disabled the function of the R `set.seed` function, to avoid that methods false make their output deterministic.
+
+### Effect of prior information
+
+To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test.
+P-values were controlled for multiple testing using Benjamini-Hochberg correction.
+
+### Effect of algorithm components
+To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score.
+To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test.
+We only investigated algorithm components which were part of at least 4 different methods in our evaluation study.
+P-values were controlled for multiple testing using Benjamini-Hochberg correction.
+
+## Method quality control
+
+```{r}
+qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation")) %>% 
+  .$references %>% 
+  str_split("[, \n]+") %>% 
+  unlist() %>% 
+  unique() %>% 
+  discard(is.na) %>%
+  glue::collapse(., ";") %>%
+  paste0("[", ., "]")
+```
+
+We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
+The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
+The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
+The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
+The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system.
+The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
+The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
+Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
+
+Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions), the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing), or indications that the tool will be broadly applicable on new datasets (such as being open-source, containing a good tutorial and support system, and being thoroughly evaluated in the study where it was published).
+
+Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
+Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("table", "implementations")`).
+For calculating the final score, we weighed each of the six categories equally.
+
+## Trajectory types
+
+We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
+These trajectory types start from the most general trajectory type, a disconnected directed graph, and moves down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
+For every directed trajectory type, a corresponding undirected trajectory type can also be defined (`r ref("sfig", "trajectory_type_tree", "b")`).
+In most cases, a method which can detect a complex trajectory type, was also be able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).
+
+```{r}
+add_sfig(
+  figure_file("trajectory_type_trees.svg","2-dataset_characterisation/2-trajectory_types"), 
+  "trajectory_type_tree", 
+  "Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types.", 
+  15, 
+  10
 )
@@
 This paragraph may be expanded
-Rephrase -_-
-mention timecourse prior
+Done (stolen from results section)
+🤔
+Add the timecourse prior to the figure
+LINK TO
+
+### Trajectory types and topology complexity
+LINK TO
+
+### Trajectory types and topology complexity
+It's already a bit better, though could be improved still!
+Expand, but maybe it should be moved to discussion anyway
+ADD CAPTION
 😉
-For V2/3: we should check what happens if wrong or noise prior information is given to the methods!
-I wish we were writing V2 instead of this temporary, half-decent version :p
-In V2, "randomise" the prior information so that the positive effect of giving it to a method can be directly assessed, instead of this indirect evidence
 TODO
-TODO
+ADD CAPTION
+FILL IN SIGNIFICANCE LEVELS
 to do: add metric figure

### 2018-03-05 analysis/paper/paper.Rmd
update to figures

@@
 ### Method selection
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`, als), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We welcome any contributions by creating an issue at [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.

### 2018-03-05 analysis/paper/paper_latex.Rmd
update to figures

@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-When cells are profiled from a population in which cells are at different unknown points in the dynamic process, this process  processes can be modelled using trajectory inference (TI) methods  [@cannoodt_computational_2016].
-These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex.
-TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017].
+Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), by profiling cells from a population in which the cells are at different unknown points in the dynamic process  [@cannoodt_computational_2016].
+These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex tree or graph structures.
+Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
@@
   "implementation_components", 
-"TI methods frequently share steps"
+  "TI methods frequently share steps."
 )
@@
   figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
-"synthetic_data_generation",  "Synthetic data generation workflow"
+  "synthetic_data_generation",
+  "Synthetic data generation workflow."
 )
@@
 We not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation.
-We scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
 This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
-
-Using both the benchmark and quality control results, we will finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+Finally, using both the benchmark results and quality control, we finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
 
@@
 add_sfig(
-        {figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview')}, 
-"extended_evaluation_overview", "Extended overview ", 15, 7.5
+  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
+  "extended_evaluation_overview",
+  "Extended overview", 
+  15, 
+  7.5
 )
@@
 <!-- §3: real versus synthetic -->
-A high correlation can be observed between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`).
-This confirms both the relevance of the synthetic data and the accuracy of the real datasets.
+We observed a high correlation (0.7-0.9) between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`).
+This confirms both the relevance of the synthetic data and the accuracy of the gold standard in the real datasets.
+However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available which contain such topologies (`r ref("stable", "datasets")`).
 
@@
 add_sfig(
-figure_file('real_synthetic_comparison.tmp.pdf', "5-optimise_parameters/5-real_synthetic_comparison"), 
+  figure_file(
+    "real_synthetic_comparison.tmp.pdf",
+    "5-optimise_parameters/5-real_synthetic_comparison"
+  ), 
   "real_synthetic_comparison", 
@@
 <!-- §4: Compare metrics -->
-As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is unsurprising to observe differences in rankings of TI methods amongst the different metrics.
-While there is no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also seed to obtain a higher RF MSE score (`r ref("sfig", "metric_comparison")`).
-In addition, methods able to detect more complex trajectory types seem to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
-This issue will also be investigated in more detail in Section SEC_TRAJTYPE.
+As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics.
+While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`).
+In addition, methods able to detect more complex trajectory types seemed to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
+We will explore this issue in more detail in a further section SEC_TRAJTYPE.
 
@@
 ```{r}
-add_sfig({
-        figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology")
-}, 
+add_sfig(
+  figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology"), 
   "trajtype_handle_comparison", 
-""
+  "ADD CAPTION"
 )
@@
 ```{r}
-add_sfig({
-        figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology")
-}, 
+add_sfig(
+  figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology"), 
   "linear_split_comparison", 
   "Linear methods have a higher correlation score on non-linear datasets.
-Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
-"
+Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods."
 )
@@
 ![](`r figure_file("topology_complexity_comparison.tmp.pdf","8-compare_topology")`)
-`r ref("fig", "topology_complexity_comparison")`
+
+`r ref("fig", "topology_complexity_comparison")` **INSERT CAPTION.** Caption subdetails.
 
 ```{r}
-add_sfig({
-        figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology")
-}, 
+add_sfig(
+  figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology"), 
   "topology_sensitivity", 
-""
+  "ADD CAPTION"
 )
@@
 ```{r}
-add_sfig({
-        figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information")
-}, 
+add_sfig(
+  figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information"), 
   "required_priors_performance_comparison", 
-"Comparison of method performance "
+  "Comparison of method performance."
 )
@@
 add_stable(
-read_rds(figure_file("component_importance.rds","6-developer_guidelines"))[[params$table_format]], 
+  read_rds(figure_file(
+    "component_importance.rds",
+    "6-developer_guidelines"
+  ))[[params$table_format]], 
   "component_importance", 
-"Importance values for algorithmic components"
+  "Importance values for algorithmic components."
 )
@@
 ![](`r figure_file("implementation_qc_overview.tmp.pdf","4-method_characterisation")`)
-`r ref("fig", "implementation_qc_overview")` Overview of the quality control results for every method.
+
+`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.**
 
@@
 add_sfig(
-figure_file("qc_performance_comparison.tmp.pdf","5-optimise_parameters/6-qc_performance_comparison"), 
+  figure_file(
+    "qc_performance_comparison.tmp.pdf",
+    "5-optimise_parameters/6-qc_performance_comparison"
+  ), 
   "qc_performance_comparison", 
@@
 We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynalysis/issues).
+We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues).
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
 ### Method input
-
 As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
@@
 For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
-For ties, we chose one random cell.
+For ties, one random cell was chosen.
 For cyclic datasets, a random cell was chosen.
@@
 add_sfig(
-        {figure_file('wrapping.tmp.pdf', '4-method_characterisation')},
-"wrapping",  "Our wrapping workflow", 15, 7.5
+  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
+  "wrapping", 
+  "Our wrapping workflow.", 
+  15, 
+  7.5
 )
@@
   "datasets", 
-"Real datasets used in this study"
+  "Real datasets used in this study."
 )
@@
 ```{r}
-add_sfig({
-read_rds(figure_file("modulenets.rds", experiment_id="2-dataset_characterisation/3-synthetic"))
-}, 
+add_sfig(
+  read_rds(figure_file(
+    "modulenets.rds",
+     experiment_id = "2-dataset_characterisation/3-synthetic"
+   )), 
   "modulenets", 
@@
 add_sfig({
-read_rds(figure_file("score_aspects.rds", experiment_id="3-score_rules/5-score_aspects"))
+  read_rds(figure_file(
+    "score_aspects.rds", 
+    experiment_id="3-score_rules/5-score_aspects"))
   }, 
@@
 ```{r}
-add_sfig({
-figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip")
-}, 
+add_sfig(
+  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip"), 
   "edge_flip_comparison", 
@@
 ```{r}
-add_sfig({
-figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview")
-}, 
+add_sfig(
+  figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview"), 
   "aggregation_of_results", 
-"Aggregation methodology of metric scores", 
+  "Aggregation methodology of metric scores.", 
   15, 
@@
 ### Pipeline
-* Look at dyneval methods to write this paragraph
+<!-- Look at dyneval methods to write this paragraph -->
 
@@
 ```{r}
-qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
+qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation")) %>% 
+  .$references %>% 
   str_split("[, \n]+") %>% 
@@
 Expand, but maybe it should be moved to discussion anyway
+ADD CAPTION
 😉
+TODO
+ADD CAPTION
+FILL IN SIGNIFICANCE LEVELS
 to do: add metric figure

### 2018-03-05 analysis/paper/paper.Rmd
update rmd

@@
 ³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
-⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
-⁵ Université Claude Bernard Lyon 1, Lyon, France.
+⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, LYON, France
 
@@
 
-Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
+Using single-cell transcriptomics or proteomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
 Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-As a result, comparing between trajectory inference methods is often challenging, and a comprehensive assessment of the performance and robustness of each method is still lacking.
-We developed a common data structure for trajectory inference methods in order to make their results comparable to each other and to a gold standard.
-Using this framework, we compare the results from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets.
-We employ several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We conclude with practical guidelines for users to select appropriate methods for a given use case.
-We believe our evaluation framework can be used to spearhead the development for new methods.
-Our evaluation pipeline is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+As a result, choosing a method to infer trajectories is often challenging, since a comprehensive assessment of the performance and robustness of each method is still lacking.
+In order to facilitate the comparison of the results of these methods to each other and to a gold standard, we developed a global framework to benchmark trajectory inference tools.
+.
+Using this framework, we compared the results from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real and synthetic datasets.
+We evaluate methods using several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
 
-We believe this to be the first comprehensive assessment of trajectory inference methods.
-For now, we only evaluated the methods on default parameters, but plan to add a detailed parameter tuning procedure in the future.
+To our knowledge, this is the first comprehensive assessment of trajectory inference methods.
+For now, we exclusively evaluated the methods on their default parameters, but plan to add a detailed parameter tuning procedure in the future.
 We gladly welcome any discussion and feedback on key decisions made as part of this study, including the metrics used in the benchmark, the quality control checklist, and the implementation of the method wrappers.
@@
 <!-- §1: Single-cell omics and TI --->
-Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before.
+Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017].
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), by profiling cells from a population in which the cells are at different unknown points in the dynamic process  [@cannoodt_computational_2016].
+Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), which use single-cell profiles from a population in which the cells are at different unknown points in the dynamic process  [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
 These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex tree or graph structures.
-Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
+Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
@@
   read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  "Overview of current trajectory inference methods, and whether they were included in this current evaluation study."
+  "Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated at https://goo.gl/FnsTxG."
 )
@@
 Early TI methods typically fixed the topology algorithmically (e.g.
-linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]) or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
-The focus of these methods was therefore mainly to correctly order the cells along this fixed topology.
-Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows a method to be more broadly applicable on more use cases.
+linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
+These methods therefore mainly focused on correctly ordering the cells along this fixed topology.
+Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods  to be broadly applicable on more use cases.
 Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).
-Most noticeable of these methods is AGA[@wolf_graphabstractionreconciles_2017], which is the only TI method currently able to predict disconnected graphs.
+A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is the only TI method currently able to deal with disconnected graphs.
 
 <!-- §4: Prior information -->
-Another key characteristic is the selection of prior information that a method requires or can optionally exploit.
-Examples of prior informations are supplying a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
-Giving prior information to a TI method can be both a blessing and a curse.
+Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
+Prior information can be supplied as  a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
+Providing prior information to a TI method can be both a blessing and a curse.
 In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
 On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
-Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, limiting the applicability of such TI methods to only well studied systems.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.
 
 <!-- §5: Components -->
-A reductionist approach to characterising TI methods is dissecting it into a set of algorithmic components, as any one component can have a significant impact on the performance, scalability, and output data structures.
+A reductionist approach to characterising TI methods consists in dissecting them into a set of algorithmic components, as any component can have a significant impact on the performance, scalability, and output data structures.
 Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
 Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-MST in particular, used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is used in almost half of all methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
 
@@
 <!-- §6: Problem statement -->
-Given this diversity in TI methods, an important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
-Many attempts at addressing this issue has already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, performing a comprehensive benchmarking evaluation of TI methods is still lacking.
-This is problematic, as new users to the field are confronted with a wide array of TI methods without a clue as to what the most optimal method available for their problem could be. 
-Moreover, the strengths and weakness of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
+Given the  diversity in TI methods, an important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
+Many attempts at tackling  this issue have already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, a comprehensive benchmarking evaluation of TI methods is still lacking.
+This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would be the most optimal for their problem. 
+Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
 
@@
 In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (Overview: `r ref("fig", "evaluation_overview", "a")`, Extended overview: `r ref("sfig", "extended_evaluation_overview")`).
-The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`).
-Currently, only methods published before june 2017 are included in the current version of the evaluation, while more recent methods will be included in the next version.
-The evaluation comprises of three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) the assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
+The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the date of publication  (`r ref("table", "implementations")`).
+Only methods published before june 2017 are included in the current version of the evaluation, while more recent methods will be added in the next version.
+The evaluation comprised three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
 
 In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
-In this model the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
 We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
@@
 
-We not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation.
+As the aim of this study is to provide both guidelines for the use of TI methods and a base for the development of new TI methods, we not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation.
 To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
 This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
-Finally, using both the benchmark results and quality control, we finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+Finally, using both the benchmark results and quality control, we  produced a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
 
@@
 c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
+
+----
 
@@
 
+-------
+
 <!-- §2: Overall and topology types -->
-Ordering all methods by their overall benchmarking score, we found that Slingshot predicts the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
-Observing the benchmark scores per trajectory type, Slingshot seems to be the only method that performs well across most trajectory types.
-However, several methods are seemingly specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, Monocle DDRTree for rooted trees, and, surprisingly Wishbone for multifurcations.
+Having ordered all methods by their overall benchmarking score, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
+When we looked at the benchmark scores per trajectory type, Slingshot was to be the only method that performs well across most trajectory types.
+However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, Monocle DDRTree for rooted trees, and Wishbone for multifurcations.
+This last result is unexpected, as Wishbone is not able to output any multifurcating topologies.
 We will discuss the effect of topology types and constraints on inferrable trajectory types placed by the method in Section SEC_TRAJTYPE.
@@
 This confirms both the relevance of the synthetic data and the accuracy of the gold standard in the real datasets.
-However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available which contain such topologies (`r ref("stable", "datasets")`).
+However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available for such topologies (`r ref("stable", "datasets")`).
 
@@
 As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics.
-While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`).
-In addition, methods able to detect more complex trajectory types seemed to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
+While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtained a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`).
+In addition, the methods that were able to detect more complex trajectory types also obtained higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
 We will explore this issue in more detail in a further section SEC_TRAJTYPE.
@@
 
-During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded (>6h) or an error was produced, a zero score would be returned for that execution.
+During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
 If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
@@
 Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largests of datasets (reCAT, SCOUP, StemID).
-For a few methods, the time or memory limits were exceeded to frequently that insufficient results were generated to compare it to other methods (SCIMITAR, Ouija, Pseudogp).
-
-Comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
-We hope the identifications of the main strengths and weaknesses of existing TI methods will lead to the next few state-of-the-art improvements.
-
-```{r}
-add_sfig(
-  figure_file('leading_method.svg', "9-main_figure"), 
-  "leading_method", 
-  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
-)
-```
+Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
+For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
 
@@
 
-In most cases, methods which can handle a particular trajectory type also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
-These methods had typically better edge flip scores - as can be expected - and RF MSE scores compared to methods not able to handle the particular trajectory type.
+In most cases, the methods which were specifically designed to  handle a particular trajectory type, also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+These methods had typically better edge flip scores - as can be expected - and RF MSE scores, compared to methods not able to handle the particular trajectory type.
 However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-To further investigate, we divided all methods in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
-While there were no significant differences on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`).
+To further investigate the effect of trajectory complexity on performance, we divided them in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
+While there were no significant differences in performance on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`).
 Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
@@
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree for example, frequently predicted a tree topology, even when only cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
-Trajectories generated by Monocle DDRTree (at default parameters) tend to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an edge on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
+Trajectories detected using the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree, for example frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories generated by Monocle DDRTree (at default parameters) tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
 Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
 Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
-Therefore, inferring the correct topology without parameter tuning is still an open challenge.
-Conversely, when the data contains a complex trajectory structure, TI currently will still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
+Inferring the correct topology without parameter tuning therefore still is an open challenge.
+Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
 
@@
 `r ref("fig", "topology_complexity_comparison")` **INSERT CAPTION.** Caption subdetails.
+
+----
 
@@
 
-In the current version of the evaluation, we only provided prior information when it was required by a method.
+In the current version of the evaluation, we only provided prior information when a method required it .
 We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`).
-Rather, methods who received prior information were on average positioned in the middle of the ranking.
+Rather, methods which received prior information were on average positioned in the middle of the ranking.
 Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.
@@
 We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
-The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
+The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, method behaviour and the depth by which the method was presented in its study) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
 These categorisations can help current developers to improve their tool, and guide the selection of users and developers to use these tools for their purpose.
@@
 `r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.**
+
+----
 
@@
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
-Here, to our knowledge, only ElPiGraph.R [https://github.com/Albluca/ElPiGraph.R](github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
+Here, to our knowledge, only ElPiGraph.R [github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
 
@@
 
+----
+
 # Discussion
@@
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
+Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
 It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
 In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters on real and synthetic data.
+Comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
+We believe that the benchmarking presented in this study will pave the way to the next few state-of-the-art improvements.
 
-<!-- The implications of the results -->
+```{r}
+add_sfig(
+  figure_file('leading_method.svg', "9-main_figure"), 
+  "leading_method", 
+  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
+)
+```
+
+<!-- Future work -->
+In this study, we presented our first version of evaluation of TI methods.
+We are convinced that these results are already useful for both method users and tool developers, as we provide clear practical guidelines for users depending on their current knowledge of the trajectory, as well as an objective benchmark on which new methods can be tested.
+Nonetheless, our evaluation can be expanded on several points, all of which we will try to tackle in the near future:
+
+* Inclusion of more recent methods, especially those which can detect more complex trajectory types such as general graphs `r cite_methods(c("wadot", "calista"))` and disconnected graphs `r cite_methods("aga")`.
+* A parameter tuning for each method, both across all trajectory types, as well as on specific trajectory types.
+* A test of robustness on noise, parameter changes and dataset size
+* An evaluation of the stability of method's results when running the same parameter setting on the same dataset multiple times
+* Evaluate the methods when given optional priors, and compare with performance without priors
+* Assess the effect of noisy or incorrect prior information
+* Test methods on datasets with no clear trajectory present
+* Include feedback on the quality control scoring scheme, and update the scoring when method get updated
+* Test the scalability of each method, both in the number of cells and the number of features
+* Include more real datasets with complex trajectory types, as they become available
+* Evaluate methods on other single-cell -omics datasets, such as proteomics and epigenomics data
+
+<!-- Are certain methods better than others? -->
 Our evaluation indicated a large difference between current TI methods, with Slingshot, followed by TSCAN and Monocle DDRTree, towering above all other methods.
@@
 ### Method selection
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`, als), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We welcome any contributions by creating an issue at [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
+We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynalysis/issues).
+
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
 
-### Common output model
-
-Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with rarely any overlap.
-We therefore created wrappers around each method in R (available at [https://www.github.com/dynverse/dynmethods](github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping")`).
-This model consisted of two parts.
-The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
-The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.
+### Common trajectory model
+Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap.
+We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping", "a")`).
+This model consisted of three parts.
+(i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
+(ii) The milestone percentages contain for each cell its position between milestones, and sums for each cell to one.
+(iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones.
+Regions must be explicitly defined in the trajectory model.
+Per region, one milestone must be directly connected to all other milestones in the network.
 
@@
   "wrapping", 
-  "Our wrapping workflow.", 
+  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the five wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
   15, 
@@
 
-Depending on the output of a method, we used different strategies to convert the output to our model (in some cases combined with a special conversion denoted by a \*):
+Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "wrapping", "b")`). Special conversions are denoted by an \*.
 
@@
 
-* Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
-* Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`.
-Similarly, the branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
-* Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`.
+* **Type 1, direct:** Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`.
+The branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
+* **Type 2, linear pseudotime:** Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
+* **Type 3, end state probability:** Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`.
 We use a single start state and add an edge to every end milestone each representing an end state.
 The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
-* Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`.
+* **Type 4, cluster assignment:** Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`.
 The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
-* Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`.
+* **Type 5, projection onto nearest branch:** Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`.
 We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction.
@@
 We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")` and `r ref("sfig", "dataset_characterisation")`).
-The scripts to download and process these datasets will be made available n our repository ([http://www.github.com/dynverse/dynalysis](github.com/dynverse/dynalysis)).
+The scripts to download and process these datasets will be made available n our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
 When available, we preferred to start from the raw counts data.
@@
 ```
-This paragraph may be expanded
-Done (stolen from results section)
-🤔
+I would avoid making "absolute" statements like this one.
 Add the timecourse prior to the figure
+Changed the tenses here to past (results usually only include past tense, except when referring to figures or later in the text)
+why surprisingly?
+because wishbone is only able to detect bifurcations. Should mention this in the text :)
+Added an explanation in the following sentence
+But still more should be added, I think this simply indicates that none of the methods are performing well on multifurcating datasets.
+But we should show this, eg.
+looking at the raw performance on multifurcating datasets versus other datasets.
+If we don't have the time to show this, then we should not mention wishbone here
 LINK TO
@@
 ### Trajectory types and topology complexity
-It's already a bit better, though could be improved still!
-Expand, but maybe it should be moved to discussion anyway
 ADD CAPTION
-😉
 TODO
@@
 FILL IN SIGNIFICANCE LEVELS
+? Unclear what you mean by this.
+It's already a bit better, though could be improved still!
 to do: add metric figure

### 2018-03-05 analysis/paper/paper.Rmd
update paper & references

@@
 ```
-`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³")), ", ")`, Helena Todorov¹ ² ⁴ ⁵, Yvan Saeys¹ ²
+`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³")), ", ")`, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
 
@@
 Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
-Inferring the correct topology without parameter tuning therefore still is an open challenge.
+Inferring the correct topology without parameter tuning, is therefore still an open challenge.
 Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
@@
 
-We assessed whether the components of an algorithm can be predictive for method performance, using both random forest classification and statistical testing.
-Methods which include principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+We assessed whether the components of an algorithm could  be predictive of a method’s performance, using both random forest classification and statistical testing.
+Methods which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
 On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
@@
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, since good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
-We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
-The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, method behaviour and the depth by which the method was presented in its study) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric.
+Good unit testing assures that  the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+We therefore looked at the implementation of each method, and assessed its  quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
+The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, method's behaviour at runtime and the depth by which the method was presented in its study) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
 These categorisations can help current developers to improve their tool, and guide the selection of users and developers to use these tools for their purpose.
@@
 Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
-As method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
+As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each such use case, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contains a complex tree, and Slingshot performing better when the trajectory could as well be less complex.
-For the use case where the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA`r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
-If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and Slingshot performing equally well on less complex trajectories .
+No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
+In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although Slingshot could return other topologies if it would fit the data more accurately.
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
@@
 Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
-In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters on real and synthetic data.
-Comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
-We believe that the benchmarking presented in this study will pave the way to the next few state-of-the-art improvements.
+It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
+In this preprint, we present an  initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters.
+When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
+We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
 
@@
 <!-- Future work -->
-In this study, we presented our first version of evaluation of TI methods.
-We are convinced that these results are already useful for both method users and tool developers, as we provide clear practical guidelines for users depending on their current knowledge of the trajectory, as well as an objective benchmark on which new methods can be tested.
+In this study, we presented our first version of the evaluation of TI methods.
+We are convinced that the results we provided will be useful for both method users and tool developers, as we provide clear practical guidelines for users depending on their current knowledge of the trajectory, as well as an objective benchmark on which new methods can be tested.
 Nonetheless, our evaluation can be expanded on several points, all of which we will try to tackle in the near future:
 
-* Inclusion of more recent methods, especially those which can detect more complex trajectory types such as general graphs `r cite_methods(c("wadot", "calista"))` and disconnected graphs `r cite_methods("aga")`.
+* Inclusion of more recent methods
 * A parameter tuning for each method, both across all trajectory types, as well as on specific trajectory types.
 * A test of robustness on noise, parameter changes and dataset size
-* An evaluation of the stability of method's results when running the same parameter setting on the same dataset multiple times
+* An evaluation of the stability of a method's results when running the same parameter setting on the same dataset multiple times
 * Evaluate the methods when given optional priors, and compare with performance without priors
@@
 <!-- Are certain methods better than others? -->
-Our evaluation indicated a large difference between current TI methods, with Slingshot, followed by TSCAN and Monocle DDRTree, towering above all other methods.
-Nonetheless, we found that methods which did not perform well across all settings, can still be useful in certain specific use cases.
+Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, TSCAN, and Monocle DDRTree, towering above all other methods.
+Nonetheless, we found that methods which did not perform well across all settings could still be useful in certain specific use cases.
 Indeed, on data containing more than one bifurcations, Monocle DDRTree clearly performed better than other methods.
-We found that this was mainly because the default parameters of Monocle DDRTree prefer are more likely to detect tree topologies, while those of Slingshot prefer to find linear and bifurcation trajectories.
+We found that this particular result was mainly caused by the fact that the default parameters of Monocle DDRTree preferably led to the detection of tree topologies, while those of Slingshot preferably found linear and bifurcation trajectories.
 
@@
 We managed to wrap the output of all methods into one common format.
-This not only allowed us to compare different methods with a gold standard, but could also be useful for TI users because it allows the user to test multiple methods without manual conversion of input and output.
-Furthermore, it makes it possible to compare the output of different methods, which opens up possibilities for new comparative visualisation techniques or ensemble methods.
+This not only allowed us to compare different methods with a gold standard, but could also be useful for TI users, as  it allows the user to test multiple methods on the same data and compare the results without manual conversion of input and output.
+Furthermore, it makes it possible to directly compare the output of different methods, which opens up possibilities for new comparative visualisation techniques or ensemble methods.
 However, we acknowledge that our model has some limitations.
-It can currently not take into account uncertainty of the cell's position.
-This can be both locally, for example uncertainty of its position within one branch, or globally, for example uncertainty over connections between branches.
-Some current methods already model this uncertainty in some way, mainly locally `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
+Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
+when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
+when the connections between branches are uncertain).
+Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
 
@@
 As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the five wrapper functions to transform its output to the common trajectory model.
+After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
 This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
@@
 
-* **Type 1, direct:** Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`.
+* **Type 1, direct:** `r conversion("trajectory")`.
+These methods assign each cell to a branch together with a pseudotime value and a branch network.
 The branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
-* **Type 2, linear pseudotime:** Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
-* **Type 3, end state probability:** Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`.
+* **Type 2, linear pseudotime:** `r conversion("linear")`. These methods return a single pseudotime value. A single edge between two milestones, its percentage proportional to the pseudotime.
+* **Type 3, end state probability:** `r conversion("end_state_probability")`.
+These methods return a global pseudotime value and a probability for every end state.
 We use a single start state and add an edge to every end milestone each representing an end state.
 The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
-* **Type 4, cluster assignment:** Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`.
+* **Type 4, cluster assignment:** `r conversion("cluster_graph")`.
+These methods return a cluster assignment and a cluster network.
 The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
-* **Type 5, projection onto nearest branch:** Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`.
+* **Type 5, projection onto nearest branch:** `r conversion("dimred_projection")`.
+These methods returning a cluster assignment, cluster network and dimensionality reduction.
 We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction.
 This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.
+* **Type 6, cell graph:** `r conversion("cell_graph")`. These methods return a network of connected cells, and determine which cells are part of the "backbone". Cell
 
@@
 This was then processed as described above.
-* **SLICER** Although this method returns a branch network and a global pseudotime, the method did not return a position of each cell on the individual branches.
-We therefore used the cell graph provided by the method to calculate the shortest paths between the start and end cells found by the method, and then used this graph as the backbone for all other cells, creating a local pseudotime for every cell within its branch.
-The output was then further processed as above.
 * **SLICE** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
@@
 I would avoid making "absolute" statements like this one.
+maybe add something like "which we believe to be the largest list of TI methods to be compared in a single study"?
+Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
 Add the timecourse prior to the figure
-Changed the tenses here to past (results usually only include past tense, except when referring to figures or later in the text)
 why surprisingly?
@@
 FILL IN SIGNIFICANCE LEVELS
-? Unclear what you mean by this.
 It's already a bit better, though could be improved still!
+do you talk about the synthetic data or the workflow?
+Because I'm not sure if the wrappers, comparison metrics..
+could be used as they are to compare NI, clustering or normalisation methods.
+Or if you are convinced of this statement, then I would add one or two sentences to explain how it could be possible.
 to do: add metric figure

### 2018-03-05 analysis/paper/paper_latex.Rmd

@@
 Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largests of datasets (reCAT, SCOUP, StemID).
-For a few methods, the time or memory limits were exceeded to frequently that insufficient results were generated to compare it to other methods (SCIMITAR, Ouija, Pseudogp).
-
-Comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
-We hope the identifications of the main strengths and weaknesses of existing TI methods will lead to the next few state-of-the-art improvements.
-
-```{r}
-add_sfig(
-  figure_file('leading_method.tmp.pdf', "9-main_figure"), 
-  "leading_method", 
-  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
-)
-```
+Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
+For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
 
@@
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree for example, frequently predicted a tree topology, even when only cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
-Trajectories generated by Monocle DDRTree (at default parameters) tend to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an edge on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
+Trajectories detected using the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree, for example frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories generated by Monocle DDRTree (at default parameters) tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
 Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
 Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
-Therefore, inferring the correct topology without parameter tuning is still an open challenge.
-Conversely, when the data contains a complex trajectory structure, TI currently will still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
+Inferring the correct topology without parameter tuning, is therefore still an open challenge.
+Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
 
@@
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric, since good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
-We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
-The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric.
+Good unit testing assures that  the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+We therefore looked at the implementation of each method, and assessed its  quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
+The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, method's behaviour at runtime and the depth by which the method was presented in its study) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
 These categorisations can help current developers to improve their tool, and guide the selection of users and developers to use these tools for their purpose.
@@
 Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
-As method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
+As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each such use case, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contains a complex tree, and Slingshot performing better when the trajectory could as well be less complex.
-For the use case where the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA`r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
-If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data.
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and Slingshot performing equally well on less complex trajectories .
+No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
+In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although Slingshot could return other topologies if it would fit the data more accurately.
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
-Here, to our knowledge, only ElPiGraph.R [https://github.com/Albluca/ElPiGraph.R](github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
+Here, to our knowledge, only ElPiGraph.R [github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
 
@@
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017] .
-It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice.
-In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters on real and synthetic data.
+Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
+It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
+In this preprint, we present an  initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters.
+When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
+We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
 
-<!-- The implications of the results -->
-Our evaluation indicated a large difference between current TI methods, with Slingshot, followed by TSCAN and Monocle DDRTree, towering above all other methods.
-Nonetheless, we found that methods which did not perform well across all settings, can still be useful in certain specific use cases.
+```{r}
+add_sfig(
+  figure_file('leading_method.tmp.pdf', "9-main_figure"), 
+  "leading_method", 
+  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
+)
+```
+
+<!-- Future work -->
+In this study, we presented our first version of the evaluation of TI methods.
+We are convinced that the results we provided will be useful for both method users and tool developers, as we provide clear practical guidelines for users depending on their current knowledge of the trajectory, as well as an objective benchmark on which new methods can be tested.
+Nonetheless, our evaluation can be expanded on several points, all of which we will try to tackle in the near future:
+
+* Inclusion of more recent methods
+* A parameter tuning for each method, both across all trajectory types, as well as on specific trajectory types.
+* A test of robustness on noise, parameter changes and dataset size
+* An evaluation of the stability of a method's results when running the same parameter setting on the same dataset multiple times
+* Evaluate the methods when given optional priors, and compare with performance without priors
+* Assess the effect of noisy or incorrect prior information
+* Test methods on datasets with no clear trajectory present
+* Include feedback on the quality control scoring scheme, and update the scoring when method get updated
+* Test the scalability of each method, both in the number of cells and the number of features
+* Include more real datasets with complex trajectory types, as they become available
+* Evaluate methods on other single-cell -omics datasets, such as proteomics and epigenomics data
+
+<!-- Are certain methods better than others? -->
+Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, TSCAN, and Monocle DDRTree, towering above all other methods.
+Nonetheless, we found that methods which did not perform well across all settings could still be useful in certain specific use cases.
 Indeed, on data containing more than one bifurcations, Monocle DDRTree clearly performed better than other methods.
-We found that this was mainly because the default parameters of Monocle DDRTree prefer are more likely to detect tree topologies, while those of Slingshot prefer to find linear and bifurcation trajectories.
+We found that this particular result was mainly caused by the fact that the default parameters of Monocle DDRTree preferably led to the detection of tree topologies, while those of Slingshot preferably found linear and bifurcation trajectories.
 
@@
 ### Method selection
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues).
+We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
+We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynalysis/issues).
+
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
@@
   "wrapping", 
-  "Our wrapping workflow.", 
+  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
   15, 
@@
 
-* Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
-* Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`.
-Similarly, the branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
-* Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`.
+* **Type 1, direct:** `r conversion("trajectory")`.
+These methods assign each cell to a branch together with a pseudotime value and a branch network.
+The branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
+* **Type 2, linear pseudotime:** `r conversion("linear")`. These methods return a single pseudotime value. A single edge between two milestones, its percentage proportional to the pseudotime.
+* **Type 3, end state probability:** `r conversion("end_state_probability")`.
+These methods return a global pseudotime value and a probability for every end state.
 We use a single start state and add an edge to every end milestone each representing an end state.
 The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
-* Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`.
+* **Type 4, cluster assignment:** `r conversion("cluster_graph")`.
+These methods return a cluster assignment and a cluster network.
 The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
-* Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`.
+* **Type 5, projection onto nearest branch:** `r conversion("dimred_projection")`.
+These methods returning a cluster assignment, cluster network and dimensionality reduction.
 We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction.
 This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.
+* **Type 6, cell graph:** `r conversion("cell_graph")`. These methods return a network of connected cells, and determine which cells are part of the "backbone". Cell
 
@@
 ```
-This paragraph may be expanded
-Done (stolen from results section)
-🤔
+I would avoid making "absolute" statements like this one.
+maybe add something like "which we believe to be the largest list of TI methods to be compared in a single study"?
+Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
 Add the timecourse prior to the figure
+why surprisingly?
+because wishbone is only able to detect bifurcations. Should mention this in the text :)
+Added an explanation in the following sentence
+But still more should be added, I think this simply indicates that none of the methods are performing well on multifurcating datasets.
+But we should show this, eg.
+looking at the raw performance on multifurcating datasets versus other datasets.
+If we don't have the time to show this, then we should not mention wishbone here
 LINK TO
@@
 FILL IN SIGNIFICANCE LEVELS
+It's already a bit better, though could be improved still!
+do you talk about the synthetic data or the workflow?
+Because I'm not sure if the wrappers, comparison metrics..
+could be used as they are to compare NI, clustering or normalisation methods.
+Or if you are convinced of this statement, then I would add one or two sentences to explain how it could be possible.
 to do: add metric figure

### 2018-03-05 analysis/paper/paper.Rmd
commit latest version of rmd

@@
 ```
-`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³")), ", ")`, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
+`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³"), prob=c(9999, 1)), ", ")`, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
 
@@
 ³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
-⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, LYON, France
+⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
 
@@
 In order to facilitate the comparison of the results of these methods to each other and to a gold standard, we developed a global framework to benchmark trajectory inference tools.
-.
 Using this framework, we compared the results from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real and synthetic datasets.
 We evaluate methods using several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
+We found that some methods, including Slingshot, TSCAN and Monocle DDRTree, clearly outperform.
+Based on these results, we developed a set of guidelines for method users.
+However, our analysis also indicated that there is still a lot of room for improvement for methods detecting complex trajectory topologies.
+Our evaluation pipeline can therefore be used to spearhead the development of new scalable and more accurate methods.
+Our evaluation pipeline is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
@@
   "methods_over_time", 
-  "New TI methods over time. a) Number of methods published or in preprint. b) Number of methods which can handle a particular type of trajectory."
+  "New TI methods over time.
+a) Number of methods published or in preprint.
+b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
+c) Number of methods which can handle a particular type of trajectory."
+)
+```
+
+```{r, results='asis' }
+add_table(
+  "implementations",
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
+  "Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated [online](https://goo.gl/FnsTxG)."
 )
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`).
-We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` TI methods (`r ref("table", "implementations")`), which we believe to be the most complete list to date.
-It is perhaps surprising that almost all of them have a unique combination of characteristics, in terms of the required inputs (prior information), produced outputs (topology fixing and trajectory type) and methodology used.
-
-```{r, results='asis' }
-add_table(
-  "implementations",
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  "Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated at https://goo.gl/FnsTxG."
-)
-```
+It is perhaps surprising that of the `r sum(implementations$is_ti, na.rm=TRUE)` methods in existence today, almost all methods have a unique combination of characteristics (`r ref("table", "implementations")`), in terms of the required inputs (prior information), produced outputs (topology fixing and trajectory type) and methodology used (not shown).
 <!-- §3: Topology inference -->
@@
   "implementation_components", 
-  "TI methods frequently share steps."
+  "Common algorithmic components shared by different TI methods.
+Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
+a) Wordcloud of the common components.
+b) Number of times a component was shared between methods.
+c) Number of methods containing a particular category."
 )
@@
   "dataset_characterisation",  
-  "Characteristics of the real datasets used for the evaluation."
+  "Characteristics of the real datasets used for the evaluation.
+Top: distributions of number of cells and genes (after filtering) and the date these datasets were published.
+Bottom: Diversity of technologies which were used to generate the dataset, the organism, whether we extracted a silver or a gold standard from the data, the type of dynamic process and the trajectory type present in the data."
 )
@@
   "synthetic_data_generation",
-  "Synthetic data generation workflow."
+  "Workflow to generate the synthetic data.
+Starting from a state network, we extracted a regulatory network using real networks which is expected to induce such a set of state transitions when simulated.
+We simulated this network using a detailed model of gene regulation which includes the thermodynamics of transcription factor binding.
+Finally, we simulated the single-cell RNA-seq (scRNA-seq) experiment itself, by matching the distribution of expression and dropouts to a real reference dataset."
 )
@@
 
-`r ref("fig", "evaluation_overview")` **An overview of several key aspects of the evaluation.** a) .
+`r ref("fig", "evaluation_overview")` **An overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
 b) In order to make any two trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
@@
   "extended_evaluation_overview",
-  "Extended overview", 
+  "Extended overview of our evaluation pipeline.
+From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
+We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
+Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
+We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
   15, 
@@
 When we looked at the benchmark scores per trajectory type, Slingshot was to be the only method that performs well across most trajectory types.
-However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, Monocle DDRTree for rooted trees, and Wishbone for multifurcations.
-This last result is unexpected, as Wishbone is not able to output any multifurcating topologies.
-We will discuss the effect of topology types and constraints on inferrable trajectory types placed by the method in Section SEC_TRAJTYPE.
+However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for rooted trees.
 
@@
   "real_synthetic_comparison", 
-  "Comparison between performance on real and synthetic datasets across trajectory types."
+  "Comparison between performance on real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
 )
@@
 In addition, the methods that were able to detect more complex trajectory types also obtained higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
-We will explore this issue in more detail in a further section SEC_TRAJTYPE.
+We will explore this issue in more detail in a further section.
 
@@
   "metric_comparison", 
-  "Comparison between the different metrics across trajectory types."
+  "Comparison between the different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
 )
@@
   "trajtype_handle_comparison", 
-  "ADD CAPTION"
+  "Comparison between the performance of methods able to handle a trajectory type with those who don't.
+Shown is the performance on those datasets containing a particular trajectory type (top), according to several metrics (right)."
 )
@@
   "linear_split_comparison", 
-  "Linear methods have a higher correlation score on non-linear datasets.
-Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods."
+  "Comparison of the performance between linear and non-linear methods.
+Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
+Their performance is shown on linear and non-linear datasets according to several performance metrics (right).
+NS: Not significant, *: p-value < 0.05, **: p-value < 0.001, ***: p-value < 0.0001"
 )
@@
 
-`r ref("fig", "topology_complexity_comparison")` **INSERT CAPTION.** Caption subdetails.
+`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** a) % of datasets on which a TI method detects a particular trajectory type, compared with the correct trajectory type.
+b) Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
 
@@
   "topology_sensitivity", 
-  "ADD CAPTION"
+  "Sensitivity of each TI method in predicting the correct trajectory topology.
+Shown are the % of datasets where a method is able to predict the exact correct trajectory topology, depending on the trajectory type present in the data.
+When a method is able to predict a particular trajectory type, the box is surrounded by a black border."
 )
@@
   "required_priors_performance_comparison", 
-  "Comparison of method performance."
+  "Comparison of method performance depending on whether the method required a certain prior or not. The top performing method which requires a particular type of prior information is labelled."
 )
@@
   "method_components_scores", 
-  "Score of methods containing particular algorithmic components. NS: Not significant. *: "
+  "Score of methods containing particular algorithmic components.
+Shown are the algorithmic components present in four or more methods, and whether the peformance of methods containing a certain component was significantly higher than all the other methods.
+NS: Not significant.
+*: corrected p-value < 0.05 "
 )
@@
   "component_importance", 
-  "Importance values for algorithmic components."
+  "Importance values for algorithmic components.
+The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
 )
@@
   "qc_checks", 
-  "Scoring scheme for method quality control."
+  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
 )
@@
 
-`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.**
+`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each methods on every item from our quality controle score sheet (`r ref("stable", "qc_checks")`).
+Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
+Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation.
+This weight is represented in the plot as distance on the y-axis.
+Top: Average QC score for each method.
+Right: The average score of each quality control item.
+Shown into more detail are those items which had a score lower than 0.5.
 
@@
   "qc_performance_comparison", 
-  "Comparison between method qc (x-axis) and method performance (y-axis)."
+  "Comparison between method quality control scores and overall method performance."
 )
@@
 
-`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.**
+`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.** As the performance of a method most heavily depended on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
+We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
+Methods to the right are ranked according to overall performance.
+Further to the right are shown the user friendly scores (++: ≥ 0.9, +: ≥ 0.8, ± ≥ 0.65, - ≥ 0.5) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
 
 ----
+
+When choosing a method, it is important to take two further points into account.
+First, it is important that a trajectory and the downstream results and/or hypotheses originating from it are confirmed by multiple TI methods.
+This to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type.
+Second, even if the expected topology is known, it can be beneficial to also try out methods which make the less assumptions about the trajectory topology.
+When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the data.
+When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
 
@@
 
-* Inclusion of more recent methods
+* Inclusion of methods published during of after june 2017
 * A parameter tuning for each method, both across all trajectory types, as well as on specific trajectory types.
@@
 * Evaluate methods on other single-cell -omics datasets, such as proteomics and epigenomics data
+* Provide functionality for visually interpreting and comparing predicted trajectories
 
@@
 <!-- Synthetic data: observations and possibilities -->
-The use of synthetic data to evaluate methods can be problematic, because the model to generate the data is not necessarily similar to the way a true biological system works.
-This could bias an evaluation towards methods for which the underlying model best fits the model used for generating data.
+The use of synthetic data to evaluate TI methods offers the advantage of having an exact gold standard to which the methods’ results can be compared.
+However, the use of synthetic data can also be questionable, because the model which generates the data does not necessarily reflect the intrinsic characteristics of true biological systems.
+This could bias an evaluation towards methods for which the underlying model best fits the model used for generating the data.
 Therefore, it is essential that results of an evaluation on synthetic data are confirmed using real data.
 In our study, the overall performance of methods was very similar between real and synthetic datasets, confirming the biological relevance of the synthetic data.
-Given this, we believe our synthetic data can now be used to effectively prototype new TI methods, which use ever more complex models, such as disconnected graphs.
-Furthermore, it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
-Synthetic data based on our workflow could therefore be used to spearhead the development of these methods, given that currently only a limited number of real datasets are available for which the methods could be useful.
-Furthermore, we believe our evaluation workflow could also be used to evaluate other types of single-cell modelling techniques, such as single-cell network inference, clustering and normalisation.
+Given this, we believe that our synthetic data can now be used to effectively prototype new TI methods for more complex trajectories such as disconnected graphs, for which the availability of real datasets is poor.
+Furthermore, it is expected that in the future, methods will be able to model even more complex cellular behaviors, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
+Synthetic data generated with  our workflow could therefore be used to spearhead the development of these methods, given that currently only a limited number of real datasets are available for which the methods could be useful.
+Furthermore, we believe that our data generation workflow could also be used to evaluate other types of single-cell modelling techniques, such as single-cell network inference, clustering and normalisation.
 We sincerely hope that such efforts will lead to a more rapid development of accurate methods, and will in the near future provide a package which can be used to simulate synthetic data for a wide variety of single-cell modelling problems.
@@
 For real data, the states from the gold/silver standard.
-For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
+For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
 * **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
-* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.
+* **Time course** For each cell a time point from which it was sampled.
+If available, directly extracted from the gold standard.
+For synthetic data: four timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
+.
 
@@
 We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping", "a")`).
-This model consisted of three parts.
+This model consists of three parts.
 (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
-(ii) The milestone percentages contain for each cell its position between milestones, and sums for each cell to one.
+(ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one.
 (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones.
@@
 
-Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "wrapping", "b")`). Special conversions are denoted by an \*.
+Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "wrapping", "b")`).
+Special conversions are denoted by an \*, and will be explained in more detail below.
 
@@
    pull(text) %>% 
+    sort() %>%
    list_sentence()
@@
 The branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
-* **Type 2, linear pseudotime:** `r conversion("linear")`. These methods return a single pseudotime value. A single edge between two milestones, its percentage proportional to the pseudotime.
+* **Type 2, linear pseudotime:** `r conversion("linear")`.
+These methods return a pseudotime value for each cell.
+The milestone network will consist of a single edge between two milestones, where cells are positioned on the transition proportional to their pseudotime value.
 * **Type 3, end state probability:** `r conversion("end_state_probability")`.
@@
 This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.
-* **Type 6, cell graph:** `r conversion("cell_graph")`. These methods return a network of connected cells, and determine which cells are part of the "backbone". Cell
+* **Type 6, cell graph:** `r conversion("cell_graph")`.
+These methods return a network of connected cells, and determine which cells are part of the "backbone".
+One milestone is created for each cell that is part of the backbone and has a degree $\neq$ 2.
+Cells are positioned on the closest segment that is part of the backbone.
 
@@
 We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")` and `r ref("sfig", "dataset_characterisation")`).
-The scripts to download and process these datasets will be made available n our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
-When available, we preferred to start from the raw counts data.
+The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
+Whenever possible, we preferred to start from the raw counts data.
 These raw counts were all normalised and filtered using a common pipeline, discussed later.
@@
 Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "synthetic_data_generation")`).
-At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
+At every step, we took great care to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
 For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
@@
 
-One of the main process involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcations seem to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009].
+One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
+The exact way a cell choses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
+One driver of bifurcation seems to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009].
 Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
@@
 
-To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
-Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated trajectories which simple undirected graphs (bifurcating loop and bifurcating convergence).
+To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
+Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple undirected graphs (bifurcating loop and bifurcating convergence).
 
-From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks.
+From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added.
 For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
-5% of genes were assigned to be part of the main, and were randomly distributed among all modules (with at least one gene per module).
-We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module.
+5% of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module).
+We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module.
 To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.
@@
 add_sfig(
-  read_rds(figure_file(
-    "modulenets.rds",
-     experiment_id = "2-dataset_characterisation/3-synthetic"
-   )), 
+  figure_file("modulenets.svg", experiment_id = "2-dataset_characterisation/3-synthetic"),
   "modulenets", 
@@
 
-### Simulation of thermodynamics
+### Simulation of gene regulatory systems using thermodynamic models
 
-To simulate the gene regulatory network, we used a system of differential equations similar to which has been standard in evaluations of network inference [@marbach_wisdom_2012].
-In this model, changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations [@schaffter_genenetweaversilicobenchmark_2011] (ODEs):
+To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods [@marbach_wisdom_2012].
+In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations [@schaffter_genenetweaversilicobenchmark_2011] (ODEs):
 
@@
 
+We converted each ODE to an SDE by adding a chemical Langevin equation, as described in [@schaffter_genenetweaversilicobenchmark_2011].
 These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$.
@@
 
-Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward.
-Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable.
+Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward.
+Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
 To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
@@
 
-## Common normalisation pipeline
+## Expression normalisation pipeline
 We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016].
 The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3).
-We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05.
+We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05.
 Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5.
-As a final filter, we removed both all zero genes and cells until convergence.
+As a final filter, we removed both zero genes and cells until convergence.
 
 ## Evaluation metrics
-We used three different metrics to assess the performance of each TI method (`r ref("fig", "evaluation_overview")`).
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-Indeed, each of our metrics looks at the correspondence between prediction and gold standard in a different way: the correlation between geodesic distances assesses the overall ordering, the RF prediction error assesses the cell neighbourhood, and the edge flip score assesses the milestone network topology (`r ref("sfig", "score_aspects")`).
+We defined three metrics for comparing the likeness of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`).
+Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
 
 ```{r}
-add_sfig({
-  read_rds(figure_file(
-    "score_aspects.rds", 
-    experiment_id="3-score_rules/5-score_aspects"))
-  }, 
+add_sfig(
+  figure_file("score_aspects.svg", experiment_id="3-score_rules/5-score_aspects"), 
   "score_aspects", 
@@
 ### Correlation between geodesic distances
-The accuracy of cell ordering was assessed by first calculating the geodesic distances $d(c_i, c_j)$ between all cells in both the gold standard and the prediction.
-The geodesic distance is calculated by finding the shortest path between two cells through the milestone network, while taking into account the edge lengths as returned by most methods.
-To calculate this for every pair of cells, we first calculated the shortest path lengths $d(m_a, m_b)$ between all milestones using Dijkstra's algorithm, as implemented in the R *igraph* package.
-We also calculated $d(m_a, c_i)$ for every cell to its neighbouring milestones by calculating the earth's movers distance using the shortlist method [@gottschlich_shortlistmethodfast_2014], as implemented in the R *transport* package.
-Then we again used Dijkstra's algorithm to calculate the shortest path lengths between every pair of cells which minimises:
+The similarity in cell ordering between two trajectories is assessed by calculating the geodesic distances between each pair of cells for both trajectories.
+The definition of a geodesic distance between two cells part of the common trajectory model will be demonstrated using a toy example (`r ref("sfig", "metric_geodesic")`).
 
-$$
-d(c_i, c_j) = min_{a \in A, b \in B} d(m_a, c_i) + d(m_a, m_b) + d(m_b, c_j)
-$$
+```{r}
+add_sfig(
+  figure_file("metric_geodesic.svg", experiment_id="evaluation_overview"), 
+  "metric_geodesic", 
+  "**The modified geodesic distances demonstrated on a toy example.** a) A toy example containing four milestones (W to Z) and five cells (a to e).
+b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
+c) The calculations made for calculating the pairwise geodesic distances.
+d) A heatmap representation of the pairwise geodesic distances."
+)
+```
 
-where A and B contain all milestones adjacent to respectively $c_i$ and $c_j$.
-The final score was then calculated the Spearman's rank correlation between all $d(c_i, c_j)$ in gold standard and prediction.
+The geodesic distance between two cells depends on whether they are (i) on the same transition, or (ii) in different transitions.
+In the first case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on.
+For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
+In the latter case, the distances between the cells and all of their neighbouring milestones will be calculated.
+These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
+For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
 
-In some cases, methods are unable to place some cells on a trajectory, mainly because they classify these cells as noisy.
-In these cases, we add these cells to a new milestone and put this milestone at a distance of 5 times the maximal distance between two milestones within the network.
+According to the defined common trajectory model (`r ref("fig", "evaluation_overview", "b")`), cells are also allowed to have a delayed commitment.
+In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network.
+In this case, the distance between two cells both inside a region of delayed commitment is calculated as the manhattan distances between the milestone percentages weighted by the transition weights from the milestone network.
+For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
+The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
+
+Finally, calculating all pairwise distances between cells would scale poorly for trajectories with large numbers of cells.
+For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
+The waypoints are determined by viewing each milestone, transition and region of delayed commitment as a collection of cells, and sampling cells from the different collections weighted by the total number of cells within that collection.
+For calculating the correlation of geodesic distances between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
+For the benchmark evaluation, the total number of waypoints sampled from a trajectory was one hundred.
 
@@
 We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard.
-We used a Random Forest regression, implemented in the ranger package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of every cell in the gold standard, using the milestone percentages of these cells in the prediction as features.
+We used a Random Forest regression, implemented in the *ranger* package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of every cell in the gold standard, using the milestone percentages of these cells in the prediction as features.
 We then used the out-of-bag mean-squared error on these percentages to score each method's capability of predicting the correct neighbourhood of each cell.
@@
 
-As a third independent score, we assessed the similarity between the milestone network topology.
-We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks.
-Because we are interested , the network was made undirected.
+As a third independent score, we assessed the similarity between the milestone network topologies.
+We first simplified each network, by merging consecutive linear edges into one edge, and adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks.
+Because we are interested in the overall similarity between two topologies irrespective of the direction of the edges, the network was made undirected.
 Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
-This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no scalable solutions [@bahiense_maximumcommonedge_2012].
-We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes.
-To further limit the search space, we make sure the degree distributions between the two networks are similar, before assessing whether the two networks are isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
+This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012].
+We implemented a branch and bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, we tried out all solutions with additional two edge changes.
+To further limit the search space, we made sure the degree distributions between the two networks were  similar, before assessing whether the two networks were isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
+
+A comparison of edge flip scores between common trajectory topologies is illustrated in `r ref("fig", "edge_flip_comparison")`.
 
@@
   "edge_flip_comparison", 
-  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets")), 
+  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets"), "."), 
   15, 
@@
 ### Aggregation of scores
+`r ref("fig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
+In order to make the scores produced by the different metrics comparable to one another, across datasets of varying difficulty, the raw scores are percentile rank transformed per metric per dataset to a $[0,1]$ range.
+The arithmetic mean is calculated across replicates.
+In order to ensure a method only obtains a high score if it scores well on all three metrics, the harmonic mean across metrics is calculated.
+At this point, the aggregated scores of all methods across all datasets is combined.
+As there can be an overrepresentation of datasets of a certain trajectory type, first an arithmetic mean is calculated per trajectory type, followed by an overall arithmetic mean across all trajectory types, thus obtaining a ranking of the methods.
+
 ```{r}
@@
 
-### Pipeline
-<!-- Look at dyneval methods to write this paragraph -->
-
 ### Method execution
-We ran each method in a controlled environment, where we limited the memory usage (max.
-32GB) and total run time (max 24h).
-Moreover, we disabled the function of the R `set.seed` function, to avoid that methods false make their output deterministic.
+Each execution of a method on a dataset was performed in a separate task as part of a gridengine job.
+Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 0 @ 2.40GHz, and was provided a maximum of 32GB in memory and 6 hours of wall time.
+One R session was started for each task, with the environment variable `R_MAX_NUM_DLLS` set to 500.
+The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust.
+Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.
 
 ### Effect of prior information
-
 To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test.
@@
 The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
-Finally, we also assessed the certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset on which the usefulness of the method was shown, and the scope of method evaluation in the paper.
+Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset in which the usefulness of the method was shown, and the scope of method evaluation in the paper.
 
@@
 
-Each quality aspect received a weight by how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
-This was to make sure that more important aspect, such as the open source availability of the method, outweigh other aspects, such as the availability of a graphical user interface.
+Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface.
 Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("table", "implementations")`).
@@
 ## Trajectory types
-
 We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
-These trajectory types start from the most general trajectory type, a disconnected directed graph, and moves down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`).
+These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical (`r ref("sfig", "trajectory_type_tree", "a")`).
 For every directed trajectory type, a corresponding undirected trajectory type can also be defined (`r ref("sfig", "trajectory_type_tree", "b")`).
-In most cases, a method which can detect a complex trajectory type, was also be able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).
+In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).
 
@@
   "trajectory_type_tree", 
-  "Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types.", 
+  "Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types.
+a) The graph structure for undirected and directed trajectory types.
+Shown are the topological properties which change when going from a particular trajectory type to a less complex type.
+b) Combined structure for directed and undirected trajectory types.",
   15, 
@@
 
-# Acknowledgements
-
 # Supplementary Material
-
 ```{r "supplementary", echo=FALSE, results='asis'}
@@
 ```
-
-The current Git commit details are:
-
-```{r}
-git2r::repository(".")
-```
-I would avoid making "absolute" statements like this one.
-maybe add something like "which we believe to be the largest list of TI methods to be compared in a single study"?
-Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
-Add the timecourse prior to the figure
-why surprisingly?
-because wishbone is only able to detect bifurcations. Should mention this in the text :)
-Added an explanation in the following sentence
-But still more should be added, I think this simply indicates that none of the methods are performing well on multifurcating datasets.
-But we should show this, eg.
-looking at the raw performance on multifurcating datasets versus other datasets.
-If we don't have the time to show this, then we should not mention wishbone here
-LINK TO
-
-### Trajectory types and topology complexity
-LINK TO
-
-### Trajectory types and topology complexity
-ADD CAPTION
-TODO
-ADD CAPTION
-FILL IN SIGNIFICANCE LEVELS
-It's already a bit better, though could be improved still!
-do you talk about the synthetic data or the workflow?
-Because I'm not sure if the wrappers, comparison metrics..
-could be used as they are to compare NI, clustering or normalisation methods.
-Or if you are convinced of this statement, then I would add one or two sentences to explain how it could be possible.
-to do: add metric figure
-needs to be updated with waypoints
-this is not used anymore
-TODO
-More to add here?
-* Look at dynmethods::execute_method to write this paragraph
-Are there more exectuion details?

### 2018-03-05 analysis/paper/paper_latex.Rmd
commit latest version of rmd

@@
 add_sfig(
-  read_rds(figure_file(
-    "modulenets.rds",
-     experiment_id = "2-dataset_characterisation/3-synthetic"
-   )), 
+  figure_file("modulenets.tmp.pdf", experiment_id = "2-dataset_characterisation/3-synthetic"),
   "modulenets", 
@@
 
-## Common normalisation pipeline
+## Expression normalisation pipeline
 We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016].
 The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3).
-We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05.
+We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05.
 Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5.
-As a final filter, we removed both all zero genes and cells until convergence.
+As a final filter, we removed both zero genes and cells until convergence.
 
 ## Evaluation metrics
-We used three different metrics to assess the performance of each TI method (`r ref("fig", "evaluation_overview")`).
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-Indeed, each of our metrics looks at the correspondence between prediction and gold standard in a different way: the correlation between geodesic distances assesses the overall ordering, the RF prediction error assesses the cell neighbourhood, and the edge flip score assesses the milestone network topology (`r ref("sfig", "score_aspects")`).
+We defined three metrics for comparing the likeness of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`).
+Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
 
 ```{r}
-add_sfig({
-  read_rds(figure_file(
-    "score_aspects.rds", 
-    experiment_id="3-score_rules/5-score_aspects"))
-  }, 
+add_sfig(
+  figure_file("score_aspects.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
   "score_aspects", 
@@
 ### Correlation between geodesic distances
-The accuracy of cell ordering was assessed by first calculating the geodesic distances $d(c_i, c_j)$ between all cells in both the gold standard and the prediction.
-The geodesic distance is calculated by finding the shortest path between two cells through the milestone network, while taking into account the edge lengths as returned by most methods.
-To calculate this for every pair of cells, we first calculated the shortest path lengths $d(m_a, m_b)$ between all milestones using Dijkstra's algorithm, as implemented in the R *igraph* package.
-We also calculated $d(m_a, c_i)$ for every cell to its neighbouring milestones by calculating the earth's movers distance using the shortlist method [@gottschlich_shortlistmethodfast_2014], as implemented in the R *transport* package.
-Then we again used Dijkstra's algorithm to calculate the shortest path lengths between every pair of cells which minimises:
+The similarity in cell ordering between two trajectories is assessed by calculating the geodesic distances between each pair of cells for both trajectories.
+The definition of a geodesic distance between two cells part of the common trajectory model will be demonstrated using a toy example (`r ref("sfig", "metric_geodesic")`).
 
-$$
-d(c_i, c_j) = min_{a \in A, b \in B} d(m_a, c_i) + d(m_a, m_b) + d(m_b, c_j)
-$$
+```{r}
+add_sfig(
+  figure_file("metric_geodesic.tmp.pdf", experiment_id="evaluation_overview"), 
+  "metric_geodesic", 
+  "**The modified geodesic distances demonstrated on a toy example.** a) A toy example containing four milestones (W to Z) and five cells (a to e).
+b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
+c) The calculations made for calculating the pairwise geodesic distances.
+d) A heatmap representation of the pairwise geodesic distances."
+)
+```
 
-where A and B contain all milestones adjacent to respectively $c_i$ and $c_j$.
-The final score was then calculated the Spearman's rank correlation between all $d(c_i, c_j)$ in gold standard and prediction.
+The geodesic distance between two cells depends on whether they are (i) on the same transition, or (ii) in different transitions.
+In the first case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on.
+For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
+In the latter case, the distances between the cells and all of their neighbouring milestones will be calculated.
+These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
+For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
 
-In some cases, methods are unable to place some cells on a trajectory, mainly because they classify these cells as noisy.
-In these cases, we add these cells to a new milestone and put this milestone at a distance of 5 times the maximal distance between two milestones within the network.
+According to the defined common trajectory model (`r ref("fig", "evaluation_overview", "b")`), cells are also allowed to have a delayed commitment.
+In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network.
+In this case, the distance between two cells both inside a region of delayed commitment is calculated as the manhattan distances between the milestone percentages weighted by the transition weights from the milestone network.
+For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
+The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
+
+Finally, calculating all pairwise distances between cells would scale poorly for trajectories with large numbers of cells.
+For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
+The waypoints are determined by viewing each milestone, transition and region of delayed commitment as a collection of cells, and sampling cells from the different collections weighted by the total number of cells within that collection.
+For calculating the correlation of geodesic distances between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
+For the benchmark evaluation, the total number of waypoints sampled from a trajectory was one hundred.
 

### 2018-03-05 analysis/paper/paper_latex.Rmd
update paper

@@
   - \usepackage[labelformat=empty]{caption}
-  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript}
+  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
 mainfont: Open Sans
@@
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory networks using a thermodynamic model [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
+To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
 
@@
 ```
+Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
+We will do that in V2 for sure! Thanks!
+And maybe maybe if someone finds the time this evening :)
+In the next version of the paper, an overview of all the errors produced by a method and an overview of all of the timings should be provided.
+For the next version of the paper, explain the different metrics (and the common trajectory format) with one supplementary figure, like figure evaluation_overview/metric_geodesic.tmp.pdf

### 2018-03-05 analysis/paper/paper.Rmd
update paper

@@
 ```
+Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
+We will do that in V2 for sure! Thanks!
+And maybe maybe if someone finds the time this evening :)
+In the next version of the paper, an overview of all the errors produced by a method and an overview of all of the timings should be provided.
+For the next version of the paper, explain the different metrics (and the common trajectory format) with one supplementary figure, like figure evaluation_overview/metric_geodesic.svg

### 2018-03-05 analysis/paper/paper.Rmd

@@
 In order to facilitate the comparison of the results of these methods to each other and to a gold standard, we developed a global framework to benchmark trajectory inference tools.
-Using this framework, we compared the results from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real and synthetic datasets.
+Using this framework, we compared the trajectories from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real and synthetic datasets.
 We evaluate methods using several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We found that some methods, including Slingshot, TSCAN and Monocle DDRTree, clearly outperform.
-Based on these results, we developed a set of guidelines for method users.
-However, our analysis also indicated that there is still a lot of room for improvement for methods detecting complex trajectory topologies.
-Our evaluation pipeline can therefore be used to spearhead the development of new scalable and more accurate methods.
-Our evaluation pipeline is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+We found that some methods, including Slingshot, TSCAN and Monocle DDRTree, clearly outperform other methods, although their performance depended on the type of trajectory present in the data.
+Based on our benchmarking results, we therefore developed a set of guidelines for method users.
+However, our analysis also indicated that there is still a lot of room for improvement, especially for methods detecting complex trajectory topologies.
+Our evaluation pipeline can therefore be used to spearhead the development of new scalable and more accurate methods, and is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 

### 2018-03-05 analysis/paper/paper.Rmd
remove comments from end

@@
 
-In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
-In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
-The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
+In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources
 
@@
 
-During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
-If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
-Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
-Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
-For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
+During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error
 
@@
 ```
-Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
-We will do that in V2 for sure! Thanks!
-And maybe maybe if someone finds the time this evening :)
-In the next version of the paper, an overview of all the errors produced by a method and an overview of all of the timings should be provided.
-For the next version of the paper, explain the different metrics (and the common trajectory format) with one supplementary figure, like figure evaluation_overview/metric_geodesic.svg

### 2018-03-05 analysis/paper/paper_latex.Rmd
remove comments from end

@@
 ```
-Maybe add a figure here showing the real outputs of some methods(which are very method specific and not so easy to compare), and how they can nicely be compared when using your common visualisation tool?
-We will do that in V2 for sure! Thanks!
-And maybe maybe if someone finds the time this evening :)
-In the next version of the paper, an overview of all the errors produced by a method and an overview of all of the timings should be provided.
-For the next version of the paper, explain the different metrics (and the common trajectory format) with one supplementary figure, like figure evaluation_overview/metric_geodesic.tmp.pdf

### 2018-03-05 analysis/paper/update_paper_from_google_drive.R
remove comments ²

@@ -10,7 +10,7 @@ system(pritt("cat {drive$local_path} > analysis/paper/paper.Rmd"))
 # add wip
 read_file("analysis/paper/paper.Rmd") %>%
   str_replace_all("§(.*?)\n", "<p class='wip'>\\1</p>") %>%
-  str_replace_all("\\[[a-z]\\][^\n]*", "") %>%
+  str_replace_all("\n\\[[a-z]\\][^\n]*", "") %>%
   str_replace_all("\\[[a-z]\\]", "") %>%
   str_replace_all("→", "-->") %>%
   str_replace_all("\n#", "\n\n#") %>% # add double new line before (sub)titles

### 2018-03-12 analysis/paper/paper.Rmd

@@
 Tips for working in this document:
-* Go to tools -> perferences and remove Use smart quotes
+* Go to tools -> preferences and remove Use smart quotes
 * Enable automatic zotero syncing to bibtex
 * Make sure to pin the zotero bibtex references after first using them!
-
-TITLES: 
-Not all trajectory inference methods are created equal
-Trajectory inference: work in progress
-Trajectory inference: we’re not quite there yet
-Paving the way for trajectories
-Benchmarking trajectory inference methods
-Bringing single-cell trajectory inference methods to the next level
-Resolving disarray amongst trajectory inference methods with a comprehensive evaluation
-A comparison of trajectory inference methods
-Trajectory inference 2.0
-We’re on the trajectory to better trajectory inference methods
-On the road to more accurate and robust single-cell trajectory inference methods
-Towards more accurate and robust single-cell trajectory inference methods
-On the trajectory to more accurate,robust and trajectory-like single-cell trajectory inference method trajectories
-A comparison of single-cell trajectory inference methods: towards more accurate and robust tools
 -->
@@
 
-Using single-cell transcriptomics or proteomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes.
+Using single-cell -omics data, it is now possible to computationally order cells along trajectories, allowing the unbiased study of cellular dynamic processes.
 Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
@@
 Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), which use single-cell profiles from a population in which the cells are at different unknown points in the dynamic process  [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex tree or graph structures.
+These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or a more complex tree or graph structure.
 Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
@@
   read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  "Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated [online](https://goo.gl/FnsTxG)."
+  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof.
+This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
 )
@@
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single cell profiling experiment of different single-cell RNA-seq technologies (`r ref("sfig", "synthetic_data_generation")`).
+These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
+To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets (`r ref("sfig", "synthetic_data_generation")`).
 
@@
 
-`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the maximum complexity trajectory they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally make use of.
+`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
 b) The methods are ordered according to the overall score achieved in the benchmark evaluation.
@@
 Having ordered all methods by their overall benchmarking score, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
-When we looked at the benchmark scores per trajectory type, Slingshot was to be the only method that performs well across most trajectory types.
-However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for rooted trees.
+When we looked at the benchmark scores per trajectory type, Slingshot was the only method that performed well across most trajectory types.
+However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for trees.
 
@@
 Their performance is shown on linear and non-linear datasets according to several performance metrics (right).
-NS: Not significant, *: p-value < 0.05, **: p-value < 0.001, ***: p-value < 0.0001"
+NS: Not significant, \\*: p-value < 0.05, \\*\\*: p-value < 0.001, \\*\\*\\*: p-value < 0.0001"
 )
@@
   "Score of methods containing particular algorithmic components.
-Shown are the algorithmic components present in four or more methods, and whether the peformance of methods containing a certain component was significantly higher than all the other methods.
+Shown are the algorithmic components present in four or more methods, and whether the performance of methods containing a certain component was significantly higher than all the other methods.
 NS: Not significant.
@@
 
-`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each methods on every item from our quality controle score sheet (`r ref("stable", "qc_checks")`).
+`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each methods on every item from our quality control score sheet (`r ref("stable", "qc_checks")`).
 Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
@@
 Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
+It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
 In this preprint, we present an  initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters.
@@
 ### Method selection
-We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
 A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
@@
 For synthetic data: four timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
-.
 
@@
 
-A comparison of edge flip scores between common trajectory topologies is illustrated in `r ref("fig", "edge_flip_comparison")`.
+A comparison of edge flip scores between common trajectory topologies is illustrated in `r ref("sfig", "edge_flip_comparison")`.
 
@@
 ### Aggregation of scores
-`r ref("fig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
+`r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
 In order to make the scores produced by the different metrics comparable to one another, across datasets of varying difficulty, the raw scores are percentile rank transformed per metric per dataset to a $[0,1]$ range.

### 2018-03-12 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
add labels for qc categories

@@
     scale_x_continuous("", breaks=NULL, expand = c(0, 0)) +
-    scale_fill_manual("Category", values=set_names(qc_categories$color, qc_categories$category), label=label_long, guide=guide_legend(ncol=2)) +
+    scale_fill_manual("Category", values=set_names(qc_categories$color, qc_categories$category), labels=set_names(qc_categories$label, qc_categories$category), guide=guide_legend(ncol=2)) +
     scale_alpha_continuous("Score", guide=guide_legend(ncol=5, label.position = "bottom")) +

### 2018-03-12 analysis/paper/paper.Rmd
update paper

@@
 add_sfig(
-  figure_file("score_aspects.svg", experiment_id="3-score_rules/5-score_aspects"), 
+  figure_file("score_aspects_v1.svg", experiment_id="3-score_rules/5-score_aspects"), 
   "score_aspects", 
@@
 Each execution of a method on a dataset was performed in a separate task as part of a gridengine job.
-Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 0 @ 2.40GHz, and was provided a maximum of 32GB in memory and 6 hours of wall time.
+Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and was provided a maximum of 32GB in memory and 6 hours of wall time.
 One R session was started for each task, with the environment variable `R_MAX_NUM_DLLS` set to 500.

### 2018-03-14 analysis/paper/paper_latex.Rmd
update paper

@@
 ```
-`r glue::collapse(sample(c("Wouter Saelens* ¹ ²", "Robrecht Cannoodt* ¹ ² ³"), prob=c(9999, 1)), ", ")`, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
+Wouter Saelens* ¹ ², Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
 
@@
 
-Using single-cell -omics data, it is now possible to computationally order cells along trajectories, allowing the unbiased study of cellular dynamic processes.
-Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-As a result, choosing a method to infer trajectories is often challenging, since a comprehensive assessment of the performance and robustness of each method is still lacking.
-In order to facilitate the comparison of the results of these methods to each other and to a gold standard, we developed a global framework to benchmark trajectory inference tools.
-Using this framework, we compared the trajectories from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real and synthetic datasets.
-We evaluate methods using several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We found that some methods, including Slingshot, TSCAN and Monocle DDRTree, clearly outperform other methods, although their performance depended on the type of trajectory present in the data.
-Based on our benchmarking results, we therefore developed a set of guidelines for method users.
-However, our analysis also indicated that there is still a lot of room for improvement, especially for methods detecting complex trajectory topologies.
-Our evaluation pipeline can therefore be used to spearhead the development of new scalable and more accurate methods, and is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+Since 2014, at least `r sum(implementations$is_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+As a result, a comprehensive assessment of the performance and robustness of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
 
-To our knowledge, this is the first comprehensive assessment of trajectory inference methods.
-For now, we exclusively evaluated the methods on their default parameters, but plan to add a detailed parameter tuning procedure in the future.
-We gladly welcome any discussion and feedback on key decisions made as part of this study, including the metrics used in the benchmark, the quality control checklist, and the implementation of the method wrappers.
-These discussions can be held at [github.com/dynverse/dynverse/issues](https://www.github.com/dynverse/dynverse/issues).
+In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness.
+Our results indicate that Slingshot is the most accurate method to date, but other leading methods (e.g.
+TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases.
+We therefore developed a set of guidelines to aid users select the most appropriate method.
+While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases.
+We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
 # Introduction
-<!-- §1: Single-cell omics and TI --->
 Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017].
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), which use single-cell profiles from a population in which the cells are at different unknown points in the dynamic process  [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or a more complex tree or graph structure.
-Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
+Such dynamic processes can be computationally modelled by ordering cells along a trajectory, also called trajectory inference (TI) or pseudotemporal ordering [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs.
+TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
-
-```{r}
-add_sfig(
-  figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
-  "methods_over_time", 
-  "New TI methods over time.
-a) Number of methods published or in preprint.
-b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
-c) Number of methods which can handle a particular type of trajectory."
-)
-```
-
-```{r, results='asis' }
-add_table(
-  "implementations",
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof.
-This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
-)
-```
 
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`).
-It is perhaps surprising that of the `r sum(implementations$is_ti, na.rm=TRUE)` methods in existence today, almost all methods have a unique combination of characteristics (`r ref("table", "implementations")`), in terms of the required inputs (prior information), produced outputs (topology fixing and trajectory type) and methodology used (not shown).
-<!-- §3: Topology inference -->
-One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
+All methods have their own unique set of characteristics (`r ref("stable", "implementations")`) in terms of underlying algorithm, the prior information they require and the outputs they produce.
+Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect.
 Early TI methods typically fixed the topology algorithmically (e.g.
@@
 These methods therefore mainly focused on correctly ordering the cells along this fixed topology.
-Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods  to be broadly applicable on more use cases.
+Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods  to be applicable on more use cases.
 Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).
@@
 
-<!-- §4: Prior information -->
-Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
-Prior information can be supplied as  a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
-Providing prior information to a TI method can be both a blessing and a curse.
-In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
-On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
-Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.
-
-<!-- §5: Components -->
-A reductionist approach to characterising TI methods consists in dissecting them into a set of algorithmic components, as any component can have a significant impact on the performance, scalability, and output data structures.
-Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-```{r}
-add_sfig(
-  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
-  "implementation_components", 
-  "Common algorithmic components shared by different TI methods.
-Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
-a) Wordcloud of the common components.
-b) Number of times a component was shared between methods.
-c) Number of methods containing a particular category."
-)
-```
-
-<!-- §6: Problem statement -->
-Given the  diversity in TI methods, an important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
-Many attempts at tackling  this issue have already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, a comprehensive benchmarking evaluation of TI methods is still lacking.
-This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would be the most optimal for their problem. 
+<!-- §3: Problem statement -->
+Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness..
+Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods is still lacking.
+This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem.
 Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
@@
 # Results
-In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (Overview: `r ref("fig", "evaluation_overview", "a")`, Extended overview: `r ref("sfig", "extended_evaluation_overview")`).
+
+In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`, and `r ref("sfig", "extended_evaluation_overview")`).
 The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the date of publication  (`r ref("table", "implementations")`).
-Only methods published before june 2017 are included in the current version of the evaluation, while more recent methods will be added in the next version.
+Only methods published before june 2017 are included in the current version of the evaluation.
 The evaluation comprised three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
@@
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
@@
 
-As the aim of this study is to provide both guidelines for the use of TI methods and a base for the development of new TI methods, we not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation.
-To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
-This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
 Finally, using both the benchmark results and quality control, we produced a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
@@
 ## Evaluation of trajectory inference methods
-<!-- §1: Introduce figure --> 
-An overview of the main results from this study is shown in `r ref("fig", "results_overview")`.
-This includes an overview of the results obtained from the method characterisation (`r ref("fig", "results_overview", "a")`), the benchmarking evaluation (`r ref("fig", "results_overview", "b")`), and the quality control evaluation (`r ref("fig", "results_overview", "c")`).
 
@@
 
-<!-- §2: Overall and topology types -->
-Having ordered all methods by their overall benchmarking score, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
-When we looked at the benchmark scores per trajectory type, Slingshot was the only method that performed well across most trajectory types.
+<!-- §1: Overall performance -->
+Across all datasets and evaluation metrics, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, slingshot was the only method that performed well across most trajectory types, followed by TSCAN and Monocle DDRTree.
 However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for trees.
+This was also reflected in the rankings of different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, performing better at our metric investigating topological similarity, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("sfig", "metric_comparison")`).
 
-<!-- §3: real versus synthetic -->
-We observed a high correlation (0.7-0.9) between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`).
-This confirms both the relevance of the synthetic data and the accuracy of the gold standard in the real datasets.
-However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available for such topologies (`r ref("stable", "datasets")`).
+<!-- §2: The stability and robustness of the evaluation/method ranking -->
+```{r}
+correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
+filter(n_tasks == round(max(n_tasks)/10)) %>% 
+          pull(cor) %>% 
+          mean()
+top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
+filter(method_id == "slngsht") %>% 
+filter(n_tasks == round(max(n_tasks)/10)) %>%
+pull(rank_match)
+random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
+filter(method_id == "scoup") %>% 
+pull(max_n_tasks_top)
+
+```
+
+We further investigated the robustness of our evaluation across datasets and dataset sources.
+Method performance was moderately to highly correlated (0.5-0.9) between real and synthetic datasets (`r ref("fig", "results_overview", "b")` and `r ref("sfig", "real_synthetic_comparison")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
+Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with most methods significantly outperforming the next method in the ranking `r ref("sfig", "performance_dataset_variability_combined")`.
+To further illustrate this, if we looked at the scores a random 10% subset of all datasets, the overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
+On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
+For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top`datasets on which it would be the top ranking method.
+Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
+```{r}
+add_sfig(
+  figure_file(
+    "real_synthetic_comparison.tmp.pdf",
+    "11-evaluation_robustness"
+  ), 
+  "real_synthetic_comparison", 
+  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
+)
+```
 
@@
   figure_file(
-    "real_synthetic_comparison.tmp.pdf",
-    "5-optimise_parameters/5-real_synthetic_comparison"
+    "performance_dataset_variability_combined.tmp.pdf",
+    "11-evaluation_robustness"
   ), 
-  "real_synthetic_comparison", 
-  "Comparison between performance on real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
+  "performance_dataset_variability_combined", 
+  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
 )
 ```
-
-<!-- §4: Compare metrics -->
-As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics.
-While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtained a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`).
-In addition, the methods that were able to detect more complex trajectory types also obtained higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
-We will explore this issue in more detail in a further section.
 
@@
   "metric_comparison", 
-  "Comparison between the different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
+  "Comparison of method performance between different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
 )
 ```
+
+### Trajectory types and topology complexity
 
@@
 
-In the current version of the evaluation, we only provided prior information when a method required it .
+In the current version of the evaluation, we only provided prior information when a method required it.
 We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`).
@@
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric.
-Good unit testing assures that  the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further.
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
+To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
+
 We therefore looked at the implementation of each method, and assessed its  quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
@@
 
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "implementation_qc_overview")`).
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`).
 We found that recent methods had a slightly better quality than older methods (`r ref("sfig", "qc_over_time")`).
@@
 
-Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`.
+Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
 As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
@@
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
-Here, to our knowledge, only ElPiGraph.R [github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.
+Here, to our knowledge, only ElPiGraph.R ([github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R)), which is not yet published, can be used.
 
@@
 # Discussion
-<!-- The context -->
-Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
-In this preprint, we present an  initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters.
-When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
-We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
-
-```{r}
-add_sfig(
-  figure_file('leading_method.tmp.pdf', "9-main_figure"), 
-  "leading_method", 
-  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
-)
-```
-
-<!-- Future work -->
-In this study, we presented our first version of the evaluation of TI methods.
-We are convinced that the results we provided will be useful for both method users and tool developers, as we provide clear practical guidelines for users depending on their current knowledge of the trajectory, as well as an objective benchmark on which new methods can be tested.
-Nonetheless, our evaluation can be expanded on several points, all of which we will try to tackle in the near future:
-
-* Inclusion of methods published during of after june 2017
-* A parameter tuning for each method, both across all trajectory types, as well as on specific trajectory types.
-* A test of robustness on noise, parameter changes and dataset size
-* An evaluation of the stability of a method's results when running the same parameter setting on the same dataset multiple times
-* Evaluate the methods when given optional priors, and compare with performance without priors
-* Assess the effect of noisy or incorrect prior information
-* Test methods on datasets with no clear trajectory present
-* Include feedback on the quality control scoring scheme, and update the scoring when method get updated
-* Test the scalability of each method, both in the number of cells and the number of features
-* Include more real datasets with complex trajectory types, as they become available
-* Evaluate methods on other single-cell -omics datasets, such as proteomics and epigenomics data
-* Provide functionality for visually interpreting and comparing predicted trajectories
-
-<!-- Are certain methods better than others? -->
-Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, TSCAN, and Monocle DDRTree, towering above all other methods.
-Nonetheless, we found that methods which did not perform well across all settings could still be useful in certain specific use cases.
-Indeed, on data containing more than one bifurcations, Monocle DDRTree clearly performed better than other methods.
-We found that this particular result was mainly caused by the fact that the default parameters of Monocle DDRTree preferably led to the detection of tree topologies, while those of Slingshot preferably found linear and bifurcation trajectories.
-
-<!-- Our common output model, advantages & limitations -->
-We managed to wrap the output of all methods into one common format.
-This not only allowed us to compare different methods with a gold standard, but could also be useful for TI users, as  it allows the user to test multiple methods on the same data and compare the results without manual conversion of input and output.
-Furthermore, it makes it possible to directly compare the output of different methods, which opens up possibilities for new comparative visualisation techniques or ensemble methods.
-However, we acknowledge that our model has some limitations.
-Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
-when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
-when the connections between branches are uncertain).
-Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
-
-<!-- Synthetic data: observations and possibilities -->
-The use of synthetic data to evaluate TI methods offers the advantage of having an exact gold standard to which the methods’ results can be compared.
-However, the use of synthetic data can also be questionable, because the model which generates the data does not necessarily reflect the intrinsic characteristics of true biological systems.
-This could bias an evaluation towards methods for which the underlying model best fits the model used for generating the data.
-Therefore, it is essential that results of an evaluation on synthetic data are confirmed using real data.
-In our study, the overall performance of methods was very similar between real and synthetic datasets, confirming the biological relevance of the synthetic data.
-Given this, we believe that our synthetic data can now be used to effectively prototype new TI methods for more complex trajectories such as disconnected graphs, for which the availability of real datasets is poor.
-Furthermore, it is expected that in the future, methods will be able to model even more complex cellular behaviors, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
-Synthetic data generated with  our workflow could therefore be used to spearhead the development of these methods, given that currently only a limited number of real datasets are available for which the methods could be useful.
-Furthermore, we believe that our data generation workflow could also be used to evaluate other types of single-cell modelling techniques, such as single-cell network inference, clustering and normalisation.
-We sincerely hope that such efforts will lead to a more rapid development of accurate methods, and will in the near future provide a package which can be used to simulate synthetic data for a wide variety of single-cell modelling problems.
 
@@
 This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).
+
+<!-- PASTED FROM INTRODUCTION, need to rework -->
+<!-- §4: Prior information -->
+Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
+Prior information can be supplied as  a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
+Providing prior information to a TI method can be both a blessing and a curse.
+In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
+On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.
+
+<!-- §5: Components -->
+A reductionist approach to characterising TI methods consists in dissecting them into a set of algorithmic components, as any component can have a significant impact on the performance, scalability, and output data structures.
+Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
+
+```{r}
+add_sfig(
+  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
+  "implementation_components", 
+  "Common algorithmic components shared by different TI methods.
+Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
+a) Wordcloud of the common components.
+b) Number of times a component was shared between methods.
+c) Number of methods containing a particular category."
+)
+```
 
@@
 add_sfig(
-  figure_file("score_aspects.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
+  figure_file("score_aspects_v1.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
   "score_aspects", 
@@
 ```{r}
-qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation")) %>% 
+qc_references <- readRDS(result_file("checks.rds", "4-method_characterisation")) %>% 
   .$references %>% 
@@
 # Supplementary Material
+```{r}
+add_sfig(
+  figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
+  "methods_over_time", 
+  "New TI methods over time.
+a) Number of methods published or in preprint.
+b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
+c) Number of methods which can handle a particular type of trajectory."
+)
+```
+
+```{r, results='asis' }
+add_stable(
+  "implementations",
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
+  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof.
+This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
+)
+```
+
 ```{r "supplementary", echo=FALSE, results='asis'}

### 2018-03-14 analysis/paper/paper.Rmd

@@
 
-Using single-cell -omics data, it is now possible to computationally order cells along trajectories, allowing the unbiased study of cellular dynamic processes.
-Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics.
-As a result, choosing a method to infer trajectories is often challenging, since a comprehensive assessment of the performance and robustness of each method is still lacking.
-In order to facilitate the comparison of the results of these methods to each other and to a gold standard, we developed a global framework to benchmark trajectory inference tools.
-Using this framework, we compared the trajectories from a total of  `r sum(methods$evaluated)` trajectory inference methods, on a large collection of real and synthetic datasets.
-We evaluate methods using several metrics, including accuracy of the inferred ordering, correctness of the network topology, code quality and user friendliness.
-We found that some methods, including Slingshot, TSCAN and Monocle DDRTree, clearly outperform other methods, although their performance depended on the type of trajectory present in the data.
-Based on our benchmarking results, we therefore developed a set of guidelines for method users.
-However, our analysis also indicated that there is still a lot of room for improvement, especially for methods detecting complex trajectory topologies.
-Our evaluation pipeline can therefore be used to spearhead the development of new scalable and more accurate methods, and is available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+Since 2014, at least `r sum(implementations$is_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+As a result, a comprehensive assessment of the performance and robustness of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
 
-To our knowledge, this is the first comprehensive assessment of trajectory inference methods.
-For now, we exclusively evaluated the methods on their default parameters, but plan to add a detailed parameter tuning procedure in the future.
-We gladly welcome any discussion and feedback on key decisions made as part of this study, including the metrics used in the benchmark, the quality control checklist, and the implementation of the method wrappers.
-These discussions can be held at [github.com/dynverse/dynverse/issues](https://www.github.com/dynverse/dynverse/issues).
+In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness.
+Our results indicate that Slingshot is the most accurate method to date, but other leading methods (e.g.
+TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases.
+We therefore developed a set of guidelines to aid users select the most appropriate method.
+While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases.
+We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
 # Introduction
-<!-- §1: Single-cell omics and TI --->
 Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017].
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), which use single-cell profiles from a population in which the cells are at different unknown points in the dynamic process  [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or a more complex tree or graph structure.
-Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
+Such dynamic processes can be computationally modelled by ordering cells along a trajectory, also called trajectory inference (TI) or pseudotemporal ordering [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs.
+TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
-
-```{r}
-add_sfig(
-  figure_file("methods_timeline.svg", "4-method_characterisation"),
-  "methods_over_time", 
-  "New TI methods over time.
-a) Number of methods published or in preprint.
-b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
-c) Number of methods which can handle a particular type of trajectory."
-)
-```
-
-```{r, results='asis' }
-add_table(
-  "implementations",
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof.
-This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
-)
-```
 
@@
 
-<!-- §4: Prior information -->
-Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
-Prior information can be supplied as  a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
-Providing prior information to a TI method can be both a blessing and a curse.
-In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
-On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
-Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.
-
-<!-- §5: Components -->
-A reductionist approach to characterising TI methods consists in dissecting them into a set of algorithmic components, as any component can have a significant impact on the performance, scalability, and output data structures.
-Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-```{r}
-add_sfig(
-  figure_file("implementation_components.svg","4-method_characterisation"), 
-  "implementation_components", 
-  "Common algorithmic components shared by different TI methods.
-Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
-a) Wordcloud of the common components.
-b) Number of times a component was shared between methods.
-c) Number of methods containing a particular category."
-)
-```
-
-<!-- §6: Problem statement -->
-Given the  diversity in TI methods, an important issue to address is a quantitative assessment of the performance and robustness of the existing TI methods.
-Many attempts at tackling  this issue have already been made[@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but due to the high number of TI methods available today and the great diversity in the outputted data structures, a comprehensive benchmarking evaluation of TI methods is still lacking.
-This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would be the most optimal for their problem. 
+<!-- §3: Problem statement -->
+Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness..
+Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods is still lacking.
+This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem.
 Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
@@
 
-<!-- §2: Overall and topology types -->
-Having ordered all methods by their overall benchmarking score, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree.
-When we looked at the benchmark scores per trajectory type, Slingshot was the only method that performed well across most trajectory types.
+<!-- §1: Overall performance -->
+Across all datasets and evaluation metrics, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, slingshot was the only method that performed well across most trajectory types, followed by TSCAN and Monocle DDRTree.
 However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for trees.
+This was also reflected in the rankings of different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, performing better at our metric investigating topological similarity, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("sfig", "metric_comparison")`).
 
-<!-- §3: real versus synthetic -->
-We observed a high correlation (0.7-0.9) between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`).
-This confirms both the relevance of the synthetic data and the accuracy of the gold standard in the real datasets.
-However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available for such topologies (`r ref("stable", "datasets")`).
+<!-- §2: The stability and robustness of the evaluation/method ranking -->
+```{r}
+correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
+filter(n_tasks == round(max(n_tasks)/10)) %>% 
+          pull(cor) %>% 
+          mean()
+top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
+filter(method_id == "slngsht") %>% 
+filter(n_tasks == round(max(n_tasks)/10)) %>%
+pull(rank_match)
+random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
+filter(method_id == "scoup") %>% 
+pull(max_n_tasks_top)
+
+```
+
+We further investigated the robustness of our evaluation across datasets and dataset sources.
+Method performance was moderately to highly correlated (0.5-0.9) between real and synthetic datasets (`r ref("fig", "results_overview", "b")` and `r ref("sfig", "real_synthetic_comparison")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
+Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with most methods significantly outperforming the next method in the ranking `r ref("sfig", "performance_dataset_variability_combined")`.
+To further illustrate this, if we looked at the scores a random 10% subset of all datasets, the overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
+On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
+For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top`datasets on which it would be the top ranking method.
+Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
+```{r}
+add_sfig(
+  figure_file(
+    "real_synthetic_comparison.svg",
+    "11-evaluation_robustness"
+  ), 
+  "real_synthetic_comparison", 
+  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
+)
+```
 
@@
   figure_file(
-    "real_synthetic_comparison.svg",
-    "5-optimise_parameters/5-real_synthetic_comparison"
+    "performance_dataset_variability_combined.svg",
+    "11-evaluation_robustness"
   ), 
-  "real_synthetic_comparison", 
-  "Comparison between performance on real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
+  "performance_dataset_variability_combined", 
+  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
 )
 ```
-
-<!-- §4: Compare metrics -->
-As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics.
-While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtained a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`).
-In addition, the methods that were able to detect more complex trajectory types also obtained higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology.
-We will explore this issue in more detail in a further section.
 
@@
 # Discussion
-<!-- The context -->
-Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
-In this preprint, we present an  initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using the default parameters.
-When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art (`r ref("sfig", "leading_method")`).
-We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
-
-```{r}
-add_sfig(
-  figure_file('leading_method.svg', "9-main_figure"), 
-  "leading_method", 
-  "Method performance compared to the time the method was published. The red line indicates the optimal performance up to a point in time."
-)
-```
-
-<!-- Future work -->
-In this study, we presented our first version of the evaluation of TI methods.
-We are convinced that the results we provided will be useful for both method users and tool developers, as we provide clear practical guidelines for users depending on their current knowledge of the trajectory, as well as an objective benchmark on which new methods can be tested.
-Nonetheless, our evaluation can be expanded on several points, all of which we will try to tackle in the near future:
-
-* Inclusion of methods published during of after june 2017
-* A parameter tuning for each method, both across all trajectory types, as well as on specific trajectory types.
-* A test of robustness on noise, parameter changes and dataset size
-* An evaluation of the stability of a method's results when running the same parameter setting on the same dataset multiple times
-* Evaluate the methods when given optional priors, and compare with performance without priors
-* Assess the effect of noisy or incorrect prior information
-* Test methods on datasets with no clear trajectory present
-* Include feedback on the quality control scoring scheme, and update the scoring when method get updated
-* Test the scalability of each method, both in the number of cells and the number of features
-* Include more real datasets with complex trajectory types, as they become available
-* Evaluate methods on other single-cell -omics datasets, such as proteomics and epigenomics data
-* Provide functionality for visually interpreting and comparing predicted trajectories
-
-<!-- Are certain methods better than others? -->
-Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, TSCAN, and Monocle DDRTree, towering above all other methods.
-Nonetheless, we found that methods which did not perform well across all settings could still be useful in certain specific use cases.
-Indeed, on data containing more than one bifurcations, Monocle DDRTree clearly performed better than other methods.
-We found that this particular result was mainly caused by the fact that the default parameters of Monocle DDRTree preferably led to the detection of tree topologies, while those of Slingshot preferably found linear and bifurcation trajectories.
-
-<!-- Our common output model, advantages & limitations -->
-We managed to wrap the output of all methods into one common format.
-This not only allowed us to compare different methods with a gold standard, but could also be useful for TI users, as  it allows the user to test multiple methods on the same data and compare the results without manual conversion of input and output.
-Furthermore, it makes it possible to directly compare the output of different methods, which opens up possibilities for new comparative visualisation techniques or ensemble methods.
-However, we acknowledge that our model has some limitations.
-Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
-when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
-when the connections between branches are uncertain).
-Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
-
-<!-- Synthetic data: observations and possibilities -->
-The use of synthetic data to evaluate TI methods offers the advantage of having an exact gold standard to which the methods’ results can be compared.
-However, the use of synthetic data can also be questionable, because the model which generates the data does not necessarily reflect the intrinsic characteristics of true biological systems.
-This could bias an evaluation towards methods for which the underlying model best fits the model used for generating the data.
-Therefore, it is essential that results of an evaluation on synthetic data are confirmed using real data.
-In our study, the overall performance of methods was very similar between real and synthetic datasets, confirming the biological relevance of the synthetic data.
-Given this, we believe that our synthetic data can now be used to effectively prototype new TI methods for more complex trajectories such as disconnected graphs, for which the availability of real datasets is poor.
-Furthermore, it is expected that in the future, methods will be able to model even more complex cellular behaviors, such as multiple dynamic processes happening at parallel in a single cell, the integration of datasets from different patients or trajectories in a spatial context [@tanay_scaling_2017; @cannoodt_computational_2016].
-Synthetic data generated with  our workflow could therefore be used to spearhead the development of these methods, given that currently only a limited number of real datasets are available for which the methods could be useful.
-Furthermore, we believe that our data generation workflow could also be used to evaluate other types of single-cell modelling techniques, such as single-cell network inference, clustering and normalisation.
-We sincerely hope that such efforts will lead to a more rapid development of accurate methods, and will in the near future provide a package which can be used to simulate synthetic data for a wide variety of single-cell modelling problems.
 
@@
 This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).
+
+<!-- PASTED FROM INTRODUCTION, need to rework -->
+<!-- §4: Prior information -->
+Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
+Prior information can be supplied as  a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
+Providing prior information to a TI method can be both a blessing and a curse.
+In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
+On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.
+
+<!-- §5: Components -->
+A reductionist approach to characterising TI methods consists in dissecting them into a set of algorithmic components, as any component can have a significant impact on the performance, scalability, and output data structures.
+Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
+
+```{r}
+add_sfig(
+  figure_file("implementation_components.svg","4-method_characterisation"), 
+  "implementation_components", 
+  "Common algorithmic components shared by different TI methods.
+Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
+a) Wordcloud of the common components.
+b) Number of times a component was shared between methods.
+c) Number of methods containing a particular category."
+)
+```
 
@@
 # Supplementary Material
+```{r}
+add_sfig(
+  figure_file("methods_timeline.svg", "4-method_characterisation"),
+  "methods_over_time", 
+  "New TI methods over time.
+a) Number of methods published or in preprint.
+b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
+c) Number of methods which can handle a particular type of trajectory."
+)
+```
+
+```{r, results='asis' }
+add_stable(
+  "implementations",
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
+  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof.
+This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
+)
+```
+
 ```{r "supplementary", echo=FALSE, results='asis'}

### 2018-03-22 analysis/paper/supp_notes/supp_note_trajectory_inference_methods.Rmd
add supp note for acos

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
bibliography: ../references.bib
csl: ../nature-biotechnology.csl
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage[normalem]{ulem}
  - \usepackage{lscape}
  - \usepackage{pbox}
  - \usepackage[labelformat=empty]{caption}
  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
mainfont: Open Sans
output: 
  pdf_document: 
    latex_engine: xelatex
    dev: png
    keep_tex: true
    citation_package: biblatex
---

*`r format(Sys.time(), '%d %B, %Y')`*

```{r setup, include=FALSE}
# knitr options
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/supp_notes/.scratch/")
)

# load libraries
library(dynalysis)
library(dynplot)
library(tidyverse)
library(cowplot)
```
Wouter Saelens* ¹ ², Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²

\* Equal contribution  
¹ Data mining and Modelling for Biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France

# Supplementary Note: Trajectory Inference Wrapppers

## `acos`: Arc cosine
This is a control method for inferring cyclic trajectories in a very simple approach,
consisting of a dimensionality reduction step and the computation of the arc cosine.

### Parameters
There is only one parameter, namely the dimensionality reduction method used.
```{r}
par_set <- description_acos()$par_set
par_set
```

The default parameters are the following:
```{r}
defprm <- ParamHelpers::dfRowToList(ParamHelpers::generateDesignOfDefaults(par_set, trafo = TRUE), par_set, 1)
defprm
```

### Example dataset
We will be using a cyclic toy dataset to demonstrate this method with.

```{r}
toy <- dyntoy::toy_tasks %>% filter(model == "cycle") %>% extract_row_to_list(row_id = 2)
expression <- toy$expression

plot_default(toy)
```

### Wrapper
In the first step, a dimensionality reduction to two dimensions is computed. By default, this is PCA. 
Here, it is assumed that the dimensionality reduction will be centered around the origin 0.

```{r}
dimred <- dynmethods:::dimred(expression, method = defprm$dimred, ndim = 2)

# Plot dimred
dimred_df <- data.frame(dimred) %>% rownames_to_column("cell_id")
ggplot(dimred_df) + geom_point(aes(Comp1, Comp2)) + coord_equal()

```

The ordering of the cells is computed by calculating the bearing of each cell with respect to the origin. The bearing is transformed to a [0,1] range.

```{r}
pseudotimes <- atan2(dimred[,2], dimred[,1]) / 2 / pi + .5

dimred_df <- dimred_df %>% mutate(pseudotime = pseudotimes[cell_id])
ggplot(dimred_df) + geom_point(aes(Comp1, Comp2, colour = pseudotime)) + coord_equal() + viridis::scale_color_viridis()
```

The prediction is wrapped using the `add_cycling_trajectory_to_wrapper` function. 
```{r}
prediction <- wrap_prediction_model(
  cell_ids = rownames(expression)
) %>% add_cyclic_trajectory_to_wrapper(
  pseudotimes = pseudotimes,
  do_scale_minmax = FALSE
) %>% add_dimred_to_wrapper(
  dimred = dimred
)

plot_default(prediction)
```

The figure below shows a comparison of the orderings in the toy versus the prediction.
```{r}
plot_strip_connections(toy, prediction)
```

# Colophon
This report was generated on `r Sys.time()` using `r devtools:::platform_info()$version` and the following packages:
```{r colophon, cache = FALSE}
devtools::session_info()$packages %>% knitr::kable()
```

### 2018-03-22 analysis/paper/supp_notes/supp_note_trajectory_inference_methods.Rmd
update results

@@
     latex_engine: xelatex
-    dev: png
     keep_tex: true

### 2018-03-22 analysis/paper/supp_notes/supp_note_trajectory_inference_methods.Rmd
update

@@
 ---
-title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
+title: "Supplementary Note: Trajectory Inference Wrappers"
 bibliography: ../references.bib
@@
   pdf_document: 
+    citation_package: biblatex
+    fig_height: 6
+    fig_width: 10
+    keep_tex: yes
     latex_engine: xelatex
-    keep_tex: true
-    citation_package: biblatex
+    toc: yes
+editor_options: 
+  chunk_output_type: console
 ---
-
-*`r format(Sys.time(), '%d %B, %Y')`*
 
@@
   message = FALSE,
+  out.width = "0.5\\textwidth",
+  fig.align = "center",
   fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/supp_notes/.scratch/")
@@
 ```
-Wouter Saelens* ¹ ², Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
 
-\* Equal contribution  
-¹ Data mining and Modelling for Biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
-² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
-³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
-⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
+# `aga`: AGA
+<insert description AGA>
 
-# Supplementary Note: Trajectory Inference Wrapppers
+## Parameters
 
-## `acos`: Arc cosine
-This is a control method for inferring cyclic trajectories in a very simple approach,
-consisting of a dimensionality reduction step and the computation of the arc cosine.
-
-### Parameters
-There is only one parameter, namely the dimensionality reduction method used.
 ```{r}
-par_set <- description_acos()$par_set
+method <- description_aga()
+par_set <- method$par_set
 par_set
@@
 ```{r}
-defprm <- ParamHelpers::dfRowToList(ParamHelpers::generateDesignOfDefaults(par_set, trafo = TRUE), par_set, 1)
+defprm <- ParamHelpers::dfRowToList(
+  ParamHelpers::generateDesignOfDefaults(par_set, trafo = TRUE),
+  par_set, 
+  i = 1
+)
 defprm
@@
 
-### Example dataset
+## Example dataset
+We will be using a consecutive bifurcating toy dataset to demonstrate this method with.
+
+```{r}
+toy <- dyntoy::toy_tasks %>% filter(model == "consecutive_bifurcating") %>% extract_row_to_list(row_id = 1)
+expression <- toy$expression
+
+plot_default(toy)
+```
+
+## Wrapper
+
+Since AGA is implemented in Python, we created an R package for it. 
+```{r}
+aga_out <- aga::aga(
+  counts = expression,
+  start_cell = NULL,
+  grouping_assignment = NULL,
+  n_neighbours = defprm$n_neighbours,
+  n_pcs = defprm$n_pcs,
+  n_dcs = defprm$n_dcs,
+  resolution = defprm$resolution,
+  tree_based_confidence = defprm$tree_based_confidence,
+  verbose = FALSE,
+  num_cores = 1
+)
+```
+
+<Talk about aga_out$adj>
+```{r}
+aga_out$adj %>% head(6) %>% knitr::kable()
+```
+
+<Talk about aga_out$obs>
+```{r}
+aga_out$obs %>% head(6) %>% knitr::kable()
+```
+
+Reformat everything into milestone_network and milestone_percentages
+```{r}
+cell_ids <- rownames(expression)
+
+milestone_assignment_cells <- setNames(aga_out$obs$group_id, aga_out$obs$cell_id)
+
+milestone_network <- aga_out$adj %>%
+  mutate_at(vars(from, to), as.character) %>%
+  filter(aga_adjacency_tree_confidence > 0) %>%
+  mutate(
+    length = 1,
+    directed = TRUE
+  ) %>%
+  select(from, to, length, directed)
+
+milestone_ids <- sort(unique(milestone_assignment_cells))
+```
+
+The prediction is wrapped using the `add_trajectory_to_wrapper` function. 
+```{r}
+prediction <- wrap_prediction_model(
+  cell_ids = rownames(expression)
+) %>% add_cluster_graph_to_wrapper(
+  milestone_ids = milestone_ids,
+  milestone_network = milestone_network,
+  milestone_assignment_cells = milestone_assignment_cells,
+  aga_out = aga_out
+)
+
+plot_default(prediction)
+method$plot_fun(prediction)
+plot_strip_connections(toy, prediction)
+```
+
+# `atan`: Arc-tangent
+This is a control method for inferring cyclic trajectories in a very simple approach,
+consisting of a dimensionality reduction step and followed by calculating the arc-tangent
+between the first and the second component.
+
+## Parameters
+There is only one parameter, namely the dimensionality reduction method used.
+```{r}
+method <- description_atan()
+par_set <- method$par_set
+par_set
+```
+
+The default parameters are the following:
+```{r}
+defprm <- ParamHelpers::dfRowToList(
+  ParamHelpers::generateDesignOfDefaults(par_set, trafo = TRUE),
+  par_set, 
+  i = 1
+)
+defprm
+```
+
+## Example dataset
 We will be using a cyclic toy dataset to demonstrate this method with.
@@
 
-### Wrapper
+## Wrapper
 In the first step, a dimensionality reduction to two dimensions is computed. By default, this is PCA. 
@@
 dimred_df <- dimred_df %>% mutate(pseudotime = pseudotimes[cell_id])
-ggplot(dimred_df) + geom_point(aes(Comp1, Comp2, colour = pseudotime)) + coord_equal() + viridis::scale_color_viridis()
+ggplot(dimred_df) +
+  geom_point(aes(Comp1, Comp2, colour = pseudotime)) +
+  coord_equal() + 
+  viridis::scale_color_viridis()
 ```
@@
 plot_default(prediction)
-```
-
-The figure below shows a comparison of the orderings in the toy versus the prediction.
-```{r}
+method$plot_fun(prediction)
 plot_strip_connections(toy, prediction)

### 2018-03-23 analysis/paper/supp_notes/supp_note_trajectory_inference_methods.Rmd
move aga supp note information

@@
 
-# load libraries
-library(dynalysis)
-library(dynplot)
-library(tidyverse)
-library(cowplot)
+source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
 ```
 
-# `aga`: AGA
-<insert description AGA>
+This document serves to provide insight into how each trajectory inference method was used and wrapped.
+
+```{r}
+library(dynmethods)
+library(dynwrap)
+library(dynutils)
+library(dynplot)
+```
+
+<!-- ################################################## aga ################################################## -->
+See [https://github.com/dynverse/dynmethods/blob/master/vignettes/aga.Rmd](dynmethods).
+
+<!-- ################################################## agapt ################################################## -->
+
+# `agapt`: AGA pseudotime
+TODO: move to dynmethods.
+
+When a start cell is given, AGA supports ordering the cells along branches.
 
 ## Parameters
-
 ```{r}
-method <- description_aga()
-par_set <- method$par_set
-par_set
-```
-
-The default parameters are the following:
-```{r}
-defprm <- ParamHelpers::dfRowToList(
-  ParamHelpers::generateDesignOfDefaults(par_set, trafo = TRUE),
-  par_set, 
-  i = 1
-)
-defprm
+method <- description_agapt()
+method$par_set
 ```
@@
 ## Example dataset
-We will be using a consecutive bifurcating toy dataset to demonstrate this method with.
+We will be using the same consecutive bifurcating toy dataset.
 
 ```{r}
-toy <- dyntoy::toy_tasks %>% filter(model == "consecutive_bifurcating") %>% extract_row_to_list(row_id = 1)
+toy <- dyntoy::toy_tasks %>% 
+  filter(model == "consecutive_bifurcating") %>% 
+  extract_row_to_list(row_id = 1)
 expression <- toy$expression
+start_cells <- toy$prior_information$start_cells
 
@@
 
-Since AGA is implemented in Python, we created an R package for it. 
+Since AGA is implemented in Python, we created an R package for it, available at [github.com/dynverse/aga](https://github.com/dynverse/aga).
+The R package will write the expression values and any parameters to a file, execute a python script, and read the results from output files produced by AGA.
 ```{r}
+defprm <- get_default_parameters(method)
+
+# sample one start cell
+start_cell <- sample(start_cells, 1)
+
 aga_out <- aga::aga(
-  counts = expression,
-  start_cell = NULL,
+  expression = expression,
+  start_cell = start_cell,
   grouping_assignment = NULL,
@@
   tree_based_confidence = defprm$tree_based_confidence,
-  verbose = FALSE,
+  verbose = TRUE,
   num_cores = 1
@@
 
-<Talk about aga_out$adj>
+After building a kNN graph (not shown), the kNN graph is clustered into louvain groups:
+```{r}
+aga_out$obs %>% head(6) %>% knitr::kable()
+
+# cell_ids <- rownames(expression)
+# milestone_assignment_cells <- setNames(aga_out$obs$group_id, aga_out$obs$cell_id)
+```
+
+Several tests are used to assess which transitions exist between the louvain groups. 
+We use the `aga_adjacency_tree_confidence` to construct the milestone network.
 ```{r}
@@
 
-<Talk about aga_out$obs>
-```{r}
-aga_out$obs %>% head(6) %>% knitr::kable()
-```
-
-Reformat everything into milestone_network and milestone_percentages
-```{r}
-cell_ids <- rownames(expression)
-
-milestone_assignment_cells <- setNames(aga_out$obs$group_id, aga_out$obs$cell_id)
-
-milestone_network <- aga_out$adj %>%
-  mutate_at(vars(from, to), as.character) %>%
-  filter(aga_adjacency_tree_confidence > 0) %>%
-  mutate(
-    length = 1,
-    directed = TRUE
-  ) %>%
-  select(from, to, length, directed)
-
-milestone_ids <- sort(unique(milestone_assignment_cells))
-```
-
-The prediction is wrapped using the `add_trajectory_to_wrapper` function. 
-```{r}
-prediction <- wrap_prediction_model(
-  cell_ids = rownames(expression)
-) %>% add_cluster_graph_to_wrapper(
-  milestone_ids = milestone_ids,
-  milestone_network = milestone_network,
-  milestone_assignment_cells = milestone_assignment_cells,
-  aga_out = aga_out
-)
-
-plot_default(prediction)
-method$plot_fun(prediction)
-plot_strip_connections(toy, prediction)
-```
+<!-- ################################################## atan ################################################## -->
 
 # `atan`: Arc-tangent
+TODO: move to dynmethods.
+
 This is a control method for inferring cyclic trajectories in a very simple approach,
@@
 ## Parameters
-There is only one parameter, namely the dimensionality reduction method used.
 ```{r}
 method <- description_atan()
-par_set <- method$par_set
-par_set
-```
-
-The default parameters are the following:
-```{r}
-defprm <- ParamHelpers::dfRowToList(
-  ParamHelpers::generateDesignOfDefaults(par_set, trafo = TRUE),
-  par_set, 
-  i = 1
-)
-defprm
+method$par_set
 ```
@@
 ```{r}
-toy <- dyntoy::toy_tasks %>% filter(model == "cycle") %>% extract_row_to_list(row_id = 2)
+toy <- dyntoy::toy_tasks %>% 
+  filter(model == "cycle") %>% 
+  extract_row_to_list(row_id = 2)
 expression <- toy$expression
@@
 ```{r}
+defprm <- get_default_parameters(method)
+
 dimred <- dynmethods:::dimred(expression, method = defprm$dimred, ndim = 2)

### 2018-03-27 analysis/paper/paper_latex.Rmd

@@
 In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness.
-Our results indicate that Slingshot is the most accurate method to date, but other leading methods (e.g.
+Our results indicate that slingshot is the most accurate method to date, but other leading methods (e.g.
 TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases.
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be computationally modelled by ordering cells along a trajectory, also called trajectory inference (TI) or pseudotemporal ordering [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+Such dynamic processes can be modelled computationally using using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs.
@@
 <!-- §2: Plethora of new TI methods -->
-A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`).
-All methods have their own unique set of characteristics (`r ref("stable", "implementations")`) in terms of underlying algorithm, the prior information they require and the outputs they produce.
+A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`).
+All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
 Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect.
@@
 Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases.
-Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).
-A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is the only TI method currently able to deal with disconnected graphs.
+Methods that perform topology inference are still in the minority, though current trends suggest this will soon change.
+A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.
 
 <!-- §3: Problem statement -->
-Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness..
-Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods is still lacking.
+Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness.
+Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation, which compares a large set of methods against a large set of datasets, of TI methods is still lacking.
 This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem.
@@
 
-In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`, and `r ref("sfig", "extended_evaluation_overview")`).
-The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the date of publication  (`r ref("table", "implementations")`).
-Only methods published before june 2017 are included in the current version of the evaluation.
-The evaluation comprised three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
-
-In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
-In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
-The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets (`r ref("sfig", "synthetic_data_generation")`).
-
-```{r}
-add_sfig(
-  figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
-  "dataset_characterisation",  
-  "Characteristics of the real datasets used for the evaluation.
-Top: distributions of number of cells and genes (after filtering) and the date these datasets were published.
-Bottom: Diversity of technologies which were used to generate the dataset, the organism, whether we extracted a silver or a gold standard from the data, the type of dynamic process and the trajectory type present in the data."
-)
-```
-
-```{r}
-add_sfig(
-  figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
-  "synthetic_data_generation",
-  "Workflow to generate the synthetic data.
-Starting from a state network, we extracted a regulatory network using real networks which is expected to induce such a set of state transitions when simulated.
-We simulated this network using a detailed model of gene regulation which includes the thermodynamics of transcription factor binding.
-Finally, we simulated the single-cell RNA-seq (scRNA-seq) experiment itself, by matching the distribution of expression and dropouts to a real reference dataset."
-)
-```
-
-Finally, using both the benchmark results and quality control, we produced a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`).
+The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`).
+The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.
 
@@
 
-----
+## Evaluation of trajectory inference methods
 
-```{r}
-add_sfig(
-  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
-  "extended_evaluation_overview",
-  "Extended overview of our evaluation pipeline.
-From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
-We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
-Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
-We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
-  15, 
-  7.5
-)
-```
-
-## Evaluation of trajectory inference methods
+<!-- Evaluation overview -->
+In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
+To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets.
+<!-- Overall evaluation results -->
+Across all datasets and evaluation metrics, we found that slingshot, TSCAN and Monocle DDRTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returns the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees.
+This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
 
@@
 
--------
+----------
 
-<!-- §1: Overall performance -->
-Across all datasets and evaluation metrics, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview", "b")`).
-When we focused on the benchmark scores per trajectory type, slingshot was the only method that performed well across most trajectory types, followed by TSCAN and Monocle DDRTree.
-However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for trees.
-This was also reflected in the rankings of different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, performing better at our metric investigating topological similarity, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("sfig", "metric_comparison")`).
+<!-- Errors-->
+Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
+Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
+Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
+For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
 
-<!-- §2: The stability and robustness of the evaluation/method ranking -->
+<!-- Evaluation robustness -->
 ```{r}
@@
 We further investigated the robustness of our evaluation across datasets and dataset sources.
-Method performance was moderately to highly correlated (0.5-0.9) between real and synthetic datasets (`r ref("fig", "results_overview", "b")` and `r ref("sfig", "real_synthetic_comparison")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
-Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with most methods significantly outperforming the next method in the ranking `r ref("sfig", "performance_dataset_variability_combined")`.
-To further illustrate this, if we looked at the scores a random 10% subset of all datasets, the overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
+Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
+Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
+To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets.
+The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
 On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
-For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top`datasets on which it would be the top ranking method.
+For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
 Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
-```{r}
-add_sfig(
-  figure_file(
-    "real_synthetic_comparison.tmp.pdf",
-    "11-evaluation_robustness"
-  ), 
-  "real_synthetic_comparison", 
-  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
-)
-```
 
-```{r}
-add_sfig(
-  figure_file(
-    "performance_dataset_variability_combined.tmp.pdf",
-    "11-evaluation_robustness"
-  ), 
-  "performance_dataset_variability_combined", 
-  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
-)
-```
+## Relationships between method characteristics to performance
 
-```{r}
-add_sfig(
-  figure_file('metrics_comparison.tmp.pdf', "5-optimise_parameters/7-metric_comparison"), 
-  "metric_comparison", 
-  "Comparison of method performance between different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
-)
-```
-
-### Trajectory types and topology complexity
-
-During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
-If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
-Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
-Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
-For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
-
-### Trajectory types and topology complexity
-
-In most cases, the methods which were specifically designed to handle a particular trajectory type, also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
-These methods had typically better edge flip scores - as can be expected - and RF MSE scores, compared to methods not able to handle the particular trajectory type.
-However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-To further investigate the effect of trajectory complexity on performance, we divided them in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
-While there were no significant differences in performance on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`).
+<!-- Inverse relation between edge flip and correlation -->
+In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
+However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
+To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods.
+While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`).
 Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
@@
 
-```{r}
-add_sfig(
-  figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology"), 
-  "trajtype_handle_comparison", 
-  "Comparison between the performance of methods able to handle a trajectory type with those who don't.
-Shown is the performance on those datasets containing a particular trajectory type (top), according to several metrics (right)."
-)
-```
-
-```{r}
-add_sfig(
-  figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology"), 
-  "linear_split_comparison", 
-  "Comparison of the performance between linear and non-linear methods.
-Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
-Their performance is shown on linear and non-linear datasets according to several performance metrics (right).
-NS: Not significant, \\*: p-value < 0.05, \\*\\*: p-value < 0.001, \\*\\*\\*: p-value < 0.0001"
-)
-```
-
+<!-- Topological complexity versus performance -->
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected using the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree, for example frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
-Trajectories generated by Monocle DDRTree (at default parameters) tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
-Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
-Inferring the correct topology without parameter tuning, is therefore still an open challenge.
-Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
+Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
+Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases.
+Inferring the correct topology without parameter tuning is therefore still an open challenge.
+Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
 
@@
 
-----
-
-```{r}
-add_sfig(
-  figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology"), 
-  "topology_sensitivity", 
-  "Sensitivity of each TI method in predicting the correct trajectory topology.
-Shown are the % of datasets where a method is able to predict the exact correct trajectory topology, depending on the trajectory type present in the data.
-When a method is able to predict a particular trajectory type, the box is surrounded by a black border."
-)
-```
-
-### Effect of prior information
-
-In the current version of the evaluation, we only provided prior information when a method required it.
-We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`).
-Rather, methods which received prior information were on average positioned in the middle of the ranking.
-Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.
-
-```{r}
-add_sfig(
-  figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information"), 
-  "required_priors_performance_comparison", 
-  "Comparison of method performance depending on whether the method required a certain prior or not. The top performing method which requires a particular type of prior information is labelled."
-)
-```
-
-### Algorithm components
-
-We assessed whether the components of an algorithm could  be predictive of a method’s performance, using both random forest classification and statistical testing.
-Methods which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+<!-- Additional factors which could influence performance -->
+Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method.
+Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
+When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`).
 On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
-
-```{r}
-add_sfig(
-  figure_file("method_components_scores.tmp.pdf","6-developer_guidelines"), 
-  "method_components_scores", 
-  "Score of methods containing particular algorithmic components.
-Shown are the algorithmic components present in four or more methods, and whether the performance of methods containing a certain component was significantly higher than all the other methods.
-NS: Not significant.
-*: corrected p-value < 0.05 "
-)
-```
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "component_importance.rds",
-    "6-developer_guidelines"
-  ))[[params$table_format]], 
-  "component_importance", 
-  "Importance values for algorithmic components.
-The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
-)
-```
 
@@
 While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
-To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review.
+To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_checks")`).
 This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
 
-We therefore looked at the implementation of each method, and assessed its  quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`).
-The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, method's behaviour at runtime and the depth by which the method was presented in its study) or which purpose(s) they serve (user friendliness, developer friendliness or future proof).
-These categorisations can help current developers to improve their tool, and guide the selection of users and developers to use these tools for their purpose.
-After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "qc_checks.rds",
-    "4-method_characterisation"
-  ))[[params$table_format]], 
-  "qc_checks", 
-  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
-)
-```
-
 We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`).
-We found that recent methods had a slightly better quality than older methods (`r ref("sfig", "qc_over_time")`).
-However, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
+While recent methods had a slightly better quality than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
 Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
@@
 
-----
-
-```{r}
-add_sfig(
-  figure_file("qc_over_time.tmp.pdf","4-method_characterisation"), 
-  "qc_over_time", 
-  "Quality control scores and number of citations over time."
-)
-```
-We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`).
+We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")`).
 We could also not find any quality aspect which was significantly associated with method performance.
-
-```{r}
-add_sfig(
-  figure_file(
-    "qc_performance_comparison.tmp.pdf",
-    "5-optimise_parameters/6-qc_performance_comparison"
-  ), 
-  "qc_performance_comparison", 
-  "Comparison between method quality control scores and overall method performance."
-)
-```
 
@@
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and Slingshot performing equally well on less complex trajectories .
-No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories.
-In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although Slingshot could return other topologies if it would fit the data more accurately.
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and slingshot performing equally well on less complex trajectories .
+No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")`.
+should be able to handle graph trajectories or even disconnected trajectories .
+In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, slingshot and cellTree for linear trajectories, and slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
@@
 
-----
-
 When choosing a method, it is important to take two further points into account.
@@
 
-# Methods
+<!-- The context -->
+Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
+Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
+It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
+In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
+When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art.
+We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
+
+# Online Methods
 
@@
 For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-```{r}
-add_sfig(
-  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
-  "implementation_components", 
-  "Common algorithmic components shared by different TI methods.
-Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
-a) Wordcloud of the common components.
-b) Number of times a component was shared between methods.
-c) Number of methods containing a particular category."
-)
-```
 
@@
 Per region, one milestone must be directly connected to all other milestones in the network.
-
-```{r}
-add_sfig(
-  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
-  "wrapping", 
-  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
-As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
-This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
-  15, 
-  7.5
-)
-```
 
@@
 
+#### cellTree
+
 ## Real datasets
-We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")` and `r ref("sfig", "dataset_characterisation")`).
+We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
 The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
@@
 
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "synthetic_data_generation")`).
+Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
 At every step, we took great care to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
@@
 
-To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`).
+To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
 Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple undirected graphs (bifurcating loop and bifurcating convergence).
@@
 To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.
-
-```{r}
-add_sfig(
-  figure_file("modulenets.tmp.pdf", experiment_id = "2-dataset_characterisation/3-synthetic"),
-  "modulenets", 
-  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
-  15, 
-  5
-)
-```
 
@@
 add_sfig(
-  figure_file("score_aspects_v1.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
+  figure_file("score_aspects_v2.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
   "score_aspects", 
@@
 Each execution of a method on a dataset was performed in a separate task as part of a gridengine job.
-Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and was provided a maximum of 32GB in memory and 6 hours of wall time.
-One R session was started for each task, with the environment variable `R_MAX_NUM_DLLS` set to 500.
+Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task.
+During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
+If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
+The environment variable `R_MAX_NUM_DLLS` set to 500.
 The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust.
@@
 We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
-These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical (`r ref("sfig", "trajectory_type_tree", "a")`).
-For every directed trajectory type, a corresponding undirected trajectory type can also be defined (`r ref("sfig", "trajectory_type_tree", "b")`).
+These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical.
+A disconnected directed graph is a graph in which only one edge can exist between two nodes.
+A directed graph is a disconnected graph in which all nodes are (indirectly) connected.
+A directed acyclic graph is a graph containing no cycles.
+A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1).
+A rooted binary tree is a rooted tree with every node having maximally two outgoing edges.
+A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one.
+A multifurcation is a rooted tree in which only one node has a degree larger than one.
+A bifurcation is a multifurcation in which only one node has a degree equal to 3.
+A directed linear graph is a graph in which no node has a degree larger than 3.
+Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle.
+For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork.
 In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).
+
+# Supplementary Material
 
@@
 add_sfig(
-  figure_file("trajectory_type_trees.tmp.pdf","2-dataset_characterisation/2-trajectory_types"), 
-  "trajectory_type_tree", 
-  "Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types.
-a) The graph structure for undirected and directed trajectory types.
-Shown are the topological properties which change when going from a particular trajectory type to a less complex type.
-b) Combined structure for directed and undirected trajectory types.",
+  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
+  "extended_evaluation_overview",
+  "Extended overview of our evaluation pipeline.
+From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
+We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
+Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
+We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
   15, 
-  10
+  7.5
 )
@@
 
-# Supplementary Material
 ```{r}
 add_sfig(
-  figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
-  "methods_over_time", 
-  "New TI methods over time.
-a) Number of methods published or in preprint.
-b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
-c) Number of methods which can handle a particular type of trajectory."
+  figure_file(
+    "robustness.tmp.pdf",
+    "11-evaluation_robustness"
+  ), 
+  "robustness_comparison", 
+  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
+Given in the top-left corner of each panel is the Pearson's correlation.
+b) Violin plots of the overall performance.
+For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
+)
+```
+
+```{r}
+add_sfig(
+  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
+  "wrapping", 
+  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
+  15, 
+  7.5
 )
@@
   read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof.
-This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
+  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
 )
@@
 
-```{r "supplementary", echo=FALSE, results='asis'}
-source("analysis/paper/sfigs_stables.R")
+```{r}
+add_stable(
+  read_rds(figure_file(
+    "component_importance.rds",
+    "6-developer_guidelines"
+  ))[[params$table_format]], 
+  "component_importance", 
+  "Importance values for algorithmic components.
+The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
+)
 ```
 
-# Colophon
+```{r}
+add_stable(
+  read_rds(figure_file(
+    "qc_checks.rds",
+    "4-method_characterisation"
+  ))[[params$table_format]], 
+  "qc_checks", 
+  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
+)
+```
 
-This report was generated on `r Sys.time()` using `r devtools:::platform_info()$version` and the following packages:
-```{r colophon, cache = FALSE}
-devtools::session_info()$packages %>% knitr::kable()
-```
+1) method performance highly depends on trajectory type
+2) methods which are specific to a certain trajectory type generally perform better on datasets containing these trajectory types
+3) the complexity of the topologies returned by the different methods is highly variable 
+4) prior information doesn't increase performance in our settings

### 2018-03-27 analysis/paper/paper.Rmd

@@
 
-In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`, and `r ref("sfig", "extended_evaluation_overview")`).
-The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the date of publication  (`r ref("table", "implementations")`).
-Only methods published before june 2017 are included in the current version of the evaluation.
-The evaluation comprised three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.
-
-In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
-In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
-The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets (`r ref("sfig", "synthetic_data_generation")`).
-
-```{r}
-add_sfig(
-  figure_file('dataset_characterisation.svg', '2-dataset_characterisation'),
-  "dataset_characterisation",  
-  "Characteristics of the real datasets used for the evaluation.
-Top: distributions of number of cells and genes (after filtering) and the date these datasets were published.
-Bottom: Diversity of technologies which were used to generate the dataset, the organism, whether we extracted a silver or a gold standard from the data, the type of dynamic process and the trajectory type present in the data."
-)
-```
-
-```{r}
-add_sfig(
-  figure_file('synthetic_data_generation.svg', "2-dataset_characterisation/3-synthetic"),
-  "synthetic_data_generation",
-  "Workflow to generate the synthetic data.
-Starting from a state network, we extracted a regulatory network using real networks which is expected to induce such a set of state transitions when simulated.
-We simulated this network using a detailed model of gene regulation which includes the thermodynamics of transcription factor binding.
-Finally, we simulated the single-cell RNA-seq (scRNA-seq) experiment itself, by matching the distribution of expression and dropouts to a real reference dataset."
-)
-```
-
-Finally, using both the benchmark results and quality control, we produced a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.
+In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`).
+The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`).
+The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.
 
@@
 
-----
+## Evaluation of trajectory inference methods
 
-```{r}
-add_sfig(
-  figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview'),
-  "extended_evaluation_overview",
-  "Extended overview of our evaluation pipeline.
-From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
-We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
-Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
-We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
-  15, 
-  7.5
-)
-```
-
-## Evaluation of trajectory inference methods
+<!-- Evaluation overview -->
+In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
+To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets.
+<!-- Overall evaluation results -->
+Across all datasets and evaluation metrics, we found that slingshot, TSCAN and Monocle DDRTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returns the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees.
+This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
 
@@
 We further investigated the robustness of our evaluation across datasets and dataset sources.
-Method performance was moderately to highly correlated (0.5-0.9) between real and synthetic datasets (`r ref("fig", "results_overview", "b")` and `r ref("sfig", "real_synthetic_comparison")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
-Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with most methods significantly outperforming the next method in the ranking `r ref("sfig", "performance_dataset_variability_combined")`.
-To further illustrate this, if we looked at the scores a random 10% subset of all datasets, the overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
+Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
+Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
+To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets.
+The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
 On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
-For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top`datasets on which it would be the top ranking method.
+For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
 Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
-```{r}
-add_sfig(
-  figure_file(
-    "real_synthetic_comparison.svg",
-    "11-evaluation_robustness"
-  ), 
-  "real_synthetic_comparison", 
-  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
-)
-```
 
-```{r}
-add_sfig(
-  figure_file(
-    "performance_dataset_variability_combined.svg",
-    "11-evaluation_robustness"
-  ), 
-  "performance_dataset_variability_combined", 
-  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
-)
-```
+## Relationships between method characteristics to performance
 
-```{r}
-add_sfig(
-  figure_file('metrics_comparison.svg', "5-optimise_parameters/7-metric_comparison"), 
-  "metric_comparison", 
-  "Comparison of method performance between different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
-)
-```
-
-### Trajectory types and topology complexity
-
-During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
-If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
-Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
-Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
-For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
-
-### Trajectory types and topology complexity
-
-In most cases, the methods which were specifically designed to handle a particular trajectory type, also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
-These methods had typically better edge flip scores - as can be expected - and RF MSE scores, compared to methods not able to handle the particular trajectory type.
-However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-To further investigate the effect of trajectory complexity on performance, we divided them in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods.
-While there were no significant differences in performance on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`).
+<!-- Inverse relation between edge flip and correlation -->
+In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
+However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
+To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods.
+While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`).
 Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
@@
 
-```{r}
-add_sfig(
-  figure_file('trajtype_handle_comparison.svg', "8-compare_topology"), 
-  "trajtype_handle_comparison", 
-  "Comparison between the performance of methods able to handle a trajectory type with those who don't.
-Shown is the performance on those datasets containing a particular trajectory type (top), according to several metrics (right)."
-)
-```
-
-```{r}
-add_sfig(
-  figure_file('linear_split_comparison.svg', "8-compare_topology"), 
-  "linear_split_comparison", 
-  "Comparison of the performance between linear and non-linear methods.
-Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods.
-Their performance is shown on linear and non-linear datasets according to several performance metrics (right).
-NS: Not significant, \\*: p-value < 0.05, \\*\\*: p-value < 0.001, \\*\\*\\*: p-value < 0.0001"
-)
-```
-
+<!-- Topological complexity versus performance -->
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected using the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree, for example frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
-Trajectories generated by Monocle DDRTree (at default parameters) tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
-Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`).
-Inferring the correct topology without parameter tuning, is therefore still an open challenge.
-Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.
+Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
+Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
+Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases.
+Inferring the correct topology without parameter tuning is therefore still an open challenge.
+Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
 
@@
 
-----
-
-```{r}
-add_sfig(
-  figure_file('topology_sensitivity.svg', "8-compare_topology"), 
-  "topology_sensitivity", 
-  "Sensitivity of each TI method in predicting the correct trajectory topology.
-Shown are the % of datasets where a method is able to predict the exact correct trajectory topology, depending on the trajectory type present in the data.
-When a method is able to predict a particular trajectory type, the box is surrounded by a black border."
-)
-```
-
-### Effect of prior information
-
-In the current version of the evaluation, we only provided prior information when a method required it.
-We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`).
-Rather, methods which received prior information were on average positioned in the middle of the ranking.
-Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.
-
-```{r}
-add_sfig(
-  figure_file('required_priors_performance_comparison.svg', "10-prior_information"), 
-  "required_priors_performance_comparison", 
-  "Comparison of method performance depending on whether the method required a certain prior or not. The top performing method which requires a particular type of prior information is labelled."
-)
-```
-
-### Algorithm components
-
-We assessed whether the components of an algorithm could  be predictive of a method’s performance, using both random forest classification and statistical testing.
-Methods which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`).
+<!-- Additional factors which could influence performance -->
+Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method.
+Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
+When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`).
 On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
-
-```{r}
-add_sfig(
-  figure_file("method_components_scores.svg","6-developer_guidelines"), 
-  "method_components_scores", 
-  "Score of methods containing particular algorithmic components.
-Shown are the algorithmic components present in four or more methods, and whether the performance of methods containing a certain component was significantly higher than all the other methods.
-NS: Not significant.
-*: corrected p-value < 0.05 "
-)
-```
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "component_importance.rds",
-    "6-developer_guidelines"
-  ))[[params$table_format]], 
-  "component_importance", 
-  "Importance values for algorithmic components.
-The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
-)
-```
 
@@
 
-----
-
-```{r}
-add_sfig(
-  figure_file("qc_over_time.svg","4-method_characterisation"), 
-  "qc_over_time", 
-  "Quality control scores and number of citations over time."
-)
-```
-We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`).
+We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")`).
 We could also not find any quality aspect which was significantly associated with method performance.
-
-```{r}
-add_sfig(
-  figure_file(
-    "qc_performance_comparison.svg",
-    "5-optimise_parameters/6-qc_performance_comparison"
-  ), 
-  "qc_performance_comparison", 
-  "Comparison between method quality control scores and overall method performance."
-)
-```
 
@@
 For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-```{r}
-add_sfig(
-  figure_file("implementation_components.svg","4-method_characterisation"), 
-  "implementation_components", 
-  "Common algorithmic components shared by different TI methods.
-Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue).
-a) Wordcloud of the common components.
-b) Number of times a component was shared between methods.
-c) Number of methods containing a particular category."
-)
-```
 
@@
 Per region, one milestone must be directly connected to all other milestones in the network.
-
-```{r}
-add_sfig(
-  figure_file('wrapping.svg', '4-method_characterisation'),
-  "wrapping", 
-  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
-As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
-This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
-  15, 
-  7.5
-)
-```
 
@@
 To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.
-
-```{r}
-add_sfig(
-  figure_file("modulenets.svg", experiment_id = "2-dataset_characterisation/3-synthetic"),
-  "modulenets", 
-  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
-  15, 
-  5
-)
-```
 
@@
 add_sfig(
-  figure_file("score_aspects_v1.svg", experiment_id="3-score_rules/5-score_aspects"), 
+  figure_file("score_aspects_v2.svg", experiment_id="3-score_rules/5-score_aspects"), 
   "score_aspects", 
@@
 add_sfig(
-  figure_file("trajectory_type_trees.svg","2-dataset_characterisation/2-trajectory_types"), 
-  "trajectory_type_tree", 
-  "Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types.
-a) The graph structure for undirected and directed trajectory types.
-Shown are the topological properties which change when going from a particular trajectory type to a less complex type.
-b) Combined structure for directed and undirected trajectory types.",
+  figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview'),
+  "extended_evaluation_overview",
+  "Extended overview of our evaluation pipeline.
+From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
+We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
+Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
+We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
   15, 
-  10
+  7.5
 )
@@
 
-# Supplementary Material
 ```{r}
 add_sfig(
-  figure_file("methods_timeline.svg", "4-method_characterisation"),
-  "methods_over_time", 
-  "New TI methods over time.
-a) Number of methods published or in preprint.
-b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter).
-c) Number of methods which can handle a particular type of trajectory."
+  figure_file(
+    "robustness.svg",
+    "11-evaluation_robustness"
+  ), 
+  "robustness_comparison", 
+  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
+Given in the top-left corner of each panel is the Pearson's correlation.
+b) Violin plots of the overall performance.
+For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
+)
+```
+
+```{r}
+add_sfig(
+  figure_file('wrapping.svg', '4-method_characterisation'),
+  "wrapping", 
+  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
+  15, 
+  7.5
 )

### 2018-03-27 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
cleanup method characterisation

@@
 
-g2 <- ggplot(implementations_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(date, Citations)) +
+g2 <- ggplot(implementations_evaluated %>% mutate(ncitations = ifelse(ncitations == 0, 0.5, ncitations)), aes(date, ncitations)) +
     geom_smooth(span=2) +
@@
 
-g3 <- ggplot(implementations_evaluated %>% mutate(Citations = ifelse(Citations == 0, 0.5, Citations)), aes(Citations, qc_score)) +
+g3 <- ggplot(implementations_evaluated %>% mutate(ncitations = ifelse(ncitations == 0, 0.5, ncitations)), aes(ncitations, qc_score)) +
     geom_smooth(span=2) +

### 2018-03-27 analysis/analyses/4-method_characterisation/8_implementation_tables.R

@@ -7,12 +7,12 @@ experiment("4-method_characterisation")
 implementations <- read_rds(derived_file("implementations.rds"))
 
 implementations <- implementations %>%
-  filter(is_ti) %>%
+  filter(contains_ti) %>%
   arrange(date)
 
 # process priors
 prdf <- implementations %>%
-  select(implementation_id, start_cells:marker_feature_ids) %>%
+  select(implementation_id, !!priors$prior_id) %>%
   gather(prior_id, value, -implementation_id) %>%
   left_join(priors, by = "prior_id")
 pr_required <- prdf %>%
@@ -70,7 +70,6 @@ citation_format <- c(
 imp_table <- map(c("latex", "html"), function(format) {
   implementations_table <-
     implementations %>%
-    filter(is_ti) %>%
     arrange(date) %>%
     mutate(
       evaluated = ifelse(evaluated, "Yes", map_chr(non_inclusion_reasons_footnotes, ~paste0("No" , superscript[[format]](paste0(., collapse=" "))))),
@@ -86,13 +85,15 @@ imp_table <- map(c("latex", "html"), function(format) {
         kableExtra::cell_spec(
           label_simple_trajectory_types(maximal_trajectory_type),
           format,
-          background = toupper(set_names(trajectory_types$color, trajectory_types$id)[maximal_trajectory_type]),
+          background =
+            toupper(set_names(trajectory_types$color, trajectory_types$id)[maximal_trajectory_type]) %>%
+            map(~replace(., is.na(.), "#666666")),
           color = "#FFFFFF"
         ),
       implementation_name = kableExtra::cell_spec(
         implementation_name,
         format,
-        link = Code
+        link = code_location
       ),
       reference = kableExtra::cell_spec(
         map_chr(bibtex, citation_format[[format]]),
@@ -159,7 +160,7 @@ columns <- tribble(
   "Priors optional", "prior_optional", label_tbd,
   "Evaluated", "evaluated", label_long,
   "Reference" , "DOI", cite,
-  "Code/package", "Code", url
+  "Code/package", "code_location", url
 )
 
 # the order in the spreadsheet is the reference, as then conditional formatting is copied over :)

### 2018-03-27 analysis/analyses/4-method_characterisation/5_component_characterisation.R
cleanup method characterisation

@@ -18,7 +18,7 @@ component_categories <- tribble(
   ~component_id, ~category,
   "principal curves", "ordering",
   "PCA", "dimensionality reduction",
-  "knn graph", "graph building",
+  "kNN", "graph building",
   "MST", "graph building",
   "k-medoids", "clustering",
   "ICA", "dimensionality reduction",
@@ -58,7 +58,7 @@ component_categories <- tribble(
   "Morphing Gaussian Mixture", "generative model",
   "shortest path to start", "ordering",
   "latent variable model", "generative model",
-  "PQ tree", "graph building"
+  "PQ tree", "graph building",
 ) %>% group_by(component_id) %>%
   filter(row_number() == 1)
 

### 2018-03-28 analysis/paper/paper.Rmd
update paper

@@
 
-*`r format(Sys.time(), '%d %B, %Y')`*
 <!--
 Tips for working in this document:
-* Go to tools -> preferences and remove Use smart quotes
-* Enable automatic zotero syncing to bibtex
-* Make sure to pin the zotero bibtex references after first using them!
+Go to tools -> preferences and remove Use smart quotes
+Enable automatic zotero syncing to bibtex
+Make sure to pin the zotero bibtex references after first using them!
 -->
@@
 source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
+
+# knitr options
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  comment = "#>",
+  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
+)
 ```
+
+*`r format(Sys.time(), '%d %B, %Y')`*
+
 Wouter Saelens* ¹ ², Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
@@
 Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
-Since 2014, at least `r sum(implementations$is_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
 As a result, a comprehensive assessment of the performance and robustness of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
@@
 Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases.
-Methods that perform topology inference are still in the minority, though current trends suggest this will soon change.
+Methods that perform topology inference are still in the minority, though current trends suggest this will change soon.
 A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.
@@
 In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`).
-The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`).
-The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.
+The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("table", "implementations")`).
+The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on more than 200 real and synthetic datasets, (ii) a quality control of the provided software and documentation, and (iii) a combination of both the performance and QC evaluations to create a set of practical guidelines for users.
 
@@
 
-`r ref("fig", "evaluation_overview")` **An overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
-b) In order to make any two trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
+`r ref("fig", "evaluation_overview")` **Overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
+b) In order to make the trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
 c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
@@
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets.
+These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
+In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.
 <!-- Overall evaluation results -->
-Across all datasets and evaluation metrics, we found that slingshot, TSCAN and Monocle DDRTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
-When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returns the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees.
+Across all datasets and evaluation metrics, we found that slingshot, TSCAN, Scorpius and CellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees.
 This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
@@
 Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
-Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID).
-For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).
+Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions.
+Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID, MFA, cellTree Gibbs).
+For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).
 
@@
 Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
-Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
+Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
 To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets.
@@
 
-## Relationships between method characteristics to performance
+## Relationships between method characteristics and performance
 
 <!-- Inverse relation between edge flip and correlation -->
-In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type(`r ref("fig", "results_overview", "b")`).
 These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
-However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
+However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
 To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods.
-While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`).
+While there were no significant differences in performance on linear datasets, branching methods produced a significantly better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`).
 Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
-A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.
+A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.
 
@@
 Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
+It is therefore critical that these methods, now reaching `r sum(implementations$contains_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
 In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
@@
 We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
+
+In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
+In order to make the outputs of the TI methods more comparable, we developed a new common format.
+This not only allowed us to assess the performance of the methods compared to a gold standard, but may also facilitate the use of these methods by TI users, as multiple methods could be run on the same dataset and their results could be compared without manual conversion of input and output.
+Furthermore, the common structure opens up possibilities for new comparative visualisation techniques or ensemble methods.
+However, we acknowledge that our model has some limitations.
+Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
+when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
+when the connections between branches are uncertain).
+Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
+
+Although this study included the comparison of more TI methods than ever before (cite other methods which compared themselves to 3-4 others), the field keeps evolving, requiring a constant upgrade of the benchmarking.
+We believe that the common structure for TI results and the comparative metrics that we generated for this study, as well as the high number of datasets that we both gathered from the literature and generated synthetically represent a solid potential for benchmarking new methods.
+We will make the datasets available on “github..” and the tools necessary for the comparison of methods available on “github..
+Dyneval”, therefore providing the material for authors of new TI methods to robustly assess the performance of their method.
+
+Discuss advantages of synthetic data and how it could be used for other studies?
+
+Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, TSCAN, Monocle DDRTree, SCORPIUS and cellTree towering above all other methods.
+Nonetheless, we found that methods which did not perform well across all settings, such as reCAT, could still be highly performant in certain specific use cases.
+By comparing TI methods on a large set of datasets, we were able to extract characteristics concerning which methods performed best in which case.
+Currently, all TI methods were compared by using the default parameters set in their implementation.
+We decided to use this setting for two reasons.
+1) the setting of several parameters requires a certain knowledge of the data structure, which is not always known of TI users and 2) as certain methods require more parameters than others, we meant to prevent these methods from being over-ranked.
+Nonetheless, we plan to apply parameter tuning to all the methods in the near future, in order to benchmark the TI methods’ behaviour with optimal parameters.
+
+Discuss code quality, how it tends to improve with time, and how we believe good code quality leads to “better science”?
 
@@
 ### Method selection
-We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
 A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
@@
 
-1) method performance highly depends on trajectory type
-2) methods which are specific to a certain trajectory type generally perform better on datasets containing these trajectory types
-3) the complexity of the topologies returned by the different methods is highly variable 
-4) prior information doesn't increase performance in our settings
+# Supplementary Material
+```{r "supplementary", echo=FALSE, results='asis'}
+source("analysis/paper/sfigs_stables.R")
+```

### 2018-03-29 README.Rmd
add librsvg to rmd

@@
 ```
-sudo apt-get install libudunits2-dev libgsl-dev libsdl1.2-dev libreadline-dev imagemagick libfftw3-dev libudunits2-dev -y
+sudo apt-get install libudunits2-dev libgsl-dev libsdl1.2-dev libreadline-dev imagemagick libfftw3-dev libudunits2-dev librsvg2-dev -y
 sudo apt-get install python2.7-dev python3-dev python3-pip
@@
 ```
-sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip python3-matplotlib-tk openblas-devel lapack-devel
+sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip python3-matplotlib-tk openblas-devel lapack-devel librsvg2-devel
 pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn
@@
 ```
+
+macOS:
+
+Install SDL2 and SDL1.2 manually.
+
+Install `librsvg` with brew.
 

### 2018-03-29 README.Rmd
update readme

--- 2017-12-13 README.Rmd
@@
 ![travis](https://travis-ci.com/Zouter/dynalysis.svg?token=pzNqJjk3yapU1Hk7sDg8&branch=master)
-
-![overview](analysis/figures/toy_method_plots/grid.png)
-
-The goal of dynalysis is to ...
-
-### Citation
-
-Please cite this compendium as:
-
-> Authors, (`r format(Sys.Date(), "%Y")`). _Title of compendium_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
-
-### Dependencies
-
-You might have to install a few packages.
-
-Ubuntu/Debian:
-```
-sudo apt-get install libudunits2-dev libgsl-dev libsdl1.2-dev libreadline-dev imagemagick libfftw3-dev libudunits2-dev -y
-sudo apt-get install python2.7-dev python3-dev python3-pip
-pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn
-```
-
-Fedora/CentOS:
-```
-sudo dnf install openssl-devel libcurl-devel udunits2-devel libxml2-devel gsl-devel SDL2-devel readline-devel ImageMagick-c++-devel SDL-devel python2-devel python3-devel python3-pip python3-matplotlib-tk openblas-devel lapack-devel
-pip3 install --user virtualenv numpy matplotlib pandas six jinja2 python-dateutil pytz pyparsing cycler tqdm python-igraph rpy2 Cython scipy statsmodels sklearn seaborn
-Rscript -e 'install.packages("udunits2", configure.args =  c(udunits2 = '--with-udunits2-include=/usr/include/udunits2'))'
-```
 

### 2018-03-30 analysis/paper/paper.Rmd
update paper

@@
 <!-- Overall evaluation results -->
-Across all datasets and evaluation metrics, we found that slingshot, TSCAN, Scorpius and CellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
-When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees.
-This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
+Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, Scorpius and CellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Arctangent, Periodic principal curves (two control methods) and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
+This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Arctangent and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
 
@@
 
-`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
+`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and nrow(filter(methods, evaluated, type == "control")) control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
 b) The methods are ordered according to the overall score achieved in the benchmark evaluation.
@@
 top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
-filter(method_id == "slngsht") %>% 
+filter(method_short_name == "slngsht") %>% 
 filter(n_tasks == round(max(n_tasks)/10)) %>%
@@
 random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
-filter(method_id == "scoup") %>% 
+filter(method_short_name == "wishbone") %>% 
 pull(max_n_tasks_top)
-
 ```
@@
 To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets.
-The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
+The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, Slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
 On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
-For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
-Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
+For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
+Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, such as the more than 200 included in our evaluation, is necessary to robustly compare the overall performance of methods.
 
@@
 <!-- Inverse relation between edge flip and correlation -->
-In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type(`r ref("fig", "results_overview", "b")`).
+In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`).
 These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
 However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods.
-While there were no significant differences in performance on linear datasets, branching methods produced a significantly better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`).
-Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
+This indicates that current non-linear methods potentially contain less accurate ordering algorithms.
 A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.
@@
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
-Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
-Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases.
+Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN, AGA Pseudotime and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+AGA Pseudotime and Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data.
+Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 11% of the cases  (`r ref("stable", "topology_sensitivity")`).
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
-Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
+When the data contains a complex trajectory structure, TI will currently require a considerable guidance by the user, either to optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
 
@@
 Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
-When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`).
+When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`).
 On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
@@
 We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`).
-While recent methods had a slightly better quality than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
+While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
 Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
@@
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and slingshot performing equally well on less complex trajectories .
-No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")`.
-should be able to handle graph trajectories or even disconnected trajectories .
-In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, slingshot and cellTree for linear trajectories, and slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and SCOUP performing best when the data contained a complex tree, and Slingshot performing well on less complex trajectories .
+No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` should be able to handle graph trajectories or even disconnected trajectories .
+In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, Slingshot and cellTree for linear trajectories, and Slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
 The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
@@
 Methods to the right are ranked according to overall performance.
-Further to the right are shown the user friendly scores (++: ≥ 0.9, +: ≥ 0.8, ± ≥ 0.65, - ≥ 0.5) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
+Further to the right are shown the user friendliness scores (++: ≥ 0.9, +: ≥ 0.8, ± ≥ 0.65, - ≥ 0.5) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
 
@@
 
-Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, TSCAN, Monocle DDRTree, SCORPIUS and cellTree towering above all other methods.
+Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, SCORPIUS, TSCAN, AGA,  and cellTree obtaining the best overall scores.
 Nonetheless, we found that methods which did not perform well across all settings, such as reCAT, could still be highly performant in certain specific use cases.
@@
 
-### Method selection
+### Method selection and characterisation
 We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
@@
 
-* **DPT** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
+* **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
 This was then processed as described above.
-* **Sincell** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
+* **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
 This was then processed as described above.
-* **SLICE** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
-* **MFA** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
-
-#### cellTree
+* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
+* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
 
@@
 ```
+_Assigned to Wouter Saelens_

### 2018-04-03 analysis/paper/paper.Rmd
update paper

@@
   - \usepackage[labelformat=empty]{caption}
+  - \usepackage{hyperref}
+  - \hypersetup{colorlinks=true}
   - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
@@
 Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
-Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
-As a result, a comprehensive assessment of the performance and robustness of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
+Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
 
-In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness.
-Our results indicate that slingshot is the most accurate method to date, but other leading methods (e.g.
-TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases.
-We therefore developed a set of guidelines to aid users select the most appropriate method.
-While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases.
+In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness.
+Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data.
+We therefore developed a set of guidelines to aid users select the most appropriate method, based on the user's prior knowledge of the trajectory present in the data.
+While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be modelled computationally using using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs.
+Such dynamic processes can be modelled computationally using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic or disconnected graphs.
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`).
+For no other single-cell analyses workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
 All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
@@
 
+```{r, results='asis' }
+add_stable(
+  "implementations",
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
+  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
+)
+```
+
 <!-- §3: Problem statement -->
 Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness.
-Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation, which compares a large set of methods against a large set of datasets, of TI methods is still lacking.
-This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem.
+Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
+This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about which method would optimally solve their problem.
 Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
 
-# Results
-
-In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`).
+<!-- §4: Our study -->
+In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`).
 The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("table", "implementations")`).
-The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on more than 200 real and synthetic datasets, (ii) a quality control of the provided software and documentation, and (iii) a combination of both the performance and QC evaluations to create a set of practical guidelines for users.
+In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories.
+The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation.
+Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.
 
@@
 `r ref("fig", "evaluation_overview")` **Overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
-b) In order to make the trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
+b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
 c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
+
+# Results
 
@@
 <!-- Evaluation overview -->
-In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
@@
 In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.
+
 <!-- Overall evaluation results -->
-Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, Scorpius and CellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
-When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Arctangent, Periodic principal curves (two control methods) and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
+Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, SCORPIUS, AGA pseudotime and cellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
 This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Arctangent and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
@@
 
-`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and nrow(filter(methods, evaluated, type == "control")) control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
-b) The methods are ordered according to the overall score achieved in the benchmark evaluation.
-Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
-c) Overall performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
+`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and `r nrow(filter(methods, evaluated, type == "control"))` control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
+b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
+Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, as these simple methods return relatively accurate trajectories, to which other TI methods can be compared c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
 Also listed are the quality control scores aggregated according to practicality and the different categories.
@@
 For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
-Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, such as the more than 200 included in our evaluation, is necessary to robustly compare the overall performance of methods.
+Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, as we included in our evaluation, is necessary to robustly compare the overall performance of methods.
+
+```{r}
+add_sfig(
+  figure_file(
+    "robustness.svg",
+    "11-evaluation_robustness"
+  ), 
+  "robustness_comparison", 
+  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
+Given in the top-left corner of each panel is the Pearson's correlation.
+b) Violin plots of the overall performance.
+For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
+)
+```
 
@@
 AGA Pseudotime and Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data.
-Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 11% of the cases  (`r ref("stable", "topology_sensitivity")`).
+Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`).
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in only 11% of the cases  (`r ref("stable", "topology_sensitivity")`).
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
@@
 
-`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** a) % of datasets on which a TI method detects a particular trajectory type, compared with the correct trajectory type.
-b) Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
+`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
 
@@
 Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method.
-Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
-When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`).
-On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
+Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
+When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`).
+
+```{r}
+add_stable(
+  read_rds(figure_file(
+    "component_importance.rds",
+    "6-performance_predictors"
+  ))[[params$table_format]], 
+  "component_importance", 
+  "Importance values for algorithmic components.
+The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
+)
+```
 
@@
 
-`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each methods on every item from our quality control score sheet (`r ref("stable", "qc_checks")`).
+`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each method on every item from our quality control score sheet (`r ref("stable", "qc_checks")`).
 Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
 Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation.
-This weight is represented in the plot as distance on the y-axis.
+This height is represented in the plot as the block heighton the y-axis.
 Top: Average QC score for each method.
@@
 
-## Practical guidelines
+# Discussion
 
+<!-- Overview of evaluation -->
+
+<!-- Practical guidelines -->
 Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
-As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
-For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and SCOUP performing best when the data contained a complex tree, and Slingshot performing well on less complex trajectories .
-No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` should be able to handle graph trajectories or even disconnected trajectories .
-In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, Slingshot and cellTree for linear trajectories, and Slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
-The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
-Here, to our knowledge, only ElPiGraph.R ([github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R)), which is not yet published, can be used.
+As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
+For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex tree-like trajectories, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
+In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories.
+although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
 
@@
 Methods to the right are ranked according to overall performance.
-Further to the right are shown the user friendliness scores (++: ≥ 0.9, +: ≥ 0.8, ± ≥ 0.65, - ≥ 0.5) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
+Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
 
@@
 Second, even if the expected topology is known, it can be beneficial to also try out methods which make the less assumptions about the trajectory topology.
-When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the data.
+When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
 When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
 
-# Discussion
+<!-- The robustness of the evaluation -->
+While we have shown that the  Large subset of datasets on which a method is best. This raises important issues, even methods in the sub top deserve to be published.
 
@@
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
 It is therefore critical that these methods, now reaching `r sum(implementations$contains_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
@@
 Nonetheless, we found that methods which did not perform well across all settings, such as reCAT, could still be highly performant in certain specific use cases.
-By comparing TI methods on a large set of datasets, we were able to extract characteristics concerning which methods performed best in which case.
+By comparing TI methods on `r sum(tasks_info$task_source == "real")` real datasets and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, we were able to extract characteristics concerning which methods performed best in which case.
 Currently, all TI methods were compared by using the default parameters set in their implementation.
@@
 ## Trajectory inference methods
+<!-- Overview -->
+We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
+We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynverse/issues).
 
-### Method selection and characterisation
-We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
-We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynalysis/issues).
+<!-- Inclusion criterion -->
+Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
+This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation.
+<!-- Components -->
+The selected methods were dissected into a set of algorithmic components.
+Each component can have a significant impact on the performance, scalability, and output data structures.
+Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
 
-Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).
-
-<!-- PASTED FROM INTRODUCTION, need to rework -->
-<!-- §4: Prior information -->
-Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
+### Method input
+As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
+<!-- Prior information -->
+A large portion of the methods requires some form of prior information (e.g.
+a start cell) in order to be executable.
+Other methods allow to exploit certain prior information, though this is not required.
 Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
@@
 
-<!-- §5: Components -->
-A reductionist approach to characterising TI methods consists in dissecting them into a set of algorithmic components, as any component can have a significant impact on the performance, scalability, and output data structures.
-Across all TI methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
-
-### Method input
-As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
-Furthermore, when required, we also provided a maximum of 7 types of prior information.
-This prior information was extracted from the gold/silver standards as follows:
+The prior information was extracted from the gold/silver standards as follows:
 
 * **Start cells** The identity of one or more start cells.
-For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges.
+For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges.
 For ties, one random cell was chosen.
@@
 We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping", "a")`).
-This model consists of three parts.
-(i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
+This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
 (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one.
@@
 Per region, one milestone must be directly connected to all other milestones in the network.
+
+```{r}
+add_sfig(
+  figure_file('wrapping.svg', '4-method_characterisation'),
+  "wrapping", 
+  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
+  15, 
+  7.5
+)
+```
 
@@
 list_sentence <- function(x) {
-  pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))
+  if (length(x) == 1) {
+    x
+  } else {
+    paste(head(x, -1), collapse = ", ") %>% paste0(., " and ", tail(x, 1), sep = "")
+  }
 }
@@
  methods_evaluated %>% 
+    filter(type %in% c("algorithm", "control")) %>%
    filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% 
@@
 * **Type 1, direct:** `r conversion("trajectory")`.
-These methods assign each cell to a branch together with a pseudotime value and a branch network.
-The branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
+The wrapped method directly returns a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone.
+This usually indicates the transformations required for the method were specific to that method only and thus required a custom implementation.
 * **Type 2, linear pseudotime:** `r conversion("linear")`.
-These methods return a pseudotime value for each cell.
-The milestone network will consist of a single edge between two milestones, where cells are positioned on the transition proportional to their pseudotime value.
-* **Type 3, end state probability:** `r conversion("end_state_probability")`.
-These methods return a global pseudotime value and a probability for every end state.
-We use a single start state and add an edge to every end milestone each representing an end state.
-The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
-* **Type 4, cluster assignment:** `r conversion("cluster_graph")`.
-These methods return a cluster assignment and a cluster network.
-The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
-* **Type 5, projection onto nearest branch:** `r conversion("dimred_projection")`.
-These methods returning a cluster assignment, cluster network and dimensionality reduction.
-We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction.
-This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.
-* **Type 6, cell graph:** `r conversion("cell_graph")`.
-These methods return a network of connected cells, and determine which cells are part of the "backbone".
-One milestone is created for each cell that is part of the backbone and has a degree $\neq$ 2.
-Cells are positioned on the closest segment that is part of the backbone.
+The method returns a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
+* **Type 3, cyclical pseudotime:** `r conversion("cyclic")`.
+The method returns a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones.
+* **Type 4, end state probability:** `r conversion("end_state_probability")`.
+The method returns a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state.
+This is translated into a star-shaped milestone network, with regions of delayed commitment between all transitions.
+* **Type 5, cluster assignment:** `r conversion("cluster_graph")`.
+The method returns a milestone network, and an assignment of each cell to a specific milestone.
+Cells are positioned onto the milestones they are assigned to.
+* **Type 6, project onto nearest branch:** `r conversion("dimred_projection")`.
+The method returns a milestone network, and a dimensionality reduction of the cells and milestones.
+The cells are projected onto the closest nearest segment, thus determining the cells position along the milestone network.
+* **Type 7, cell graph:** `r conversion("cell_graph")`. The method returns a network of cells, 
+and which cell-cell transitions are part of the 'backbone' structure.
+Backbone cells with degree $\neq$ 2 are regarded as milestones, and all other cells are placed on transitions between the milestones.
 
@@
 
+The dynmethods github repository contains one vignette for each of the methods, discussing how to use the package to apply a method to a dataset, the specific implementation of the wrapper, and how the quality control scores of each TI method were derived.
+
+### Control methods
+For comparison purposes, we added three control methods, namely "Random", "Component 1", and "Angle".
+
+* **Random:** A random network with 50 milestones according to the Barabási–Albert model is generated, and cells are placed randomly along the network.
+This method provides a baseline performance to compare against; each other TI method should be able to obtain a higher benchmark score than this control method.
+* **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
+This method was included as some researchers indeed use this approach to obtain a pseudotime ordering of the cells.
+* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
+As reCAT was the only method specifically designed to produce cyclical trajectories, we believed it would be important to add this simple approach.
+
 ## Real datasets
-We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
+We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
 The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
@@
 Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
-At every step, we took great care to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable.
+At every step, we took great care to mimic real cellular regulatory networks as good as possible, while keeping the model simple and easily extendable.
 For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
@@
 Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
-Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into models) repressing each other [@yosef_dynamicregulatorynetwork_2013].
+Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other [@yosef_dynamicregulatorynetwork_2013].
 
@@
 
-Where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
+where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
 $$
@@
 `r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
-In order to make the scores produced by the different metrics comparable to one another, across datasets of varying difficulty, the raw scores are percentile rank transformed per metric per dataset to a $[0,1]$ range.
-The arithmetic mean is calculated across replicates.
-In order to ensure a method only obtains a high score if it scores well on all three metrics, the harmonic mean across metrics is calculated.
-At this point, the aggregated scores of all methods across all datasets is combined.
-As there can be an overrepresentation of datasets of a certain trajectory type, first an arithmetic mean is calculated per trajectory type, followed by an overall arithmetic mean across all trajectory types, thus obtaining a ranking of the methods.
+
+**Step 1:** In order to make the scores produced by the different metrics comparable to one another, across datasets of varying difficulty, the raw scores are transformed per metric per dataset to a $[0,1]$ range as follows:
+
+<!--
+$$
+\begin{aligned}
+\text{scale}(x) &= \text{sigmoid}\left( \frac{x - \textrm{mean}(x)}{\textrm{max}(\textrm{abs}(x - \textrm{mean}(x))) \times 5} \right)
+\end{aligned}
+$$
+-->
+
+$$
+\begin{aligned}
+\text{scale}(x) &= \text{sigmoid}\left( \frac{x - \bar{x}}{\textrm{max}(\textrm{abs}(x - \bar{x})) \times 5} \right)\text{, where } \bar{x} \text{ is the arithmetic mean of } x.
+\end{aligned}
+$$
+
+Here, the mean is subtracted from the raw score in order to center the distribution around zero.
+The denominator is used to scale the range of the distribution such that, after sigmoid transformation, the minimum and maximum transformed scores lie close to 0 and/or 1, respectively.
+
+**Step 2:** After transformation, the arithmetic mean is calculated across replicates.
+
+**Step 3:** The normalised scores is combined across all methods and all datasets.
+
+**Step 4:** In order to remove effects from unbalanced trajectory types (e.g. an overrepresentation of linear datasets), the arithmetic mean is first computed per trajectory type.
+
+**Step 5:** An overall score is computed by calculating the arithmetic mean across all trajectory types.
+
+**Step 6:** A ranking of the methods is obtained by calculating the harmonic mean between the different metrics, thus requiring a method to obtain a high score on all three metrics in order to obtain a high overall score.
 
@@
 
+```{r}
+add_stable(
+  read_rds(figure_file(
+    "qc_checks.rds",
+    "4-method_characterisation"
+  ))[[params$table_format]], 
+  "qc_checks", 
+  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
+)
+```
+
 Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions), the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing), or indications that the tool will be broadly applicable on new datasets (such as being open-source, containing a good tutorial and support system, and being thoroughly evaluated in the study where it was published).
@@
 
-```{r}
-add_sfig(
-  figure_file(
-    "robustness.svg",
-    "11-evaluation_robustness"
-  ), 
-  "robustness_comparison", 
-  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
-Given in the top-left corner of each panel is the Pearson's correlation.
-b) Violin plots of the overall performance.
-For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
-)
-```
-
-```{r}
-add_sfig(
-  figure_file('wrapping.svg', '4-method_characterisation'),
-  "wrapping", 
-  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
-As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
-This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
-  15, 
-  7.5
-)
-```
-
-```{r, results='asis' }
-add_stable(
-  "implementations",
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
-)
-```
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "component_importance.rds",
-    "6-developer_guidelines"
-  ))[[params$table_format]], 
-  "component_importance", 
-  "Importance values for algorithmic components.
-The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
-)
-```
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "qc_checks.rds",
-    "4-method_characterisation"
-  ))[[params$table_format]], 
-  "qc_checks", 
-  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
-)
-```
-
-# Supplementary Material
 ```{r "supplementary", echo=FALSE, results='asis'}
@@
 ```
-_Assigned to Wouter Saelens_

### 2018-04-03 analysis/paper/paper_latex.Rmd
update paper

@@
 Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
-Since 2014, at least `r sum(implementations$is_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
-As a result, a comprehensive assessment of the performance and robustness of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
+Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
 
-In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness.
-Our results indicate that slingshot is the most accurate method to date, but other leading methods (e.g.
-TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases.
-We therefore developed a set of guidelines to aid users select the most appropriate method.
-While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases.
+In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness.
+Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data.
+We therefore developed a set of guidelines to aid users select the most appropriate method, based on the user's prior knowledge of the trajectory present in the data.
+While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
@@
 Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases.
-Methods that perform topology inference are still in the minority, though current trends suggest this will soon change.
+Methods that perform topology inference are still in the minority, though current trends suggest this will change soon.
 A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.
+
+```{r, results='asis' }
+add_stable(
+  "implementations",
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
+  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
+)
+```
 
@@
 Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness.
-Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation, which compares a large set of methods against a large set of datasets, of TI methods is still lacking.
-This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem.
+Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
+This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about which method would optimally solve their problem.
 Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
 
-# Results
-
-In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`).
-The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`).
-The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.
+<!-- §4: Our study -->
+In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`).
+The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("table", "implementations")`).
+In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories.
+The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation.
+Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.
 
@@
 
-`r ref("fig", "evaluation_overview")` **An overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
-b) In order to make any two trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
+`r ref("fig", "evaluation_overview")` **Overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
+b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
 c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
+
+# Results
 
@@
 <!-- Evaluation overview -->
-In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
-These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets.
+These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
+In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.
+
 <!-- Overall evaluation results -->
-Across all datasets and evaluation metrics, we found that slingshot, TSCAN and Monocle DDRTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
-When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returns the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees.
-This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
+Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, SCORPIUS, AGA pseudotime and cellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
+When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
+This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Arctangent and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
 
@@
 
-`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
-b) The methods are ordered according to the overall score achieved in the benchmark evaluation.
-Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
-c) Overall performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
+`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and `r nrow(filter(methods, evaluated, type == "control"))` control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
+b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
+Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, as these simple methods return relatively accurate trajectories, to which other TI methods can be compared c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
 Also listed are the quality control scores aggregated according to practicality and the different categories.
@@
 Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
-Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
+Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
 To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets.
-The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
+The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, Slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
 On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
-For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
-Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
+For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
+Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, as we included in our evaluation, is necessary to robustly compare the overall performance of methods.
 
-## Relationships between method characteristics to performance
+```{r}
+add_sfig(
+  figure_file(
+    "robustness.tmp.pdf",
+    "11-evaluation_robustness"
+  ), 
+  "robustness_comparison", 
+  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
+Given in the top-left corner of each panel is the Pearson's correlation.
+b) Violin plots of the overall performance.
+For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
+)
+```
+
+## Relationships between method characteristics and performance
 
 <!-- Inverse relation between edge flip and correlation -->
-In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`).
+In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`).
 These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
-However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods.
-While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`).
-Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms.
-A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.
+However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
+This indicates that current non-linear methods potentially contain less accurate ordering algorithms.
+A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.
 
@@
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`).
-Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets.
-Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies.
-Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases.
+Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN, AGA Pseudotime and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+AGA Pseudotime and Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data.
+Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`).
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in only 11% of the cases  (`r ref("stable", "topology_sensitivity")`).
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
-Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
+When the data contains a complex trajectory structure, TI will currently require a considerable guidance by the user, either to optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
 
@@
 Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method.
-Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
-When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`).
-On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.
+Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
+When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`).
+
+```{r}
+add_stable(
+  read_rds(figure_file(
+    "component_importance.rds",
+    "6-performance_predictors"
+  ))[[params$table_format]], 
+  "component_importance", 
+  "Importance values for algorithmic components.
+The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
+)
+```
 
@@
 
-## Practical guidelines
+# Discussion
 
+<!-- Overview of evaluation -->
+
+<!-- Practical guidelines -->
 Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
-As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data.
-For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and slingshot performing equally well on less complex trajectories .
-No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")`.
-should be able to handle graph trajectories or even disconnected trajectories .
-In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, slingshot and cellTree for linear trajectories, and slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
-The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory.
-Here, to our knowledge, only ElPiGraph.R ([github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R)), which is not yet published, can be used.
+As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
+For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex tree-like trajectories, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
+In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories.
+although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
 
@@
 Methods to the right are ranked according to overall performance.
-Further to the right are shown the user friendly scores (++: ≥ 0.9, +: ≥ 0.8, ± ≥ 0.65, - ≥ 0.5) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
+Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
 
@@
 Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
+It is therefore critical that these methods, now reaching `r sum(implementations$contains_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
 In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
@@
 
+In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
+In order to make the outputs of the TI methods more comparable, we developed a new common format.
+This not only allowed us to assess the performance of the methods compared to a gold standard, but may also facilitate the use of these methods by TI users, as multiple methods could be run on the same dataset and their results could be compared without manual conversion of input and output.
+Furthermore, the common structure opens up possibilities for new comparative visualisation techniques or ensemble methods.
+However, we acknowledge that our model has some limitations.
+Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
+when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
+when the connections between branches are uncertain).
+Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
+
+Although this study included the comparison of more TI methods than ever before (cite other methods which compared themselves to 3-4 others), the field keeps evolving, requiring a constant upgrade of the benchmarking.
+We believe that the common structure for TI results and the comparative metrics that we generated for this study, as well as the high number of datasets that we both gathered from the literature and generated synthetically represent a solid potential for benchmarking new methods.
+We will make the datasets available on “github..” and the tools necessary for the comparison of methods available on “github..
+Dyneval”, therefore providing the material for authors of new TI methods to robustly assess the performance of their method.
+
+Discuss advantages of synthetic data and how it could be used for other studies?
+
+Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, SCORPIUS, TSCAN, AGA,  and cellTree obtaining the best overall scores.
+Nonetheless, we found that methods which did not perform well across all settings, such as reCAT, could still be highly performant in certain specific use cases.
+By comparing TI methods on `r sum(tasks_info$task_source == "real")` real datasets and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, we were able to extract characteristics concerning which methods performed best in which case.
+Currently, all TI methods were compared by using the default parameters set in their implementation.
+We decided to use this setting for two reasons.
+1) the setting of several parameters requires a certain knowledge of the data structure, which is not always known of TI users and 2) as certain methods require more parameters than others, we meant to prevent these methods from being over-ranked.
+Nonetheless, we plan to apply parameter tuning to all the methods in the near future, in order to benchmark the TI methods’ behaviour with optimal parameters.
+
+Discuss code quality, how it tends to improve with time, and how we believe good code quality leads to “better science”?
+
 # Online Methods
@@
 ## Trajectory inference methods
+<!-- Overview -->
+We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
+We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynverse/issues).
 
-### Method selection
-We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
-We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynalysis/issues).
+<!-- Inclusion criterion -->
+Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
+This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation.
+<!-- Components -->
+The selected methods were dissected into a set of algorithmic components.
+Each component can have a significant impact on the performance, scalability, and output data structures.
+Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
+Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
 
-Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).
-
-<!-- PASTED FROM INTRODUCTION, need to rework -->
-<!-- §4: Prior information -->
-Another key characteristic of TI methods is the selection of prior information that a method requires or can optionally exploit.
+### Method input
+As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
+<!-- Prior information -->
+A large portion of the methods requires some form of prior information (e.g.
+a start cell) in order to be executable.
+Other methods allow to exploit certain prior information, though this is not required.
 Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
@@
 Per region, one milestone must be directly connected to all other milestones in the network.
+
+```{r}
+add_sfig(
+  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
+  "wrapping", 
+  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
+  15, 
+  7.5
+)
+```
 
@@
 
-* **DPT** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
+* **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
 This was then processed as described above.
-* **Sincell** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
+* **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
 This was then processed as described above.
-* **SLICE** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
-* **MFA** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
+* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
+* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
 
-#### cellTree
+The dynmethods github repository contains one vignette for each of the methods, discussing how to use the package to apply a method to a dataset, the specific implementation of the wrapper, and how the quality control scores of each TI method were derived.
+
+### Control methods
+For comparison purposes, we added three control methods, namely "Random", "Component 1", and "Angle".
+
+* **Random:** A random network with 50 milestones according to the Barabási–Albert model is generated, and cells are placed randomly along the network.
+This method provides a baseline performance to compare against; each other TI method should be able to obtain a higher benchmark score than this control method.
+* **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
+This method was included as some researchers indeed use this approach to obtain a pseudotime ordering of the cells.
+* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
+As reCAT was the only method specifically designed to produce cyclical trajectories, we believed it would be important to add this simple approach.
 
 ## Real datasets
-We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
+We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
 The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
@@
 
-```{r}
-add_sfig(
-  figure_file(
-    "robustness.tmp.pdf",
-    "11-evaluation_robustness"
-  ), 
-  "robustness_comparison", 
-  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
-Given in the top-left corner of each panel is the Pearson's correlation.
-b) Violin plots of the overall performance.
-For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
-)
+```{r "supplementary", echo=FALSE, results='asis'}
+source("analysis/paper/sfigs_stables.R")
 ```
-
-```{r}
-add_sfig(
-  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
-  "wrapping", 
-  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
-As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
-This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
-  15, 
-  7.5
-)
-```
-
-```{r, results='asis' }
-add_stable(
-  "implementations",
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
-)
-```
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "component_importance.rds",
-    "6-developer_guidelines"
-  ))[[params$table_format]], 
-  "component_importance", 
-  "Importance values for algorithmic components.
-The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
-)
-```
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "qc_checks.rds",
-    "4-method_characterisation"
-  ))[[params$table_format]], 
-  "qc_checks", 
-  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
-)
-```
-
-1) method performance highly depends on trajectory type
-2) methods which are specific to a certain trajectory type generally perform better on datasets containing these trajectory types
-3) the complexity of the topologies returned by the different methods is highly variable 
-4) prior information doesn't increase performance in our settings

### 2018-04-03 analysis/paper/paper_latex.Rmd
update paper

@@
 ```{r}
+
+# the correlation when subsampling
 correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
@@
           mean()
+
+# how often is a method top when subsampling
 top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
@@
 pull(rank_match)
+
+# how many datasets are necessary to get a method on top
 random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
@@
 The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, Slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
-On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best.
+On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best  (`r ref("sfig", "robustness_comparison", "c")`).
 For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
@@
 <!-- Topological complexity versus performance -->
+```{r}
+
+# sensitivity of topology prediction
+topology_sensitivity <- read_rds(result_file("topology_sensitivity.rds"))
+```
+
 Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN, AGA Pseudotime and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-AGA Pseudotime and Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data.
+Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
+AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data.
 Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`).
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in only 11% of the cases  (`r ref("stable", "topology_sensitivity")`).
+Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot best discriminating linear from bifurcating topologies (sensitivity of respectively ), and Monocle DDRTree and SCUBA best predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`).
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
@@
 Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method.
-Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "b")`).
+Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a-b")`).
 When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`).
@@
 Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation.
-This height is represented in the plot as the block heighton the y-axis.
+This is represented in the plot as the height on the y-axis.
 Top: Average QC score for each method.
@@
 The selected methods were dissected into a set of algorithmic components.
-Each component can have a significant impact on the performance, scalability, and output data structures.
+Each component can have a significant impact on the performance, scalability, and output.
 Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
@@
 
-The prior information was extracted from the gold/silver standards as follows:
+The prior information was extracted from the gold standard as follows:
 
@@
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-We defined three metrics for comparing the likeness of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`).
+We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`).
 Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
@@
 ```{r "supplementary", echo=FALSE, results='asis'}
-source("analysis/paper/sfigs_stables.R")
+source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
 ```

### 2018-04-03 analysis/analyses/4-method_characterisation/7_quality_control.Rmd
update qc

@@
   arrange(-qc_score) %>% 
+  filter(!is.na(qc_score)) %>% 
   pull(implementation_id)
@@
   mutate(implementation_id = factor(implementation_id, implementation_order)) %>% 
-  ggplot(aes(implementation_id, qc_score)) +
+  ggplot(aes(as.numeric(implementation_id), qc_score)) +
     geom_bar(aes(fill=qc_score), stat="identity") +
@@
     scale_y_continuous(NULL, expand=c(0, 0), limits=c(0, 1)) +
-    scale_x_discrete(NULL, breaks=c(0, 100), expand=c(0, 0)) +
+    scale_x_continuous("", breaks=NULL, expand = c(0, 0)) +
     viridis::scale_fill_viridis(label_long("qc_score"), option="D", guide=guide_colorbar(title.position = "top", title.hjust=0.5, barwidth = unit(2, "inches"))) +
@@
       ), 
-      nudge_x = 6, 
+      nudge_x = 0.5, 
       direction="y", 
@@
       lineheight=0.8, 
-      min.segment.length = 0
+      min.segment.length = 0,
+      hjust=0
     ) +

### 2018-04-03 analysis/paper/paper.Rmd

@@
 add_stable(
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
 "implementations",
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
-  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
+"Overview of the trajectory inference methods included in this study."
 )
@@
 In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`).
-The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("table", "implementations")`).
+The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`).
 In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories.
@@
 When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
-This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Arctangent and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
+This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Angle and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
 
@@
 b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
-Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, as these simple methods return relatively accurate trajectories, to which other TI methods can be compared c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, which provide a baseline performance.
+c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
 d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
@@
 Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID, MFA, cellTree Gibbs).
+Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID, cellTree Gibbs).
 For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).
@@
 # sensitivity of topology prediction
-topology_sensitivity <- read_rds(result_file("topology_sensitivity.rds"))
+topology_sensitivity <- read_rds(result_file("topology_sensitivity.rds", "8-compare_topology"))
 ```
@@
 Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`).
-Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot best discriminating linear from bifurcating topologies (sensitivity of respectively ), and Monocle DDRTree and SCUBA best predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`).
+Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`).
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
@@
 
-`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
+```{r, results='asis' }
+add_stable(
+  read_rds(figure_file("topology_sensitivity_table.rds", "8-compare_topology"))[[params$table_format]],
+"topology_sensitivity",
+"Overview of the trajectory inference methods included in this study."
+)
+```
+
+`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
 
@@
 <!-- Overview -->
-We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("stable", "implementations")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
 A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
@@
 
-Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "simplementations")`).
 This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface.

### 2018-04-03 analysis/paper/supplementary_latex.Rmd
add supplementary_latex.Rmd

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
bibliography: references.bib
csl: nature-biotechnology.csl
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage[normalem]{ulem}
  - \usepackage{lscape}
  - \usepackage{pbox}
  - \usepackage[labelformat=empty]{caption}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true}
  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
  - \newcommand{\hideFromPandoc}[1]{#1}
  - \hideFromPandoc{\let\Begin\begin}
  - \hideFromPandoc{\let\End\end}
  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
mainfont: Open Sans

---

<!--
Tips for working in this document:
Go to tools -> preferences and remove Use smart quotes
Enable automatic zotero syncing to bibtex
Make sure to pin the zotero bibtex references after first using them!
-->

```{r setup, include=FALSE}
source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))

# knitr options
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
)
```

```{r "supplementary", echo=FALSE, results='asis', warning=FALSE}
list2env(read_rds(derived_file("sfigs_stables.rds", "paper")), environment())
source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
```

### 2018-04-03 analysis/paper/paper.Rmd
update paper

@@
   - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
+  - \newcommand{\hideFromPandoc}[1]{#1}
+  - \hideFromPandoc{\let\Begin\begin}
+  - \hideFromPandoc{\let\End\end}
+  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
 mainfont: Open Sans
@@
 Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data.
-We therefore developed a set of guidelines to aid users select the most appropriate method, based on the user's prior knowledge of the trajectory present in the data.
+We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data.
 While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be modelled computationally using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic or disconnected graphs.
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`).
-For no other single-cell analyses workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
+For no other single-cell analysis workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
 All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
@@
 
-`r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview")`)
-
-`r ref("fig", "evaluation_overview")` **Overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview"),
+  ref_id = "evaluation_overview", 
+  caption_main = "Overview of several key aspects of the evaluation.", 
+  caption_text = "a) A schematic overview of our evaluation pipeline.
 b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
-c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
+c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
+)
+```
 
@@
 
-`r anchor("fig", "results_overview")`
-![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
-
-`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and `r nrow(filter(methods, evaluated, type == "control"))` control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("overview.svg", experiment_id = "9-main_figure"),
+  ref_id = "results_overview", 
+  caption_main = glue::glue("Overview of the results on the evaluation of {nrow(filter(methods, evaluated, type == 'algorithm'))} TI methods and {nrow(filter(methods, evaluated, type == 'control'))} control methods."),
+  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
 b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
@@
 d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
-Also listed are the quality control scores aggregated according to practicality and the different categories.
-
-----------
+Also listed are the quality control scores aggregated according to practicality and the different categories."
+)
+```
 
@@
   "robustness_comparison", 
-  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types.
+  "The robustness of our evaluation.",
+  "a) Comparison of method performance between real and synthetic datasets across trajectory types.
 Given in the top-left corner of each panel is the Pearson's correlation.
-b) Violin plots of the overall performance.
-For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
+\"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types.
+b) Violin plots of the overall performance on each individual dataset.
+For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05).
+c) Estimate of the size of largest subset of datasets on which the method would be the average top performer.
+For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether after aggregating all scores leads to the method being ranked first.
+This number is therefore only a        lower bound, and the actual size of largest subset could be higher."
 )
@@
 
-`r anchor("fig", "topology_complexity_comparison")`
-![](`r figure_file("topology_complexity_comparison.svg","8-compare_topology")`)
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("topology_complexity_comparison.svg", experiment_id = "8-compare_topology"),
+  ref_id = "topology_complexity_comparison", 
+  caption_main = "Comparing the ability of the top TI methods to detect the correct trajectory type.", 
+  caption_text = "Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
+)
+```
 
@@
 "topology_sensitivity",
-"Overview of the trajectory inference methods included in this study."
+"Sensitivity of topology prediction of each TI method.
+Given is how often a method is able to correctly predict the topology of the trajectory present in the dataset, between different trajectory types."
 )
 ```
-
-`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
 
@@
 Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a-b")`).
-When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`).
+When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`), although these tendencies were not significant.
 
@@
 
-`r anchor("fig", "implementation_qc_overview")`
-![](`r figure_file("implementation_qc_overview.svg","4-method_characterisation")`)
-
-`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each method on every item from our quality control score sheet (`r ref("stable", "qc_checks")`).
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "4-method_characterisation"),
+  ref_id = "implementation_qc_overview", 
+  caption_main = "Overview of the quality control results for every method.", 
+  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_checks')}).
 Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
@@
 Right: The average score of each quality control item.
-Shown into more detail are those items which had a score lower than 0.5.
+Shown into more detail are those items which had a score lower than 0.5.")
+)
+```
 
@@
 
-<!-- Overview of evaluation -->
+<!-- The context -->
+In this study, we presented a large scale evaluation of the performance and code of  `r nrow(filter(methods, evaluated, type == "algorithm"))`  TI methods.
+By using a common structure and three metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types.
+We also assessed the code quality of these methods, to provide information on important scientific and software development practices for these methods.
 
+<!-- FOR METHOD USERS -->
 <!-- Practical guidelines -->
-Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
-As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
+Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
+We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex tree-like trajectories, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
-In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories.
-although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex trajectory structures with at most one bifurcation, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
+In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories, although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
 
-`r anchor("fig", "user_guidelines")`
-![](`r figure_file("tree.svg", experiment_id="7-user_guidelines")`)
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("tree.svg", experiment_id = "7-user_guidelines"),
+  ref_id = "user_guidelines", 
+  caption_main = "Practical guidelines for method users.", 
+  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
+We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
+Methods to the right are ranked according to their performance on a particular (set of) trajectory type.
+Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
+)
+```
 
-`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.** As the performance of a method most heavily depended on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
-We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
-Methods to the right are ranked according to overall performance.
-Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
+<!-- FOR METHOD DEVELOPERS -->
+<!-- New tools: challenges and possibilities -->
+Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories.
+Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories.
+Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods.
+Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers.
+To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
 
-When choosing a method, it is important to take two further points into account.
-First, it is important that a trajectory and the downstream results and/or hypotheses originating from it are confirmed by multiple TI methods.
-This to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type.
-Second, even if the expected topology is known, it can be beneficial to also try out methods which make the less assumptions about the trajectory topology.
-When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
-When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
-
-<!-- The robustness of the evaluation -->
-While we have shown that the  Large subset of datasets on which a method is best. This raises important issues, even methods in the sub top deserve to be published.
-
-<!-- The context -->
-Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-It is therefore critical that these methods, now reaching `r sum(implementations$contains_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
-In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
-When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art.
-We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
-
-In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
-In order to make the outputs of the TI methods more comparable, we developed a new common format.
-This not only allowed us to assess the performance of the methods compared to a gold standard, but may also facilitate the use of these methods by TI users, as multiple methods could be run on the same dataset and their results could be compared without manual conversion of input and output.
-Furthermore, the common structure opens up possibilities for new comparative visualisation techniques or ensemble methods.
-However, we acknowledge that our model has some limitations.
-Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
-when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
-when the connections between branches are uncertain).
-Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
-
-Although this study included the comparison of more TI methods than ever before (cite other methods which compared themselves to 3-4 others), the field keeps evolving, requiring a constant upgrade of the benchmarking.
-We believe that the common structure for TI results and the comparative metrics that we generated for this study, as well as the high number of datasets that we both gathered from the literature and generated synthetically represent a solid potential for benchmarking new methods.
-We will make the datasets available on “github..” and the tools necessary for the comparison of methods available on “github..
-Dyneval”, therefore providing the material for authors of new TI methods to robustly assess the performance of their method.
-
-Discuss advantages of synthetic data and how it could be used for other studies?
-
-Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, SCORPIUS, TSCAN, AGA,  and cellTree obtaining the best overall scores.
-Nonetheless, we found that methods which did not perform well across all settings, such as reCAT, could still be highly performant in certain specific use cases.
-By comparing TI methods on `r sum(tasks_info$task_source == "real")` real datasets and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, we were able to extract characteristics concerning which methods performed best in which case.
-Currently, all TI methods were compared by using the default parameters set in their implementation.
-We decided to use this setting for two reasons.
-1) the setting of several parameters requires a certain knowledge of the data structure, which is not always known of TI users and 2) as certain methods require more parameters than others, we meant to prevent these methods from being over-ranked.
-Nonetheless, we plan to apply parameter tuning to all the methods in the near future, in order to benchmark the TI methods’ behaviour with optimal parameters.
-
-Discuss code quality, how it tends to improve with time, and how we believe good code quality leads to “better science”?
+<!-- The robustness of the evaluation & development of new methods-->
+We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
+Nonetheless, we also found that it is possible to define for almost every method a relatively large subset of datasets on which this method on average performs the best.
+These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, as there is a high risk that a certain new methodology would not improve upon the state-of-the-art.
+Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, provide a unique insight in specific use cases, or return a more scalable solution.
+We feel that such methods also deserve to be published, and encourage their use provided that they are contained in a user friendly tool.
 
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation.
+This resulted in the inclusion of  `r nrow(filter(methods, evaluated, type == "algorithm"))`  methods in the evaluation.
 <!-- Components -->
@@
 Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Interestingly, components are frequently shared between different algorithms (`r ref("sfig", "implementation_components")`).
-For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated  (`r ref("sfig", "implementation_components", "b")`).
+Interestingly, components are frequently shared between different algorithms.
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.
 
@@
   "wrapping", 
-  "**A common interface for TI methods.** a) The input and output of each TI method is standardised.
+  "A common interface for TI methods.", 
+  "a) The input and output of each TI method is standardised.
 As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
@@
 Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
-To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone.
+To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, as a backbone.
 We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
@@
   "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
+  "",
   15, 
@@
   "metric_geodesic", 
-  "**The modified geodesic distances demonstrated on a toy example.** a) A toy example containing four milestones (W to Z) and five cells (a to e).
+  "The modified geodesic distances demonstrated on a toy example.", 
+  "a) A toy example containing four milestones (W to Z) and five cells (a to e).
 b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
@@
   "edge_flip_comparison", 
-  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets"), "."), 
+  paste0("Edge flip score values for common trajectory topologies"), 
   15, 
@@
   "Aggregation methodology of metric scores.",
+  "",
   15, 
@@
   "extended_evaluation_overview",
-  "Extended overview of our evaluation pipeline.
-From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
+  "Extended overview of our evaluation pipeline.", 
+  "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
 We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.

### 2018-04-03 analysis/paper/paper_latex.Rmd
update paper

@@
 
-`r anchor("fig", "evaluation_overview")`
-![](`r figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="evaluation_overview")`)
-
-`r ref("fig", "evaluation_overview")` **Overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline.
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="evaluation_overview"),
+  ref_id = "evaluation_overview", 
+  caption_main = "Overview of several key aspects of the evaluation.", 
+  caption_text = "a) A schematic overview of our evaluation pipeline.
 b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
-c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.
+c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
+)
+```
 
@@
 When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
-This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Arctangent and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
+This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Angle and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
 
-`r anchor("fig", "results_overview")`
-![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
-
-`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and `r nrow(filter(methods, evaluated, type == "control"))` control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("overview.tmp.pdf", experiment_id = "9-main_figure"),
+  ref_id = "results_overview", 
+  caption_main = glue::glue("Overview of the results on the evaluation of {nrow(filter(methods, evaluated, type == 'algorithm'))} TI methods and {nrow(filter(methods, evaluated, type == 'control'))} control methods."),
+  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
 b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
-Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, as these simple methods return relatively accurate trajectories, to which other TI methods can be compared c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, which provide a baseline performance.
+c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
 d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
-Also listed are the quality control scores aggregated according to practicality and the different categories.
-
-----------
+Also listed are the quality control scores aggregated according to practicality and the different categories."
+)
+```
 
@@
 
-`r anchor("fig", "topology_complexity_comparison")`
-![](`r figure_file("topology_complexity_comparison.tmp.pdf","8-compare_topology")`)
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("topology_complexity_comparison.tmp.pdf", experiment_id = "8-compare_topology"),
+  ref_id = "topology_complexity_comparison", 
+  caption_main = "Comparing the ability of the top TI methods to detect the correct trajectory type.", 
+  caption_text = "Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
+)
+```
 
-`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.
+```{r, results='asis' }
+add_stable(
+  read_rds(figure_file("topology_sensitivity_table.rds", "8-compare_topology"))[[params$table_format]],
+"topology_sensitivity",
+"Sensitivity of topology prediction of each TI method.
+Given is how often a method is able to correctly predict the topology of the trajectory present in the dataset, between different trajectory types."
+)
+```
 
@@
 
-`r anchor("fig", "implementation_qc_overview")`
-![](`r figure_file("implementation_qc_overview.tmp.pdf","4-method_characterisation")`)
-
-`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.** Shown is the score given for each method on every item from our quality control score sheet (`r ref("stable", "qc_checks")`).
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "4-method_characterisation"),
+  ref_id = "implementation_qc_overview", 
+  caption_main = "Overview of the quality control results for every method.", 
+  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_checks')}).
 Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
@@
 
-<!-- Overview of evaluation -->
+<!-- The context -->
+In this study, we presented a large scale evaluation of the performance and code of  `r nrow(filter(methods, evaluated, type == "algorithm"))`  TI methods.
+By using a common structure and three metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types.
+We also assessed the code quality of these methods, to provide information on important scientific and software development practices for these methods.
 
+<!-- FOR METHOD USERS -->
 <!-- Practical guidelines -->
-Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
-As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
+Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
+We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex tree-like trajectories, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
-In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories.
-although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
+In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex trajectory structures with at most one bifurcation, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
+In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories, although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
 
-`r anchor("fig", "user_guidelines")`
-![](`r figure_file("tree.tmp.pdf", experiment_id="7-user_guidelines")`)
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("tree.tmp.pdf", experiment_id = "7-user_guidelines"),
+  ref_id = "user_guidelines", 
+  caption_main = "Practical guidelines for method users.", 
+  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
+We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
+Methods to the right are ranked according to their performance on a particular (set of) trajectory type.
+Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
+)
+```
 
-`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.** As the performance of a method most heavily depended on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
-We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
-Methods to the right are ranked according to overall performance.
-Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).
+<!-- FOR METHOD DEVELOPERS -->
+<!-- New tools: challenges and possibilities -->
+Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories.
+Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories.
+Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods.
+Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers.
+To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
 
-When choosing a method, it is important to take two further points into account.
-First, it is important that a trajectory and the downstream results and/or hypotheses originating from it are confirmed by multiple TI methods.
-This to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type.
-Second, even if the expected topology is known, it can be beneficial to also try out methods which make the less assumptions about the trajectory topology.
-When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
-When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
-
-<!-- The robustness of the evaluation -->
-While we have shown that the  Large subset of datasets on which a method is best. This raises important issues, even methods in the sub top deserve to be published.
-
-<!-- The context -->
-Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data.
-It is therefore critical that these methods, now reaching `r sum(implementations$contains_ti, na.rm=TRUE)`, are evaluated to guide users in their choice.
-In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
-When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art.
-We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
-
-In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods.
-In order to make the outputs of the TI methods more comparable, we developed a new common format.
-This not only allowed us to assess the performance of the methods compared to a gold standard, but may also facilitate the use of these methods by TI users, as multiple methods could be run on the same dataset and their results could be compared without manual conversion of input and output.
-Furthermore, the common structure opens up possibilities for new comparative visualisation techniques or ensemble methods.
-However, we acknowledge that our model has some limitations.
-Currently, It cannot take into account uncertainty of a cell's position, which can both occur on the cellular ordering (e.g.
-when the position of a cell is uncertain within a branch)  or on the trajectory topology (e.g.
-when the connections between branches are uncertain).
-Some current methods already model this uncertainty in some way, mainly on the cellular ordering `r cite_methods("ouija")`, and in the future we will adapt our output model to also allow this uncertainty.
-
-Although this study included the comparison of more TI methods than ever before (cite other methods which compared themselves to 3-4 others), the field keeps evolving, requiring a constant upgrade of the benchmarking.
-We believe that the common structure for TI results and the comparative metrics that we generated for this study, as well as the high number of datasets that we both gathered from the literature and generated synthetically represent a solid potential for benchmarking new methods.
-We will make the datasets available on “github..” and the tools necessary for the comparison of methods available on “github..
-Dyneval”, therefore providing the material for authors of new TI methods to robustly assess the performance of their method.
-
-Discuss advantages of synthetic data and how it could be used for other studies?
-
-Our evaluation indicated a large heterogeneity in the performance of the current TI methods, with Slingshot, SCORPIUS, TSCAN, AGA,  and cellTree obtaining the best overall scores.
-Nonetheless, we found that methods which did not perform well across all settings, such as reCAT, could still be highly performant in certain specific use cases.
-By comparing TI methods on `r sum(tasks_info$task_source == "real")` real datasets and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, we were able to extract characteristics concerning which methods performed best in which case.
-Currently, all TI methods were compared by using the default parameters set in their implementation.
-We decided to use this setting for two reasons.
-1) the setting of several parameters requires a certain knowledge of the data structure, which is not always known of TI users and 2) as certain methods require more parameters than others, we meant to prevent these methods from being over-ranked.
-Nonetheless, we plan to apply parameter tuning to all the methods in the near future, in order to benchmark the TI methods’ behaviour with optimal parameters.
-
-Discuss code quality, how it tends to improve with time, and how we believe good code quality leads to “better science”?
+<!-- The robustness of the evaluation & development of new methods-->
+We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
+Nonetheless, we also found that it is possible to define for almost every method a relatively large subset of datasets on which this method on average performs the best.
+These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, as there is a high risk that a certain new methodology would not improve upon the state-of-the-art.
+Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, provide a unique insight in specific use cases, or return a more scalable solution.
+We feel that such methods also deserve to be published, and encourage their use provided that they are contained in a user friendly tool.
 

### 2018-04-03 analysis/paper/supplementary_latex.Rmd
move list2env

@@
 )
+
+list2env(read_rds(derived_file("sfigs_stables.rds", "paper")), environment())
 ```
 
-```{r "supplementary", echo=FALSE, results='asis', warning=FALSE}
-list2env(read_rds(derived_file("sfigs_stables.rds", "paper")), environment())
+```{r "supplementary", echo=FALSE, results='asis'}
 source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))

### 2018-04-03 analysis/paper/paper.Rmd
whatever

@@
 
-# Supplementary Material
+## Data and code availability
+The datasets used in this study are deposited in zenodo [doi.org/10.5281/zenodo.1211533](doi.org/10.5281/zenodo.1211533).
+Code is available as several R packages on github [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
 
@@
 ```{r "supplementary", echo=FALSE, results='asis'}
+if (params$table_format == "html") 
   source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
+write_rds(lst(sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
 ```

### 2018-04-03 analysis/analyses/4-method_characterisation/8_implementation_tables.R

@@ -70,6 +70,7 @@ citation_format <- c(
 imp_table <- map(c("latex", "html"), function(format) {
   implementations_table <-
     implementations %>%
+    filter(type == "algorithm") %>%
     arrange(date) %>%
     mutate(
       evaluated = ifelse(evaluated, "Yes", map_chr(non_inclusion_reasons_footnotes, ~paste0("No" , superscript[[format]](paste0(., collapse=" "))))),

### 2018-04-03 analysis/paper/paper.Rmd
update paper

@@
 ## Data and code availability
-The datasets used in this study are deposited in zenodo [doi.org/10.5281/zenodo.1211533](doi.org/10.5281/zenodo.1211533).
-Code is available as several R packages on github [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
+The datasets used in this study are deposited in zenodo [doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533).
+Code is available as several R packages on github [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
 

### 2018-04-04 analysis/paper/paper.Rmd

@@
 
+<!-- Further considerations -->
+When choosing a method, it is important to take two further points into account.
+First, it is important that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
+This is to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type.
+Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
+When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
+When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
+
 <!-- FOR METHOD DEVELOPERS -->
@@
 
-<!-- The robustness of the evaluation & development of new methods-->
+<!-- The robustness of the evaluation & development of new methods -->
 We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
@@
 Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, provide a unique insight in specific use cases, or return a more scalable solution.
-We feel that such methods also deserve to be published, and encourage their use provided that they are contained in a user friendly tool.
+We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.
 
@@
 For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
-The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
+The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
 
@@
 
-Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "simplementations")`).
+Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
 This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface.

### 2018-04-04 analysis/paper/paper.Rmd

@@
 "implementations",
-"Overview of the trajectory inference methods included in this study."
+paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::collapse("; "))
 )
@@
 <!-- §4: Our study -->
-In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`).
+In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
 The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`).
@@
 add_fig(
-  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="evaluation_overview"),
-  ref_id = "evaluation_overview", 
+  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="manual_figures"),
+  ref_id = "ti_evaluation_overview", 
   caption_main = "Overview of several key aspects of the evaluation.", 
@@
 c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
+)
+```
+
+```{r}
+add_sfig(
+  fig_path = figure_file("overview_evaluation_expanded_v1.svg", "manual_figures"),
+  ref_id = "overview_evaluation_expanded",
+  caption_main = "Extended overview of our evaluation pipeline.", 
+  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
+We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
+Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
+We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
+  width = 15, 
+  height = 7.5
 )
@@
 <!-- Evaluation overview -->
-In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`).
+In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`).
+We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "c")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
@@
 b) Violin plots of the overall performance on each individual dataset.
-For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05).
-c) Estimate of the size of largest subset of datasets on which the method would be the average top performer.
-For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether after aggregating all scores leads to the method being ranked first.
+For each method, the arch indicates the first method with which a method performs significantly higher (using a pairwise wilcoxon test corrected for multiple testing; p-value < 0.05).
+c) Estimates of the size of largest subset of datasets on which the method would be the average top performer.
+For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether this method ranks first when aggregating the scores over these datasets.
 This number is therefore only a lower bound, and the actual size of largest subset could be higher."
@@
   ref_id = "topology_complexity_comparison", 
-  caption_main = "Comparing the ability of the top TI methods to detect the correct trajectory type.", 
-  caption_text = "Distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
+  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the gold standard.", 
+  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
 )
@@
 Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap.
-We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "wrapping", "a")`).
+We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`).
 This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
@@
 add_sfig(
-  figure_file('wrapping.svg', '4-method_characterisation'),
-  "wrapping", 
+  figure_file("overview_wrapping_v1.svg", "manual_figures"),
+  "overview_wrapping", 
   "A common interface for TI methods.", 
@@
 As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the six wrapper functions to transform its output to the common trajectory model.
+After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model.
 This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-b) The specific transformations performed by each of the wrapper functions is explained in more detail.",
+b) The specific transformations performed by each of the wrapper functions is shown in more detail.",
   15, 
@@
 
-Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "wrapping", "b")`).
+Depending on the output of a method, we used different strategies to convert the output to our model (`r ref("sfig", "overview_wrapping", "b")`).
 Special conversions are denoted by an \*, and will be explained in more detail below.
@@
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`).
+We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`).
 Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
@@
 add_sfig(
-  figure_file("score_aspects_v2.svg", experiment_id="3-score_rules/5-score_aspects"), 
+  figure_file("aspects_v2.svg", experiment_id="manual_figures/metrics"), 
   "score_aspects", 
@@
 add_sfig(
-  figure_file("metric_geodesic.svg", experiment_id="evaluation_overview"), 
+  figure_file("geodesic.svg", experiment_id="manual_figures/metrics"), 
   "metric_geodesic", 
@@
 
-According to the defined common trajectory model (`r ref("fig", "evaluation_overview", "b")`), cells are also allowed to have a delayed commitment.
+According to the defined common trajectory model (`r ref("fig", "ti_evaluation_overview", "b")`), cells are also allowed to have a delayed commitment.
 In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network.
@@
 add_sfig(
-  figure_file("metric_aggregation_overview_v1.svg", experiment_id="evaluation_overview"), 
+  figure_file("overview_aggregation_v1.svg", experiment_id="manual_figures/metrics"), 
   "aggregation_of_results", 
@@
 
-```{r}
-add_sfig(
-  figure_file('expanded_ti_evaluation_overview_v1.svg', 'evaluation_overview'),
-  "extended_evaluation_overview",
-  "Extended overview of our evaluation pipeline.", 
-  "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
-We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
-Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
-We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
-  15, 
-  7.5
-)
-```
-
 ```{r "supplementary", echo=FALSE, results='asis'}

### 2018-04-04 analysis/paper/paper_latex.Rmd

@@
 add_fig(
-  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="evaluation_overview"),
-  ref_id = "evaluation_overview", 
+  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="manual_figures"),
+  ref_id = "ti_evaluation_overview", 
   caption_main = "Overview of several key aspects of the evaluation.", 
@@
 c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
+)
+```
+
+```{r}
+add_sfig(
+  fig_path = figure_file("overview_evaluation_expanded_v1.tmp.pdf", "manual_figures"),
+  ref_id = "overview_evaluation_expanded",
+  caption_main = "Extended overview of our evaluation pipeline.", 
+  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
+We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
+Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
+We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
+  width = 15, 
+  height = 7.5
 )
@@
 add_sfig(
-  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
-  "wrapping", 
+  figure_file("overview_wrapping_v1.tmp.pdf", "manual_figures"),
+  "overview_wrapping", 
   "A common interface for TI methods.", 
@@
 add_sfig(
-  figure_file("score_aspects_v2.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
+  figure_file("aspects_v2.tmp.pdf", experiment_id="manual_figures/metrics"), 
   "score_aspects", 
@@
 add_sfig(
-  figure_file("metric_geodesic.tmp.pdf", experiment_id="evaluation_overview"), 
+  figure_file("geodesic.tmp.pdf", experiment_id="manual_figures/metrics"), 
   "metric_geodesic", 
@@
 add_sfig(
-  figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview"), 
+  figure_file("overview_aggregation_v1.tmp.pdf", experiment_id="manual_figures/metrics"), 
   "aggregation_of_results", 
@@
 
-```{r}
-add_sfig(
-  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
-  "extended_evaluation_overview",
-  "Extended overview of our evaluation pipeline.", 
-  "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
-We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
-Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
-We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
-  15, 
-  7.5
-)
-```
-
 ```{r "supplementary", echo=FALSE, results='asis'}

### 2018-04-04 analysis/analyses/3-metric_characterisation/4-edge_flip.Rmd

--- 2018-02-06 analysis/analyses/3-score_rules/flip_score_characterisation.Rmd
@@
 
-```{r setup, include=FALSE}
+```{r setup}
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+library(dynalysis)
 library(tidyverse)
-library(dynmethods)
 library(dyneval)
-library(dynalysis)
+library(cowplot)
+```
 
-experiment("score_rules/edge_flip")
+```{r}
+experiment("3-score_rules/4-edge_flip")
 
@@
 write_rds(edge_flip_comparison, figure_file("edge_flip_comparison.rds"))
+ggsave(figure_file("edge_flip_comparison.svg"), edge_flip_comparison, width=7, height=7)
 ```

### 2018-04-05 analysis/paper/paper.Rmd
update paper

@@
 While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
-We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
@@
 Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
-This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about which method would optimally solve their problem.
+This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea about which method would optimally solve their problem.
 Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
@@
 <!-- §4: Our study -->
-In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
+In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
 The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`).
@@
 add_fig(
-  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="manual_figures"),
+  fig_path = figure_file("overview_evaluation_v4.svg", experiment_id="manual_figures"),
   ref_id = "ti_evaluation_overview", 
@@
 b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
-c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
+c) Trajectories are automatically classified into one of nine trajectory types.
+d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
 )
@@
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "c")`).
+The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`).
+We defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
@@
 Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions.
-Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID, cellTree Gibbs).
+Other methods (reCAT, SCOUP, StemID, cellTree Gibbs) exceeded the pre-defined time or memory limits when applied to the largest datasets, and therefore were not scalable to large numbers of cells or genes.
 For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).
@@
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
-When the data contains a complex trajectory structure, TI will currently require a considerable guidance by the user, either to optimise the parameters or to choose the method whose output best fits the user's expert knowledge.
+Currently, inferring a complex trajectory structure still requires a considerable user input, either by setting optimal parameters, or by subjectively choosing a method whose results best fit the user’s expert knowledge.
 
@@
 <!-- Additional factors which could influence performance -->
-Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method.
+Apart from the trajectory types, we also investigated several other factors that could influence the performance of a method.
 Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a-b")`).
@@
 Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers.
-To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
+To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
 
@@
 * **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
-This method was included as some researchers indeed use this approach to obtain a pseudotime ordering of the cells.
+We added this method because indeed in some studies the first component of a dimensionality reduction is being used to pseudotemporally order cells [@kouno_temporaldynamicstranscriptional_2013,@zeng_pseudotemporalorderingsingle_2017].
 * **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
@@
 We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
+The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
 Whenever possible, we preferred to start from the raw counts data.
@@
 add_sfig(
-  figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip"), 
+  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 
@@
 
-```{r}
-qc_references <- readRDS(result_file("checks.rds", "4-method_characterisation")) %>% 
-  .$references %>% 
-  str_split("[, \n]+") %>% 
-  unlist() %>% 
-  unique() %>% 
-  discard(is.na) %>%
-  glue::collapse(., ";") %>%
-  paste0("[", ., "]")
-```
-
-We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` .
+We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_checks")`).
 The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
@@
 ## Trajectory types
-We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria.
+We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`).
 These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical.
@@
 ## Data and code availability
-The datasets used in this study are deposited in zenodo [doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533).
-Code is available as several R packages on github [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
+The datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)).
+The code to fully reproduce the results from this study has been split into several R packages and is available on github ([github.com/dynverse/dynverse](https://github.com/dynverse/dynverse)).
 
@@
   source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
-write_rds(lst(sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
+write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
 ```

### 2018-04-05 analysis/paper/paper_latex.Rmd
update paper

@@
   - \usepackage{hyperref}
-  - \hypersetup{colorlinks=true}
+  - \definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
+  - \definecolor{cerulean}{rgb}{0.0, 0.48, 0.65}
+  - \definecolor{burntorange}{rgb}{0.8, 0.33, 0.0}
+  - \hypersetup{colorlinks=true,linkcolor=airforceblue,citecolor=cerulean,urlcolor=burntorange}
   - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
@@
 In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness.
-Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data.
+Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
+Instead, the choice of method depends on the trajectory topology present in the data.
 We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data.
 While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
-We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
 
@@
 add_fig(
-  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="manual_figures"),
+  fig_path = figure_file("overview_evaluation_v4.tmp.pdf", experiment_id="manual_figures"),
   ref_id = "ti_evaluation_overview", 
@@
 First, it is important that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
-This is to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type.
+This is to make sure the prediction is not biased due to the given parameter setting or the particular algorithms underlying a TI method.
 Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
@@
 When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
+
+<!-- A common trajectory analysis framework -->
+Critical to the broad application of TI methods is that both the user's input and the method's output is standardised, so that users can easily and quickly try out different methods, compare their output, and run common analysis pipelines such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018].
+A common framework to analyse trajectories is especially important when the output of methods contradict, as they often do.
+We illustrate this in  `r ref("fig", "example_predictions")`.
+Here, we predicted trajectories on a datasets containing fibroblasts which were reprogrammed towards neurons and myocytes, a bifurcating trajectory [@treutlein_dissectingdirectreprogramming_2016].
+The topology of all five methods differs greatly, with Slingshot underestimating the complexity of the topology, while AGA pseudotime and Monocle DDRTree overestimated the complexity, at least compared to what was expected in the data.
+
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("bifurcating_example.tmp.pdf", experiment_id = "14-example_predictions"),
+  ref_id = "example_predictions", 
+  caption_main = "An example gold standard and prediction.", 
+  caption_text = "As an illustration to our practical guidelines"
+)
+```
 
@@
 add_sfig(
-  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip"), 
+  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 

### 2018-04-05 analysis/paper/paper.Rmd
update paper

@@
 In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness.
-Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data.
+Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
+Instead, the choice of method depends on the trajectory topology present in the data.
 We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data.
@@
 When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
+
+<!-- A common trajectory analysis framework -->
+Critical to the broad application of TI methods is that both the user's input and the method's output is standardised, so that users can easily and quickly try out different methods, compare their output, and run common analysis pipelines such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018].
+A common framework to analyse trajectories is especially important when the output of methods contradict, as they often do.
+We illustrate this in  `r ref("fig", "example_predictions")`.
+Here, we predicted trajectories on a datasets containing fibroblasts which were reprogrammed towards neurons and myocytes, a bifurcating trajectory [@treutlein_dissectingdirectreprogramming_2016].
+The topology of all five methods differs greatly, with Slingshot underestimating the complexity of the topology, while AGA pseudotime and Monocle DDRTree overestimated the complexity, at least compared to what was expected in the data.
+
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
+  ref_id = "example_predictions", 
+  caption_main = "An example gold standard and prediction.", 
+  caption_text = "As an illustration to our practical guidelines"
+)
+```
 

### 2018-04-05 analysis/paper/update_paper_from_google_drive.R

@@ -90,3 +90,8 @@ render(
 drive_update("dynverse/paper.pdf", "analysis/paper/paper_latex.pdf")
 drive_update("dynverse/supplementary.pdf", "analysis/paper/supplementary_latex.pdf")
 
+# prepare for submission
+paper_folder <- "../../dyndocs/20180401_submission_nat_biotech/"
+file.copy("analysis/paper/paper_latex.pdf", paste0(paper_folder, "paper.pdf"))
+
+figs$fig_path

### 2018-04-05 analysis/paper/paper_latex.Rmd
update paper

@@
 Inferring the correct topology without parameter tuning is therefore still an open challenge.
-Currently, inferring a complex trajectory structure still requires a considerable user input, either by setting optimal parameters, or by subjectively choosing a method whose results best fit the user’s expert knowledge.
 
@@
 <!-- A common trajectory analysis framework -->
-Critical to the broad application of TI methods is that both the user's input and the method's output is standardised, so that users can easily and quickly try out different methods, compare their output, and run common analysis pipelines such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018].
-A common framework to analyse trajectories is especially important when the output of methods contradict, as they often do.
-We illustrate this in  `r ref("fig", "example_predictions")`.
-Here, we predicted trajectories on a datasets containing fibroblasts which were reprogrammed towards neurons and myocytes, a bifurcating trajectory [@treutlein_dissectingdirectreprogramming_2016].
-The topology of all five methods differs greatly, with Slingshot underestimating the complexity of the topology, while AGA pseudotime and Monocle DDRTree overestimated the complexity, at least compared to what was expected in the data.
+Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018].
+Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`).
+In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values.
+In addition, further debate is required in order to arrive at a consensus concerning a common interface for trajectory models.
 
@@
   ref_id = "example_predictions", 
-  caption_main = "An example gold standard and prediction.", 
-  caption_text = "As an illustration to our practical guidelines"
+  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
+  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process  [@treutlein_dissectingdirectreprogramming_2016].
+b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions.
+The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology but obtains a suboptimal ordering of the cells.
+c) More insight can be gained into which cells ended up on which branch by placing the trajectory segments one after the other.
+d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed.
+e) While only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
 )
@@
 These were combined with the expression levels of the genes simulated within a trajectory.
-Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$.
+Next, true counts were simulated using $Y'_{i,j} \sim \text{Poisson}(\lambda_{i,j})$.
 Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
@@
   "edge_flip_comparison", 
-  paste0("Edge flip score values for common trajectory topologies"), 
-  15, 
-  5
+ "Edge flip score values for common trajectory topologies",
+  ""
 )

### 2018-04-05 analysis/paper/paper.Rmd
update paper

@@
 Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic or disconnected graphs.
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected [@wolf_graphabstractionreconciles_2017] graphs.
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
@@
 linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
-These methods therefore mainly focused on correctly ordering the cells along this fixed topology.
+These methods therefore mainly focused on correctly ordering the cells along the fixed topology.
 Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases.
 Methods that perform topology inference are still in the minority, though current trends suggest this will change soon.
-A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.
 
@@
 <!-- §4: Our study -->
-In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
+In this study, we performed a comprehensive evaluation (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`) of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods.
 The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`).
@@
 The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`).
-We defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`).
+We defined a set of metrics for calculating the similarity of two trajectories represented in this common model, each assessing a different aspect of a trajectory: the cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
@@
 Nonetheless, we also found that it is possible to define for almost every method a relatively large subset of datasets on which this method on average performs the best.
-These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, as there is a high risk that a certain new methodology would not improve upon the state-of-the-art.
-Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, provide a unique insight in specific use cases, or return a more scalable solution.
+These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, given that there is a high risk that a certain new methodology would not improve upon the state-of-the-art.
+Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution, or provide a unique insight in specific use cases.
+Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns an measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
 We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.
@@
 write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
+write_rds(figs, derived_file("figs.rds", "paper"))
 ```

### 2018-04-05 analysis/paper/paper.Rmd
update paper

@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected [@wolf_graphabstractionreconciles_2017] graphs.
@@
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`).
-We defined a set of metrics for calculating the similarity of two trajectories represented in this common model, each assessing a different aspect of a trajectory: the cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`).
+The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs).
+We defined a set of metrics for calculating the similarity of two trajectories represented in the common model, each assessing a different aspect of a trajectory: the cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
@@
 Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, SCORPIUS, AGA pseudotime and cellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
-When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
+When focusing on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS and cellTree maptpx returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
 This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Angle and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).

### 2018-04-05 analysis/paper/paper.Rmd
fix uncurl reference

@@
   caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the gold standard.", 
-  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data."
+  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies.
+A positive difference means that the topology predicted by the method is more complex than the one in the gold standard."
 )
@@
 "Sensitivity of topology prediction of each TI method.
-Given is how often a method is able to correctly predict the topology of the trajectory present in the dataset, between different trajectory types."
+Given is how often a method is able to correctly predict the topology of the trajectory present in the dataset, within a particular trajectory type."
 )
@@
 Apart from the trajectory types, we also investigated several other factors that could influence the performance of a method.
-Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a-b")`).
+Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a")`).
 When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`), although these tendencies were not significant.
@@
   "Importance values for algorithmic components.
-The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
+The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a p-value using a two-tailed Mann–Whitney U test (corrected for multiple testing using the method of Benjamini-Hochberg) and the increase in node purity importance measure using random forest classification."
 )
@@
 When choosing a method, it is important to take two further points into account.
-First, it is important that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
+First, it is critical that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
 This is to make sure the prediction is not biased due to the given parameter setting or the particular algorithms underlying a TI method.
@@
 In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values.
-In addition, further debate is required in order to arrive at a consensus concerning a common interface for trajectory models.
+In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.
 
@@
   caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process  [@treutlein_dissectingdirectreprogramming_2016].
-b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions.
-The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology but obtains a suboptimal ordering of the cells.
-c) More insight can be gained into which cells ended up on which branch by placing the trajectory segments one after the other.
+A multi-dimensional scaling of dataset is shown at the bottom.
+MEF: mouse embryonic fibroblast b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions.
+The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology.
+c) More insight can be gained into the ordering of the cells by placing the branches one after the other.
+The predicted topology is shown in the bottom of each plot.
 d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed.
-e) While only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
+e) Our metrics can be used to compare the output to a gold standard or to each other.
+In this case, while only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
 )
@@
 ### Control methods
-For comparison purposes, we added three control methods, namely "Random", "Component 1", and "Angle".
+To define baseline performance, we added three control methods, namely "Random", "Component 1", and "Angle".
 
 * **Random:** A random network with 50 milestones according to the Barabási–Albert model is generated, and cells are placed randomly along the network.
-This method provides a baseline performance to compare against; each other TI method should be able to obtain a higher benchmark score than this control method.
+This method provides a bottom line performance; each other TI method should be able to obtain a higher benchmark score than this control method.
 * **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
-We added this method because indeed in some studies the first component of a dimensionality reduction is being used to pseudotemporally order cells [@kouno_temporaldynamicstranscriptional_2013,@zeng_pseudotemporalorderingsingle_2017].
+Although conceptually very simple, it has been used in certain studies [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
 * **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
-As reCAT was the only method specifically designed to produce cyclical trajectories, we believed it would be important to add this simple approach.
 
@@
 These raw counts were all normalised and filtered using a common pipeline, discussed later.
-We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).
+We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample; gold standard) (`r ref("stable", "datasets")`).
+These datasets are deposited in Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].
 
@@
 ## Data and code availability
-The datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)).
+The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].
 The code to fully reproduce the results from this study has been split into several R packages and is available on github ([github.com/dynverse/dynverse](https://github.com/dynverse/dynverse)).

### 2018-05-24 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
rename description

@@
 
-method_descriptions <- map(method_names, ~get(paste0("description_", .))()) %>% set_names(method_names)
+method_ti_methods <- map(method_names, ~get(paste0("ti_", .))()) %>% set_names(method_names)
 
@@
 ```{r}
-# results <- PRISM::qsub_lapply(names(method_descriptions), function(method_name) {
+# results <- PRISM::qsub_lapply(names(method_ti_methods), function(method_name) {
 results <- purrr::map(method_names, function(method_name) {
@@
   cat("Processing ", method_name, "\n", sep="")
-  method <- method_descriptions[[method_name]]
+  method <- method_ti_methods[[method_name]]
   params <- method_params[[method_name]]
@@
 })
-# }, qsub_environment = list2env(lst(method_descriptions, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
+# }, qsub_environment = list2env(lst(method_ti_methods, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
 ```

### 2018-05-24 analysis/analyses/3-metric_characterisation/4-edge_flip.Rmd
rename description

@@
 ```{r}
-descriptions <- list(description_scorpius())
+ti_methods <- list(ti_scorpius())
 data <- dyntoy::generate_toy_datasets(c("linear"), num_replicates=1, num_cells=30, num_genes=10) %>% dynutils::extract_row_to_list(1)
 
-results <- map(descriptions, ~.$run_fun(data$counts))
+results <- map(ti_methods, ~.$run_fun(data$counts))
 
-networks_methods <- tibble(topology=map_chr(descriptions, "name"), milestone_network=map(results, "milestone_network"))
+networks_methods <- tibble(topology=map_chr(ti_methods, "name"), milestone_network=map(results, "milestone_network"))
 ```

### 2018-05-24 analysis/analyses/6-benchmark/1-submit_jobs.R

@@ -4,7 +4,7 @@ library(tidyverse)
 experiment("5-optimise_parameters/3-evaluate_parameters")
 
 # settings
-methods <- get_descriptions() %>% filter(!short_name %in% c("manual", "scorspar"))
+methods <- get_ti_methods() %>% filter(!short_name %in% c("manual", "scorspar"))
 metrics <- c("correlation", "rf_mse", "edge_flip")
 timeout_per_execution <- 60 * 60 * 6
 num_repeats <- 4

### 2018-06-08 analysis/analyses/3-metric_characterisation/4-edge_flip.Rmd
rename dyntoy function

@@
 ```{r}
-topologies <- eval(formals(dyntoy:::generate_toy_milestone_network)$model)
+topologies <- eval(formals(dyntoy:::generate_milestone_network)$model)
 topologies <- topologies[topologies != "BA"]
-networks_synthetic <- map(topologies, dyntoy:::generate_toy_milestone_network) %>% tibble(milestone_network = ., topology = topologies)
+networks_synthetic <- map(topologies, dyntoy:::generate_milestone_network) %>% tibble(milestone_network = ., topology = topologies)
 ```

### 2018-06-24 analysis/analyses/evaluate_toy/1-methods-toys.Rmd

@@
   # factory(function() {
-    dyneval:::execute_evaluation(tasks, method, params, metrics = metric_names, timeout = 100)
+    dyneval:::evaluate_ti_method(tasks, method, params, metrics = metric_names, timeout = 100)
   # }

### 2018-06-27 scripts/evaluate_toy/1-methods-toys.Rmd
reorganise dynalysis

--- 2018-06-24 analysis/analyses/evaluate_toy/1-methods-toys.Rmd
@@
 ```{r}
-# results <- PRISM::qsub_lapply(names(method_ti_methods), function(method_name) {
+# results <- qsub::qsub_lapply(names(method_ti_methods), function(method_name) {
 results <- purrr::map(method_names, function(method_name) {
@@
 })
-# }, qsub_environment = list2env(lst(method_ti_methods, tasks, metric_names, method_params)), qsub_config = PRISM::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
+# }, qsub_environment = list2env(lst(method_ti_methods, tasks, metric_names, method_params)), qsub_config = qsub::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
 ```

### 2018-06-27 scripts/evaluate_toy/2-methods-plots.Rmd
reorganise dynalysis

--- 2017-09-15 analysis/analyses/evaluate_toy/2-methods-plots.Rmd
@@
   print(method_name)
-  plot_fun <- method_descriptions[[method_name]]$plot_fun
+  plot_fun <- method_ti_methods[[method_name]]$plot_fun
 
@@
 
-method_descriptions <- map(scores$method_id, ~get(paste0("description_", .))()) %>% set_names(scores$method_id)
+method_ti_methods <- map(scores$method_id, ~get(paste0("ti_", .))()) %>% set_names(scores$method_id)
 plots <- map2(scores$model, scores$method_id, plot_method)

### 2018-06-27 scripts/manual_ti/guide.Rmd
reorganise dynalysis

--- 2018-02-12 analysis/analyses/manual_ti/guide.Rmd
@@
 When we assume that cells go through this process gradually, the expression changes will follow a continuum, and this will be reflected in the 2D visualisation of the cells.
-Cells will follow paths within this 2D visualisation, but due to heterogeneity, will not exactly follow the path.
-See for example the red path in the following visualisation;
+Cells will follow paths within this 2D visualisation, but due to heterogeneity, will not exactly lie on this "average" path.
+See for example the red path in the following visualisation:
 
@@
 
-You don't have to define an end or start point. When drawing cycles, splits or convergences, two points don't have to lie at exactly the same position, as they will be mergen computationally.
+You don't have to define an end or start point. When drawing cycles, splits or convergences, two points don't have to lie at exactly the same position, as they will be merged computationally.
 
 When ready, send the svg file back to wouter.saelens@ugent.be
+
+Thank you!

### 2018-06-27 R/google_scholar.R

@@ -1,68 +1,0 @@
-#' Find the number of citations for a given google scholar id
-#'
-#' @param cluster_id The Google Scholar cluster id to query
-#'
-#' @export
-#'
-#' @importFrom httr GET add_headers
-#' @importFrom xml2 read_html
-#' @importFrom rvest html_nodes html_attr html_text
-#' @importFrom stats runif
-google_scholar <- function(cluster_id) {
-  base_url <- "https://scholar.google.com/scholar"
-
-  Sys.sleep(stats::runif(1, 0.05, .1))
-
-  random_user_agent <- paste0(sample(c("Mozilla", "Chrome"), 1), "/", sample.int(100, 1), ".", sample.int(100, 1))
-
-  site_html <- httr::GET(base_url, query = list(hl = "en", cluster = cluster_id, pagesize = "100", view_op = "list_works", sortby = "pubdate"), httr::add_headers("user_agent" = random_user_agent))
-  webpage <- xml2::read_html(site_html)
-
-  results <- webpage %>% html_nodes(".gs_ri")
-
-  map_df(results, function(result) {
-    title <- result %>%
-      html_nodes(".gs_rt") %>%
-      html_text() %>%
-      str_replace_all("\\[[^\\]]*\\] *", "")
-    url <- result %>%
-      html_nodes(".gs_rt a") %>%
-      html_attr(name = "href")
-    text_fields <- result %>%
-      html_nodes(".gs_fl a") %>%
-      html_text()
-    num_citations <- text_fields[grepl("Cited by", text_fields)] %>%
-      str_replace("Cited by ", "") %>%
-      as.integer() %>%
-      ifelse(length(.) == 0, NA, .)
-    web_of_science <- text_fields[grepl("Web of Science", text_fields)] %>%
-      str_replace("Web of Science: ", "") %>%
-      as.integer() %>%
-      ifelse(length(.) == 0, NA, .)
-
-    data_frame(
-      cluster_id,
-      title,
-      url,
-      num_citations,
-      web_of_science
-    )
-  })
-}
-
-#' Find the number of citations for a given google scholar id
-#'
-#' @param cluster_id The Google Scholar cluster id to query
-#'
-#' @export
-#'
-#' @importFrom xml2 read_html
-#' @importFrom rvest html_nodes html_attr html_text
-google_scholar_num_citations <- function(cluster_id) {
-  df <- google_scholar(cluster_id)
[…]

### 2018-06-27 analysis/analyses/4-method_characterisation/10_implementation_supnote.R

@@ -1,44 +1,0 @@
-library(googlesheets)
-library(tidyverse)
-library(dynalysis)
-
-experiment("4-method_characterisation")
-
-methods <- read_rds(derived_file("methods_evaluated.rds"))
-methods$run_fun <- paste0("ti_", methods$method_id) %>% map(get, "package:dynmethods") %>% map(~.()) %>% map("run_fun")
-
-methods$ti_methods <- paste0("ti_", methods$method_id) %>% map(get, "package:dynmethods") %>% map(~.())
-methods$inputs <- methods$ti_methods %>% map("run_fun") %>% map(formals)
-methods$par_set <- methods$ti_methods %>% map("par_set")
-
-get_input_expression <- function(inputs) {
-  possible_inputs <- c("expression", "counts", "task")
-  intersect(names(inputs), possible_inputs) %>% first()
-}
-methods$input_expression <- map_chr(methods$inputs, get_input_expression)
-
-param_print_functions <- list(
-  discrete = function(x) {
-    glue::collapse(x$values, ", ")
-  },
-  integer = function(x) {
-    seq(x$lower, x$upper) %>% glue::collapse(", ") %>% str_replace("([0-9]*, [0-9]*, )[0-9, ]*(, [0-9]*)", "\\1...\\2")
-  },
-  numeric = function(x) {
-    glue::glue("\\[{x$lower}, {x$upper}\\]")
-  }
-)
-
-
-create_detectable_trajectory_types_table <- function(implementation) {
-  trajectory_types %>%
-    filter(directedness == "undirected") %>%
-
-}
-
-
-create_method_table <- function(implementation) {
-
-}
-
-

### 2018-06-27 analysis/analyses/4-method_characterisation/1_update_citations.R

@@ -1,22 +1,0 @@
-# This file will update the number of citations in the google sheets
-library(tidyverse)
-library(googlesheets)
-library(dynalysis)
-
-experiment("4-method_characterisation")
-
-# Downloading -----------------------
-# # If it's your first time running this script, run this:
-# gs_auth()
-
-sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE")
-
-methods <- sheet %>%
-  gs_read(ws = "Implementations", col_types = cols(gscholar_cluster_id = "c"), skip = 1)
-
-ncitations <- pbapply::pbsapply(methods$gscholar_cluster_id, google_scholar_num_citations)
-# ncitations <- pbapply::pbsapply(methods$doi, rcrossref::cr_citation_count)
-ncitations[is.na(ncitations)] <- ""
-
-gs_edit_cells(sheet, ws = "Implementations", input = ncitations, byrow = FALSE, anchor = which(colnames(methods) == "ncitations") %>% {paste0(LETTERS[floor(./26)], LETTERS[. %% 26], "3")})
-

### 2018-06-27 manuscript/setup.R

@@ -1,0 +1,142 @@
+
+# load libraries
+library(dynalysis)
+library(tidyverse)
+library(cowplot)
+
+# refering to figures & supplementary figures
+refs <- tibble(ref_id = character(), name = character(), ref_type = character())
+ref <- function(ref_type, ref_id, suffix = "", anchor = FALSE, pattern = "[**{ref_full_name}**](#{ref_type}_{ref_id})") {
+  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
+    refs <<- refs %>% bind_rows(tibble(
+      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
+      ref_type = ref_type,
+      ref_id = ref_id
+    ))
+  }
+  ref_name <- refs %>% filter(ref_id == !!ref_id) %>%
+    pull(name)
+  ref_full_name <- paste0(ref_name, glue::collapse(suffix, ','))
+  ref <- pritt(pattern)
+  if (anchor)  ref <- pritt("{ref}{anchor(ref_type, ref_id)}")
+  ref
+}
+anchor <- function(ref_type, ref_id) {
+  case_when(
+    params$table_format == "html" ~ pritt("<a name = '{ref_type}_{ref_id}'></a>"),
+    params$table_format == "latex" ~ pritt("\\protect\\hypertarget{{{ref_type}_{ref_id}}}{{}}")
+  )
+}
+create_names <- list(
+  sfig = function(i) pritt("Supplementary Figure {i}"),
+  fig = function(i) pritt("Figure {i}"),
+  snote = function(i) pritt("Supplementary Note {i}"),
+  table = function(i) pritt("Table {i}"),
+  stable = function(i) pritt("Supplementary Table {i}"),
+  section = function(i) pritt("")
+)
+sfigs <- tibble(ref_id = character(), fig_path = character(), caption_main = character(), caption_text = character(), width = numeric(), height = numeric())
+add_sfig <- function(fig_path, ref_id, caption_main, caption_text, width = 15, height = 10) {
+  sfigs <<- sfigs %>% add_row(
+    ref_id = ref_id,
+    fig_path = fig_path,
+    caption_main = caption_main,
+    caption_text = caption_text,
+    width = width,
+    height = height
+  )
+}
+
+figs <- tibble(ref_id = character(), fig_path = character(), caption_main = character(), caption_text = character(), width = numeric(), height = numeric())
[…]

### 2018-06-27 package/tests/testthat.R

@@ -1,0 +1,10 @@
+library(testthat)
+library(dynutils)
+library(dplyr)
+library(ggplot2)
+library(purrr)
+
+Sys.setenv("R_TESTS" = "")
+
+test_check("dynalysis")
+

### 2018-06-27 scripts/1-datasets/real/1_run_all_datasets.R

@@ -1,0 +1,39 @@
+library(dynalysis)
+
+experiment("1-datasets/real/run_all_datasets")
+
+dataset_scripts <- list.files(path = "analysis/analyses/1-datasets/real", pattern = "^dataset_.*\\.R", full.names = TRUE)
+
+remote <- FALSE
+
+if (remote) {
+  ## Local
+  for (scr in dataset_scripts) {
+    do.call(source, list(scr))
+  }
+} else {
+  ## Remote
+  # Make sure all packages are installed on the cluster; i.e. GEOquery, MultiAssayExperiment, tidyverse, and dynalysis.
+  handle <- qsub::qsub_lapply(
+    X = dataset_scripts,
+    qsub_environment = list2env(list()),
+    qsub_config = qsub::override_qsub_config(
+      name = "dynreal",
+      memory = "30G",
+      wait = FALSE,
+      r_module = NULL,
+      execute_before = "",
+      stop_on_error = FALSE,
+      remove_tmp_folder = FALSE
+    ),
+    qsub_packages = c("GEOquery", "MultiAssayExperiment", "tidyverse", "dynalysis"),
+    FUN = function(dataset_script) {
+      cat("Running ", sQuote(dataset_script), "\n", sep = "")
+      oldwd <- getwd()
+      setwd("/group/irc/shared/dynalysis/")
+      source(paste0(dynalysis::get_dynalysis_folder(), "/", dataset_script))
+      setwd(oldwd)
+      TRUE
+    }
+  )
+}

### 2018-06-27 scripts/1-datasets/real/dataset_conquer_varia.R

@@ -1,0 +1,209 @@
+library(tidyverse)
+library(dynalysis)
+library(dynwrap)
+
+dataset_preprocessing("real/conquer_varia")
+
+requireNamespace("MultiAssayExperiment")
+
+addcols <- function(x) x %>% mutate(length = 1, directed = TRUE)
+
+conquer_infos <- list(
+  list(
+    id = "real/cell-cycle_buettner",
+    rds_name = "EMTAB2805",
+    milestone_source = "cell_cycle_stage",
+    milestone_network = tribble(
+      ~from, ~to,
+      "G1", "S",
+      "S", "G2M",
+      "G2M", "G1"
+    ) %>% addcols,
+    source_id = "conquer"
+  ),
+  list(
+    id = "real/human-embryos_petropoulos",
+    rds_name = "EMTAB3929",
+    milestone_source = "Characteristics.developmental.stage.",
+    milestone_network = tribble(
+      ~from, ~to,
+      "embryonic day 3", "embryonic day 4",
+      "embryonic day 4", "embryonic day 5",
+      "embryonic day 5", "embryonic day 6",
+      "embryonic day 6", "embryonic day 7"
+    ) %>% addcols,
+    source_id = "conquer"
+  ),
+  list(
+    id = "real/NKT-differentiation_engel",
+    rds_name = "GSE74596",
+    milestone_source = "characteristics_ch1.5",
+    milestone_network = tribble(
+      ~from, ~to,
+      "vα14 inkt thymocyte subset: NKT0", "vα14 inkt thymocyte subset: NKT1",
+      "vα14 inkt thymocyte subset: NKT0", "vα14 inkt thymocyte subset: NKT17",
+      "vα14 inkt thymocyte subset: NKT0", "vα14 inkt thymocyte subset: NKT2"
+    ) %>% addcols,
+    source_id = "conquer",
+    remove_spike_ins = TRUE
+  ),
+  list(
+    id = "real/cell-cycle_leng",
+    rds_name = "GSE64016",
+    milestone_source = function(cell_info) gsub("(.*)_.*", "\\1", as.character(cell_info$title)),
+    milestone_network = tribble(
+      ~from, ~to,
+      "G1", "S",
+      "S", "G2",
+      "G2", "G1"
+    ) %>% addcols,
+    remove_spike_ins = TRUE
+  ),
+  list(
+    id = "real/mesoderm-development_loh",
[…]

### 2018-06-27 scripts/1-datasets/real/dataset_cortical-interneuron-differentiation_frazer.R

@@ -1,0 +1,55 @@
+library(dynalysis)
+library(tidyverse)
+options('download.file.method.GEOquery'='curl')
+
+dataset_preprocessing("real/cortical-interneuron-differentiation_frazer")
+
+txt_location <- download_dataset_file(
+  "GSE90860_SupplementaryData1_revised_julien.tsv.gz",
+  "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE90860&format=file&file=GSE90860%5FSupplementaryData1%5Frevised%5Fjulien%2Etsv%2Egz"
+)
+
+counts_df <- read_tsv(txt_location)
+counts <- counts_df %>% select(-X1) %>% as.matrix %>% t %>% magrittr::set_colnames(counts_df$X1)
+
+geo <- GEOquery::getGEO(GEO = "GSE90860", destdir = dataset_preproc_file(""))
+cell_info <- geo[[1]] %>%
+  Biobase::phenoData() %>%
+  as("data.frame") %>%
+  mutate(
+    cell_id = rownames(counts),
+    milestone_id = paste0(gsub("age: (.*)", "\\1", characteristics_ch1), "#", gsub("assigned_subgroup: (.*)", "\\1", characteristics_ch1.1))
+  ) %>%
+  select(cell_id, milestone_id) %>%
+  mutate_all(funs(as.character))
+
+milestone_network <- tribble(
+  ~from, ~to, ~length,
+  "E18#0", "P2#1", 4,
+  "E18#0", "P2#2", 4,
+  "E18#0", "P2#3", 4,
+  "P2#1", "P5#1", 3,
+  "P2#2", "P5#2", 3,
+  "P2#3", "P5#3", 3
+) %>% mutate(directed = TRUE)
+milestone_ids <- unique(c(milestone_network$from, milestone_network$to))
+
+cell_info <- cell_info %>% filter(milestone_id %in% milestone_ids)
+counts <- counts[cell_info$cell_id, ]
+cell_ids <- cell_info$cell_id
+
+cell_grouping <- cell_info %>% select(cell_id, milestone_id) %>% rename(group_id = milestone_id)
+milestone_percentages <- cell_grouping %>% rename(milestone_id=group_id) %>% mutate(percentage=1)
+
+feature_info <- tibble(feature_id = colnames(counts))
+
+preprocess_dataset(
+  counts = counts,
+  cell_ids = cell_ids,
+  milestone_ids = milestone_ids,
[…]

### 2018-06-27 scripts/1-datasets/real/dataset_dentate-gyrus-neurogenesis_hochgerner.R

@@ -1,0 +1,72 @@
+library(dynalysis)
+library(tidyverse)
+library(GEOquery)
+library(Biobase)
+options('download.file.method.GEOquery'='curl')
+
+dataset_preprocessing("real/dentate-gyrus-neurogenesis_hochgerner")
+
+
+# get cell info
+geo <- GEOquery::getGEO("GSE95315", destdir = dataset_preproc_file(""))[[1]]
+
+cell_info_all <- phenoData(geo) %>% as("data.frame") %>% select(title, characteristics_ch1.2, characteristics_ch1.4) %>%
+  rename(
+    cell_id = title,
+    time_id = characteristics_ch1.2,
+    population_id = characteristics_ch1.4
+  ) %>%
+  mutate_if(is.factor, as.character) %>%
+  mutate(population_id = gsub("cell cluster: (.*)", "\\1", population_id)) %>%
+  mutate(time_id = gsub("postnatal day: (.*)", "\\1", time_id))
+
+# merge some clusters as in the paper
+merge_clusters <- setNames(unique(cell_info_all$population_id), unique(cell_info_all$population_id))
+merge_clusters["nIPC2"] <- "nIPC"
+cell_info_all$population_id <- merge_clusters[cell_info_all$population_id]
+
+# load and process all counts
+tab_location <- download_dataset_file(
+  "GSE95315_10X_expression_data.tab.gz",
+  "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95315&format=file&file=GSE95315%5F10X%5Fexpression%5Fdata%2Etab%2Egz"
+)
+
+counts_all_preproc <- read_tsv(tab_location)
+gene_ids <- counts_all_preproc$cellid
+counts_all_preproc <- as.matrix(counts_all_preproc[, -1])
+rownames(counts_all_preproc) <- gene_ids
+counts_all <- t(counts_all_preproc)
+
+# construct dataset
+milestone_network <- tribble(
+  ~from, ~to, ~length, ~directed,
+  "nIPC", "Neuroblast", 1, T,
+  "Neuroblast", "Neuroblast2", 1, T,
+  "Neuroblast2", "Granule-immature", 1, T,
+  "Granule-immature", "Granule-mature", 1, T
+)
+
+milestone_ids <- unique(c(milestone_network$from, milestone_network$to))
+
+cell_info <- cell_info_all %>% slice(match(rownames(counts_all), cell_id)) %>%
+  rename(milestone_id = population_id) %>%
[…]

### 2018-06-27 scripts/1-datasets/real/dataset_germline-human_li.R

@@ -1,0 +1,125 @@
+library(tidyverse)
+library(dynalysis)
+
+dataset_preprocessing("real/germline-human_li")
+
+tar_location <- download_dataset_file(
+  "GSE86146_RAW.tar",
+  "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE86146&format=file"
+)
+
+utils::untar(tar_location, exdir = dataset_preproc_file(""))
+
+allcounts <- list.files(dataset_preproc_file("")) %>%
+  keep(~ startsWith(., "GSM")) %>%
+  map(~ read_tsv(dataset_preproc_file(.), col_types = cols(.default = "d", Gene = "c")) %>% gather(Sample, Expression, -Gene)) %>%
+  bind_rows() %>%
+  spread(Gene, Expression) %>%
+  as.data.frame() %>%
+  magrittr::set_rownames(., .$Sample) %>%
+  select(-Sample) %>%
+  as.matrix
+
+mmc2_location <- download_dataset_file(
+  "mmc2.xlsx",
+  "http://www.sciencedirect.com/science/MiamiMultiMediaURL/1-s2.0-S1934590917300784/1-s2.0-S1934590917300784-mmc2.xlsx/274143/html/S1934590917300784/139b580f8ca22e965c646ac00b373e93/mmc2.xlsx"
+)
+
+allcell_info <- readxl::read_xlsx(mmc2_location, sheet = 3) %>%
+  rename(cell_id = Cell, cluster=Cluster) %>%
+  filter(cell_id %in% rownames(allcounts)) %>%
+  mutate(
+    week = as.numeric(gsub("[FM]_(.+)W_.*", "\\1", cell_id)),
+    gender = gsub("([FM]).*", "\\1", cell_id),
+    type = gsub(".*_(Soma|FGC).*", "\\1", cluster),
+    weekgendertype = paste0(gender, "#", week, "#", type)
+  )
+
+milestone_order_to_network <- function(order) {
+  tibble(from = order[-length(order)], to = order[-1])
+}
+
+settings <- list(
+  list(
+    id = "real/germline-human-male_li",
+    milestone_network = tribble(
+      ~from, ~to,
+      "Male_FGC#1", "Male_FGC#2",
+      "Male_FGC#2", "Male_FGC#3"
+    ),
+    milestone_source = "cluster"
+  ),
+  list(
+    id = "real/germline-human-female_li",
+    milestone_network = tribble(
+      ~from, ~to,
[…]

### 2018-06-27 scripts/1-datasets/real/dataset_hematopoiesis_olsson.R

@@ -1,0 +1,101 @@
+library(tidyverse)
+library(dynalysis)
+options('download.file.method.GEOquery'='curl')
+
+dataset_preprocessing("real/hematopoiesis_olsson")
+
+files_df <- tribble(
+  ~location, ~url,
+  "GSE70240_Gmp.txt.gz", "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE70240&format=file&file=GSE70240%5FGmp%2Etxt%2Egz",
+  "GSE70243_LK.CD34+.txt.gz", "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE70243&format=file&file=GSE70243%5FLK%2ECD34%2B%2Etxt%2Egz",
+  "GSE70244_Lsk.txt.gz", "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE70244&format=file&file=GSE70244%5FLsk%2Etxt%2Egz",
+  "GSE70236_Cmp.txt.gz", "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE70236&format=file&file=GSE70236%5FCmp%2Etxt%2Egz"
+) %>%
+  rowwise() %>%
+  mutate(
+    preproc_loc = download_dataset_file(location, url)
+  ) %>%
+  ungroup()
+
+allexpression <- files_df$preproc_loc %>%
+  map(~ read_tsv(., col_types = cols(uid = "c", .default = "d")) %>% gather(feature, expression, -uid)) %>%
+  bind_rows() %>%
+  spread(feature, expression) %>%
+  as.data.frame() %>%
+  magrittr::set_rownames(NULL) %>%
+  column_to_rownames("uid") %>%
+  as.matrix %>%
+  t
+
+txt_location <- download_dataset_file(
+  "nature19348-f1.xlsx",
+  "https://images.nature.com/original/nature-assets/nature/journal/v537/n7622/source_data/nature19348-f1.xlsx"
+)
+
+small_expression <- readxl::read_xlsx(txt_location, 1)
+subclusters <- small_expression[1, -c(1, 2)] %>% as.list() %>% unlist()
+subclusters_mapper <- c("HSCP-1", "HSCP-2", "Meg", "Eryth", "Multi-Lin", "MDP", "Mono", "Gran", "Myelocyte")
+subclusters <- subclusters_mapper[subclusters] %>% set_names(names(subclusters))
+
+allcell_info <- tibble(
[…]

### 2018-06-27 scripts/1-datasets/synthetic/0_load_on_cluster.R

@@ -1,0 +1,12 @@
+saveRDS(qsub_environment, "env.rds")
+qsub:::rsync_remote("", "env.rds", "prism", "/group/irc/personal/wouters/env.rds")
+saveRDS(paramsets, "paramsets.rds")
+qsub:::rsync_remote("", "paramsets.rds", "prism", "/group/irc/personal/wouters/paramsets.rds")
+
+
+##
+
+list2env(as.list(readRDS("env.rds")), .GlobalEnv)
+paramsets <- readRDS("paramsets.rds")
+prepare_environment()
+params <- paramsets[[1]]

### 2018-06-27 scripts/4-method_characterisation/10_implementation_supnote.R

@@ -1,0 +1,44 @@
+library(googlesheets)
+library(tidyverse)
+library(dynalysis)
+
+experiment("4-method_characterisation")
+
+methods <- read_rds(derived_file("methods_evaluated.rds"))
+methods$run_fun <- paste0("ti_", methods$method_id) %>% map(get, "package:dynmethods") %>% map(~.()) %>% map("run_fun")
+
+methods$ti_methods <- paste0("ti_", methods$method_id) %>% map(get, "package:dynmethods") %>% map(~.())
+methods$inputs <- methods$ti_methods %>% map("run_fun") %>% map(formals)
+methods$par_set <- methods$ti_methods %>% map("par_set")
+
+get_input_expression <- function(inputs) {
+  possible_inputs <- c("expression", "counts", "task")
+  intersect(names(inputs), possible_inputs) %>% first()
+}
+methods$input_expression <- map_chr(methods$inputs, get_input_expression)
+
+param_print_functions <- list(
+  discrete = function(x) {
+    glue::collapse(x$values, ", ")
+  },
+  integer = function(x) {
+    seq(x$lower, x$upper) %>% glue::collapse(", ") %>% str_replace("([0-9]*, [0-9]*, )[0-9, ]*(, [0-9]*)", "\\1...\\2")
+  },
+  numeric = function(x) {
+    glue::glue("\\[{x$lower}, {x$upper}\\]")
+  }
+)
+
+
+create_detectable_trajectory_types_table <- function(implementation) {
+  trajectory_types %>%
+    filter(directedness == "undirected") %>%
+
+}
+
+
+create_method_table <- function(implementation) {
+
+}
+
+

### 2018-06-27 scripts/4-method_characterisation/4_process_combined.R

@@ -1,0 +1,41 @@
+library(tidyverse)
+library(dynalysis)
+library(googlesheets)
+
+experiment("4-method_characterisation")
+
+implementations_tidy <- read_rds(derived_file("implementations_tidy.rds"))
+methods_tidy <- read_rds(derived_file("methods_tidy.rds"))
+
+# combine with qc scores
+implementation_qc_scores <- readRDS(derived_file("implementation_qc_scores.rds"))
+
+# merge qc scores with implementations tibble
+implementations <- implementations_tidy %>%
+  select(-matches("qc_score")) %>%
+  left_join(implementation_qc_scores, "implementation_id")
+
+methods <- methods_tidy %>%
+  select(-matches("qc_score")) %>%
+  left_join(implementation_qc_scores, "implementation_id")
+
+# filter evaluated
+methods$evaluated <- (methods$evaluation_inclusion == "TRUE") %>% coalesce(FALSE)
+implementations$evaluated <- (implementations$evaluation_inclusion == "TRUE") %>% coalesce(FALSE)
+
+methods_evaluated <- methods %>%
+  filter(evaluated)
+
+implementations_evaluated <- implementations %>%
+  filter(evaluated)
+
+# save
+write_rds(implementations, derived_file("implementations.rds"))
+write_rds(implementations_evaluated, derived_file("implementations_evaluated.rds"))
+write_rds(methods, derived_file("methods.rds"))
+write_rds(methods_evaluated, derived_file("methods_evaluated.rds"))
+
+write_rds(implementations, result_file("implementations.rds"))
+write_rds(implementations_evaluated, result_file("implementations_evaluated.rds"))
+write_rds(methods, result_file("methods.rds"))
+write_rds(methods_evaluated, result_file("methods_evaluated.rds"))

### 2018-06-27 scripts/4-method_characterisation/6_implementation_characterisation.R

@@ -1,0 +1,349 @@
+#   ____________________________________________________________________________
+#   Loading                                                                 ####
+
+library(tidyverse)
+library(dynalysis)
+library(cowplot)
+
+experiment("4-method_characterisation")
+
+implementations_evaluated <- read_rds(derived_file("implementations_evaluated.rds"))
+implementations <- read_rds(derived_file("implementations.rds"))
+implementation_qc <- read_rds(derived_file("implementation_qc.rds"))
+
+implementation_id_to_name <- setNames(implementations$implementation_name, implementations$implementation_id)
+
+start_date <- as.Date("2014-01-01")
+end_date <- as.Date("2018-09-01")
+
+
+implementations$publication_date <- implementations$publication_date + runif(length(implementations$publication_date), 0, 1)
+
+
+#   ____________________________________________________________________________
+#   Figure generation                                                       ####
+
+##  ............................................................................
+##  Publication timeline of implementations                                         ####
+
+implementation_publication_data <- implementations %>%
+  gather("publication_type", "publication_date", c(publication_date, preprint_date)) %>%
+  select(implementation_id, publication_type, publication_date) %>%
+  drop_na() %>%
+  mutate(counter = 1)
+
+# add published preprints
+implementation_publication_data <- implementation_publication_data %>%
+  group_by(implementation_id) %>%
+  filter(n() == 2) %>%
+  filter(publication_type == "publication_date") %>%
+  mutate(publication_type = "preprint_date", counter = -1) %>%
+  bind_rows(implementation_publication_data) %>%
+  mutate(publication_type = factor(publication_type, levels = c("publication_date", "preprint_date")))
+
+# summarise by publication type how many implementations there are
+publication_cumulative_by_type <- implementation_publication_data %>%
+  group_by(publication_type) %>%
+  arrange(publication_date) %>%
+  mutate(n_implementations = cumsum(counter)) %>%
+  ungroup()
+# add first and last
+earliest_date <- start_date
+latest_date <- end_date
+earliest <- publication_cumulative_by_type %>%
+  group_by(publication_type) %>%
+  arrange(publication_date) %>%
+  filter(row_number() == 1) %>%
[…]

### 2018-06-27 scripts/6-benchmark/3-generate_figures.R

@@ -1,0 +1,131 @@
+library(dynalysis)
+library(tidyverse)
+library(dynplot)
+
+experiment("5-optimise_parameters/3-evaluate_parameters")
+
+# Rsync latest results!
+# qsub:::rsync_remote(
+#   remote_src = "prism",
+#   path_src = paste0("/group/irc/shared/dynalysis/analysis/data/derived_data/", getOption("dynalysis_experiment_id"), "/"),
+#   remote_dest = "",
+#   path_dest = derived_file(),
+#   verbose = T
+# )
+
+############################################################
+############### PART THREE: GENERATE FIGURES ###############
+############################################################
+
+olist <- read_rds(derived_file("outputs_postprocessed.rds"))
+
+# get ordering of methods
+method_ord <- olist$outputs_summtrajtype_totalsx2 %>%
+  filter(task_source == "mean", trajectory_type == "overall") %>%
+  arrange(desc(harm_mean)) %>%
+  .$method_name
+
+# create method_name_f factor in all data structures
+for (oname in str_subset(names(olist), "outputs")) {
+  olist[[oname]] <- olist[[oname]] %>%
+    mutate(method_name_f = factor(method_name, levels = rev(method_ord)))
+}
+
+# load all outputs in environment and remove outputs_list
+list2env(olist, environment())
+rm(olist)
+
+# collect which methods use which prior information
+prior_df <- outputs_ind %>% select(method_name, prior_str) %>% distinct()
+
+
+############### OVERALL COMPARISON ###############
+metr_lev <- c(
+  "harm_mean", "norm_correlation", "norm_edge_flip", "norm_rf_mse",
+  "real", "synthetic", "time_method", "rank_time_method",
+  "pct_errored", "pct_time_exceeded", "pct_memory_exceeded", "num_setseed_calls"
+)
+
+oc1 <-
+  outputs_summtrajtype_totalsx2 %>%
+  filter(task_source == "mean", trajectory_type_f == "overall") %>%
+  select(method_name, method_short_name, method_name_f, paramset_id, one_of(metr_lev)) %>%
+  gather(metric, score, -method_name:-paramset_id)
+oc2 <-
+  outputs_summtrajtype_totalsx2 %>%
[…]

### 2018-06-27 scripts/manual_ti/process_svg.R

@@ -1,0 +1,240 @@
+library(dynalysis)
+library(png)
+library(tidyverse)
+library(GNG)
+library(tidygraph)
+library(ggraph)
+
+experiment("manual_ti")
+
+
+##  ............................................................................
+##  Load run                                                                ####
+
+run_id <- "mds_robrechtc_1"
+run <- read_rds(derived_file(paste0(run_id, ".rds")))
+
+##  ............................................................................
+##  Load bitmap                                                             ####
+
+processed_file <- derived_file(c(run_id, '_run', '.png'))
+processx::run(commandline=dynutils::pritt("inkscape -z {derived_file(c(run_id, '.svg'))} -e={processed_file} -d=100"))
+
+png <- readPNG(processed_file)
+red <- png[, , 1] - png[, , 2]
+
+
+##  ............................................................................
+##  Get coordinates from bitmap                                             ####
+coordinates <- red %>%
+  reshape2::melt(varnames=c("y", "x"), value.var=value) %>%
+  as_tibble() %>%
+  filter(value > 0.5)
+
+box_height <- dim(red)[[1]]/run$nrow
+box_width <- dim(red)[[2]]/run$ncol
+
+coordinates <- coordinates %>%
+  mutate(
+    col = floor((x-1)/box_width) + 1,
+    row = floor((y-1)/box_height),
+    box_id =  (row * run$ncol) + col
+  )
+coordinates %>% filter(y<200) %>%  ggplot() + geom_point(aes(x, -y, color=factor(box_id)))
+
+
+
+##  ............................................................................
+##  Generate initial GNG from coordinates                                   ####
+process_coordinates <- function(coordinates) {
+  data <- coordinates %>% select(x, y) %>% as.matrix()
+  result <- GNG::gng(data, max_nodes=50, assign_cluster=F)
+
+  graph <- result$edges %>% as_tbl_graph() %>%
+    activate(nodes) %>%
+    left_join(result$nodes %>% mutate(index=as.character(index)) %>% rename(id=name), by=c("name"="index")) %>%
+    left_join(result$node_space %>% as.data.frame() %>% tibble::rownames_to_column("id"), "id")
+  graph
+}
+
+coordinates_split <- coordinates %>% split(coordinates$box_id)
[…]

### 2018-06-27 scripts/manual_ti/process_svg_paths.R

@@ -1,0 +1,314 @@
+library(tidyverse)
+library(cowplot)
+library(googlesheets)
+library(dynalysis)
+
+library(xml2)
+library(stringr)
+
+experiment("manual_ti")
+
+run_id <- "mds_robrechtc_1"
+run <- read_rds(derived_file(paste0(run_id, ".rds")))
+run$spaces <- run$spaces[[1]]
+
+run$spaces$id <- tasks$id[pmatch(run$spaces$id %>% gsub(".*/(.*)", "\\1", .), tasks$id %>% gsub(".*/(.*)", "\\1", .))]
+
+
+
+##  ............................................................................
+##  Process svg                                                             ####
+# make sure to
+# - remove beziers using 1. select all paths 2. extensions -> modify paths -> flatten bezier 3. Move all up and down to refresh data
+# - The images dimensions are in points
+
+svg <- read_xml(derived_file(glue::glue("{run$run_id}.svg")))
+layer <- svg %>% xml_find_first("//svg:g[@inkscape:label = 'segments']")
+paths_svg <- layer %>% xml_find_all("svg:path") %>% xml_attr("d")
+
+split_coords <- function(x) as.numeric(str_split(x, ",")[[1]])
+
+# calculate dimensions of svg and layer transformation
+dimensions <- c(
+  svg %>% xml_find_first("svg:g[@inkscape:label = 'back']") %>% xml_child() %>% xml_attr("width") %>% as.numeric(),
+  svg %>% xml_find_first("svg:g[@inkscape:label = 'back']") %>% xml_child() %>% xml_attr("height") %>% as.numeric()
+)
+if (any(is.na(dimensions))) {stop("Dimensions should be in pt!!")}
+base_coords <- layer %>% xml_attr("transform") %>% str_extract("\\(.*\\)") %>% str_sub(2, -2) %>% split_coords()
+if(is.na(base_coords)) base_coords <- c(0, 0)
+
+
+
+### . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . ..
+### Process paths                                                           ####
+path <- paths_svg[[1]]
+
+extract_first_consecutive <- function(x) {
+  if(all(diff(x) == 1)) {
+    x
+  } else {
+    x[1:first(which(diff(x) != 1))]
+  }
+}
+
+process_path <- function(path, base_coords = c(0, 0)) {
[…]

### 2018-06-27 scripts/old6-performance_predictors/component_importance.R

@@ -1,0 +1,211 @@
+library(cowplot)
+library(tidyverse)
+library(dynalysis)
+
+experiment("6-performance_predictors")
+
+read_rds(derived_file("evaluation_algorithm.rds", "5-optimise_parameters/10-aggregations")) %>% list2env(.GlobalEnv)
+
+method_components <- read_rds(derived_file("method_components.rds", experiment_id = "4-method_characterisation"))
+component_categories <- read_rds(derived_file("component_categories.rds", experiment_id = "4-method_characterisation"))
+component_category_colors <- read_rds(derived_file("component_category_colors.rds", experiment_id = "4-method_characterisation"))
+
+#   ____________________________________________________________________________
+#   Component importance                                                    ####
+
+
+# join categories as regular components
+method_components_category <- method_components %>%
+  left_join(component_categories, "component_id") %>%
+  filter(!is.na(category)) %>%
+  group_by(method_id, category) %>%
+  summarise() %>%
+  ungroup() %>%
+  rename(component_id = category) %>%
+  mutate(component_id = paste0("any ", component_id))
+
+method_components <- bind_rows(
+  method_components,
+  method_components_category
+)
+component_categories <- bind_rows(
+  component_categories,
+  tibble(category = unique(component_categories$category), component_id = paste0("any ", category))
+)
+
+components <- tibble(component_id = unique(method_components$component_id))
+
+method_features_components <- method_components %>%
+  mutate(method_short_name = factor(method_id, levels = overall_scores$method_short_name)) %>%
+  select(-method_id) %>%
+  group_by(method_short_name, component_id) %>%
+  filter(row_number() == 1) %>%
+  ungroup() %>%
+  mutate(contains = T) %>%
+  spread("component_id", "contains", FALSE, drop = F) %>%
+  slice(match(overall_scores$method_short_name, method_short_name))
+
+is_variable_feature <- function(x) sum(table(x) > 1) >= 4
+method_features <- method_features_components %>%
+  mutate(method_short_name = method_features_components$method_short_name)
+
+components <- components %>%
+  filter(component_id %in% colnames(method_features)) %>%
+  mutate(n = colSums(method_features[,component_id])) %>%
+  filter(n >= 4)
+
+components$method_short_names <- map(components$component_id, ~as.character(method_features$method_short_name[which(method_features[[.]])]))
+
[…]

### 2018-06-27 scripts/3-metric_characterisation/1old-generate_toys.Rmd
reorganise dynalysis

--- 2018-02-07 analysis/analyses/3-score_rules/1-generate_toys.Rmd
@@
 ```{r}
-source("analysis/analyses/3-score_rules/generation.R")
-source("analysis/analyses/3-score_rules/perturbation.R")
+source("analysis/analyses/3-metric_characterisation/generation.R")
+source("analysis/analyses/3-metric_characterisation/perturbation.R")
 
-experiment("3-score_rules/1-generate_toys")
+experiment("3-metric_characterisation/1-generate_toys")
 ```

### 2018-06-27 README.Rmd
update readme

@@
 
-```{r, echo = FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  fig.path = "README-"
-)
-```
+# dynbenchmark
 
-# An evaluation framework for trajectory inference methods
+[![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)
 
-![travis](https://travis-ci.com/Zouter/dynalysis.svg?token=pzNqJjk3yapU1Hk7sDg8&branch=master)
+Scripts to reproduce the manuscript
 
-### Installation
+A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
+Wouter Saelens\*, Robrecht Cannoodt\*, Helena Todorov, Yvan Saeys  
+bioRxiv 276907; doi: https://doi.org/10.1101/276907
 
-You can install dynalysis from github with:
+## Datasets
 
-```{r gh-installation, eval = FALSE}
-install.packages("devtools")
-setRepositories(ind = 1:2)
-devtools::install_github("Zouter/dynalysis", dependencies = T, upgrade = T)
-```
+We include the following datasets. When using any of these datasets, please also cite our preprint: [![DOI](https://zenodo.org/badge/DOI/10.1101/276907.svg)](https://doi.org/10.1101/276907)
 
-### Running
-
-To run locally, install `docker` and run `docker build -t dynalysis .`
-
-### Licenses
-
-**Text and figures:**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
-
-**Code:** See the [DESCRIPTION](DESCRIPTION) file
-
-$\textbf{Data:}$ [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
-
-### Contributions
-
-We welcome contributions from everyone.
-Before you get started, please see our [contributor guidelines](CONTRIBUTING.md).
-Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md).
-By participating in this project you agree to abide by its terms.
+* [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1211533.svg)](https://doi.org/10.5281/zenodo.1211533) Single-cell -omics datasets, both real and synthetic, used to evaluated the trajectory inference methods
+* **Coming soon** Main results of the evaluation, used to rank the methods and construct practical guidelines

### 2018-06-27 manuscript/paper_latex.Rmd
rename all mentions of dynalysis to dynbenchmark

--- 2018-04-05 analysis/paper/paper_latex.Rmd
@@
 ```{r setup, include=FALSE}
-source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
+source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/setup.R"))
 
@@
 We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
+The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynbenchmark](http://www.github.com/dynverse/dynbenchmark)).
 Whenever possible, we preferred to start from the raw counts data.
@@
 if (params$table_format == "html") 
-  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
+  source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/sfigs_stables.R"))
 write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))

### 2018-06-27 manuscript/supplementary_latex.Rmd
rename all mentions of dynalysis to dynbenchmark

--- 2018-04-03 analysis/paper/supplementary_latex.Rmd
@@
 ```{r "supplementary", echo=FALSE, results='asis'}
-source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
+source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/sfigs_stables.R"))
 ```

### 2018-06-27 scripts/3-metric_characterisation/3old-rules.Rmd
rename all mentions of dynalysis to dynbenchmark

--- 2017-12-05 analysis/analyses/3-score_rules/2-rules.Rmd
@@
 knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
-library(dynalysis)
+library(dynbenchmark)
 library(tidyverse)
@@
 ```{r}
-experiment("score_rules")
+experiment("3-metric_characterisation/3-rules")
 
@@
 
-source("analysis/analyses/3-score_rules/rules.R")
+source("analysis/analyses/3-metric_characterisation/rules.R")
 rules_info$check_func <- map(paste0("check_", rules_info$id), get)

### 2018-06-27 package/R/preprocess_dataset.R

@@ -116,7 +116,7 @@ preprocess_dataset <- function(
 
 convert_to_symbol <- function(counts) {
   colnames(counts) <- tibble(gene_id = colnames(counts)) %>%
-    left_join(dynalysis::id_mapper, by = c("gene_id" = "ensembl")) %>%
+    left_join(dynbenchmark::id_mapper, by = c("gene_id" = "ensembl")) %>%
     mutate(gene_id = ifelse(is.na(symbol), gene_id, symbol)) %>%
     pull(gene_id)
   filtered <- names(which(table(colnames(counts)) == 1))

### 2018-06-27 scripts/1-datasets/real/dataset_aging-hsc_kowalczyk.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 options('download.file.method.GEOquery'='curl')
 
 dataset_preprocessing("real/aging-hsc_kowalczyk")

### 2018-06-27 scripts/1-datasets/real/dataset_hematopoiesis_olsson.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 options('download.file.method.GEOquery'='curl')
 
 dataset_preprocessing("real/hematopoiesis_olsson")

### 2018-06-27 scripts/4-method_characterisation/11_create_github_issue.R

@@ -1,6 +1,6 @@
 # code to generate the issues for authors on the dynmethods github
 
-library(dynalysis)
+library(dynbenchmark)
 library(tidyverse)
 library(cowplot)
 library(googlesheets)

### 2018-06-27 scripts/1-datasets/1-download_from_zenodo.R
rename all mentions of dynalysis to dynbenchmark

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 
 experiment("1-datasets")
 
@@ -51,7 +51,7 @@ tasks <- list_as_tibble(map(task_ids, function(task_id) {
     assign("task_id", task_id, env)
     assign("col", col, env)
     task[[col]] <- function() {
-      dynalysis::load_dataset(task_id)[[col]]
+      dynbenchmark::load_dataset(task_id)[[col]]
     }
     environment(task[[col]]) <- env
   }
@@ -70,10 +70,10 @@ tasks %>% map_int(~ pryr::object_size(.) %>% as.integer) %>% sort(decreasing = T
 # upload to remote, if relevant
 # this assumes you have a gridengine cluster at hand
 remote_config <- qsub::override_qsub_config()
-remote_dynalysis_path <- qsub:::run_remote("echo $DYNALYSIS_PATH", remote = remote_config$remote)$stdout
+remote_dynbenchmark_path <- qsub:::run_remote("echo $DYNALYSIS_PATH", remote = remote_config$remote)$stdout
 qsub:::rsync_remote(
   remote_src = "",
   path_src = derived_file(),
   remote_dest = remote_config$remote,
-  path_dest = derived_file() %>% str_replace(get_dynalysis_folder(), remote_dynalysis_path)
+  path_dest = derived_file() %>% str_replace(get_dynbenchmark_folder(), remote_dynbenchmark_path)
 )

### 2018-06-27 scripts/1-datasets/real/dataset_neonatal-inner-ear_burns.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 
 dataset_preprocessing("real/neonatal-inner-ear_burns")
 

### 2018-06-27 scripts/1-datasets/real/dataset_oligodendrocyte-differentiation_marques.R

@@ -1,4 +1,4 @@
-library(dynalysis)
+library(dynbenchmark)
 library(tidyverse)
 options('download.file.method.GEOquery'='curl')
 

### 2018-06-27 scripts/1-datasets/real/dataset_psc-astrocyte-maturation_sloan.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 
 dataset_preprocessing("real/pancreatic-cell-maturation_zhang")
 

### 2018-06-27 scripts/1-datasets/real/dataset_stimulated-dendritic-cells_shalek.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 options('download.file.method.GEOquery'='curl')
 
 dataset_preprocessing("real/stimulated-dendritic-cells_shalek")

### 2018-06-27 scripts/old11-evaluation_robustness/create_figure.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-library(dynalysis)
+library(dynbenchmark)
 
 experiment("11-evaluation_robustness")
 

### 2018-06-27 scripts/generate_hypotheses_mindmap/hypotheses_tikz.tex

  \myhypothesis{ti_evaluation}{ti\_evaluation}{102,102,102}{Dynamic processes can be modelled and studied with TI methods}{* Characterise and organise TI methodology @characterisation\\
    * Thoroughly set up experiment design to ensure correct scientific reasoning @experiment\_design\\
    * Train TI methods on synthetic data @synthetic\_testing\\
    * Evaluate them on biological data @biological\_validation\\
    * Provide reader with a set of instructions on which TI method to use on which kinds of datasets @guidelines}{Some TI methods will perform well on some datasets.\\
    Performance will depend on parameters.}{An extensive TI evaluation will:\\
    * Stimulate the adoption of TI methods\\
    * Stimulate the development of more robust and accurate TI methods\\
    * Stimulate the evolution of the TI to tackle more complex tasks}

  \myhypothesis{characterisation}{characterisation}{52,152,219}{We characterised and organised TI methodology in terms of which biological question is being asked, how a TI method tries to solve this question, and what characteristics it has}{We need to characterise \\
    * the tasks / biological questions @characterise\_tasks\\
    * the platforms @characterise\_platforms\\
    * the methods @characterise\_methods}{}{There is a lot of confusion about what TI is, what problems can be solved with it, and what sort of solution you might expect from a TI method. The characterisation will provide a structure to link everything together and clarify ambiguities}

  \myhypothesis{characterise_tasks}{characterise\_tasks}{52,152,219}{We characterised the different TI tasks}{TODO}{}{}

  \myhypothesis{characterise_platforms}{characterise\_platforms}{52,152,219}{We collected properties of possible datasets and determined which TI task could be used on the dataset}{TODO}{}{}

  \myhypothesis{characterise_methods}{characterise\_methods}{52,152,219}{We characterised TI methods in terms of its inputs, outputs, methodology and type of TI task it tries to solve}{TODO}{}{}

  \myhypothesis{experiment_design}{experiment\_design}{231,76,60}{Our evaluation of TI methods is scientifically sound}{Construct and evaluate the:\\
    * synthetic gold standard\\
    * biological gold standard\\
    * benchmark pipeline}{}{Good and scientifically sound TI evaluation methodology is required in order for the field to progress further}

  \myhypothesis{construct_synthetic_gold}{construct\_synthetic\_gold}{231,76,60}{We generated realistic synthetic expression data of cells as part of a dynamic process}{Determine:\\
    * the gene regulation model @regulation\_model\\
    * the transcription @transcription\_model\\
    * the dynamic process model @dynamic\_process\_model\\
    * the sequencing model @platforms\_model}{}{}

  \myhypothesis{regulation_model}{regulation\_model}{231,76,60}{We use GRNs as a model for gene regulation. These GRNs are derived from real regulatory networks}{TODO}{}{}

  \myhypothesis{transcription_model}{transcription\_model}{231,76,60}{We can generate expression given a GRN}{TODO}{}{}

  \myhypothesis{dynamic_process_model}{dynamic\_process\_model}{231,76,60}{We can translate a dynamic process into a GRN such that generated expression therefrom resembles observations of cells following this dynamic process}{TODO}{}{}

  \myhypothesis{platforms_model}{platforms\_model}{231,76,60}{We determined a model of the technical noise generated by observing the state of a cell through RNA sequencing}{Compare real data across platforms with the simulated data across platforms\\
    - Drop-out rates\\
    - Distributions\\
    - Compare with "ideal" platform}{}{Platforms could induce a bias in the performance of certain methods, so it is important that our platform data is comparable to real platform data\\
    This will allow us to investigate the advantages and disadvantages of different platforms}

  \myhypothesis{generate_synthetic_data}{generate\_synthetic\_data}{231,76,60}{We generated synthetic data in accordance to our models: \\
     * @regulation\_model\\
     * @transcription\_model\\
     * @dynamic\_process\_model\\
     * @platforms\_model}{We simulate using SDEs (with enough noise) and Gillespie SSA (with noise and single-molecule) which takes into account\\
    * Technical bias introduced by platforms @simulating\_platform\_noise\\
    * Biological\\
     - Modular\\
     - Dim red and example of single-cell paths}{N/A}{We need to generate realistic (or at least useful) expression data from GRNs in order to be able to prioritise TI methods for real datasets}

  \myhypothesis{timings_synthetic_generator}{timings\_synthetic\_generator}{231,76,60}{Our SSA implementation is /blazing fast/.\\
    Goal: simulate 100'000 cells with 20'000 genes in <24 hours.\\
    10-fold sampling of single cells can be allowed.}{Perform timings experiments on our implementation of fastgssa versus that of GillespieSSA.}{fastgssa is very faster, but still outputs very similar data for the same given network}{If we want to generate realistic data, we need to be able to simulate as many cells and as many genes as would hopefully occur in near-future experiments}

  \myhypothesis{construct_biological_gold}{construct\_biological\_gold}{231,76,60}{We collected high-quality datasets containing cells along a dynamic process, of which we know the state of each of the cells}{* Collect data\\
    * Perform some kind of tests to ensure there is enough signal-to-noise?}{}{}

  \myhypothesis{benchmarking_pipeline}{benchmarking\_pipeline}{231,76,60}{We use a benchmarking pipeline to train the parameters of the methods and ensure that the datasets the methods are evaluated on are unrelated to avoid overfitting}{For each of the TI tasks, we train the parameters of the methods on synthetic datasets and evaluate them on biological datasets}{}{}

  \myhypothesis{metrics}{metrics}{231,76,60}{We have several good metrics to evaluate different aspects of predicting a good trajectory:\\
      * the main structure of the predicted trajectory should be good   \\
      * the ordering of the cells should be good\\
      * metric with distance from origin}{Unit test metrics with toy examples}{Each metric makes sense theoretically and fulfill the requirements of the toy example, although they can focus different aspects (or levels) of the TI model}{Good metrics are the foundation of a good evaluation}

  \myhypothesis{method_fairness}{method\_fairness}{231,76,60}{We treat each TI method fairly}{* No TI method is left out\\
    * Used implementation from original authors if available\\
    * Made own implementation if necessary; contacted original authors to verify implementation\\
    * Described all sorts of parameters, which will be optimised later}{}{}

  \myhypothesis{parameter_crossvalidation}{parameter\_crossvalidation}{231,76,60}{For each of the tasks, we executed the different methods and trained their parameters using the metrics specified by @metrics}{Train parameters for each of the TI tasks either by generating a grid of parameters or by using a smarter parameter optimisation algorithm}{}{}

  \myhypothesis{benchmarking}{benchmarking}{164,204,46}{We benchmarked TI methods using the real and simulated data and constructed a set of guidelines for the user to follow}{We need to:\\
    * generate synthetic data @generate\_synthetic\_data\\
    * perform parameter cross-validation @parameter\_crossvalidation\\
    * find methods (\& parameters) which work well for certain datasets/task characteristics @characteristics\_vs\_performance\\
    * construct a set of guidelines for the reader with which to decide which method to use @construct\_guidelines}{}{}

  \myhypothesis{characteristics_vs_performance}{characteristics\_vs\_performance}{164,204,46}{We could link performance to dataset and method metadata}{Use dataset and method metadata to determine which methods and parameters work well for certain TI tasks}{We hope to see links between the performance and the metadata}{We need to do this in order to start constructing guidelines}

  \myhypothesis{construct_guidelines}{construct\_guidelines}{164,204,46}{We constructed a set of usage guidelines}{Attempt to construct TI usage guidelines depending on the TI task.}{This will likely not be feasible, but would be extremely useful.}{Guide the biologists in their selection of methods\\
    Guide the bio-informatician in the choice of future methods}

  \myhypothesis{biological_validation}{biological\_validation}{164,204,46}{Results from real and simulated data are similar, validating the simulated data, especially on more complex tasks for which no decent real datasets are available. The selected datasets cover the range of TI tasks as well as possible.}{Apply guidelines on real data and show that the selected methods work well in comparison to when the guidelines are not followed. A TI method that obtains a good relative performance on synthetic data should also obtain a good performance on the biological data}{}{}

  \myhypothesis{future}{future}{254,179,8}{We also look to the future of single-cell technologies and TI}{We simulate data which will be available soon\\
    We also generate data which can't be modelled yet by current methods}{N/A}{We have only seen the tip of the iceberg of which can be modeled using current TI methods, and we thus stimulate the development of future methods and the generation of data from more complex biological settings}

  \myhypothesis{simulate_inferred_networks}{simulate\_inferred\_networks}{254,179,8}{We can generate expression data on inferred networks of developing cells}{0. Collect expression of cells in a dynamic process\\
    1. Perform NI\\
    2. Generate expression of GRN\\
    3. Apply TI on real expression data\\
    4. Apply TI on synthetic expression data\\
    5. TI results on real expression data should be comparable to synthetic expression data}{Data looks similar, as long as we can trust the NI}{It is a validation of the whole synthetic data generation workflow}

  \myhypothesis{tradeoff_cost_performance}{tradeoff\_cost\_performance}{254,179,8}{We compared current platforms, estimated the cost to improve certain parameters, and predicted which parameters to improve to obtain the best performance increase versus cost increase}{We should attempt to quantify the cost associated with improving each of the sources of technical noise, and make a trade-off between TI performance and total cost}{}{scRNA-seq platforms will improve. We can find out what the most cost-effective way of choosing/improving a platform is in order to study dynamic processes.}

  \myhypothesis{mrna_vs_protein}{mrna\_vs\_protein}{254,179,8}{Observing protein levels might improve TI}{Add flowcyto and proseq platforms. Evaluate}{As cells differentiate, mRNA levels increase first, protein levels increase later.\\
    Observing protein expression levels might thus improve TI.}{Integrative TI with index sorting}

  \myhypothesis{multiple_processes}{multiple\_processes}{254,179,8}{Most existing methods will have problems detecting multiple processes. These might be synchronised or not.}{Generate networks and expression data containing multiple dynamic processes}{Watch TI methods fail horribly}{Promoting the development of better methods}

  \myhypothesis{intercellular_communication}{intercellular\_communication}{254,179,8}{We generated data in which multiple cells interact with eachother, influencing eachothers expression. And develop methods to synchronize multiple cells}{Add networks between cells\\
    Slightly different metric to comparison}{✓}{scRNA-seq will in the future allow the inference of networks between cells}

  \myhypothesis{modular_ti}{modular\_ti}{254,179,8}{Each TI approach can be disassembled into smaller components, new TI approaches can be constructed from a combination from components}{TODO}{}{}

  \myhypothesis{communication}{communication}{155,89,182}{We communicated our results to a broad audience in an interesting manner}{TODO}{}{}

  \myhypothesis{manuscript}{manuscript}{155,89,182}{We communicated to the 100 most interested people}{Write a clear and consise manuscript}{}{}

  \myhypothesis{large_communication}{large\_communication}{155,89,182}{We communicated to the 1000 most interested people}{We make our evaluation code available as a package.\\
    We post the results on the github page. When then evaluation pipeline is modified (i.e. new methods or datasets have been added), the results are also updated}{}{}

  \myhypothesis{mass_communication}{mass\_communication}{155,89,182}{We communicated to 10000+ people}{Go to conferences and/or contact the press?}{Fame and glory}{}

### 2018-06-27 README.Rmd
update readme

@@
 
+## Structure
+
+This repository is structured as follows:
+
+* data: raw data files required by the scripts
+* derived: data files produced by the scripts
+* figures: figures created by the scrips
+* manuscript: source files for producing the manuscript
+* package: an R package with helper functions
+* results: data objects outputted by the scripts
+* scripts: code required to reproduce the analyses of the benchmark
+
 ## Datasets

### 2018-07-05 scripts/evaluate_toy/1-methods-toys.Rmd
dataset -> task

@@
 
-data("toy_tasks")
-tasks <- toy_tasks
+data("toy_datasets")
+datasets <- toy_datasets
 ```
@@
 ```{r}
-tasks$counts <- map(tasks$counts, function(counts) {
+datasets$counts <- map(datasets$counts, function(counts) {
   cbind2(counts, matrix(sample(1:100, nrow(counts)*100, replace=TRUE), nrow=nrow(counts))) %>% {magrittr::set_colnames(., paste0("G", 1:ncol(.)))}
 })
-task <- dynutils::extract_row_to_list(tasks, 3)
-pheatmap::pheatmap(task$counts)
+dataset <- dynutils::extract_row_to_list(datasets, 3)
+pheatmap::pheatmap(dataset$counts)
 
 source("~/thesis/projects/dynverse/dynmodular/dimred_wrappers.R")
-# space <- task$counts %>% dimred_dp(2, neigen=2) %>% as.data.frame
-space <- task$counts %>% dimred_mds(2) %>% as.data.frame
-# space <- task$counts %>% dimred_lle(2) %>% as.data.frame
-space <- space %>% bind_cols(task$prior_information$grouping_assignment %>% slice(match(rownames(task$counts), cell_id)))
+# space <- dataset$counts %>% dimred_dp(2, neigen=2) %>% as.data.frame
+space <- dataset$counts %>% dimred_mds(2) %>% as.data.frame
+# space <- dataset$counts %>% dimred_lle(2) %>% as.data.frame
+space <- space %>% bind_cols(dataset$prior_information$grouping_assignment %>% slice(match(rownames(dataset$counts), cell_id)))
 space %>%  ggplot() + geom_point(aes(Comp1, Comp2, color=group_id))
@@
   # factory(function() {
-    dyneval:::evaluate_ti_method(tasks, method, params, metrics = metric_names, timeout = 100)
+    dyneval:::evaluate_ti_method(datasets, method, params, metrics = metric_names, timeout = 100)
   # }
@@
 })
-# }, qsub_environment = list2env(lst(method_ti_methods, tasks, metric_names, method_params)), qsub_config = qsub::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
+# }, qsub_environment = list2env(lst(method_ti_methods, datasets, metric_names, method_params)), qsub_config = qsub::override_qsub_config(memory = "10G", execute_before=c("module load python/x86_64/3.5.1"), wait=F))
 ```
@@
 ```{r fig.height=10, fig.width=10}
-scores %>% left_join(tasks %>% select(id, trajectory_type), by=c("task_id"="id")) %>% 
+scores %>% left_join(datasets %>% select(id, trajectory_type), by=c("dataset_id"="id")) %>% 
   dplyr::select(-starts_with("time")) %>%
-  gather(score_id, score, -method_name, -method_short_name, -method_id, -task_id, -error, -warning, -model, -trajectory_type) %>%
+  gather(score_id, score, -method_name, -method_short_name, -method_id, -dataset_id, -error, -warning, -model, -trajectory_type) %>%
   filter(score_id %in% c("errored", "correlation", "robbie_network_score", "mean_R_nx")) %>% 

### 2018-07-05 scripts/manual_ti/guide.Rmd
dataset -> task

@@
 ```{r setup, include=FALSE}
-library(dynalysis)
+library(dynbenchmark)
 library(tidyverse)
 
-tasks_control <- read_rds(derived_file("tasks.rds", "1-datasets/control"))
-space <- tasks_control %>% filter(id == "control_bifurcating") %>% pull(space) %>% first()
+datasets_control <- read_rds(derived_file("datasets.rds", "1-datasets/control"))
+space <- datasets_control %>% filter(id == "control_bifurcating") %>% pull(space) %>% first()
 ```

### 2018-07-05 manuscript/setup.R

@@ -105,7 +105,7 @@ add_stable <- function(table, ref_id, caption) {
 methods <- read_rds(result_file("methods.rds", experiment_id = "4-method_characterisation"))
 methods_evaluated <- read_rds(result_file("methods_evaluated.rds", experiment_id = "4-method_characterisation"))
 implementations <- read_rds(result_file("implementations.rds", experiment_id = "4-method_characterisation"))
-tasks_info <- read_rds(result_file("tasks_info.rds", experiment_id = "5-optimise_parameters/3-evaluate_parameters"))
+datasets_info <- read_rds(result_file("datasets_info.rds", experiment_id = "5-optimise_parameters/3-evaluate_parameters"))
 
 # citate
 cite_methods <- function(method_ids) {

### 2018-07-05 scripts/3-metric_characterisation/perturbation.R

@@ -1,37 +1,37 @@
 ## Again gold standard
-perturb_gs <- function(task) {
-  task
+perturb_gs <- function(dataset) {
+  dataset
 }
 
 ## Switch cells
-perturb_switch_n_cells <- function(task, n = length(task$cell_ids)) {
-  the_chosen_ones <- sample(task$cell_ids, n)
+perturb_switch_n_cells <- function(dataset, n = length(dataset$cell_ids)) {
+  the_chosen_ones <- sample(dataset$cell_ids, n)
 
-  mapper <- set_names(task$cell_ids, task$cell_ids)
+  mapper <- set_names(dataset$cell_ids, dataset$cell_ids)
   mapper[match(the_chosen_ones, mapper)] <- rev(the_chosen_ones)
 
-  task$milestone_percentages$cell_id <- mapper[task$milestone_percentages$cell_id]
-  task$progression$cell_id <- mapper[task$progression$cell_id]
+  dataset$milestone_percentages$cell_id <- mapper[dataset$milestone_percentages$cell_id]
+  dataset$progression$cell_id <- mapper[dataset$progression$cell_id]
 
-  recreate_task(task)
+  recreate_dataset(dataset)
 }
 
-perturb_switch_two_cells <- function(task) perturb_switch_n_cells(task, 2)
+perturb_switch_two_cells <- function(dataset) perturb_switch_n_cells(dataset, 2)
 
-perturb_switch_all_cells <- function(task) perturb_switch_n_cells(task, length(unique(task$progressions$cell_id)))
+perturb_switch_all_cells <- function(dataset) perturb_switch_n_cells(dataset, length(unique(dataset$progressions$cell_id)))
 
 percs <- seq(0, 1, 0.1)
-perc_perturbators <- map(percs, function(perc) {function(task) perturb_switch_n_cells(task, length(task$cell_id) * perc)}) %>%
+perc_perturbators <- map(percs, function(perc) {function(dataset) perturb_switch_n_cells(dataset, length(dataset$cell_id) * perc)}) %>%
   set_names(paste0("perturb_switch_", percs * 100))
 perc_perturbators %>% list2env(.GlobalEnv)
 
 
 # Break cycle
 # this should be easier, no?
-perturb_break_cycles <- function(task) {
-  # task <- generate_cycle()
+perturb_break_cycles <- function(dataset) {
+  # dataset <- generate_cycle()
 
[…]

### 2018-07-05 scripts/5-optimise_parameters/1-submit_jobs.R

@@ -16,7 +16,7 @@ execute_before <- "export DYNBENCHMARK_PATH=/group/irc/shared/dynbenchmark/; sin
 verbose <- TRUE
 
 # define important folders
-task_ids <- load_datasets() %>% filter(task_source == "synthetic") %>% pull(id)
+dataset_ids <- load_datasets() %>% filter(dataset_source == "synthetic") %>% pull(id)
 local_output_folder <- derived_file("suite/")
 remote_output_folder <- paste0("/scratch/irc/shared/dynverse_derived/", getOption("dynbenchmark_experiment_id"), "/")
 
@@ -34,11 +34,11 @@ methods$short_name
 # save benchmark configuration and start it
 write_rds(lst(
   methods, metrics, timeout_paramoptim, num_repeats, num_init_params, num_iterations, max_memory_per_core, num_cores,
-  execute_before, verbose, local_output_folder, remote_output_folder, task_ids
+  execute_before, verbose, local_output_folder, remote_output_folder, dataset_ids
 ), derived_file("config.rds"))
 
 paramoptim_submit(
-  task_ids = task_ids,
+  dataset_ids = dataset_ids,
   methods = methods,
   timeout_paramoptim = timeout_paramoptim,
   max_memory_per_core = max_memory_per_core,

### 2018-07-05 scripts/3-metric_characterisation/1old-generate_toys.Rmd
rename all mentions of dynalysis to dynbenchmark
dataset -> task

@@
 
-toys$gs_grouped <- parallel::mclapply(toys$gs, group_task, mc.cores=8)
+toys$gs_grouped <- parallel::mclapply(toys$gs, group_dataset, mc.cores=8)
 

### 2018-07-05 scripts/1-datasets/real/1_download_from_sources.R

@@ -11,25 +11,26 @@ dataset_scripts <- list.files(path = "scripts/1-datasets/real/1_download_from_so
 script_contents <- map_chr(dataset_scripts, read_file)
 
 
+qsub_config <- qsub::override_qsub_config(
+  name = "dynreal",
+  memory = "30G",
+  wait = FALSE,
+  stop_on_error = FALSE
+)
+
 # Make sure all packages are installed on the cluster; i.e. GEOquery, MultiAssayExperiment, tidyverse, and dynbenchmark.
 handle <- qsub::qsub_lapply(
-  X = script_contents,
-  qsub_environment = character(),
-  qsub_config = qsub::override_qsub_config(
-    name = "dynreal",
-    memory = "30G",
-    wait = FALSE,
-    stop_on_error = FALSE
-  ),
-  qsub_packages = "dynbenchmark",
-  FUN = function(script_content) {
+  script_contents,
+  function(script_content) {
     dataset_script <- tempfile()
     readr::write_file(script_content, dataset_script)
 
     source(dataset_script)
 
     TRUE
-  }
+  },
+  qsub_environment = NA,
+  qsub_config = qsub_config
 )
 
 write_rds(handle, "handle.rds")
@@ -40,13 +41,4 @@ names(results) <- dataset_scripts
 
 results <- results %>%
   keep(is.na) %>%
-  enframe("dataset_script", "result") %>%
-  mutate(error = map_chr(result, ~attr(., "qsub_error")))
-
-
-remote_dynbenchmark_folder <- qsub::qsub_lapply(1, function(x) dynbenchmark::get_dynbenchmark_folder(), qsub_environment = character())[[1]]
-local_dynbenchmark_folder <- dynbenchmark::get_dynbenchmark_folder()
-local_folder_sync <- dynbenchmark::derived_file("", "1-datasets/real/")
-remote_folder_sync <- gsub(local_dynbenchmark_folder, remote_dynbenchmark_folder, local_folder_sync)
-
-qsub::rsync_remote("prism", remote_folder_sync, "", local_folder_sync)
+  enframe("dataset_script", "result")

### 2018-07-05 scripts/1-datasets/synthetic/1_settings.R

@@ -5,10 +5,10 @@ restart <- FALSE
 
 if (restart) {
   unlink(dataset_file(), recursive = TRUE);dir.create(dataset_file(), recursive = TRUE, showWarnings = FALSE)
-  unlink(dataset_preproc_file(), recursive = TRUE);dir.create(dataset_preproc_file(), recursive = TRUE, showWarnings = FALSE)
+  unlink(dataset_source_file(), recursive = TRUE);dir.create(dataset_source_file(), recursive = TRUE, showWarnings = FALSE)
   qsub:::run_remote(glue::glue("rm -r {paste0(remote_folder, dataset_file(relative = TRUE))}"), "prism")
-  qsub:::run_remote(glue::glue("rm -r {paste0(remote_folder, dataset_preproc_file(relative = TRUE))}"), "prism")
-  qsub:::run_remote(glue::glue("mkdir {paste0(remote_folder, dataset_preproc_file(relative = TRUE))}"), "prism")
+  qsub:::run_remote(glue::glue("rm -r {paste0(remote_folder, dataset_source_file(relative = TRUE))}"), "prism")
+  qsub:::run_remote(glue::glue("mkdir {paste0(remote_folder, dataset_source_file(relative = TRUE))}"), "prism")
   qsub:::run_remote(glue::glue("mkdir {paste0(remote_folder, dataset_file(relative = TRUE))}"), "prism")
 }
 
@@ -59,5 +59,5 @@ paramsets <- map(seq_len(nrow(settings)), function(row_id) {
 })
 
 # save paramsets ----------------------------------------
-saveRDS(paramsets, dataset_preproc_file("paramsets.rds"))
-qsub:::rsync_remote("", dataset_preproc_file(), "prism", paste0(remote_folder, dataset_preproc_file(relative = TRUE)))
+saveRDS(paramsets, dataset_source_file("paramsets.rds"))
+qsub:::rsync_remote("", dataset_source_file(), "prism", paste0(remote_folder, dataset_source_file(relative = TRUE)))

### 2018-07-05 scripts/5-optimise_parameters/2-retrieve_results.R

@@ -13,18 +13,18 @@ paramoptim_fetch_results(derived_file("suite/"))
 # bind results in one data frame (without models)
 outputs <- paramoptim_bind_results(derived_file("suite/"))
 
-# load tasks info
+# load datasets info
 list2env(read_rds(derived_file("config.rds")), environment())
-tasks <- load_datasets()
+datasets <- load_datasets()
 
 ###################################################
 ############### CREATE AGGREGATIONS ###############
 ###################################################
 
 outputs_ind <- outputs %>%
-  left_join(tasks_info, by = "task_id") %>%
-  filter(task_source != "toy") %>%
-  group_by(task_id) %>%
+  left_join(datasets_info, by = "dataset_id") %>%
+  filter(dataset_source != "toy") %>%
+  group_by(dataset_id) %>%
   mutate(
     rank_correlation = percent_rank(correlation),
     rank_rf_mse = percent_rank(-rf_mse),
@@ -36,14 +36,14 @@ outputs_ind <- outputs %>%
 
 # process trajtype grouped evaluation
 outputs_summtrajtype <- outputs_ind %>%
-  group_by(method_name, method_short_name, task_source, param_i, repeat_i, trajectory_type) %>%
+  group_by(method_name, method_short_name, dataset_source, param_i, repeat_i, trajectory_type) %>%
   mutate(n = n()) %>%
   summarise_if(is.numeric, mean) %>%
   ungroup()
 
 # process overall evaluation
 outputs_summmethod <- outputs_summtrajtype %>%
-  group_by(method_name, method_short_name, task_source, param_i, repeat_i) %>%
+  group_by(method_name, method_short_name, dataset_source, param_i, repeat_i) %>%
   mutate(n = n()) %>%
   summarise_if(is.numeric, mean) %>%
   ungroup()
@@ -55,7 +55,7 @@ outputs_summmethod_totals <-
       group_by(method_name, method_short_name, param_i, repeat_i) %>%
       summarise_if(is.numeric, mean) %>%
       ungroup() %>%
-      mutate(task_source = "mean")
+      mutate(dataset_source = "mean")
 
 ggplot(outputs_summmethod_totals) + geom_point(aes(param_i, harm_mean)) +
   facet_wrap(~method_short_name, scales = "free") +

### 2018-07-05 scripts/7-aggregate/0-aggregate_data.R

@@ -10,9 +10,9 @@ eval_param_config <- read_rds(derived_file("config.rds", "5-optimise_parameters/
 # read param eval results
 eval_param_outputs <- read_rds(derived_file("outputs_postprocessed.rds", "5-optimise_parameters/3-evaluate_parameters"))
 
-# read tasks
-tasks <- map_df(
-  paste0(eval_param_config$local_tasks_folder, "/", eval_param_config$task_ids, ".rds"),
+# read datasets
+datasets <- map_df(
+  paste0(eval_param_config$local_datasets_folder, "/", eval_param_config$dataset_ids, ".rds"),
   function(file_path) {
     read_rds(file_path) %>% mutate(ncell = nrow(expression[[1]]), nfeat = ncol(expression[[1]]))
   }
@@ -32,7 +32,7 @@ qc_application_scores <- read_rds(derived_file("implementation_qc_application_sc
 ## START GATHERING COLUMNS
 part_overall_mean <-
   eval_param_outputs$outputs_summtrajtype_totalsx2 %>%
-  filter(trajectory_type == "overall", task_source == "mean") %>%
+  filter(trajectory_type == "overall", dataset_source == "mean") %>%
   select(
     method_short_name, harm_mean, norm_correlation, norm_edge_flip, norm_rf_mse, time_method, rank_time_method,
     num_files_created, num_setseed_calls, pct_errored, pct_time_exceeded, pct_memory_exceeded, pct_allerrored, pct_stochastic
@@ -41,15 +41,15 @@ part_overall_mean <-
 
 part_sources <-
   eval_param_outputs$outputs_summtrajtype_totalsx2 %>%
-  filter(trajectory_type == "overall", task_source != "mean") %>%
-  select(method_short_name, task_source, harm_mean) %>%
-  mutate(task_source = paste0("source_", task_source)) %>%
-  spread(task_source, harm_mean) %>%
+  filter(trajectory_type == "overall", dataset_source != "mean") %>%
+  select(method_short_name, dataset_source, harm_mean) %>%
+  mutate(dataset_source = paste0("source_", dataset_source)) %>%
+  spread(dataset_source, harm_mean) %>%
   mutate(overall_source = apply(.[,colnames(.)[grepl("^source_", colnames(.))]], 1, mean))
 
 part_trajtypes <-
   eval_param_outputs$outputs_summtrajtype_totalsx2 %>%
-  filter(trajectory_type != "overall", task_source == "mean") %>%
[…]

### 2018-07-05 scripts/gource/1_gource_google_drive.R

@@ -33,7 +33,7 @@ links <- tribble(
   "dynverse_ideas", "https://docs.google.com/document/d/18PuTgnpUtbGwZvbTyiTsJQCaGRXratjbmsjv9JsnYUc/edit",
   "manuscript_v0", "https://docs.google.com/document/d/1DNuhEHJR03yT8yaAXHc8LYpV7EJDOMOTMdGl76J3zLg/edit",
   "real_datasets", "https://docs.google.com/spreadsheets/d/1SALZ2jt7TZJQJMEvvOwSR2r5yIl50qcGAZ-K5AC4DJo/edit#gid=0",
-  "hypotheses_and_tasks", "https://docs.google.com/spreadsheets/d/13jHFuMwie7oIxGcyLO95-onFri4M_GJjqgMEUnZliQI/edit#gid=0",
+  "hypotheses_and_datasets", "https://docs.google.com/spreadsheets/d/13jHFuMwie7oIxGcyLO95-onFri4M_GJjqgMEUnZliQI/edit#gid=0",
   "trajectory_inference_methods_public", "https://docs.google.com/spreadsheets/d/1xeRtfa4NZR-FBZOosPv6nY-alfoGE1ZCEP2ceQg6GH0/edit?usp=drive_web&ouid=105833423392613772145",
   "trajectory_inference_methods_private", "https://docs.google.com/spreadsheets/d/1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE/edit?usp=drive_web&ouid=105833423392613772145",
   "manuscript_v1_overview", "https://docs.google.com/document/d/12xrM1GDq4Lk3mHHQzazgL5eYq7yttrKFtUakgU2AvGs/edit",

### 2018-07-05 scripts/old11-evaluation_robustness/1-real_synthetic.R

@@ -11,15 +11,15 @@ read_rds(derived_file("evaluation_algorithm.rds", "5-optimise_parameters/10-aggr
 
 # create trajectory type
 trajtype_source_scores <- trajtype_scores %>%
-  filter(task_source %in% c("real", "synthetic")) %>%
-  select(method_short_name, trajectory_type, task_source, harm_mean) %>%
-  spread("task_source", "harm_mean") %>%
+  filter(dataset_source %in% c("real", "synthetic")) %>%
+  select(method_short_name, trajectory_type, dataset_source, harm_mean) %>%
+  spread("dataset_source", "harm_mean") %>%
   drop_na()
 
 # create overall
 source_scores <- source_scores %>%
-  select(method_short_name, task_source, harm_mean) %>%
-  spread("task_source", "harm_mean") %>%
+  select(method_short_name, dataset_source, harm_mean) %>%
+  spread("dataset_source", "harm_mean") %>%
   drop_na() %>%
   mutate(trajectory_type = "overall")
 

### 2018-07-09 scripts/03-metric_characterisation/1old-generate_toys.Rmd
aaaaa! renaming scripts folders!

--- 2018-07-05 scripts/3-metric_characterisation/1old-generate_toys.Rmd
@@
 ```{r}
-source("analysis/analyses/3-metric_characterisation/generation.R")
-source("analysis/analyses/3-metric_characterisation/perturbation.R")
+source("analysis/analyses/03-metric_characterisation/generation.R")
+source("analysis/analyses/03-metric_characterisation/perturbation.R")
 
-experiment("3-metric_characterisation/1-generate_toys")
+experiment("03-metric_characterisation/1-generate_toys")
 ```

### 2018-07-09 scripts/02-dataset_characterisation/1-synthetic.R

@@ -1,0 +1,113 @@
+library(dynbenchmark)
+library(tidyverse)
+
+library(tidygraph)
+library(ggraph)
+
+experiment("02-dataset_characterisation/1-synthetic")
+datasets <- load_datasets()
+
+
+#   ____________________________________________________________________________
+#   Plotting the different models                                           ####
+
+modulenet_names <- datasets$settings %>% map("modulenet_name") %>% unlist() %>% unique()
+
+params <- dyngen::simple_params
+
+scale_edge_burn <- scale_edge_linetype_manual(values = c("FALSE" = "solid", "TRUE" = "dashed"))
+scale_edge_transition <- scale_edge_colour_brewer(palette = "Set2")
+theme_without_legend <- theme(legend.position = "none")
+
+plots <- map(modulenet_names, function(modulenet_name) {
+  model <- dyngen:::load_modulenet(modulenet_name)
+
+  # Module net graph
+  modulenet_graph <- tbl_graph(model$modulenodes, model$modulenet)%>%
+    activate(edges) %>%
+    mutate(effect = as.character(effect)) %>%
+    activate(nodes) %>%
+    mutate(a0 = as.character(a0))
+
+  modulenet_plot <- modulenet_graph %>%
+    activate(edges) %>%
+    arrange(from == to) %>%
+    ggraph(layout = 'linear', circular = TRUE) +
+    geom_edge_link(aes(color = effect), arrow = arrow(type = "closed"), end_cap = circle(3, 'mm')) +
+    geom_edge_loop(aes(color = effect), arrow = arrow(type = "closed"), end_cap = circle(3, 'mm')) +
+    geom_node_label(aes(fill = a0, label = module_id)) +
+    scale_edge_color_manual(values = c("-1" = "#0074D9", "1" = "#FF4136")) +
+    scale_fill_manual(values = c("1" = "#FF4136", "0" = "white")) +
+    theme_graph() +
+    theme_without_legend +
+    ggtitle(label_long(modulenet_name), "Module network")
+
+  # Milestone graph
+  edge_operations <- model$edge_operations %>%
+    mutate(edge_id = factor(seq_len(n())))
+
+  milestone_graph <- edge_operations %>%
+    as_tbl_graph()
+
+  milestone_plot <- ggraph(milestone_graph, layout = "fr") +
+    geom_node_point(aes()) +
[…]

### 2018-07-09 scripts/04-method_characterisation/3_process_qc.R

@@ -1,0 +1,84 @@
+library(tidyverse)
+library(googlesheets)
+library(dynbenchmark)
+
+experiment("04-method_characterisation")
+
+##  ............................................................................
+##  QC sheet processing                                                     ####
+# Download qc & initial processing
+implementation_qc_sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE") %>%
+  gs_read(ws = "QualityControl")
+
+categories <- unique(implementation_qc_sheet$category) %>% discard(is.na)
+applications <- c("developer_friendly", "user_friendly", "good_science")
+
+implementation_qc_converted <- implementation_qc_sheet %>%
+  filter(active) %>%
+  tidyr::fill(aspect_id) %>%
+  mutate(aspect_id = factor(aspect_id, levels = implementation_qc_sheet$aspect_id %>% unique() %>%
+                              keep(~!is.na(.)))) %>%
+  group_by(aspect_id) %>%
+  tidyr::fill(name, category, weight, aspect, !!applications, references) %>%
+  mutate_at(vars(!!applications), ~ifelse(is.na(.), FALSE, TRUE)) %>%
+  ungroup()
+
+implementation_cols <- colnames(implementation_qc_converted)[seq(which(colnames(implementation_qc_converted) == "item")+1, ncol(implementation_qc_converted))]
+
+implementation_qc_converted <- implementation_qc_converted %>% mutate(check_id = row_number())
+
+# create checks
+checks <- implementation_qc_converted[, colnames(implementation_qc_converted) %>% keep(~!(. %in% implementation_cols))]
+
+implementation_qc_molten <- implementation_qc_converted %>%
+  gather("implementation_id", "answer", !!implementation_cols)
+
+# process answer, NA coercion warnings are normal here due to check of
+implementation_qc_processed <- implementation_qc_molten %>%
+  mutate(
+    answer = ifelse(is.na(answer), "0", answer),
+    answer_first_char = str_replace(answer, "([\\d\\.]*).*", "\\1"),
+    answer_description = str_replace(answer, "[\\d\\.]*(.*)", "\\1") %>% trimws()
+  ) %>%
+  mutate(answer = as.numeric(ifelse(answer_first_char == "", "0", answer_first_char)))
+
+implementation_qc_processed <- implementation_qc_processed %>%
+  group_by(implementation_id) %>%
+  filter(!any(is.na(answer))) %>%
+  ungroup()
+
[…]

### 2018-07-09 scripts/7-aggregate/0-aggregate_data.R

@@ -1,161 +1,0 @@
-library(dynbenchmark)
-library(tidyverse)
-library(dynplot)
-
-experiment("9-main_figure")
-
-# read param eval config
-eval_param_config <- read_rds(derived_file("config.rds", "5-optimise_parameters/3-evaluate_parameters"))
-
-# read param eval results
-eval_param_outputs <- read_rds(derived_file("outputs_postprocessed.rds", "5-optimise_parameters/3-evaluate_parameters"))
-
-# read datasets
-datasets <- map_df(
-  paste0(eval_param_config$local_datasets_folder, "/", eval_param_config$dataset_ids, ".rds"),
-  function(file_path) {
-    read_rds(file_path) %>% mutate(ncell = nrow(expression[[1]]), nfeat = ncol(expression[[1]]))
-  }
-)
-
-# read method meta info
-meta <- read_rds(derived_file("methods_evaluated.rds", "4-method_characterisation")) %>%
-  mutate(
-    prior_mini_id = paste0("prior_mini_", group_indices(group_by_at(., vars(!!setdiff(priors$prior_id, "time")))))
-  )
-
-# read qc scores
-qc_category_scores <- read_rds(derived_file("implementation_qc_category_scores.rds", "4-method_characterisation"))
-qc_application_scores <- read_rds(derived_file("implementation_qc_application_scores.rds", "4-method_characterisation"))
-
-
-## START GATHERING COLUMNS
-part_overall_mean <-
-  eval_param_outputs$outputs_summtrajtype_totalsx2 %>%
-  filter(trajectory_type == "overall", dataset_source == "mean") %>%
-  select(
-    method_short_name, harm_mean, norm_correlation, norm_edge_flip, norm_rf_mse, time_method, rank_time_method,
-    num_files_created, num_setseed_calls, pct_errored, pct_time_exceeded, pct_memory_exceeded, pct_allerrored, pct_stochastic
-  ) %>%
-  mutate(overall_metric = apply(.[,colnames(.)[grepl("^norm_", colnames(.))]], 1, mean))
-
-part_sources <-
-  eval_param_outputs$outputs_summtrajtype_totalsx2 %>%
-  filter(trajectory_type == "overall", dataset_source != "mean") %>%
-  select(method_short_name, dataset_source, harm_mean) %>%
-  mutate(dataset_source = paste0("source_", dataset_source)) %>%
-  spread(dataset_source, harm_mean) %>%
[…]

### 2018-07-09 scripts/7-aggregate/old10-aggregations.R

@@ -1,31 +1,0 @@
-library(cowplot)
-library(tidyverse)
-library(dynbenchmark)
-
-library(tidygraph)
-library(ggraph)
-
-experiment("5-optimise_parameters/10-aggregations")
-
-outputs_list <- read_rds(derived_file("outputs_postprocessed.rds", "5-optimise_parameters/3-evaluate_parameters"))
-aggregated_data <- read_rds(result_file("aggregated_data.rds", "9-main_figure"))
-
-overall_scores <- aggregated_data$method_tib %>%
-  arrange(-overall_benchmark)
-method_order <- overall_scores$method_short_name
-source_scores <- outputs_list$outputs_summmethod %>%
-  inner_join(methods, "method_short_name")
-trajtype_scores <- outputs_list$outputs_summtrajtype_totals %>%
-  inner_join(methods, "method_short_name")
-indrep_scores <- outputs_list$outputs_summrepl %>%
-  inner_join(methods, "method_short_name")
-ind_scores <- outputs_list$outputs_ind %>%
-  inner_join(methods, "method_short_name")
-
-methods <- read_rds(derived_file("methods.rds", experiment_id = "4-method_characterisation")) %>%
-  filter(type %in% c("algorithm", "control")) %>%
-  filter(method_id %in% overall_scores$method_short_name)
-
-evaluation_algorithm <- lst(overall_scores, method_order, trajtype_scores, indrep_scores, ind_scores, methods)
-
-evaluation_algorithm %>% write_rds(derived_file("evaluation_algorithm.rds"))

### 2018-07-09 scripts/7-user_guidelines/create_data.R
aaaaa! renaming scripts folders!

@@ -1,39 +1,0 @@
-library(cowplot)
-library(tidyverse)
-library(dynbenchmark)
-
-experiment("7-user_guidelines")
-
-read_rds(derived_file("evaluation_algorithm.rds", "5-optimise_parameters/10-aggregations")) %>% list2env(.GlobalEnv)
-
-methods <-
-  left_join(
-    methods,
-    read_rds(derived_file("implementation_qc_application_scores.rds", "4-method_characterisation")) %>%
-      spread("application", "score"),
-    "implementation_id"
-  ) %>%
-  filter(type %in% c("algorithm", "control"))
-
-
-
-scoring <- list(
-  methods = methods,
-  trajtype_scores = trajtype_scores,
-  metrics = metrics %>% filter(metric_id %in% c("correlation", "rf_mse", "edge_flip"))
-)
-
-process_prior_information <- function(scoring, answer = NULL) {
-  scoring
-}
-
-add_surveyjs_data <- function(l) {
-  l$name = l$question_id
-  if(!"title" %in% names(l)) {
-    l$title = label_long(l$question_id)
-  }
-  l
-}
-
-
-saveRDS("")

### 2018-07-09 manuscript/paper_latex.Rmd

@@
 add_stable(
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
+  read_rds(figure_file("implementations_table.rds", "04-method_characterisation"))[[params$table_format]],
 "implementations",
@@
 add_fig(
-  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "4-method_characterisation"),
+  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "04-method_characterisation"),
   ref_id = "implementation_qc_overview", 
@@
 add_fig(
-  fig_path = figure_file("tree.tmp.pdf", experiment_id = "7-user_guidelines"),
+  fig_path = figure_file("tree.tmp.pdf", experiment_id = "09-user_guidelines"),
   ref_id = "user_guidelines", 
@@
 add_stable(
-  read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
+  read_rds(figure_file("datasets.rds","02-dataset_characterisation"))[[params$table_format]], 
   "datasets", 
@@
     "samplers.rds",
-    "2-dataset_characterisation/3-synthetic")
+    "02-dataset_characterisation/3-synthetic")
   )[[params$table_format]], 
@@
 add_sfig(
-  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-metric_characterisation/4-edge_flip"), 
+  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="03-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 
@@
     "qc_checks.rds",
-    "4-method_characterisation"
+    "04-method_characterisation"
   ))[[params$table_format]], 

### 2018-07-09 scripts/evaluate_toy/1-methods-toys.Rmd

@@
     tibble(
-      model = attr(result, "extras")$.models
+      model = result$models
     ),
-    attr(result, "extras")$.summary
+    result$summary
   )

### 2018-07-09 manuscript/paper.Rmd

@@
 add_fig(
-  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "4-method_characterisation"),
+  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "04-method_characterisation"),
   ref_id = "implementation_qc_overview", 
@@
 add_fig(
-  fig_path = figure_file("tree.svg", experiment_id = "7-user_guidelines"),
+  fig_path = figure_file("tree.svg", experiment_id = "09-user_guidelines"),
   ref_id = "user_guidelines", 
@@
 add_sfig(
-  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
+  figure_file("edge_flip_comparison.svg", experiment_id="03-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 

### 2018-07-09 scripts/manual_ti/guide.Rmd

@@
 
-datasets_control <- read_rds(derived_file("datasets.rds", "1-datasets/control"))
+datasets_control <- read_rds(derived_file("datasets.rds", "01-datasets/control"))
 space <- datasets_control %>% filter(id == "control_bifurcating") %>% pull(space) %>% first()

### 2018-07-13 README.Rmd
update readme

@@
 
-# dynbenchmark
+[![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <img src="package/man/figures/logo.png" align="right" />
 
-[![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
-![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)
+# Benchmarking single-cell trajectory inference methods
 

### 2018-07-31 README.Rmd
update readme

@@
 <!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r setup, echo = FALSE}
+knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE, echo = FALSE)
+```
+
+```{r}
+library(tidyverse)
+library(dynbenchmark)
+```
 
@@
 
-We include the following datasets. When using any of these datasets, please also cite our preprint: [![DOI](https://zenodo.org/badge/DOI/10.1101/276907.svg)](https://doi.org/10.1101/276907)
+The benchmarking pipeline generates (and uses) the following datasets:
 
-* [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1211533.svg)](https://doi.org/10.5281/zenodo.1211533) Single-cell -omics datasets, both real and synthetic, used to evaluated the trajectory inference methods
-* **Coming soon** Main results of the evaluation, used to rank the methods and construct practical guidelines
+* **Gold standard single-cell datasets**, both real and synthetic, used to evaluated the trajectory inference methods [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1211533.svg)](https://doi.org/10.5281/zenodo.1211533)
+
+```{r, echo = FALSE, message=FALSE}
+datasets <- list_datasets() %>% group_by(dataset_source) %>% sample_n(1) %>% pull(dataset_id) %>% load_datasets()
+
+datasets$plot_dimred <- mapdf(datasets, ~plot_dimred(., dimred = dyndimred::dimred_landmark_mds) + ggtitle(.$dataset_source) + theme(plot.title = element_text(size = 10)))
+p <- patchwork::wrap_plots(datasets$plot_dimred, ncol = nrow(datasets))
+ggplot2::ggsave("package/man/figures/datasets.png", width = nrow(datasets)*3, height = 1*3)
+```
+
+![datasets](package/man/figures/datasets.png)
+
+* **The performance of methods** used in [dynguidelines](https://www.github.com/dynverse/dynguidelines). _Not yet available_
+
+* **General information about trajectory inference methods**, available as a data frame in `dynmethods::methods`
+
+## Scripts
+
+The scripts folder contains all the scripts necessary to fully reproduce the benchmarking manuscript in chronologically ordered subfolders.
+Each subfolder contains a readme file with further explanations of the different scripts and what they do:
+
+```{r}
+scripts <- read_rds(derived_file("scripts.rds", experiment_id = "12-varia/01-scripts"))
+
+scripts %>% filter(
+  level == 1
+) %>%
+  arrange(ix) %>%
+  mutate(
+    name = str_glue("[{id}](https://github.com/dynverse/dynbenchmark/tree/master/{location})")
+  ) %>% 
+  select(name, title) %>%
+  knitr::kable()
+
+```
+
+## Benchmarking your own method
+
+Explanation coming soon. Feel free to make an issue if you want your method to be included.

### 2018-07-31 README.Rmd
even more reorganisation

@@
   mutate(
-    name = str_glue("[{id}](https://github.com/dynverse/dynbenchmark/tree/master/{location})")
+    name = str_glue("[{id}](https://github.com/dynverse/dynbenchmark/tree/master/{location})"),
+    order = ifelse(is.na(ix), "", ix)
   ) %>% 
-  select(name, title) %>%
+  select(order, subfolder = name, description = title) %>%
   knitr::kable()
@@
 
-Explanation coming soon. Feel free to make an issue if you want your method to be included.
+Explanation coming soon. Feel free to make an issue or send us an e-mail if you want your method to be included.

### 2018-07-31 README.Rmd

@@
 
-# Benchmarking single-cell trajectory inference methods
+# Benchmarking trajectory inference methods
 
-Scripts to reproduce the manuscript
+This repo contains the scripts to reproduce the manuscript
 
@@
 ```{r, echo = FALSE, message=FALSE}
+set.seed(1)
+
 datasets <- list_datasets() %>% group_by(dataset_source) %>% sample_n(1) %>% pull(dataset_id) %>% load_datasets()

### 2018-08-02 README.Rmd
rename dataset_id to id and dataset_source to source

@@
 
-datasets <- list_datasets() %>% group_by(dataset_source) %>% sample_n(1) %>% pull(dataset_id) %>% load_datasets()
+datasets <- list_datasets() %>% group_by(source) %>% sample_n(1) %>% pull(id) %>% load_datasets()
 
-datasets$plot_dimred <- mapdf(datasets, ~plot_dimred(., dimred = dyndimred::dimred_landmark_mds) + ggtitle(.$dataset_source) + theme(plot.title = element_text(size = 10)))
+datasets$plot_dimred <- mapdf(datasets, ~plot_dimred(., dimred = dyndimred::dimred_landmark_mds) + ggtitle(.$source) + theme(plot.title = element_text(size = 10)))
 p <- patchwork::wrap_plots(datasets$plot_dimred, ncol = nrow(datasets))

### 2018-08-02 scripts/01-datasets/01-real/02_filter_and_normalise.R
rename dataset_id to id and dataset_source to source

@@ -19,9 +19,9 @@ list_datasets_remote <- function() {
 
 dataset_ids_to_process <- setdiff(list_raw_datasets_remote(), list_datasets_remote())
 
-process_raw_dataset_wrapper <- function(dataset_id) {
-  raw_dataset <- readr::read_rds(dynbenchmark::dataset_raw_file(dataset_id))
-  raw_dataset$id <- dataset_id
+process_raw_dataset_wrapper <- function(id) {
+  raw_dataset <- readr::read_rds(dynbenchmark::dataset_raw_file(id))
+  raw_dataset$id <- id
 
   do.call(dynbenchmark::process_raw_dataset, raw_dataset)
   TRUE

### 2018-08-02 scripts/01-datasets/02-synthetic/01-estimate_platform.R

@@ -10,7 +10,7 @@ experiment("01-platforms")
 rm_remote(derived_file("", remote = TRUE), remote = TRUE, recursive = TRUE)
 
 # use all real dataset for platform estimation
-dataset_ids <- list_datasets() %>% filter(dataset_source == "real") %>% pull(dataset_id)
+dataset_ids <- list_datasets() %>% filter(source == "real") %>% pull(id)
 
 # run on cluster
 qsub_config <- override_qsub_config(

### 2018-08-02 README.Rmd
add scripts README

@@
 output: github_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-## Scripts
-
-The scripts folder contains all the scripts necessary to fully reproduce the benchmarking manuscript in chronologically ordered subfolders.
-Each subfolder contains a readme file with further explanations of the different scripts and what they do:
-
-```{r}
-scripts <- read_rds(derived_file("scripts.rds", experiment_id = "12-varia/01-scripts"))
-
-scripts %>% filter(
-  level == 1
-) %>%
-  arrange(ix) %>%
-  mutate(
-    name = str_glue("[{id}](https://github.com/dynverse/dynbenchmark/tree/master/{location})"),
-    order = ifelse(is.na(ix), "", ix)
-  ) %>% 
-  select(order, subfolder = name, description = title) %>%
-  knitr::kable()
-
+```{r child='scripts/README.Rmd'}
 ```

### 2018-08-02 scripts/README.Rmd
add scripts README

---
output: github_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE, echo = FALSE)
```

```{r}
library(tidyverse)
library(dynbenchmark)
```

## Scripts

The scripts folder contains all the scripts necessary to fully reproduce the benchmarking manuscript in chronologically ordered subfolders. Each subfolder contains a readme file with further explanations of the different scripts and what they do:

```{r}
scripts <- read_rds(derived_file("scripts.rds", experiment_id = "12-varia/01-scripts"))

scripts %>% filter(
  level == 1
) %>%
  arrange(ix) %>%
  mutate(
    name = str_glue("[{id}](https://github.com/dynverse/dynbenchmark/tree/master/{location})"),
    order = ifelse(is.na(ix), "", ix)
  ) %>%
  select(order, subfolder = name, description = title) %>%
  knitr::kable()

```

### 2018-08-03 scripts/03-metric_characterisation/01-metric_conformity/03-assess_rules.R
update metric characterisation

@@ -1,47 +1,0 @@
-## Assess whether the metrics follow certain rules
-
-library(tidyverse)
-library(dynbenchmark)
-
-experiment("03-metric_characterisation/01-reference_vs_perturbation")
-
-dataset_design <- read_rds(derived_file("dataset_design.rds"))
-
-# load rules
-source(scripts_file("helper-02-rules.R"))
-
-scores <- read_rds(derived_file("scores.rds"))
-
-filter_scores <- function(crossing, scores) {
-  scores %>% inner_join(crossing, colnames(crossing))
-}
-
-assess_rule <- function(rule, scores) {
-  scores <- filter_scores(rule$crossing, scores)
-
-  # extract only the relevant parts
-  if (nrow(scores) < nrow(rule$crossing)) {
-    warning("Missing some results for ", rule$id, call. = FALSE)
-  }
-
-  rule$assessment(scores, rule)
-}
-
-
-assessments <- mapdf(rules, assess_rule, scores=scores) %>% list_as_tibble() %>% mutate(rule_id = rules$id)
-assessments$plot_datasets
-
-assessments %>%
-  unnest(assessment) %>%
-  ggplot(aes(rule_id, metric_id)) +
-  geom_tile(aes(fill = check)) +
-  scale_fill_manual(values = c(`TRUE` = "#2ECC40", `FALSE` = "#FF4136"))
-
-
-
-assessments$plot_scores
-
-
-## test/create one new rule
-rule <- rules %>% extract_row_to_list(which(id == "merge_bifurcation"))
-scores <- filter_scores(rule$crossing, scores)

### 2018-08-03 scripts/03-metric_characterisation/01-metric_conformity/helper-02-perturbations.R

@@ -34,18 +34,20 @@ perturb_switch_n_edges <- function(dataset, switch_n = Inf) {
 
   switch_n = min(switch_n, nrow(dataset$milestone_network))
 
+  browser()
+
   # create an edge id and add it to the progressions
   milestone_network <- dataset$milestone_network %>%
     mutate(edge_id = row_number())
 
   progressions <- dataset$progressions %>%
-    left_join(milestone_network, c("from", "to")) %>%
+    left_join(milestone_network %>% select(from, to, edge_id), c("from", "to")) %>%
     select(-from, -to)
 
   # shuffle edges
   the_chosen_ones <- sample(milestone_network$edge_id, switch_n)
   mapper <- set_names(milestone_network$edge_id, milestone_network$edge_id)
-  mapper[match(the_chosen_ones, mapper)] <- sample(the_chosen_ones)
+  mapper[match(the_chosen_ones, mapper)] <- sample(the_chosen_ones, length(the_chosen_ones))
 
   # change in milestone_network and join with progressions
   milestone_network$edge_id <- mapper[as.character(milestone_network$edge_id)]
@@ -64,11 +66,30 @@ perturb_switch_n_edges <- function(dataset, switch_n = Inf) {
 }
 
 perturb_switch_edges <- function(dataset, switch_perc = 1) {
-  perturb_switch_n_edges(dataset, switch_n = nrow(dataset$milestone_network) * switch_perc)
+  perturb_switch_n_edges(dataset, switch_n = round(nrow(dataset$milestone_network) * switch_perc))
 }
 
+
+## Remove cells
+perturb_remove_cells <- function(dataset, remove_perc = 0.6) {
+  remove_cell_ids <- sample(dataset$cell_ids, length(dataset$cell_ids) * remove_perc)
+  progressions <- dataset$progressions %>% filter(!(cell_id %in% remove_cell_ids))
+
+  dataset %>%
+    add_trajectory(
+      milestone_network = dataset$milestone_network,
+      progressions = progressions,
+      divergence_regions = dataset$divergence_regions
+    )
+}
+
+
+
+
+#   ____________________________________________________________________________
+#   Changing the topology                                                   ####
 ##  ............................................................................
-##  Chaning bifurcating trajectories                                        ####
+##  Changing bifurcating trajectories                                        ####
 # Merge the branch after bifurcations
 perturb_merge_bifurcation <- function(dataset) {
[…]

### 2018-08-08 manuscript/supplementary_notes/metric_conformity/assessment.Rmd
add score conformity note

---
title: "assessment"
output: html_document
params:
  assessment: NULL
---

```{r}
rule <- dynutils::extract_row_to_list(rules, which(rules$id == assessment$rule_id))
```

## `r knitr::include_graphics(rule$image_location)` `r label_long(assessment$rule_id)`

`r rule$description`

```{r}
width <- ifelse(is.null(assessment$plot_datasets$width), 1, assessment$plot_datasets$width)
height <- ifelse(is.null(assessment$plot_datasets$height), 1, assessment$plot_datasets$height)
```
```{r, fig.width=width, fig.height=height}
assessment$plot_datasets
```

```{r}
width <- ifelse(is.null(assessment$plot_scores$width), 1, assessment$plot_scores$width)
height <- ifelse(is.null(assessment$plot_scores$height), 1, assessment$plot_scores$height)
```
```{r, fig.width=width, fig.height=height}
assessment$plot_scores
```

### 2018-08-08 manuscript/supplementary_notes/metric_conformity/metric_conformity.Rmd
add score conformity note

---
title: "Method conformity"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(dynbenchmark)

experiment("03-metric_characterisation/01-metric_conformity")
```

```{r}
assessments <- read_rds(derived_file("assessments.rds"))
rules <- read_rds(derived_file("rules.rds"))

rules <- rules %>% 
  mutate(
    image_location = map_chr(id, function(id) figure_file(paste0("images/", id, ".png"))),
    image_found = file.exists(image_location)
  )

if (any(!rules$image_found)) {stop("Files not found: ", basename(rules$image_location[!rules$image_found]))}
```

```{r results='asis'}
walkdf(assessments, function(assessment) {
  cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))
  cat("\n------------------------------\n")
})
```

### 2018-08-09 manuscript/supplementary_notes/metric_conformity/assessment.Rmd
update metric characterisation

@@
 title: "assessment"
-output: html_document
 params:
@@
 
-## `r knitr::include_graphics(rule$image_location)` `r label_long(assessment$rule_id)`
+\begin{center}
+```{r, out.width="2in", out.height="10in", dpi = 300}
+knitr::include_graphics(rule$image_location)
+```
+\end{center}
 
 `r rule$description`
+
+A metric conforms to this rule if: $`r rule$conforms_if`$
+
+```{r}
+assessment$conformity %>% 
+  select(metric_id, conforms) %>% 
+  arrange(metric_id) %>% 
+  mutate(
+    metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder(),
+    conforms = kableExtra::cell_spec(
+      ifelse(conforms, "\U2714", "\U2716"), 
+      background = ifelse(conforms, "#2ECC40", "#FF4136"),
+      format = "latex"
+    )
+  ) %>% 
+  spread(metric_id, conforms) %>% 
+  knitr::kable(format = "latex", escape = FALSE) %>% 
+  kableExtra::kable_styling() %>% 
+  kableExtra::row_spec(0, hline_after = FALSE)
+```
 
@@
 ```
-```{r, fig.width=width, fig.height=height}
+```{r A, fig.width=width, fig.height=height, out.width=paste0(width/2, "in")}
 assessment$plot_datasets
@@
 ```
-```{r, fig.width=width, fig.height=height}
+```{r B, fig.width=width, fig.height=height, out.width=paste0(width/2, "in")}
 assessment$plot_scores
 ```
+
+----------------------------------------------------------------------------
+
+\pagebreak

### 2018-08-09 manuscript/supplementary_notes/metric_conformity/metric_conformity.Rmd
update metric characterisation

@@
 ---
-title: "Method conformity"
-output: html_document
+title: "Metric conformity"
+output: 
+  pdf_document:
+    latex_engine: xelatex
+    keep_tex: true
+mainfont: DejaVu Sans
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(
+  echo = FALSE
+)
+options(knitr.duplicate.label = 'allow')
 
@@
 rules <- read_rds(derived_file("rules.rds"))
-
-rules <- rules %>% 
-  mutate(
-    image_location = map_chr(id, function(id) figure_file(paste0("images/", id, ".png"))),
-    image_found = file.exists(image_location)
-  )
-
-if (any(!rules$image_found)) {stop("Files not found: ", basename(rules$image_location[!rules$image_found]))}
 ```
 
+## Summary
+
+```{r}
+assessments %>%
+  left_join(rules %>% select(id, name), by = c("rule_id" = "id")) %>%
+  unnest(conformity) %>%
+  mutate(
+    metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder(),
+    conforms = kableExtra::cell_spec(
+      ifelse(conforms, "\U2714", "\U2716"),
+      background = ifelse(conforms, "#2ECC40", "#FF4136"),
+      format = "latex"
+    )
+  ) %>%
+  spread(metric_id, conforms) %>%
+  select(-rule_id) %>% 
+  knitr::kable(format = "latex", escape = FALSE) %>%
+  kableExtra::kable_styling(row_label_position = "r")
+```
+
+\pagebreak
+
 ```{r results='asis'}
-walkdf(assessments, function(assessment) {
+# assessment <- dynutils::extract_row_to_list(assessments, 1)
+
+walkdf(assessments[1, ], function(assessment) {
   cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))
-  cat("\n------------------------------\n")
 })

### 2018-08-09 scripts/03-metric_characterisation/01-metric_conformity/03-assess_conformity.R

@@ -70,7 +70,7 @@ assessments %>%
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
 # test one rule
-rule <- rules %>% extract_row_to_list(which(rules$id == "combined_local_global_position_change"))
+rule <- rules %>% extract_row_to_list(which(rules$id == "cell_gathering"))
 assessment <- assess_conformity(rule, scores, models)
 assessment$conformity
 assessment$plot_scores
@@ -86,6 +86,17 @@ assessment$plot_datasets
 dev.off()
 
 
-# save
+# save assessment
 write_rds(assessments, derived_file("assessments.rds"))
+
+# add image location to rules and save
+# source(script_file("helper-create_perturbation_images.R))
+rules <- rules %>%
+  mutate(
+    image_location = map_chr(id, function(id) figure_file(paste0("images/", id, ".png"))),
+    image_found = file.exists(image_location)
+  )
+
+if (any(!rules$image_found)) {stop("Files not found: ", basename(rules$image_location[!rules$image_found]))}
+
 write_rds(rules, derived_file("rules.rds"))

### 2018-08-09 scripts/03-metric_characterisation/01-metric_conformity/helper-rules.R

@@ -40,7 +40,9 @@ is_monotonic_changing <- function(x, y, decreasing = TRUE) {
 
 equal_identity <- lst(
   id = "equal_identity",
-  description = "The score should always be the same when comparing the gold standard to itself",
+  name = "Same score on identity",
+  description = "The score should be approximately the same when comparing the trajectory to itself",
+  conforms_if = "1 \\leqslant \\mathit{score} \\leqslant 0.99",
   crossing = crossing(
     dataset_id = dataset_design %>% filter(topology_id %in% names(topologies), num_cells > 50) %>% pull(dataset_id),
     method_id = "identity"
@@ -102,13 +104,16 @@ equal_identity <- lst(
 # create a rule which checks whether the score decreases compared to gold standard
 rule_lower <- function(
   id,
+  name,
   description,
   dataset_ids,
   method_id
 ) {
   lst(
     id = id,
+    name = name,
     description = description,
+    conforms_if = "\\mathit{score}_{\\textit{identity}} > \\mathit{score}_{\\textit{prediction}}",
     crossing = crossing(
       dataset_id = dataset_ids,
       method_id = c("identity", method_id)
@@ -188,27 +193,31 @@ rule_lower <- function(
 
 switch_cells_edgewise <- rule_lower(
   id = "switch_cells_edgewise",
-  description = "Switching the positions of cells within each edge should lower the score. This can be seen as a local randomisation in cellular ordering",
+  name = "Local cell shuffling",
+  description = "Shuffling the positions of cells within each edge should lower the score. This is equivalent to shuffling the cellular position locally.",
   dataset_ids = dataset_design %>% filter(topology_id %in% names(topologies), cell_positioning != "milestones", num_cells > 50) %>% pull(dataset_id),
   method_id = "switch_cells_edgewise"
 )
 
 merge_bifurcation <- rule_lower(
   id = "merge_bifurcation",
-  description = "Merging the two branches after a bifurcation should lower the score",
+  name = "Bifurcation merging",
+  description = "Merging the two branches after a bifurcation point should lower the score",
   dataset_ids = dataset_design %>% filter(topology_id == "bifurcation_simple") %>% pull(dataset_id),
   method_id = "merge_bifurcation"
 )
 
 concatenate_bifurcation <- rule_lower(
   id = "concatenate_bifurcation",
[…]

### 2018-08-09 scripts/01-datasets/README.Rmd
add dataset readme

---
title: "Datasets"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The datasets are split in real datasets and synthetic datasets. The real datasets are downloaded and preprocessed first, and characteristics from these datasets (such as the number of cells and genes, library sizes, dropout probabilities, ...) are used to generate synthetic datasets. Finally, these dataset are uploaded to zenodo.

All datasets can be downloaded from zenodo using the [00-download_from_zenodo.R](00-download_from_zenodo.R) script.

```{r child='01-real/README.Rmd'}

```

```{r child='02-synthetic/README.Rmd'}

```

### 2018-08-09 scripts/01-datasets/02-synthetic/README.Rmd
add dataset readme

---
output: github_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synthetic datasets

### 2018-08-10 scripts/01-datasets/01-real/README.Rmd

---
output: github_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Real datasets

The generation of the real datasets is divided in two parts:

1. The datasets are downloaded from their respective sources (such as GEO) using sets of scripts specific for each dataset in [01-download_from_sources.R](01-download_from_sources.R) (these scripts are located in [helpers-download_from_sources](helpers-download_from_sources))
2. These datasets are filtered and normalised in [02-filter_and_normalise.R](02-filter_and_normalise.R)

### 2018-08-10 scripts/01-datasets/02-synthetic/README.Rmd
add some readme to the datasets

@@
 ## Synthetic datasets
+
+Each synthetic dataset is based on some characteristics of some real datasets. These characteristics include:
+
+* The number of cells and features
+* The number of features which are differentially expressed in the trajectory
+* Estimates of the distribution of the library sizes, average expression, dropout probabilities, ... estimated by [Splatter](https://github.com/Oshlack/splatter)
+
+These are estimated in [01-estimate_platform.R](01-estimate_platform.R).
+
+Next, we simulate datasets using different simulators:
+
+* [Dyngen](https://github.com/dynverse/dyngen), simulations of regulatory networks which will produce a particular trajectory
+* [PROSSTT](https://github.com/soedinglab/prosstt), simulations of tree topologies using random walks
+* [Splatter](https://github.com/Oshlack/splatter), simulations of non-linear paths between different states
+* [Dynplot](https://github.com/dynverse/dynplot), simulations of toy data using random expression gradients in a reduced space

### 2018-08-10 scripts/03-metric_characterisation/01-metric_conformity/04-create_table.R
large update to metric conformity, add new scores and rules

@@ -1,23 +1,0 @@
-library(dynbenchmark)
-library(tidyverse)
-
-experiment("03-metric_characterisation/01-metric_conformity")
-
-assessments <- read_rds(derived_file("assessments.rds"))
-rules <- read_rds(derived_file("rules.rds"))
-
-assessments %>%
-  left_join(rules %>% select(id, description), by = c("rule_id" = "id")) %>%
-  unnest(conformity) %>%
-  mutate(
-    metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder(),
-    conforms = kableExtra::cell_spec(
-      ifelse(conforms, "\U2714", "\U2716"),
-      background = ifelse(conforms, "#2ECC40", "#FF4136"),
-      format = "latex"
-    )
-  ) %>%
-  mutate(metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder()) %>%
-  spread(metric_id, conforms) %>%
-  knitr::kable(format = "latex") %>%
-  kableExtra::kable_styling()

### 2018-08-10 manuscript/supplementary_notes/metric_conformity/metric_conformity.Rmd

@@
     keep_tex: true
+header-includes:
+  - \usepackage{tabularx}
 mainfont: DejaVu Sans
@@
       format = "latex"
-    )
+    ),
+    rule_id = forcats::fct_inorder(rule_id)
   ) %>%
@@
   select(-rule_id) %>% 
-  knitr::kable(format = "latex", escape = FALSE) %>%
-  kableExtra::kable_styling(row_label_position = "r")
+  knitr::kable(format = "latex", escape = FALSE, booktabs = TRUE) %>%
+  kableExtra::kable_styling(full_width = FALSE) %>% 
+  kableExtra::column_spec(1, width="15em") %>% 
+  kableExtra::column_spec(2:20, width = "3em") %>% 
+  kableExtra::row_spec(0, angle = 30) %>% 
+  gsub("\\\\addlinespace", "", .)
 ```
@@
 
-walkdf(assessments[1, ], function(assessment) {
+walkdf(assessments[, ], function(assessment) {
   cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))

### 2018-08-10 scripts/01-datasets/02-synthetic/README.Rmd

@@
 
-These are estimated in [01-estimate_platform.R](01-estimate_platform.R).
+These are estimated in [01-estimate_platform.R](01-estimate_platform.R) and are called "platforms".
 
@@
 * [Dynplot](https://github.com/dynverse/dynplot), simulations of toy data using random expression gradients in a reduced space
+
+Each simulation script (02a-02d) first creates a design dataframe, which links particular platforms, different topologies, seeds and other parameters specific for a simulator.
+
+Next it calls the wrappers around the simulators (see [/package/R/simulators.R](/package/R/simulators.R))

### 2018-08-10 scripts/03-metric_characterisation/02-metric_examples/4-edge_flip.Rmd

--- 2018-06-08 analysis/analyses/3-metric_characterisation/4-edge_flip.Rmd
@@
 ```{r setup}
+knitr::opts_chunk$set(echo = TRUE)
 knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
-library(dynalysis)
+```
+
+```{r}
+library(dynbenchmark)
 library(tidyverse)
@@
 library(cowplot)
-```
 
-```{r}
-experiment("3-score_rules/4-edge_flip")
-
-knitr::opts_chunk$set(echo = TRUE)
+experiment("03-score_rules/02-metric_examples")
 ```
@@
 topologies <- eval(formals(dyntoy:::generate_milestone_network)$model)
-topologies <- topologies[topologies != "BA"]
-networks_synthetic <- map(topologies, dyntoy:::generate_milestone_network) %>% tibble(milestone_network = ., topology = topologies)
-```
+networks <- map(topologies, dyntoy:::generate_milestone_network) %>% tibble(milestone_network = ., topology = topologies)
 
-Add SCORPIUS predictions
-```{r}
-ti_methods <- list(ti_scorpius())
-data <- dyntoy::generate_toy_datasets(c("linear"), num_replicates=1, num_cells=30, num_genes=10) %>% dynutils::extract_row_to_list(1)
-
-results <- map(ti_methods, ~.$run_fun(data$counts))
-
-networks_methods <- tibble(topology=map_chr(ti_methods, "name"), milestone_network=map(results, "milestone_network"))
-```
-
-```{r}
-networks <- bind_rows(networks_synthetic, networks_methods)
 design <- crossing(networks, networks)
@@
   start <- Sys.time()
-  score <- dyneval:::calculate_edge_flip(milestone_network, milestone_network1)
+  score <- dyneval:::calculate_edge_flip(milestone_network, milestone_network1, limit_flips = 10)
   time <- Sys.time() - start

### 2018-08-10 scripts/01-datasets/05-dataset_characterisation/3-trajectory_type_dag.R
02-dataset_characterisation -> 01/datasets/05-dataset_characterisation

@@ -3,7 +3,7 @@ library(tidygraph)
 library(ggraph)
 library(tidyverse)
 
-experiment("02-dataset_characterisation/3-trajectory_type_dag")
+experiment("01-datasets/05-dataset_characterisation/3-trajectory_type_dag")
 
 trajectory_type_ancestors <- trajectory_type_dag %>% igraph::ego(99999999, mode = "out") %>% map(names) %>% setNames(names(igraph::V(trajectory_type_dag)))
 

### 2018-08-10 scripts/01-datasets/01-real/README.Rmd

@@
 The datasets are downloaded from their respective sources (such as GEO) using sets of scripts specific for each dataset in [01-download_from_sources.R](01-download_from_sources.R) (these scripts are located in [helpers-download_from_sources](helpers-download_from_sources))
-2. These datasets are filtered and normalised in [02-filter_and_normalise.R](02-filter_and_normalise.R)
+2. These datasets are filtered and normalised in [02-filter_and_normalise.R](02-filter_and_normalise.R). This is done using the dynbenchmark::process_raw_dataset function.
+
+Following datasets were ultimately generated:
+
+```{r}
+# library(tidyverse)
+# library(dynbenchmark)
+# list_datasets() %>% 
+#   filter(source == "real") %>% 
+#   select(name)
+```

### 2018-08-10 manuscript/paper.Rmd

@@
 add_stable(
-  read_rds(figure_file("datasets.rds","02-dataset_characterisation"))[[params$table_format]], 
+  read_rds(figure_file("datasets.rds","01-datasets/05-dataset_characterisation"))[[params$table_format]], 
   "datasets", 
@@
     "samplers.rds",
-    "02-dataset_characterisation/3-synthetic")
+    "01-datasets/05-dataset_characterisation/3-synthetic")
   )[[params$table_format]], 

### 2018-08-10 manuscript/paper_latex.Rmd
rename method_characterisation

@@
 add_stable(
-  read_rds(figure_file("implementations_table.rds", "04-method_characterisation"))[[params$table_format]],
+  read_rds(figure_file("implementations_table.rds", "03-method_characterisation"))[[params$table_format]],
 "implementations",
@@
 add_fig(
-  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "04-method_characterisation"),
+  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "03-method_characterisation"),
   ref_id = "implementation_qc_overview", 
@@
     "qc_checks.rds",
-    "04-method_characterisation"
+    "03-method_characterisation"
   ))[[params$table_format]], 

### 2018-08-10 scripts/08-aggregate/old10-aggregations.R

@@ -22,7 +22,7 @@ indrep_scores <- outputs_list$outputs_summrepl %>%
 ind_scores <- outputs_list$outputs_ind %>%
   inner_join(methods, "method_short_name")
 
-methods <- read_rds(derived_file("methods.rds", experiment_id = "04-method_characterisation")) %>%
+methods <- read_rds(derived_file("methods.rds", experiment_id = "03-method_characterisation")) %>%
   filter(type %in% c("algorithm", "control")) %>%
   filter(method_id %in% overall_scores$method_short_name)
 

### 2018-08-10 scripts/old/old6-performance_predictors/prior_information_importance.R

@@ -6,7 +6,7 @@ experiment("6-performance_predictors")
 
 read_rds(derived_file("evaluation_algorithm.rds", "06-optimise_parameters/10-aggregations")) %>% list2env(.GlobalEnv)
 
-methods <- read_rds(derived_file("methods.rds", experiment_id = "04-method_characterisation"))
+methods <- read_rds(derived_file("methods.rds", experiment_id = "03-method_characterisation"))
 
 method_priors <- methods %>%
   filter(method_short_name %in% overall_scores$method_short_name) %>%

### 2018-08-10 manuscript/paper.Rmd
rename method_characterisation

@@
 add_fig(
-  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "04-method_characterisation"),
+  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "03-method_characterisation"),
   ref_id = "implementation_qc_overview", 

### 2018-08-10 scripts/03-method_characterisation/10_create_github_issue.R

@@ -5,7 +5,7 @@ library(tidyverse)
 library(cowplot)
 library(googlesheets)
 
-experiment("04-method_characterisation")
+experiment("03-method_characterisation")
 
 
 #   ____________________________________________________________________________

### 2018-08-10 scripts/03-method_characterisation/11_create_singularity_images.R

@@ -3,7 +3,7 @@ library(dynwrap)
 library(dynbenchmark)
 library(tidyverse)
 
-experiment("04-method_characterisation/singularity_images")
+experiment("03-method_characterisation/singularity_images")
 
 # get methods
 data("methods", package = "dynmethods")

### 2018-08-10 scripts/old/old8-topology_detection/2-topology_sensitivity.R

@@ -7,7 +7,7 @@ library(ggraph)
 
 experiment("8-compare_topology")
 
-methods <- read_rds(derived_file("methods.rds", experiment_id = "04-method_characterisation"))
+methods <- read_rds(derived_file("methods.rds", experiment_id = "03-method_characterisation"))
 method_trajtypes <- methods %>%
   gather(trajectory_type, can_handle_trajectory_type, !!trajectory_types$id[trajectory_types$directedness == "directed"])
 

### 2018-08-10 manuscript/paper.Rmd
rename metric characterisation

@@
 add_sfig(
-  figure_file("edge_flip_comparison.svg", experiment_id="03-metric_characterisation/4-edge_flip"), 
+  figure_file("edge_flip_comparison.svg", experiment_id="02-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 

### 2018-08-10 scripts/02-metric_characterisation/01-metric_conformity/01-generate_reference_datasets.R

@@ -4,7 +4,7 @@ library(tidyverse)
 library(dynbenchmark)
 library(dyntoy)
 
-experiment("03-metric_characterisation/01-metric_conformity")
+experiment("02-metric_characterisation/01-metric_conformity")
 
 source(scripts_file("helper-topologies.R"))
 

### 2018-08-10 scripts/02-metric_characterisation/02-metric_examples/6-find_examples.R

@@ -3,7 +3,7 @@ library(dynbenchmark)
 library(dynplot)
 library(PRISM)
 
-experiment("03-metric_characterisation/6-find_examples")
+experiment("02-metric_characterisation/6-find_examples")
 
 cells_per_group <- 5
 num_groups <- 3

### 2018-08-10 manuscript/paper_latex.Rmd
rename metric characterisation

@@
 add_sfig(
-  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="03-metric_characterisation/4-edge_flip"), 
+  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="02-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 

### 2018-08-10 scripts/04-method_testing/README.Rmd

---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Method testing

Here, we test the methods on some small real and toy datasets, to capture errors early on and make sure all methods run fine on these "simple" datasets.

These testing results are in the end pushed to [https://dynverse.github.io/dynmethods_automated_testing/index.html](https://dynverse.github.io/dynmethods_automated_testing/index.html)

### 2018-08-10 scripts/README.Rmd

@@
 
-```{r setup, echo = FALSE}
+```{r echo = FALSE}
 knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE, echo = FALSE)
@@
   knitr::kable()
-
 ```

### 2018-08-13 manuscript/supplementary_notes/metrics/assessment.Rmd
update to metric characterisation

--- 2018-08-09 manuscript/supplementary_notes/metric_conformity/assessment.Rmd
@@
 
+\subsection{`r rule$name`}
+
 \begin{center}
@@
 ```
+
 \end{center}
@@
 
+`r rule$observation`
+
 ```{r}
-width <- ifelse(is.null(assessment$plot_datasets$width), 1, assessment$plot_datasets$width)
-height <- ifelse(is.null(assessment$plot_datasets$height), 1, assessment$plot_datasets$height)
+width <- ifelse(is.null(assessment$plot_datasets$width), 10, assessment$plot_datasets$width)
+height <- ifelse(is.null(assessment$plot_datasets$height), 10, assessment$plot_datasets$height)
 ```
@@
 ```{r}
-width <- ifelse(is.null(assessment$plot_scores$width), 1, assessment$plot_scores$width)
-height <- ifelse(is.null(assessment$plot_scores$height), 1, assessment$plot_scores$height)
+width <- ifelse(is.null(assessment$plot_scores$width), 10, assessment$plot_scores$width)
+height <- ifelse(is.null(assessment$plot_scores$height), 10, assessment$plot_scores$height)
 ```
@@
 
-----------------------------------------------------------------------------
-
 \pagebreak

### 2018-08-13 scripts/02-metric_characterisation/01-metric_conformity/02-evaluate_perturbations.R

@@ -4,16 +4,9 @@ library(dynbenchmark)
 experiment("02-metric_characterisation/01-metric_conformity")
 
 dataset_design <- read_rds(derived_file("dataset_design.rds"))
-datasets <- read_rds(derived_file("datasets.rds"))
 
 # load perturbations = dynwrap::ti_methods
 source(scripts_file("helper-perturbations.R"))
-perturbation_methods <- ls() %>% str_subset("^perturb_") %>% map(function(x) {
-  id <- str_replace(x, "perturb_(.*)", "\\1")
-  run_fun <- get(x)
-
-  create_ti_method(id = id, run_fun = run_fun)()
-})
 
 # load rules
 source(scripts_file("helper-rules.R"))
@@ -32,7 +25,7 @@ crossing <- mapdf(rules, function(rule) {
 crossing <- crossing[!duplicated(crossing),]
 
 # join all parameters from the different rules and remove duplicates
-parameters <- map(perturbation_methods %>% map_chr("id"), function(method_id) {
+parameters <- map(perturbation_methods$id, function(method_id) {
   parameters <- rules %>% mapdf(
     function(rule) {
       if(method_id %in% names(rule$parameters)) {
@@ -43,27 +36,36 @@ parameters <- map(perturbation_methods %>% map_chr("id"), function(method_id) {
     }) %>% bind_rows()
   parameters <- parameters[!duplicated(parameters),]
   parameters
-}) %>% set_names(perturbation_methods %>% map_chr("id"))
+}) %>% set_names(perturbation_methods$id)
+
+# get dataset functions
+datasets <- dynbenchmark::process_datasets_design(dataset_design %>% select(dataset_id, dataset) %>% deframe())
 
 design <- benchmark_generate_design(
   datasets = datasets,
   methods = perturbation_methods,
   parameters = parameters,
   num_repeats = 1,
-  crossing = crossing# %>% filter(method_id == "remove_divergence_regions") %>% sample_n(3)
+  crossing = crossing# %>% filter(method_id == "identity") %>% sample_n(3)
 )
 
 
 ##  ............................................................................
 ##  Run the crossing                                                        ####
-metrics <- dyneval::metrics$metric_id
+metric_ids <- metrics_characterised$metric_id
 
 # run evaluation on cluster
-benchmark_submit(design, metrics = metrics, qsub_params = list(memory = "2G", timeout = 3600), qsub_grouping = "{method_id}")
+benchmark_submit(
[…]

### 2018-08-13 scripts/02-metric_characterisation/01-metric_conformity/helper-perturbations.R

@@ -513,3 +513,17 @@ perturb_shuffle_cells_and_merge_bifurcation <- function(dataset, shuffle_perc = 
 
   dataset
 }
+
+
+
+
+##  ............................................................................
+##  Combine all perturbation methods                                        ####
+perturbation_methods <- ls() %>% str_subset("^perturb_") %>% map(function(x) {
+  id <- str_replace(x, "perturb_(.*)", "\\1")
+  run_fun <- get(x)
+
+  create_ti_method(id = id, run_fun = run_fun)()
+})
+
+perturbation_methods <- dynbenchmark::process_methods_design(perturbation_methods)

### 2018-08-13 manuscript/supplementary_notes/metrics/metrics.Rmd
update to metric characterisation

--- 2018-08-10 manuscript/supplementary_notes/metric_conformity/metric_conformity.Rmd
@@
 ---
-title: "Metric conformity"
+title: "Evaluation metrics"
 output: 
@@
 
-experiment("03-metric_characterisation/01-metric_conformity")
+experiment("02-metric_characterisation/01-metric_conformity")
 ```
@@
 ```
+
+A trajectory, as defined in our evaluation, is a model with multiple abstractions.
+The top abstraction is the topology, represented by a milestone network, which contains information about the paths each cell can take from their starting point.
+More complex abstractions involve the mapping of each cell to a particular branch within this network, and the ordering of each cells within these branches (both represented in the milestone percentages).
+
+When designing metrics to compare trajectories, we therefore
+
+- A good overall metric should capture all these different abstractions, so that the resulting trajectory has the best topology, branch assignment and cellular ordering.
+- IndividuSome users are more interested in particular abstractions. For example, the actual
+
+Here we define several metrics
 
@@
 
-walkdf(assessments[, ], function(assessment) {
+walkdf(assessments, function(assessment) {
   cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))

### 2018-08-20 manuscript/supplementary_notes/metrics/metrics.Rmd

--- 2018-08-10 manuscript/supplementary_notes/metric_conformity/metric_conformity.Rmd
@@
 ---
-title: "Metric conformity"
+title: "Supplementary Note 1: Evaluation metrics"
 output: 
@@
     keep_tex: true
+    number_sections: true
 header-includes:
   - \usepackage{tabularx}
+  - \newcommand{\hideFromPandoc}[1]{#1}
+  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
 mainfont: DejaVu Sans
@@
 )
-options(knitr.duplicate.label = 'allow')
+options(knitr.duplicate.label='allow')
 
@@
 
-experiment("03-metric_characterisation/01-metric_conformity")
+experiment("02-metric_characterisation/01-metric_conformity")
+
+refs <- setup_refs()
+figs <- setup_figs()
 ```
+
+A trajectory, as defined in our evaluation, is a model with multiple abstractions.
+The top abstraction is the topology which contains information about the paths each cell can take from their starting point.
+Deeper abstractions involve the mapping of each cell to a particular branch within this network, and the position (or ordering) of each cells within these branches.
+Internally, the topology is represented by the milestone network and regions of delayed commitment, the branch assignment and cellular positions are represented by the milestone percentages (`r ref("fig", "trajectory_model_example")`).
+
+```{r, results = "asis"}
+add_fig(
+  figure_file("trajectory_model_example.svg", "manual_figures"),
+  "trajectory_model_example",
+  "An example trajectory that will be used throughout this section.",
+  "It contains contains four milestones (W to Z) and five cells (a to e)."
+)
+```
+
+Given the multilayered complexity of a trajectory model, it is not trivial to compare two trajectory models (for example a gold standard a prediction).
+We therefore sought to use different comparison metrics, each belonging to one particular type:
+
+- **Specific metrics** investigate one particular aspect of the trajectory.
+This would allow us to find particular weak points for methods, e.g.
+that a method is very good at ordering but does not frequently find the correct topology.
+Moreover, this makes it possible to create personalised rankings of methods, for example for users which are primarily interested in using the method correct topology.
+- **Overall metrics** should capture all the different abstractions, i.e.
+it measures whether the resulting trajectory has a good topology, that the cells belong to similar branches _and_ that they are ordered correctly.
+- **Application metrics** focus on the quality of a downstream analysis using the trajectory.
+For example, it measures whether the trajectory can be used to find accurate differentially expressed genes.
+
+Here, we first described and illustrate several possible metrics which we defined (`r ref("section", "metrics")`).
+Next, we tested these metrics on several test cases, to make sure they were robustly identify different type of "wrong" trajectory predictions (`r ref("section", "conformity")`).
+Based on this robustness assessment, we chose 4 metrics for the final evaluation (`r ref("section", "final")`).
+
+# Metrics
+
+All metrics described here were implemented within our dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).
+
+## $`r label_metric("correlation", "latex")`$: Correlation between geodesic distances
+
+When the position of a cell is the same in both the gold standard and the prediction, its distances to all other cells in the trajectory should also be the same.
+This observation is the basis for the $`r label_metric("correlation", "latex")`$ metric.
+
+```{r, results = "asis"}
+add_fig(
+  figure_file("metrics_geodesic.svg", "manual_figures"),
+  "metrics_geodesic",
+  "The calculation of geodesic distances on a small example trajectory.",
+  "a) A toy example containing four milestones (W to Z) and five cells (a to e).
+b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
+c) The calculations made for calculating the pairwise geodesic distances.
+d) A heatmap representation of the pairwise geodesic distances."
+)
+```
+
+The geodesic distance is the distance a cell has to through the trajectory space to get from one position to another (`r ref("fig", "metrics_geodesic")`).
+The way this distance is calculated depends on how two cells are positioned:
+
+- **Both cells are on the same edge in the milestone network.** In this case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on.
+The geodesic distance is defined as the product of the difference in milestone percentages and the length of their common edge.
+For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
+- **Cells reside on different edges in the milestone network.** First, the distance of the cell to all its nearby milestones is calculated, based on its percentage within the edge and the length of the edge.
+These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
+For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
+
+The geodesic distance can be easily extended towards cells within regions of delayed commitment.
+When both cells are part of the same region of delayed commitment, the geodesic distance was defined as the manhattan distances between the milestone percentages weighted by the lengths from the milestone network.
+For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
+The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
+
+Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets.
+For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
+The waypoints are determined by viewing each milestone, transition and region of delayed commitment as a collection of cells, and sampling cells from the different collections weighted by the total number of cells within that collection.
+For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
+For the benchmark evaluation, the total number of waypoints sampled from a trajectory was set to 100.
+
+## $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory
+
+An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory.
+If the prediction error for a particular cell is low, the more similar its positions are between the two trajectories.
+
+Specifically, we implemented two metrics which predict the milestone percentages from the gold standards by using the predicted milestone percentages as features  (`r ref("fig", "metrics_prediction")`).
+We did this with two regression methods, linear regression ($\textit{lm}$, using the R `lm` function) and Random Forest ($\textit{rf}$, implemented in the *ranger* package [@wright_rangerfastimplementation_2017]).
+In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error.
+Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average.
+We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$.
+We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$ scores.
+
+```{r, results = "asis"}
+add_fig(
+  figure_file("metrics_prediction.svg", "manual_figures"),
+  "metrics_prediction",
+  glue::glue("The calculation of ${label_metric('lm_nmse', 'latex')}$ distances on a small example trajectory."),
+  ""
+)
+```
+
+## $`r label_metric("edgeflip", "latex")`$: Number of edge additions and removals to convert one topology to another
+
+We first simplified each network, by merging consecutive linear edges into one edge, and adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks.
+Because we are interested in the overall similarity between two topologies irrespective of the direction of the edges, the network was made undirected.
+Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
+This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012].
+We implemented a branch and bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, we tried out all solutions with additional two edge changes.
+To further limit the search space, we made sure the degree distributions between the two networks were  similar, before assessing whether the two networks were isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
+
+## $`r label_metric("edgeflip", "latex")`$: Number of edge additions and removals to convert one topology to another
+
+# Metric conformity
 
@@
 ```
-
-## Summary
 
@@
 
-walkdf(assessments[, ], function(assessment) {
+walkdf(assessments[1,], function(assessment) {
   cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))

### 2018-08-23 manuscript/supplementary_notes/metrics/metrics.Rmd
git-annex in wouters

@@
 add_fig(
-  figure_file("trajectory_model_example.svg", "manual_figures"),
+  result_file("trajectory_model_example.svg", "manual_figures"),
   "trajectory_model_example",
@@
 add_fig(
-  figure_file("metrics_geodesic.svg", "manual_figures"),
+  result_file("metrics_geodesic.svg", "manual_figures"),
   "metrics_geodesic",
@@
 For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
-The waypoints are determined by viewing each milestone, transition and region of delayed commitment as a collection of cells, and sampling cells from the different collections weighted by the total number of cells within that collection.
+These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells.
+We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection.
 For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
-For the benchmark evaluation, the total number of waypoints sampled from a trajectory was set to 100.
+
+The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$.
+To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$.
 
@@
 add_fig(
-  figure_file("metrics_prediction.svg", "manual_figures"),
+  result_file("metrics_prediction.svg", "manual_figures"),
   "metrics_prediction",
@@
 
-## $`r label_metric("edgeflip", "latex")`$: Number of edge additions and removals to convert one topology to another
+## $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("isomorphic", "latex")`$: Number of edge additions and removals to convert one topology to another
 
-We first simplified each network, by merging consecutive linear edges into one edge, and adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks.
-Because we are interested in the overall similarity between two topologies irrespective of the direction of the edges, the network was made undirected.
-Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
+We used three different scores to assess the similarity in the topology between two trajectories, iregardless of where the cells were positioned.
+
+For all three scores, we first simplified the topology of the trajectory to make both graph structures comparable:
+
+- As we are only interested in the main structure of the topology without start or end, the graph was made undirected.
+- All nodes with degree 2 were removed. For example in the topology A ⇨ B ⇨ C ⇨ D, C ⇨ D, the nodes B was removed
+- A linear topology was converted to A ⇨ B ⇨ C
+- A cyclical topology such as A ⇨ B ⇨ C ⇨ D or A ⇨ B ⇨ A were all simplified to A ⇨ B ⇨ C ⇨ A
+- Duplicated edges such as A ⇨ B, A ⇨ B were decoupled to A ⇨ B, A ⇨ C ⇨ B
+
+The $`r label_metric("isomorphic", "latex")`$ score returns 1 if two graphs are isomorphic, and 0 if they were not.
+For this, we used the used the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
+
+The $`r label_metric("edge_flip", "latex")`$ score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
 This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012].
-We implemented a branch and bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, we tried out all solutions with additional two edge changes.
-To further limit the search space, we made sure the degree distributions between the two networks were  similar, before assessing whether the two networks were isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
+We implemented a branch and bound approach for this problem, using several heuristics to make speed up the search:
 
-## $`r label_metric("edgeflip", "latex")`$: Number of edge additions and removals to convert one topology to another
+- First check all possible edge additions and removals corresponding to the number of different edges between the two graphs.
+If no solution is found, check all possible solutions with two extra edge additions/removals.
+- For each possible solution, first check whether:
+  - The maximal degree is the same
+  - The minimal degree is the same
+  - All degrees are the same after sorting
+- Only then check if the two graphs are isomorphic as described earlier.
+
+The $`r label_metric("him", "latex")`$ metric (Hamming-Ipsen-Mikhailov distance) [@jurmanHIMGlocalMetric2015] which was adopted from the R nettools package [https://github.com/filosi/nettools](https://github.com/filosi/nettools).
+It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network.
+Conceptually, $`r label_metric("him", "latex")`$ is a linear combination of:
+
+- The normalised Hamming distance [@doughertyValidationGeneRegulatory2011], which calculates the distance between two graphs by matching individual edges in the adjacency matrix, but disregards overall structural similarity.
+- The normalised Ipsen-Mikhailov distance [@ipsenEvolutionaryReconstructionNetworks2002], which calculates the overall distance of two graphs based on matches between its degree and adjacency matrix, while disregarding local structural similarities.
+It requires a $\gamma$ parameter, which is usually estimated based on the number of nodes in the graph, but which we fixed at $0.1$ so as to make the score comparable across different graph sizes.
+
+We compared the three scores for several common topologies `r ref("fig", "topology_scores_overview")`.
+Although conceptually different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores `r ref("fig", "topology_scores_overview", "b")`, although the $`r label_metric("him", "latex")`$ tends to punish cycles more, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations `r ref("fig", "topology_scores_overview", "c")`.
+
+```{r, results = "asis"}
+add_fig(
+  result_file("topology_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
+  "topology_scores_overview",
+  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ ."),
+  "",
+  width = 12,
+  height = 12
+)
+```
+
+The different topology based scores are useful for different scenarios:
+
+- If the two trajectories are only comparable when the topology is exactly the same, the $`r label_metric("isomorphic", "latex")`$ should be used.
+- If it is important that the topologies are similar (but possible not isomorphic), the $`r label_metric("edge_flip", "latex")`$ is most appropriate.
+- If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the $`r label_metric("HIM", "latex")`$ is most appropriate.
 

### 2018-08-23 scripts/02-metric_characterisation/02-individual_metrics/01-correlation.R
git-annex in wouters

@@ -1,42 +1,21 @@
 library(dynbenchmark)
 library(tidyverse)
 
-original_dataset_design <- crossing(
-  num_cells = 1000
-)
+experiment("02-metric_characterisation/02-individual_metrics")
 
-original_dataset_design$dataset <- original_dataset_design %>% pmap(function(num_cells, ...)  {
-  print(num_cells)
-  dataset <- dyntoy::generate_trajectory(model = "bifurcating", num_cells = num_cells)
-})
+# select the most complex dataset of all our datasets (excluding disconnected)
+dataset <- load_datasets(list_datasets() %>% filter(startsWith(source, "synthetic")) %>% pull(id)) %>%
+  filter(trajectory_type != "directed_disconnected_graph") %>%
+  mutate(n_edges = map_int(milestone_network, nrow)) %>%
+  top_n(1, n_edges) %>%
+  extract_row_to_list(1)
 
-# add varying number of cell waypoints
 dataset_design <- crossing(
-  original_dataset_design,
-  n_cell_waypoints = c(2, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 150, 200, NA),
-  repeat_ix = 1:3,
-  seed = repeat_ix
-)
-
-dataset_design$dataset <- dataset_design %>% pmap(function(seed, dataset, n_cell_waypoints, ...)  {
-  set.seed(seed)
-  if (is.na(n_cell_waypoints)) {
-    n_cell_waypoints <- length(dataset$cell_ids)
-  }
-
-  dataset %>% add_cell_waypoints(n_cell_waypoints)
-})
-
-# calculate geodesic distances
-dataset_design$geodesic_distances <- map(dataset_design$dataset, function(x) {
-  print(length(x$cell_ids))
-  start <- Sys.time()
-  distances <- dynwrap::compute_tented_geodesic_distances(x, waypoint_cells = x$waypoint_cells)
-  time <- Sys.time() - start
-
-  tibble(distances = list(distances), time = as.numeric(time))
-})
-dataset_design <- dataset_design %>% unnest(geodesic_distances)
+  n_cell_waypoints = c(2, 5, 10, 20, 50, 100, 200, 500, 1000),
+  repeat_ix = 1:100,
+  shuffle_perc = c(0, 0.2, 0.4, 0.6, 0.8, 1)
+) %>%
+  mutate(seed = repeat_ix)
 
 # do the same for perturbations
[…]

### 2018-08-23 scripts/02-metric_characterisation/README.Rmd

---
output: github_document
---

## Metrics characterisation and testing

Good metrics are crucial for an unbiased and comprehensive benchmarking. In this experiment, we test several possible metrics for comparing two trajectories.

### Metric conformity

Differences between two datasets should be reflected in certain changes in the metrics. This can be formalised in a set of rules, for example:

* If the position of some cells are different than in the gold standard, the score should decrease.
* If the topology of the network is different than that in the gold standard, the score should not be perfect.
* The more cells are filtered from the trajectory, the more the score should decrease.

Here, we assess whether metrics conforms these rules by perturbing some datasets in several ways, and assessing how the scores are affected.

If not one metric satisfies all the rules, we try to find a combination which conforms to all rules.

### Metric examples

### 2018-08-23 manuscript/supplementary_notes/metrics/metrics.Rmd
update metric characterisation

@@
 
-Given the multilayered complexity of a trajectory model, it is not trivial to compare two trajectory models (for example a gold standard a prediction).
-We therefore sought to use different comparison metrics, each belonging to one particular type:
+Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models.
+We therefore sought to use different comparison metrics, each serving a different purpose:
 
@@
 Moreover, this makes it possible to create personalised rankings of methods, for example for users which are primarily interested in using the method correct topology.
+- **Application metrics** focus on the quality of a downstream analysis using the trajectory.
+For example, it measures whether the trajectory can be used to find accurate differentially expressed genes.
 - **Overall metrics** should capture all the different abstractions, i.e.
 it measures whether the resulting trajectory has a good topology, that the cells belong to similar branches _and_ that they are ordered correctly.
-- **Application metrics** focus on the quality of a downstream analysis using the trajectory.
-For example, it measures whether the trajectory can be used to find accurate differentially expressed genes.
 
-Here, we first described and illustrate several possible metrics which we defined (`r ref("section", "metrics")`).
-Next, we tested these metrics on several test cases, to make sure they were robustly identify different type of "wrong" trajectory predictions (`r ref("section", "conformity")`).
-Based on this robustness assessment, we chose 4 metrics for the final evaluation (`r ref("section", "final")`).
+Here, we first described and illustrate several specific, application and overall metrics which we defined.
+Next, we tested these metrics on several test cases, to make sure they were robustly identify different wrong trajectory predictions.
+Based on this robustness assessment, we chose 4 metrics for the final evaluation.
 
-# Metrics
+# Specific metrics
 
@@
 - As we are only interested in the main structure of the topology without start or end, the graph was made undirected.
-- All nodes with degree 2 were removed. For example in the topology A ⇨ B ⇨ C ⇨ D, C ⇨ D, the nodes B was removed
+- All milestones with degree 2 were removed. For example in the topology A ⇨ B ⇨ C ⇨ D, C ⇨ D, the B milestone was removed
 - A linear topology was converted to A ⇨ B ⇨ C
@@
 We compared the three scores for several common topologies `r ref("fig", "topology_scores_overview")`.
-Although conceptually different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores `r ref("fig", "topology_scores_overview", "b")`, although the $`r label_metric("him", "latex")`$ tends to punish cycles more, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations `r ref("fig", "topology_scores_overview", "c")`.
+While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores `r ref("fig", "topology_scores_overview", "b")`, although the $`r label_metric("him", "latex")`$ tends to punish cycles more, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations `r ref("fig", "topology_scores_overview", "c")`.
+Their main difference is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in `r ref("fig", "topology_scores_overview", "d")`.
 
@@
 
-The different topology based scores are useful for different scenarios:
+To summarise, the different topology based scores are useful for different scenarios:
 
-- If the two trajectories are only comparable when the topology is exactly the same, the $`r label_metric("isomorphic", "latex")`$ should be used.
-- If it is important that the topologies are similar (but possible not isomorphic), the $`r label_metric("edge_flip", "latex")`$ is most appropriate.
+- If the two trajectories should only be compared when the topology is exactly the same, the $`r label_metric("isomorphic", "latex")`$ should be used.
+- If it is important that the topologies are similar (but possible exactly isomorphic), the $`r label_metric("edge_flip", "latex")`$ is most appropriate.
 - If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the $`r label_metric("HIM", "latex")`$ is most appropriate.
+
+## $`r label_metric("F1_branches", "latex")`$ and $`r label_metrics("F1_milestones", "latex")`$: Comparing how well the cells are clustered in the trajectory
+
+# Application metrics
+
+Although most metrics described above already assess some aspects directly relevant to the user (eg. whether the method is good at finding the right topology),
+
+## $`r label_metric("featureimp_cor", "latex")`$: The similarity between features differentially expressed along the trajectory
 
@@
 
-walkdf(assessments[1,], function(assessment) {
+walkdf(assessments, function(assessment) {
   cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))
@@
 ```
+
+# Metrics used in the evaluation
+
+## $`r label_metric("harm_mean", "latex")`$ The harmonic mean

### 2018-08-24 manuscript/supplementary_notes/metrics/metrics.Rmd
update metrics

@@
 
-Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models.
+Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models using only one metric.
 We therefore sought to use different comparison metrics, each serving a different purpose:
@@
 
-## $`r label_metric("correlation", "latex")`$: Correlation between geodesic distances
-
-When the position of a cell is the same in both the gold standard and the prediction, its distances to all other cells in the trajectory should also be the same.
-This observation is the basis for the $`r label_metric("correlation", "latex")`$ metric.
-
-```{r, results = "asis"}
-add_fig(
-  result_file("metrics_geodesic.svg", "manual_figures"),
-  "metrics_geodesic",
-  "The calculation of geodesic distances on a small example trajectory.",
-  "a) A toy example containing four milestones (W to Z) and five cells (a to e).
-b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
-c) The calculations made for calculating the pairwise geodesic distances.
-d) A heatmap representation of the pairwise geodesic distances."
-)
-```
-
-The geodesic distance is the distance a cell has to through the trajectory space to get from one position to another (`r ref("fig", "metrics_geodesic")`).
-The way this distance is calculated depends on how two cells are positioned:
-
-- **Both cells are on the same edge in the milestone network.** In this case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on.
-The geodesic distance is defined as the product of the difference in milestone percentages and the length of their common edge.
-For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
-- **Cells reside on different edges in the milestone network.** First, the distance of the cell to all its nearby milestones is calculated, based on its percentage within the edge and the length of the edge.
-These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
-For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
-
-The geodesic distance can be easily extended towards cells within regions of delayed commitment.
-When both cells are part of the same region of delayed commitment, the geodesic distance was defined as the manhattan distances between the milestone percentages weighted by the lengths from the milestone network.
-For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
-The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
-
-Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets.
-For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
-These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells.
-We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection.
-For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
-
-The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$.
-To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$.
-
-## $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory
-
-An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory.
-If the prediction error for a particular cell is low, the more similar its positions are between the two trajectories.
-
-Specifically, we implemented two metrics which predict the milestone percentages from the gold standards by using the predicted milestone percentages as features  (`r ref("fig", "metrics_prediction")`).
-We did this with two regression methods, linear regression ($\textit{lm}$, using the R `lm` function) and Random Forest ($\textit{rf}$, implemented in the *ranger* package [@wright_rangerfastimplementation_2017]).
-In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error.
-Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average.
-We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$.
-We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$ scores.
-
-```{r, results = "asis"}
-add_fig(
-  result_file("metrics_prediction.svg", "manual_figures"),
-  "metrics_prediction",
-  glue::glue("The calculation of ${label_metric('lm_nmse', 'latex')}$ distances on a small example trajectory."),
-  ""
-)
-```
-
-## $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("isomorphic", "latex")`$: Number of edge additions and removals to convert one topology to another
+## $`r label_metric("isomorphic", "latex")`$, $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$: Edit distance between two trajectory topologies
 
@@
 
+Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone _or_ branch `r ref("fig", "clustering_scores_overview")`.
+These clusters of cells can then be compared using one of the many external cluster evaluation measures [@saelensComprehensiveEvaluationModule2018].
+When selecting such a metric, we had two conditions:
+
+- Because we allow some methods to filter cells in the trajectory, the metric should be able to handle "non-exhaustive assignment", where some cells are not assigned to any cluster.
+- The metric should give each cluster equal weight, so that rare cell stages are equally important as large stages.
+
+Based on these requirements, and on the analysis in [@saelensComprehensiveEvaluationModule2018], we chose the $\textrm{F1}$ score between the $\textrm{Recovery}$ and $\textrm{Relevance}$. If
+
+```{r, results = "asis"}
+add_fig(
+  result_file("plot_clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
+  "clustering_scores_overview",
+  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ .")
+)
+```
+
+## $`r label_metric("correlation", "latex")`$: Correlation between geodesic distances
+
+When the position of a cell is the same in both the gold standard and the prediction, its distances to all other cells in the trajectory should also be the same.
+This observation is the basis for the $`r label_metric("correlation", "latex")`$ metric.
+
+```{r, results = "asis"}
+add_fig(
+  result_file("metrics_geodesic.svg", "manual_figures"),
+  "metrics_geodesic",
+  "The calculation of geodesic distances on a small example trajectory.",
+  "a) A toy example containing four milestones (W to Z) and five cells (a to e).
+b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
+c) The calculations made for calculating the pairwise geodesic distances.
+d) A heatmap representation of the pairwise geodesic distances."
+)
+```
+
+The geodesic distance is the distance a cell has to through the trajectory space to get from one position to another (`r ref("fig", "metrics_geodesic")`).
+The way this distance is calculated depends on how two cells are positioned:
+
+- **Both cells are on the same edge in the milestone network.** In this case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on.
+The geodesic distance is defined as the product of the difference in milestone percentages and the length of their common edge.
+For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
+- **Cells reside on different edges in the milestone network.** First, the distance of the cell to all its nearby milestones is calculated, based on its percentage within the edge and the length of the edge.
+These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
+For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
+
+The geodesic distance can be easily extended towards cells within regions of delayed commitment.
+When both cells are part of the same region of delayed commitment, the geodesic distance was defined as the manhattan distances between the milestone percentages weighted by the lengths from the milestone network.
+For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
+The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
+
+Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets.
+For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
+These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells.
+We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection.
+For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
+
+The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$.
+To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$.
+TODO
+
+## $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory
+
+An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory.
+If the prediction error for a particular cell is low, the more similar its positions are between the two trajectories.
+
+Specifically, we implemented two metrics which predict the milestone percentages from the gold standards by using the predicted milestone percentages as features  (`r ref("fig", "metrics_prediction")`).
+We did this with two regression methods, linear regression ($\textit{lm}$, using the R `lm` function) and Random Forest ($\textit{rf}$, implemented in the *ranger* package [@wright_rangerfastimplementation_2017]).
+In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error.
+Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average.
+We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$.
+We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$ scores.
+
+```{r, results = "asis"}
+add_fig(
+  result_file("metrics_prediction.svg", "manual_figures"),
+  "metrics_prediction",
+  glue::glue("The calculation of ${label_metric('lm_nmse', 'latex')}$ distances on a small example trajectory."),
+  ""
+)
+```
+
 # Application metrics
 
-Although most metrics described above already assess some aspects directly relevant to the user (eg. whether the method is good at finding the right topology),
+Although most metrics described above already assess some aspects directly relevant to the user, such as whether the method is good at finding the right topology, these metrics do not assess the quality of downstream analyses and hypotheses which can be generated from these models.
 
-## $`r label_metric("featureimp_cor", "latex")`$: The similarity between features differentially expressed along the trajectory
+## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of differentially expressed features/genes along the trajectory
+
+One of the main advantages of studying cellular dynamic processes using single-cell -omics data is that the
 

### 2018-08-24 scripts/02-metric_characterisation/02-individual_metrics/01-correlation.R

@@ -91,6 +91,15 @@ load(derived_file("01-correlation.rda"))
 
 scores <- qsub::qsub_retrieve(handle) %>% bind_rows()
 
+save(
+  scores,
+  dataset,
+  dataset_design,
+  file = derived_file("01-correlation.rda")
+)
+
+load(derived_file("01-correlation.rda"))
+
 results <- bind_cols(
   bind_rows(scores),
   dataset_design

### 2018-08-24 manuscript/supplementary_notes/metrics/metrics.Rmd
update metrics vignette

@@
 ---
-title: "Supplementary Note 1: Evaluation metrics"
+title: "Supplementary Note 1: Metrics to compare two trajectories"
 output: 
@@
 
+All metrics described here were implemented within the dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).
+
 # Specific metrics
-
-All metrics described here were implemented within our dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).
 
@@
 This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012].
-We implemented a branch and bound approach for this problem, using several heuristics to make speed up the search:
+We implemented a branch and bound approach for this problem, using several heuristics to speed up the search:
 
@@
 
-We compared the three scores for several common topologies `r ref("fig", "topology_scores_overview")`.
-While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores `r ref("fig", "topology_scores_overview", "b")`, although the $`r label_metric("him", "latex")`$ tends to punish cycles more, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations `r ref("fig", "topology_scores_overview", "c")`.
-Their main difference is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in `r ref("fig", "topology_scores_overview", "d")`.
+We compared the three scores for several common topologies (`r ref("fig", "topology_scores_overview")`).
+While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores (`r ref("fig", "topology_scores_overview", "b")`).
+The $`r label_metric("him", "latex")`$ tends to punish the detection of cycles, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations (`r ref("fig", "topology_scores_overview", "c")`).
+The main difference however is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in (`r ref("fig", "topology_scores_overview", "d")`).
 
@@
 
-Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone _or_ branch `r ref("fig", "clustering_scores_overview")`.
+Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone _or_ branch (`r ref("fig", "clustering_scores_overview")`).
 These clusters of cells can then be compared using one of the many external cluster evaluation measures [@saelensComprehensiveEvaluationModule2018].
@@
 
-Based on these requirements, and on the analysis in [@saelensComprehensiveEvaluationModule2018], we chose the $\textrm{F1}$ score between the $\textrm{Recovery}$ and $\textrm{Relevance}$. If
+Based on these requirements, and on the analysis in [@saelensComprehensiveEvaluationModule2018], we chose the $\textrm{F1}$ score between the $\textrm{Recovery}$ and $\textrm{Relevance}$.
+If $C$ and $C'$ are two cell clusters:
+
+$$
+Jaccard(c, c') = \frac{|c \cap c'|}{|c \cup c'|}
+$$
+
+$$
+\textrm{Recovery} = \frac{1}{|C|} \sum_{c \in C}{\max_{c' \in C'}{\textrm{Jaccard(c, c')}}}
+$$
+$$
+\textrm{Relevance} = \frac{1}{|C'|} \sum_{c' \in C'}{\max_{c \in C}{\textrm{Jaccard(c, c')}}}
+$$
+$$
+\textrm{F1} = \frac{2}{\frac{1}{\textrm{Recovery}} + \frac{1}{\textrm{Relevance}}}
+$$
 
@@
 add_fig(
-  result_file("plot_clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
+  result_file("clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
   "clustering_scores_overview",
-  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ .")
+  glue::glue("Mapping cells to their closest milestone or branch for the calculation of the ${label_metric('F1_branches', 'latex')}$, ${label_metric('F1_edges', 'latex')}$ and ${label_metric('F1_branches', 'latex')}$ ."),
+  "",
+  width = 12,
+  height = 5
 )
@@
 Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets.
-For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
+For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories (`r ref("fig", "waypoints_overview", "a")`).
 These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells.
@@
 
-The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$.
-To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$.
-TODO
+To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$.
+To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the $`r label_metric("correlation", "latex")`$ at different levels of both cell shuffling and number of cell waypoints (`r ref("fig", "waypoints_overview", "b")`).
+We found that using cell waypoints does not induce a systematic bias in the $`r label_metric("correlation", "latex")`$, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.
+
+```{r, results = "asis"}
+add_fig(
+  result_file("waypoints_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
+  "waypoints_overview",
+  glue::glue("Determination of cell waypoints"),
+  glue::glue("a) Illustration of the stratified cell sampling using an example dataset (top).
+Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom).
+b) Accuracy versus time to calculate ${label_metric('correlation', 'latex')}$"),
+  width = 12, height = 8
+)
+```
 
@@
 An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory.
-If the prediction error for a particular cell is low, the more similar its positions are between the two trajectories.
+If the prediction error for a particular cell is low, the higher the similarity between the cellular positions.
 
@@
 
-## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of differentially expressed features/genes along the trajectory
+## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of dynamical differentially expressed features/genes.
 
-One of the main advantages of studying cellular dynamic processes using single-cell -omics data is that the
+Perhaps the main advantages of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome.
+This allows the study of dynamic regulatory networks and gene expression modules.
+Such analyses rely on a "good-enough" cellular ordering, so that the cellular ordering can be used to identify dynamical differentially expressed genes.
+
+To calculate the $`r label_metric("featureimp_cor", "latex")`$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory.
+Specifically, we calculated the geodesic distances for each cell to all milestones in the trajectory.
+We then trained a Random Forest regression model (implemented in the R _ranger_ package [@wright_rangerfastimplementation_2017], [https://github.com/imbs-hl/ranger](https://github.com/imbs-hl/ranger)) to predict these distances for each milestone, based on the expression of genes within each cell.
+We then extracted feature importances using the Mean Decrease in Impurity (`importance = 'impurity'` parameter of the `ranger` function), as illustrated in (`r ref("fig", "featureimp_overview.rds")`).
+The overall importance of a feature (gene) was then equal to the mean importance over all milestones.
+
+```{r, results = "asis"}
+add_fig(
+  result_file("featureimp_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
+  "featureimp_overview",
+  glue::glue(""),
+  glue::glue(""),
+  width = 12, height = 6
+)
+```
+
+Random forest regression has two main hyperparameters.
+The number of trees to be fitted (`num_trees` parameter) was fixed to `10000` to provide accurate estimates of the feature importance.
+The number of features on which can be split (`mtry` parameter) was set to 1% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated feautres (omnipresent in transcriptomics data) can all get higher on the ranking.
 

### 2018-08-27 manuscript/supplementary_notes/metrics/metrics.Rmd
update metrics

@@
 
-experiment("02-metric_characterisation/01-metric_conformity")
-
 refs <- setup_refs()
 figs <- setup_figs()
+tables <- setup_tables()
 ```
@@
 
-All metrics described here were implemented within the dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).
+All metrics described here were implemented within the _dyneval_ R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).
 
@@
 
-## $`r label_metric("isomorphic", "latex")`$, $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$: Edit distance between two trajectory topologies
+## `r label_metric("isomorphic")`, `r label_metric("edge_flip")` and `r label_metric("him")`: Edit distance between two trajectory topologies
 
@@
 
-The $`r label_metric("isomorphic", "latex")`$ score returns 1 if two graphs are isomorphic, and 0 if they were not.
+The `r label_metric("isomorphic")` score returns 1 if two graphs are isomorphic, and 0 if they were not.
 For this, we used the used the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
 
-The $`r label_metric("edge_flip", "latex")`$ score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
+The `r label_metric("edge_flip")` score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
 This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012].
@@
 
-The $`r label_metric("him", "latex")`$ metric (Hamming-Ipsen-Mikhailov distance) [@jurmanHIMGlocalMetric2015] which was adopted from the R nettools package [https://github.com/filosi/nettools](https://github.com/filosi/nettools).
+The `r label_metric("him")` metric (Hamming-Ipsen-Mikhailov distance) [@jurmanHIMGlocalMetric2015] which was adopted from the R nettools package [https://github.com/filosi/nettools](https://github.com/filosi/nettools).
 It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network.
-Conceptually, $`r label_metric("him", "latex")`$ is a linear combination of:
+Conceptually, `r label_metric("him")` is a linear combination of:
 
@@
 We compared the three scores for several common topologies (`r ref("fig", "topology_scores_overview")`).
-While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores (`r ref("fig", "topology_scores_overview", "b")`).
-The $`r label_metric("him", "latex")`$ tends to punish the detection of cycles, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations (`r ref("fig", "topology_scores_overview", "c")`).
-The main difference however is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in (`r ref("fig", "topology_scores_overview", "d")`).
+While conceptually very different, the `r label_metric("edge_flip")`r and `r label_metric("him")`r still produce similar scores (`r ref("fig", "topology_scores_overview", "b")`).
+The `r label_metric("him")`r tends to punish the detection of cycles, while the `r label_metric("edge_flip")`r is more harsh for differences in the number of bifurcations (`r ref("fig", "topology_scores_overview", "c")`).
+The main difference however is that the `r label_metric("him")`r takes into account edge lengths when comparing two trajectories, as illustrated in (`r ref("fig", "topology_scores_overview", "d")`).
 
@@
   "topology_scores_overview",
-  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ ."),
+  glue::glue("{label_metric('isomorphic')}, {label_metric('edge_flip')} and {label_metric('him')} ."),
   "",
@@
 
-- If the two trajectories should only be compared when the topology is exactly the same, the $`r label_metric("isomorphic", "latex")`$ should be used.
-- If it is important that the topologies are similar (but possible exactly isomorphic), the $`r label_metric("edge_flip", "latex")`$ is most appropriate.
-- If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the $`r label_metric("HIM", "latex")`$ is most appropriate.
+- If the two trajectories should only be compared when the topology is exactly the same, the `r label_metric("isomorphic")` should be used.
+- If it is important that the topologies are similar (but possible exactly isomorphic), the `r label_metric("edge_flip")` is most appropriate.
+- If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the `r label_metric("HIM")` is most appropriate.
 
-## $`r label_metric("F1_branches", "latex")`$ and $`r label_metrics("F1_milestones", "latex")`$: Comparing how well the cells are clustered in the trajectory
+## `r label_metric("F1_branches")` and `r label_metric("F1_milestones")`: Comparing how well the cells are clustered in the trajectory
 
@@
   "clustering_scores_overview",
-  glue::glue("Mapping cells to their closest milestone or branch for the calculation of the ${label_metric('F1_branches', 'latex')}$, ${label_metric('F1_edges', 'latex')}$ and ${label_metric('F1_branches', 'latex')}$ ."),
+  glue::glue("Mapping cells to their closest milestone or branch for the calculation of the {label_metric('F1_milestones')} and {label_metric('F1_branches')} ."),
   "",
@@
 
-## $`r label_metric("correlation", "latex")`$: Correlation between geodesic distances
+## `r label_metric("correlation")`: Correlation between geodesic distances
 
 When the position of a cell is the same in both the gold standard and the prediction, its distances to all other cells in the trajectory should also be the same.
-This observation is the basis for the $`r label_metric("correlation", "latex")`$ metric.
+This observation is the basis for the `r label_metric("correlation")` metric.
 
@@
 We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection.
-For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
+For calculating the `r label_metric("correlation")` between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
 
-To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$.
-To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the $`r label_metric("correlation", "latex")`$ at different levels of both cell shuffling and number of cell waypoints (`r ref("fig", "waypoints_overview", "b")`).
-We found that using cell waypoints does not induce a systematic bias in the $`r label_metric("correlation", "latex")`$, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.
+To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate `r label_metric("correlation")`.
+To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the `r label_metric("correlation")` at different levels of both cell shuffling and number of cell waypoints (`r ref("fig", "waypoints_overview", "b")`).
+We found that using cell waypoints does not induce a systematic bias in the `r label_metric("correlation")`, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.
 
@@
 Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom).
-b) Accuracy versus time to calculate ${label_metric('correlation', 'latex')}$"),
+b) Accuracy versus time to calculate {label_metric('correlation')}"),
   width = 12, height = 8
@@
 
-## $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory
+## `r label_metric("rf_nmse")` and `r label_metric("lm_nmse")`: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory
 
@@
 We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$.
-We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$ scores.
+We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the `r label_metric("rf_nmse")` and `r label_metric("lm_nmse")` scores.
 
@@
   "metrics_prediction",
-  glue::glue("The calculation of ${label_metric('lm_nmse', 'latex')}$ distances on a small example trajectory."),
+  glue::glue("The calculation of {label_metric('lm_nmse')} distances on a small example trajectory."),
   ""
@@
 
-## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of dynamical differentially expressed features/genes.
+## `r label_metric("featureimp_cor")`: The accuracy of dynamical differentially expressed features/genes.
 
 Perhaps the main advantages of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome.
-This allows the study of dynamic regulatory networks and gene expression modules.
-Such analyses rely on a "good-enough" cellular ordering, so that the cellular ordering can be used to identify dynamical differentially expressed genes.
+This can be used to construct other models such as dynamic regulatory networks and gene expression modules.
+Such analyses rely on a "good-enough" cellular ordering, so that it can be used to identify dynamical differentially expressed genes.
 
-To calculate the $`r label_metric("featureimp_cor", "latex")`$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory.
+To calculate the `r label_metric("featureimp_cor")` we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory.
 Specifically, we calculated the geodesic distances for each cell to all milestones in the trajectory.
-We then trained a Random Forest regression model (implemented in the R _ranger_ package [@wright_rangerfastimplementation_2017], [https://github.com/imbs-hl/ranger](https://github.com/imbs-hl/ranger)) to predict these distances for each milestone, based on the expression of genes within each cell.
+Next, we trained a Random Forest regression model (implemented in the R _ranger_ package [@wright_rangerfastimplementation_2017], [https://github.com/imbs-hl/ranger](https://github.com/imbs-hl/ranger)) to predict these distances for each milestone, based on the expression of genes within each cell.
 We then extracted feature importances using the Mean Decrease in Impurity (`importance = 'impurity'` parameter of the `ranger` function), as illustrated in (`r ref("fig", "featureimp_overview.rds")`).
@@
 Random forest regression has two main hyperparameters.
-The number of trees to be fitted (`num_trees` parameter) was fixed to `10000` to provide accurate estimates of the feature importance.
-The number of features on which can be split (`mtry` parameter) was set to 1% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated feautres (omnipresent in transcriptomics data) can all get higher on the ranking.
+The number of trees to be fitted (`num_trees` parameter) was fixed to `10000` to provide accurate and stable estimates of the feature importance (`r ref("fig", "featureimp_cor_distributions")`).
+The number of features on which can be split (`mtry` parameter) was set to 1% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated feautres (omnipresent in transcriptomics data) are not suppressed in the ranking.
+
+```{r, results = "asis"}
+add_fig(
+  result_file("feautreimp_cor_distributions.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
+  "feautreimp_cor_distributions",
+  glue::glue(""),
+  glue::glue(""),
+  width = 12, height = 6
+)
+```
+
+# Overall metrics
+
+An overall metric should not only focus on
 
 # Metric conformity
+
+```{r}
+experiment("02-metric_characterisation/01-metric_conformity")
+
+dataset_design <- read_rds(derived_file("dataset_design.rds"))
+```
+
+Although most metrics described in the previous section make sense intuitively, this does not necessarily mean that these metrics are robust and will generate reasonable results when used for benchmarking.
+This is because different methods and datasets will all lead to a varied set of trajectory models:
+
+- Real datasets have all cells grouped onto milestones
+- Some methods place all cells in a region of delayed commitment, others never generate a region of delayed commitment
+- Some methods always return a linear trajectory, even if a bifurcation is present in the data
+- Some methods filter cells
+
+A good metric, especially a good overall metric, should work in all these circumstances.
+To test this, we designed a set of rules to which a good metric should conform, and assessed empirically whether a metric conforms to these rules.
+
+We generated a panel of toy datasets (using our _dyntoy_ package (https://github.com/dynverse/dyntoy)) with all possible combinations of:
+
+- \# cells: `r sort(unique(dataset_design$num_cells)) %>% label_vector()`
+- \# features: 200
+- topologies: linear, bifurcation, multifurcating, tree, cycle, connected graph and disconnected graph
+- Whether cells are placed on the milestones (as in real data) or on the edges/regions of delayed commitment between the milestones (as in synthetic data)
+
+We then perturbed the trajectories in these datasets in certain ways, and tested whether the scores follow an expected pattern.
+An overview of the conformity of every metric is first given in `r ref("table", "conformity_overview")`.
+The individual rules and metric behaviour are discussed more into detail after that.
 
@@
 
-```{r}
-assessments %>%
+```{r, results="asis"}
+table <- map(c("html", "latex", "markdown"), function(format) {
+  table <- assessments %>%
     left_join(rules %>% select(id, name), by = c("rule_id" = "id")) %>%
@@
     mutate(
-    metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder(),
+      metric_id = label_metrics(metric_id) %>% as.character() %>% forcats::fct_inorder()
+    )
+
+  if (format != "markdown") {
+    table <- table %>% 
+      mutate(
         conforms = kableExtra::cell_spec(
@@
           background = ifelse(conforms, "#2ECC40", "#FF4136"),
-      format = "latex"
+          format = format
         ),
         rule_id = forcats::fct_inorder(rule_id)
-  ) %>%
+      ) 
+  } else {
+    table <- table %>% 
+      mutate(
+        conforms = ifelse(conforms, "\U2714", "\U2716")
+      )
+  }
+
+  table <- table %>%
     spread(metric_id, conforms) %>%
-  select(-rule_id) %>% 
-  knitr::kable(format = "latex", escape = FALSE, booktabs = TRUE) %>%
+    select(-rule_id)
+
+  if (format != "markdown") {
+    knitr::kable(table, format = format, escape = FALSE, booktabs = TRUE) %>% 
       kableExtra::kable_styling(full_width = FALSE) %>% 
       kableExtra::column_spec(1, width="15em") %>% 
-  kableExtra::column_spec(2:20, width = "3em") %>% 
+      kableExtra::column_spec(2:ncol(table), width = "3em") %>% 
       kableExtra::row_spec(0, angle = 30) %>% 
       gsub("\\\\addlinespace", "", .)
+  } else {
+    knitr::kable(table, format = format)
+  }
+}) %>% set_names(c("html", "latex", "markdown"))
+
+add_table(
+  table,
+  "conformity_overview",
+  "Overview of whether a particular metric conforms to a particular rule"
+)
 ```
@@
 
-walkdf(assessments, function(assessment) {
+walkdf(assessments[1, ], function(assessment) {
   cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))
@@
 ```
-
-# Metrics used in the evaluation
-
-## $`r label_metric("harm_mean", "latex")`$ The harmonic mean

### 2018-08-27 manuscript/supplementary_notes/metrics/assessment.Rmd
move figures to manual figures
update metrics

@@
 ```{r}
-rule <- dynutils::extract_row_to_list(rules, which(rules$id == assessment$rule_id))
+rule_ix <- which(rules$id == assessment$rule_id)
+rule <- dynutils::extract_row_to_list(rules, rule_ix)
 ```
 
-\subsection{`r rule$name`}
+## `r rule$name`
 
@@
   mutate(
-    metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder(),
+    metric_id = label_metrics(metric_id, format = "latex") %>% as.character() %>% forcats::fct_inorder(),
     conforms = kableExtra::cell_spec(

### 2018-08-29 manuscript/paper.Rmd

@@
 ```{r setup, include=FALSE}
-source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/setup.R"))
+source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
 
@@
 add_stable(
-  read_rds(figure_file("implementations_table.rds", "03-method_characterisation"))[[params$table_format]],
+  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
 "implementations",
@@
 add_fig(
-  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "03-method_characterisation"),
+  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "4-method_characterisation"),
   ref_id = "implementation_qc_overview", 
@@
 add_fig(
-  fig_path = figure_file("tree.svg", experiment_id = "09-user_guidelines"),
+  fig_path = figure_file("tree.svg", experiment_id = "7-user_guidelines"),
   ref_id = "user_guidelines", 
@@
 We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynbenchmark](http://www.github.com/dynverse/dynbenchmark)).
+The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
 Whenever possible, we preferred to start from the raw counts data.
@@
 add_stable(
-  read_rds(figure_file("datasets.rds","01-datasets/05-dataset_characterisation"))[[params$table_format]], 
+  read_rds(figure_file("datasets.rds","2-dataset_characterisation/2-real"))[[params$table_format]], 
   "datasets", 
@@
     "samplers.rds",
-    "01-datasets/05-dataset_characterisation/3-synthetic")
+    "2-dataset_characterisation/1-synthetic")
   )[[params$table_format]], 
@@
 add_sfig(
-  figure_file("edge_flip_comparison.svg", experiment_id="02-metric_characterisation/4-edge_flip"), 
+  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 
@@
     "qc_checks.rds",
-    "03-method_characterisation"
+    "4-method_characterisation"
   ))[[params$table_format]], 
@@
 if (params$table_format == "html") 
-  source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/sfigs_stables.R"))
+  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
 write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))

### 2018-08-29 scripts/02-metric_characterisation/02-metric_conformity/README.Rmd
update metric characterisation readmes

--- 2018-08-23 scripts/02-metric_characterisation/README.Rmd
@@
 
-## Metrics characterisation and testing
+# Metric conformity
 
-Good metrics are crucial for an unbiased and comprehensive benchmarking. In this experiment, we test several possible metrics for comparing two trajectories.
+```{r include=FALSE}
+library(dynbenchmark)
+library(tidyverse)
 
-### Metric conformity
+knitr::opts_chunk$set(echo = FALSE)
+```
 
@@
 
-Here, we assess whether metrics conforms these rules by perturbing some datasets in several ways, and assessing how the scores are affected.
+Here, we assess whether metrics conforms such rules empirically:
 
-If not one metric satisfies all the rules, we try to find a combination which conforms to all rules.
-
-### Metric examples
+```{r, results = "asis"}
+extract_scripts_documentation(".", recursive = FALSE) %>% 
+  arrange(ix) %>% 
+    mutate(
+    script = glue::glue("[`{id}`]({file})"),
+    order = ifelse(is.na(ix), "", ix)
+  ) %>%
+  select(order, script, description = title) %>%
+  knitr::kable()
+```

### 2018-08-29 scripts/02-metric_characterisation/README.Rmd
update metric characterisation readmes

@@
 
-## Metrics characterisation and testing
+# Metric characterisation and testing
 
-Good metrics are crucial for an unbiased and comprehensive benchmarking. In this experiment, we test several possible metrics for comparing two trajectories.
+Good metrics are crucial for an unbiased and comprehensive benchmarking. Here we test several possible metrics for comparing two trajectories.
 
-### Metric conformity
+```{r include=FALSE}
+library(dynbenchmark)
+library(tidyverse)
 
-Differences between two datasets should be reflected in certain changes in the metrics. This can be formalised in a set of rules, for example:
+knitr::opts_chunk$set(echo = FALSE)
+```
 
-* If the position of some cells are different than in the gold standard, the score should decrease.
-* If the topology of the network is different than that in the gold standard, the score should not be perfect.
-* The more cells are filtered from the trajectory, the more the score should decrease.
+```{r, results = "asis"}
+knit_child_readme("01-individual_metrics")
+```
 
-Here, we assess whether metrics conforms these rules by perturbing some datasets in several ways, and assessing how the scores are affected.
-
-If not one metric satisfies all the rules, we try to find a combination which conforms to all rules.
-
-### Metric examples
+```{r, results = "asis"}
+knit_child_readme("02-metric_conformity")
+```

### 2018-08-29 scripts/02-metric_characterisation/01-individual_metrics/02-topology.R
add documentation to metric characterisation readme

@@ -1,3 +1,5 @@
+#' Testing the `r label_metric("isomorphic")`, `r label_metric("edge_flip")` and `r label_metric("him")`
+
 library(dynbenchmark)
 library(tidyverse)
 library(furrr)

### 2018-08-29 scripts/02-metrics/02-metric_conformity/README.Rmd
metric characterisation -> metrics

--- 2018-08-29 scripts/02-metric_characterisation/02-metric_conformity/README.Rmd
@@
 ---
-output: github_document
+output: 
+  github_document:
+    pandoc_args: --atx-headers
 ---
@@
 ```{r, results = "asis"}
-extract_scripts_documentation(".", recursive = FALSE) %>% 
-  arrange(ix) %>% 
-    mutate(
-    script = glue::glue("[`{id}`]({file})"),
-    order = ifelse(is.na(ix), "", ix)
-  ) %>%
-  select(order, script, description = title) %>%
-  knitr::kable()
+render_scripts_documentation()
 ```

### 2018-08-29 scripts/02-metrics/01-individual_metrics/02-topology.R

@@ -4,14 +4,14 @@ library(dynbenchmark)
 library(tidyverse)
 library(furrr)
 
-experiment("02-metric_characterisation/01-individual_metrics")
+experiment("02-metrics/01-individual_metrics")
 
 # use all topology based metrics
 metric_ids <- dyneval::metrics %>% filter(category %in% "topology") %>% pull(metric_id)
 
 ##  ............................................................................
 ##  Compare topologies                                                      ####
-source(scripts_file("helper-topologies.R", experiment = "02-metric_characterisation/02-metric_conformity"))
+source(scripts_file("helper-topologies.R", experiment = "02-metrics/02-metric_conformity"))
 
 plan(multiprocess)
 

### 2018-08-29 scripts/02-metrics/02-metric_conformity/helper-create_perturbation_images.R

@@ -3,7 +3,7 @@
 library(dynbenchmark)
 library(esfiji)
 
-experiment("02-metric_characterisation/02-metric_conformity")
+experiment("02-metrics/02-metric_conformity")
 
 svg_location <- result_file("perturbations.svg", experiment_id = "manual_figures")
 

### 2018-08-29 scripts/02-metrics/02-metric_conformity/helper-perturbations.R

@@ -29,7 +29,7 @@ perturb_shuffle_n_cells <- function(dataset, shuffle_n = Inf, seed = NULL) {
 }
 
 perturb_shuffle_cells <- function(dataset, shuffle_perc = 1, seed = NULL) {
-  source(scripts_file("helper-perturbations.R", experiment_id = "02-metric_characterisation/02-metric_conformity"))
+  source(scripts_file("helper-perturbations.R", experiment_id = "02-metrics/02-metric_conformity"))
 
   perturb_shuffle_n_cells(dataset, shuffle_n = length(dataset$cell_ids) * shuffle_perc, seed = seed)
 }
@@ -74,7 +74,7 @@ perturb_shuffle_n_edges <- function(dataset, shuffle_n = Inf) {
 }
 
 perturb_shuffle_edges <- function(dataset, shuffle_perc = 1) {
-  source(scripts_file("helper-perturbations.R", experiment_id = "02-metric_characterisation/02-metric_conformity"))
+  source(scripts_file("helper-perturbations.R", experiment_id = "02-metrics/02-metric_conformity"))
 
   perturb_shuffle_n_edges(dataset, shuffle_n = round(nrow(dataset$milestone_network) * shuffle_perc))
 }
@@ -463,7 +463,7 @@ perturb_shuffle_lengths <- function(dataset) {
 ##  ............................................................................
 ##  Direct topological changes                                              ####
 perturb_change_topology <- function(dataset, topology_id = "linear") {
-  source(scripts_file("helper-topologies.R", experiment_id = "02-metric_characterisation/02-metric_conformity"))
+  source(scripts_file("helper-topologies.R", experiment_id = "02-metrics/02-metric_conformity"))
 
   if (nrow(dataset$milestone_network) != 5) {stop("Can only change topology if there are 5 edges")}
   if (nrow(dataset$divergence_regions)) {stop("To change the topology, dataset cannot have divergence regions")}
@@ -499,7 +499,7 @@ perturb_change_topology <- function(dataset, topology_id = "linear") {
 ##  ............................................................................
 ##  Combined perturbations                                                  ####
 perturb_shuffle_cells_and_add_connecting_edges <- function(dataset, shuffle_perc = 0.2, n_edges = 1) {
-  source(scripts_file("helper-perturbations.R", experiment_id = "02-metric_characterisation/02-metric_conformity"))
[…]

### 2018-08-29 scripts/01-datasets/README.Rmd
add some readme to the datasets
update readmes

@@
 ---
-title: "Datasets"
 output: github_document
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# Dataset processing and characterisation
+
+```{r include=FALSE}
+library(dynbenchmark)
+library(tidyverse)
+
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 
-All datasets can be downloaded from zenodo using the [00-download_from_zenodo.R](00-download_from_zenodo.R) script.
-
-```{r child='01-real/README.Rmd'}
-
+```{r}
+dynbenchmark::render_scripts_documentation()
 ```
 
-```{r child='02-synthetic/README.Rmd'}
+```{r results = 'asis'}
+knit_child_readme("01-real")
+```
 
+```{r results = 'asis'}
+knit_child_readme("02-synthetic")
 ```
+
+```{r results = 'asis'}
+knit_child_readme("05-dataset_characterisation")
+```

### 2018-08-29 scripts/03-methods/README.Rmd
hi methods

---
output: github_document
---

```{r include = FALSE}
library(tidyverse)
library(dynbenchmark)

knitr::opts_chunk$set(echo = FALSE)
```

# Method characterisation

Here we

- Compile all the information we have about TI methods
- Group methods into sets of "tools"
- Download and process the information from the QC spread sheet (https://docs.google.com/spreadsheets/d/1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE)
- Interpret the method quality control

```{r}
dynbenchmark::render_scripts_documentation()
```

### 2018-08-29 scripts/03-methods/01-gather_methods_tools/02-process_quality_control.R

@@ -5,7 +5,7 @@ library(tidyverse)
 library(googlesheets)
 library(dynbenchmark)
 
-experiment("03-method_characterisation")
+experiment("03-methods")
 
 ##  ............................................................................
 ##  QC sheet processing                                                     ####

### 2018-08-29 scripts/03-methods/05-component_characterisation.R

@@ -4,7 +4,7 @@ library(patchwork)
 library(tidytext)
 library(svglite)
 
-experiment("03-method_characterisation")
+experiment("03-methods")
 
 methods <- read_rds(derived_file("methods.rds")) %>%
   filter(type == "algorithm") # only really interested in algorithms

### 2018-08-29 scripts/03-methods/10-create_github_issue.R

@@ -5,7 +5,7 @@ library(tidyverse)
 library(cowplot)
 library(googlesheets)
 
-experiment("03-method_characterisation")
+experiment("03-methods")
 
 
 #   ____________________________________________________________________________

### 2018-08-29 scripts/03-methods/README.Rmd
update readmes

@@
 
-# Method characterisation
+# Trajectory inference methods
 
@@
 - Download and process the information from the QC spread sheet (https://docs.google.com/spreadsheets/d/1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE)
-- Interpret the method quality control
+- Create tables and figures to interpret the methods and quality control
 

### 2018-08-29 scripts/README.Rmd
update readmes

@@
 
-```{r echo = FALSE}
+```{r include = FALSE}
 knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE, echo = FALSE)
@@
 
-## Scripts
+# Scripts
 
@@
 ```{r}
-scripts <- read_rds(derived_file("scripts.rds", experiment_id = "12-varia/01-scripts"))
-
-scripts %>% filter(
-  level == 1
-) %>%
-  arrange(ix) %>%
-  mutate(
-    name = str_glue("[{id}](https://github.com/dynverse/dynbenchmark/tree/master/{location})"),
-    order = ifelse(is.na(ix), "", ix)
-  ) %>%
-  select(order, subfolder = name, description = title) %>%
-  knitr::kable()
+dynbenchmark::render_scripts_documentation()
 ```

### 2018-08-29 manuscript/paper_latex.Rmd

@@
 add_stable(
-  read_rds(figure_file("implementations_table.rds", "03-method_characterisation"))[[params$table_format]],
+  read_rds(figure_file("implementations_table.rds", "03-methods"))[[params$table_format]],
 "implementations",
@@
 add_fig(
-  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "03-method_characterisation"),
+  fig_path = figure_file("implementation_qc_overview.tmp.pdf", experiment_id = "03-methods"),
   ref_id = "implementation_qc_overview", 
@@
 add_sfig(
-  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="02-metric_characterisation/4-edge_flip"), 
+  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="02-metrics/4-edge_flip"), 
   "edge_flip_comparison", 
@@
     "qc_checks.rds",
-    "03-method_characterisation"
+    "03-methods"
   ))[[params$table_format]], 

### 2018-08-29 scripts/01-datasets/01-real/README.Rmd

--- 2018-08-10 scripts/01-datasets/01-real/README.Rmd
@@
 
+# Real datasets
+
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
-
-## Real datasets
 
@@
 
-1.
-The datasets are downloaded from their respective sources (such as GEO) using sets of scripts specific for each dataset in [01-download_from_sources.R](01-download_from_sources.R) (these scripts are located in [helpers-download_from_sources](helpers-download_from_sources))
-2. These datasets are filtered and normalised in [02-filter_and_normalise.R](02-filter_and_normalise.R)
+```{r}
+dynbenchmark::render_scripts_documentation()
+```

### 2018-08-29 scripts/01-datasets/02-synthetic/README.Rmd
02-dataset_characterisation -> 01/datasets/05-dataset_characterisation

--- 2018-08-09 scripts/01-datasets/02-synthetic/README.Rmd
@@
 
+# Synthetic datasets
+
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
 
-## Synthetic datasets
+Each synthetic dataset is based on some characteristics of some real datasets. These characteristics include:
+
+* The number of cells and features
+* The number of features which are differentially expressed in the trajectory
+* Estimates of the distribution of the library sizes, average expression, dropout probabilities, ... estimated by [Splatter](https://github.com/Oshlack/splatter).
+
+Here we estimate the parameters of these "platforms" and use them to simulate datasets using different simulators.
+Each simulation script first creates a design dataframe, which links particular platforms, different topologies, seeds and other parameters specific for a simulator.
+
+The data is then simulated using wrappers around the simulators (see [/package/R/simulators.R](/package/R/simulators.R)), so that they all return datasets in a format consistent with dynwrap.
+
+```{r}
+dynbenchmark::render_scripts_documentation()
+```

### 2018-08-29 scripts/05-scaling/README.Rmd

---
output: github_document
---

```{r include = FALSE}
library(tidyverse)
library(dynbenchmark)

knitr::opts_chunk$set(echo = FALSE)
```

# Scalability of methods

Here we test how well a method scales with increasing number of features (genes) and/or cells.

Each method is run on randomly generated data with increasing gene set and cell set sizes, and the execution times and memory usages are modelled using linear models.

```{r}
dynbenchmark::render_scripts_documentation()
```

### 2018-08-29 scripts/02-metrics/01-metric_characterisation/README.Rmd
add ability to link to results

--- 2018-08-29 scripts/02-metrics/02-metric_conformity/README.Rmd
@@
 
-# Metric conformity
+# Metric characterisation
 
@@
 
-Differences between two datasets should be reflected in certain changes in the metrics. This can be formalised in a set of rules, for example:
+A first characterisation of the metrics. For each metric we:
 
-* If the position of some cells are different than in the gold standard, the score should decrease.
-* If the topology of the network is different than that in the gold standard, the score should not be perfect.
-* The more cells are filtered from the trajectory, the more the score should decrease.
-
-Here, we assess whether metrics conforms such rules empirically:
+- generate some examples to get some intuition on how the metric works
+- test the robustness to a metric to stochasticity or parameters when appropriate
 
@@
 ```
+
+The results of this experiment are available [here](`r link_to_results()`)

### 2018-08-29 package/R/text_document_scripts.R

@@ -105,3 +105,22 @@ knit_child_readme <- function(folder) {
 
   invisible()
 }
+
+
+
+
+#' Generates a link to the results
+#'
+#' @param scripts_folder Folder from which to calculate the experiment id
+#' @export
+link_to_results <- function(scripts_folder = getwd()) {
+  experiment_id <- scripts_folder %>% gsub(".*/scripts/(.*)", "\\1", .)
+  results_dir <- scripts_folder %>% gsub("scripts", "results", .)
+  results_location <- paste0("https://github.com/dynverse/dynbenchmark_results/tree/master/", experiment_id)
+
+  if (fs::is_absolute_path(experiment_id) || !dir.exists(results_dir)) {
+    stop(results_dir, " does not exist! Cannot link to results")
+  }
+
+  results_location
+}

### 2018-08-30 scripts/01-datasets/01-real/README.Rmd
update readmes with github_markdown_nested format

@@
 ---
-output: github_document
+output: dynbenchmark::github_markdown_nested
 ---
@@
 
-The generation of the real datasets is divided in two parts:
+The generation of the real datasets is divided in two parts.
+We first download all the (annotated) expression files from sites such as GEO.
+Next, we filter and normalise all datasets, and wrap them into the common trajectory format of [dynwrap](https://www.github.com/dynverse/dynwrap).
 

### 2018-08-30 scripts/01-datasets/README.Rmd
update readmes with github_markdown_nested format
rerender readmes using knit_nest

@@
 The real datasets are downloaded and preprocessed first, and characteristics from these datasets (such as the number of cells and genes, library sizes, dropout probabilities, ...) are used to generate synthetic datasets.
-Finally, these dataset are uploaded to zenodo.
+The datasets are then characterised, after which they are uploaded to Zenodo.
 
@@
 ```{r results = 'asis'}
-knit_child_readme("01-real")
+knit_nest("01-real")
 ```
@@
 ```{r results = 'asis'}
-knit_child_readme("02-synthetic")
+knit_nest("02-synthetic")
 ```
@@
 ```{r results = 'asis'}
-knit_child_readme("05-dataset_characterisation")
+knit_nest("04-dataset_characterisation")
 ```

### 2018-08-30 scripts/02-metrics/README.Rmd
metric characterisation -> metrics
update readmes with github_markdown_nested format
rerender readmes using knit_nest

--- 2018-08-29 scripts/02-metric_characterisation/README.Rmd
@@
 ---
-output: github_document
+output: dynbenchmark::github_markdown_nested
 ---
 
-# Metric characterisation and testing
+# Metrics for comparing two trajectories
 
@@
 
-```{r include=FALSE}
-library(dynbenchmark)
-library(tidyverse)
-
-knitr::opts_chunk$set(echo = FALSE)
+```{r, results = "asis"}
+knit_nest("01-metric_characterisation")
 ```
@@
 ```{r, results = "asis"}
-knit_child_readme("01-individual_metrics")
+knit_nest("02-metric_conformity")
 ```
-
-```{r, results = "asis"}
-knit_child_readme("02-metric_conformity")
-```

### 2018-08-30 README.Rmd
update readmes with github_markdown_nested format
update readme with guidelines

@@
 ---
-output: github_document
+output: dynbenchmark::github_markdown_nested
 editor_options: 
@@
 
-```{r child='scripts/README.Rmd'}
+```{r results = 'asis'}
+knit_nest("scripts")
 ```
@@
 Explanation coming soon. Feel free to make an issue or send us an e-mail if you want your method to be included.
+
+## Guidelines
+
+Based on the results of the benchmark, we provide context-dependent user guidelines.
+If you want to run methods based on these guidelines, see the [dynorepository](https://github.com/dynverse/dyno).
+For the guidelines itself, see the [dynguidelines ](https://github.com/dynverse/dynguidelines).

### 2018-08-30 README.Rmd
update github readme

@@
 [![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
-![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <img src="package/man/figures/logo.png" align="right" />
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <a href = "package/man/figures/logo.svg"><img src="package/man/figures/logo.png" align="right" /></a>
 
@@
 
-## Structure
+## Experiments
 
-This repository is structured as follows:
+From start to finish, the repository is divided into several experiments, each with their own scripts and results.
+The scripts and results are accompanied by documentation using github READMEs and can be easily explored by going to the appropriate folders:
 
+```{r results = 'asis'}
+extract_scripts_documentation("scripts", recursive = FALSE) %>% 
+  mutate(
+    scripts = location,
+    results = map_chr(scripts, dynbenchmark::link_to_results)
+  ) %>% 
+  mutate(
+    ix = case_when(is.na(ix) ~ "", TRUE ~ as.character(ix)),
+    id = label_long(id),
+    scripts = paste0("[", id, " scripts](", scripts, ")"),
+    results = case_when(is.na(results) ~ "", TRUE ~ paste0("[", id, " results](", results, ")"))
+  ) %>% 
+  select(`\\#` = ix, id, scripts, results) %>% 
+  knitr::kable()
+```
+
+We also have several additional subfolders:
+
+* manuscript: source files for producing the manuscript
+* package: an R package with several helper functions
 * data: raw data files required by the scripts
-* derived: data files produced by the scripts
-* figures: figures created by the scrips
-* manuscript: source files for producing the manuscript
-* package: an R package with helper functions
-* results: data objects outputted by the scripts
-* scripts: code required to reproduce the analyses of the benchmark
+* derived: intermediate data files produced by the scripts
 
@@
 
-```{r results = 'asis'}
-knit_nest("scripts")
-```
-
 ## Benchmarking your own method

### 2018-08-30 scripts/03-methods/README.Rmd
update readmes with github_markdown_nested format
cleanup methods

@@
 ---
-output: github_document
+output: dynbenchmark::github_markdown_nested
 ---
-
-```{r include = FALSE}
-library(tidyverse)
-library(dynbenchmark)
-
-knitr::opts_chunk$set(echo = FALSE)
-```
 
@@
 - Compile all the information we have about TI methods
-- Group methods into sets of "tools"
-- Download and process the information from the QC spread sheet (https://docs.google.com/spreadsheets/d/1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE)
-- Create tables and figures to interpret the methods and quality control
+- Characterise the methods with regards to user and developer friendliness (method quality control)
+- Characterise the methods with regards to prior information, underlying algorithm, possible detectable trajectory types, ...
 
@@
 ```
+
+```{r, results = "asis"}
+dynbenchmark::knit_nest("01-gather_methods_information")
+```

### 2018-08-30 scripts/03-methods/01-gather_methods_information/03-add_quality_control.R

@@ -1,4 +1,4 @@
-## Add QC scores to methods and tools tibble
+#' Add QC scores to methods and tools tibble
 
 library(tidyverse)
 library(dynbenchmark)

### 2018-08-30 scripts/03-methods/01-gather_methods_information/README.Rmd

---
output: dynbenchmark::github_markdown_nested
---

# Gathering all the information we have about the methods

Most information of the methods are contained within their respective containers (see the [dynmethods](https://github.com/dynverse/dynmethods) repository, [https://github.com/dynverse/dynmethods](https://github.com/dynverse/dynmethods)). We gather additional information from our google sheets (https://docs.google.com/spreadsheets/d/1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE), which also contains the quality control for each methods.

```{r}
dynbenchmark::render_scripts_documentation()
```

### 2018-08-31 scripts/02-metrics/01-metric_characterisation/README.Rmd
update readmes with github_markdown_nested format
fix completely sensible error for when dynbenchmark is not imported in readmes

@@
 ---
-output: 
-  github_document:
-    pandoc_args: --atx-headers
+output: dynbenchmark::github_markdown_nested
 ---
@@
 # Metric characterisation
-
-```{r include=FALSE}
-library(dynbenchmark)
-library(tidyverse)
-
-knitr::opts_chunk$set(echo = FALSE)
-```
 
@@
 
-```{r, results = "asis"}
-render_scripts_documentation()
+```{r}
+dynbenchmark::render_scripts_documentation()
 ```
 
-The results of this experiment are available [here](`r link_to_results()`)
+The results of this experiment are available [here](`r dynbenchmark::link_to_results()`)

### 2018-08-31 scripts/02-metrics/01-metric_characterisation/02-topology.R

@@ -1,4 +1,4 @@
-#'  Characterisation of the `r label_metric("isomorphic")`, `r label_metric("edge_flip")` and `r label_metric("him")`
+#'  Characterisation of the `r dynbenchmark::label_metric("isomorphic")`, `r dynbenchmark::label_metric("edge_flip")` and `r dynbenchmark::label_metric("him")`
 
 library(dynbenchmark)
 library(tidyverse)

### 2018-09-13 scripts/02-metrics/02-metric_conformity/README.Rmd
add ability to link to results
update readmes with github_markdown_nested format
update readmes to nested github document

@@
 # Metric conformity
-
-```{r include=FALSE}
-library(dynbenchmark)
-library(tidyverse)
-
-knitr::opts_chunk$set(echo = FALSE)
-```
 
@@
 ```{r, results = "asis"}
-render_scripts_documentation()
+dynbenchmark::render_scripts_documentation()
 ```
+
+The results of this experiment are available [here](`r dynbenchmark::link_to_results()`).

### 2018-09-13 README.Rmd
do not show conflicts

@@
 
-```{r}
+```{r, warning=FALSE, message=FALSE}
 library(tidyverse)
@@
 From start to finish, the repository is divided into several experiments, each with their own scripts and results.
-The scripts and results are accompanied by documentation using github READMEs and can be easily explored by going to the appropriate folders:
+These are accompanied by documentation using github READMEs and can be easily explored by going to the appropriate folders:
 

### 2018-09-17 scripts/04-method_testing/README.Rmd
update readmes with github_markdown_nested format
update readmes to nested github document
update method testing scripts readme

@@
 ---
-output: github_document
+output: dynbenchmark::github_markdown_nested
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# Method testing
+
+Each method is tested on a few small real and toy datasets. Ideally, each method should be able to execute successfully on each of these datasets.
+If a method is not able to infer a trajectory on any of the datasets, it will not be included in any of the next experiments.
+
+Overview experiment:
+```{r}
+dynbenchmark::render_scripts_documentation()
 ```
-
-## Method testing
-
-Here, we test the methods on some small real and toy datasets, to capture errors early on and make sure all methods run fine on these "simple" datasets.
-
-These testing results are in the end pushed to [https://dynverse.github.io/dynmethods_automated_testing/index.html](https://dynverse.github.io/dynmethods_automated_testing/index.html)

### 2018-09-17 scripts/04-method_testing/0-pull_singularity_images.R

@@ -1,4 +1,4 @@
-#' Download singularity images if TI methods on the cluster
+#' Update singularity images on the cluster
 
 library(dynmethods)
 library(dynwrap)
@@ -6,7 +6,6 @@ library(dynbenchmark)
 library(tidyverse)
 
 experiment("singularity_images")
-
 
 # If nothing is running on the cluster, run this on the login node to clear temporary folders:
 # rm -rf /tmp/* /data/*

### 2018-09-17 manuscript/paper.Rmd
update paper for v3, it builds! :confetti:

@@
 title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
-bibliography: references.bib
-csl: nature-biotechnology.csl
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage[table]{xcolor}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{tabu}
-  - \usepackage[normalem]{ulem}
-  - \usepackage{lscape}
-  - \usepackage{pbox}
-  - \usepackage[labelformat=empty]{caption}
-  - \usepackage{hyperref}
-  - \definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
-  - \definecolor{cerulean}{rgb}{0.0, 0.48, 0.65}
-  - \definecolor{burntorange}{rgb}{0.8, 0.33, 0.0}
-  - \hypersetup{colorlinks=true,linkcolor=airforceblue,citecolor=cerulean,urlcolor=burntorange}
-  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
-  - \newcommand{\hideFromPandoc}[1]{#1}
-  - \hideFromPandoc{\let\Begin\begin}
-  - \hideFromPandoc{\let\End\end}
-  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
-mainfont: Open Sans
-
+output: dynbenchmark::pdf_manuscript
 ---
 
-<!--
-Tips for working in this document:
-Go to tools -> preferences and remove Use smart quotes
-Enable automatic zotero syncing to bibtex
-Make sure to pin the zotero bibtex references after first using them!
--->
+```{r setup, include=FALSE}
+library(dynbenchmark)
+library(tidyverse)
 
-```{r setup, include=FALSE}
-source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))
-
-# knitr options
-knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
-  comment = "#>",
-  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
-)
+tools <- read_rds(result_file("tools.rds", "03-methods"))
+methods <- read_rds(result_file("methods.rds", "03-methods"))
 ```
@@
 
+```{r}
+n_tools <- tools %>% filter(type == "algorithm") %>% nrow()
+n_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("algorithm", "algorithm_test")))
+n_control_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("control", "control_test")))
+n_datasets_real <- 10
+n_datasets_synthetic <- 20
+```
 # Abstract
@@
 Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
-Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+Since 2014, at least → `r n_tools` ←tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
 As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
 
-In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness.
+In this study, we evaluated a total of → `r n_methods_evaluated` ←methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
 Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
 Instead, the choice of method depends on the trajectory topology present in the data.
-We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data.
-While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
-We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).
+→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`,← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
 
@@
 <!-- §2: Plethora of new TI methods -->
-A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`).
-For no other single-cell analysis workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017].
-All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
-Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect.
+A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
+→In several repositories listing single-cell tools,  omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
+Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type(s) of graph topologies they can detect.
 Early TI methods typically fixed the topology algorithmically (e.g.
-linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016].
+linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016].
 These methods therefore mainly focused on correctly ordering the cells along the fixed topology.
-Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases.
-Methods that perform topology inference are still in the minority, though current trends suggest this will change soon.
+→Most recent methods also infer the topology computationally, which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←
 
-```{r, results='asis' }
+```{r, eval = FALSE}
 add_stable(
-  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
-"implementations",
-paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::collapse("; "))
+  read_rds(result_file("tools_table.rds", "03-methods"))[[params$table_format]],
+"tools",
+paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::glue_collapse("; "))
 )
@@
 <!-- §4: Our study -->
-In this study, we performed a comprehensive evaluation (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`) of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods.
-The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`).
-In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories.
-The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation.
-Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.
+In this study, we performed a comprehensive evaluation of  →`r n_methods_evaluated`← TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
+The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
+→The evaluated each methods on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) their stability across reruns stability.
+Finally, we combined these metrics to provide a set of practical guidelines for users.
+←
 
-```{r echo=FALSE, results="asis"}
+```{r}
 add_fig(
-  fig_path = figure_file("overview_evaluation_v4.svg", experiment_id="manual_figures"),
+  fig_path = "figures/overview_evaluation_v4.svg",
   ref_id = "ti_evaluation_overview", 
@@
 b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
-c) Trajectories are automatically classified into one of nine trajectory types.
-d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
+c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity.
+d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.",
+   width = 12,
+  height = 8, 
 )
@@
 add_sfig(
-  fig_path = figure_file("overview_evaluation_expanded_v1.svg", "manual_figures"),
+  fig_path = "figures/overview_evaluation_expanded_v1.svg",
   ref_id = "overview_evaluation_expanded",
@@
 The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs).
-We defined a set of metrics for calculating the similarity of two trajectories represented in the common model, each assessing a different aspect of a trajectory: the cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`).
-The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+→We defined several possible metrics to compare a prediction to the reference trajectory, and based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose a set of 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.
+→For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulates single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.←
 
 <!-- Overall evaluation results -->
-Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, SCORPIUS, AGA pseudotime and cellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).
-When focusing on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS and cellTree maptpx returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees.
-This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Angle and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).
-
 ```{r echo=FALSE, results="asis"}
 add_fig(
-  fig_path = figure_file("overview.svg", experiment_id = "9-main_figure"),
+  fig_path = result_file("overview.pdf", experiment_id = "08-summary"),
   ref_id = "results_overview", 
-  caption_main = glue::glue("Overview of the results on the evaluation of {nrow(filter(methods, evaluated, type == 'algorithm'))} TI methods and {nrow(filter(methods, evaluated, type == 'control'))} control methods."),
+  caption_main = glue::glue("Overview of the results on the evaluation of {n_methods_evaluated} TI methods and {n_control_methods_evaluated} control methods."), 
   caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
 b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
-Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, which provide a baseline performance.
-c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+`r {n_control_methods_evaluated}` control methods were also included as an indication of baseline performance c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
 d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
-Also listed are the quality control scores aggregated according to practicality and the different categories."
+Also listed are the quality control scores aggregated according to practicality and the different categories.",
+  width = 15,
+  height = 20
 )
@@
 
-<!-- Errors-->
-Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g.
-Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions.
-Other methods (reCAT, SCOUP, StemID, cellTree Gibbs) exceeded the pre-defined time or memory limits when applied to the largest datasets, and therefore were not scalable to large numbers of cells or genes.
-For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).
-
 <!-- Evaluation robustness -->
-```{r}
-
-# the correlation when subsampling
-correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
-filter(n_tasks == round(max(n_tasks)/10)) %>% 
-          pull(cor) %>% 
-          mean()
-
-# how often is a method top when subsampling
-top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
-filter(method_short_name == "slngsht") %>% 
-filter(n_tasks == round(max(n_tasks)/10)) %>%
-pull(rank_match)
-
-# how many datasets are necessary to get a method on top
-random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
-filter(method_short_name == "wishbone") %>% 
-pull(max_n_tasks_top)
-```
-
-We further investigated the robustness of our evaluation across datasets and dataset sources.
-Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard.
-Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`).
-To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets.
-The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, Slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets.
-On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best  (`r ref("sfig", "robustness_comparison", "c")`).
-For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method.
-Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, as we included in our evaluation, is necessary to robustly compare the overall performance of methods.
-
-```{r}
-add_sfig(
-  figure_file(
-    "robustness.svg",
-    "11-evaluation_robustness"
-  ),
-  "robustness_comparison", 
-  "The robustness of our evaluation.",
-  "a) Comparison of method performance between real and synthetic datasets across trajectory types.
-Given in the top-left corner of each panel is the Pearson's correlation.
-\"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types.
-b) Violin plots of the overall performance on each individual dataset.
-For each method, the arch indicates the first method with which a method performs significantly higher (using a pairwise wilcoxon test corrected for multiple testing; p-value < 0.05).
-c) Estimates of the size of largest subset of datasets on which the method would be the average top performer.
-For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether this method ranks first when aggregating the scores over these datasets.
-This number is therefore only a lower bound, and the actual size of largest subset could be higher."
-)
-```
 
@@
 <!-- Inverse relation between edge flip and correlation -->
-In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`).
-These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
-However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-This indicates that current non-linear methods potentially contain less accurate ordering algorithms.
-A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.
 
 <!-- Topological complexity versus performance -->
-```{r}
 
-# sensitivity of topology prediction
-topology_sensitivity <- read_rds(result_file("topology_sensitivity.rds", "8-compare_topology"))
-```
-
-Despite their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`).
-AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data.
-Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`).
-Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`).
-Inferring the correct topology without parameter tuning is therefore still an open challenge.
-
-```{r echo=FALSE, results="asis"}
-add_fig(
-  fig_path = figure_file("topology_complexity_comparison.svg", experiment_id = "8-compare_topology"),
-  ref_id = "topology_complexity_comparison", 
-  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the gold standard.", 
-  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies.
-A positive difference means that the topology predicted by the method is more complex than the one in the gold standard."
-)
-```
-
-```{r, results='asis' }
-add_stable(
-  read_rds(figure_file("topology_sensitivity_table.rds", "8-compare_topology"))[[params$table_format]],
-"topology_sensitivity",
-"Sensitivity of topology prediction of each TI method.
-Given is how often a method is able to correctly predict the topology of the trajectory present in the dataset, within a particular trajectory type."
-)
-```
-
-<!-- Additional factors which could influence performance -->
-Apart from the trajectory types, we also investigated several other factors that could influence the performance of a method.
-Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a")`).
-When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`), although these tendencies were not significant.
-
-```{r}
-add_stable(
-  read_rds(figure_file(
-    "component_importance.rds",
-    "6-performance_predictors"
-  ))[[params$table_format]], 
-  "component_importance", 
-  "Importance values for algorithmic components.
-The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a p-value using a two-tailed Mann–Whitney U test (corrected for multiple testing using the method of Benjamini-Hochberg) and the increase in node purity importance measure using random forest classification."
-)
-```
+## Method stability and scalability
 
 ## Method quality control
-
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
-To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_checks")`).
-This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.
-
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`).
-While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
-
-```{r echo=FALSE, results="asis"}
-add_fig(
-  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "4-method_characterisation"),
-  ref_id = "implementation_qc_overview", 
-  caption_main = "Overview of the quality control results for every method.", 
-  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_checks')}).
-Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
-Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation.
-This is represented in the plot as the height on the y-axis.
-Top: Average QC score for each method.
-Right: The average score of each quality control item.
-Shown into more detail are those items which had a score lower than 0.5.")
-)
-```
-
-We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")`).
-We could also not find any quality aspect which was significantly associated with method performance.
 
@@
 add_fig(
-  fig_path = figure_file("tree.svg", experiment_id = "7-user_guidelines"),
+  fig_path = ggplot(),#figure_file("tree.svg", experiment_id = "7-user_guidelines"),
   ref_id = "user_guidelines", 
@@
 add_fig(
-  fig_path = figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
+  fig_path = ggplot(),#figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
   ref_id = "example_predictions", 
   caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
-  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process  [@treutlein_dissectingdirectreprogramming_2016].
+  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process.
 A multi-dimensional scaling of dataset is shown at the bottom.
@@
 # Online Methods
+## Data and code availability
+→ 
+The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].
 
+The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("github.com/dynverse/dynverse")`.
+These include packages for wrapping trajectories (`r print_url("github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("github.com/dynverse/dynguidelines")`).
+The main analysis repository is available at `r print_url("github.com/dynverse/dynbenchmark")` and is divided into several experiments.
+In this repository, we provide READMEs for both the scripts and the results, which can be easily browsed and explored on the github website.
+←
 ## Trajectory inference methods
 <!-- Overview -->
-We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("stable", "implementations")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-A continuously updated list can also be found [online](https://goo.gl/FnsTxG)).
-We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynverse/issues).
+We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We welcome any contributions by creating an issue at →`r print_url("github.com/dynverse/dynmethods/issues")`←.
 
 <!-- Inclusion criterion -->
-Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-This resulted in the inclusion of  `r nrow(filter(methods, evaluated, type == "algorithm"))`  methods in the evaluation.
+Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
+This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.
+
 <!-- Components -->
@@
 Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Interestingly, components are frequently shared between different algorithms.
+Components are frequently shared between different algorithms.
 For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.
-
 ### Method input
 As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
+
 <!-- Prior information -->
@@
 For synthetic data: four timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
-
 ### Common trajectory model
 Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap.
-We therefore created wrappers around each method (available at [github.com/dynverse/dynmethods](https://www.github.com/dynverse/dynmethods)) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`).
+We therefore created wrappers around each method (available at `r print_url("https://www.github.com/dynverse/dynmethods")`) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`).
 This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
@@
 ```{r}
-add_sfig(
-  figure_file("overview_wrapping_v1.svg", "manual_figures"),
-  "overview_wrapping", 
-  "A common interface for TI methods.", 
-  "a) The input and output of each TI method is standardised.
-As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
+#add_sfig(
+#  figure_file("overview_wrapping_v1.svg", "manual_figures"),
+#  "overview_wrapping", 
+#  "A common interface for TI methods.", 
+#  "a) The input and output of each TI method is standardised.
+As input, each TI method #receives either raw or normalised counts, several parameters, and a selection of prior information.
 After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model.
@@
 b) The specific transformations performed by each of the wrapper functions is shown in more detail.",
-  15, 
-  7.5
-)
+#  15, 
+#  7.5
+#)
 ```
@@
 ```{r}
-
+methods_evaluated <- read_rds(result_file("methods_evaluated.rds", "03-methods"))
 # create functions to list the methods within a particular conversion category
-list_sentence <- function(x) {
-  if (length(x) == 1) {
-    x
-  } else {
-    paste(head(x, -1), collapse = ", ") %>% paste0(., " and ", tail(x, 1), sep = "")
-  }
-}
 conversion <- function(conversion_oi) {
   methods_evaluated %>% 
-    filter(type %in% c("algorithm", "control")) %>%
-   filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% 
-   mutate(text = paste0(method_name, ifelse(conversion_special, "\\*", ""))) %>%
-   pull(text) %>% 
+    filter(type %in% c("algorithm", "algorithm_test")) %>%
+    filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
+    pull(name) %>% 
     sort() %>%
-   list_sentence()
+    label_vector()
 }
@@
 * **Type 1, direct:** `r conversion("trajectory")`.
-The wrapped method directly returns a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone.
-This usually indicates the transformations required for the method were specific to that method only and thus required a custom implementation.
-* **Type 2, linear pseudotime:** `r conversion("linear")`.
-The method returns a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
-* **Type 3, cyclical pseudotime:** `r conversion("cyclic")`.
-The method returns a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones.
-* **Type 4, end state probability:** `r conversion("end_state_probability")`.
-The method returns a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state.
-This is translated into a star-shaped milestone network, with regions of delayed commitment between all transitions.
+The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone.
+This usually indicates that additional transformations were required for the method not covered by any of the following output formats.
+* **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`.
+The method returned a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
+* **Type 3, cyclical pseudotime:** `r conversion("cyclic_trajectory")`.
+The method returned a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones.
+→These milestones were positioned at pseudotime 0, ⅓ and ⅔ .←
+* **Type 4, end state probability:** `r conversion("end_state_probabilities")`.
+The method returned a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state.
+→This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$) with regions of delayed commitment between all milestones.
+The milestone percentage of a cell to one of the outer milestones is equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone is equal to $1-\textrm{pseudotime}$ .←
 * **Type 5, cluster assignment:** `r conversion("cluster_graph")`.
-The method returns a milestone network, and an assignment of each cell to a specific milestone.
-Cells are positioned onto the milestones they are assigned to.
+The method returned a milestone network, and an assignment of each cell to a specific milestone.
+Cells were positioned onto the milestones they are assigned to.
 * **Type 6, project onto nearest branch:** `r conversion("dimred_projection")`.
-The method returns a milestone network, and a dimensionality reduction of the cells and milestones.
-The cells are projected onto the closest nearest segment, thus determining the cells position along the milestone network.
-* **Type 7, cell graph:** `r conversion("cell_graph")`. The method returns a network of cells, 
+The method returned a milestone network, and a dimensionality reduction of the cells and milestones.
+The cells were projected onto the closest nearest segment, thus determining the cells position along the milestone network.
+→If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell.
+For these methods, we usually wrote two wrappers, one which included the projection, and one without.←
+* →**Type 7, branch trajectory:** `r conversion("branch_trajectory")`.
+The method returned an assignment of cells to a particular branch, a pseudotime for each cell within the branch, and a network connecting these different branch.
+We converted this branch network to a milestone network by calculating the line graph of the network.
+The cells were then placed between each two milestones using the provided pseudotime.
+←
+* **Type 8, cell graph:** `r conversion("cell_graph")`. The method returned a network of cells, 
 and which cell-cell transitions are part of the 'backbone' structure.
-Backbone cells with degree $\neq$ 2 are regarded as milestones, and all other cells are placed on transitions between the milestones.
+Backbone cells with degree $\neq$ 2 were regarded as milestones, and all other cells were placed on transitions between the milestones.
+→If methods did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones.
+Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells, and then divided the distance of each pair of cells with the total distance to get the milestone percentages.←
 
@@
 The dynmethods github repository contains one vignette for each of the methods, discussing how to use the package to apply a method to a dataset, the specific implementation of the wrapper, and how the quality control scores of each TI method were derived.
+→
+### Control methods
+We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the gold standard, and **shuffle**, which will place the cells randomly on the gold standard milestone network.
+### Off-the-shelf methods
+We added several additional TI methods which can be run using just a couple of commands in R. These "off-the-shelf" methods were used to provide baseline performance.←
 
-### Control methods
-To define baseline performance, we added three control methods, namely "Random", "Component 1", and "Angle".
-
-* **Random:** A random network with 50 milestones according to the Barabási–Albert model is generated, and cells are placed randomly along the network.
-This method provides a bottom line performance; each other TI method should be able to obtain a higher benchmark score than this control method.
-* **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
+* **Component 1:** This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
 Although conceptually very simple, it has been used in certain studies [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
 * **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
-
+→
+* **Periodic PrinCurve:** Periodic principal curves, using the R princurve package
+* **MST:** Dimensionality reduction with PCA, followed by clustering using the R mclust package, after which the clusters are connected using a minimal spanning tree.
+The cells are then projected onto the edges of the MST.←
 ## Real datasets
-We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)).
+We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
+The scripts to download and process these datasets are available on our repository ([https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets](https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets)).
 Whenever possible, we preferred to start from the raw counts data.
 These raw counts were all normalised and filtered using a common pipeline, discussed later.
-We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample; gold standard) (`r ref("stable", "datasets")`).
-These datasets are deposited in Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].
+
+→For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network).
+The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample or the time it was sampled) or silver standard otherwise (usually by clustering the expression values).
+To connect these cell groups, we used the original study to determine the network which the authors proposed or otherwise found to be the most likely.
+In the end, each group of cells was placed on a milestone, having percentage = 1 for that particular milestone.
+The known connections between these groups were used to construct the milestone network.
+If there was time data available, we used this as the length of the edge, otherwise we set all the lengths equal to one.
+←
 
 ```{r}
-add_stable(
-  read_rds(figure_file("datasets.rds","2-dataset_characterisation/2-real"))[[params$table_format]], 
-  "datasets", 
-  "Real datasets used in this study."
-)
+#add_stable(
+#  read_rds(
+#     figure_file("datasets.rds","2-dataset_characterisation/2-real")
+#  )[[params$table_format]], 
+#  "datasets", 
+#  "Real datasets used in this study."
+#)
 ```
+## Synthetic datasets
+→
+To generate synthetic datasets, we used four different synthetic data simulators:
 
-## Synthetic datasets
+* **dyngen:** Simulations of gene regulatory networks, available at `r print_url("https://github.com/dynverse/dyngen")`
+* **dyntoy:** Random gradients of expression in the reduced space, available at `r print_url("https://github.com/dynverse/dyntoy")`
+* **PROSSTT:** Expression is sampled from a linear model which depends on pseudotime [@papadopoulosPROSSTTProbabilisticSimulation2018]
+* **Splatter:** Simulations of non-linear paths between different expression states
 
-Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
-At every step, we took great care to mimic real cellular regulatory networks as good as possible, while keeping the model simple and easily extendable.
+For every simulator, we took great care with trying to make the datasets look like real data as  best as possible.
+To do this, we extracted several parameters from all real datasets.
+We calculated the number of differentially expressed features within a trajectory using a wilcoxon test between every pair of cell groups.
+We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package [@zappia_splattersimulationsinglecell_2017].
+These parameters were then given to the simulators when applicable, as described below.
+←
+### dyngen
+
+The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
 For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
-
-### Network generation
+#### Network generation
 
@@
 
-To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
-Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple undirected graphs (bifurcating loop and bifurcating convergence).
+→To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
+Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating converging), (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).←
 
@@
 
-### Simulation of gene regulatory systems using thermodynamic models
+#### Simulation of gene regulatory systems using thermodynamic models
 
@@
 ```{r}
-add_stable(
-  read_rds(figure_file(
-    "samplers.rds",
-    "2-dataset_characterisation/1-synthetic")
-  )[[params$table_format]], 
-  "samplers", 
-  "Distributions from which each parameter in the thermodynamic model was sampled."
-)
+#add_stable(
+#  read_rds(figure_file(
+#    "samplers.rds",
+#    "2-dataset_characterisation/1-synthetic")
+#  )[[params$table_format]], 
+#  "samplers", 
+#  "Distributions from which each parameter in the thermodynamic model was sampled."
+#)
 ```
@@
 
-### Simulation of the single-cell RNA-seq experiment
+#### Simulation of the single-cell RNA-seq experiment
 
@@
 Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
+→Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.←
 
-### Gold standard extraction
+#### Gold standard extraction
 
@@
 
-## Expression normalisation pipeline
+## Dataset filtering and normalisation
 We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016].
@@
 As a final filter, we removed both zero genes and cells until convergence.
-
 ## Evaluation metrics
@@
 add_sfig(
-  figure_file("aspects_v2.svg", experiment_id="manual_figures/metrics"), 
+  ggplot(),#figure_file("aspects_v2.svg", experiment_id="manual_figures/metrics"), 
   "score_aspects", 
@@
 ```
-
 ### Correlation between geodesic distances
@@
 add_sfig(
-  figure_file("geodesic.svg", experiment_id="manual_figures/metrics"), 
+  ggplot(),#figure_file("geodesic.svg", experiment_id="manual_figures/metrics"), 
   "metric_geodesic", 
@@
 We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard.
-We used a Random Forest regression, implemented in the *ranger* package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of every cell in the gold standard, using the milestone percentages of these cells in the prediction as features.
+We used a Random Forest regression, implemented in the *ranger* package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of each cell in the gold standard, using the milestone percentages of this cells in the prediction as features.
 We then used the out-of-bag mean-squared error on these percentages to score each method's capability of predicting the correct neighbourhood of each cell.
@@
 add_sfig(
-  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
+  ggplot(),#figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
   "edge_flip_comparison", 
@@
 ```
-
 ### Aggregation of scores
@@
 add_sfig(
-  figure_file("overview_aggregation_v1.svg", experiment_id="manual_figures/metrics"), 
+  ggplot(),#figure_file("overview_aggregation_v1.svg", experiment_id="manual_figures/metrics"), 
   "aggregation_of_results", 
@@
 ```
-
 ## Benchmark
-
 ### Method execution
@@
 Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.
-
 ### Effect of prior information
@@
 P-values were controlled for multiple testing using Benjamini-Hochberg correction.
-
 ### Effect of algorithm components
@@
 P-values were controlled for multiple testing using Benjamini-Hochberg correction.
-
 ## Method quality control
@@
 ```{r}
-add_stable(
-  read_rds(figure_file(
-    "qc_checks.rds",
-    "4-method_characterisation"
-  ))[[params$table_format]], 
-  "qc_checks", 
-  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
-)
+#add_stable(
+#  read_rds(figure_file(
+#    "qc_checks.rds",
+#    "4-method_characterisation"
+#  ))[[params$table_format]], 
+#  "qc_checks", 
+#  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
+#)
 ```
@@
 For calculating the final score, we weighed each of the six categories equally.
-
 ## Trajectory types
@@
 
-## Data and code availability
-The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].
-The code to fully reproduce the results from this study has been split into several R packages and is available on github ([github.com/dynverse/dynverse](https://github.com/dynverse/dynverse)).
+```{r "supplementary", echo=FALSE, results='asis'}
+#if (params$table_format == "html") 
+#  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
+#write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
+#write_rds(figs, derived_file("figs.rds", "paper"))
+```
 
-```{r "supplementary", echo=FALSE, results='asis'}
-if (params$table_format == "html") 
-  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
-write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
-write_rds(figs, derived_file("figs.rds", "paper"))
-```
+<!--
+Tips for working in this document:
+Go to tools -> preferences and remove Use smart quotes
+Enable automatic zotero syncing to bibtex
+Make sure to pin the zotero bibtex references after first using them!
+-->

### 2018-09-17 README.Rmd
update readme

@@
 
-Based on the results of the benchmark, we provide context-dependent user guidelines.
-If you want to run methods based on these guidelines, see the [dynorepository](https://github.com/dynverse/dyno).
-For the guidelines itself, see the [dynguidelines ](https://github.com/dynverse/dynguidelines).
+Based on the results of the benchmark, we provide context-dependent user guidelines, [available as a shiny app]https://github.com/dynverse/dynguidelines).
+This app is integrated within the a whole pipeline [dyno repository](https://github.com/dynverse/dyno), which also includes the wrappers used in the benchmarking and other packages for visualising and interpreting the results.

### 2018-09-17 README.Rmd
updated readme

--- 2018-09-13 README.Rmd
@@
 
-A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
-Wouter Saelens\*, Robrecht Cannoodt\*, Helena Todorov, Yvan Saeys  
-bioRxiv 276907; doi: https://doi.org/10.1101/276907
+> A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
+<strong> Wouter Saelens\* </strong> <a href='https://orcid.org/0000-0002-7114-6248'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/orcid_logo.svg' height='16'></a> <a href='https://github.com/zouter'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
+<strong> Robrecht Cannoodt\* </strong> <a href='https://orcid.org/0000-0003-3641-729X'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/orcid_logo.svg' height='16'></a> <a href='https://github.com/rcannood'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
+Helena Todorov <a href='https://github.com/Helena-todd'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
+<em> Yvan Saeys </em> <a href='https://github.com/saeyslab'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>  
+[bioRxiv:276907](https://www.biorxiv.org/content/early/2018/03/05/276907) [doi:10.1101/276907](https://doi.org/10.1101/276907)
 
@@
 
-Based on the results of the benchmark, we provide context-dependent user guidelines.
-If you want to run methods based on these guidelines, see the [dynorepository](https://github.com/dynverse/dyno).
-For the guidelines itself, see the [dynguidelines ](https://github.com/dynverse/dynguidelines).
+Based on the results of the benchmark, we provide context-dependent user guidelines, [available as a shiny app](https://github.com/dynverse/dynguidelines).
+This app is integrated within the [dyno pipeline](https://github.com/dynverse/dyno), which also includes the wrappers used in the benchmarking and other packages for visualising and interpreting the results.
+
+![dynguidelines](https://github.com/dynverse/dynguidelines/raw/master/man/figures/demo.gif)

### 2018-09-17 README.Rmd
cosmetics ~

--- 2018-09-13 README.Rmd
@@
 
-A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
-Wouter Saelens\*, Robrecht Cannoodt\*, Helena Todorov, Yvan Saeys  
-bioRxiv 276907; doi: https://doi.org/10.1101/276907
+> A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
+<strong> Wouter Saelens\* </strong> <a href='https://orcid.org/0000-0002-7114-6248'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/orcid_logo.svg?sanitize=true' height='16'></a> <a href='https://github.com/zouter'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
+<strong> Robrecht Cannoodt\* </strong> <a href='https://orcid.org/0000-0003-3641-729X'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/orcid_logo.svg?sanitize=true' height='16'></a> <a href='https://github.com/rcannood'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
+Helena Todorov <a href='https://github.com/Helena-todd'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
+<em> Yvan Saeys </em> <a href='https://github.com/saeyslab'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>  
+[bioRxiv:276907](https://www.biorxiv.org/content/early/2018/03/05/276907) [doi:10.1101/276907](https://doi.org/10.1101/276907)
 
@@
 
-Based on the results of the benchmark, we provide context-dependent user guidelines.
-If you want to run methods based on these guidelines, see the [dynorepository](https://github.com/dynverse/dyno).
-For the guidelines itself, see the [dynguidelines ](https://github.com/dynverse/dynguidelines).
+Based on the results of the benchmark, we provide context-dependent user guidelines, [available as a shiny app](https://github.com/dynverse/dynguidelines).
+This app is integrated within the [dyno pipeline](https://github.com/dynverse/dyno), which also includes the wrappers used in the benchmarking and other packages for visualising and interpreting the results.
+
+[![dynguidelines](https://github.com/dynverse/dynguidelines/raw/master/man/figures/demo.gif)](https://github.com/dynverse/dynguidelines)

### 2018-09-18 README.Rmd
update readme

@@
 [![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
-![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <a href = "package/man/figures/logo.svg"><img src="package/man/figures/logo.png" align="right" /></a>
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <a href = "package/man/figures/logo.svg"><img src="package/man/figures/logo.png" align="right" /></a> [![doi](https://zenodo.org/badge/doi/10.1101/276907.svg)](https://doi.org/10.1101/276907)
 

### 2018-09-18 manuscript/paper.Rmd

@@
 Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
-Since 2014, at least → `r n_tools` ←tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+Since 2014, at least →`r n_tools`← tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
 As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
 
-In this study, we evaluated a total of → `r n_methods_evaluated` ←methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
+In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
 Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→In several repositories listing single-cell tools,  omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
+→In several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
 Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type(s) of graph topologies they can detect.
@@
 ### Off-the-shelf methods
-We added several additional TI methods which can be run using just a couple of commands in R. These "off-the-shelf" methods were used to provide baseline performance.←
+For baseline performance, we added several "off-the-shelf" TI methods which can be run using just a couple of commands in R.←
 
@@
 * **PROSSTT:** Expression is sampled from a linear model which depends on pseudotime [@papadopoulosPROSSTTProbabilisticSimulation2018]
-* **Splatter:** Simulations of non-linear paths between different expression states
+* **Splatter:** Simulations of non-linear paths between different expression states [@zappia_splattersimulationsinglecell_2017]
 
@@
 For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
+
 #### Network generation
@@
 
+### dyntoy
+
+### PROSSTT
+
+### Splatter
 ## Dataset filtering and normalisation
@@
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`).
+Furthermore, We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`).
 Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
@@
 
-Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("table", "implementations")`).
+Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("stable", "qc_checks")`).
 This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface.
-Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("table", "implementations")`).
+Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_checks")`).
 For calculating the final score, we weighed each of the six categories equally.
+→
+## Scalability
+←
 ## Trajectory types

### 2018-09-18 scripts/01-datasets/README.Rmd
update readmes to nested github document

@@
 library(tidyverse)
-
-knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 ```{r results = 'asis'}
-knit_nest("01-real")
+dynbenchmark::knit_nest("01-real")
 ```
@@
 ```{r results = 'asis'}
-knit_nest("02-synthetic")
+dynbenchmark::knit_nest("02-synthetic")
 ```
@@
 ```{r results = 'asis'}
-knit_nest("04-dataset_characterisation")
+dynbenchmark::knit_nest("04-dataset_characterisation")
 ```

### 2018-09-18 README.Rmd
update the readme

@@
 From start to finish, the repository is divided into several experiments, each with their own scripts and results.
-These are accompanied by documentation using github READMEs and can be easily explored by going to the appropriate folders:
+These are accompanied by documentation using github readmes and can thus be easily explored by going to the appropriate folders:
 
@@
 
-* manuscript: source files for producing the manuscript
-* package: an R package with several helper functions
-* data: raw data files required by the scripts
-* derived: intermediate data files produced by the scripts
+* [Manuscript](manuscript): Source files for producing the manuscript.
+* [Package](package): An R package with several helper functions.
+* [Data](data): Raw data files required by the scripts.
+* [Derived](derived): Intermediate data files produced by the scripts. These files or not git committed.
+
+## Guidelines
+
+Based on the results of the benchmark, we provide context-dependent user guidelines, [available as a shiny app](https://github.com/dynverse/dynguidelines).
+This app is integrated within the [dyno pipeline](https://github.com/dynverse/dyno), which also includes the wrappers used in the benchmarking and other packages for visualising and interpreting the results.
+
+[![dynguidelines](https://github.com/dynverse/dynguidelines/raw/master/man/figures/demo.gif)](https://github.com/dynverse/dynguidelines)
 
@@
 
-* **The performance of methods** used in [dynguidelines](https://www.github.com/dynverse/dynguidelines). _Not yet available_
+* **The performance of methods** used for the [results overview figure](`r dynbenchmark::link_to_results("scripts/08-summary/overview.pdf")`) and the [dynguidelines](https://www.github.com/dynverse/dynguidelines) app.
+_Not yet available_, but can be accessed in the results folder.
 
@@
 Explanation coming soon. Feel free to make an issue or send us an e-mail if you want your method to be included.
-
-## Guidelines
-
-Based on the results of the benchmark, we provide context-dependent user guidelines, [available as a shiny app](https://github.com/dynverse/dynguidelines).
-This app is integrated within the [dyno pipeline](https://github.com/dynverse/dyno), which also includes the wrappers used in the benchmarking and other packages for visualising and interpreting the results.
-
-[![dynguidelines](https://github.com/dynverse/dynguidelines/raw/master/man/figures/demo.gif)](https://github.com/dynverse/dynguidelines)

### 2018-09-18 manuscript/paper.Rmd
update paper

@@
 ⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
-
 ```{r}
@@
 The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs).
-→We defined several possible metrics to compare a prediction to the reference trajectory, and based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose a set of 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+→We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`).
+Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-→For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulates single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.←
+→For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011].
+We paired each simulated datasets with a real dataset to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←
 
@@
 ## Method stability and scalability
-
 ## Method quality control
@@
 
+________________
 # Online Methods
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.
+The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")` This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.
 
@@
 For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.
+
+→
+### Method wrappers
+Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`).
+These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
+For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors.
+This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below).
+To wrap each method, we based ourselves on example scripts or tutorials which are mentioned in the comments of each wrapper script.
+We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).
+
+We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control.
+These discussion can be viewed on github (`r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`).
+About ⅓ of authors responded, and we improved the wrappers based on their feedback.
+←
 ### Method input
 As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
-
 <!-- Prior information -->
@@
 a start cell) in order to be executable.
-Other methods allow to exploit certain prior information, though this is not required.
+Other methods optionally allow to exploit certain prior information, though this is not required.
 Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
@@
 * **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
-* **Time course** For each cell a time point from which it was sampled.
-If available, directly extracted from the gold standard.
-For synthetic data: four timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
+→
+* **Discrete time course** For each cell a time point from which it was sampled.
+If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used.
+For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
+* **Continuous time course** For each cell a time point from which it was sampled.
+For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
+←
 ### Common trajectory model
-Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap.
-We therefore created wrappers around each method (available at `r print_url("https://www.github.com/dynverse/dynmethods")`) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`).
-This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
+Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps.
+→We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`).
+This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
 (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one.
@@
 Regions must be explicitly defined in the trajectory model.
-Per region, one milestone must be directly connected to all other milestones in the network.
+Per region, one milestone must be directly connected to all other milestones →of the region.←
 
@@
 methods_evaluated <- read_rds(result_file("methods_evaluated.rds", "03-methods"))
+special_methods <- c("mfa", "dpt", "slice", "sincell", "calista", "urd")
 # create functions to list the methods within a particular conversion category
@@
     filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
+    mutate(name = paste0(name, ifelse(id %in% special_methods, "\\*", ""))) %>% 
     pull(name) %>% 
@@
 The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone.
-This usually indicates that additional transformations were required for the method not covered by any of the following output formats.
+→In some cases, this← indicates that additional transformations were required for the method not covered by any of the following output formats.
 * **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`.
@@
 →This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$) with regions of delayed commitment between all milestones.
-The milestone percentage of a cell to one of the outer milestones is equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone is equal to $1-\textrm{pseudotime}$ .←
+The milestone percentage of a cell to one of the outer milestones was equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .←
 * **Type 5, cluster assignment:** `r conversion("cluster_graph")`.
@@
 For these methods, we usually wrote two wrappers, one which included the projection, and one without.←
-* →**Type 7, branch trajectory:** `r conversion("branch_trajectory")`.
+* →**Type 7, branch trajectory[a]:** `r conversion("branch_trajectory")`.
 The method returned an assignment of cells to a particular branch, a pseudotime for each cell within the branch, and a network connecting these different branch.
@@
 * **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
-This was then processed as described above.
+→This was then processed as a type 1, direct,  method.←
 * **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
-This was then processed as described above.
-* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
-* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.
+→This was then processed as a type 8, cell graph, method.←
+* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. →This was then processed as a type 1, direct, method.←
+* **mfa:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed →as a type 4, end state probabilities, method.←
+→
+* **CALISTA:** We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest.
+The cluster probabilities of the two selected milestones were then used as milestone percentages.
+This was then processed as a type 1, direct, method.
+* **URD:** We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.
 
-The dynmethods github repository contains one vignette for each of the methods, discussing how to use the package to apply a method to a dataset, the specific implementation of the wrapper, and how the quality control scores of each TI method were derived.
+More information on how each method was wrapped can be found within the comments of each wrapper script.
+←
+
 →
@@
 Although conceptually very simple, it has been used in certain studies [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
-* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
+* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA, and uses this angle as pseudotime for a cyclical trajectory.
 →
@@
 * **MST:** Dimensionality reduction with PCA, followed by clustering using the R mclust package, after which the clusters are connected using a minimal spanning tree.
-The cells are then projected onto the edges of the MST.←
+The cells are then projected onto the edges of the MST.
+←
+
 ## Real datasets
@@
 For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
-
 #### Network generation
@@
 
+→
 ### dyntoy
@@
 ### Splatter
+←
 ## Dataset filtering and normalisation
@@
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-Furthermore, We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`).
+→Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
+We therefore defined several possible metrics for comparing trajectories, each investigating different layers.
+These are all discussed in `r ref("snote", "metrics")` along with examples and robustness analyses when appropriate.
+
+Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (`r ref("snote", "metrics")`).
+Based on this analysis, we chose 4 metrics for ←
+
+  We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`).
 Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
@@
 -->
+
+[a]Gotta update the figure to reflect this

### 2018-09-18 manuscript/paper.Rmd
update paper

@@
 <!-- A common trajectory analysis framework -->
-Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018].
+Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017] or module detection [@saelens_comprehensiveevaluationmodule_2018].
 Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`).
@@
 Components are frequently shared between different algorithms.
-For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.
+For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.
 
@@
 These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
-For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors.
+For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
 This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below).
-To wrap each method, we based ourselves on example scripts or tutorials which are mentioned in the comments of each wrapper script.
 We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).
@@
 We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control.
-These discussion can be viewed on github (`r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`).
 About ⅓ of authors responded, and we improved the wrappers based on their feedback.
+These discussion can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`.
 ←

### 2018-09-18 README.Rmd
update readme

@@
 
+```{r}
+get_altmetric_badge <- function() {
+  curl::curl_fetch_memory("https://api.altmetric.com/v1/doi/10.1101/276907")$content %>%
+    rawToChar() %>% 
+    jsonlite::fromJSON() %>% 
+    purrr::pluck("images") %>%
+    purrr::pluck("medium") %>% 
+    paste0("&style=bar")
+}
+```
+
 > A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
@@
 <em> Yvan Saeys </em> <a href='https://github.com/saeyslab'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>  
-[bioRxiv:276907](https://www.biorxiv.org/content/early/2018/03/05/276907) [doi:10.1101/276907](https://doi.org/10.1101/276907)
+[bioRxiv:276907](https://www.biorxiv.org/content/early/2018/03/05/276907) [doi:10.1101/276907](https://doi.org/10.1101/276907) [![altmetric](`r get_altmetric_badge()`)](https://altmetric.com/details/33972849)
 

### 2018-09-18 manuscript/paper.Rmd
update paper

@@
 Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
-Instead, the choice of method depends on the trajectory topology present in the data.
-→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`,← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+Instead, the choice of method →mostly← depends on the trajectory topology present in the data.
+→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")` This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.
+→The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")`.
+In the end, we included  `r n_tools` tools in the evaluation, constituting `r n_methods_evaluated` methods.←
 
@@
 For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.
-
 →
@@
 For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
-This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below).
+This script reads in the input data (as described below), runs the method, and outputs the files required to construct a trajectory (as described below).
 We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).
@@
 About ⅓ of authors responded, and we improved the wrappers based on their feedback.
-These discussion can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`.
+These discussions can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`.
 ←
 ### Method input
-As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
+As input, we provided each method with either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the method documentation or from the study describing the method.
 <!-- Prior information -->
@@
 a start cell) in order to be executable.
-Other methods optionally allow to exploit certain prior information, though this is not required.
+Other methods optionally allow to exploit certain prior information.
 Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states.
@@
 This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
-(ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one.
-(iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones.
-Regions must be explicitly defined in the trajectory model.
-Per region, one milestone must be directly connected to all other milestones →of the region.←
+(ii) The milestone percentages contains, for each cell, its position between milestones, and sums for each cell to one.
+(iii) →The regions of delayed commitment, which are connection between three or more milestones.
+These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←
 
@@
 ←
-
 →
@@
 We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets are available on our repository ([https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets](https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets)).
+The scripts to download and process these datasets are available on our repository (`r print_url("https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets")`).
 Whenever possible, we preferred to start from the raw counts data.
@@
 The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample or the time it was sampled) or silver standard otherwise (usually by clustering the expression values).
-To connect these cell groups, we used the original study to determine the network which the authors proposed or otherwise found to be the most likely.
+To connect these cell groups, we used the original study to determine the network which the authors validated or otherwise found to be the most likely.
 In the end, each group of cells was placed on a milestone, having percentage = 1 for that particular milestone.
@@
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
-For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
+→We simulated a total of ←
 #### Network generation
@@
 Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
-To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, as a backbone.
-We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
-We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
+→We therefore first constructed a backbone expression profile for each branch within the trajectory.
+To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.
+We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
+← We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
 In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
@@
 For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork.
-In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).
+In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →the exceptions being being DPT (limited to bifurcating trajectories), and STEMNET and FateID (unable to return linear trajectories).←
 

### 2018-09-19 manuscript/paper.Rmd
update figure 1 and paper

@@
 
-*`r format(Sys.time(), '%d %B, %Y')`*
-
 Wouter Saelens* ¹ ², Robrecht Cannoodt* ¹ ² ³, Helena Todorov¹ ² ⁴, Yvan Saeys¹ ²
@@
 ⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
+
+*`r format(Sys.time(), '%d %B, %Y')`*
+
+`dynverse/dynbenchmark/commit/`r git2r::revparse_single(".","HEAD")$sha``
 ```{r}
-n_tools <- tools %>% filter(type == "algorithm") %>% nrow()
-n_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("algorithm", "algorithm_test")))
-n_control_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("control", "control_test")))
-n_datasets_real <- 10
-n_datasets_synthetic <- 20
+n_tools <- nrow(tools)
+n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("tool", "adaptation", "offtheshelf")))
+n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
+n_datasets_real <- nrow(list_datasets("real"))
+n_datasets_synthetic <- nrow(list_datasets("synthetic"))
 ```
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→In several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
+→Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
 Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type(s) of graph topologies they can detect.
@@
 The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
-→The evaluated each methods on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) their stability across reruns stability.
+→We evaluated each methods on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) the stability of the predictions on the same dataset.
 Finally, we combined these metrics to provide a set of practical guidelines for users.
@@
 add_fig(
-  fig_path = "figures/overview_evaluation_v4.svg",
+  fig_path = raw_file("overview_evaluation_v6.svg", ""),
   ref_id = "ti_evaluation_overview", 
@@
 c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity.
-d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.",
-   width = 12,
+d) →We defined four metrics each assessing the similarity of a different aspect of the trajectory: the topology, the branch assignment, the cell positions and the important features.←",
+   width = 13,
   height = 8, 
@@
 add_sfig(
-  fig_path = "figures/overview_evaluation_expanded_v1.svg",
+  fig_path = raw_file("overview_evaluation_expanded_v1.svg", ""),
   ref_id = "overview_evaluation_expanded",
@@
 <!-- The context -->
-In this study, we presented a large scale evaluation of the performance and code of  `r nrow(filter(methods, evaluated, type == "algorithm"))`  TI methods.
+In this study, we presented a large scale evaluation of the performance and code of  `r n_methods_evaluated`  TI methods.
 By using a common structure and three metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types.
@@
   methods_evaluated %>% 
-    filter(type %in% c("algorithm", "algorithm_test")) %>%
+    filter(source %in% c("tool", "offtheshelf", "adaptation")) %>%
     filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
@@
 The method returned a milestone network, and an assignment of each cell to a specific milestone.
-Cells were positioned onto the milestones they are assigned to.
+Cells were positioned onto the milestones they are assigned to→, with milestone percentage equal to 1.←
 * **Type 6, project onto nearest branch:** `r conversion("dimred_projection")`.
@@
 
+```{r}
+design_dyngen <- read_rds(result_file("design_dyngen.rds", "01-datasets/02-synthetic"))
+```
+
 The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
-→We simulated a total of ←
+→We simulated a total of `r nrow(design_dyngen)`, with `r length(unique(design_dyngen$modulenet_name))` different topologies.←
 #### Network generation
@@
 For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in.
-Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
+→These cells were sampled uniformly across the different steps of the 32 simulations.
+← Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
 We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$.
@@
 add_sfig(
-  ggplot(),#figure_file("overview_aggregation_v1.svg", experiment_id="manual_figures/metrics"), 
+  raw_file("overview_aggregation_v1.svg", "08-summary"), 
   "aggregation_of_results", 

### 2018-09-19 manuscript/paper.Rmd
Add PROSSTT, dyntoy and splatter discussion to methods section.

@@
 Instead, the choice of method →mostly← depends on the trajectory topology present in the data.
-→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://dynverse.duckdns.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
@@
 →In some cases, this← indicates that additional transformations were required for the method not covered by any of the following output formats.
+→Some methods (`r conversion("branch_trajectory")`) returned a branch network instead of a milestone network, and this conversion was done by calculating the line graph of the branch network.←
 * **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`.
@@
 The method returned a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state.
-→This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$) with regions of delayed commitment between all milestones.
+→This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$), with regions of delayed commitment between all milestones.
 The milestone percentage of a cell to one of the outer milestones was equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .←
@@
 For these methods, we usually wrote two wrappers, one which included the projection, and one without.←
-* →**Type 7, branch trajectory[a]:** `r conversion("branch_trajectory")`.
-The method returned an assignment of cells to a particular branch, a pseudotime for each cell within the branch, and a network connecting these different branch.
-We converted this branch network to a milestone network by calculating the line graph of the network.
-The cells were then placed between each two milestones using the provided pseudotime.
-←
-* **Type 8, cell graph:** `r conversion("cell_graph")`. The method returned a network of cells, 
+* **Type 7, cell graph:** `r conversion("cell_graph")`. The method returned a network of cells, 
 and which cell-cell transitions are part of the 'backbone' structure.
@@
 * **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
-→This was then processed as a type 8, cell graph, method.←
+→This was then processed as a type 7, cell graph, method.←
 * **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. →This was then processed as a type 1, direct, method.←
@@
 
-More information on how each method was wrapped can be found within the comments of each wrapper script.
-←
+More information on how each method was wrapped can be found within the comments of each wrapper script, available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`.←
+
 →
@@
 These parameters were then given to the simulators when applicable, as described below.
+To select a diverse set of $k$ reference real datasets, we clustered all the parameters of each real dataset, and used the $k$ cluster centers from a pam clustering (implemented in the R cluster package);
 ←
@@
 
-The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+The dyngen ( `r print_url("https://github.com/dynverse/dyngen")`) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
-→We simulated a total of `r nrow(design_dyngen)`, with `r length(unique(design_dyngen$modulenet_name))` different topologies.←
+→We simulated a total of `r nrow(design_dyngen)` datasets, with `r length(unique(design_dyngen$modulenet_name))` different topologies.←
 #### Network generation
@@
 
+This count matrix was then filtered and normalised using the pipeline described below.
+
 #### Gold standard extraction
@@
 ### dyntoy
+```{r}
+design_dyntoy <- read_rds(result_file("design_dyntoy.rds", "01-datasets/02-synthetic"))
+```
 
+For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (`r print_url("https://github.com/dynverse/dyntoy")`) .
+We created `r length(unique(design_dyntoy$topology_model))` topology generators (described below), and with   `r nrow(design_dyntoy)/length(unique(design_dyntoy$topology_model))` datasets per generator, we created `r nrow(design_dyntoy)` datasets.
+
+We created a set of topology generators:
+
+* Linear and cyclic, with number of milestones $\sim B(10, 0.25)$
+* Bifurcating and converging, with four milestones
+* Binary tree, with number of branching points $\sim U(3, 6)$
+* Tree, with number of branching points $\sim U(3, 6)$ and maximal degree $\sim U(3, 6)$
+
+For more complex topologies we first calculated a random number of "modifications" $\sim U(3, 6)$ and a $\textit{deg}_{\textit{max}} \sim B(10, 0.25) + 1$.
+For each type of topology, we defined what kind of modifications are possible: divergences, loops, convergences and divergence-convergence.
+We then iteratively constructed the topology by uniformly sampling from the set of possible modifications, and adding this modification to the existing topology.
+For a divergence, we connected an existing milestone to a number of a new milestones.
+Conversely, for a convergence we connected a number of new nodes to an existing node.
+For a loop, we connected two existing milestones with a number of milestones in between.
+Finally for a divergence-convergence we connected an existing milestone to several new milestones which again converged on a new milestone.
+The number of nodes was sampled from $\sim B(\textit{deg}_{\textit{max}} - 3, 0.25) + 2$
+
+* Diverging-converging, allowed divergence and converging modifications
+* Diverging with loops, allowed divergence and loop modifications
+* Multiple looping, allowed looping modifications
+* Connected, allowed looping, divergence and convergence modifications
+* Disconnected, number of components sampled from $\sim B(5, 0.25) + 2$, for each component we randomly chose a topology from the ones listed above
+
+After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$.
+We added regions of delayed commitment to a divergence by virtually flipping a coin.
+[a]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[b]
+
+For each gene (same number as from the reference real dataset), we calculated the Kamada-Kawai layout in 2 dimensions, with edge weight equal to the length of the edge.
+For this gene, we then extracted for each cell a density value using a bivariate normal distribution with $\mu \sim U(x_{\textit{min}}, x_{\textit{min}})$ and $\sigma \sim U(x_{\textit{min}}/10, x_{\textit{min}}/8)$.
+We used this density as input for a zero-inflated negative binomial distribution with $\mu ~ U(100, 1000) \times \textit{density}$, $k ~ U(\mu / 10, \mu / 4)$ and $pi$ from the parameters of the reference real dataset, to get the final count values.
+
+This count matrix was then filtered and normalised using the pipeline described below.
 ### PROSSTT
+```{r}
+design_prosstt <- read_rds(result_file("design_dyntoy.rds", "01-datasets/02-synthetic"))
+```
 
+PROSSTT is a recent data simulator [@papadopoulosPROSSTTProbabilisticSimulation2018], which simulates expression using linear mixtures of expression programs and random walks through the trajectory.
+We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets, leading to a total of 50 datasets.
+
+Using the `simulate_lineage` function, we simulated the lineage expression, with parameters $\a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$.
+These parameter distributions were chosen very broad so as to make sure both easy and difficult datasets are simulated.
+After simulating base gene expression with `simulate_base_gene_exp`, we used the `sample_density` function to finally simulate expression values of a number of cells (the same as from the reference real dataset), with $\alpha \sim \textit{Lognormal}$ ($\mu = 0.3$ and $\sigma = 1.5$) and $\beta \sim \textit{Lognormal}$ ($\mu = 2$ and $\sigma = 1.5$).
+Each of these parameters were centered around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.
+
+This count matrix was then filtered and normalised using the pipeline described below.
 ### Splatter
+Splatter [@zappia_splattersimulationsinglecell_2017] simulates expression values by constructing non-linear paths between different states, each having a distinct expression profile.
+We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets, leading to a total of 50 datasets.
+
+We used the `splatSimulatePaths` function from Splatter to simulate datasets, with number of cells and genes equal to those in the reference real dataset, and with parameters  $\textit{nonlinearProb}$, $\textit{sigmaFac}$ and $\textit{skew}$ all sampled from $U(0, 1)$.
+
+This count matrix was then filtered and normalised using the pipeline described below.
+
 ←
@@
 We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016].
-The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3).
-We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05.
+The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (→which we set to 3←).
+We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.02.
 Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5.
-As a final filter, we removed both zero genes and cells until convergence.
+As a final filter, we removed both all-zero genes and cells until convergence.
 ## Evaluation metrics
@@
 Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (`r ref("snote", "metrics")`).
-Based on this analysis, we chose 4 metrics for ←
-
-  We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`).
-Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.
-
-```{r}
-add_sfig(
-  ggplot(),#figure_file("aspects_v2.svg", experiment_id="manual_figures/metrics"), 
-  "score_aspects", 
-  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
-  "",
-  15, 
-  8
-)
-```
+Based on this analysis, we chose 4 metrics for the evaluation, each assessing a different aspect of the trajectory: (i) the `r label_metric("him")` measures the topological similarity, (ii) the  `r label_metric("F1_branches")` compares the branch assignment, (iii) the  `r label_metric("correlation")` assesses the similarity in pairwise cell-cell distances and thus the cellular positions, and (iv) the  `r label_metric("featureimp_cor")` looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
+←
 ### Correlation between geodesic distances
@@
 
-[a]Gotta update the figure to reflect this
+[a]Don't put this in the paper
+[b]Factually incorrect but ¯\_(ツ)_/¯

### 2018-09-19 manuscript/paper.Rmd
update paper with discussion on the metrics added

@@
 ### Method wrappers
-Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`).
+Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynmethods")`).
 These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
@@
 
-More information on how each method was wrapped can be found within the comments of each wrapper script, available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`.←
+More information on how each method was wrapped can be found within the comments of each wrapper script, available at `r print_url("https://github.com/dynverse/dynmethods")`.←
 
@@
 For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (`r print_url("https://github.com/dynverse/dyntoy")`) .
-We created `r length(unique(design_dyntoy$topology_model))` topology generators (described below), and with   `r nrow(design_dyntoy)/length(unique(design_dyntoy$topology_model))` datasets per generator, we created `r nrow(design_dyntoy)` datasets.
+We created `r length(unique(design_dyntoy$topology_model))` topology generators (described below), and with   `r nrow(design_dyntoy)/length(unique(design_dyntoy$topology_model))` datasets per generator, this lead to a total of `r nrow(design_dyntoy)` datasets.
 
@@
 
+* Looping, allowed loop modifications
 * Diverging-converging, allowed divergence and converging modifications
@@
 Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (`r ref("snote", "metrics")`).
-Based on this analysis, we chose 4 metrics for the evaluation, each assessing a different aspect of the trajectory: (i) the `r label_metric("him")` measures the topological similarity, (ii) the  `r label_metric("F1_branches")` compares the branch assignment, (iii) the  `r label_metric("correlation")` assesses the similarity in pairwise cell-cell distances and thus the cellular positions, and (iv) the  `r label_metric("featureimp_cor")` looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
+Based on this analysis, we chose 4 metrics for the evaluation, each assessing a different aspect of the trajectory: (i) the `r label_metric("him")` measures the topological similarity, (ii) the   `r label_metric("F1_branches")`compares the branch assignment, (iii) the  `r label_metric("correlation")` assesses the similarity in pairwise cell-cell distances and thus the cellular positions, and (iv) the  `r label_metric("featureimp_wcor")` looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
+
+### The Hamming-Ipsen-Mikhailov metric
+The Hamming-Ipsen-Mikhailov (HIM) [@jurman_HIM_Glocal_Metric_2015] uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
+It is a linear combination of the normalised Hamming distance, which gives an indication of the differences in edge lengths, and the normalised Ipsen-Mikhailov distance, which assesses the similarity in degree distributions.
+The latter has a parameter $\gamma$, which was fixed at 0.1 as to make the scores comparable between datasets.
+We illustrate the metric and discuss alternatives in `r ref("snote", "metrics")`.
+
+### The F1 between branch assignments
+To compare branch assignment, we borrowed the $F_1$ metric from the field of expression biclustering [@saelens_comprehensiveevaluationmodule_2018].
+To calculate this metric, we first calculated the similarity of all pairs of branches between the two trajectories using the Jaccard similarity.
+Next, we define the $\textit{Recovery}$ (respectively $\textit{Relevance}$) as the average maximal similarity of all branches in the reference dataset (respectively $\textit{prediction}$).
+The `r label_metric("F1_branches")` was then defined as the harmonic mean between $\textit{Recovery}$ and $\textit{Relevance}$.
+We illustrate this metric further in  `r ref("snote", "metrics")`.
 ←
 ### Correlation between geodesic distances
-The similarity in cell ordering between two trajectories is assessed by calculating the geodesic distances between each pair of cells for both trajectories.
-The definition of a geodesic distance between two cells part of the common trajectory model will be demonstrated using a toy example (`r ref("sfig", "metric_geodesic")`).
+→When the position of a cell is the same in both the gold standard and the prediction, its relative distances to all other cells in the trajectory should also be the same.
+This observation is the basis for the `r label_metric("correlation")` metric.
+To calculate the `r label_metric("correlation")`, we first sampled 100 waypoint cells in both the prediction and the reference dataset, using stratified sampling between the different milestones, edges and regions of delayed commitment, weighted by the number of cells in each collection.
+We then calculated the geodesic distances between the union of waypoint cells from both datasets and all other cells.
+The calculation of the geodesic distance depended on the location of the two cells within the trajectory, further discussed in `r ref("snote", "metrics")` and was weighted by the length of the edge in the milestone network.
+Finally, the `r label_metric("correlation")` was defined as the Spearman rank correlation between the distances of both datasets.
+We illustrate the metric and assess the effect of the number of waypoint cells in `r ref("snote", "metrics")`.
+### The correlation between important features
+The `r label_metric("featureimp_wcor")` assesses whether similar dynamically differentially expressed features are found between the prediction and model.
+To calculate this metric, we used Random Forest regression (implemented in the R ranger package), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
+We then extract feature importance values for each feature, and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset as to give more weight to differentially expressed features then non-differentially expressed features.
+As hyperparameters we set the number of trees to 10000, and the number of features on which can be split to 1% of all available features.
+We illustrate this metric and assess the effect of its hyperparameters in `r ref("snote", "metrics")`.
 
-```{r}
-add_sfig(
-  ggplot(),#figure_file("geodesic.svg", experiment_id="manual_figures/metrics"), 
-  "metric_geodesic", 
-  "The modified geodesic distances demonstrated on a toy example.", 
-  "a) A toy example containing four milestones (W to Z) and five cells (a to e).
-b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
-c) The calculations made for calculating the pairwise geodesic distances.
-d) A heatmap representation of the pairwise geodesic distances."
-)
-```
-
-The geodesic distance between two cells depends on whether they are (i) on the same transition, or (ii) in different transitions.
-In the first case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on.
-For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
-In the latter case, the distances between the cells and all of their neighbouring milestones will be calculated.
-These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
-For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
-
-According to the defined common trajectory model (`r ref("fig", "ti_evaluation_overview", "b")`), cells are also allowed to have a delayed commitment.
-In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network.
-In this case, the distance between two cells both inside a region of delayed commitment is calculated as the manhattan distances between the milestone percentages weighted by the transition weights from the milestone network.
-For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$.
-The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
-
-Finally, calculating all pairwise distances between cells would scale poorly for trajectories with large numbers of cells.
-For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories.
-The waypoints are determined by viewing each milestone, transition and region of delayed commitment as a collection of cells, and sampling cells from the different collections weighted by the total number of cells within that collection.
-For calculating the correlation of geodesic distances between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
-For the benchmark evaluation, the total number of waypoints sampled from a trajectory was one hundred.
-
-### Random forest prediction error
-
-Although the correlation between geodesic distances directly assesses the position of the cells in the trajectory, a bad correlation does not directly imply that similar cells were not grouped together by the method, as illustrated in `r ref("sfig", "score_aspects", c("b", "c"))`.
-For example, certain methods will inevitably reach an incorrect ordering because they cannot handle the correct trajectory type, but these methods could still correctly place similar cells next to each other.
-We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard.
-We used a Random Forest regression, implemented in the *ranger* package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of each cell in the gold standard, using the milestone percentages of this cells in the prediction as features.
-We then used the out-of-bag mean-squared error on these percentages to score each method's capability of predicting the correct neighbourhood of each cell.
-
-### Edge flip score
-
-As a third independent score, we assessed the similarity between the milestone network topologies.
-We first simplified each network, by merging consecutive linear edges into one edge, and adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks.
-Because we are interested in the overall similarity between two topologies irrespective of the direction of the edges, the network was made undirected.
-Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
-This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012].
-We implemented a branch and bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, we tried out all solutions with additional two edge changes.
-To further limit the search space, we made sure the degree distributions between the two networks were  similar, before assessing whether the two networks were isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.
-
-A comparison of edge flip scores between common trajectory topologies is illustrated in `r ref("sfig", "edge_flip_comparison")`.
-
-```{r}
-add_sfig(
-  ggplot(),#figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
-  "edge_flip_comparison", 
- "Edge flip score values for common trajectory topologies",
-  ""
-)
-```
-### Aggregation of scores
+←
+### Aggregation of scores[c]
 `r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
@@
 We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_checks")`).
-The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The goal of this quality control is in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
 The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
@@
 
-Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions), the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing), or indications that the tool will be broadly applicable on new datasets (such as being open-source, containing a good tutorial and support system, and being thoroughly evaluated in the study where it was published).
+Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions) or the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing).
 
@@
 →
-## Scalability
+## Scalability[d]
+## [e]Stability
 ←
@@
 [b]Factually incorrect but ¯\_(ツ)_/¯
+[c]Robrecht, you know what to do ;)
+[d]I will try to do my best in adding this if you don't do it first Robrecht
+[e]I will try to do my best in adding this if you don't do it first Robrecht

### 2018-09-19 manuscript/supplementary.Rmd
the supplementary works again :sunglasses:

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
output: 
  dynbenchmark::pdf_manuscript:
      render_change: FALSE
---

# Supplementary material

```{r setup, include=FALSE}
library(dynbenchmark)
library(tidyverse)
```

```{r}
supplementary <- read_rds("supplementary.rds")
sfigs <- supplementary$sfigs
```

# Supplementary figures
```{r, results="asis"}
sfigs %>% pmap(dynbenchmark:::plot_fig, ref_type = "sfig") %>% walk(cat)
```

# Supplementary tables

```{r setup, include=FALSE}
# source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/setup.R"))
# 
# # knitr options
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
# knitr::opts_chunk$set(
#   collapse = TRUE,
#   warning = FALSE,
#   message = FALSE,
#   echo = FALSE,
#   comment = "#>",
#   fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
# )
# 
# list2env(read_rds(derived_file("sfigs_stables.rds", "paper")), environment())
```

```{r "supplementary", echo=FALSE, results='asis'}
# source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/sfigs_stables.R"))
```

### 2018-09-19 package/R/text_ref.R

@@ -91,7 +91,7 @@ add_fig <- function(
   # save it because why not
   figs <<- figs %>% add_row(
     ref_id = ref_id,
-    fig_path = fig_path,
+    fig_path = if(is.character(fig_path)) {fig_path} else {list(fig_path)},
     caption_main = caption_main,
     caption_text = caption_text,
     width = width,
@@ -112,10 +112,10 @@ add_sfig <- function(
   height = 7,
   format = get_default_format()
 ) {
-  # save it because why not
-  figs <<- sfigs %>% add_row(
+  # save it because it's necessary
+  sfigs <<- sfigs %>% add_row(
     ref_id = ref_id,
-    fig_path = fig_path,
+    fig_path = if(is.character(fig_path)) {fig_path} else {list(fig_path)},
     caption_main = caption_main,
     caption_text = caption_text,
     width = width,

### 2018-09-19 manuscript/paper.Rmd
update paper with link to modulenetworks

@@
 →To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
-Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating converging), (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).←
+Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating converging), (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
+The structure of these module networks is available at `r print_url("https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks")`.←
 

### 2018-09-19 manuscript/supplementary.Rmd
add ability to include supplementary tables as excel

@@
 sfigs <- supplementary$sfigs
+stables <- supplementary$stables
+refs <- supplementary$refs
 ```
@@
 # Supplementary tables
+
+```{r, results="asis"}
+stables %>% pmap(dynbenchmark:::show_table, ref_type = "stable") %>% walk(cat)
+```
 

### 2018-09-19 manuscript/paper.Rmd
update

@@
 These methods therefore mainly focused on correctly ordering the cells along the fixed topology.
-→Most recent methods also infer the topology computationally, which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←
+→Most recent methods also infer the topology [@street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←
 
-```{r, eval = FALSE}
+```{r}
 add_stable(
-  read_rds(result_file("tools_table.rds", "03-methods"))[[params$table_format]],
+ result_file("tools.xlsx", "03-methods"),
   "tools",
-paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::glue_collapse("; "))
+paste0("→Overview of available trajectory inference tools, and whether they were included in this study.←"),
+  ""
 )
@@
 → 
-The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].
+The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[a])) [@cannoodt_goldstandardsinglecell_].
 
-The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("github.com/dynverse/dynverse")`.
-These include packages for wrapping trajectories (`r print_url("github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("github.com/dynverse/dynguidelines")`).
-The main analysis repository is available at `r print_url("github.com/dynverse/dynbenchmark")` and is divided into several experiments.
-In this repository, we provide READMEs for both the scripts and the results, which can be easily browsed and explored on the github website.
+The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("https://www.github.com/dynverse/dynverse")`.
+These include packages for wrapping trajectories (`r print_url("https://www.github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("https://www.github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("https://www.github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("https://www.github.com/dynverse/dynguidelines")`).
+The main analysis repository is available at `r print_url("https://www.github.com/dynverse/dynbenchmark")` and is divided into several experiments.
+In this repository, we provide readmes for both the scripts and the results, which can be easily browsed and explored on the github website.
 ←
@@
 <!-- Overview -->
-We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
-We welcome any contributions by creating an issue at →`r print_url("github.com/dynverse/dynmethods/issues")`←.
+We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: `r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We welcome any contributions by creating an issue at →`r print_url("https://www.github.com/dynverse/dynmethods/issues")`←.
 
@@
 ```{r}
-#add_stable(
-#  read_rds(
-#     figure_file("datasets.rds","2-dataset_characterisation/2-real")
-#  )[[params$table_format]], 
-#  "datasets", 
-#  "Real datasets used in this study."
-#)
+add_stable(
+  result_file("metadata.xlsx","01-datasets/01-real"), 
+  "datasets", 
+  "Real datasets used in this study."
+)
 ```
@@
 We added regions of delayed commitment to a divergence by virtually flipping a coin.
-[a]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[b]
+[b]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[c]
 
@@
 ←
-### Aggregation of scores[c]
+### Aggregation of scores[d]
 `r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
@@
 →
-## Scalability[d]
-## [e]Stability
+## Scalability[e]
+## [f]Stability
 ←
@@
 
-[a]Don't put this in the paper
-[b]Factually incorrect but ¯\_(ツ)_/¯
-[c]Robrecht, you know what to do ;)
-[d]I will try to do my best in adding this if you don't do it first Robrecht
+[a]Update the doi when reuploaded
+[b]Don't put this in the paper
+[c]Factually incorrect but ¯\_(ツ)_/¯
+[d]Robrecht, you know what to do ;)
 [e]I will try to do my best in adding this if you don't do it first Robrecht
+[f]I will try to do my best in adding this if you don't do it first Robrecht

### 2018-09-21 manuscript/supplementary.Rmd
Update supplementary RMD

@@
 ---
-title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
+title: "Supplementary material for: \nA comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
 output: 
@@
 ---
-
-# Supplementary material
 

### 2018-09-21 manuscript/paper.Rmd
Update paper

@@
 n_tools <- nrow(tools)
+n_tools_evaluated <- nrow(filter(tools, evaluated))
 n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("tool", "adaptation", "offtheshelf")))
@@
 Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected [@wolf_graphabstractionreconciles_2017] graphs.
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected graphs [@wolf_graphabstractionreconciles_2017].
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
@@
 →Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
-Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type(s) of graph topologies they can detect.
-Early TI methods typically fixed the topology algorithmically (e.g.
-linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016].
+Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←.
+Early TI methods typically fixed the topology algorithmically (such as linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016].
 These methods therefore mainly focused on correctly ordering the cells along the fixed topology.
@@
 →We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`).
-Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
@@
 In this study, we presented a large scale evaluation of the performance and code of  `r n_methods_evaluated`  TI methods.
-By using a common structure and three metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types.
-We also assessed the code quality of these methods, to provide information on important scientific and software development practices for these methods.
+By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types.
+→We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on the same datasets.←
 
-<!-- FOR METHOD USERS -->
+<!-- FOR METHOD USERS -->[a][b]
 <!-- Practical guidelines -->
@@
 → 
-The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[a])) [@cannoodt_goldstandardsinglecell_].
+The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[c])) [@cannoodt_goldstandardsinglecell_].
 
@@
 <!-- Overview -->
-We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: `r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We gathered a list of →`r n_tools`← trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: →`r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018]←.
 We welcome any contributions by creating an issue at →`r print_url("https://www.github.com/dynverse/dynmethods/issues")`←.
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-→The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")`.
-In the end, we included  `r n_tools` tools in the evaluation, constituting `r n_methods_evaluated` methods.←
+→The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:\"won't wrap\"")`.
+In the end, we included  `r n_tools_evaluated` tools in the evaluation, constituting `r n_methods_evaluated` methods.←
 
@@
 ### Method wrappers
-Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynmethods")`).
+Each method was wrapped within docker and singularity containers (available at `r print_url("https://github.com/dynverse/dynmethods")`).
 These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
@@
 We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control.
-About ⅓ of authors responded, and we improved the wrappers based on their feedback.
+About one third of the authors responded, and we improved the wrappers based on their feedback.
 These discussions can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`.
@@
 (ii) The milestone percentages contains, for each cell, its position between milestones, and sums for each cell to one.
-(iii) →The regions of delayed commitment, which are connection between three or more milestones.
+(iii) →The regions of delayed commitment, connections between three or more milestones.
 These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←
@@
 ```{r}
-#add_sfig(
-#  figure_file("overview_wrapping_v1.svg", "manual_figures"),
-#  "overview_wrapping", 
-#  "A common interface for TI methods.", 
-#  "a) The input and output of each TI method is standardised.
+add_sfig(
+  raw_file("overview_wrapping_v2.svg", "03-methods"),
+  "overview_wrapping", 
+  "A common interface for TI methods.", 
+  "a) The input and output of each TI method is standardised.
 As input, each TI method #receives either raw or normalised counts, several parameters, and a selection of prior information.
@@
 b) The specific transformations performed by each of the wrapper functions is shown in more detail.",
-#  15, 
-#  7.5
-#)
+  13, 
+  7.5
+)
 ```
@@
 ## Real datasets
-We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
+We gathered real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
 The scripts to download and process these datasets are available on our repository (`r print_url("https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets")`).
@@
 These raw counts were all normalised and filtered using a common pipeline, discussed later.
+→Some datasets contained more than one trajectory, in which we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
+In the end, we included `r sum(list_datasets()$source == "real")` for this evaluation.←
 
@@
 ```{r}
-#add_stable(
-#  read_rds(figure_file(
-#    "samplers.rds",
-#    "2-dataset_characterisation/1-synthetic")
-#  )[[params$table_format]], 
-#  "samplers", 
-#  "Distributions from which each parameter in the thermodynamic model was sampled."
-#)
+add_stable(
+  read_rds(result_file("samplers.rds","01-datasets/02-synthetic")), 
+  "samplers", 
+  "Distributions from which each parameter in the thermodynamic model was sampled."
+)
 ```
@@
 We added regions of delayed commitment to a divergence by virtually flipping a coin.
-[b]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[c]
+[d]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[e]
 
@@
 ←
-### Aggregation of scores[d]
+### Aggregation of scores[f]
 `r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
@@
 
-We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_checks")`).
+We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`).
 The goal of this quality control is in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
@@
 ```{r}
-#add_stable(
-#  read_rds(figure_file(
-#    "qc_checks.rds",
-#    "4-method_characterisation"
-#  ))[[params$table_format]], 
-#  "qc_checks", 
-#  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
-#)
+add_stable(
+ result_file("qc_scoresheet.xlsx", "03-methods/02-tool_qc"),
+  "qc_scoresheet",
+paste0("→Scoring sheet for method quality control.
+Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.←"),
+  ""
+)
 ```
@@
 
-Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("stable", "qc_checks")`).
+Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("stable", "qc_scoresheet")`).
 This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface.
-Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_checks")`).
+Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_scoresheet")`).
 For calculating the final score, we weighed each of the six categories equally.
 →
-## Scalability[e]
-## [f]Stability
+## Scalability[g]
+## [h]Stability
 ←
@@
 
-```{r "supplementary", echo=FALSE, results='asis'}
-#if (params$table_format == "html") 
-#  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
-#write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
-#write_rds(figs, derived_file("figs.rds", "paper"))
-```
-
 <!--
@@
 
-[a]Update the doi when reuploaded
-[b]Don't put this in the paper
-[c]Factually incorrect but ¯\_(ツ)_/¯
-[d]Robrecht, you know what to do ;)
-[e]I will try to do my best in adding this if you don't do it first Robrecht
-[f]I will try to do my best in adding this if you don't do it first Robrecht
+[a]Discussion still has to be updated from here onwards, but I think the structure can stay the same
+[b]Perhaps one thing which can be added is the docker/singularity wrapping of methods.
+[c]Update the doi when reuploaded
+[d]Don't put this in the paper
+[e]Factually incorrect but ¯\_(ツ)_/¯
+[f]Robrecht, you know what to do ;)
+[g]I will try to do my best in adding this if you don't do it first Robrecht
+[h]I will try to do my best in adding this if you don't do it first Robrecht

### 2018-09-25 manuscript/supplementary.Rmd

@@
 ```{r, results="asis"}
-sfigs %>% pmap(dynbenchmark:::plot_fig, ref_type = "sfig") %>% walk(cat)
+sfigs %>% pmap(dynbenchmark:::plot_fig, ref_type = "sfig") %>% modify(paste0, "\n\n") %>% walk(cat)
 ```
@@
 ```{r, results="asis"}
-stables %>% pmap(dynbenchmark:::show_table, ref_type = "stable") %>% walk(cat)
+stables %>% pmap(dynbenchmark:::show_table, ref_type = "stable") %>% modify(paste0, "\n\n") %>% walk(cat)
 ```

### 2018-09-27 manuscript/paper.Rmd
update on metrics

@@
 ² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
-³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
+³ Center for Medical Genetics, Ghent plot_topology_complexity %>% write_rds(result_file("topology_complexity.rdsUniversity Hospital, Ghent, Belgium.  
 ⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
@@
 
-`dynverse/dynbenchmark/commit/`r git2r::revparse_single(".","HEAD")$sha``
+`www.github.com/dynverse/dynbenchmark/commit/`r git2r::revparse_single(".","HEAD")$sha``
 ```{r}
@@
 Since 2014, at least →`r n_tools`← tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
-As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.
+As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for users in the field.
 
@@
 Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
-Instead, the choice of method →mostly← depends on the trajectory topology present in the data.
-→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://dynverse.duckdns.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+Instead, the choice toof method →mostly← depends on the trajectory topology present in the data.
+→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
+→Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
 Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←.
@@
 In this study, we performed a comprehensive evaluation of  →`r n_methods_evaluated`← TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
-The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
-→We evaluated each methods on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) the stability of the predictions on the same dataset.
+→We included methods primarily based← on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
+→We evaluated each method on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) the stability of the predictions across multiple bootstrap replicates.
 Finally, we combined these metrics to provide a set of practical guidelines for users.
@@
   caption_text = "a) A schematic overview of our evaluation pipeline.
-b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
+b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
 c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity.
@@
 We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
-Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories.
+Using several similarity metrics, we compared the →reference trajectory← of the real and synthetic datasets with the inferred trajectories.
 We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
@@
 <!-- Evaluation overview -->
-In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
+In order to make →the outputs from different methods directly comparable to each other←, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
@@
 →We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`).
-Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance.
+Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`)[a].←The data compendium consisted of both synthetic datasets, which offer the most exact →reference trajectory←, and real datasets, which offer the highest biological relevance.
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
 →For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011].
-We paired each simulated datasets with a real dataset to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←
+For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←
 
 <!-- Overall evaluation results -->
-```{r echo=FALSE, results="asis"}
+```{r}
 add_fig(
@@
 b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
-`r {n_control_methods_evaluated}` control methods were also included as an indication of baseline performance c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
-d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation.
+c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+d) Performance in the quality control evaluation is highly variabl[b]e, even amongst the highest ranked methods according to the benchmark evaluation.
+Also listed are the quality control scores aggregated according to practicality and the different categories.",
+  width = 15,
+  height = 20
+)
+```
+
+```{r}
+add_fig(
+  fig_path = result_file("overview.pdf", experiment_id = "08-summary"),
+  ref_id = "results_overview_detailed", 
+  caption_main = glue::glue("Overview of the results on the evaluation of {n_methods_evaluated} TI methods and {n_control_methods_evaluated} control methods."), 
+  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
+b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
+c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
+d) Performance in the quality control evaluation is highly variabl[c]e, even amongst the highest ranked methods according to the benchmark evaluation.
 Also listed are the quality control scores aggregated according to practicality and the different categories.",
@@
 
-## Relationships between method characteristics and performance
-
+## Topology inference
 <!-- Inverse relation between edge flip and correlation -->
+In[d] most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`).
+These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
+However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
+This indicates that current non-linear methods potentially contain less accurate ordering algorithms.
+A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.
 
 <!-- Topological complexity versus performance -->
+Despite[e] their similar overall performance, the topologies predicted by the top methods differed considerably.
+Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity")`).
+AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data.
+Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology.
+Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively and of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively and of the cases.
+Inferring the correct topology without parameter tuning is therefore still an open challenge.
+
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = result_file("topology_complexity.rds", experiment_id = "10-topologies"),
+  ref_id = "topology_complexity", 
+  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the reference trajectory.", 
+  caption_text = "Shown are the distributions of the difference in size between predicted and →reference← topologies.
+A positive difference means that the topology predicted by the method is more complex than the one in the →reference←.",
+  width = 12,
+  height = 4
+)
+```
 
@@
 ## Method quality control
+While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
+To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scorsheet")`).
+
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("sfig", "qc_overview")`).
+While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers.
+Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+
+```{r}
+add_sfig(
+  fig_path = result_file("qc_overview.rds", experiment_id = "03-methods/02-tool_qc"),
+  ref_id = "qc_overview", 
+  caption_main = "Overview of the quality control results for every method.", 
+  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_scoresheet')}).
+Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
+Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation.
+This is represented in the plot as the height on the y-axis.
+Top: Average QC score for each method.
+Right: The average score of each quality control item.
+Shown into more detail are those items which had an average score lower than 0.5."),
+   width = 15,
+   height = 18
+)
+```
+
+```{r}
+add_stable(
+ result_file("qc_scoresheet.xlsx", "03-methods/02-tool_qc"),
+  "qc_scoresheet",
+paste0("→Scoring sheet for method quality control.
+Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.←"),
+  ""
+)
+```
+
+We observed no clear relation between method quality and method performance (`r ref("sig", "results_overview")`).
+We could also not find any quality aspect which was significantly associated with method performance.
 
@@
 <!-- The context -->
-In this study, we presented a large scale evaluation of the performance and code of  `r n_methods_evaluated`  TI methods.
-By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on different trajectory types.
-→We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on the same datasets.←
+In this study, we presented a large scale evaluation of the performance of  `r n_methods_evaluated`  TI methods.
+By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on →different types of trajectories.
+We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on the same datasets.←
 
-<!-- FOR METHOD USERS -->[a][b]
+<!-- FOR METHOD USERS -->
 <!-- Practical guidelines -->
-Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`).
+Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("guidelines.dynverse.org")`).
 We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
-In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex trajectory structures with at most one bifurcation, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories.
-In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories, although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).
+In each of these use cases, →our evaluation suggests a different set of optimal methods, as shown in `r ref("fig", "user_guidelines")`.
+Several other factors will also impact the choice of methods, such as the dimensions of the dataset, and the prior information which is available.
+These factors and several others can all be changed in our interactive app (`r print_url("http://dynverse.org")`).
+This app can also be used to explore the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←
 
-```{r echo=FALSE, results="asis"}
+```{r}
 add_fig(
-  fig_path = ggplot(),#figure_file("tree.svg", experiment_id = "7-user_guidelines"),
+  fig_path = result_file("guidelines.svg", experiment_id = "09-guidelines"),
   ref_id = "user_guidelines", 
@@
 Methods to the right are ranked according to their performance on a particular (set of) trajectory type.
-Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
+Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information.",
+  width = 13,
+  height = 10
 )
@@
 When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
-When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.
+When a more complex topology is produced, →this could indicate that the underlying biology is much more complex than expected by the user.←
 
 <!-- A common trajectory analysis framework -->
-Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017] or module detection [@saelens_comprehensiveevaluationmodule_2018].
-Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`).
-In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values.
-In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.
+Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as find genes important for the trajectory, network inference [@aibar_scenicsinglecellregulatory_2017] or finding modules of genes [@saelens_comprehensiveevaluationmodule_2018].
+Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods →on some a linear, a tree, a cyclic and a disconnected graph dataset (`r ref("fig", "example_predictions")`).
+Using our framework, this figure can be recreated by using only a couple of lines of R code (`r print_url("https://www.github.com/dynverse/dynmethods")`).
+This framework should still be extended to allow easier downstream analyses.
+In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.←
 
-```{r echo=FALSE, results="asis"}
+```{r}
 add_fig(
-  fig_path = ggplot(),#figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
+  fig_path = result_file("example_predictions.rds", experiment_id = "11-example_predictions"),
   ref_id = "example_predictions", 
   caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
-  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process.
-A multi-dimensional scaling of dataset is shown at the bottom.
-MEF: mouse embryonic fibroblast b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions.
-The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology.
-c) More insight can be gained into the ordering of the cells by placing the branches one after the other.
-The predicted topology is shown in the bottom of each plot.
-d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed.
-e) Our metrics can be used to compare the output to a gold standard or to each other.
-In this case, while only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
+  caption_text = "",
+  width = 13,
+  height = 8
 )
@@
 <!-- New tools: challenges and possibilities -->
-Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories.
+Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (as we illustrate with two datasets in `r ref("fig", "example_predictions", "a and b")`).
 Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories.
-Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods.
-Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers.
+Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly fail both on real and synthetic data, even if the trajectory can be easily identified in a dimensionality reduction (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
+← Furthermore, higher standards for code assurance, documentation and scalability could help the adoption of these tools across single-cell -omics field.
 To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
@@
 Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution, or provide a unique insight in specific use cases.
-Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns an measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
+Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns a measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
 We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.
@@
 → 
-The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[c])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[f])) [@cannoodt_goldstandardsinglecell_].
 
-The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("https://www.github.com/dynverse/dynverse")`.
-These include packages for wrapping trajectories (`r print_url("https://www.github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("https://www.github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("https://www.github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("https://www.github.com/dynverse/dynguidelines")`).
-The main analysis repository is available at `r print_url("https://www.github.com/dynverse/dynbenchmark")` and is divided into several experiments.
-In this repository, we provide readmes for both the scripts and the results, which can be easily browsed and explored on the github website.
+The main analysis repository is available at `r print_url("benchmark.dynverse.org")` and is divided into several experiments.
+Each experiment has its own set of scripts and results, each accompanied by an illustrated readme which can be easily browsed and explored on the github website.
+
+The analysis scripts call several other R packages, of which an overview is available at `r print_url("dynverse.org")`.
+These packages include **dynwrap**, used to wrap the output of methods into the common trajectory model, **dyneval**, which contains the evaluation metrics, **dynguidelines**, the guidelines app, and **dynplot** for plotting trajectories.
 ←
@@
 We gathered a list of →`r n_tools`← trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: →`r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018]←.
-We welcome any contributions by creating an issue at →`r print_url("https://www.github.com/dynverse/dynmethods/issues")`←.
+We welcome any contributions by creating an issue at →`r print_url("methods.dynverse.org")`←.
 
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-→The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:\"won't wrap\"")`.
+→The discussions on why these methods were excluded can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:\"won't wrap\"")`.
 In the end, we included  `r n_tools_evaluated` tools in the evaluation, constituting `r n_methods_evaluated` methods.←
@@
 ### Method wrappers
-Each method was wrapped within docker and singularity containers (available at `r print_url("https://github.com/dynverse/dynmethods")`).
+Each method was wrapped within docker and singularity containers (available at `r print_url("methods.dynverse.org")`).
 These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
@@
 
-The prior information was extracted from the gold standard as follows:
+The prior information was extracted from the →reference trajectory← as follows:
 
@@
 * **Discrete time course** For each cell a time point from which it was sampled.
-If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used.
+If available, directly extracted from the reference trajectory, otherwise the geodesic distance from the root milestone was used.
 For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
@@
   "a) The input and output of each TI method is standardised.
-As input, each TI method #receives either raw or normalised counts, several parameters, and a selection of prior information.
+As input, each TI method receives either raw or normalised counts, several parameters, and a selection of prior information.
 After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model.
@@
 ### Control methods
-We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the gold standard, and **shuffle**, which will place the cells randomly on the gold standard milestone network.
+We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the reference trajectory, and **shuffle**, which will place the cells randomly on the reference milestone network.
 ### Off-the-shelf methods
@@
 These raw counts were all normalised and filtered using a common pipeline, discussed later.
-→Some datasets contained more than one trajectory, in which we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
+→Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
 In the end, we included `r sum(list_datasets()$source == "real")` for this evaluation.←
@@
 
-For every simulator, we took great care with trying to make the datasets look like real data as  best as possible.
+For every simulator, we took great care to make the datasets as realistic as possible.
 To do this, we extracted several parameters from all real datasets.
 We calculated the number of differentially expressed features within a trajectory using a wilcoxon test between every pair of cell groups.
+These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5% of cells, and had at least a fold-change of 2.
 We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package [@zappia_splattersimulationsinglecell_2017].
-These parameters were then given to the simulators when applicable, as described below.
-To select a diverse set of $k$ reference real datasets, we clustered all the parameters of each real dataset, and used the $k$ cluster centers from a pam clustering (implemented in the R cluster package);
+These parameters were then given to the simulators when applicable, as described for each simulator below.
+To select a diverse set of $k$ reference real datasets, we clustered all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (implemented in the R cluster package) to generate synthetic data.
 ←
@@
 After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$.
-We added regions of delayed commitment to a divergence by virtually flipping a coin.
-[d]We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, depending on the length of the edges in the milestone network.[e]
+We added regions of delayed commitment to a divergence by in half of the cases.
+We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, based on the length of the edges in the milestone network.
 
@@
 PROSSTT is a recent data simulator [@papadopoulosPROSSTTProbabilisticSimulation2018], which simulates expression using linear mixtures of expression programs and random walks through the trajectory.
-We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets, leading to a total of 50 datasets.
+We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets.
+However, due to frequent crashes of the tool, only 19 datasets created output and were thus used in the evaluation.
 
@@
 ### Correlation between geodesic distances
-→When the position of a cell is the same in both the gold standard and the prediction, its relative distances to all other cells in the trajectory should also be the same.
+→When the position of a cell is the same in both the reference and the prediction, its relative distances to all other cells in the trajectory should also be the same.
 This observation is the basis for the `r label_metric("correlation")` metric.
@@
 ←
-### Aggregation of scores[f]
+### Aggregation of scores[g]
 `r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
@@
 
-```{r}
-add_stable(
- result_file("qc_scoresheet.xlsx", "03-methods/02-tool_qc"),
-  "qc_scoresheet",
-paste0("→Scoring sheet for method quality control.
-Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.←"),
-  ""
-)
-```
-
-Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions) or the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing).
+Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions) and/or the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing).
 
@@
 →
-## Scalability[g]
-## [h]Stability
+## Scalability[h]
+To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section.
+We up or down-scaled these datasets … OKAY I GIVE UP generate_dataset.R
+
+## [i]Stability
 ←
@@
 
-[a]Discussion still has to be updated from here onwards, but I think the structure can stay the same
-[b]Perhaps one thing which can be added is the docker/singularity wrapping of methods.
-[c]Update the doi when reuploaded
-[d]Don't put this in the paper
-[e]Factually incorrect but ¯\_(ツ)_/¯
-[f]Robrecht, you know what to do ;)
-[g]I will try to do my best in adding this if you don't do it first Robrecht
+[a]De nomenclatuur in deze figuur moet nog aangepast worden aan de nieuwe maten...
+[b]Als we zoiets zeggen moeten we misschien ook SDs tonen op de plots ?
+[c]Als we zoiets zeggen moeten we misschien ook SDs tonen op de plots ?
+[d]Update when benchmarking results
+[e]Update when benchmarking results
+[f]Update the doi when reuploaded
+[g]TODO
 [h]I will try to do my best in adding this if you don't do it first Robrecht
+[i]I will try to do my best in adding this if you don't do it first Robrecht

### 2018-09-27 scripts/02-metrics/02-metric_conformity/helper-rules.R

@@ -518,7 +518,7 @@ combined_local_global_position_change <- rule_combined(
   ),
   method_ids = c("shuffle_edges", "shuffle_cells_edgewise", "shuffle_cells"),
   dataset_ids = dataset_design %>% filter(topology_id %in% names(topologies), cell_positioning != "milestones", num_cells > 50) %>% pull(dataset_id),
-  observation = "Most metrics that look at the position of each cell conform to this rule."
+  observation = "Because the topology remains the same, the topology scores do not conform to this rule. Also the clustering based scores have some difficulties with this rule."
 )
 
 filter_cells <- rule_monotonic(
@@ -531,7 +531,7 @@ filter_cells <- rule_monotonic(
   varied_parameter_id = "filter_perc",
   varied_parameter_name = "Filtered cells",
   varied_parameter_labeller = scales::percent,
-  observation = "Only metrics which look only at the topology do not conform to this rule."
+  observation = "Metrics which look at the topology do not conform to this rule."
 )
 
 remove_divergence_regions <- rule_lower(
@@ -544,19 +544,19 @@ remove_divergence_regions <- rule_lower(
     num_cells >= 100
   ) %>% pull(dataset_id),
   method_id = "remove_divergence_regions",
-  observation = glue::glue("Both {label_metric('F1_branches', 'latex')} and {label_metric('edgeflip', 'latex')} fail here because neither the topology nor the branche assignment changes.")
+  observation = glue::glue("Both {label_metric('F1_branches', 'latex')} and {label_metric('edge_flip', 'latex')} fail here because neither the topology nor the branch assignment changes. Moreover, the decreases in score are relatively minor for all metrics, given that the impact of the positions of the cells is only minimal.")
 )
 
 time_warping_start <- rule_monotonic(
   id = "time_warping_start",
   name = "Move cells to start milestone",
-  description = "Moving the cells closer to their start milestone should lower the score. Cells were moved closer to the start milestone by doing $\\textit{percentage}^{\\textit{warp magnitude}}",
+  description = "Moving the cells closer to their start milestone should lower the score. Cells were moved closer to the start milestone using $\\textit{percentage}_{\\textit{new}} = \\textit{percentage}^{\\textit{warp magnitude}}$",
   dataset_ids = dataset_design %>% filter(topology_id %in% names(topologies)) %>% pull(dataset_id),
[…]

### 2018-09-27 scripts/02-metrics/README.Rmd

@@
 ```{r, results = "asis"}
-knit_nest("01-metric_characterisation")
+dynbenchmark::knit_nest("01-metric_characterisation")
 ```
@@
 ```{r, results = "asis"}
-knit_nest("02-metric_conformity")
+dynbenchmark::knit_nest("02-metric_conformity")
 ```
+
+```{r, results = "asis"}
+dynbenchmark::knit_nest("03-aggregation")
+```

### 2018-09-28 README.Rmd

@@
 
+(Note that this preprint is currently outdated, but cannot be updated because the paper is under revision)
+
 ## Experiments
@@
     id = label_long(id),
-    scripts = paste0("[", id, " scripts](", scripts, ")"),
-    results = case_when(is.na(results) ~ "", TRUE ~ paste0("[", id, " results](", results, ")"))
+    scripts = paste0("[\U1F4C4](", scripts, ")"),
+    results = case_when(is.na(results) ~ "", TRUE ~ paste0("[\U1F4CA](", results, ")"))
   ) %>% 
@@
 * [Package](package): An R package with several helper functions.
-* [Data](data): Raw data files required by the scripts.
+* [Raw](raw): Files generated by hand, such as figures and spreadsheets.
 * [Derived](derived): Intermediate data files produced by the scripts. These files or not git committed.
@@
 
+## Methods
+
+All wrapped methods are wrapped as both docker and singularity containers. These can be easily run using [*dyn*methods](https://github.com/dynverse/dynmethods).
+
 ## Benchmarking your own method

### 2018-09-28 scripts/02-metrics/02-metric_conformity/README.Rmd

@@
 
-* If the position of some cells are different than in the gold standard, the score should decrease.
-* If the topology of the network is different than that in the gold standard, the score should not be perfect.
+* If the position of some cells are different than in the reference, the score should decrease.
+* If the topology of the network is different than that in the reference, the score should not be perfect.
 * The more cells are filtered from the trajectory, the more the score should decrease.

### 2018-09-29 manuscript/paper.Rmd
update paper

@@
 ² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
-³ Center for Medical Genetics, Ghent plot_topology_complexity %>% write_rds(result_file("topology_complexity.rdsUniversity Hospital, Ghent, Belgium.  
+³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
 ⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
@@
 # Introduction
+<!-- §1: TI overview -->
 Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017].
@@
 
-<!-- §4: Our study -->
-In this study, we performed a comprehensive evaluation of  →`r n_methods_evaluated`← TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
-→We included methods primarily based← on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
-→We evaluated each method on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) the stability of the predictions across multiple bootstrap replicates.
-Finally, we combined these metrics to provide a set of practical guidelines for users.
-←
-
 ```{r}
@@
 
+→
+
 # Results
+## Method overview
+<!-- Method characterisation -->
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
+We included methods primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
+In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
+In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+The output of each method was converted to this model using different wrapper types (`r ref("fig", "results_summary", "a")`).
 
-## Evaluation of trajectory inference methods
+When required, we also provided prior information to the method, which ranges from "soft" priors, such as start or end cells, to "harder" priors such as the known grouping of the cells (`r ref("fig", "results_summary", "a")`).
 
-<!-- Evaluation overview -->
-In order to make →the outputs from different methods directly comparable to each other←, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
-In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs).
-→We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`).
-Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`)[a].←The data compendium consisted of both synthetic datasets, which offer the most exact →reference trajectory←, and real datasets, which offer the highest biological relevance.
-These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-→For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011].
-For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←
+The main difference between trajectory inference methods is whether the method fixes the topology, and if it is free, what kind of topologies it can detect.
+We defined nine possible types of topology, ranging from the very basic (linear, cyclical and bifurcating) to the most complex (connected and disconnected graphs).
+Most methods either focus on inferring linear trajectories, or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (`r ref("fig", "results_summary", "a")`).
 
 <!-- Overall evaluation results -->
+Each method was evaluated on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions across multiple bootstrap replicates, and (iv) the user friendliness of the provided software and documentation.
+Overall, we found a large diversity across all four evaluation criteria, with only a few methods, such as PAGA, GNG, Slingshot, Projected TSCAN and SCORPIUS, performing well across the board  (`r ref("fig", "results_summary", "b")`).
+We will first discuss each evaluation criterion in detail, after which we conclude with guidelines for method users and future perspectives for method developers.
 ```{r}
 add_fig(
-  fig_path = result_file("overview.pdf", experiment_id = "08-summary"),
-  ref_id = "results_overview", 
-  caption_main = glue::glue("Overview of the results on the evaluation of {n_methods_evaluated} TI methods and {n_control_methods_evaluated} control methods."), 
-  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
-b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
-c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
-d) Performance in the quality control evaluation is highly variabl[b]e, even amongst the highest ranked methods according to the benchmark evaluation.
-Also listed are the quality control scores aggregated according to practicality and the different categories.",
+  fig_path = result_file("results_summary.pdf", experiment_id = "08-summary"),
+  ref_id = "results_summary", 
+  caption_main = glue::glue("A characterisation of the {n_methods_evaluated} methods evaluated in this study, and their overall evaluation results."), 
+  caption_text = "(a) We characterised the methods according to their required priors, the types of inferreable topologies, whether or not the inferred topology is constrained by the algorithm or a parameter.
+The methods are grouped vertically based on the most complex trajectory type they can infer.
+(b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset bootstraps, and the quality control of the tool which implemented the method.
+",
   width = 15,
-  height = 20
+  height = 18
 )
@@
 add_fig(
-  fig_path = result_file("overview.pdf", experiment_id = "08-summary"),
-  ref_id = "results_overview_detailed", 
-  caption_main = glue::glue("Overview of the results on the evaluation of {n_methods_evaluated} TI methods and {n_control_methods_evaluated} control methods."), 
-  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use.
-b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation.
-c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced.
-d) Performance in the quality control evaluation is highly variabl[c]e, even amongst the highest ranked methods according to the benchmark evaluation.
-Also listed are the quality control scores aggregated according to practicality and the different categories.",
+  fig_path = result_file("results_detailed.pdf", experiment_id = "08-summary"),
+  ref_id = "results_detailed", 
+  caption_main = glue::glue("Detailed results of the four main evaluation criteria: benchmarking, scalability, stability and quality control."), 
+  caption_text = "(a) Results of the benchmark across metrics, dataset sources and dataset trajectory types.
+The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
+Also shown is how often a method errored, and what the reasons were for these errors.
+(b) Scalability results, with increasing number of cells and features.
+Shown are predicted times at a fixed set of cells (10.000) or features (10.000).
+(c) Stability results (d) Quality control results across six categories.",
   width = 15,
-  height = 20
+  height = 18
+)
+```
+## Benchmark
+<!-- Benchmark methods -->
+We defined several possible metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`).
+Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which offer the highest biological relevance.
+These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
+Some real datasets were classified as gold if the reference trajectory was defined using other data than the expression data, such as via cellular sorting.
+All other real datasets were classified as silver.
+For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011].
+For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.
+
+<!-- Benchmark overview -->
+We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset `r ref("sfig", "benchmark_interpretation", "a")`.
+Even methods which can detect most of the trajectory types, such as PAGA or GNG, were not the best methods across all trajectory types `r ref("fig", "results_detailed", "a")`.
+The average performance between the different dataset sources was moderately to highly similar correlated, with the performance on all dataset sources being moderately to highly correlated (0.5 - 0.9) with the one on real gold datasets `r ref("sfig", "benchmark_interpretation", "b")`.
+On the other hand, the different metrics frequently disagreed with each other, with Monocle scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches `r ref("fig", "results_detailed", "a")`.
+
+<!-- Topologies-->
+The performance of a method was strongly dependent on the type of trajectory present in the data `r ref("fig", "results_detailed", "a")`.
+Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories.
+This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, while those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (`r ref("sfig", "benchmark_interpretation", "c")`).
+
+<!-- Complementarity between methods -->
+The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
+To test this, we looked for subsets of methods at which the probability of getting a "top" model (defined as having a score at least 95% as the best model) was highest.
+On data which contains a tree trajectory, the use of the best method, RaceID / StemID, would result in getting top model about 50% of the time.
+This probability rapidly increased to up to 90% with the addition of 2 to 5 other methods (`r ref("fig", "complementarity", "a")`).
+We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`).
+When the datasets from different trajectory types were pooled, the number of methods necessary to achieve more than 90% top models increased to more than 8.
+This pooling also resulted in a more diverse set of methods, containing both strictly linear methods such as SCORPIUS, and methods with a broad trajectory type range such as PAGA.
+Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out different methods on their data.
+
+```{r echo=FALSE, results="asis"}
+add_fig(
+  fig_path = result_file("complementarity.rds", experiment_id = "10-benchmark_interpretation"),
+  ref_id = "complementarity", 
+  caption_main = "Complementarity between different TI methods", 
+  caption_text = "",
+  width = 12,
+  height = 6
 )
@@
 
-<!-- Evaluation robustness -->
-
-## Topology inference
-<!-- Inverse relation between edge flip and correlation -->
-In[d] most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`).
-These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type.
-However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
-This indicates that current non-linear methods potentially contain less accurate ordering algorithms.
-A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.
-
-<!-- Topological complexity versus performance -->
-Despite[e] their similar overall performance, the topologies predicted by the top methods differed considerably.
-Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity")`).
-AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data.
-Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology.
-Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively and of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively and of the cases.
-Inferring the correct topology without parameter tuning is therefore still an open challenge.
-
 ```{r echo=FALSE, results="asis"}
-add_fig(
-  fig_path = result_file("topology_complexity.rds", experiment_id = "10-topologies"),
-  ref_id = "topology_complexity", 
-  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the reference trajectory.", 
-  caption_text = "Shown are the distributions of the difference in size between predicted and →reference← topologies.
-A positive difference means that the topology predicted by the method is more complex than the one in the →reference←.",
+add_sfig(
+  fig_path = result_file("benchmark_interpretation.rds", experiment_id = "10-benchmark_interpretation"),
+  ref_id = "benchmark_interpretation", 
+  caption_main = "Robustness of the benchmark", 
+  caption_text = "Shown are the distributions of the difference in size between predicted and reference topologies.
+A positive difference means that the topology predicted by the method is more complex than the one in the reference.",
   width = 12,
-  height = 4
+  height = 15
 )
@@
 
-## Method stability and scalability
-## Method quality control
+## Scalability
+Given the ever increasing 
+## Stability
+←
+## Quality control
+
 While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
@@
 
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("sfig", "qc_overview")`).
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_detailed", "c")` and `r ref("sfig", "qc_overview")`).
 While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers.
@@
 
-We observed no clear relation between method quality and method performance (`r ref("sig", "results_overview")`).
+We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary")`).
 We could also not find any quality aspect which was significantly associated with method performance.
@@
 Using our framework, this figure can be recreated by using only a couple of lines of R code (`r print_url("https://www.github.com/dynverse/dynmethods")`).
-This framework should still be extended to allow easier downstream analyses.
+This framework should still be extended to allow more input data, such as spatial and RNA velocity information [@manno_rnavelocitysingle_2018], and easier downstream analyses.
 In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.←
@@
 → 
-The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[f])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[a])) [@cannoodt_goldstandardsinglecell_].
 
@@
 
+### Aggregation of scores
+To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics.
+This aggregation strategy is explained into more detail in `r ref("snote", "metrics")`.
+
+To make sure easy and difficult datasets have equal influence on the final score, we first normalised the scores on each dataset across the different methods.
+We shifted and scaled the scores to $\sigma = 1$ and $\mu = 0$, and then applied the unit probability density function on these values to get the scores back into the $[0, 1]$ range.
+
+Next, we calculated for each metric, trajectory type and dataset source an average score using an arithmetic mean.
+
+Next, we aggregated the scores across trajectory types and dataset source (real/gold, real/silver, synthetic/dyngen, synthetic/splatter, synthetic/prosstt and synthetic/dyntoy), so that each of these had an equal weight on the final score:
+
+1. Calculate a score for each method, dataset source and trajectory type
 ←
-### Aggregation of scores[g]
-`r ref("sfig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods.
-
-**Step 1:** In order to make the scores produced by the different metrics comparable to one another, across datasets of varying difficulty, the raw scores are transformed per metric per dataset to a $[0,1]$ range as follows:
-
-<!--
-$$
-\begin{aligned}
-\text{scale}(x) &= \text{sigmoid}\left( \frac{x - \textrm{mean}(x)}{\textrm{max}(\textrm{abs}(x - \textrm{mean}(x))) \times 5} \right)
-\end{aligned}
-$$
--->
-
-$$
-\begin{aligned}
-\text{scale}(x) &= \text{sigmoid}\left( \frac{x - \bar{x}}{\textrm{max}(\textrm{abs}(x - \bar{x})) \times 5} \right)\text{, where } \bar{x} \text{ is the arithmetic mean of } x.
-\end{aligned}
-$$
-
-Here, the mean is subtracted from the raw score in order to center the distribution around zero.
-The denominator is used to scale the range of the distribution such that, after sigmoid transformation, the minimum and maximum transformed scores lie close to 0 and/or 1, respectively.
-
-**Step 2:** After transformation, the arithmetic mean is calculated across replicates.
-
-**Step 3:** The normalised scores is combined across all methods and all datasets.
-
-**Step 4:** In order to remove effects from unbalanced trajectory types (e.g. an overrepresentation of linear datasets), the arithmetic mean is first computed per trajectory type.
-
-**Step 5:** An overall score is computed by calculating the arithmetic mean across all trajectory types.
-
-**Step 6:** A ranking of the methods is obtained by calculating the harmonic mean between the different metrics, thus requiring a method to obtain a high score on all three metrics in order to obtain a high overall score.
-
-```{r}
-add_sfig(
-  raw_file("overview_aggregation_v1.svg", "08-summary"), 
-  "aggregation_of_results", 
-  "Aggregation methodology of metric scores.",
-  "",
-  15, 
-  5
-)
-```
 ## Benchmark
@@
 P-values were controlled for multiple testing using Benjamini-Hochberg correction.
+→
+## Scalability[b]
+To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section.
+We up or down-scaled these datasets … OKAY I GIVE UP generate_dataset.R
+
+## [c]Stability
+←
 ## Method quality control
@@
 For calculating the final score, we weighed each of the six categories equally.
-→
-## Scalability[h]
-To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section.
-We up or down-scaled these datasets … OKAY I GIVE UP generate_dataset.R
-
-## [i]Stability
-←
 ## Trajectory types
@@
 
+# Acknowledgements
+We would like to thank all the original authors of the methods for their feedback and improvements on the method wrappers.
+
 <!--
@@
 
-[a]De nomenclatuur in deze figuur moet nog aangepast worden aan de nieuwe maten...
-[b]Als we zoiets zeggen moeten we misschien ook SDs tonen op de plots ?
-[c]Als we zoiets zeggen moeten we misschien ook SDs tonen op de plots ?
-[d]Update when benchmarking results
-[e]Update when benchmarking results
-[f]Update the doi when reuploaded
-[g]TODO
-[h]I will try to do my best in adding this if you don't do it first Robrecht
-[i]I will try to do my best in adding this if you don't do it first Robrecht
+[a]Update the doi when reuploaded
+[b]I will try to do my best in adding this if you don't do it first Robrecht
+[c]I will try to do my best in adding this if you don't do it first Robrecht

### 2018-09-29 scripts/varia/render/render_manuscript.R

@@ -15,6 +15,7 @@ rmarkdown::render("manuscript/supplementary.Rmd", output_format = pdf_manuscript
 
 # browse
 system("/usr/bin/xdg-open manuscript/paper.pdf")
+system("/usr/bin/xdg-open manuscript/paper_changes.pdf")
 system("/usr/bin/xdg-open manuscript/supplementary.pdf")
 
 # upload to google drive

### 2018-09-30 scripts/03-methods/varia/create_github_issue.R
update some docs

@@ -1,4 +1,4 @@
-# code to generate the issues for authors on the dynmethods github
+#' Add & update the method discussion issues on the dynmethods github
 
 library(dynbenchmark)
 library(tidyverse)

### 2018-10-01 README.Rmd

@@
     id = label_long(id),
-    scripts = paste0("[\U1F4C4](", scripts, ")"),
-    results = case_when(is.na(results) ~ "", TRUE ~ paste0("[\U1F4CA](", results, ")"))
+    scripts = paste0("[\U1F4C4\U27A1](", scripts, ")"),
+    results = case_when(is.na(results) ~ "", TRUE ~ paste0("[\U1F4CA\U27A1](", results, ")"))
   ) %>% 

### 2018-10-01 manuscript/paper.Rmd
update paper with scalability assessment

@@
 Instead, the choice toof method →mostly← depends on the trajectory topology present in the data.
-→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce.
+→Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the required prior information and the produced outputs.
 Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←.
@@
 # Results
-## Method overview
+## Trajectory inference methods
 <!-- Method characterisation -->
-In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`).
-We included methods primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`).
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")` and `r ref("sfig", "overview_evaluation_expanded")`).
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-The output of each method was converted to this model using different wrapper types (`r ref("fig", "results_summary", "a")`).
+Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`).
+When required, we also provided prior information to the method, which ranges from "soft" priors, such as start or end cells, to "hard" priors such as the known grouping of the cells (`r ref("fig", "results_summary", "a")`).
 
-When required, we also provided prior information to the method, which ranges from "soft" priors, such as start or end cells, to "harder" priors such as the known grouping of the cells (`r ref("fig", "results_summary", "a")`).
-
+<!-- Topology characterisation -->
 The main difference between trajectory inference methods is whether the method fixes the topology, and if it is free, what kind of topologies it can detect.
@@
 <!-- Overall evaluation results -->
-Each method was evaluated on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions across multiple bootstrap replicates, and (iv) the user friendliness of the provided software and documentation.
+We evaluated each method on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions across multiple bootstrap replicates, and (iv) the user friendliness of the provided software and documentation.
 Overall, we found a large diversity across all four evaluation criteria, with only a few methods, such as PAGA, GNG, Slingshot, Projected TSCAN and SCORPIUS, performing well across the board  (`r ref("fig", "results_summary", "b")`).
@@
   caption_main = glue::glue("A characterisation of the {n_methods_evaluated} methods evaluated in this study, and their overall evaluation results."), 
-  caption_text = "(a) We characterised the methods according to their required priors, the types of inferreable topologies, whether or not the inferred topology is constrained by the algorithm or a parameter.
+  caption_text = "(a) We characterised the methods according to the wrapper type, their required priors (soft = start or end cells, hard = cell grouping or time course), whether the inferred topology is constrained by the algorithm or a parameter and the types of inferreable topologies.
 The methods are grouped vertically based on the most complex trajectory type they can infer.
-(b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset bootstraps, and the quality control of the tool which implemented the method.
+(b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset bootstraps, and the implementation quality.
 ",
@@
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-Some real datasets were classified as gold if the reference trajectory was defined using other data than the expression data, such as via cellular sorting.
+Some real datasets were classified as gold if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or a time course experiment.
 All other real datasets were classified as silver.
@@
 <!-- Benchmark overview -->
-We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset `r ref("sfig", "benchmark_interpretation", "a")`.
-Even methods which can detect most of the trajectory types, such as PAGA or GNG, were not the best methods across all trajectory types `r ref("fig", "results_detailed", "a")`.
-The average performance between the different dataset sources was moderately to highly similar correlated, with the performance on all dataset sources being moderately to highly correlated (0.5 - 0.9) with the one on real gold datasets `r ref("sfig", "benchmark_interpretation", "b")`.
-On the other hand, the different metrics frequently disagreed with each other, with Monocle scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches `r ref("fig", "results_detailed", "a")`.
+We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset (`r ref("sfig", "benchmark_interpretation", "a")`).
+Even methods which can detect most of the trajectory types, such as PAGA or GNG, were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "a")`).
+The average performance between the different dataset sources was moderately to highly similar correlated, with the performance on all dataset sources being moderately to highly correlated (0.5 - 0.9) with the one on real gold datasets (`r ref("sfig", "benchmark_interpretation", "b")`).
+On the other hand, the different metrics frequently disagreed with each other, with Monocle scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "a")`).
 
 <!-- Topologies-->
-The performance of a method was strongly dependent on the type of trajectory present in the data `r ref("fig", "results_detailed", "a")`.
+The performance of a method was strongly dependent on the type of trajectory present in the data (`r ref("fig", "results_detailed", "a")`).
 Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories.
@@
 The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
-To test this, we looked for subsets of methods at which the probability of getting a "top" model (defined as having a score at least 95% as the best model) was highest.
+To test this, we looked for subsets of methods at which percentage of datasets with a "top" model (defined as having a score at least 95% as the best model) was highest.
 On data which contains a tree trajectory, the use of the best method, RaceID / StemID, would result in getting top model about 50% of the time.
-This probability rapidly increased to up to 90% with the addition of 2 to 5 other methods (`r ref("fig", "complementarity", "a")`).
+This rapidly increased to up to 90% with the addition of 2 to 5 other methods (`r ref("fig", "complementarity", "a")`).
 We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`).
-When the datasets from different trajectory types were pooled, the number of methods necessary to achieve more than 90% top models increased to more than 8.
+When the datasets from different trajectory types were pooled, the number of methods necessary to achieve more than 90% top models increased to more than 8, indicating that more methods are necessary when different trajectory types could be present in the data.
 This pooling also resulted in a more diverse set of methods, containing both strictly linear methods such as SCORPIUS, and methods with a broad trajectory type range such as PAGA.
-Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out different methods on their data.
+Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out a diverse set of methods on their data.
 
@@
   caption_main = "Complementarity between different TI methods", 
-  caption_text = "",
+  caption_text = "(a) Here we tested how many methods are needed to still obtain a top model (defined as a model with overall score at least 95\\% of the best model) on datasets containing a tree topology.
+Shown are the percentage of datasets fulfilling this rule, with increasing number of methods.
+At each step, we test which method would be the most optimal (as indicated), and chose this method for all next steps.
+(b) Similar as (a), but now for datasets which contain varying trajectory types.",
   width = 12,
@@
   ref_id = "benchmark_interpretation", 
-  caption_main = "Robustness of the benchmark", 
+  caption_main = "Benchmarking TI methods", 
   caption_text = "Shown are the distributions of the difference in size between predicted and reference topologies.
@@
 ## Scalability
-Given the ever increasing 
+While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million [@svensson_exponentialscalingsinglecell_2018].
+Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features [@cao_jointprofilingchromatin_2018].
+To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets.
+We then modelled the running time using a generalized additive model (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
+As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.
+
+Overall, methods which detect more complex trajectory types also have longer running times, although two exceptions which stand out from the rest: GNG and PAGA (and its derivatives Projected PAGA and PAGA tree).
+Despite their ability to detect all trajectory types, these two methods are still able to accurately detect a trajectory with 100.000 cells and 10.000 features in less than 10 minutes (`r ref("fig", "results_detailed", "b")`).
+Methods with a low running time typically have two defining aspects: they have a sublinear time complexity with respect to the features and/or cells, and adding new cells or features leads to a relatively low increase in time (`r ref("sfig", "scaling", "b")`).
+We found that over half of all methods had a linear or superlinear complexity with respect to the number of cells, which makes it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand of cells (`r ref("sfig", "scaling", "b")`).
+
+```{r echo=FALSE, results="asis"}
+add_sfig(
+  fig_path = result_file("scaling.pdf", experiment_id = "05-scaling"),
+  ref_id = "scaling", 
+  caption_main = "Scalability of TI methods", 
+  caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time based on a generalized additive model (GAM, right).
+(b) Overview of the scalability results of all methods, ordered by their average predicted running time from (a).
+We used to GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve.
+We also used the predicted timings to calculate the average increase per new 100 cells or features.
+Also shown is the average maximal memory usage..",
+  width = 12,
+  height = 15
+)
+```
 ## Stability
@@
 We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary")`).
-We could also not find any quality aspect which was significantly associated with method performance.
 
@@
 <!-- Practical guidelines -->
-Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("guidelines.dynverse.org")`).
+Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("http://guidelines.dynverse.org")`).
 We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
@@
 
-### Aggregation of scores
+### Score aggregation
 To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics.
@@
 
-Next, we calculated for each metric, trajectory type and dataset source an average score using an arithmetic mean.
+After normalisation, we aggregated step by step the scores from different datasets.
+We first aggregated the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores.
+Next, the scores were averaged over different dataset sources, using a arithmetic mean which was weighted based on how much the synthetic and silver scores correlated with the real gold scores.
+Finally, the scores were aggregated over the different trajectory types again using a arithmetic mean.
 
-Next, we aggregated the scores across trajectory types and dataset source (real/gold, real/silver, synthetic/dyngen, synthetic/splatter, synthetic/prosstt and synthetic/dyntoy), so that each of these had an equal weight on the final score:
-
-1. Calculate a score for each method, dataset source and trajectory type
+Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
 ←
-## Benchmark
-### Method execution
+## Method execution
 Each execution of a method on a dataset was performed in a separate task as part of a gridengine job.
@@
 Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.
-### Effect of prior information
-To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test.
-P-values were controlled for multiple testing using Benjamini-Hochberg correction.
-### Effect of algorithm components
-To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score.
-To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test.
-We only investigated algorithm components which were part of at least 4 different methods in our evaluation study.
-P-values were controlled for multiple testing using Benjamini-Hochberg correction.
 →

### 2018-10-02 manuscript/paper.Rmd

@@
 In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
-Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method.
-Instead, the choice toof method →mostly← depends on the trajectory topology present in the data.
-→To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+Our results highlight the complementarity of existing tools, and that there is no "one-fits-all" method.
+Instead, the choice toof method →mostly depends on the dimensions and the trajectory topology of the data.
+To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
@@
 <!-- §3: Problem statement -->
-Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness.
-Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
+Given the diversity in TI methods, an important issue to address is a quantitative assessment of their →performance, scalability, stability and robustness←.
+Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017; @wolf_graphabstractionreconciles_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
 This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea about which method would optimally solve their problem.
@@
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
-In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
+In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of coSome dataset sourcesSome dataset sourcesnnected milestones.
 Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`).
@@
 This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, while those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (`r ref("sfig", "benchmark_interpretation", "c")`).
+This analysis therefore indicated that detecting the right topology is still a difficult task for most of these methods, and that methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
 
@@
 To test this, we looked for subsets of methods at which percentage of datasets with a "top" model (defined as having a score at least 95% as the best model) was highest.
-On data which contains a tree trajectory, the use of the best method, RaceID / StemID, would result in getting top model about 50% of the time.
-This rapidly increased to up to 90% with the addition of 2 to 5 other methods (`r ref("fig", "complementarity", "a")`).
-We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`).
-When the datasets from different trajectory types were pooled, the number of methods necessary to achieve more than 90% top models increased to more than 8, indicating that more methods are necessary when different trajectory types could be present in the data.
-This pooling also resulted in a more diverse set of methods, containing both strictly linear methods such as SCORPIUS, and methods with a broad trajectory type range such as PAGA.
-Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out a diverse set of methods on their data.
+On all datasets, using one method would result in getting top model about 40% of the time.
+This increased to up to 90% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
+This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves, and methods with a broad trajectory type range such as PAGA.
+We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to get a top model on the datasets.
+Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
 
@@
 add_fig(
-  fig_path = result_file("complementarity.rds", experiment_id = "10-benchmark_interpretation"),
+  fig_path = result_file("complementarity.pdf", experiment_id = "10-benchmark_interpretation"),
   ref_id = "complementarity", 
   caption_main = "Complementarity between different TI methods", 
-  caption_text = "(a) Here we tested how many methods are needed to still obtain a top model (defined as a model with overall score at least 95\\% of the best model) on datasets containing a tree topology.
+  caption_text = "(a) Here we tested how many methods are needed to obtain at least one  'top' model (defined as a model with overall score at least 95\\% of the best model) for all datasets.
 Shown are the percentage of datasets fulfilling this rule, with increasing number of methods.
@@
 add_sfig(
-  fig_path = result_file("benchmark_interpretation.rds", experiment_id = "10-benchmark_interpretation"),
+  fig_path = result_file("benchmark_interpretation.pdf", experiment_id = "10-benchmark_interpretation"),
   ref_id = "benchmark_interpretation", 
-  caption_main = "Benchmarking TI methods", 
-  caption_text = "Shown are the distributions of the difference in size between predicted and reference topologies.
+  caption_main = "Benchmarking of trajectory inference methods", 
+  caption_text = "(a) Overall score for all methods and datasets, colored by the source of the datasets.
+(b)  hown are the distributions of the difference in size between predicted and reference topologies.
 A positive difference means that the topology predicted by the method is more complex than the one in the reference.",
@@
   ref_id = "scaling", 
-  caption_main = "Scalability of TI methods", 
+  caption_main = "Scalability of trajectory inference methods", 
   caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time based on a generalized additive model (GAM, right).
@@
   ref_id = "qc_overview", 
-  caption_main = "Overview of the quality control results for every method.", 
+  caption_main = "Quality control of trajectory inference methods.", 
   caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_scoresheet')}).
@@
 In this study, we presented a large scale evaluation of the performance of  `r n_methods_evaluated`  TI methods.
-By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets, which allowed us to highlight the strengths and weaknesses of each method on →different types of trajectories.
-We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on the same datasets.←
+By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets.
+We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on small variations of the datasets.←
 
@@
 This is to make sure the prediction is not biased due to the given parameter setting or the particular algorithms underlying a TI method.
-Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
+→The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods.← Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
 When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
@@
 Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories.
-Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly fail both on real and synthetic data, even if the trajectory can be easily identified in a dimensionality reduction (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
-← Furthermore, higher standards for code assurance, documentation and scalability could help the adoption of these tools across single-cell -omics field.
-To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).
+Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the topology within a datasets, even if the trajectory can be easily identified in a dimensionality reduction (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
+← Furthermore, higher standards for code assurance and documentation could help the adoption of these tools across single-cell -omics field.
+Finally, new tools should focus on an improved scalability, as we found that only a handful of current methods are able to handle datasets with more than 100.000 cells within a reasonable timeframe, making the others possibly obsolete very soon given the rapid technological developments.
+To support the development of these new tools, we provide a series of vignettes on how to wrap a method within our common interface, and evaluate the method on the different measures proposed in this study at `r print_url("evaluate.dynverse.org")`.
 
@@
 We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
-Nonetheless, we also found that it is possible to define for almost every method a relatively large subset of datasets on which this method on average performs the best.
-These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, given that there is a high risk that a certain new methodology would not improve upon the state-of-the-art.
+Nonetheless, we want to avoid that a race to be the first on the ranking would stifle creativity, given that there is a high risk that a certain new methodology would not improve upon the state-of-the-art.
 Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution, or provide a unique insight in specific use cases.
@@
 As a final filter, we removed both all-zero genes and cells until convergence.
-## Evaluation metrics
+## Benchmark metrics
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
@@
 If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
-The environment variable `R_MAX_NUM_DLLS` set to 500.
-The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust.
-Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.
 →
-## Scalability[b]
+## Scalability
 To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section.
-We up or down-scaled these datasets … OKAY I GIVE UP generate_dataset.R
+We up- and downscaled these datasets between 10 to 100.000 cells and 10 to 100.000 features, while never going higher than 1.000.000 values in total.
+To generate new cells or features, we first generated a 10-nearest neighbour graph of both the cells and features from the expression space.
+For every new cell or feature, we used a linear combination of one to three existing cells or features, where each cell or features was given a weight sampled from $\textit{U}(0,1)$.
 
-## [c]Stability
+We ran each method on each dataset for maximally 1 hour and gave each process 10GB of memory.
+To determine the running time of each method, we started the timer right after data loading, and the loading of any packages, and stopped the clock before post-processing and saving of the output.
+Pre- and postprocessing steps specific to a method, such as dimensionality reduction and gene filtering, were included in the time.
+To estimate the maximal memory usage, we used the max_vmem value from the qacct command provided by the gridengine cluster.
+We acknowledge that these values are very noisy, and the averages provided in this study are therefore only rough estimates.
+
+The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using generalized additive models.
+We used thin-plate regression spline as smoothing basis, with $log_{10}|cells|$ and $log_{10}|features|$ as predictor variables, and fitted this model using the gam as implemented in the R mgcv package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
+
+To classify the time complexity of each method, we predicted the running time at 10.000 features (cells) with increasing number of cells (features) from 100 to 100.000, with steps of 100.
+A method was classified as constant / low slope if on average an addition of 100 cells would increase the running time with less than 0.1 seconds.
+To classify as sublinear, linear or superlinear, we scaled the times between 0 and 1, and classified the method as sublinear if the running time was on average 5% above the diagonal, superlinear if on average 5% under the diagonal, and linear otherwise.
+
+## Stability[b]
 ←
@@
 For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork.
-In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →the exceptions being being DPT (limited to bifurcating trajectories), and STEMNET and FateID (unable to return linear trajectories).←
+In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.←
 
 # Acknowledgements
-We would like to thank all the original authors of the methods for their feedback and improvements on the method wrappers.
+We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers.
 
@@
 [a]Update the doi when reuploaded
-[b]I will try to do my best in adding this if you don't do it first Robrecht
-[c]I will try to do my best in adding this if you don't do it first Robrecht
+[b]Todo once this experiment is finished

### 2018-10-02 scripts/05-scaling/README.Rmd
update readmes with github_markdown_nested format
update readmes to nested github document

@@
 
-Each method is run on randomly generated data with increasing gene set and cell set sizes, and the execution times and memory usages are modelled using linear models.
+Each method is run on down- and upscaled datasets with increasing gene set and cell set sizes, and the execution times and memory usages are modelled using thin plate regressions splines within a generalised additivate model.
 

### 2018-10-02 manuscript/paper.Rmd
update paper

@@
 While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
-To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scorsheet")`).
+To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scorseheet")`).
 
@@
 We ran each method on each dataset for maximally 1 hour and gave each process 10GB of memory.
-To determine the running time of each method, we started the timer right after data loading, and the loading of any packages, and stopped the clock before post-processing and saving of the output.
+To determine the running time of each method, we started the timer right after data loading and the loading of any packages, and stopped the clock before post-processing and saving of the output.
 Pre- and postprocessing steps specific to a method, such as dimensionality reduction and gene filtering, were included in the time.
 To estimate the maximal memory usage, we used the max_vmem value from the qacct command provided by the gridengine cluster.
-We acknowledge that these values are very noisy, and the averages provided in this study are therefore only rough estimates.
+We acknowledge however that these values are very noisy, and the averages provided in this study are therefore only rough estimates.
 
 The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using generalized additive models.
-We used thin-plate regression spline as smoothing basis, with $log_{10}|cells|$ and $log_{10}|features|$ as predictor variables, and fitted this model using the gam as implemented in the R mgcv package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
+We used thin-plate regression spline as smoothing basis, with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the gam as implemented in the R mgcv package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
 

### 2018-10-03 manuscript/paper.Rmd
update paper

@@
 <!-- Overall evaluation results -->
-We evaluated each method on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions across multiple bootstrap replicates, and (iv) the user friendliness of the provided software and documentation.
+We evaluated each method on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions after subsampling the datasets, and (iv) the user friendliness of the provided software and documentation.
 Overall, we found a large diversity across all four evaluation criteria, with only a few methods, such as PAGA, GNG, Slingshot, Projected TSCAN and SCORPIUS, performing well across the board  (`r ref("fig", "results_summary", "b")`).
@@
 The methods are grouped vertically based on the most complex trajectory type they can infer.
-(b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset bootstraps, and the implementation quality.
+(b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset subsamples, and the implementation quality.
 ",
@@
   ref_id = "complementarity", 
-  caption_main = "Complementarity between different TI methods", 
+  caption_main = "Complementarity between different trajectory inference methods", 
   caption_text = "(a) Here we tested how many methods are needed to obtain at least one  'top' model (defined as a model with overall score at least 95\\% of the best model) for all datasets.
@@
 At each step, we test which method would be the most optimal (as indicated), and chose this method for all next steps.
-(b) Similar as (a), but now for datasets which contain varying trajectory types.",
+(b) Similar as (a), but now for datasets which contain varying trajectory types.
+For this figure, we did not include any methods requiring prior information.",
   width = 12,
@@
   caption_text = "(a) Overall score for all methods and datasets, colored by the source of the datasets.
-(b)  hown are the distributions of the difference in size between predicted and reference topologies.
+(b) Bias in the overall score towards trajectory types.
+(c) Similarity between the overall scores of all dataset sources with the real gold datasets.
+(d) Distributions of the difference in size between predicted and reference topologies.
 A positive difference means that the topology predicted by the method is more complex than the one in the reference.",
@@
 We also used the predicted timings to calculate the average increase per new 100 cells or features.
-Also shown is the average maximal memory usage..",
+Also shown is the average maximal memory usage.",
   width = 12,
@@
 add_fig(
-  fig_path = result_file("example_predictions.rds", experiment_id = "11-example_predictions"),
+  fig_path = result_file("example_predictions.pdf", experiment_id = "11-example_predictions"),
   ref_id = "example_predictions", 
   caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
-  caption_text = "",
-  width = 13,
+  caption_text = "(a) The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
+(b) The top methods applied on a dataset containing a bifurcating trajectory of ",
+  width = 14,
   height = 8
@@
 Finally, new tools should focus on an improved scalability, as we found that only a handful of current methods are able to handle datasets with more than 100.000 cells within a reasonable timeframe, making the others possibly obsolete very soon given the rapid technological developments.
-To support the development of these new tools, we provide a series of vignettes on how to wrap a method within our common interface, and evaluate the method on the different measures proposed in this study at `r print_url("evaluate.dynverse.org")`.
+To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method and on the different measures proposed in this study at `r print_url("http://dynbenchmark.dynverse.org")`.
 
@@
 
-<!-- Components -->
-The selected methods were dissected into a set of algorithmic components.
-Each component can have a significant impact on the performance, scalability, and output.
-Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016].
-Components are frequently shared between different algorithms.
-For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamics_2014], is shared by almost half of the methods we evaluated.
 →
 ### Method wrappers
-Each method was wrapped within docker and singularity containers (available at `r print_url("methods.dynverse.org")`).
+Each method was wrapped within docker and singularity containers (available at `r print_url("http://dynmethods.dynverse.org")`).
 These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
@@
 During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
-If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”.
 →
+## Complementarity
+To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score is equal or higher than 95% of the best overall score for that particular dataset.
+We then calculated for every method the weighted percentage of datasets which fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method.
+We iteratively added new methods until all methods were selected.
+For this analysis, we did not include any methods which require prior information, and only included methods which can detect at least one of the trajectory types present in the datasets.
 ## Scalability
@@
 ## Stability[b]
+We ran each method on a selection of datasets.
+This subset consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
+We subsampled each dataset 10 times with 95% of the original cells and 95% of the original features, such that each method should produce slightly different results.
 ←
@@
 
-Each aspect was further assigned to one or more applications, based on whether it influenced the user friendliness of the tool (such as code availability, good documentation or contains plotting functions) and/or the developer friendliness of the tool (such as unit testing, inline documentation and a clear licensing).
-
 Each quality aspect received a weight depending on how frequently it was found in several papers and online sources which discuss tool quality (`r ref("stable", "qc_scoresheet")`).
-This was to make sure that more important aspects, such as the open source availability of the method, outweighed other aspects, such as the availability of a graphical user interface.
+This was to make sure that more important aspects, such as the open source availability of the method, outweighed other less important aspects, such as the availability of a graphical user interface.
 Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_scoresheet")`).
@@
 We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`).
-These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical.
-A disconnected directed graph is a graph in which only one edge can exist between two nodes.
-A directed graph is a disconnected graph in which all nodes are (indirectly) connected.
-A directed acyclic graph is a graph containing no cycles.
-A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1).
-A rooted binary tree is a rooted tree with every node having maximally two outgoing edges.
+These trajectory types start from the most general trajectory type, a disconnected graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical.
+A disconnected graph is a graph in which only one edge can exist between two nodes.
+A (connected) graph is a disconnected graph in which all nodes are connected.
+An acyclic graph is a graph containing no cycles.
+A tree is an acyclic graph containing no convergences (no nodes with in-degree higher than 1).
+A binary tree is a rooted tree with every node having maximally two outgoing edges.
 A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one.
@@
 A bifurcation is a multifurcation in which only one node has a degree equal to 3.
-A directed linear graph is a graph in which no node has a degree larger than 3.
-Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle.
-For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork.
+A linear topology is a graph in which no node has a degree larger than 3.
+Finally, a cycle is a connected graph in which no node has a degree larger than 3 but contains one cycle.
 In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.←
@@
 # Acknowledgements
-We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers.
+
+# Author contributions
 

### 2018-11-12 manuscript/supplementary.Rmd
rename title in supplementary material

@@
 ---
-title: "Supplementary material for: \nA comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
+title: "Supplementary material for: \nAccuracy, scalability, robustness and usability of single-cell trajectory inference methods"
 output: 

### 2018-11-12 manuscript/supplementary.Rmd
change title again

@@
 ---
-title: "Supplementary material for: \nAccuracy, scalability, robustness and usability of single-cell trajectory inference methods"
+title: "Supplementary material for: \nA comparison of single-cell trajectory inference methods"
 output: 

### 2018-11-12 manuscript/supplementary.Rmd
update supplementary

--- 2018-09-25 manuscript/supplementary.Rmd
@@
 ```{r, results="asis"}
-sfigs %>% pmap(dynbenchmark:::plot_fig, ref_type = "sfig") %>% modify(paste0, "\n\n") %>% walk(cat)
+sfigs %>% 
+  left_join(refs %>% select(ref_id, name), "ref_id") %>% 
+  arrange(name) %>% 
+  select(-name) %>% 
+  pmap(dynbenchmark:::plot_fig, ref_type = "sfig") %>% modify(paste0, "\n\n") %>% walk(cat)
 ```
+
+\clearpage
 
@@
 
-```{r setup, include=FALSE}
-# source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/setup.R"))
-# 
-# # knitr options
-# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
-# knitr::opts_chunk$set(
-#   collapse = TRUE,
-#   warning = FALSE,
-#   message = FALSE,
-#   echo = FALSE,
-#   comment = "#>",
-#   fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
-# )
-# 
-# list2env(read_rds(derived_file("sfigs_stables.rds", "paper")), environment())
+\clearpage
+
+# Supplementary note 1
+
+```{r}
+# dynbenchmark::knit_nest("../results/02-metrics/")
 ```
-
-```{r "supplementary", echo=FALSE, results='asis'}
-# source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/sfigs_stables.R"))
-```

### 2018-11-12 manuscript/supplementary.Rmd

@@
 ---
-title: "Supplementary material for: \nA comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
+title: "Supplementary material for: \nA comparison of single-cell trajectory inference methods"
 output: 

### 2018-11-12 manuscript/paper.Rmd
update manuscript

@@
 ---
-title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
+title: "Accuracy, scalability, robustness and usability of single-cell trajectory inference methods[a][b]"
 output: dynbenchmark::pdf_manuscript
@@
 In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
-Our results highlight the complementarity of existing tools, and that there is no "one-fits-all" method.
-Instead, the choice toof method →mostly depends on the dimensions and the trajectory topology of the data.
-To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
-We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
+Our results highlight the complementarity of existing tools, and that there is no generally best performing method.
+Instead, the choice of method →mostly depends on the dataset characteristics and the trajectory topology present.
+To assist users in selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+We hope to spearhead such developments, and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
 
@@
 Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
-The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected graphs [@wolf_graphabstractionreconciles_2017].
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow identifying more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected graphs [@wolf_graphabstractionreconciles_2017].
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the required prior information and the produced outputs.
-Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←.
-Early TI methods typically fixed the topology algorithmically (such as linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016].
+→Indeed, in several repositories listing single-cell tools,  such as `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← While each method has its own unique set of characteristics in terms of underlying algorithm, the required prior information and the produced outputs, two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←.
+Early TI methods typically fixed the topology algorithmically (e.g.
+linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating trajectories [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016].
 These methods therefore mainly focused on correctly ordering the cells along the fixed topology.
@@
 <!-- §3: Problem statement -->
-Given the diversity in TI methods, an important issue to address is a quantitative assessment of their →performance, scalability, stability and robustness←.
+Given the diversity in TI methods, an important issue to address is a quantitative assessment of their →performance, scalability, robustness and usability←.
 Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017; @wolf_graphabstractionreconciles_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
@@
 c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity.
-d) →We defined four metrics each assessing the similarity of a different aspect of the trajectory: the topology, the branch assignment, the cell positions and the important features.←",
+d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory: the topology, the branch assignment, the cell positions and the important features.←",
    width = 13,
@@
 
-```{r}
-add_sfig(
-  fig_path = raw_file("overview_evaluation_expanded_v1.svg", ""),
-  ref_id = "overview_evaluation_expanded",
-  caption_main = "Extended overview of our evaluation pipeline.", 
-  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data.
-We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters.
-Using several similarity metrics, we compared the →reference trajectory← of the real and synthetic datasets with the inferred trajectories.
-We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.",
-  width = 15, 
-  height = 7.5
-)
-```
-
 →
-
 # Results
@@
 <!-- Method characterisation -->
-In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")` and `r ref("sfig", "overview_evaluation_expanded")`).
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
-In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of coSome dataset sourcesSome dataset sourcesnnected milestones.
+In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
 Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`).
-When required, we also provided prior information to the method, which ranges from "soft" priors, such as start or end cells, to "hard" priors such as the known grouping of the cells (`r ref("fig", "results_summary", "a")`).
+When required, we also provided prior information to the method, the severity of which varies (`r ref("fig", "results_summary", "a")`).
+Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[c][d]
 
 <!-- Topology characterisation -->
-The main difference between trajectory inference methods is whether the method fixes the topology, and if it is free, what kind of topologies it can detect.
-We defined nine possible types of topology, ranging from the very basic (linear, cyclical and bifurcating) to the most complex (connected and disconnected graphs).
+The main difference between trajectory inference methods is whether the method fixes the topology, and if it does not, what kind of topologies it can detect.
+We defined nine possible types of topologies, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
 Most methods either focus on inferring linear trajectories, or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (`r ref("fig", "results_summary", "a")`).
@@
 <!-- Overall evaluation results -->
-We evaluated each method on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions after subsampling the datasets, and (iv) the user friendliness of the provided software and documentation.
-Overall, we found a large diversity across all four evaluation criteria, with only a few methods, such as PAGA, GNG, Slingshot, Projected TSCAN and SCORPIUS, performing well across the board  (`r ref("fig", "results_summary", "b")`).
-We will first discuss each evaluation criterion in detail, after which we conclude with guidelines for method users and future perspectives for method developers.
+We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
+genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the quality of the tool in terms of software, documentation and the manuscript.
+Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (`r ref("fig", "results_summary", "b")`).
+
+Unfortunately, few methods seem to have been given consideration to the scalability or the usability of the tool.
+While less important than the accuracy, with increasing dataset sizes the scalability of the tool is becoming increasingly important, and a high usability score is important to ensure correct usage of the tool by a user.
+
+Methods in which the topology of the model was fixed by the algorithm or by a parameter obtain a much higher stability score.
+As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[e][f]
+
+We will discuss each evaluation criterion in more detail, after which we conclude with guidelines for method users and future perspectives for method developers.
 ```{r}
@@
   caption_main = glue::glue("A characterisation of the {n_methods_evaluated} methods evaluated in this study, and their overall evaluation results."), 
-  caption_text = "(a) We characterised the methods according to the wrapper type, their required priors (soft = start or end cells, hard = cell grouping or time course), whether the inferred topology is constrained by the algorithm or a parameter and the types of inferreable topologies.
+  caption_text = "(a) We characterised the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies.
 The methods are grouped vertically based on the most complex trajectory type they can infer.
-(b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset subsamples, and the implementation quality.
-",
+(b) The overall results of the evaluation on four criteria: benchmarking using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples, and quality of the implementation.",
   width = 15,
@@
   caption_main = glue::glue("Detailed results of the four main evaluation criteria: benchmarking, scalability, stability and quality control."), 
-  caption_text = "(a) Results of the benchmark across metrics, dataset sources and dataset trajectory types.
+  caption_text = glue::glue("(a) The names of the methods.
+(b) Results of the benchmark across metrics, dataset sources and dataset trajectory types.
 The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
-Also shown is how often a method errored, and what the reasons were for these errors.
-(b) Scalability results, with increasing number of cells and features.
-Shown are predicted times at a fixed set of cells (10.000) or features (10.000).
-(c) Stability results (d) Quality control results across six categories.",
+(c) Predicted execution times for varying numbers of cells and features (# cells × # features).
+Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
+(d) Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method.
+(e) Usability scores of the tool and corresponding manuscript, grouped per category."),
   width = 15,
@@
 ```
-## Benchmark
-<!-- Benchmark methods -->
+
+```{r}
+add_sfig(
+  fig_path = result_file("results_suppfig.pdf", experiment_id = "08-summary"),
+  ref_id = "results_suppfig", 
+  caption_main = glue::glue(""), 
+  caption_text = glue::glue(""),
+  width = 30,
+  height = 18,
+  integrate = FALSE
+)
+```
+## Accuracy
+<!-- Accuracy methods -->
 We defined several possible metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`).
-Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which offer the highest biological relevance.
+Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose four metrics each assessing a different aspect of a trajectory (`r ref("fig", "ti_evaluation_overview", "d")`): the topology (Hamming-Ipsen-Mikhailov, `r label_metric("him")`), the quality of clustering of the assignment of cells to branches (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`).
+The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance.
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-Some real datasets were classified as gold if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or a time course experiment.
-All other real datasets were classified as silver.
-For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011].
-For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.
+Some real datasets were classified as "gold standard" if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing [@tian_scrnaseqmixologybetter_2018].
+All other real datasets were classified as "silver standard".
+For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation[@schaffter_genenetweaversilicobenchmark_2011].
+For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties[@zappia_splattersimulationsinglecell_2017].
 
@@
 We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset (`r ref("sfig", "benchmark_interpretation", "a")`).
-Even methods which can detect most of the trajectory types, such as PAGA or GNG, were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "a")`).
-The average performance between the different dataset sources was moderately to highly similar correlated, with the performance on all dataset sources being moderately to highly correlated (0.5 - 0.9) with the one on real gold datasets (`r ref("sfig", "benchmark_interpretation", "b")`).
-On the other hand, the different metrics frequently disagreed with each other, with Monocle scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "a")`).
+Even methods which can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "b")`).
+The overall score between the different dataset sources was moderately to highly correlated (0.5 - 0.9) with the scores on real datasets with a gold standard (`r ref("sfig", "benchmark_interpretation", "b")`), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data.
+On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "b")`).
 
 <!-- Topologies-->
-The performance of a method was strongly dependent on the type of trajectory present in the data (`r ref("fig", "results_detailed", "a")`).
-Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories.
-This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, while those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (`r ref("sfig", "benchmark_interpretation", "c")`).
-This analysis therefore indicated that detecting the right topology is still a difficult task for most of these methods, and that methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
+The performance of a method was strongly dependent on the type of trajectory present in the data (`r ref("fig", "results_detailed", "b")`).
+Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories (`r ref("sfig", "benchmark_interpretation", "c")`).
+This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, while those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (`r ref("sfig", "benchmark_interpretation", "d")`).
+This analysis therefore indicates that detecting the right topology is still a difficult task for most of these methods, because methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
 
@@
 The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
-To test this, we looked for subsets of methods at which percentage of datasets with a "top" model (defined as having a score at least 95% as the best model) was highest.
-On all datasets, using one method would result in getting top model about 40% of the time.
+To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
+A top model in this case was defined as a model with an overall score at least 95% as the best model.
+On all datasets, using one method would result in getting a top model in about 40% of the time.
 This increased to up to 90% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
 This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves, and methods with a broad trajectory type range such as PAGA.
-We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to get a top model on the datasets.
-Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
+We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
+Altogether, this shows that there is considerable complementarity between the different methods, and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
+Moreover, it also opens up the possibilities for new ensemble methods which utilize this complementarity.
 
@@
   caption_main = "Complementarity between different trajectory inference methods", 
-  caption_text = "(a) Here we tested how many methods are needed to obtain at least one  'top' model (defined as a model with overall score at least 95\\% of the best model) for all datasets.
-Shown are the percentage of datasets fulfilling this rule, with increasing number of methods.
-At each step, we test which method would be the most optimal (as indicated), and chose this method for all next steps.
-(b) Similar as (a), but now for datasets which contain varying trajectory types.
+  caption_text = "(a) We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with overall score at least 95\\% of the best model) when applied to all datasets.
+(b) The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types.
 For this figure, we did not include any methods requiring prior information.",
-  width = 12,
+  width = 14,
   height = 6
@@
   caption_text = "(a) Overall score for all methods and datasets, colored by the source of the datasets.
+(c) Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard.
 (b) Bias in the overall score towards trajectory types.
-(c) Similarity between the overall scores of all dataset sources with the real gold datasets.
 (d) Distributions of the difference in size between predicted and reference topologies.
@@
 To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets.
-We then modelled the running time using a generalized additive model (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
-As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.
+We then modelled the running time using a generalized additive model[@wood_thinplateregression_2003] (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
+As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[g][h]
 
-Overall, methods which detect more complex trajectory types also have longer running times, although two exceptions which stand out from the rest: GNG and PAGA (and its derivatives Projected PAGA and PAGA tree).
+Overall, methods which detect more complex trajectory types also have longer running times, although two exceptions stand out: GNG and PAGA (and its derivatives Projected PAGA and PAGA tree).
 Despite their ability to detect all trajectory types, these two methods are still able to accurately detect a trajectory with 100.000 cells and 10.000 features in less than 10 minutes (`r ref("fig", "results_detailed", "b")`).
 Methods with a low running time typically have two defining aspects: they have a sublinear time complexity with respect to the features and/or cells, and adding new cells or features leads to a relatively low increase in time (`r ref("sfig", "scaling", "b")`).
-We found that over half of all methods had a linear or superlinear complexity with respect to the number of cells, which makes it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand of cells (`r ref("sfig", "scaling", "b")`).
+We found that over half of all methods had a linear or superlinear complexity with respect to the number of cells, which makes it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).
 
@@
 ## Stability
+It is not only important that a method is able to infer an accurate model, but also that it produces more-or-less the same model when the input data is perturbed slightly.
+In order to test the stability of each method, we executed each method on 10 different bootstraps of the datasets (95% cells, 95% features), and calculated the average similarity between each pair of models (`r ref("fig", "results_detailed", "d")`).
+
 ←
-## Quality control
+## Implementation quality and user friendliness
 
-While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018].
-To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scorseheet")`).
+While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how easy it is for a biological user to use [@_doesyourcode_2018].
+We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scorseheet")`).
 
@@
 In this study, we presented a large scale evaluation of the performance of  `r n_methods_evaluated`  TI methods.
-By using a common structure and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than a two-hundred datasets.
-We also assessed several other important quality measures, such as quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on small variations of the datasets.←
+By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than two-hundred datasets.
+We also assessed several other important quality measures, such as the quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on small variations of the datasets.←
 
@@
 Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("http://guidelines.dynverse.org")`).
-We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data.
+We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should currently still be primarily driven by the anticipated trajectory topology in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
@@
 Several other factors will also impact the choice of methods, such as the dimensions of the dataset, and the prior information which is available.
-These factors and several others can all be changed in our interactive app (`r print_url("http://dynverse.org")`).
-This app can also be used to explore the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←
+These factors and several others can all be dynamically explored in our interactive app (`r print_url("https://dynverse.org")`).
+This app can also be used to query the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←
 
@@
 add_fig(
-  fig_path = result_file("guidelines.svg", experiment_id = "09-guidelines"),
+  fig_path = result_file("guidelines.pdf", experiment_id = "09-guidelines"),
   ref_id = "user_guidelines", 
   caption_main = "Practical guidelines for method users.", 
-  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
+  caption_text = "As the performance of a method mostly depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data.
 We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
@@
 →The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods.← Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
-When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system.
-When a more complex topology is produced, →this could indicate that the underlying biology is much more complex than expected by the user.←
+When the expected topology is confirmed using such a method, it provides additional evidence to the user.
+When a more complex topology is produced, →this could indicate that the underlying biology is much more complex than anticipated by the user.←
 
 <!-- A common trajectory analysis framework -->
-Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as find genes important for the trajectory, network inference [@aibar_scenicsinglecellregulatory_2017] or finding modules of genes [@saelens_comprehensiveevaluationmodule_2018].
-Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods →on some a linear, a tree, a cyclic and a disconnected graph dataset (`r ref("fig", "example_predictions")`).
-Using our framework, this figure can be recreated by using only a couple of lines of R code (`r print_url("https://www.github.com/dynverse/dynmethods")`).
-This framework should still be extended to allow more input data, such as spatial and RNA velocity information [@manno_rnavelocitysingle_2018], and easier downstream analyses.
+Critical to the broad applicability of TI methods is the standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as finding genes important for the trajectory, network inference [@aibar_scenicsinglecellregulatory_2017] or finding modules of genes [@saelens_comprehensiveevaluationmodule_2018].
+Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods →on datasets containing a linear, tree, cyclic and disconnected graph topology (`r ref("fig", "example_predictions")`).
+Using our framework, this figure can be recreated using only a couple of lines of R code (`r print_url("https://methods.dynverse.org")`).
+In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information [@manno_rnavelocitysingle_2018], and easier downstream analyses.
 In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.←
@@
   caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
-  caption_text = "(a) The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
-(b) The top methods applied on a dataset containing a bifurcating trajectory of ",
+  caption_text = "Trajectories inferred by each method were projected to a common dimensionality reduction.
+(a) The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
+(b) The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts.
+(c) A synthetic dataset generated by dyntoy, containing four disconnected trajectories.
+(d) A synthetic dataset generated by dyngen, containing a cyclic trajectory..",
   width = 14,
@@
 Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (as we illustrate with two datasets in `r ref("fig", "example_predictions", "a and b")`).
-Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories.
-Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the topology within a datasets, even if the trajectory can be easily identified in a dimensionality reduction (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
-← Furthermore, higher standards for code assurance and documentation could help the adoption of these tools across single-cell -omics field.
-Finally, new tools should focus on an improved scalability, as we found that only a handful of current methods are able to handle datasets with more than 100.000 cells within a reasonable timeframe, making the others possibly obsolete very soon given the rapid technological developments.
-To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method and on the different measures proposed in this study at `r print_url("http://dynbenchmark.dynverse.org")`.
+However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analysing single-cell -omics datasets with complex trajectories.
+Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory can be easily identified in a dimensionality reduction (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
+← Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell -omics field.
+Finally, new tools should be designed to scale well with the increasing number of cells and features.
+We found that only a handful of current methods are able to handle datasets with more than 10.000 cells within a reasonable timeframe, making the others possibly obsolete very soon given the rapid technological developments.
+To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method and on the different measures proposed in this study at `r print_url("https://benchmark.dynverse.org")`.
 
@@
 → 
-The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[a])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[i])) [@cannoodt_goldstandardsinglecell_].
 
@@
 ### Method wrappers
-Each method was wrapped within docker and singularity containers (available at `r print_url("http://dynmethods.dynverse.org")`).
+To make easy to run each method in a reproducible manner, each method was wrapped within docker and singularity containers (available at `r print_url("https://methods.dynverse.org")`).
 These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
 For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
-This script reads in the input data (as described below), runs the method, and outputs the files required to construct a trajectory (as described below).
+This script reads in the input data , runs the method, and outputs the files required to construct a trajectory.
 We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).
@@
 For cyclic datasets, a random cell was chosen.
-* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
-* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
+* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only incoming edges.
+* **# end states** Number of terminal states. Number of milestones with only incoming edges.
 * **Grouping** For each cell a label to which state/cluster/branch it belongs.
@@
 This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
-(ii) The milestone percentages contains, for each cell, its position between milestones, and sums for each cell to one.
-(iii) →The regions of delayed commitment, connections between three or more milestones.
+(ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one.
+(iii) →The regions of delayed commitment, which defines connections between three or more milestones.
 These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←
@@
 * **Type 1, direct:** `r conversion("trajectory")`.
-The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone.
-→In some cases, this← indicates that additional transformations were required for the method not covered by any of the following output formats.
-→Some methods (`r conversion("branch_trajectory")`) returned a branch network instead of a milestone network, and this conversion was done by calculating the line graph of the branch network.←
+The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell it is given to what extent it belongs to a milestone.
+→In some cases, this← indicates that additional transformations were required for the method, not covered by any of the following output formats.
+→Some methods (`r conversion("branch_trajectory")`) returned a branch network instead of a milestone network, and this network was converted by calculating the line graph of the branch network.←
 * **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`.
@@
 →This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$), with regions of delayed commitment between all milestones.
-The milestone percentage of a cell to one of the outer milestones was equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .←
+The milestone percentage of a cell to one of the outer milestones was equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$.
+The milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .←
 * **Type 5, cluster assignment:** `r conversion("cluster_graph")`.
@@
 The method returned a milestone network, and a dimensionality reduction of the cells and milestones.
-The cells were projected onto the closest nearest segment, thus determining the cells position along the milestone network.
+The cells were projected to the closest nearest segment, thus determining the cells' position along the milestone network.
 →If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell.
@@
 →
-### Control methods
-We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the reference trajectory, and **shuffle**, which will place the cells randomly on the reference milestone network.
 ### Off-the-shelf methods
@@
 The cells are then projected onto the edges of the MST.
+* **GNG**: Dimensionality reduction with PCA, followed by modelling of the topology with a growing neural gas using the R gng package (`r print_url("https://github.com/rcannood/gng")`).
+The cells are then projected onto the edges of the network.
 ←
+
+## Trajectory types
+We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`).
+These trajectory types start from the most general trajectory type, a disconnected graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical.
+A disconnected graph is a graph in which only one edge can exist between two nodes.
+A (connected) graph is a disconnected graph in which all nodes are connected.
+An acyclic graph is a graph containing no cycles.
+A tree is an acyclic graph containing no convergences (no nodes with in-degree higher than 1).
+A binary tree is a rooted tree with every node having maximally two outgoing edges.
+A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one.
+A multifurcation is a rooted tree in which only one node has a degree larger than one.
+A bifurcation is a multifurcation in which only one node has a degree equal to 3.
+A linear topology is a graph in which no node has a degree larger than 3.
+Finally, a cycle is a connected graph in which no node has a degree larger than 3 but contains one cycle.
+In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.←
 
@@
 Whenever possible, we preferred to start from the raw counts data.
-These raw counts were all normalised and filtered using a common pipeline, discussed later.
+These raw counts were all normalised and filtered using a common pipeline, as discussed later.
 →Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
-In the end, we included `r sum(list_datasets()$source == "real")` for this evaluation.←
+In the end, we included `r nrow(list_datasets("real"))` for this evaluation.←
 
@@
 These parameters were then given to the simulators when applicable, as described for each simulator below.
-To select a diverse set of $k$ reference real datasets, we clustered all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (implemented in the R cluster package) to generate synthetic data.
+Not every real dataset was selected to serve as a reference for a synthetic dataset.
+Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (implemented in the R cluster package) to generate synthetic data.
 ←
@@
 One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell choses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
+The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
 One driver of bifurcation seems to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009].
@@
 →To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
-Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating converging), (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
+Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
 The structure of these module networks is available at `r print_url("https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks")`.←
@@
 
-We created a set of topology generators:
+We created a set of topology generators, were $B(n, p)$ denotes a binomial distribution, and $U(a, b)$ denotes a uniform distribution:
 
@@
 ### The Hamming-Ipsen-Mikhailov metric
-The Hamming-Ipsen-Mikhailov (HIM) [@jurman_HIM_Glocal_Metric_2015] uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
+The Hamming-Ipsen-Mikhailov (HIM) metric [@jurman_HIM_Glocal_Metric_2015] uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
 It is a linear combination of the normalised Hamming distance, which gives an indication of the differences in edge lengths, and the normalised Ipsen-Mikhailov distance, which assesses the similarity in degree distributions.
@@
 ### The F1 between branch assignments
-To compare branch assignment, we borrowed the $F_1$ metric from the field of expression biclustering [@saelens_comprehensiveevaluationmodule_2018].
+To compare branch assignment, we used an $F_1$ score, also used used for comparing biclustering methods [@saelens_comprehensiveevaluationmodule_2018].
 To calculate this metric, we first calculated the similarity of all pairs of branches between the two trajectories using the Jaccard similarity.
@@
 ### The correlation between important features
-The `r label_metric("featureimp_wcor")` assesses whether similar dynamically differentially expressed features are found between the prediction and model.
+The `r label_metric("featureimp_wcor")` assesses whether similar dynamically differentially expressed features are found between the prediction and the model.
 To calculate this metric, we used Random Forest regression (implemented in the R ranger package), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
-We then extract feature importance values for each feature, and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset as to give more weight to differentially expressed features then non-differentially expressed features.
-As hyperparameters we set the number of trees to 10000, and the number of features on which can be split to 1% of all available features.
+We then extract feature importance values for each feature, and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset as to give more weight to large differences.
+As hyperparameters we set the number of trees to 10000, and the number of features on which to split to 1% of all available features.
 We illustrate this metric and assess the effect of its hyperparameters in `r ref("snote", "metrics")`.
@@
 
-Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
+Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[j][k][l][m]
 ←
@@
 Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task.
-During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution.
+During the execution of a method on a dataset, if the time limit (>1h) or memory limit (16GB) was exceeded, or an error was produced, a zero score was returned for that execution.
 →
@@
 The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using generalized additive models.
-We used thin-plate regression spline as smoothing basis, with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the gam as implemented in the R mgcv package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
+As smoothing bases, we used a thin-plate regression spline [@wood_thinplateregression_2003] , with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the gam as implemented in the R mgcv package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
 
@@
 
-## Stability[b]
-We ran each method on a selection of datasets.
-This subset consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
+## Stability[n][o]
+We ran each method on a selection of datasets; which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
 We subsampled each dataset 10 times with 95% of the original cells and 95% of the original features, such that each method should produce slightly different results.
@@
 We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`).
-The goal of this quality control is in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
+The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
 The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items.
@@
 The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
-Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset in which the usefulness of the method was shown, and the scope of method evaluation in the paper.
+Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of datasets in which the usefulness of the method was shown, and the scope of method evaluation in the paper.
 
@@
 This was to make sure that more important aspects, such as the open source availability of the method, outweighed other less important aspects, such as the availability of a graphical user interface.
-Within each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_scoresheet")`).
+For each aspect, we also assigned a weight to the individual questions being investigated (`r ref("stable", "qc_scoresheet")`).
 For calculating the final score, we weighed each of the six categories equally.
-## Trajectory types
-We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`).
-These trajectory types start from the most general trajectory type, a disconnected graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical.
-A disconnected graph is a graph in which only one edge can exist between two nodes.
-A (connected) graph is a disconnected graph in which all nodes are connected.
-An acyclic graph is a graph containing no cycles.
-A tree is an acyclic graph containing no convergences (no nodes with in-degree higher than 1).
-A binary tree is a rooted tree with every node having maximally two outgoing edges.
-A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one.
-A multifurcation is a rooted tree in which only one node has a degree larger than one.
-A bifurcation is a multifurcation in which only one node has a degree equal to 3.
-A linear topology is a graph in which no node has a degree larger than 3.
-Finally, a cycle is a connected graph in which no node has a degree larger than 3 but contains one cycle.
-In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.←
 
 # Acknowledgements
-
+We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers. This study was supported by Fond Wetenschappelijk Onderzoek (W.S. and R.C.).
 # Author contributions
+W.S., R.C., H.T.
+and Y.S.
+designed the study, performed the experiments and analysed the data.
+W.S., R.C., and H.T.
+implemented software packages.
+W.S., R.C., Y.S.
+and H.T.
+prepared the manuscript.
+Y.S.
+supervised the project.
+# Competing financial interests
+The authors declare no competing financial interests
 
@@
 
-[a]Update the doi when reuploaded
-[b]Todo once this experiment is finished
+[a]?
+[b]we changed the title :)
+[c]Try to rewrite a little bi
+[d]somehing like his?
+[e]TODO: rework enumeration into a paragraph
+[f]something like this?
+[g](Not shown), or include in Supp Fig 4?
+[h]Yeah we don't include this yet, should be in sup fig 4
+[i]Update when final
+[j]This might give the impression of being a bit "overengineered".
+Do we also show somewhere how much the resulting ranking using this aggregation would differ from a very simple "mean" ranking of the individual metrics/datasets ?
+[k]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
+[l]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
+[m]I adapted the text a little bit to better explain the 'why'
+[n]Todo once this experiment is finished
+[o]Good enough?

### 2018-11-12 manuscript/supplementary.Rmd
update paper & supplementary

@@
   select(-name) %>% 
-  pmap(dynbenchmark:::plot_fig, ref_type = "sfig") %>% modify(paste0, "\n\n") %>% walk(cat)
+  pmap(dynbenchmark:::plot_fig) %>% modify(paste0, "\n\n") %>% walk(cat)
 ```
@@
 ```{r, results="asis"}
-stables %>% pmap(dynbenchmark:::show_table, ref_type = "stable") %>% modify(paste0, "\n\n") %>% walk(cat)
+stables %>% 
+  left_join(refs %>% select(ref_id, name), "ref_id") %>% 
+  arrange(name) %>% 
+  select(-name) %>% 
+  pmap(dynbenchmark:::show_table) %>% modify(paste0, "\n\n") %>% walk(cat)
 ```

### 2018-11-12 manuscript/paper.Rmd

@@
 ---
-title: "Accuracy, scalability, robustness and usability of single-cell trajectory inference methods[a][b]"
+title: "A comparison of single-cell trajectory inference methods"
 output: dynbenchmark::pdf_manuscript
@@
 
-`www.github.com/dynverse/dynbenchmark/commit/`r git2r::revparse_single(".","HEAD")$sha``
+`www.github.com/dynverse/dynbenchmark/tree/`r git2r::revparse_single(".","HEAD")$sha``
 ```{r}
@@
 n_tools_evaluated <- nrow(filter(tools, evaluated))
-n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("tool", "adaptation", "offtheshelf")))
+n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% "tool" | id == "paga_tree"))
 n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
@@
 When required, we also provided prior information to the method, the severity of which varies (`r ref("fig", "results_summary", "a")`).
-Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[c][d]
+Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[a][b]
 
@@
 Methods in which the topology of the model was fixed by the algorithm or by a parameter obtain a much higher stability score.
-As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[e][f]
+As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[c][d]
 
-We will discuss each evaluation criterion in more detail, after which we conclude with guidelines for method users and future perspectives for method developers.
+We will discuss each evaluation criterion in more detail (`r ref("fig", "results_summary")` and `r ref("sfig", "results")`), after which we conclude with guidelines for method users and future perspectives for method developers.
 ```{r}
@@
   caption_main = glue::glue("Detailed results of the four main evaluation criteria: benchmarking, scalability, stability and quality control."), 
-  caption_text = glue::glue("(a) The names of the methods.
+  caption_text = glue::glue("(a) The names of the methods, ordered as in {ref('fig', 'results_summary')}.
 (b) Results of the benchmark across metrics, dataset sources and dataset trajectory types.
@@
   fig_path = result_file("results_suppfig.pdf", experiment_id = "08-summary"),
-  ref_id = "results_suppfig", 
+  ref_id = "results", 
   caption_main = glue::glue(""), 
@@
 We then modelled the running time using a generalized additive model[@wood_thinplateregression_2003] (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
-As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[g][h]
+As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[e][f]
 
@@
 
+## Usability
 ←
-## Implementation quality and user friendliness
+While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how easy it is for a biological user to use [@_doesyourcode_2018].
+We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scoresheet")`).
 
-While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how easy it is for a biological user to use [@_doesyourcode_2018].
-We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scorseheet")`).
-
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_detailed", "c")` and `r ref("sfig", "qc_overview")`).
+We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`).
 While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers.
@@
   ref_id = "qc_overview", 
-  caption_main = "Quality control of trajectory inference methods.", 
-  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_scoresheet')}).
+  caption_main = "Usability of trajectory inference methods.", 
+  caption_text = glue::glue("Shown is the score given for each method on every item from our usability score sheet ({ref('stable', 'qc_scoresheet')}).
 Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
@@
   "qc_scoresheet",
-paste0("→Scoring sheet for method quality control.
+paste0("→Scoring sheet for assessing usability of trajectory inference methods.
 Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.←"),
@@
 → 
-The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[i])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[g])) [@cannoodt_goldstandardsinglecell_].
 
@@
 This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-b) The specific transformations performed by each of the wrapper functions is shown in more detail.",
+b) Illustrations of the specific transformations performed by each of the wrapper functions.",
   13, 
@@
 add_stable(
-  result_file("metadata.xlsx","01-datasets/01-real"), 
+  result_file("real_datasets.xlsx","01-datasets/01-real"), 
   "datasets", 
@@
 
-Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[j][k][l][m]
+Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[h][i][j][k]
 ←
@@
 
-## Stability[n][o]
+## Stability[l][m]
 We ran each method on a selection of datasets; which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
@@
 ←
-## Method quality control
+## Usability
 
-We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`).
+We created a transparent scoring scheme to quantify the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`).
 The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
@@
 
-[a]?
-[b]we changed the title :)
-[c]Try to rewrite a little bi
-[d]somehing like his?
-[e]TODO: rework enumeration into a paragraph
-[f]something like this?
-[g](Not shown), or include in Supp Fig 4?
-[h]Yeah we don't include this yet, should be in sup fig 4
-[i]Update when final
-[j]This might give the impression of being a bit "overengineered".
+[a]Try to rewrite a little bi
+[b]somehing like his?
+[c]TODO: rework enumeration into a paragraph
+[d]something like this?
+[e](Not shown), or include in Supp Fig 4?
+[f]Yeah we don't include this yet, should be in sup fig 4
+[g]Update when final
+[h]This might give the impression of being a bit "overengineered".
 Do we also show somewhere how much the resulting ranking using this aggregation would differ from a very simple "mean" ranking of the individual metrics/datasets ?
-[k]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
-[l]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
-[m]I adapted the text a little bit to better explain the 'why'
-[n]Todo once this experiment is finished
-[o]Good enough?
+[i]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
+[j]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
+[k]I adapted the text a little bit to better explain the 'why'
+[l]Todo once this experiment is finished
+[m]Good enough?

### 2018-11-12 manuscript/paper.Rmd

@@
 
-`www.github.com/dynverse/dynbenchmark/tree/`r git2r::revparse_single(".","HEAD")$sha``
+Commit: `r print_url(paste0("www.github.com/dynverse/dynbenchmark/tree/", git2r::revparse_single(".","HEAD")$sha))`
 ```{r}
@@
 
-In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
+In this study, we evaluated a total of →`r n_methods_evaluated`← m[a]ethods in terms of cellular →ordering, topology, scalability, and user friendliness←.
 Our results highlight the complementarity of existing tools, and that there is no generally best performing method.
@@
   ref_id = "ti_evaluation_overview", 
-  caption_main = "Overview of several key aspects of the evaluation.", 
-  caption_text = "a) A schematic overview of our evaluation pipeline.
+  caption_main = "Overview of several key aspects of the evaluation[b].", 
+  caption_text = glue::glue("a) A schematic overview of our evaluation pipeline.
 b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
 c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity.
-d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory: the topology, the branch assignment, the cell positions and the important features.←",
-   width = 13,
-  height = 8, 
+d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory.
+The Hamming-Ipsen-Mikhailov (HIM) score will assess the similarity between the two topologies, taking into account differences in edge lengths and degree distributions.
+The {label_metric('F1_branches')} assesses the similarity of the assignment of cells onto branches.
+The {label_metric('correlation')} quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances.
+Finally, {label_metric('featureimp_wcor')} quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.←"),
+   width = 14,
+  height = 8.5, 
 )
@@
 <!-- Method characterisation -->
-In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated`[c] TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
@@
 When required, we also provided prior information to the method, the severity of which varies (`r ref("fig", "results_summary", "a")`).
-Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[a][b]
+Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[d][e][f]
 
@@
 <!-- Overall evaluation results -->
-We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
+We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real`[g] real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
 genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the quality of the tool in terms of software, documentation and the manuscript.
@@
 
-Unfortunately, few methods seem to have been given consideration to the scalability or the usability of the tool.
+Unfortunately, only few methods seem to have been given consideration to the scalability or the usability of the tool.
 While less important than the accuracy, with increasing dataset sizes the scalability of the tool is becoming increasingly important, and a high usability score is important to ensure correct usage of the tool by a user.
@@
 Methods in which the topology of the model was fixed by the algorithm or by a parameter obtain a much higher stability score.
-As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[c][d]
+As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[h][i]
 
@@
 <!-- Accuracy methods -->
-We defined several possible metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`).
-Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose four metrics each assessing a different aspect of a trajectory (`r ref("fig", "ti_evaluation_overview", "d")`): the topology (Hamming-Ipsen-Mikhailov, `r label_metric("him")`), the quality of clustering of the assignment of cells to branches (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`).
+We defined several metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`).
+Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose four metrics each assessing a different aspect of a trajectory (`r ref("fig", "ti_evaluation_overview", "d")`): the topology (Hamming-Ipsen-Mikhailov, `r label_metric("him")`), the quality of the assignment of cells to branches (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the accuracy of the differentially expressed features along the trajectory (`r label_metric("featureimp_wcor")`).
 The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance.
@@
 Even methods which can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "b")`).
-The overall score between the different dataset sources was moderately to highly correlated (0.5 - 0.9) with the scores on real datasets with a gold standard (`r ref("sfig", "benchmark_interpretation", "b")`), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data.
+The overall score between the different dataset sources was moderately to highly correlated (0.5 - 0.9) with the scores on real datasets containing a gold standard (`r ref("sfig", "benchmark_interpretation", "b")`), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data.
 On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "b")`).
@@
 To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
-A top model in this case was defined as a model with an overall score at least 95% as the best model.
+A top model in this case was defined as a model with an overall score of at least 95% as the best model.
 On all datasets, using one method would result in getting a top model in about 40% of the time.
 This increased to up to 90% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
-This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves, and methods with a broad trajectory type range such as PAGA.
+This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves[j][k], and methods with a broad trajectory type range such as PAGA.
 We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
@@
 To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets.
-We then modelled the running time using a generalized additive model[@wood_thinplateregression_2003] (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
-As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[e][f]
+We modelled the running time and memory usage using a Shape Constrained Additive Model[@pya_shapeconstrainedadditive_2015] (`r ref("sfig", "scaling", "a")`).
+We assessed whether the time / memory complexity was sublinear, linear ,quadratic, or superquadratic with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
+As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[l][m][n]
 
-Overall, methods which detect more complex trajectory types also have longer running times, although two exceptions stand out: GNG and PAGA (and its derivatives Projected PAGA and PAGA tree).
-Despite their ability to detect all trajectory types, these two methods are still able to accurately detect a trajectory with 100.000 cells and 10.000 features in less than 10 minutes (`r ref("fig", "results_detailed", "b")`).
-Methods with a low running time typically have two defining aspects: they have a sublinear time complexity with respect to the features and/or cells, and adding new cells or features leads to a relatively low increase in time (`r ref("sfig", "scaling", "b")`).
-We found that over half of all methods had a linear or superlinear complexity with respect to the number of cells, which makes it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).
+Overall, methods which detected more complex trajectory types also had longer running times, except for PAGA and PAGA Tree (`r ref("fig", "results_detailed", "c")`).
+We looked at the running times of some common use cases, such as datasets with a lot of cells (100 thousand to 1 million) but only a limited number of features (100 - 1000), which is usually the case for droplet based datasets, and datasets with only a limited number of cells but a lot of features, relevant for flow and mass cytometry datasets[o].
+Only a handful of methods had an acceptable running time on these datasets, with PAGA tree, GrandPrix and several linear methods able to finish in under 10 minutes on a dataset with one hundred thousand cells and one thousand features.
+Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (`r ref("sfig", "scaling", "b")`).
+We found that more than half of all methods had a linear or superlinear complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).
 
@@
 (b) Overview of the scalability results of all methods, ordered by their average predicted running time from (a).
-We used to GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve.
+We used the GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve.
 We also used the predicted timings to calculate the average increase per new 100 cells or features.
@@
 While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers.
-Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
 
@@
 This is represented in the plot as the height on the y-axis.
-Top: Average QC score for each method.
+Top: Average usability score for each method.
 Right: The average score of each quality control item.
@@
 → 
-The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[g])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[p])) [@cannoodt_goldstandardsinglecell_].
 
@@
 
-Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[h][i][j][k]
+Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[q][r][s][t]
 ←
@@
 
-## Stability[l][m]
+## Stability[u][v]
 We ran each method on a selection of datasets; which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
@@
 
-[a]Try to rewrite a little bi
-[b]somehing like his?
-[c]TODO: rework enumeration into a paragraph
-[d]something like this?
-[e](Not shown), or include in Supp Fig 4?
-[f]Yeah we don't include this yet, should be in sup fig 4
-[g]Update when final
-[h]This might give the impression of being a bit "overengineered".
+[a]Check that this is now without the control methods
+[b]Mentions "55" methods instead of the "45" mentioned in the introduction. Also, the things written in the figure b are very small.
+[c]check consistency with Figure 1
+[d]Try to rewrite a little bi
+[e]somehing like his?
+[f]Hehe ;). Something like this?
+[g]Does not fit with the number in Figure 1
+[h]TODO: rework enumeration into a paragraph
+[i]something like this?
+[j]Isn't this a control method ?
+[k]yes, should be removed!
+[l](Not shown), or include in Supp Fig 4?
+[m]Yeah we don't include this yet, should be in sup fig 4
+[n]TODO: Will be added as sup fig
+[o]I would not mention flow/mass cytoemtry here
+[p]Update when final
+[q]This might give the impression of being a bit "overengineered".
 Do we also show somewhere how much the resulting ranking using this aggregation would differ from a very simple "mean" ranking of the individual metrics/datasets ?
-[i]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
-[j]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
-[k]I adapted the text a little bit to better explain the 'why'
-[l]Todo once this experiment is finished
-[m]Good enough?
+[r]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
+[s]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
+[t]I adapted the text a little bit to better explain the 'why'
+[u]Todo once this experiment is finished
+[v]Good enough?

### 2018-11-12 manuscript/paper.Rmd
update manuscript

@@
 
-In this study, we evaluated a total of →`r n_methods_evaluated`← m[a]ethods in terms of cellular →ordering, topology, scalability, and user friendliness←.
+In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
 Our results highlight the complementarity of existing tools, and that there is no generally best performing method.
@@
   ref_id = "ti_evaluation_overview", 
-  caption_main = "Overview of several key aspects of the evaluation[b].", 
+  caption_main = "Overview of several key aspects of the evaluation.", 
   caption_text = glue::glue("a) A schematic overview of our evaluation pipeline.
@@
 <!-- Method characterisation -->
-In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated`[c] TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
@@
 Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`).
-When required, we also provided prior information to the method, the severity of which varies (`r ref("fig", "results_summary", "a")`).
-Namely, having to provide a start cell or an end cell is relatively easy to do and will only introduce moderate bias into the model, whereas having to provide a clustering or time course information will likely introduce heavy bias into the model.[d][e][f]
+When required, we also provided prior information to the method, which can range from weak priors which are easier to acquire, such as the start cells, to strong priors, such as a known grouping of cells, which are much harder to know a priori and which can potentially introduce a large bias into the analysis (`r ref("fig", "results_summary", "a")`).
 
@@
 <!-- Overall evaluation results -->
-We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real`[g] real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
+We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
 genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the quality of the tool in terms of software, documentation and the manuscript.
 Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (`r ref("fig", "results_summary", "b")`).
-
-Unfortunately, only few methods seem to have been given consideration to the scalability or the usability of the tool.
-While less important than the accuracy, with increasing dataset sizes the scalability of the tool is becoming increasingly important, and a high usability score is important to ensure correct usage of the tool by a user.
-
-Methods in which the topology of the model was fixed by the algorithm or by a parameter obtain a much higher stability score.
-As topology-free methods have more degrees of freedom, the stability of the method is also typically lower ─ with a few exceptions.[h][i]
-
 We will discuss each evaluation criterion in more detail (`r ref("fig", "results_summary")` and `r ref("sfig", "results")`), after which we conclude with guidelines for method users and future perspectives for method developers.
@@
 A top model in this case was defined as a model with an overall score of at least 95% as the best model.
-On all datasets, using one method would result in getting a top model in about 40% of the time.
-This increased to up to 90% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
-This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods such as periodic principal curves[j][k], and methods with a broad trajectory type range such as PAGA.
+On all datasets, using one method would result in getting a top model in about 27% of the time.
+This increased to up to 76% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
+This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
 We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
@@
 We assessed whether the time / memory complexity was sublinear, linear ,quadratic, or superquadratic with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
-As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.[l][m][n]
+As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method, depending on to what extent the execution of a method succeeded during the scalability experiments.[a][b][c]
 
 Overall, methods which detected more complex trajectory types also had longer running times, except for PAGA and PAGA Tree (`r ref("fig", "results_detailed", "c")`).
-We looked at the running times of some common use cases, such as datasets with a lot of cells (100 thousand to 1 million) but only a limited number of features (100 - 1000), which is usually the case for droplet based datasets, and datasets with only a limited number of cells but a lot of features, relevant for flow and mass cytometry datasets[o].
+We looked at the running times of the most immediate use case: datasets with a lot of cells (100 thousand to 1 million) but only a limited number of features (100 - 1000), which is usually the case for droplet based datasets, and datasets with only a limited number of cells but a lot of features.
 Only a handful of methods had an acceptable running time on these datasets, with PAGA tree, GrandPrix and several linear methods able to finish in under 10 minutes on a dataset with one hundred thousand cells and one thousand features.
 Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (`r ref("sfig", "scaling", "b")`).
-We found that more than half of all methods had a linear or superlinear complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).
+We found that more than half of all methods had a quadratic or superquadratic complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).
 
@@
   caption_main = "Scalability of trajectory inference methods", 
-  caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time based on a generalized additive model (GAM, right).
+  caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time (right).
 (b) Overview of the scalability results of all methods, ordered by their average predicted running time from (a).
-We used the GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve.
-We also used the predicted timings to calculate the average increase per new 100 cells or features.
-Also shown is the average maximal memory usage.",
+We predicted execution times and memory usage for each method with increasing number of features or cells, and used these values to classify each method into sublinear, linear, quadratic and superquadratic based on the shape of the curve.",
   width = 12,
@@
 However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analysing single-cell -omics datasets with complex trajectories.
-Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory can be easily identified in a dimensionality reduction (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
+Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be  identified using a dimensionality reduction method (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
 ← Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell -omics field.
@@
 → 
-The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[p])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[d])) [@cannoodt_goldstandardsinglecell_].
 
@@
   methods_evaluated %>% 
-    filter(source %in% c("tool", "offtheshelf", "adaptation")) %>%
+    filter(source %in% c("tool")) %>%
     filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
@@
 More information on how each method was wrapped can be found within the comments of each wrapper script, available at `r print_url("https://github.com/dynverse/dynmethods")`.←
-
-→
-### Off-the-shelf methods
-For baseline performance, we added several "off-the-shelf" TI methods which can be run using just a couple of commands in R.←
-
-* **Component 1:** This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
-Although conceptually very simple, it has been used in certain studies [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
-* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA, and uses this angle as pseudotime for a cyclical trajectory.
-→
-* **Periodic PrinCurve:** Periodic principal curves, using the R princurve package
-* **MST:** Dimensionality reduction with PCA, followed by clustering using the R mclust package, after which the clusters are connected using a minimal spanning tree.
-The cells are then projected onto the edges of the MST.
-* **GNG**: Dimensionality reduction with PCA, followed by modelling of the topology with a growing neural gas using the R gng package (`r print_url("https://github.com/rcannood/gng")`).
-The cells are then projected onto the edges of the network.
-←
 
@@
 Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
-The structure of these module networks is available at `r print_url("https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks")`.←
+The structure of these module networks is available at[e] `r print_url("https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks")`.←
 
@@
 
-After normalisation, we aggregated step by step the scores from different datasets.
+Since there is a bias in dataset source and trajectory type (e.g.
+there are many more linear datasets), we aggregated the scores per method and dataset in multiple steps.
 We first aggregated the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores.
@@
 
-Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.[q][r][s][t]
+Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
 ←
@@
 Pre- and postprocessing steps specific to a method, such as dimensionality reduction and gene filtering, were included in the time.
-To estimate the maximal memory usage, we used the max_vmem value from the qacct command provided by the gridengine cluster.
+To estimate the maximal memory usage, we used the `max_vmem` value from the `qacct` command provided by the gridengine cluster.
 We acknowledge however that these values are very noisy, and the averages provided in this study are therefore only rough estimates.
@@
 
-To classify the time complexity of each method, we predicted the running time at 10.000 features (cells) with increasing number of cells (features) from 100 to 100.000, with steps of 100.
+To classify the time complexity of each method, we predicted the running time at 10.000 features with increasing number of cells from 100 to 100.000, with steps of 100.
 A method was classified as constant / low slope if on average an addition of 100 cells would increase the running time with less than 0.1 seconds.
-To classify as sublinear, linear or superlinear, we scaled the times between 0 and 1, and classified the method as sublinear if the running time was on average 5% above the diagonal, superlinear if on average 5% under the diagonal, and linear otherwise.
+To classify as sublinear, linear, quadratic, or superquadratic, we scaled the times between 0 and 1, and classified the method as sublinear if the running time was on average 5% above the diagonal, superlinear if on average 5% under the diagonal, and linear otherwise.
 
-## Stability[u][v]
-We ran each method on a selection of datasets; which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
-We subsampled each dataset 10 times with 95% of the original cells and 95% of the original features, such that each method should produce slightly different results.
+## Stability
+In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different.
+However, running the method multiple times on the same input data would not be the ideal approach to assess the stability of a method, given that a lot of tools artificially are artificially deterministic by resetting the pseudorandom number generator (for example, using the set.seed function in R).
+To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
+We subsampled each dataset 10 times with 95% of the original cells and 95% of the original features.
+We ran every method on each of the bootstraps, and assessed the stability by calculating the benchmarking scores between each pair of subsequent models (run $i$ is compared to run $i+1$).
+For the `r label_metric("correlation")` and `r label_metric("F1_branches")`, we only used the intersection between the cells of two datasets, while the intersection of the features was used for the `r label_metric("featureimp_wcor")`.
 ←
@@
 and Y.S.
-designed the study, performed the experiments and analysed the data.
+designed the study.
+W.S.
+and R.C.
+performed the experiments and analysed the data.
 W.S., R.C., and H.T.
@@
 
-[a]Check that this is now without the control methods
-[b]Mentions "55" methods instead of the "45" mentioned in the introduction. Also, the things written in the figure b are very small.
-[c]check consistency with Figure 1
-[d]Try to rewrite a little bi
-[e]somehing like his?
-[f]Hehe ;). Something like this?
-[g]Does not fit with the number in Figure 1
-[h]TODO: rework enumeration into a paragraph
-[i]something like this?
-[j]Isn't this a control method ?
-[k]yes, should be removed!
-[l](Not shown), or include in Supp Fig 4?
-[m]Yeah we don't include this yet, should be in sup fig 4
-[n]TODO: Will be added as sup fig
-[o]I would not mention flow/mass cytoemtry here
-[p]Update when final
-[q]This might give the impression of being a bit "overengineered".
-Do we also show somewhere how much the resulting ranking using this aggregation would differ from a very simple "mean" ranking of the individual metrics/datasets ?
-[r]No we don't do this yet. If we do this, perhaps we should just add a "benchmarking robustness" supplementary note.
-[s]The results do change when we simply use a mean, but then they are very biased, eg. because 1/4 of our datasets are linear, the linear methods perform well..
-[t]I adapted the text a little bit to better explain the 'why'
-[u]Todo once this experiment is finished
-[v]Good enough?
+[a](Not shown), or include in Supp Fig 4?
+[b]Yeah we don't include this yet, should be in sup fig 4
+[c]TODO: Will be added as sup fig
+[d]Update when final
+[e]the following link renders as verbatim code!

### 2018-11-13 manuscript/paper.Rmd

@@
 n_tools_evaluated <- nrow(filter(tools, evaluated))
-n_methods_evaluated <- nrow(filter(methods, evaluated, source %in% "tool" | id == "paga_tree"))
-n_control_methods_evaluated <- nrow(filter(methods, evaluated, source %in% c("control")))
+n_methods_evaluated <- nrow(filter(methods, evaluated, source == "tool"))
+n_control_methods_evaluated <- nrow(filter(methods, evaluated, source == "control"))
 n_datasets_real <- nrow(list_datasets("real"))
@@
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable [@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
 
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→Indeed, in several repositories listing single-cell tools,  such as `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← While each method has its own unique set of characteristics in terms of underlying algorithm, the required prior information and the produced outputs, two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←.
+→Indeed, in several repositories listing single-cell tools,  such as `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies they can detect←.
 Early TI methods typically fixed the topology algorithmically (e.g.
@@
 Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017; @wolf_graphabstractionreconciles_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
-This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea about which method would optimally solve their problem.
+This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which method would optimally solve their problem.
 Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
@@
   caption_text = glue::glue("(a) The names of the methods, ordered as in {ref('fig', 'results_summary')}.
-(b) Results of the benchmark across metrics, dataset sources and dataset trajectory types.
+(b) Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types.
 The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
@@
   ref_id = "results", 
-  caption_main = glue::glue(""), 
-  caption_text = glue::glue(""),
+  caption_main = glue::glue("Results from the evaluation, for all methods and across all evaluation criteria."), 
+  caption_text = glue::glue("(a) We characterised the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies.
+The methods are grouped vertically based on the most complex trajectory type they can infer.
+(b) The overall results of the evaluation on four criteria: benchmarking using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples, and quality of the implementation.
+(c) Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types.
+The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
+(d) Predicted execution times and memory usage for varying numbers of cells and features (# cells × # features).
+Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
+(e) Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method.
+(f) Usability scores of the tool and corresponding manuscript, grouped per category."),
   width = 30,
@@
 On all datasets, using one method would result in getting a top model in about 27% of the time.
-This increased to up to 76% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
+This increased to up to 74% with the addition of 6 other methods (`r ref("fig", "complementarity", "a")`).
 This resulted in a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
@@
   ref_id = "benchmark_interpretation", 
-  caption_main = "Benchmarking of trajectory inference methods", 
+  caption_main = "Accuracy of trajectory inference methods.", 
   caption_text = "(a) Overall score for all methods and datasets, colored by the source of the datasets.
@@
 To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets.
-We modelled the running time and memory usage using a Shape Constrained Additive Model[@pya_shapeconstrainedadditive_2015] (`r ref("sfig", "scaling", "a")`).
-We assessed whether the time / memory complexity was sublinear, linear ,quadratic, or superquadratic with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`).
+We modelled the running time and memory usage using a Shape Constrained Additive Model [@pya_shapeconstrainedadditive_2015] (`r ref("sfig", "scaling", "a")`).
 As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method, depending on to what extent the execution of a method succeeded during the scalability experiments.[a][b][c]
 
-Overall, methods which detected more complex trajectory types also had longer running times, except for PAGA and PAGA Tree (`r ref("fig", "results_detailed", "c")`).
-We looked at the running times of the most immediate use case: datasets with a lot of cells (100 thousand to 1 million) but only a limited number of features (100 - 1000), which is usually the case for droplet based datasets, and datasets with only a limited number of cells but a lot of features.
-Only a handful of methods had an acceptable running time on these datasets, with PAGA tree, GrandPrix and several linear methods able to finish in under 10 minutes on a dataset with one hundred thousand cells and one thousand features.
+We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with 10 thousand cells and 10 thousand features (`r ref("fig", "results_detailed", "c")`), the size of a typical droplet based single-cell omics dataset [@svensson_exponentialscalingsinglecell_2018].
+Running times increased further with increasing number of cells, with only a handful of graph/tree methods running within a day on a million of cells (PAGA, PAGA Tree, Monocle DDRTree, Stemnet and GrandPrix).
+Some methods, such as Monocle DDRTree and GrandPrix, also suffered from unsatisfactory running times when given a high number of features.
+
 Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (`r ref("sfig", "scaling", "b")`).
 We found that more than half of all methods had a quadratic or superquadratic complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (`r ref("sfig", "scaling", "b")`).
+
+We also assessed the memory requirements of each methods (`r ref("sfig", "results", "c")`).
+Most methods had reasonable memory requirements for modern workstations or computer clusters (≤12GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or high number of features.
+Notably, the memory requirements were very high for several methods, such as RaceID/StemID, pCreode, Monocle DDRTree and SLICE, as the number of cells or features increases.
+
+Altogether, the scalability analysis indicated that the dimensions of the data should be important factor in the choice of method and that method development should pay more attention to maintaining reasonable running times and memory usage.
 
@@
   ref_id = "scaling", 
-  caption_main = "Scalability of trajectory inference methods", 
+  caption_main = "Scalability of trajectory inference methods.", 
   caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time (right).
@@
 ## Stability
-It is not only important that a method is able to infer an accurate model, but also that it produces more-or-less the same model when the input data is perturbed slightly.
-In order to test the stability of each method, we executed each method on 10 different bootstraps of the datasets (95% cells, 95% features), and calculated the average similarity between each pair of models (`r ref("fig", "results_detailed", "d")`).
+It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data.
+In order to test the stability of each method, we executed each method on 10 different subsamples of the datasets (95% of the cells, 95% of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (`r ref("fig", "results_detailed", "d")`).
 
+Given that the trajectories of methods which fix the topology either algorithmically or through a parameter is already very constrained, it is to be expected that such methods tend to generate very stable results.
+Nonetheless, some fixed topology methods still produced slightly more stable results, such as SCORPIUS and MATCHER for linear methods, and MFA for multifurcating methods.
+Stability was much more diverse among methods with a free topology.
+Slingshot produced more stable models than PAGA (Tree), which in turn produced more stable results than pCreode and Monocle DDRTree.
 ## Usability
-←
+
 While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how easy it is for a biological user to use [@_doesyourcode_2018].
 We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scoresheet")`).
+Important to note here is that we did not include a substantial set of tools in our analyses, because of issues with installation, code availability and executability on a freely available platform (which excludes MATLAB), all discussed at our repository (`r print_url('http://methods.dynverse.org/issues?q=label:\"won\'t+wrap\"')`).
+Installation issues seem to be quite general in bioinformatics [@mangul_comprehensiveanalysisusability_2018], and the trajectory inference field is no exception.
 
-We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`).
+We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and basic code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`).
 While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers.
 Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
+Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration.
 
@@
 ```
+
+←
 
@@
 → 
-The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533[d])) [@cannoodt_goldstandardsinglecell_].
+The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1443566](https://doi.org/10.5281/zenodo.1443566)) [@cannoodt_goldstandardsinglecell_].
 
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
-→The discussions on why these methods were excluded can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:\"won't wrap\"")`.
+→The discussions on why these methods were excluded can be found at `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"won\'t wrap"')`.
 In the end, we included  `r n_tools_evaluated` tools in the evaluation, constituting `r n_methods_evaluated` methods.←
@@
 ```{r}
-methods_evaluated <- read_rds(result_file("methods_evaluated.rds", "03-methods"))
+methods_evaluated <- load_methods()
 special_methods <- c("mfa", "dpt", "slice", "sincell", "calista", "urd")
@@
   methods_evaluated %>% 
-    filter(source %in% c("tool")) %>%
     filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
@@
 
+→
+### Off-the-shelf methods
+For baseline performance, we added several "off-the-shelf" TI methods which can be run using a few lines of code in R.
+
+* **Component 1:** This method returns the first component of a PCA dimensionality reduction as a linear trajectory.
+This method is especially relevant as it has been used in a few studies already [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
+* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two dimensional PCA, and uses this angle as a pseudotime for generating a cyclical trajectory.
+* **MST:** This method performs PCA dimensionality reduction, followed by clustering using the R mclust package, after which the clusters are connected using a minimum spanning tree.
+The trees are orthogonally projected to the nearest segment of the tree.
+This baseline is highly relevant as many methods follow the same methodology: dimensionality reduction, clustering, topology inference, project cells to topology.
+
+←
 ## Trajectory types
@@
 Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
-The structure of these module networks is available at[e] `r print_url("https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks")`.←
+The structure of these module networks is available at https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks.←
 
@@
 
-The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using generalized additive models.
-As smoothing bases, we used a thin-plate regression spline [@wood_thinplateregression_2003] , with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the gam as implemented in the R mgcv package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
+The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using shape constrained additive models, with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the `scam` function as implemented in the R scam package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
 
-To classify the time complexity of each method, we predicted the running time at 10.000 features with increasing number of cells from 100 to 100.000, with steps of 100.
-A method was classified as constant / low slope if on average an addition of 100 cells would increase the running time with less than 0.1 seconds.
-To classify as sublinear, linear, quadratic, or superquadratic, we scaled the times between 0 and 1, and classified the method as sublinear if the running time was on average 5% above the diagonal, superlinear if on average 5% under the diagonal, and linear otherwise.
+To classify the time complexity of each method with respect to the number of cells, we predicted the running time at 10.000 features with increasing number of cells from 100 to 100.000, with steps of 100.
+We trained a generalised linear model with the following function: `y ~ log(x) + sqrt(x) + x + I(x^2) + I(x^3)`.
+The time complexity of a method was then classified as follows:
+$$
+\begin{cases}
+  \text{superquadratic} & \text{if } w_{x^3} > 0.25,\\
+  \text{quadratic} & \text{if } w_{x^2} > 0.25,\\
+  \text{linear} & \text{if } w_x > 0.25,\\
+  \text{sublinear} & \text{if } w_{\text{log}(x)} > 0.25 \text{ or } w_{\text{sqrt}(x)} > 0.25,\\
+  \text{case with highest weight} & \text{else.}
+\end{cases}
+$$
+
+This process was repeated for classifying the time complexity with respect to the number of features, and the memory complexity both with respect to the number of cells and features.
 
@@
 For calculating the final score, we weighed each of the six categories equally.
-
 # Acknowledgements
@@
 [c]TODO: Will be added as sup fig
-[d]Update when final
-[e]the following link renders as verbatim code!

### 2018-11-14 README.Rmd
update readme
update readmes

@@
 
-(Note that this preprint is currently outdated, but cannot be updated because the paper is under revision)
+(Note that this preprint is currently very outdated, see below for more recent results)
 
@@
 ```{r, echo = FALSE, message=FALSE}
-set.seed(1)
+set.seed(3)
 
@@
 
-## Benchmarking your own method
+## Installation
 
-Explanation coming soon. Feel free to make an issue or send us an e-mail if you want your method to be included.
+dynbenchmark has been tested using R version 3.5.1 on Linux.
+While running the methods also works on on Windows and Mac (see [dyno](https://github.com/dynverse/dyno)), running the benchmark is currently not supported on these operating system, given that a lot of commands are linux specific.
+
+In R, you can install the dependencies of dynbenchmark from github using:
+
+``` r
+# install.packages("devtools")
+devtools::install_github("dynverse/dynbenchmark")
+```
+
+This will install several other "dynverse" packages. Depending on the number of R packages already installed, this installation should take approximately 5 to 30 minutes.
+
+On Linux, you will need to install udunits and ImageMagick:
+
+* Debian / Ubuntu / Linux Mint: `sudo apt-get install libudunits2-dev imagemagick`
+* Fedora / CentOS / RHEL: `sudo dnf install udunits2-devel ImageMagick-c++-devel`
+
+[Docker](https://docs.docker.com/install) has to be installed to run TI methods.
+See https://docs.docker.com/install for instructions.
+For windows 10 you can install [Docker CE](https://store.docker.com/editions/community/docker-ce-desktop-windows), older windows installations require the [Docker toolbox](https://docs.docker.com/toolbox/overview/).
+
+You can test whether docker is correctly installed by running:
+
+```{r, message = TRUE, echo = TRUE}
+dynwrap::test_docker_installation(detailed = TRUE)
+```
+
+This command will give helpful tips if some parts of the installation are missing.

### 2018-11-14 scripts/07-stability/0-generate_data.R
add documentation to a lot of scripts + add readmes + rerender readmes

@@ -1,3 +1,7 @@
+#' Generate the subsampled datasets
+#' First choose a set of datasets to be used
+#' Then subsample each dataset
+
 library(dynbenchmark)
 library(tidyverse)
 

### 2018-11-14 scripts/09-guidelines/01-aggregate.R

@@ -1,4 +1,4 @@
-#' Aggregation of the different bnchmarking results for user guidelines
+#' Aggregation of the different bnchmarking results for dynguidelines
 #' One big methods tibble is created, which contains the information for each method in each row
 #' This tibble is nested if a particular experiment generates multiple data points for a method
 #' Otherwise, the result is given by using {experiment_id}_{metric_id}

### 2018-11-14 scripts/10-benchmark_interpretation/02-dataset_sources.R

@@ -1,3 +1,5 @@
+#' Similarity of the results across dataset sources
+
 library(tidyverse)
 library(dynbenchmark)
 

### 2018-11-14 manuscript/supplementary.Rmd
remove toc files

@@
 output: 
-  dynbenchmark::pdf_manuscript:
+  dynbenchmark::pdf_supplementary_note:
       render_change: FALSE
@@
 ```{r}
-# dynbenchmark::knit_nest("../results/02-metrics/")
+dynbenchmark::knit_nest("../results/02-metrics/")
 ```

### 2018-11-14 README.Rmd

@@
 
-* **Gold standard single-cell datasets**, both real and synthetic, used to evaluated the trajectory inference methods [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1211533.svg)](https://doi.org/10.5281/zenodo.1211533)
+* **Gold standard single-cell datasets**, both real and synthetic, used to evaluated the trajectory inference methods [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1443566.svg)](https://doi.org/10.5281/zenodo.1443566)
 

### 2018-11-14 manuscript/paper.Rmd

@@
 
-Commit: `r print_url(paste0("www.github.com/dynverse/dynbenchmark/tree/", git2r::revparse_single(".","HEAD")$sha))`
+Commit: `r print_url(paste0("https://www.github.com/dynverse/dynbenchmark/tree/", git2r::revparse_single(".","HEAD")$sha))`
 ```{r}
@@
 
-Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+Recent technological advances allow →unbiased modelling← of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
 Since 2014, at least →`r n_tools`← tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
@@
 
-In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←.
+In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and usability←.
 Our results highlight the complementarity of existing tools, and that there is no generally best performing method.
-Instead, the choice of method →mostly depends on the dataset characteristics and the trajectory topology present.
+Instead, the choice of method →mostly depends on the dataset dimensions and the trajectory topology present in the dataset.
 To assist users in selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow identifying more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected graphs [@wolf_graphabstractionreconciles_2017].
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017].
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable [@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018; @schaum_singlecelltranscriptomics20_2018] underline the urgency for accurate, scalable [@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.
 
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→Indeed, in several repositories listing single-cell tools,  such as `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies they can detect←.
+→Indeed, in several repositories listing single-cell tools,  such as `r print_url("http://www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("http://www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies they can detect←.
 Early TI methods typically fixed the topology algorithmically (e.g.
 linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating trajectories [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016].
-These methods therefore mainly focused on correctly ordering the cells along the fixed topology.
-→Most recent methods also infer the topology [@street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←
+These methods therefore mainly focus on correctly ordering the cells along the fixed topology.
+→Most recent methods also infer the topology [@qiu_reversedgraphembedding_2017; @street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←
 
@@
   caption_text = glue::glue("a) A schematic overview of our evaluation pipeline.
-b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predicted trajectories from TI methods.
-c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity.
+b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predictions of TI methods.
+c) Trajectories are automatically classified into one of seven trajectory types, with increasing complexity.
 d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory.
@@
 ```
-
 →
@@
 <!-- Method characterisation -->
-In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")` and `r ref("fig", "results_summary")`).
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
@@
 The main difference between trajectory inference methods is whether the method fixes the topology, and if it does not, what kind of topologies it can detect.
-We defined nine possible types of topologies, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
+We defined seven possible types of topologies, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
 Most methods either focus on inferring linear trajectories, or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (`r ref("fig", "results_summary", "a")`).
@@
 We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
-genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the quality of the tool in terms of software, documentation and the manuscript.
+genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the usabiility of the tool in terms of software, documentation and the manuscript.
 Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (`r ref("fig", "results_summary", "b")`).
-We will discuss each evaluation criterion in more detail (`r ref("fig", "results_summary")` and `r ref("sfig", "results")`), after which we conclude with guidelines for method users and future perspectives for method developers.
+We will discuss each evaluation criterion in more detail (`r ref("fig", "results_detailed")` and `r ref("sfig", "results")`), after which we conclude with guidelines for method users and future perspectives for method developers.
 ```{r}
@@
 The methods are grouped vertically based on the most complex trajectory type they can infer.
-(b) The overall results of the evaluation on four criteria: benchmarking using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples, and quality of the implementation.",
+(b) The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples, and quality of the implementation.
+Methods which errored on more than 50% of the datasets were removed from this figure, and are shown in {ref('sfig', 'results')} instead.",
   width = 15,
@@
   ref_id = "results_detailed", 
-  caption_main = glue::glue("Detailed results of the four main evaluation criteria: benchmarking, scalability, stability and quality control."), 
+  caption_main = glue::glue("Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability."), 
   caption_text = glue::glue("(a) The names of the methods, ordered as in {ref('fig', 'results_summary')}.
@@
 The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
-(c) Predicted execution times for varying numbers of cells and features (# cells × # features).
+(c) Predicted execution times for varying numbers of cells and features (\\# cells × \\# features).
 Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
 (d) Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method.
-(e) Usability scores of the tool and corresponding manuscript, grouped per category."),
+(e) Usability scores of the tool and corresponding manuscript, grouped per category.
+Off-the-shelf methods were directly implemented in R and thus do not have a usability score."),
   width = 15,
@@
 The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
-(d) Predicted execution times and memory usage for varying numbers of cells and features (# cells × # features).
+(d) Predicted execution times and memory usage for varying numbers of cells and features (\\# cells × \\# features).
 Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
@@
 (b) The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types.
-For this figure, we did not include any methods requiring prior information.",
+For this figure, we did not include any methods requiring strong prior information (cell grouping or time course).",
   width = 14,
@@
 We modelled the running time and memory usage using a Shape Constrained Additive Model [@pya_shapeconstrainedadditive_2015] (`r ref("sfig", "scaling", "a")`).
-As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method, depending on to what extent the execution of a method succeeded during the scalability experiments.[a][b][c]
+As a control, we compared the predicted time (and memory) with the actual time (resp.
+memory) on all benchmarking datasets, and found that these were highly correlated overall (> 0.9, `r ref("sfig", "predcors")`), and moderately to highly correlated (0.5 - 0.9) for almost every method, depending on to what extent the execution of a method succeeded during the scalability experiments (`r ref("sfig", "results", "a")`).
 
-We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with 10 thousand cells and 10 thousand features (`r ref("fig", "results_detailed", "c")`), the size of a typical droplet based single-cell omics dataset [@svensson_exponentialscalingsinglecell_2018].
-Running times increased further with increasing number of cells, with only a handful of graph/tree methods running within a day on a million of cells (PAGA, PAGA Tree, Monocle DDRTree, Stemnet and GrandPrix).
+We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with ten thousand cells and ten thousand features (`r ref("fig", "results_detailed", "c")`), around the size of a typical droplet-based single-cell dataset [@svensson_exponentialscalingsinglecell_2018].
+Running times increased further with increasing number of cells, with only a handful of graph/tree methods completing within a day on a million of cells (PAGA, PAGA Tree, Monocle DDRTree, Stemnet and GrandPrix).
 Some methods, such as Monocle DDRTree and GrandPrix, also suffered from unsatisfactory running times when given a high number of features.
@@
 Most methods had reasonable memory requirements for modern workstations or computer clusters (≤12GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or high number of features.
-Notably, the memory requirements were very high for several methods, such as RaceID/StemID, pCreode, Monocle DDRTree and SLICE, as the number of cells or features increases.
+Notably, the memory requirements were very high for several methods on datasets with high number of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).
 
-Altogether, the scalability analysis indicated that the dimensions of the data should be important factor in the choice of method and that method development should pay more attention to maintaining reasonable running times and memory usage.
+Altogether, the scalability analysis indicated that the dimensions of the data is an important factor in the choice of method and that method development should pay more attention to maintaining reasonable running times and memory usage.
 
@@
 ```
+
+```{r}
+add_sfig(
+  fig_path = result_file("predcors.pdf", experiment_id = "10-benchmark_interpretation"),
+  ref_id = "predcors", 
+  caption_main = glue::glue("Agreement between actual values and predictions for execution times and memory usage"), 
+  caption_text = glue::glue(""),
+  width = 10,
+  height = 10
+)
+```
 ## Stability
@@
 
-While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how easy it is for a biological user to use [@_doesyourcode_2018].
+While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how easy it is for a biological user to use [@taschuk_ten_2017].
 We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scoresheet")`).
@@
 
-We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and basic code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`).
+We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (`r ref("fig", "results_detailed", "d")` and `r ref("sfig", "qc_overview")`).
 While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("sfig", "qc_overview", " right")`) and we believe that these should receive extra attention from developers.
 Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].
-Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration.
+Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration of future methods.
+← We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary", "b")`).
 
@@
 ```
-
-←
-
-We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary")`).
-
 # Discussion
@@
 Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("http://guidelines.dynverse.org")`).
-We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should currently still be primarily driven by the anticipated trajectory topology in the data.
+We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data.
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
@@
 Several other factors will also impact the choice of methods, such as the dimensions of the dataset, and the prior information which is available.
-These factors and several others can all be dynamically explored in our interactive app (`r print_url("https://dynverse.org")`).
+These factors and several others can all be dynamically explored in our interactive app (`r print_url("https://guidelines.dynverse.org")`).
 This app can also be used to query the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←
@@
 Methods to the right are ranked according to their performance on a particular (set of) trajectory type.
-Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information.",
+Further to the right are shown the accuracy (+: scaled performance ≥ 0.9, ±: > 0.6), usability scores (+: ≥ 0.9, ± ≥ 0.6), estimated running times and required prior information.",
   width = 13,
@@
   ref_id = "example_predictions", 
-  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
-  caption_text = "Trajectories inferred by each method were projected to a common dimensionality reduction.
+  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.", 
+  caption_text = glue::glue("Trajectories inferred by each method were projected to a common dimensionality reduction using multi-dimensional scaling.
+For each dataset, we also calculated a \"consensus\" prediction, by calculating the {label_metric('correlation')} between each pair of models, and picking the model with the highest score on average.
 (a) The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
@@
 (c) A synthetic dataset generated by dyntoy, containing four disconnected trajectories.
-(d) A synthetic dataset generated by dyngen, containing a cyclic trajectory..",
+(d) A synthetic dataset generated by dyngen, containing a cyclic trajectory."),
   width = 14,
@@
 Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution, or provide a unique insight in specific use cases.
-Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns a measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
+→This is also supported by our analysis of method complementarity.← Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns a measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
 We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.
@@
 
-The main analysis repository is available at `r print_url("benchmark.dynverse.org")` and is divided into several experiments.
+The main analysis repository is available at `r print_url("https://benchmark.dynverse.org")` and is divided into several experiments.
 Each experiment has its own set of scripts and results, each accompanied by an illustrated readme which can be easily browsed and explored on the github website.
 
-The analysis scripts call several other R packages, of which an overview is available at `r print_url("dynverse.org")`.
+The analysis scripts call several other R packages, of which an overview is available at `r print_url("http://dynverse.org")`.
 These packages include **dynwrap**, used to wrap the output of methods into the common trajectory model, **dyneval**, which contains the evaluation metrics, **dynguidelines**, the guidelines app, and **dynplot** for plotting trajectories.
@@
 We gathered a list of →`r n_tools`← trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: →`r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018]←.
-We welcome any contributions by creating an issue at →`r print_url("methods.dynverse.org")`←.
+We welcome any contributions by creating an issue at →`r print_url("https://methods.dynverse.org")`←.
 
 <!-- Inclusion criterion -->
-Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`.
+Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({R.utils::decapitalize(non_inclusion_reasons$footnote)}) {non_inclusion_reasons$long}"), ", ")`.
 →The discussions on why these methods were excluded can be found at `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"won\'t wrap"')`.
-In the end, we included  `r n_tools_evaluated` tools in the evaluation, constituting `r n_methods_evaluated` methods.←
+In the end, we included `r n_methods_evaluated` methods in the evaluation.←
 
@@
 For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
-This script reads in the input data , runs the method, and outputs the files required to construct a trajectory.
-We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).
+This script reads in the input data, runs the method, and outputs the files required to construct a trajectory.
+We also created a script to generate an example dataset, which is used to automatically test every method with continuous integration on travis-ci.org (`r print_url("https://travis-ci.org/dynverse")`).
 
@@
 (iii) →The regions of delayed commitment, which defines connections between three or more milestones.
-These must be explicitly defined in the trajectory model and per region, one milestone must be directly connected to all other milestones of the region.←
+These must be explicitly defined in the trajectory model and per region one milestone must be directly connected to all other milestones of the region.←
 
@@
 * **Type 4, end state probability:** `r conversion("end_state_probabilities")`.
-The method returned a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state.
+The method returned a pseudotime and for each cell and end state a probability →($\textrm{Pr}$)← for how likely a cell will end up in a certain end state.
 →This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$), with regions of delayed commitment between all milestones.
@@
 * **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. →This was then processed as a type 1, direct, method.←
-* **mfa:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed →as a type 4, end state probabilities, method.←
+* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed →as a type 4, end state probabilities, method.←
 →
@@
 
-More information on how each method was wrapped can be found within the comments of each wrapper script, available at `r print_url("https://github.com/dynverse/dynmethods")`.←
+More information on how each method was wrapped can be found within the comments of each wrapper script, listed at `r print_url("https://methods.dynverse.org")`.←
 
@@
 Finally, a cycle is a connected graph in which no node has a degree larger than 3 but contains one cycle.
-In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.←
+In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.
+
+For simplicity, we merged the bifurcation and convergence trajectory type, and the acyclic graph and connected graph trajectory type in the main figures of the paper.←
 
@@
 We gathered real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`).
-The scripts to download and process these datasets are available on our repository (`r print_url("https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets")`).
+The scripts to download and process these datasets are available on our repository (`r print_url("https://benchmark.dynverse.org/tree/master/scripts/01-datasets")`).
 Whenever possible, we preferred to start from the raw counts data.
@@
 →Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
-In the end, we included `r nrow(list_datasets("real"))` for this evaluation.←
+In the end, we included `r nrow(list_datasets("real"))` datasets for this evaluation.←
 
@@
 The known connections between these groups were used to construct the milestone network.
-If there was time data available, we used this as the length of the edge, otherwise we set all the lengths equal to one.
+If there was biological or experimental time data available, we used this as the length of the edge, otherwise we set all the lengths equal to one.
 ←
@@
 To do this, we extracted several parameters from all real datasets.
-We calculated the number of differentially expressed features within a trajectory using a wilcoxon test between every pair of cell groups.
+We calculated the number of differentially expressed features within a trajectory using a two-way Mann–Whitney U test between every pair of cell groups.
 These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5% of cells, and had at least a fold-change of 2.
@@
 Not every real dataset was selected to serve as a reference for a synthetic dataset.
-Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (implemented in the R cluster package) to generate synthetic data.
+Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (with $k = 10$, implemented in the R cluster package) to generate synthetic data.
 ←
@@
 After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$.
-We added regions of delayed commitment to a divergence by in half of the cases.
+We added regions of delayed commitment to a divergence in a random half of the cases.
 We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, based on the length of the edges in the milestone network.
@@
 Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (`r ref("snote", "metrics")`).
-Based on this analysis, we chose 4 metrics for the evaluation, each assessing a different aspect of the trajectory: (i) the `r label_metric("him")` measures the topological similarity, (ii) the   `r label_metric("F1_branches")`compares the branch assignment, (iii) the  `r label_metric("correlation")` assesses the similarity in pairwise cell-cell distances and thus the cellular positions, and (iv) the  `r label_metric("featureimp_wcor")` looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
+Based on this analysis, we chose 4 metrics for the evaluation, each assessing a different aspect of the trajectory: (i) the `r label_metric("him")` measures the topological similarity, (ii) the   `r label_metric("F1_branches")` compares the branch assignment, (iii) the  `r label_metric("correlation")` assesses the similarity in pairwise cell-cell distances and thus the cellular positions, and (iv) the  `r label_metric("featureimp_wcor")` looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
 
@@
 ### The correlation between important features
-The `r label_metric("featureimp_wcor")` assesses whether similar dynamically differentially expressed features are found between the prediction and the model.
+The `r label_metric("featureimp_wcor")` assesses whether the same trajectory differentially expressed features are found between the prediction and the model.
 To calculate this metric, we used Random Forest regression (implemented in the R ranger package), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
@@
 To make sure easy and difficult datasets have equal influence on the final score, we first normalised the scores on each dataset across the different methods.
-We shifted and scaled the scores to $\sigma = 1$ and $\mu = 0$, and then applied the unit probability density function on these values to get the scores back into the $[0, 1]$ range.
+We shifted and scaled the scores to $\sigma = 1$ and $\mu = 0$, and then applied the unit probability density function of a normal distribution on these values to get the scores back into the $[0, 1]$ range.
 
@@
 We iteratively added new methods until all methods were selected.
-For this analysis, we did not include any methods which require prior information, and only included methods which can detect at least one of the trajectory types present in the datasets.
+For this analysis, we did not include any methods which require any strong prior information, and only included methods which can detect the trajectory types present in at least one of the datasets.
 ## Scalability
@@
 To estimate the maximal memory usage, we used the `max_vmem` value from the `qacct` command provided by the gridengine cluster.
-We acknowledge however that these values are very noisy, and the averages provided in this study are therefore only rough estimates.
+We acknowledge however that these memory estimates are very noisy, and the averages provided in this study are therefore only rough estimates.
 
-The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using shape constrained additive models, with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the `scam` function as implemented in the R scam package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
+The relationship between the dimensions of a dataset and the running time or maximal memory usage was modelled using shape constrained additive models [@pya_shapeconstrainedadditive_2015], with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the `scam` function as implemented in the R scam package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
 
@@
 In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different.
-However, running the method multiple times on the same input data would not be the ideal approach to assess the stability of a method, given that a lot of tools artificially are artificially deterministic by resetting the pseudorandom number generator (for example, using the set.seed function in R).
+However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R`or the `random.seed` function in numpy).
 To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
@@
 
-We created a transparent scoring scheme to quantify the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`).
+We created a transparent scoring scheme to quantify the usability of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_scoresheet")`).
 The goal of this quality control is in the first place to stimulate the improvement of current methods, and the development of user and developer friendly new methods.
@@
 For calculating the final score, we weighed each of the six categories equally.
+
+→
+## Guidelines
+For each set of outcomes in the guidelines figure, we selected one to four methods, by first filtering the methods on those which can detect all required trajectory types, and ordering the methods according to their average accuracy score on datasets containing these trajectory types (aggregated according to the scheme presented in the accuracy section).
+
+We use the same approach for selecting the best set of methods in the guidelines app (`r print_url("https://guidelines.dynverse.org")`), developed using the R shiny package.
+This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available, and the prefered execution environment (using the dynmethods package or standalone).
+←
 # Acknowledgements
-We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers. This study was supported by Fond Wetenschappelijk Onderzoek (W.S. and R.C.).
+We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers. This study was supported by Fond Wetenschappelijk Onderzoek (R.C. and W.S.).
 # Author contributions
-W.S., R.C., H.T.
+R.C., W.S., H.T.
 and Y.S.
 designed the study.
-W.S.
-and R.C.
+R.C.
+and W.S.
 performed the experiments and analysed the data.
@@
 implemented software packages.
-W.S., R.C., Y.S.
+R.C., W.S., Y.S.
 and H.T.
@@
 -->
-
-[a](Not shown), or include in Supp Fig 4?
-[b]Yeah we don't include this yet, should be in sup fig 4
-[c]TODO: Will be added as sup fig

### 2018-11-14 manuscript/paper.Rmd
update paper

@@
 
-Recent technological advances allow →unbiased modelling← of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+Recent technological advances allow →the unbiased modelling← of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
 Since 2014, at least →`r n_tools`← tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
@@
 One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014].
-Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory by their overall similarity in -omics datasets, mainly transcriptomics [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
+Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory by their overall similarity in -omics datasets [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018].
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow identifying more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected graphs [@wolf_graphabstractionreconciles_2017].
@@
   "tools",
-paste0("→Overview of available trajectory inference tools, and whether they were included in this study.←"),
+paste0("Overview of available trajectory inference tools, and whether they were included in this study."),
   ""
@@
 <!-- Method characterisation -->
-In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")` and `r ref("fig", "results_summary")`).
+In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`).
 In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`).
 In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones.
-Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`).
-When required, we also provided prior information to the method, which can range from weak priors which are easier to acquire, such as the start cells, to strong priors, such as a known grouping of cells, which are much harder to know a priori and which can potentially introduce a large bias into the analysis (`r ref("fig", "results_summary", "a")`).
+Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups (`r ref("sfig", "overview_wrapping")`), and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")`).
+When strictly required, we also provided prior information to the method.
+These different priors can range from weak priors which are relatively easy to acquire, such as a start cell, to strong priors, such as a known grouping of cells, which are much harder to know a priori and which can potentially introduce a large bias into the analysis (`r ref("fig", "results_summary", "a")`).
 
 <!-- Topology characterisation -->
-The main difference between trajectory inference methods is whether the method fixes the topology, and if it does not, what kind of topologies it can detect.
+The largest difference between TI methods is whether a method fixes the topology, and if it does not, what kind of topologies it can detect.
 We defined seven possible types of topologies, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
@@
 We evaluated each method on four core aspects: (i) accuracy of a prediction, given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) scalability with respect to the number of cells and features (e.g.
-genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the usabiility of the tool in terms of software, documentation and the manuscript.
+genes) (iii) stability of the predictions after subsampling the datasets, and (iv) the usability of the tool in terms of software, documentation and the manuscript.
 Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (`r ref("fig", "results_summary", "b")`).
@@
 Methods which errored on more than 50% of the datasets were removed from this figure, and are shown in {ref('sfig', 'results')} instead.",
-  width = 15,
+  width = 13,
   height = 18
@@
 Off-the-shelf methods were directly implemented in R and thus do not have a usability score."),
-  width = 15,
+  width = 13,
   height = 18
@@
 These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).
-Some real datasets were classified as "gold standard" if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing [@tian_scrnaseqmixologybetter_2018].
+Real datasets were classified as "gold standard" if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing [@tian_scrnaseqmixologybetter_2018].
 All other real datasets were classified as "silver standard".
@@
 
-<!-- Benchmark overview -->
+<!-- Accuracy overview -->
 We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset (`r ref("sfig", "benchmark_interpretation", "a")`).
@@
 For this figure, we did not include any methods requiring strong prior information (cell grouping or time course).",
-  width = 14,
+  width = 13,
   height = 6
@@
 As a control, we compared the predicted time (and memory) with the actual time (resp.
-memory) on all benchmarking datasets, and found that these were highly correlated overall (> 0.9, `r ref("sfig", "predcors")`), and moderately to highly correlated (0.5 - 0.9) for almost every method, depending on to what extent the execution of a method succeeded during the scalability experiments (`r ref("sfig", "results", "a")`).
+memory) on all benchmarking datasets, and found that these were highly correlated overall (> 0.9, `r ref("sfig", "predcors")`), and moderately to highly correlated (0.5 - 0.9) for almost every method, depending on to what extent the execution of a method succeeded during the scalability experiments (`r ref("fig", "results_detailed", "c")` and `r ref("sfig", "results", "a")`).
 
@@
   ref_id = "predcors", 
-  caption_main = glue::glue("Agreement between actual values and predictions for execution times and memory usage"), 
+  caption_main = glue::glue("Agreement between actual values and predictions for execution times and memory usage."), 
   caption_text = glue::glue(""),
@@
 We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_scoresheet")`).
-Important to note here is that we did not include a substantial set of tools in our analyses, because of issues with installation, code availability and executability on a freely available platform (which excludes MATLAB), all discussed at our repository (`r print_url('http://methods.dynverse.org/issues?q=label:\"won\'t+wrap\"')`).
+Important to note here is that there is a selection bias in the tools selected for this analysis, as we did not include a substantial set of tools due to  issues with installation, code availability and executability on a freely available platform (which excludes MATLAB).
+The reason for not including certain tools are all discussed on our repository (`r print_url('https://github.com/dynverse/dynmethods/issues?q=label:\"won\'t+wrap\"')`).
 Installation issues seem to be quite general in bioinformatics [@mangul_comprehensiveanalysisusability_2018], and the trajectory inference field is no exception.
@@
   "qc_scoresheet",
-paste0("→Scoring sheet for assessing usability of trajectory inference methods.
-Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.←"),
+paste0("Scoring sheet for assessing usability of trajectory inference methods.
+Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."),
   ""
@@
 In this study, we presented a large scale evaluation of the performance of  `r n_methods_evaluated`  TI methods.
-By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the performance of the methods on more than two-hundred datasets.
+By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the →accuracy← of the methods on more than two-hundred datasets.
 We also assessed several other important quality measures, such as the quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on small variations of the datasets.←
@@
 <!-- Further considerations -->
-When choosing a method, it is important to take two further points into account.
+→When inferring a trajectory on a data set of interest, ← it is important to take two further points into account.
 First, it is critical that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
@@
 In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.←
-
-```{r}
-add_fig(
-  fig_path = result_file("example_predictions.pdf", experiment_id = "11-example_predictions"),
-  ref_id = "example_predictions", 
-  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.", 
-  caption_text = glue::glue("Trajectories inferred by each method were projected to a common dimensionality reduction using multi-dimensional scaling.
-For each dataset, we also calculated a \"consensus\" prediction, by calculating the {label_metric('correlation')} between each pair of models, and picking the model with the highest score on average.
-(a) The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
-(b) The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts.
-(c) A synthetic dataset generated by dyntoy, containing four disconnected trajectories.
-(d) A synthetic dataset generated by dyngen, containing a cyclic trajectory."),
-  width = 14,
-  height = 8
-)
-```
 
@@
 
+```{r}
+add_fig(
+  fig_path = result_file("example_predictions.pdf", experiment_id = "11-example_predictions"),
+  ref_id = "example_predictions", 
+  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.", 
+  caption_text = glue::glue("Trajectories inferred by each method were projected to a common dimensionality reduction using multi-dimensional scaling.
+For each dataset, we also calculated a \"consensus\" prediction, by calculating the {label_metric('correlation')} between each pair of models, and picking the model with the highest score on average.
+(a) The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
+(b) The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts.
+(c) A synthetic dataset generated by dyntoy, containing four disconnected trajectories.
+(d) A synthetic dataset generated by dyngen, containing a cyclic trajectory."),
+  width = 14,
+  height = 8
+)
+```
 ________________
@@
 add_sfig(
-  raw_file("overview_wrapping_v2.svg", "03-methods"),
+  raw_file("overview_wrapping_v3.svg", "03-methods"),
   "overview_wrapping", 
@@
 Cells were positioned onto the milestones they are assigned to→, with milestone percentage equal to 1.←
-* **Type 6, project onto nearest branch:** `r conversion("dimred_projection")`.
+* **Type 6, orthogonal projection:** `r conversion("dimred_projection")`.
 The method returned a milestone network, and a dimensionality reduction of the cells and milestones.
@@
 
+→
+* **CALISTA:** We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest.
+The cluster probabilities of the two selected milestones were then used as milestone percentages.
+This was then processed as a type 1, direct, method.
+←
 * **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
@@
 →
-* **CALISTA:** We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest.
-The cluster probabilities of the two selected milestones were then used as milestone percentages.
-This was then processed as a type 1, direct, method.
 * **URD:** We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.
@@
 This baseline is highly relevant as many methods follow the same methodology: dimensionality reduction, clustering, topology inference, project cells to topology.
-
 ←
@@
 The `r label_metric("featureimp_wcor")` assesses whether the same trajectory differentially expressed features are found between the prediction and the model.
-To calculate this metric, we used Random Forest regression (implemented in the R ranger package), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
+To calculate this metric, we used Random Forest regression (implemented in the R ranger package [@wright_rangerfastimplementation_2017]), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
 We then extract feature importance values for each feature, and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset as to give more weight to large differences.
@@
 Pre- and postprocessing steps specific to a method, such as dimensionality reduction and gene filtering, were included in the time.
-To estimate the maximal memory usage, we used the `max_vmem` value from the `qacct` command provided by the gridengine cluster.
+To estimate the maximal memory usage, we used the `max_vmem` value from the `qacct` command provided by a gridengine cluster.
 We acknowledge however that these memory estimates are very noisy, and the averages provided in this study are therefore only rough estimates.
@@
 In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different.
-However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R`or the `random.seed` function in numpy).
+However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R or the `random.seed` function in numpy).
 To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25% of the datasets accounting for 15% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
@@
 # Acknowledgements
-We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers. This study was supported by Fond Wetenschappelijk Onderzoek (R.C. and W.S.).
+We would like to thank the original authors of the methods for their feedback and improvements on the method wrappers.
+This study was supported by the Fonds Wetenschappelijk Onderzoek (FWO, R.C.
+and W.S.) and BOF (Ghent University, H.T).
 # Author contributions

### 2019-01-11 manuscript/paper.Rmd

@@
 
-Recent technological advances allow →the unbiased modelling← of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
-Since 2014, at least →`r n_tools`← tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
+Recent technological advances allow the unbiased modelling of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+Since 2014, at least `r n_tools` tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare.
 As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for users in the field.
 
-In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and usability←.
+In this study, we evaluated a total of `r n_methods_evaluated` methods in terms of cellular ordering, topology, scalability, and usability.
 Our results highlight the complementarity of existing tools, and that there is no generally best performing method.
-Instead, the choice of method →mostly depends on the dataset dimensions and the trajectory topology present in the dataset.
-To assist users in selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
+Instead, the choice of method mostly depends on the dataset dimensions and the trajectory topology present in the dataset.
+To assist users in selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.
+While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases.
 We hope to spearhead such developments, and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.
@@
 A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`).
-→Indeed, in several repositories listing single-cell tools,  such as `r print_url("http://www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("http://www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies they can detect←.
+Indeed, in several repositories listing single-cell tools,  such as `r print_url("http://www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("http://www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.
+While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory, and what type(s) of graph topologies they can detect.
 Early TI methods typically fixed the topology algorithmically (e.g.
@@
 These methods therefore mainly focus on correctly ordering the cells along the fixed topology.
-→Most recent methods also infer the topology [@qiu_reversedgraphembedding_2017; @street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←
+Most recent methods also infer the topology [@qiu_reversedgraphembedding_2017; @street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.
 
@@
 <!-- §3: Problem statement -->
-Given the diversity in TI methods, an important issue to address is a quantitative assessment of their →performance, scalability, robustness and usability←.
+Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance, scalability, robustness and usability.
 Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017; @wolf_graphabstractionreconciles_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking.
@@
   caption_text = glue::glue("a) A schematic overview of our evaluation pipeline.
-b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent →reference trajectories← from the real and synthetic datasets, as well as any predictions of TI methods.
+b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent reference trajectories from the real and synthetic datasets, as well as any predictions of TI methods.
 c) Trajectories are automatically classified into one of seven trajectory types, with increasing complexity.
-d) →We defined four metrics, each assessing the quality of a different aspect of the trajectory.
+d) We defined four metrics, each assessing the quality of a different aspect of the trajectory.
 The Hamming-Ipsen-Mikhailov (HIM) score will assess the similarity between the two topologies, taking into account differences in edge lengths and degree distributions.
@@
 The {label_metric('correlation')} quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances.
-Finally, {label_metric('featureimp_wcor')} quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.←"),
+Finally, {label_metric('featureimp_wcor')} quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory."),
    width = 14,
@@
 ```
-→
+
 # Results
@@
 Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration of future methods.
-← We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary", "b")`).
+We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary", "b")`).
 
@@
 In this study, we presented a large scale evaluation of the performance of  `r n_methods_evaluated`  TI methods.
-By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the →accuracy← of the methods on more than two-hundred datasets.
-We also assessed several other important quality measures, such as the quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on small variations of the datasets.←
+By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the accuracy of the methods on more than two-hundred datasets.
+We also assessed several other important quality measures, such as the quality of the method's implementation, the scalability to hundreds of thousands of cells, and the stability of the output on small variations of the datasets.
 
@@
 For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
-In each of these use cases, →our evaluation suggests a different set of optimal methods, as shown in `r ref("fig", "user_guidelines")`.
+In each of these use cases, our evaluation suggests a different set of optimal methods, as shown in `r ref("fig", "user_guidelines")`.
 Several other factors will also impact the choice of methods, such as the dimensions of the dataset, and the prior information which is available.
 These factors and several others can all be dynamically explored in our interactive app (`r print_url("https://guidelines.dynverse.org")`).
-This app can also be used to query the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←
+This app can also be used to query the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.
 
@@
 <!-- Further considerations -->
-→When inferring a trajectory on a data set of interest, ← it is important to take two further points into account.
+When inferring a trajectory on a data set of interest,  it is important to take two further points into account.
 First, it is critical that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
 This is to make sure the prediction is not biased due to the given parameter setting or the particular algorithms underlying a TI method.
-→The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods.← Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
+The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods.
+Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology.
 When the expected topology is confirmed using such a method, it provides additional evidence to the user.
-When a more complex topology is produced, →this could indicate that the underlying biology is much more complex than anticipated by the user.←
+When a more complex topology is produced, this could indicate that the underlying biology is much more complex than anticipated by the user.
 
@@
 Critical to the broad applicability of TI methods is the standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as finding genes important for the trajectory, network inference [@aibar_scenicsinglecellregulatory_2017] or finding modules of genes [@saelens_comprehensiveevaluationmodule_2018].
-Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods →on datasets containing a linear, tree, cyclic and disconnected graph topology (`r ref("fig", "example_predictions")`).
+Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (`r ref("fig", "example_predictions")`).
 Using our framework, this figure can be recreated using only a couple of lines of R code (`r print_url("https://methods.dynverse.org")`).
 In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information [@manno_rnavelocitysingle_2018], and easier downstream analyses.
-In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.←
+In addition, further discussion within the field is required in order to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.
 
@@
 However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analysing single-cell -omics datasets with complex trajectories.
-Foremost, new methods should focus on improving →the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be  identified using a dimensionality reduction method (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
-← Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell -omics field.
+Foremost, new methods should focus on improving the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be  identified using a dimensionality reduction method (illustrated with some synthetic datasets in `r ref("fig", "example_predictions", "c and d")`).
+Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell -omics field.
 Finally, new tools should be designed to scale well with the increasing number of cells and features.
@@
 Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution, or provide a unique insight in specific use cases.
-→This is also supported by our analysis of method complementarity.← Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns a measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
+This is also supported by our analysis of method complementarity.
+Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns a measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
 We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.
@@
 ## Data and code availability
-→ 
+
 The processed real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1443566](https://doi.org/10.5281/zenodo.1443566)) [@cannoodt_goldstandardsinglecell_].
@@
 These packages include **dynwrap**, used to wrap the output of methods into the common trajectory model, **dyneval**, which contains the evaluation metrics, **dynguidelines**, the guidelines app, and **dynplot** for plotting trajectories.
-←
+
 ## Trajectory inference methods
 <!-- Overview -->
-We gathered a list of →`r n_tools`← trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: →`r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018]←.
-We welcome any contributions by creating an issue at →`r print_url("https://methods.dynverse.org")`←.
+We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online: `r print_url("https://github.com/seandavi/awesome-single-cell")`  [@davis_awesomesinglecelllistsoftware_2018] and  `r print_url("https://github.com/agitter/single-cell-pseudotime")` [@gitter_singlecellpseudotimeoverviewalgorithms_2018].
+We welcome any contributions by creating an issue at `r print_url("https://methods.dynverse.org")`.
 
@@
 Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({R.utils::decapitalize(non_inclusion_reasons$footnote)}) {non_inclusion_reasons$long}"), ", ")`.
-→The discussions on why these methods were excluded can be found at `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"won\'t wrap"')`.
-In the end, we included `r n_methods_evaluated` methods in the evaluation.←
+The discussions on why these methods were excluded can be found at `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"won\'t wrap"')`.
+In the end, we included `r n_methods_evaluated` methods in the evaluation.
 
-→
 ### Method wrappers
@@
 These discussions can be viewed on github: `r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`.
-←
+
 ### Method input
@@
 
-The prior information was extracted from the →reference trajectory← as follows:
+The prior information was extracted from the reference trajectory as follows:
 
@@
 * **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
-→
+
 * **Discrete time course** For each cell a time point from which it was sampled.
@@
 For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
-←
+
 ### Common trajectory model
 Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps.
-→We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`).
+We therefore post-processed the output of each method into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`).
 This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
 (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one.
-(iii) →The regions of delayed commitment, which defines connections between three or more milestones.
-These must be explicitly defined in the trajectory model and per region one milestone must be directly connected to all other milestones of the region.←
+(iii) The regions of delayed commitment, which defines connections between three or more milestones.
+These must be explicitly defined in the trajectory model and per region one milestone must be directly connected to all other milestones of the region.
 
@@
 # create functions to list the methods within a particular conversion category
-conversion <- function(conversion_oi) {
+list_wrapper_type_methods <- function(wrapper_types_oi) {
   methods_evaluated %>% 
-    filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
+    filter(wrapper_type %in% wrapper_types_oi) %>% 
     mutate(name = paste0(name, ifelse(id %in% special_methods, "\\*", ""))) %>% 
@@
 
-* **Type 1, direct:** `r conversion("trajectory")`.
+* **Type 1, direct:** `r list_wrapper_type_methods(c("direct", "branch_trajectory"))`.
 The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell it is given to what extent it belongs to a milestone.
-→In some cases, this← indicates that additional transformations were required for the method, not covered by any of the following output formats.
-→Some methods (`r conversion("branch_trajectory")`) returned a branch network instead of a milestone network, and this network was converted by calculating the line graph of the branch network.←
-* **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`.
+In some cases, this indicates that additional transformations were required for the method, not covered by any of the following output formats.
+Some methods returned a branch network instead of a milestone network, and this network was converted by calculating the line graph of the branch network.
+* **Type 2, linear pseudotime:** `r list_wrapper_type_methods("linear")`.
 The method returned a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
-* **Type 3, cyclical pseudotime:** `r conversion("cyclic_trajectory")`.
+* **Type 3, cyclical pseudotime:** `r list_wrapper_type_methods("cyclic")`.
 The method returned a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones.
-→These milestones were positioned at pseudotime 0, ⅓ and ⅔ .←
-* **Type 4, end state probability:** `r conversion("end_state_probabilities")`.
-The method returned a pseudotime and for each cell and end state a probability →($\textrm{Pr}$)← for how likely a cell will end up in a certain end state.
-→This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$), with regions of delayed commitment between all milestones.
+These milestones were positioned at pseudotime 0, ⅓ and ⅔ .
+* **Type 4, end state probability:** `r list_wrapper_type_methods("end_state_prob")`.
+The method returned a pseudotime and for each cell and end state a probability ($\textrm{Pr}$) for how likely a cell will end up in a certain end state.
+This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$), with regions of delayed commitment between all milestones.
 The milestone percentage of a cell to one of the outer milestones was equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$.
-The milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .←
-* **Type 5, cluster assignment:** `r conversion("cluster_graph")`.
+The milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .
+* **Type 5, cluster assignment:** `r list_wrapper_type_methods("cluster_assignment")`.
 The method returned a milestone network, and an assignment of each cell to a specific milestone.
-Cells were positioned onto the milestones they are assigned to→, with milestone percentage equal to 1.←
-* **Type 6, orthogonal projection:** `r conversion("dimred_projection")`.
+Cells were positioned onto the milestones they are assigned to, with milestone percentage equal to 1.
+* **Type 6, orthogonal projection:** `r list_wrapper_type_methods("orth_proj")`.
 The method returned a milestone network, and a dimensionality reduction of the cells and milestones.
 The cells were projected to the closest nearest segment, thus determining the cells' position along the milestone network.
-→If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell.
-For these methods, we usually wrote two wrappers, one which included the projection, and one without.←
-* **Type 7, cell graph:** `r conversion("cell_graph")`. The method returned a network of cells, 
+If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell.
+For these methods, we usually wrote two wrappers, one which included the projection, and one without.
+* **Type 7, cell graph:** `r list_wrapper_type_methods("cell_graph")`. The method returned a network of cells, 
 and which cell-cell transitions are part of the 'backbone' structure.
 Backbone cells with degree $\neq$ 2 were regarded as milestones, and all other cells were placed on transitions between the milestones.
-→If methods did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones.
-Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells, and then divided the distance of each pair of cells with the total distance to get the milestone percentages.←
+If methods did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones.
+Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells, and then divided the distance of each pair of cells with the total distance to get the milestone percentages.
 
@@
 
-→
 * **CALISTA:** We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest.
@@
 This was then processed as a type 1, direct, method.
-←
+
 * **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point.
-→This was then processed as a type 1, direct,  method.←
+This was then processed as a type 1, direct,  method.
 * **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.
-→This was then processed as a type 7, cell graph, method.←
-* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. →This was then processed as a type 1, direct, method.←
-* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed →as a type 4, end state probabilities, method.←
-→
+This was then processed as a type 7, cell graph, method.
+* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This was then processed as a type 1, direct, method.
+* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
+
 * **URD:** We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.
 
-More information on how each method was wrapped can be found within the comments of each wrapper script, listed at `r print_url("https://methods.dynverse.org")`.←
+More information on how each method was wrapped can be found within the comments of each wrapper script, listed at `r print_url("https://methods.dynverse.org")`.
 
-→
 ### Off-the-shelf methods
@@
 This baseline is highly relevant as many methods follow the same methodology: dimensionality reduction, clustering, topology inference, project cells to topology.
-←
+
 ## Trajectory types
@@
 Finally, a cycle is a connected graph in which no node has a degree larger than 3 but contains one cycle.
-In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with →some exceptions shown in `r ref("fig", "results_summary", "a")`.
+In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with some exceptions shown in `r ref("fig", "results_summary", "a")`.
 
-For simplicity, we merged the bifurcation and convergence trajectory type, and the acyclic graph and connected graph trajectory type in the main figures of the paper.←
+For simplicity, we merged the bifurcation and convergence trajectory type, and the acyclic graph and connected graph trajectory type in the main figures of the paper.
 
@@
 These raw counts were all normalised and filtered using a common pipeline, as discussed later.
-→Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
-In the end, we included `r nrow(list_datasets("real"))` datasets for this evaluation.←
+Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
+In the end, we included `r nrow(list_datasets("real"))` datasets for this evaluation.
 
-→For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network).
+For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network).
 The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample or the time it was sampled) or silver standard otherwise (usually by clustering the expression values).
@@
 If there was biological or experimental time data available, we used this as the length of the edge, otherwise we set all the lengths equal to one.
-←
 
@@
 ## Synthetic datasets
-→
+
 To generate synthetic datasets, we used four different synthetic data simulators:
@@
 Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (with $k = 10$, implemented in the R cluster package) to generate synthetic data.
-←
+
 ### dyngen
@@
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
-→We simulated a total of `r nrow(design_dyngen)` datasets, with `r length(unique(design_dyngen$modulenet_name))` different topologies.←
+We simulated a total of `r nrow(design_dyngen)` datasets, with `r length(unique(design_dyngen$modulenet_name))` different topologies.
 #### Network generation
@@
 
-→To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
+To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
 Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
-The structure of these module networks is available at https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks.←
+The structure of these module networks is available at https://github.com/dynverse/dyngen/tree/master/inst/ext_data/modulenetworks.
 
@@
 For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in.
-→These cells were sampled uniformly across the different steps of the 32 simulations.
-← Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
+These cells were sampled uniformly across the different steps of the 32 simulations.
+Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
 We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$.
@@
 Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
-→Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.←
+Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.
 
@@
 Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
-→We therefore first constructed a backbone expression profile for each branch within the trajectory.
+We therefore first constructed a backbone expression profile for each branch within the trajectory.
 To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.
 We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
-← We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
+We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
 In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
 
-→
 ### dyntoy
@@
 
-Using the `simulate_lineage` function, we simulated the lineage expression, with parameters $\a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$.
+Using the `simulate_lineage` function, we simulated the lineage expression, with parameters $a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$.
 These parameter distributions were chosen very broad so as to make sure both easy and difficult datasets are simulated.
@@
 
-←
 ## Dataset filtering and normalisation
 We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016].
-The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (→which we set to 3←).
+The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (which we set to 3).
 We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.02.
@@
 The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011].
-→Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
+Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
 We therefore defined several possible metrics for comparing trajectories, each investigating different layers.
@@
 We illustrate this metric further in  `r ref("snote", "metrics")`.
-←
+
 ### Correlation between geodesic distances
-→When the position of a cell is the same in both the reference and the prediction, its relative distances to all other cells in the trajectory should also be the same.
+When the position of a cell is the same in both the reference and the prediction, its relative distances to all other cells in the trajectory should also be the same.
 This observation is the basis for the `r label_metric("correlation")` metric.
@@
 Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
-←
+
 ## Method execution
@@
 During the execution of a method on a dataset, if the time limit (>1h) or memory limit (16GB) was exceeded, or an error was produced, a zero score was returned for that execution.
-→
 ## Complementarity
@@
 For the `r label_metric("correlation")` and `r label_metric("F1_branches")`, we only used the intersection between the cells of two datasets, while the intersection of the features was used for the `r label_metric("featureimp_wcor")`.
-←
 ## Usability
@@
 
-→
 ## Guidelines
@@
 This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available, and the prefered execution environment (using the dynmethods package or standalone).
-←
 # Acknowledgements

### 2019-01-29 README.Rmd

@@
 # install.packages("devtools")
-devtools::install_github("dynverse/dynbenchmark")
+devtools::install_github("dynverse/dynbenchmark/package")
 ```
@@
 
-```{r, message = TRUE, echo = TRUE}
+```{r, message = FALSE, echo = TRUE}
 dynwrap::test_docker_installation(detailed = TRUE)

### 2019-01-30 README.Rmd
update ink to overview figure

@@
 
-* **The performance of methods** used for the [results overview figure](`r dynbenchmark::link_to_results("scripts/08-summary/overview.pdf")`) and the [dynguidelines](https://www.github.com/dynverse/dynguidelines) app.
-_Not yet available_, but can be accessed in the results folder.
+* **The performance of methods** used for the [results overview figure](`r dynbenchmark::link_to_results("scripts/08-summary/results_suppfig.pdf")`) and the [dynguidelines](http://guidelines.dynverse.org) app.
 

### 2019-01-30 scripts/01-datasets/README.Rmd
update docs
update link to results in readme
fix relative paths

@@
 
+The results of this experiment are available [here](`r dynbenchmark::link_to_results()`).
+
 ```{r results = 'asis'}
@@
 ```
-
+`
 ```{r results = 'asis'}

### 2019-02-26 README.Rmd
update readme

@@
 
-(Note that this preprint is currently very outdated, see below for more recent results)
+*Note that this preprint is currently **very** outdated. An updated manuscript will be published soon, you can browse below for more recent results*
 
@@
 * [Manuscript](manuscript): Source files for producing the manuscript.
-* [Package](package): An R package with several helper functions.
+* [Package](package): An R package with several helper functions for organizing the benchmark and rendering the manuscript.
 * [Raw](raw): Files generated by hand, such as figures and spreadsheets.
-* [Derived](derived): Intermediate data files produced by the scripts. These files or not git committed.
+* [Derived](derived): Intermediate data files produced by the scripts. These files are not git committed.
 
@@
 See https://docs.docker.com/install for instructions.
-For windows 10 you can install [Docker CE](https://store.docker.com/editions/community/docker-ce-desktop-windows), older windows installations require the [Docker toolbox](https://docs.docker.com/toolbox/overview/).
+For windows 10 you can install [Docker CE](https://store.docker.com/editions/community/docker-ce-desktop-windows), older Windows installations require the [Docker toolbox](https://docs.docker.com/toolbox/overview/).
 
@@
 
-```{r, message = FALSE, echo = TRUE}
+```{r, message = TRUE, echo = TRUE}
 dynwrap::test_docker_installation(detailed = TRUE)

### 2019-03-04 README.Rmd
add singularity to dynbenchmark readme

@@
 
-[Docker](https://docs.docker.com/install) has to be installed to run TI methods.
-See https://docs.docker.com/install for instructions.
+[Docker](https://docs.docker.com/install) or [Singularity](https://www.sylabs.io/guides/3.0/user-guide/) (version ≥ 3.0) has to be installed to run TI methods.
+We suggest docker on Windows and MacOS, while both docker and singularity are fine when running on linux.
+Singularity is strongly recommended when running the method on shared computing clusters.
+
 For windows 10 you can install [Docker CE](https://store.docker.com/editions/community/docker-ce-desktop-windows), older Windows installations require the [Docker toolbox](https://docs.docker.com/toolbox/overview/).
@@
 
-This command will give helpful tips if some parts of the installation are missing.
+Same for singularity:
+
+```{r, message = TRUE, echo = TRUE}
+dynwrap::test_singularity_installation(detailed = TRUE)
+```
+
+These commands will give helpful tips if some parts of the installation are missing.

### 2019-03-26 README.Rmd
update readme

@@
 [![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
-![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <a href = "package/man/figures/logo.svg"><img src="package/man/figures/logo.png" align="right" /></a> [![doi](https://zenodo.org/badge/doi/10.1101/276907.svg)](https://doi.org/10.1101/276907)
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <a href = "package/man/figures/logo.svg"><img src="package/man/figures/logo2.png" align="right" width="125" height="144"/></a> [![doi](https://zenodo.org/badge/doi/10.1101/276907.svg)](https://doi.org/10.1101/276907)
 
@@
 
-*Note that this preprint is currently **very** outdated. An updated manuscript will be published soon, you can browse below for more recent results*
+## Dynverse
+
+Under the hood, dynbenchmark makes use of most dynverse package for running the methods, comparing them to a gold standard, and plotting the output.
+Check out **[dynverse.org](https://dynverse.org)** for an overview!
 

### 2019-03-28 README.Rmd
update readme

@@
 [![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
-![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) <a href = "package/man/figures/logo.svg"><img src="package/man/figures/logo2.png" align="right" width="125" height="144"/></a> [![doi](https://zenodo.org/badge/doi/10.1101/276907.svg)](https://doi.org/10.1101/276907)
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)  [**ℹ️ Tutorials**](https://dynverse.org)  &nbsp; <br><img src="package/man/figures/logo.png" align="right" width="125" height="144" /> [![doi](https://zenodo.org/badge/doi/10.1038/s41587-019-0071-9.svg)](https://doi.org/10.1038/s41587-019-0071-9)
 
@@
 
-> A comparison of single-cell trajectory inference methods: towards more accurate and robust tools  
+> A comparison of single-cell trajectory inference methods 
 <strong> Wouter Saelens\* </strong> <a href='https://orcid.org/0000-0002-7114-6248'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/orcid_logo.svg?sanitize=true' height='16'></a> <a href='https://github.com/zouter'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>,
@@
 <em> Yvan Saeys </em> <a href='https://github.com/saeyslab'><img src='https://github.com/dynverse/dynmethods/raw/master/man/figures/github_logo.png' height='16'></a>  
-[bioRxiv:276907](https://www.biorxiv.org/content/early/2018/03/05/276907) [doi:10.1101/276907](https://doi.org/10.1101/276907) [![altmetric](`r get_altmetric_badge()`)](https://altmetric.com/details/33972849)
+[doi:10.1038/s41587-019-0071-9](https://doi.org/10.1038/s41587-019-0071-9) [![altmetric](`r get_altmetric_badge()`)](https://altmetric.com/details/33972849)
 

### 2019-03-28 README.Rmd
update readme

@@
 [![Build Status](https://api.travis-ci.org/dynverse/dynbenchmark.svg)](https://travis-ci.org/dynverse/dynbenchmark)
-![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)  [**ℹ️ Tutorials**](https://dynverse.org)  &nbsp; <br><img src="package/man/figures/logo.png" align="right" width="125" height="144" /> [![doi](https://zenodo.org/badge/doi/10.1038/s41587-019-0071-9.svg)](https://doi.org/10.1038/s41587-019-0071-9)
+![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg) [![doi](https://zenodo.org/badge/doi/10.1038/s41587-019-0071-9.svg)](https://doi.org/10.1038/s41587-019-0071-9) [**ℹ️ Tutorials**](https://dynverse.org)  &nbsp; <br><img src="package/man/figures/logo.png" align="right" width="125" height="144" />
 

### 2021-10-28 manuscript/paper.Rmd
update manuscript render

@@
 
-*`r format(Sys.time(), '%d %B, %Y')`*
+Published in Nat Biotechnol 37, 547–554 (2019). doi: [10.1038/s41587-019-0071-9](https://doi.org/10.1038/s41587-019-0071-9).
 
-Commit: `r print_url(paste0("https://www.github.com/dynverse/dynbenchmark/tree/", git2r::revparse_single(".","HEAD")$sha))`
 ```{r}
 n_tools <- nrow(tools)
-n_tools_evaluated <- nrow(filter(tools, evaluated))
-n_methods_evaluated <- nrow(filter(methods, evaluated, source == "tool"))
-n_control_methods_evaluated <- nrow(filter(methods, evaluated, source == "control"))
+n_tools_evaluated <- nrow(filter(tools, method_evaluated))
+n_methods_evaluated <- nrow(filter(methods, method_evaluated, method_source == "tool"))
+n_control_methods_evaluated <- nrow(filter(methods, method_evaluated, method_source == "control"))
 n_datasets_real <- nrow(list_datasets("real"))
@@
     filter(wrapper_type %in% wrapper_types_oi) %>% 
-    mutate(name = paste0(name, ifelse(id %in% special_methods, "\\*", ""))) %>% 
+    mutate(name = paste0(method_name, ifelse(method_id %in% special_methods, "\\*", ""))) %>% 
     pull(name) %>% 
